### 2008-01-16 abstract.tex
Initial commit. Basically paper.1 (Broadnets) with minor edits.

When new wireless technologies are deployed and subjected to real usage patterns, unforeseen performance problems inevitably seem to arise, to be fixed only in later generations. Why do these performance issues fail to appear in experimental settings before the technology is deployed? We believe that one of the major reasons behind the discrepancies found between experimental performance evaluations and real-world experience lies in the unrealistic workload patterns typically used in experiments. One of the significant contributions of this work is to rigorously demonstrate that common synthetic traffic models for wireless local-area networks induce drastically distorted performance metrics at every layer of the protocol stack. In order to show this, we present a testable definition of ``sufficient realism'' for traffic models, and develop the theoretical methodology necessary to interpret experimental results using this definition. Finally, we show by example that this distortion can completely invert the relative performance of protocols. The greater overall contribution of this paper, however, is the complete collection of ideas, techniques and analytical tools that will allow the development of more realistic synthetic traffic models in the future.
%
%
%Performance predictions from simulations in wireless networking rarely seem to match the behavior observed once the same technologies are deployed. We believe that one of the major factors hampering researchers' ability to make more reliable forecasts is the inability to generate realistic experimental workloads. To redress this problem, we take a fundamentally new approach to quantifying the realism of wireless traffic models. In this approach, the realism of a model is defined directly in terms of its ability to accurately reproduce the performance characteristics of actual network usage. This direct approach cuts through the Gordian knot of deciding which statistical features of traffic traces are significant. The first major contribution of this work is this new definition of workload realism, together with the analytical and statistical methodology to rigorously assess whether synthetic traffic models meet the definition. The second contribution is the conclusion that commonly used models of wireless traffic distort important metrics for performance evaluation at every layer of the wireless protocol stack. We show by example that this distortion can completely invert the relative performance of protocols. The last and most important contribution is the complete collection of ideas, techniques and analytical tools that will allow the development of more realistic synthetic traffic models in the future.

### 2008-01-16 models.tex
Initial commit. Basically paper.1 (Broadnets) with minor edits.

\begin{tabular}{|c|l|p{5in}|}
\multicolumn{1}{c}{\textbf{Behavior}} &
\multicolumn{1}{c}{\textbf{Model}} &
\multicolumn{1}{c}{\textbf{Description}} \\
\hline
\multirow{3}[13]{*}{Network}
\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
Mapping of flow end-points to wireless nodes taken directly from trace data. \\
\cline{2-3}
\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{M}arginal} &
Use marginal empirical distributions from trace. \\
\cline{2-3}
\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{S}ynthetic} &
Use synthetic distributions \textit{\`a la} Hern\'andez-Campos~\cite{Hernandez06:wlan-traffic}. \\
\cline{2-3}
\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{U}niform} &
Both end-points are uniformly randomly chosen from all of the wireless nodes. \\
\hline
\multirow{2}[11]{*}{Flow}
\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
Each flow has the actual start time, end time, total data sent, and number of packets from the trace. \\
\cline{2-3}
\bigstrut & \multirow{1}{*}{\textbf{P}roportional} &
Byte count taken from trace; sizes and intervals are CBR or VBR, on a per-flow basis. \\
\cline{2-3}
\bigstrut & \multirow{1}{*}{\textbf{U}niform} &
All flows have the same duration, volume, packet rate, and data rate as the trace average. \\
\hline
\multirow{2}[11]{*}{Packet}
\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
The sequence of packet sizes and inter-transmission intervals is taken directly from the trace. \\
\cline{2-3}
\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{S}ample} &
Packets sizes and intervals are randomly drawn from per-flow empirical distributions (VBR). \\
\cline{2-3}
\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{U}niform} &
Packets sizes are uniform and the inter-packet interval is constant (CBR). \\
\hline
\end{tabular}

### 2008-01-16 paper.tex
Initial commit. Basically paper.1 (Broadnets) with minor edits.

\documentclass[conference]{IEEEtran}

\newcommand{\thetitle}{Wireless Traffic: The Failure of CBR Modeling}

\usepackage[pdftex]{graphicx}
\usepackage[labelfont=bf,small]{caption}
\usepackage[font=small,labelfont=bf,position=top,nearskip=0em]{subfig}
\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig}
\usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle}]{hyperref}

\hyphenation{op-tical net-works semi-conduc-tor IEEEtran}
\bibliographystyle{IEEEtran}

\newcommand{\email}[1]{$\left<{\textit{#1}}\right>$}
\newcommand{\caps}[1]{{\small{#1}}}

\title{%
%\vspace{-0.2em}
\thetitle%
}
\author{
{\large{Stefan~Karpinski, Elizabeth~M.~Belding, Kevin~C.~Almeroth}}\\
\textit{\{sgk,ebelding,almeroth\}@cs.ucsb.edu}\vspace{0.5em}\\
Department of Computer Science\\
University of California, Santa Barbara
\vspace{-0.5em}
}

\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\N}{\mathcal{N}}
\DeclareMathOperator{\Def}{def}

\newcommand{\defeq}{\stackrel{\Def}{=}{}}
\newcommand{\eqnote}[1]{\text{\footnotesize{[#1]}}}

\newcommand{\X}{\mathsf{X}}
\newcommand{\M}{\mathsf{M}}
\newcommand{\E}[1]{\left<#1\right>}
\newcommand{\abs}[1]{\left|#1\right|}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathcal{Q}}

\newcommand{\TTT}{\text{TTT}}
\newcommand{\TTU}{\text{TTU}}
\newcommand{\TUU}{\text{TUU}}
\newcommand{\TUT}{\text{TUT}}
\newcommand{\UTU}{\text{UTU}}
\newcommand{\UTT}{\text{UTT}}

\newcommand{\figurename}{Figure}
\newcommand{\tablename}{Table}

\graphicspath{{graphs/}}

\begin{document}
\maketitle

\begin{abstract}
\input{abstract.tex}
\end{abstract}

\vspace{1em}
\section{Introduction}\label{sec:intro}

The experimental evaluation of new wireless technologies inevitably requires network traffic. The most commonly used models for generating wireless local-area network (\caps{WLAN}) workload are simple, treating both nodes and flows uniformly, and are generally understood to be na\"ive. They remain popular because of the lack of compelling alternatives. The scope of impact of different workloads on network performance is largely unresearched and thus, poorly understood. We believe that one of the major handicaps preventing wireless experiments from more accurately predicting real-world performance is the lack of realistic models of \caps{WLAN} workload.

%The evaluation of wireless technology requires the generation of workload to test the viability and performance of the new protocol or technique being studied. We believe that lack of realism in traffic workload generation is one of the major limiting factors that prevents simulations and experimental test-bed deployments from accurately predicting the real-world performance of wireless technologies. Today, very little is understood about the impact of different workloads on network performance. Uniform constant bit-rate traffic (\caps{CBR}) is commonly used to evaluate protocols, but there is no evidence that behavior under such workloads is an accurate predictor of performance under real usage patterns. The inability to experimentally forecast real-world performance is a severe handicap to the entire networking community. It hinders the ability to effectively develop better solutions to the many difficult problems that face emerging wireless technologies.

One of the hardest problems in modeling phenomena as complex as network traffic patterns lies in determining which characteristics of any given exemplar of behavior are essential and which can safely be abstracted away. In \caps{WLAN} traffic modeling, important characteristics have largely been determined based on intuition. In previous work~\cite{Karpinski07:realism}, we introduced a fundamentally different approach to evaluating the realism of workload models. Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics. We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values that are similar to those observed using the original trace to generate traffic. Using this new approach, we explore the space of traffic models, establishing rigorously for the first time that common but simplistic models of \caps{WLAN} traffic drastically misrepresent important performance metrics at all levels of the protocol stack.

Our analysis shows that the standard constant bit-rate (\caps{CBR}) traffic model, with uniformly chosen nodes and uniform flows of traffic, significantly misrepresents important performance metrics at every level of the network. To qualify these results in more depth, we also investigate partially synthetic variants of this standard model, allowing us to reason about which aspects of workload models have the most impact. For the most commonly used model, the delay of application data traversing the network from sender to receiver is underestimated by more than a factor of 5, \textit{on average}, and by more than a factor of 11 in a quarter of usage scenarios. Network control overhead is overestimated, on average, by a factor of about 2.5, while link control overhead is underestimated by almost a factor of 2. Even for metrics where the average and median misrepresentations are not so extreme, the statistical characteristics of error values often indicate that something is fundamentally unrealistic. None of the partially synthetic models manage to accurately represent more than one or two performance characteristics. We further use the \caps{AODV}~\cite{rfc:aodv} and \caps{OLSR}~\cite{rfc:olsr} ad~hoc routing protocols to show that the relative performance of protocols can be inverted when changing from using a simplistic traffic model to using real traffic: using random end-point, uniform \caps{CBR} traffic, \caps{AODV} appears to induce less link-layer overhead than \caps{OLSR}, whereas using real traffic, it in fact induces more.

The first major contribution of this work is the definition of ``sufficient realism,'' together with the analytical and statistical methodology to rigorously test whether synthetic traffic models meet this definition. The second contribution is the conclusion that the most commonly used model of wireless traffic drastically and consistently misrepresents some of the most important metrics for performance evaluation of wireless protocols. Many performance comparisons based on this model may need revisiting. The last and most important contribution is the collection of ideas and analytical tools necessary to create more realistic synthetic traffic models in the future. By applying the methods developed in this work, it will be possible to discover precisely which aspects of network usage affect the realism of performance results. Once this is known, it will become feasible to create models that accurately reproduce those aspects of real wireless usage.

The rest of the paper is organized as follows. In Section~\ref{sec:motivation} we present motivation and related work. Our experimental and analytical methodology is presented in Section~\ref{sec:methodology}, while the results of our experiments are explained and analyzed in Section~\ref{sec:results}. The ramifications of these results are discussed in Section~\ref{sec:discussion}. Finally, in Section~\ref{sec:conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.

\section{Motivation \& Related Work}\label{sec:motivation}\label{sec:related-work}

The history of networking research contains many examples of simplistic models that have proven not only to be inaccurate, but also to drastically skew important characteristics of network behavior. Paxson and Floyd showed that the Poisson packet arrival model, which had been standard for studying wide-area Internet traffic, failed to capture the burstiness and self-similarity of real traffic~\cite{Paxson95}. The equally common but simplistic Random Way-Point~(\caps{RWP}) mobility model was found to exhibit ``density waves'' and gradual slow-down of average node speed~\cite{Royer01,Yoon03:speed-decay}. In the worst cases, overly simplistic models can switch the relative performance of protocols, thereby invalidating the conclusions drawn from performance comparisons using those models.

The interaction of wireless user and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted. The joint pattern of traffic generation and mobility through time and space completely determines the effect of wireless usage on the lower levels of the network. This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner.%~\cite{Clark88}.
\footnote{This is violated by some quality of service (QoS) schemes. However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid. The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.} The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic and motion in the experiments are sufficiently realistic.

Paxson and Floyd observed in~\cite{Paxson95} that the interplay between end-point behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback. Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06}. This presents a fundamental difficulty, however, in that it presumes that we know the intent of end-points: what \textit{would} they have done under different conditions? While we can speculate about what an individual node might hypothetically do, we currently do not understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations. Especially in the wireless setting, a fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback. Accordingly, in this paper we attempt to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without multi-level feedback. It is important to realize that while this does not provide a final picture, we currently lack even a first-order understanding of the effect of different workloads on performance. This first-order understanding is an essential initial step.

There have been a significant number of studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}. These analyses have described a wide variety of aspects of wireless network behavior, and provide much insight into the workings of real, deployed wireless networks. These studies present a broad analysis of general system features and trends of specific corporate wireless local-area networks (\caps{WLAN}s)~\cite{Tang99,Balazinska03}, university campus \caps{WLAN}s~\cite{Tang00,Kotz02,Chinchilla04,Schwab04,Henderson04,Tuduce05}, and temporary \caps{WLAN}s at conference venues~\cite{Balachandran02,Jardosh05:ewind}. They also provide a large body of raw data for subsequent analysis and modeling research. Our work provides the methodology for turning this rich foundation of field data into usable, realistic models of workload for a wide variety of networking situations.

The choice of mobility models for mobile wireless simulations can have a drastic impact on important performance metrics~\cite{Camp02,Yoon03:speed-decay,Yoon03:sound-models,Jardosh03,Zheng04}. Moreover, commonly used but simplistic mobility models, such as \caps{RWP}, exhibit characteristics, including density waves and speed decay, that are categorically dissimilar from any known real-world behavior~\cite{Royer01,Yoon03:speed-decay,Yoon03:sound-models}. In response to this evidence, more realistic mobility models have been proposed~\cite{Jardosh03,Tuduce05,Jardosh05:voronoi}. While much of this work focuses on making models that are simply more intuitively appealing~\cite{Jardosh03,Jardosh05:voronoi}, some work has begun to capitalize on this newly created wealth of wireless field data by deriving models from observed usage behavior, rather than intuition alone~\cite{Balazinska03,Tuduce05}.
%The analysis of Balazinska \textit{et al.}~\cite{Balazinska03} introduces the notions of the prevalence and persistence of an \caps{AP} with respect to mobile nodes.\footnote{These concepts are adapted from Paxson's work on Internet routing behavior: ``\textit{prevalence} meaning the overall likelihood that a particular route is encountered, and \textit{persistence}, the likelihood that a route remains unchanged over a long period of time''~\cite{Paxson96}.} Tuduce \textit{et al.}~\cite{Tuduce05} have based their mobility model on the persistence and prevalence characteristics of real trace data, which makes their work particularly relevant to our definition of sufficient realism. They do not, however, determine the impact of prevalence and persistence characteristics on performance metrics.

In this paper, instead of mobility, we examine an even more fundamental aspect of user behavior in wireless networks: the pattern of traffic generated by users and applications. This aspect of behavior is more fundamental because it applies to all types of wireless networks, not just mobile and ad~hoc networks. Moreover, the effect of traffic patterns applies not only to simulations, but also to experimental deployments, which have become the gold standard for wireless protocol evaluation.  Experimental deployment sidesteps the issue of accurately modeling the lower layers of the network. Unless traffic and mobility are modeled realistically, however, the experimental results will still be unreliable.

There is a large and diverse body of work on traffic analysis, modeling, and generation~\cite{Paxson95,Paxson96,Sommers04,Avallone04,Hernandez06}. We are only able to discuss a small, but hopefully representative sampling of this work. Almost all of the traffic generation work has focused on wide-area Internet backbone traffic.
%Because of the comparatively unlimited capacity of wired networks, traffic workload generally does not offer interesting technical challenges at the edge of the network.
The two most prominent traffic generation frameworks are Harpoon and \caps{D-ITG}. Harpoon~\cite{Sommers04} uses a traffic trace for self-training, and can subsequently generate synthetic traffic with certain statistical properties based on the original trace. The properties reproduced are the empirical distributions of the following: ``file size, inter-connection time, source and destination IP ranges, number of active sessions.'' There is no criterion proposed to determine whether these properties characterize the original traffic adequately---we can only hope that this approximation is good enough. For many purposes, it likely is sufficient; in particular, it is probably appropriate for the intended use in generating traffic for Internet backbone simulations. Wireless networks, however, are particularly sensitive to workload conditions, and sampling from a limited set of empirical distributions does not suffice to reproduce realistic network-wide traffic.

\caps{D-ITG}~\cite{Avallone04} generates flows using an independent sampling model for packet sizes and inter-packet intervals. The framework contains pre-made models for several common types of Internet traffic. The focus of this project, however, is on providing the infrastructure to generate very large volumes of synthetic Internet-like backbone traffic. No analysis is provided for determining the realism of traffic mixes, or for choosing flow end-points realistically. In wireless networks, these factors are of crucial importance to performance, and cannot be overlooked. Both Harpoon and \caps{D-ITG} provide excellent traffic generation platforms, but do not provide a systematic framework for understanding or reproducing realistic whole-network workload in the wireless setting.

% TODO: add in the Hernandez-Cortes' dissertation.
% TODO: concluding remarks about what's missing from traffic work in general.
%   + there's no central goal or theme---because realism is ill-defined and subjective
%   + there are no rigorous tests for whether realism is achieved since it's ill-defined
%   + in wireless, where the impact is greater, there's been virtually no research of impact
% TODO: cite a bunch of papers that use the CBR traffic model.

\section{Methodology}\label{sec:methodology}

The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results. Clearly, we need not simulate subatomic particle interactions in a wireless network simulator. Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same. Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior. This requirement leads us to the following definition:

\vspace{0.5em}
\begin{samepage}
\noindent\textbf{Definition.}~A model of user behavior is \textit{sufficiently realistic} if, when compared with actual user behavior, the model, with parameter values extracted from the real data, yields statistically equivalent performance results.
\end{samepage}
\vspace{0.5em}

This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required. We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.

\subsection{Trace Data}\label{sec:trace-data}
%\subsection{Trace Data \& Simulations}\label{sec:trace-data}\label{sec:simulations}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{nodes-flows}%
\vspace{-0.75em}%
\caption{The number of active nodes and flows over time.} 
\label{fig:nodes-flows}
\end{center}
\vspace{-2em}
\end{figure}

Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of synthetic traffic models, including the standard random, uniform \caps{CBR} model. For our analysis, we use a 24-hour trace recorded in an infrastructured 802.11g wireless \caps{LAN} with 18 access points, deployed at the 60th Internet Engineering Task Force meeting (\caps{IETF60}), held in San Diego during August of 2004. The traffic trace was captured using \texttt{\small{tcpdump}} at a single router, through which all wireless traffic for the meeting was routed, including traffic between wireless nodes. The snap length of the capture was 100 bytes, allowing \caps{IP}, \caps{ICMP}, \caps{UDP} and \caps{TCP} headers to be analyzed. We limit our work to the 24-hour sub-trace recorded on Wednesday, August 4th. This trace contains a broad variety of behaviors and entails a very large volume of traffic: 2.1 million flows, 58 million packets, and 52 billion bytes.
% 2,085,563 flows, 58,377,994 packets of application data, and 51,988,245,537 bytes of application data.

We do not assume or claim that the traffic found at \caps{IETF60} is representative of conference settings in general. The observed behaviors are also unlikely to resemble those found in a typical commercial or residential setting. We have chosen this trace, however, because within it can be found behaviors resembling many different types of wireless usage cases. Figure~\ref{fig:nodes-flows} shows the wide variations in the number of active flows and nodes over the course of the trace. In the night and morning hours, the traffic patterns are similar to those one might find in a moderately trafficked business or residential area. During working group sessions, we see highly concentrated, heavy usage patterns. At the zenith of activity, over 800 users, 33 thousand flows, and 1 million packets are seen in a single 10-minute trace segment. At the nadir, a lone node sent only a single 61-byte packet in 10 minutes. All levels of activity between these extremes are represented. Moreover, the mix of traffic types observed changes dramatically over the course of the day, providing a wide representation of possible blends of behavior. This heterogeneity and extreme range of behaviors makes the \caps{IETF} data set ideal for this evaluation. The variety of activity gives us greater confidence that success or failure of traffic models is not tied to any specific network condition, but is broadly and generally applicable. %For a traffic model to be truly realistic, it must be realistic across many different usage cases.

\begin{table*}
\begin{center}\small
\input{models.tex}
\caption{The three orthogonal levels of traffic behavior, and the traffic models used for each level.}\label{tab:traffic-models}
\end{center}
\vspace{-1.75em}
\end{table*}

Before using the traces, it is necessary to extract application-level behavior from the trace header data. First, we split the trace into individual packet flows. A flow is a series of packets sharing the following five attributes: \caps{IP} and transport protocols (raw \caps{IP}, \caps{ICMP}, \caps{TCP}, \caps{UDP}); source and destination \caps{IP} addresses and \caps{TCP}/\caps{UDP} port numbers. Next, the quantity of application-initiated data contained in each packet is calculated. For non-\caps{TCP} packets, this quantity is simply the size of the transport-layer payload, but for \caps{TCP} the calculation is more complicated: only new data transfers, explicitly initiated by the application are counted. Data retransmitted by \caps{TCP} is disregarded, and empty \caps{ACK}s are ignored. \caps{SYN} and \caps{FIN} flags in packets (even empty ones) are counted as a single byte each, since they are explicitly signaled by the application. %After processing, we have a collection of flows, each with a sequence of timestamps and the amount of application-initiated data sent at that time.

\subsection{Simulations}\label{sec:simulations}

We use the Qualnet wireless network simulator to perform our experiments. We simulate a stationary multi-hop 802.11b network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol, with nodes placed randomly in a square field with sides of 1500 meters.
In addition to the active nodes corresponding to trace \caps{IP}s, equally many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays. Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings. We do not attempt to reproduce the physical environment of the original wireless network, nor do we simulate mobility. The only aspect of the original network's behavior that is reproduced is the total pattern of network-wide traffic.

There are a number of potential objections to this approach. We use single-hop trace data to drive multi-hop simulations; the physical environment, node mobility, handover behavior, and closed-loop dynamics (including \caps{TCP} feedback) of the original wireless setting are not faithfully reproduced. One must keep in mind, however, that the goal of this research is \textit{not} to understand the conditions of the original network. Rather, we are using the traffic behaviors observed as examples to help us better understand how different types of workload can affect performance metrics. In particular, we aim to understand how real workload compares with common synthetic traffic models. Of course, the reason for such objections is that networking researchers understand that the many aspects of behavior interact with each other in a complex and nearly inextricable manner. However, before we can hope to understand the interaction between workload and other features affecting network behavior, we must study traffic patterns alone, and learn to model them with reasonable accuracy in the absence of additional complicating factors. Accordingly, in this study, we detach application level traffic patterns from the other factors influencing network conditions, and study them in isolation.

The 24-hour trace is split into 144 10-minute segments, each of which serves as the basis for a set of simulations using different traffic models. The traffic models range from a completely realistic trace-driven model, to a standard \caps{CBR} traffic model. Various partially synthetic intermediate models, described in Section~\ref{sec:models-metrics}, are simulated to study the impact of different aspects of traffic behavior on network performance. To preserve the fairness of the performance comparison, we keep as many features as possible constant across different traffic models. The traffic generated by each synthetic model preserves as many characteristics from the original trace as possible, within the constraints of the model. Moreover, the following features are preserved across all models: the numbers of wireless nodes, the number of flows, the number of application-initiated data units sent, the total bytes of application data sent, and the average flow duration (and therefore the average data rate).

\subsection{Traffic Models \& Performance Metrics}\label{sec:models-metrics}

\begin{figure*}
\begin{center}
\begin{minipage}[c]{0.67\linewidth}
\subfloat[Flow Topology: Trace]
{\includegraphics[width=1.55in]{flow_topology_trace}\label{fig:flow-topology-trace}}
\subfloat[Flow Topology: Sink]
{\includegraphics[width=1.55in]{flow_topology_sink}\label{fig:flow-topology-sink}}
\subfloat[Flow Topology: Uniform]
{\includegraphics[width=1.55in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
\subfloat[Flow Behavior: Trace]
{\includegraphics[width=1.55in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
\subfloat[Flow Behavior: Uniform]
{\includegraphics[width=1.55in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
\subfloat[Packet Behavior]
{\includegraphics[width=1.55in]{flow_packet_behavior}\label{fig:packet-behavior}}
\end{minipage}\hfill
\begin{minipage}[c]{0.30\linewidth}
\caption{Examples illustrating the different traffic models for the three levels of behavior. Figures \ref{fig:flow-topology-trace}, \ref{fig:flow-topology-sink} and \ref{fig:flow-topology-uniform}, show example flow topologies. The width of each line is proportional to the logarithm of the number of flows between the nodes it connects. In each graph, node zero is the gateway to the Internet. Uniform and trace flow behavior examples are plotted in Figures \ref{fig:flow-behavior-trace} and \ref{fig:flow-behavior-uniform}. The time axis indicates when the various flows start and end; the width of each flow line is proportional to the logarithm of its data rate. Flow numbers are assigned arbitrarily. Figure \ref{fig:packet-behavior} compares packet behavior for the uniform model (i.e. {\footnotesize{CBR}}), with the trace of an actual flow. In the uniform model, the cumulative data sent increases smoothly over time (gray diagonal line). In the actual packet trace, the data transmissions are variable both in size and in inter-transmission interval, leading to a ``lumpy'' cumulative data history (black step-function).}
\label{fig:traffic-models}
\end{minipage}
\end{center}
\vspace{-1em}
\end{figure*}

We separate our traffic generation models into three orthogonal and nearly independent levels of behavior:
\begin{enumerate}
\item \textbf{Flow End-Point Topology}: which nodes communicate with each other, and how frequently; i.e. how flow end-points are mapped onto nodes in the network.
\item \textbf{Flow Behavior}: high-level parameters for each flow, including start time, end time, packets sent, bytes sent.
\item \textbf{Packet Behavior}: sizes of individual packets, and the intervals between their transmission.
\end{enumerate}
For each level of traffic behavior, we compare several different behavior models. The different levels of models and variants at each level are listed and described in Table~\ref{tab:traffic-models} and illustrated with specific examples in Figure~\ref{fig:traffic-models}.

The three levels of behavior are orthogonal and can be varied almost independently. The exception to their independence is that the trace-based packet behavior model can only be used when the flow behavior model is also trace-based. Once flow behavior is decoupled from the actual trace, there is no natural way to preserve packet behavior. This eliminates three combinations of models and leaves nine viable behaviors, abbreviated by the first letters of their flow topology, flow behavior, and packet behavior models: \caps{TTT} (fully trace-based), \caps{TTU}, \caps{TUU}, \caps{STT}, \caps{STU}, \caps{SUU} (entirely synthetic, sink topology), \caps{UTT}, \caps{UTU}, \caps{UUU} (entirely synthetic, uniform topology).

We have selected nine performance metrics at the application, network, and link layers of the protocol stack:
\begin{enumerate}
\setlength{\itemsep}{0em}
\item \textbf{Application:} average end-to-end delay, average jitter, total received throughput.
\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), \caps{RREQ}s initiated per node, routing queue drop rate.
\item \textbf{Link (\caps{MAC}):} control overhead (\caps{RTS/CTS/ACK}), packet retransmission rate, retransmission failure rate.
\end{enumerate}
These metrics are commonly used to evaluate the performance of new wireless protocols. For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
%Unfortunately, not all of the simulations were able to run to completion. In some scenarios, the number of nodes and flows caused Qualnet to simply fail. We were, however, able to get complete results for 72.5\% of the simulations, providing a clear view of overall trends.\!\footnote{That 27.5\% of simulations did not finish may seem like a cause for concern, but it does not in fact invalidate our results. The simulations that did not finish are those for which the number of flows and nodes overwhelmed the capacity of the simulator. Qualnet cannot simulate more than circa 650 nodes sending 35,000 flows. While it would be of great interest to examine results for these large-scale scenarios, they are essentially useless for comparison to other simulations studies: other studies do not simulate this quantity of nodes and flows precisely because even the best existing simulators do not support it. Once the simulators become more scalable and robust, our analysis can be extended to these scenarios as well.}

\subsection{Measures of Error}\label{sec:error-measures}

The simulations described in Section~\ref{sec:models-metrics} provide us with the raw data to compare performance metrics for synthetic traffic models with those for real traffic traces. To assess the realism of these models, however, we need a measure of how inaccurate the synthetic performance values are when compared to the real values. Let $x$ be the value of a performance metric using real traffic, and $y$ the value of the same metric using an alternate traffic model, $\M$. Some common measures of error are the difference [$y-x$], the ratio [$y/x$], and the standard error [$(y-x)/x$]. These are all reasonable measures of error; but which is most appropriate for assessing the realism of performance metrics? Instead of picking one arbitrarily, we will first consider the properties that an ideal error function should have, and then use those properties to determine the best measure of error. Moreover, we will show that the unique measure of error that exhibits these ideal properties is the  \textit{log-ratio} of metric values:
\begin{align}
\log(y/x)=\log(y)-\log(x).
\end{align}
In this discussion, $E(x,y)$ is a generic error function applied to the synthetic value, $y$, with respect to the real value, $x$.

The first property that an error function should have is  \textit{insensitivity to common factors}. That is, if both values are scaled by the same constant, the error should be unaffected:
\begin{align}
\label{eqn:normalization-invariance}
\forall\: x,y,c: E(x c,y c) &= E(x,y)
\end{align}
There are three major motivations for this requirement:
\begin{enumerate}
\item Changing units should not affect error values.
\item Error values for ``large'' and ``small'' scenarios should be directly comparable. Scenarios with large $x$ values will naturally have larger raw differences between $x$ and $y$. This requirement allows scenarios of different scales to be compared fairly and without bias.
\item Changing between metrics that differ by a known constant for each scenario should not affect error values.
\end{enumerate}
The last point is best illustrated by an example. Consider two closely related performance metrics: average throughput, $t$, and total bytes received, $r$. Suppose that there are $f$ flows in a given scenario with average duration, $d$. Since $t =r/fd$, the metrics $t$ and $r$ contain the same information---they differ only by a known constant in each simulation scenario. Equation~\ref{eqn:normalization-invariance} ensures that the errors of these metrics are the same:
\begin{align}
E(t^\TTT,t^\M)
	= E\left({\frac{r^\TTT}{fd},\frac{r^\M}{fd}}\right)
	= E(r^\TTT,r^\M).
\end{align}
The difference measure does not satisfy Equation~\ref{eqn:normalization-invariance}, but the ratio, standard error, and log-ratio error measures all do.

The second property that an ideal error function should have is  \textit{additivity of compounded errors}. If two independent causes of error each induce some factor of misrepresentation, then the combined error should be the sum of the errors caused by each factor separately:
\begin{align}
\label{eqn:linear-compounding}
\forall\: x,c_1,c_2:
	E(x,x c_1 c_2) &= E(x,x c_1) + E(x,x c_2).
\end{align}
This property allows us to compare error values meaningfully across different traffic models. For example, if flow topology and packet behavior affected some performance metric independently with no interaction effects, we would expect that
\begin{align}
E(x^\TTT,x^\UTU) \approx E(x^\TTT,x^\TTU) + E(x^\TTT,x^\UTT).
\end{align}
If these two values differ significantly, there must be some interaction between the two levels of behavior that introduces more error than can be explained by each separately. Without the property of additivity given in Equation~\ref{eqn:linear-compounding}, such a comparison would not be possible or meaningful.

Additivity of compounded errors also implies two desirable properties that are easily derived from Equation~\ref{eqn:linear-compounding}. It forces the error of an accurate representation to be zero: $E(x,x) = 0$. It also forces underestimation and overestimation to be treated symmetrically. The error of underestimating by some factor is opposite but equal to overestimating by the same factor:
\begin{align}
E(x,x/c) = - E(x,xc).
\end{align}

It is easily verified that the difference, ratio, and standard error measures do not satisfy Equation~\ref{eqn:linear-compounding}, and the difference, as noted, does not satisfy Equation~\ref{eqn:normalization-invariance}. The log-ratio is the only metric presented that satisfies both conditions. Moreover, it can be proved that $\log(y/x)$ is the \textit{only} differentiable function that satisfies both (up to a constant). In the Appendix, we present a proof of this claim. Throughout the rest of the paper, we use the log-ratio to measure the error of performance metrics.
% TODO: should Appendix be capitalized?
% TODO: hyperref the Appendix for PDF.

\subsection{Tests of Statistical Equivalence}\label{sec:statistical-equivalence}

\begin{sidewaysfigure*}
%\begin{centering}
\noindent
\subfloat[Application: Average Jitter]
{\includegraphics[width=4.6in]{Application-Average-Jitter}
\label{fig:Application-Average-Jitter}}
\subfloat[Application: Average Delay]
{\includegraphics[width=4.6in]{Application-Average-Delay}
\label{fig:Application-Average-Delay}}
\subfloat[Application: Received Throughput]
{\includegraphics[width=4.6in]{Application-Received-Throughput}
 \label{fig:Application-Received-Throughput}}
\subfloat[Network: Control Overhead]
{\includegraphics[width=4.6in]{Network-Control-Overhead}
 \label{fig:Network-Control-Overhead}}
\subfloat[Network: Packet Drop Rate]
{\includegraphics[width=4.6in]{Network-Packet-Drop-Rate}
 \label{fig:Network-Packet-Drop-Rate}}
\subfloat[Network: \caps{RREQ}s Initiated]
{\includegraphics[width=4.6in]{Network-RREQs-Initiated}
 \label{fig:Network-RREQs-Initiated}}
\subfloat[Link (MAC): Control Overhead]
{\includegraphics[width=4.6in]{MAC-Control-Overhead}
 \label{fig:MAC-Control-Overhead}}
\subfloat[Link (MAC): Retransmissions]
{\includegraphics[width=4.6in]{MAC-Retransmissions}
 \label{fig:MAC-Retransmissions}}
\subfloat[Link (\caps{MAC}): Retransmission Failures]
{\includegraphics[width=4.6in]{MAC-Retransmission-Failures}
 \label{fig:MAC-Retransmission-Failures}}
\hspace{0.035in}
\begin{minipage}[t]{4.545in}
\vspace{-2em}
\caption{Box-and-whisker plots of log-ratio error values for all metrics and traffic models. The lower axis indicates the log-ratio, while the upper axis shows raw ratio values. Each box contains the central majority of log-ratio values: the left and right bounds are at the $25^\text{th}$ and $75^\text{th}$ percentiles. The dark middle line indicates the median value, while the diamond marks the mean. The whiskers (dotted lines) extend to the furthest non-outlier values, while the points beyond that are outliers. The notches in the middle show a $95\%$ confidence interval for the actual underlying median value; if two notches do not overlap, they are very unlikely to have the same median. Numbers to the right are $p$-values for the median, mean, and 3-mean tests, respectively; $p$-values above $\alpha=0.05$ appear in bold.}
\label{fig:boxplots}
\end{minipage}
%\end{centering}
\end{sidewaysfigure*}

In this section, we consider the values of performance metrics as random variables, drawn from unknown distributions. We present three tests for the statistical equivalence of the metric values induced by synthetic and real traffic. Let $\M$ be a traffic model as before and let $\X$ be a performance metric. Let $X_k^\M$ be a random variable representing the value of $\X$ in the $k^{\text{th}}$ scenario using the traffic model $\M$. If the distribution of $X_k^\M$ is the same as that of $X_k^\TTT$, then both the median and mean values of the log-ratio $R_k^\M=\log(X_k^\M/X_k^\TTT)$ should be zero. The first two tests check the plausibility of precisely these hypotheses. The third test separates small, medium, and large simulation scenarios, and test their means separately to catch any size-dependent performance bias.

\textbf{The Median Test.} If $\M$ induces realistic performance, then the median of each log-ratio variable, $R_k^\M=\log(X_k^\M/X_k^\TTT)$, should be zero. The $k^\text{th}$ indicator variable is defined as
\begin{align}
	I_k^\M = \begin{cases}
		0 & \text{if $R_k^\M < 0$,} \\
		1 & \text{if $R_k^\M \ge 0$.}
	\end{cases}
\end{align}
If the median value of $R_k^\M$ is truly zero, then $I_k^\M$ is a Bernoulli variable with probability parameter $p=\frac{1}{2}$. The variables $I_k^\M$ are all independent since they come from separate simulations, and cannot affect each other's outcomes. Therefore, the sum $S_n^\M = \sum_{k=1}^n {I_k^\M}$ should follow a binomial distribution of $n$ trials with $p=\frac{1}{2}$. The median test applies the exact binomial cumulative distribution function (\caps{CDF}) % TODO: cite this.
 for $n$ and $p$ to the observed value of $S_n^\M$, yielding a $p$-value: the probability that such an extreme value would occur by chance under the hypothesis that the median of each $R_k^\M$ is zero. % TODO: fix the wording here.

\begin{figure*}
\begin{centering}
\subfloat[App. Average Jitter: {\footnotesize{STU}}]
{\includegraphics[width=1.75in]{application-average-jitter-stu}%
\label{fig:application-average-jitter-stu}}
\subfloat[Net. Packet Drop Rate: {\footnotesize{UTU}}]
{\includegraphics[width=1.75in]{network-packet-drop-rate-utu}%
\label{fig:network-packet-drop-rate-utu}}
\subfloat[Net. Packet Drop Rate: {\footnotesize{STU}}]
{\includegraphics[width=1.75in]{network-packet-drop-rate-stu}%
\label{fig:network-packet-drop-rate-stu}}
\subfloat[Link Retrans. Failures: {\footnotesize{SUU}}]
{\includegraphics[width=1.75in]{mac-retransmission-failures-suu}%
\label{fig:mac-retransmission-failures-suu}}
\caption{Scatter plots of various metrics for an alternate traffic model ({\footnotesize{STU}}, {\footnotesize{UTU}}, {\footnotesize{SUU}}) versus real trace traffic~({\footnotesize{TTT}}). Each data point represents a pair of matched simulations: the $x$-coordinate is the metric value for trace traffic, the $y$-coordinate is the metric value for the synthetic model. Scenarios are plotted according to the amount of flows in their trace segment: the bottom third with filled dots; the middle third with circles; the top third with squares.}
\label{fig:scatterplots}
\end{centering}
\vspace{-1em}
\end{figure*}

\textbf{The Mean Test.} We use Lyapunov's generalization of the Central Limit Theorem (\caps{LCLT}) to test the hypothesis that the mean of each $R_k^\M$ is zero. We present the theorem and its application to the series $R_k^\M$ in the Appendix. Here we simply present the resulting test statistic and its usage. If the mean of each $R_k^\M$ is zero, then \caps{LCLT} implies that the test statistic
\begin{align}
\hat{Z}_n^\M &= \frac{\sum_{k=1}^n{R_k^\M}}{\sqrt{\sum_{k=1}^n{(R_k^\M)^2}}}
\end{align}
converges to a standard normal distribution for large values of $n$, where $n$ is the number of simulated scenarios. In this case, $n=144$, which is fairly large by traditional statistical standards. The $p$-value of the mean test is given by applying the standard normal \caps{CDF} to the test statistic, $\hat{Z}_n^\M$.

\textbf{The 3-Mean Test.} Performance behavior in scenarios with a large number of nodes or flows is often very different from behavior in small scenarios. In some cases, $R_k^\M$ is skewed positively for one group, but negatively for another. In such cases, $R_k^\M$ can pass the median and mean tests even though behavior in each case is unrealistic. To catch such situations, we split the simulations into three groups by the number of flows: the lower, middle, and upper thirds. The 3-mean test simply applies the mean test to each of these groups separately and uses the minimum $p$-value of the three. This reduces the power of the test, since each group is smaller, but can catch cases where the error has size-dependent biases that cancel out on average.

\section{Results}\label{sec:results}

Our simulation results are summarized in Figure~\ref{fig:boxplots}. Each subfigure shows a single performance metric. The distribution of errors for each traffic model is visualized with a box-and-whisker plot. These plots allow immediate assessment of realism: a good traffic model should have log-ratio values that are tightly clustered around the center, with a small, evenly balanced box. Additionally, the mean and median markers should be close to the center. Complementing the visual display of summary statistics, Figure~\ref{fig:boxplots} also lists three $p$-values to the right of each box-and-whisker plot. These are, in order, the $p$-values for the median, mean and 3-mean tests described in Section~\ref{sec:statistical-equivalence}. Each test catches a different type of unrealistic statistical behavior.

The \caps{UTU} model, for example, which uses real flow behavior but synthetic flow topology and packet behavior, does a very good job of accurately reproducing realistic average end-to-end delay (Figure~\ref{fig:Application-Average-Delay}): the median and mean are both close to zero, and the $p$-values are all greater than $0.05$. The standard uniform \caps{CBR} model (\caps{UUU}), on the other hand, underestimates average end-to-end delay by between a factor of $2$ and $11$ half of the time, and by more than a factor of $11$ a quarter of the time. Its $p$-values for delay are all less than $0.005$.

Figure~\ref{fig:Application-Average-Jitter} shows that the \caps{STU} traffic model passes the mean test but not the median or 3-mean tests. This result indicates that this model tends to overestimate jitter, but in some subset of cases it significantly underestimates instead, yielding a mean near zero, but a skewed median. This analysis is verified when we examine a scatterplot of metric values for \caps{STU} versus \caps{TTT} traffic models. In Figure~\ref{fig:application-average-jitter-stu} we can see that for the majority of simulations, the \caps{STU} value for jitter exceeds the \caps{TTT} value. In a significant minority, however, the \caps{STU} model yields very small jitter values, while the \caps{TTT} model gives large values. Even though the average error is near zero, this is not a model that reproduces realistic jitter.

The \caps{UTU} traffic model accurately reproduces packet dropping behavior in routing queues. Figure~\ref{fig:network-packet-drop-rate-utu} shows a scatterplot for such a realistic model: the data points are well clustered symmetrically around the diagonal line. We conclude that the primary influence on the packet drop rate is flow behavior. However, if we use a sink topology model, as in Figure~\ref{fig:network-packet-drop-rate-stu}, the drop rate becomes inflated. The sink topology introduces an unrealistic routing bottleneck in the network, causing excessive queue overflows for all sizes of scenarios. The uniform topology model does not exhibit this bottleneck, and thereby avoids producing this unrealistic performance artifact. This demonstrates that while the uniform topology model is generally less realistic than the sink model, for certain metrics, their relative quality is the opposite.

Figure~\ref{fig:mac-retransmission-failures-suu} illustrates a case where the 3-mean test catches unrealistic behavior that the median and mean tests do not catch. From Figure~\ref{fig:MAC-Retransmission-Failures} we can see that the mean and median error values for the \caps{SUU} model are fairly small. The scatter-plot, however, shows that this model significantly underestimates the failure rate for large scenarios with many flows (squares), while overestimating the rate in smaller scenarios.

There are few, if any, positive conclusions that can be drawn from these results. The primary message is that these synthetic traffic models, especially the standard uniform \caps{CBR} model, consistently misrepresent the most important performance metrics. The traffic model that performs the best overall is the \caps{UTU} model, which uses real flow behavior with uniform flow topology and uniform packet behavior. This model, however, still fails statistical tests of realism for all but two of the metrics considered. Further development of traffic models is needed before it becomes possible to generate traffic in simulations or experimental deployments such that a single set of experiments can realistically evaluate all aspects of network performance.

\section{Discussion}\label{sec:discussion}

What are the ramifications of these results? The discovery that the most commonly used traffic model for wireless networks drastically misrepresents important performance metrics may shed some light on the lack of trust in results from wireless simulations. It is now well established that network usage behavior---both mobility, and, with this research, traffic patterns---have an impact on network performance that cannot be ignored. Even experimental deployments cannot avoid the need for more realistic traffic workload models. While using a real, physical network successfully sidesteps simulation problems below the application layer, without realistic traffic models, reliable, meaningful performance predictions remain beyond our reach.

\subsection{Relative Performance Comparisons}

Performance evaluations are primarily used to compare new protocols with existing ones. It remains possible that while misrepresenting important metrics, synthetic traffic models preserve the relative performance of protocols. In order to test this hypothesis, we have run further simulations, revisiting a classic comparison of ad~hoc routing protocols: \caps{AODV} \textit{vs.} \caps{OLSR}. % TODO: cite a previous comparison here.
We focus on the trace segments with 75 or fewer active nodes, since published performance comparisons of ad hoc routing protocols have typically not used more nodes than this. For each scenario, we have run simulations with 10 different seed values using the \caps{AODV} and \caps{OLSR} routing protocols and the \caps{UUU} and \caps{TTT} traffic models, allowing us to compare relative performance results when switching traffic models.

The key result of our comparison is that, for certain performance metrics, \textbf{the relative performance of \caps{AODV} and \caps{OLSR} is inverted, simply  by changing traffic models.} Figure~\ref{fig:mac-control-overhead-cmp} shows some of the results from this experiment. The graph plots \caps{MAC} control overhead against the number of nodes in each simulation scenario. Each data point shows the average overhead for the simulations of that scenario using ten seed values. With the standard \caps{UUU} traffic model, \caps{OLSR} consistently has much higher overhead, across the board, than \caps{AODV}. When real traffic is used, however, \caps{OLSR} performs significantly better than \caps{AODV}. A similar switch occurs for the number of \caps{MAC} retransmissions. This experiment demonstrates that the use of unrealistic traffic models can change the relative performance of different protocols, as well as skew absolute performance measurements.

\begin{figure}
\vspace{0.6em}
\begin{centering}
\includegraphics[width=3.45in]{mac-control-overhead-cmp}
\caption{Comparison of {\footnotesize{MAC}} control overhead for {\footnotesize{AODV}} and {\footnotesize{OLSR}} using both the {\footnotesize{UUU}} and {\footnotesize{TTT}} traffic models. The fit lines show locally weighted, smoothed performance trends.}
\label{fig:mac-control-overhead-cmp}
\end{centering}
\vspace{-1.5em}
\end{figure}

\subsection{Generality of Results}

% TODO: talk about other synthetic models not considered here.
% The point of our analysis is not to show that no synthetic models can be accurate, but rather, that the commonly used models, and obvious variations thereof, are not accurate.

The most significant limitation on the generality of this analysis is that it is based entirely on a single data set from \caps{IETF60}---albeit a large and varied one. It is possible that traffic in this trace happens to produce network performance that is unusually dissimilar to standard traffic models. This data set, however, represents a highly heterogeneous collection of network usage behaviors, from slow and steady off-peak usage, to extremely heavy peak usage: over 800 users, 33 thousand flows, and 1 million packets in a single 10-minute trace segment. Despite the broad variety of behaviors, the results are consistent: in all types of usage scenarios, simplistic traffic models, like uniform \caps{CBR}, systematically skew important performance measurements at all levels of the network. While the precise results for other data sets might differ, it is very unlikely that \caps{CBR} traffic models will happen to accurately reproduce realistic performance in other experiments. This paper provides strong evidence that better traffic models are needed for performance evaluations.

\subsection{Towards Realistic Models of Wireless Workload}

What would better traffic models look like? How can we create them? One possible approach is to use actual traffic traces as we have done. This approach is unsatisfactory, however, because it provides the experimenter with almost no control over experiments. Synthetic models have parameters, which can be tweaked as necessary---adjusting, for example, the number of active nodes in a simulation, without affecting other parameters. Traffic traces, on the other hand, must be used without significant alteration if they are to actually provide the desired realism. The ``messiness'' of the performance comparison from trace data in Figure~\ref{fig:mac-control-overhead-cmp} illustrates why using traces directly is not ideal: each data point differs not only in the number of nodes shown on the $x$-axis, but also in other dimensions, such as the number of flows and packets, and the average flow duration. The result is a highly noisy comparison, affected by many unseen parameters. Only by applying a local smoothing algorithm are trends somewhat elucidated.

Instead of using trace data directly, it should be possible to configure a synthetic traffic model based on observations from a real data set, and then run side-by-side simulations using the synthetic model and the real data, producing statistically equivalent performance results. This is precisely what our definition of sufficient realism entails. The work in this paper provides the tools to measure how close to this ideal a model is and in what areas it needs improvement. Without this feedback, any improvements in realism are purely guesswork. Our breakdown of traffic behavior into three orthogonal levels also allows the problem to be approached in smaller pieces, rather than being solved all at once.

The next step towards better traffic models is to investigate which aspects of real traces may be altered without detrimentally affecting the resulting performance metrics. For example, to test whether a complex time-series model of packet behavior is necessary, we randomize the order of the packet sizes and/or inter-packet intervals and compare performance using these randomized traces against performance using the original traces. If the performance is unchanged, we can conclude that no complex time-series model of packet behavior is necessary: sampling the packet sizes and inter-packet intervals from empirical distributions is sufficiently realistic. If, on the other hand, the performance characteristics are altered by shuffling packets, then some time-series model of packet behavior is needed. By partially randomizing the packet order in specific ways, the exact limits of realism necessary can be found. A similar approach will allow the development of realistic models for the other levels of network usage behavior.

\section{Conclusions}\label{sec:conclusions}

%The real significance of this paper lies not in showing that the commonly used \caps{CBR} traffic model is unrealistic. Networking researchers understand that \caps{CBR} is at best a na\"ive approximation of real network traffic. Nor is the real significance that we have quantized the high degree of inaccuracy that can be induced by using this model. The real significance of this paper lies in providing an well-defined, objective measure of realism for traffic models. This has never previously existed: evaluations of realism have relied on essentially arbitrary statistical measures of similarity to real traffic. Existing traffic generators use empirical distributions of such quantities as file size, inter-connection time, packet size, or inter-packet intervals. They can only hope that reproducing these distributions as good enough. Without an external measure of accuracy, there's simply no way to know.

%attempting to showing that existing models are unrealistic, or that new models are more realistic have always focused on arbitrary statistical measures. A

This research rigorously quantifies the impact of a variety of synthetic traffic models on performance metrics that wireless researchers use to evaluate new technologies and protocols. The first step in this assessment process was to formally define what it means for a network usage model to be sufficiently realistic. In essence, a model is considered realistic if it produces performance results that are statistically equivalent to those produced by real usage.
A well-defined, objective measure of realism for traffic models has not previously existed. Evaluations of realism have formerly relied on essentially arbitrary statistical measures of similarity to real traffic, which may or may not affect the performance metrics that researchers care about.
The definition of sufficient realism leads us to our general experimental approach: we use differential analysis comparing performance metrics derived from real traffic with those derived from synthetic traffic models. The theoretical contributions of this analysis are:
\begin{enumerate}
\item An in-depth analysis of the desirable mathematical properties of a measure of error for performance metrics.
\item Proof that the unique measure of error that satisfies these properties is the log-ratio of metric values.
\item Three rigorous tests of statistical equivalence between synthetic and real performance results.
\end{enumerate}
These analytical tools allow the evaluation of realism over a collection of drastically different usage scenarios. Evaluation over a heterogeneous collection of scenarios is essential to establishing the credibility of usage models. Moreover, these theoretical results are equally applicable to other types of usage models---for example, mobility.

On the practical side, this paper gives crucial insight into why most researchers do not trust simulation results: with the traffic models commonly used, the results are unlikely to reflect real performance. Moreover, this problem will also hamper experiments in test-bed wireless networks, so long as the same na\"ive workload models are used. The only way to address this fundamental lack of realism is to develop usage models that reproduce important performance metrics more accurately. Our theoretical results provide the tools necessary to do this. The development of better traffic models should begin with real traces, and proceed by incremental changes, checked by differential analysis.
%First, alter a small aspect of the trace, simulate, then compare. If the realism of the results is unaffected, the traffic feature altered was inessential. Otherwise, it is a feature of behavior that must be captured in a realistic traffic model.
This approach will allow the precise mapping of which aspects of traffic patterns have an impact on performance and which ones can be safely abstracted away.
% TODO: maybe cut out some of this last bit.

\section{Acknowledgments}
This work was funded in part through NSF Career Award CNS-0347886.

\vspace{-0.5em}
\appendix\footnotesize

\noindent
\textbf{Theorem.} The unique differentiable function satisfying Eqs.~\ref{eqn:normalization-invariance}~\&~\ref{eqn:linear-compounding} is $E(1,e) \ln(y/x)$. \textit{Proof:} Let $f(z) = E(1,z)$. Eq.~\ref{eqn:normalization-invariance} gives $E(x,y) = E(1,y/x) = f(y/x)$. Eq.~\ref{eqn:linear-compounding} gives: $f(z) = f(z/w) + f(w)$. Differentiation by $z$ yields $f'(z) = w^{-1} f'(z/w)$. In particular, if we choose $w=z$, we get $f'(z) = z^{-1} f'(1)$. Integration by $z$ gives:
$f(z) = f'(1) \int z^{-1} dz + c = f'(1) \ln(z) + c$.
By Eq.~\ref{eqn:linear-compounding}, $f(1)=0$, so $c=0$. Thus $f(e)=f'(1)\ln(e)=f'(1)$. We conclude that $f(z)=f(e)\ln(z)$, so $E(x,y) = f(y/x) = f(e)\ln(y/x) = E(1,e)\ln(y/x)$, as desired.\hfill\QED

\vspace{1em}\noindent
\textbf{Theorem.}~[\textit{Lyapunov's Central Limit Theorem}] Let $\{R_k\}_{k=1}^{\infty}$ be a series of independent variables with $\E{R_k}=0$. Let $s_n^2 = \sum_{k=1}^n{\E{\abs{R_k^2}}}$ and $r_n^3 = \sum_{k=1}^n{\E{\abs{R_k^3}}}$. For each $n$, let $Z_n$ be the normalized mean of $\{R_k\}_{k=1}^n$: $Z_n = \sum_{k=1}^n{R_k/s_n}$. If $\lim_{n\to\infty} {r_n/s_n} = 0$, then $\lim_{n\to\infty}{Z_n} \sim \mathcal{N}(0,1)$ (the standard normal distribution). (See~\cite{Feller68} page 229.)

\vspace{1em}
To apply the LCLT to the series $R_k^\M=\log(X_k^\M/X_k^\TTT)$, we must show that under the null hypothesis, the assumptions of the theorem are satisfied by this series. First, the null hypothesis, implies that $\E{R_k}=0$. Variables from separate simulations are independent since they cannot influence each other's values. Formally, $\Pr\left(X_k\middle|X_j\right)=\Pr\left(X_k\right)$. Therefore the log-ratios are also independent for different $k$. The last requirement is that $\lim_{n\to\infty}{r_n/s_n}=0$. To verify this, we use the estimators $\hat{s}_n^2=\sum{\abs{R_k^2}}$, and $\hat{r}_n^3=\sum{\abs{R_k^3}}$. When $\hat{r}_n/\hat{s}_n$ are plotted on a log-log scatter plot, with $n$ increasing up to the number of simulations, they asymptotically approach a downwardly sloped line as $n$ grows. Thus $\lim_{n\to\infty}{\ln(\hat{r}_n/\hat{s}_n)/\ln(n)}=c<0$. This implies that $\lim_{n\to\infty}{\ln(\hat{r}_n/\hat{s}_n)}=-\infty$, and therefore $\lim_{n\to\infty}{r_n/s_n}=\lim_{n\to\infty}{\hat{r}_n/\hat{s}_n}=0$. This test for the convergence of $r_n/s_n$ is applied to each model and metric pair. %The asymptotic approach of the ratio $\hat{r}_n/\hat{s}_n$ to a downwardly sloped line is verified visually and by comparison to a fitted regression line.

\vfill

%\pagebreak
% Hernandez06: ``complete methodology for reproducing the traffic observed on a network link in a closed-loop manner, and proposed a number of metrics for studying the realism of the generated traffic.''
\bibliography{IEEE,references}

\end{document}

### 2008-01-17 paper.tex
Minor edits.

@@
 
-The interaction of wireless user and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted.
+The interaction of wireless users and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted.
 The joint pattern of traffic generation and mobility through time and space completely determines the effect of wireless usage on the lower levels of the network.
@@
 Moreover, the effect of traffic patterns applies not only to simulations, but also to experimental deployments, which have become the gold standard for wireless protocol evaluation.
-Experimental deployment sidesteps the issue of accurately modeling the lower layers of the network.
+Experimental deployments sidestep the problem of accurately modeling the lower layers of the network.
 Unless traffic and mobility are modeled realistically, however, the experimental results will still be unreliable.
@@
 At the zenith of activity, over 800 users, 33 thousand flows, and 1 million packets are seen in a single 10-minute trace segment.
-At the nadir, a lone node sent only a single 61-byte packet in 10 minutes.
+At the nadir, a lone node sent only a single 61-byte packet over the course of 10 minutes.
 All levels of activity between these extremes are represented.
@@
 We use the Qualnet wireless network simulator to perform our experiments.
-We simulate a stationary multi-hop 802.11b network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol, with nodes placed randomly in a square field with sides of 1500 meters.
-In addition to the active nodes corresponding to trace \caps{IP}s, equally many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays.
+We simulate a stationary multi-hop 802.11b network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
+In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays.
 Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings.
@@
 \begin{enumerate}
-\item \textbf{Flow End-Point Topology}: which nodes communicate with each other, and how frequently; i.e. how flow end-points are mapped onto nodes in the network.
+\item \textbf{Network Behavior}: which nodes communicate with each other, and how frequently; \textit{i.e.} how flow end-points are mapped onto nodes in the network.
 \item \textbf{Flow Behavior}: high-level parameters for each flow, including start time, end time, packets sent, bytes sent.
@@
 
+% TODO: more here on models...
+% TODO: rewrite next paragraph too...
+
 The three levels of behavior are orthogonal and can be varied almost independently.
@@
 
-We have selected nine performance metrics at the application, network, and link layers of the protocol stack:
+We have selected a variety of performance metrics at the application, network, and link layers of the protocol stack:
 \begin{enumerate}
@@
 \item \textbf{Application:} average end-to-end delay, average jitter, total received throughput.
-\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), \caps{RREQ}s initiated per node, routing queue drop rate.
-\item \textbf{Link (\caps{MAC}):} control overhead (\caps{RTS/CTS/ACK}), packet retransmission rate, retransmission failure rate.
+\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), routing queue drop rate.
+\item \textbf{Link (\caps{MAC}):} control overhead (\caps{RTS/CTS/ACK}), packet retransmission rate.
+% TODO: are we actually using RTS/CTS?
 \end{enumerate}
@@
 
-\vfill
+%\vfill
 

### 2008-01-17 models.tex
More work on the models section.

@@
 \bigstrut & \multirow{1}{*}[-0.05em]{\textbf{S}ynthetic} &
-Use synthetic distributions \textit{\`a la} Hern\'andez-Campos~\cite{Hernandez06:wlan-traffic}. \\
+Use synthetic distributions \textit{\`a la} Hern\'andez-Campos~\textit{et~al.}~\cite{Hernandez06:wlan-traffic}. \\
 \cline{2-3}

### 2008-01-17 paper.tex
More work on the models section.

@@
 \input{models.tex}
-\caption{The three orthogonal levels of traffic behavior, and the traffic models used for each level.}\label{tab:traffic-models}
+\caption{The three levels of traffic behavior and the traffic models defined for each level.}\label{tab:traffic-models}
 \end{center}
@@
 We use the Qualnet wireless network simulator to perform our experiments.
-We simulate a stationary multi-hop 802.11b network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
+We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
 In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays.
@@
 There are a number of potential objections to this approach.
-We use single-hop trace data to drive multi-hop simulations; the physical environment, node mobility, handover behavior, and closed-loop dynamics (including \caps{TCP} feedback) of the original wireless setting are not faithfully reproduced.
+We use single-hop trace data to drive multi-hop simulations; the physical environment, node mobility, handover behavior, and closed-loop dynamics of the original wireless setting are not faithfully reproduced.
 One must keep in mind, however, that the goal of this research is \textit{not} to understand the conditions of the original network.
@@
 
-We separate our traffic generation models into three orthogonal and nearly independent levels of behavior:
-\begin{enumerate}
-\item \textbf{Network Behavior}: which nodes communicate with each other, and how frequently; \textit{i.e.} how flow end-points are mapped onto nodes in the network.
-\item \textbf{Flow Behavior}: high-level parameters for each flow, including start time, end time, packets sent, bytes sent.
-\item \textbf{Packet Behavior}: sizes of individual packets, and the intervals between their transmission.
-\end{enumerate}
-For each level of traffic behavior, we compare several different behavior models.
-The different levels of models and variants at each level are listed and described in Table~\ref{tab:traffic-models} and illustrated with specific examples in Figure~\ref{fig:traffic-models}.
+We separate traffic generation into three levels of behavior: network, flow and packet.
+The {network} level determines properties pertaining to nodes and their interaction.
+For example, it determines which nodes communicate with each other, and how frequently.
+This level can be viewed as assigning network nodes to the source and destination roles of flows, as well as determining per-node properties which affect lower level models.
+The next level of behavior determines overall properties of flows.
+This includes each flow's start time, duration, packet count and total data volume.
+Finally, the packet level of behavior determines the individual sizes of packets and the durations of intervals between their transmission.
+Despite the concreteness of the examples mentioned here, there are also less obvious, abstract properties that models at the network and flow levels determine.
+Flow models, for example, dictate the distribution of each flow's packet sizes, a property which can be used to determine individual packet sizes, but which is not itself specific to any particular packet.
+
+%\subsubsection*{Network Behavior Models}
+
+For each level of behavior we define a spectrum of models.
+The gold standard of realism at each level is provided by the ``trace'' model for that level, which replicates the node, flow or packet behavior of the original trace exactly.
+The most rudimentary model at each level is the ``uniform'' model, which treats all entities identically---be they nodes, flows or packets.
+Despite being so simplistic (or perhaps because of it), the uniform behavior models are by far the most commonly for workload generation in experiments.
+Between these extremes on the spectrum of realism, we consider one or two models with intermediate degrees of complexity and fidelity.
+At the packet level, we define the ``sample'' model, which is in essence a per-flow variable bit-rate (\caps{VBR}) model: each packet's size is determined by randomly sampling a value from the packet size distribution given by the flow model; the interval until the next packet is then sampled from the given interval distribution.\footnote{Note that the uniform packet model (\textit{i.e.} {\scriptsize{CBR}}), can be considered to be a special case of the sample model where the size and time distributions are both constant.} The intermediate model at the flow level is designated as ``proportional'': the distributions of inter-packet ...
+
+%\begin{enumerate}
+%\item \textbf{Network Behavior}: which nodes communicate with each other, and how frequently; \textit{i.e.} how flow end-points are mapped onto nodes in the network.
+%\item \textbf{Flow Behavior}: high-level parameters for each flow, including start time, end time, packets sent, bytes sent.
+%\item \textbf{Packet Behavior}: sizes of individual packets, and the intervals between their transmission.
+%\end{enumerate}
+
+%The different levels of models and variants at each level are listed and described in Table~\ref{tab:traffic-models} and illustrated with specific examples in Figure~\ref{fig:traffic-models}.
 

### 2008-01-18 paper.tex
YAML files describing models and properties defined by them.

@@
 
-\hyphenation{op-tical net-works semi-conduc-tor IEEEtran}
+\hyphenation{net-works IEEEtran}
 \bibliographystyle{IEEEtran}
@@
 
-\title{%
-%\vspace{-0.2em}
-\thetitle%
+\title{
+\thetitle\\
+\vspace{0.2em}
+\LARGE{On The Inadequacy of Uniform Modeling for WLAN Workloads}
 }
@@
 
+%We consider four models of network-level behavior. The most faithful is the ``trace'' model, which assigns each flow
+
 For each level of behavior we define a spectrum of models.
@@
 Between these extremes on the spectrum of realism, we consider one or two models with intermediate degrees of complexity and fidelity.
-At the packet level, we define the ``sample'' model, which is in essence a per-flow variable bit-rate (\caps{VBR}) model: each packet's size is determined by randomly sampling a value from the packet size distribution given by the flow model; the interval until the next packet is then sampled from the given interval distribution.\footnote{Note that the uniform packet model (\textit{i.e.} {\scriptsize{CBR}}), can be considered to be a special case of the sample model where the size and time distributions are both constant.} The intermediate model at the flow level is designated as ``proportional'': the distributions of inter-packet ...
+At the packet level, we define the ``sample'' model, which is in essence a per-flow variable bit-rate (\caps{VBR}) model: each packet's size is determined by randomly sampling a value from the packet size distribution given by the flow model; the interval until the next packet is then sampled from the given interval distribution.\footnote{Note that the uniform packet model (\textit{i.e.} {\scriptsize{CBR}}), can be considered to be a special case of the sample model where the size and time distributions are both constant.} The intermediate model at the flow level is designated as ``proportional'': the distributions of packet sizes and inter-packet intervals are taken from the trace and passed down to the packet model (which may use the actual distribution, via sampling, or simply use the mean size and intervals, to generate \caps{CBR} traffic).
 

### 2008-01-18 properties.rb
YAML parsing and ruby to generate properties table automatically.

@@ -1,0 +1,30 @@
+#!/usr/bin/ruby
+
+require 'yaml'
+
+properties = YAML.load_file("properties.yaml")
+
+print <<-'HEADER'
+\begin{table}
+\small
+\framebox[3.47in][l]{
+\begin{minipage}[t]{3.47in}
+\vspace{0.4em}
+HEADER
+properties.each do |level|
+  name = level.keys.first
+  props = level[name]
+  puts "  \\textbf{#{name}:}"
+  puts '  \begin{itemize}'
+  props.each do |prop|
+    puts "    \\item #{prop}"
+  end
+  puts '  \end{itemize}'
+end
+print <<-'FOOTER'
+\vspace{0.4em}
+\end{minipage}
+}
+\caption{Properties of simulation objects at various levels.}
+\end{table}
+FOOTER

### 2008-01-20 models.tex
Itemize-based tabels for properties & models.

--- 2008-01-16 models.tex
@@
-\begin{tabular}{|c|l|p{5in}|}
-\multicolumn{1}{c}{\textbf{Behavior}} &
-\multicolumn{1}{c}{\textbf{Model}} &
-\multicolumn{1}{c}{\textbf{Description}} \\
-\hline
-\multirow{3}[13]{*}{Network}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
-Mapping of flow end-points to wireless nodes taken directly from trace data. \\
-\cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{M}arginal} &
-Use marginal empirical distributions from trace. \\
-\cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{S}ynthetic} &
-Use synthetic distributions \textit{\`a la} Hern\'andez-Campos~\cite{Hernandez06:wlan-traffic}. \\
-\cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{U}niform} &
-Both end-points are uniformly randomly chosen from all of the wireless nodes. \\
-\hline
-\multirow{2}[11]{*}{Flow}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
-Each flow has the actual start time, end time, total data sent, and number of packets from the trace. \\
-\cline{2-3}
-\bigstrut & \multirow{1}{*}{\textbf{P}roportional} &
-Byte count taken from trace; sizes and intervals are CBR or VBR, on a per-flow basis. \\
-\cline{2-3}
-\bigstrut & \multirow{1}{*}{\textbf{U}niform} &
-All flows have the same duration, volume, packet rate, and data rate as the trace average. \\
-\hline
-\multirow{2}[11]{*}{Packet}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
-The sequence of packet sizes and inter-transmission intervals is taken directly from the trace. \\
-\cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{S}ample} &
-Packets sizes and intervals are randomly drawn from per-flow empirical distributions (VBR). \\
-\cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{U}niform} &
-Packets sizes are uniform and the inter-packet interval is constant (CBR). \\
-\hline
-\end{tabular}
+\noindent
+\textbf{Node:}
+\begin{itemize}
+  \item \textit{Individual}
+  \begin{itemize}
+    \item each node corresponds to a specific trace node with associated trace flows
+    \item session start time assigned from trace
+    \item overrides all distributions with per-node versions
+  \end{itemize}
+  \item \textit{Nonparametric}
+  \begin{itemize}
+    \item generic flows assigned to nodes by sampling
+    \item session start time assigned from pool of start times
+  \end{itemize}
+  \item \textit{Parametric}
+  \begin{itemize}
+    \item generic flows assigned to nodes according to BiPareto model
+    \item session start time randomly chosen from viable range
+  \end{itemize}
+  \item \textit{Uniform}
+  \begin{itemize}
+    \item all nodes have average number of generic flows
+    \item session start times uniformly cover viable range
+  \end{itemize}
+\end{itemize}
+\noindent
+\textbf{Flow:}
+\begin{itemize}
+  \item \textit{Individual}
+  \begin{itemize}
+    \item each flow corresponds to a specific trace flow with an associated packet trace
+    \item flow size \& start time taken from trace
+    \item overrides distributions of size/interval packet behaviors with flow-specific versions
+  \end{itemize}
+  \item \textit{Nonparametric}
+  \begin{itemize}
+    \item flow size is sampled from inherited distribution of flow sizes
+    \item start time relative to previous session flow sampled from inherited distribution
+    \item inherits distributions of size/interval packet behaviors from destination node
+  \end{itemize}
+  \item \textit{Parametric}
+  \begin{itemize}
+    \item flow size is sampled from BiPareto model fitted to inherited distribution
+    \item start time relative to previous session flow sampled Lognormal fitted to inherited
+    \item inherits distributions of size/interval packet behaviors from destination node
+  \end{itemize}
+  \item \textit{Uniform}
+  \begin{itemize}
+    \item all flows have size equal to inherited mean
+    \item average inter-flow interval used between all flows
+    \item inherits distributions of size/interval packet behaviors from destination node
+  \end{itemize}
+\end{itemize}
+\noindent
+\textbf{Packet:}
+\begin{itemize}
+  \item \textit{Individual}
+  \begin{itemize}
+    \item each packet corresponds to a specific trace packet
+  \end{itemize}
+  \item \textit{Nonparametric}
+  \begin{itemize}
+    \item packet sizes and intervals are sampled from inherited distributions
+  \end{itemize}
+  \item \textit{Parametric}
+  \begin{itemize}
+    \item packet sizes and intervals are drawn from inherited parametric distributions
+  \end{itemize}
+  \item \textit{Uniform}
+  \begin{itemize}
+    \item all packet sizes and inter-packet intervals are equal to inherited mean
+  \end{itemize}
+\end{itemize}

### 2008-01-20 paper.tex
YAML parsing and ruby to generate properties table automatically.
Itemize-based tabels for properties & models.

@@
 Using this new approach, we explore the space of traffic models, establishing rigorously for the first time that common but simplistic models of \caps{WLAN} traffic drastically misrepresent important performance metrics at all levels of the protocol stack.
+
+\begin{table}[t]
+\noindent
+\begin{tabular}[t]{|l|}
+\hline
+\begin{minipage}[l]{3.315in}
+\begin{small}
+\vspace{0.8em}
+\input{properties.tex}
+\vspace{0.7em}
+\end{small}
+\end{minipage}
+\\\hline
+\end{tabular}
+\caption{Summary of properties defined for the hierarchy of object types in the traffic object model ({\footnotesize{TOM}}).}
+\end{table}
+
+\begin{table}[t]
+\noindent
+\begin{tabular}[t]{|l|}
+\hline
+\begin{minipage}[l]{3.315in}
+\begin{small}
+\vspace{0.8em}
+\input{models.tex}
+\vspace{0.7em}
+\end{small}
+\end{minipage}
+\\\hline
+\end{tabular}
+\caption{Summary of how the models at each behavioral level override or inherit properties in the traffic object model ({\footnotesize{TOM}}).}
+\end{table}
 
@@
 
-\begin{table*}
-\begin{center}\small
-\input{models.tex}
-\caption{The three levels of traffic behavior and the traffic models defined for each level.}\label{tab:traffic-models}
-\end{center}
-\vspace{-1.75em}
-\end{table*}
+%\begin{table*}
+%\begin{center}\small
+%\input{models.tex}
+%\caption{The three levels of traffic behavior and the traffic models defined for each level.}\label{tab:traffic-models}
+%\end{center}
+%\vspace{-1.75em}
+%\end{table*}
 

### 2008-01-20 properties.tex
Itemize-based tabels for properties & models.

\noindent
\textbf{Network:}
\begin{itemize}
  \item collection of nodes
  \item collection of flows
  \item distribution of flow sizes
  \item distribution of intervals between flow start times
  \item distribution of packet sizes
  \item distribution of inter-packet intervals
\end{itemize}
\noindent
\textbf{Node:}
\begin{itemize}
  \item collection of source flows
  \item collection of destination flows
  \item session start time
  \item distribution of flow sizes
  \item distribution of intervals between flow start times
  \item distribution of packet sizes
  \item distribution of inter-packet intervals
\end{itemize}
\noindent
\textbf{Flow:}
\begin{itemize}
  \item source
  \item destination
  \item start time
  \item size (byte count)
  \item distribution of packet sizes
  \item distribution of inter-packet intervals
\end{itemize}
\noindent
\textbf{Packet:}
\begin{itemize}
  \item send time
  \item payload size (bytes)
\end{itemize}

### 2008-01-20 models.rb

@@ -1,0 +1,25 @@
+#!/usr/bin/ruby
+
+require 'yaml'
+
+models = YAML.load_file("models.yaml")
+
+models.each do |l|
+  level = l.keys.first
+  mods = l[level]
+  puts '\noindent'
+  puts "\\textbf{#{level}:}"
+  puts '\begin{itemize}'
+  mods.each do |m|
+    mod = m.keys.first
+    items = m[mod]
+    puts "  \\item \\textit{#{mod}}"
+    puts '  \begin{itemize}'
+    items.each do |item|
+      item.gsub!(/&/,'\\\&')
+      puts "    \\item #{item}"
+    end
+    puts '  \end{itemize}'
+  end
+  puts '\end{itemize}'
+end

### 2008-01-20 properties.rb

@@ -4,27 +4,14 @@ require 'yaml'
 
 properties = YAML.load_file("properties.yaml")
 
-print <<-'HEADER'
-\begin{table}
-\small
-\framebox[3.47in][l]{
-\begin{minipage}[t]{3.47in}
-\vspace{0.4em}
-HEADER
-properties.each do |level|
-  name = level.keys.first
-  props = level[name]
-  puts "  \\textbf{#{name}:}"
-  puts '  \begin{itemize}'
+properties.each do |l|
+  level = l.keys.first
+  props = l[level]
+  puts '\noindent'
+  puts "\\textbf{#{level}:}"
+  puts '\begin{itemize}'
   props.each do |prop|
-    puts "    \\item #{prop}"
+    puts "  \\item #{prop}"
   end
-  puts '  \end{itemize}'
+  puts '\end{itemize}'
 end
-print <<-'FOOTER'
-\vspace{0.4em}
-\end{minipage}
-}
-\caption{Properties of simulation objects at various levels.}
-\end{table}
-FOOTER

### 2008-01-20 models.tex
Finally tweaked the models & properties tables to satisfaction.

--- 2008-01-16 models.tex
@@
-\begin{tabular}{|c|l|p{5in}|}
-\multicolumn{1}{c}{\textbf{Behavior}} &
+\begin{tabular}{|c|c|p{4.3in}|}
+\multicolumn{1}{c}{\textbf{Level}} &
 \multicolumn{1}{c}{\textbf{Model}} &
-\multicolumn{1}{c}{\textbf{Description}} \\
+\multicolumn{1}{c}{\textbf{Description}} \\[2pt]
 \hline
-\multirow{3}[13]{*}{Network}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
-Mapping of flow end-points to wireless nodes taken directly from trace data. \\
+\multirow{8}{*}[3.0pt]{Node}
+& \multirow{1}{*}[-0.05em]{Individual} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+each node corresponds to a specific trace node with associated trace flows \\
+session start time assigned from trace \\
+overrides all distributions with per-node versions
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{M}arginal} &
-Use marginal empirical distributions from trace. \\
+& \multirow{1}{*}[-0.05em]{Nonparametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+generic flows assigned to nodes by sampling \\
+session start time assigned from pool of start times
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{S}ynthetic} &
-Use synthetic distributions \textit{\`a la} Hern\'andez-Campos~\cite{Hernandez06:wlan-traffic}. \\
+& \multirow{1}{*}[-0.05em]{Parametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+generic flows assigned to nodes according to BiPareto model \\
+session start time randomly chosen from viable range
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{U}niform} &
-Both end-points are uniformly randomly chosen from all of the wireless nodes. \\
+& \multirow{1}{*}[-0.05em]{Uniform} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+all nodes have average number of generic flows \\
+session start times uniformly cover viable range
+\vspace{2pt}
+\end{minipage} \\
 \hline
-\multirow{2}[11]{*}{Flow}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
-Each flow has the actual start time, end time, total data sent, and number of packets from the trace. \\
+\multirow{11}{*}[2.5pt]{Flow}
+& \multirow{1}{*}[-0.05em]{Individual} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+each flow corresponds to a specific trace flow with an associated packet trace \\
+flow size \& start time taken from trace \\
+overrides distributions of size/interval packet behaviors with flow-specific versions
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}{\textbf{P}roportional} &
-Byte count taken from trace; sizes and intervals are CBR or VBR, on a per-flow basis. \\
+& \multirow{1}{*}[-0.05em]{Nonparametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+flow size is sampled from inherited distribution of flow sizes \\
+start time relative to previous session flow sampled from inherited distribution \\
+inherits distributions of size/interval packet behaviors from destination node
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}{\textbf{U}niform} &
-All flows have the same duration, volume, packet rate, and data rate as the trace average. \\
+& \multirow{1}{*}[-0.05em]{Parametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+flow size is sampled from BiPareto model fitted to inherited distribution \\
+start time relative to previous session flow sampled Lognormal fitted to inherited \\
+inherits distributions of size/interval packet behaviors from destination node
+\vspace{2pt}
+\end{minipage} \\
+\cline{2-3}
+& \multirow{1}{*}[-0.05em]{Uniform} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+all flows have size equal to inherited mean \\
+average inter-flow interval used between all flows \\
+inherits distributions of size/interval packet behaviors from destination node
+\vspace{2pt}
+\end{minipage} \\
 \hline
-\multirow{2}[11]{*}{Packet}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
-The sequence of packet sizes and inter-transmission intervals is taken directly from the trace. \\
+\multirow{5}{*}[2.5pt]{Packet}
+& \multirow{1}{*}[-0.05em]{Individual} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+each packet corresponds to a specific trace packet
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{S}ample} &
-Packets sizes and intervals are randomly drawn from per-flow empirical distributions (VBR). \\
+& \multirow{1}{*}[-0.05em]{Nonparametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+packet sizes and intervals are sampled from inherited distributions
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{U}niform} &
-Packets sizes are uniform and the inter-packet interval is constant (CBR). \\
+& \multirow{1}{*}[-0.05em]{Parametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+packet sizes and intervals are drawn from inherited parametric distributions
+\vspace{2pt}
+\end{minipage} \\
+\cline{2-3}
+& \multirow{1}{*}[-0.05em]{Uniform} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+all packet sizes and inter-packet intervals are equal to inherited mean
+\vspace{2pt}
+\end{minipage} \\
 \hline

### 2008-01-20 models.rb

@@ -4,22 +4,52 @@ require 'yaml'
 
 models = YAML.load_file("models.yaml")
 
+class Object
+  def leaves; self; end
+end
+class Array
+  def leaves; collect{|x| x.leaves}.flatten; end
+end
+class Hash
+  def leaves; values.collect{|x| x.leaves}.flatten; end
+end
+
+fixups = {
+  :Node   => [ 8,'3.0pt'],
+  :Flow   => [11,'2.5pt'],
+  :Packet => [ 5,'2.5pt']
+}
+
+puts <<'HEADER'
+\begin{tabular}{|c|c|p{4.3in}|}
+\multicolumn{1}{c}{\textbf{Level}} &
+\multicolumn{1}{c}{\textbf{Model}} &
+\multicolumn{1}{c}{\textbf{Description}} \\[2pt]
+HEADER
 models.each do |l|
   level = l.keys.first
   mods = l[level]
-  puts '\noindent'
-  puts "\\textbf{#{level}:}"
-  puts '\begin{itemize}'
-  mods.each do |m|
-    mod = m.keys.first
-    items = m[mod]
-    puts "  \\item \\textit{#{mod}}"
-    puts '  \begin{itemize}'
-    items.each do |item|
+  fixup = fixups[level.to_sym]
+  puts '\hline'
+  puts "\\multirow{#{fixup.first}}{*}[#{fixup.last}]{#{level}}"
+  mods.each_with_index do |m,j|
+    model = m.keys.first
+    items = m[model]
+    items = [items] unless items.is_a?(Array)
+    puts "& \\multirow{1}{*}[-0.05em]{#{model}} &"
+    puts '\begin{minipage}[l]{4.3in}'
+    puts '\vspace{2pt}'
+    items.each_with_index do |item,k|
       item.gsub!(/&/,'\\\&')
-      puts "    \\item #{item}"
+      # puts "& & " if k > 0
+      puts "#{item}" + ((k < items.size-1) ? ' \\\\' : '')
     end
-    puts '  \end{itemize}'
+    puts '\vspace{2pt}'
+    puts '\end{minipage} \\\\'
+    puts '\cline{2-3}' if j < models.size
   end
-  puts '\end{itemize}'
 end
+puts <<'FOOTER'
+\hline
+\end{tabular}
+FOOTER

### 2008-01-20 properties.rb

@@ -7,7 +7,6 @@ properties = YAML.load_file("properties.yaml")
 properties.each do |l|
   level = l.keys.first
   props = l[level]
-  puts '\noindent'
   puts "\\textbf{#{level}:}"
   puts '\begin{itemize}'
   props.each do |prop|

### 2008-01-20 models.tex
Added subtle bullets.

--- 2008-01-17 models.tex
@@
-\begin{tabular}{|c|l|p{5in}|}
-\multicolumn{1}{c}{\textbf{Behavior}} &
+\begin{tabular}{|c|c|p{4.3in}|}
+\multicolumn{1}{c}{\textbf{Level}} &
 \multicolumn{1}{c}{\textbf{Model}} &
-\multicolumn{1}{c}{\textbf{Description}} \\
+\multicolumn{1}{c}{\textbf{Description}} \\[2pt]
 \hline
-\multirow{3}[13]{*}{Network}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
-Mapping of flow end-points to wireless nodes taken directly from trace data. \\
+\multirow{8}{*}[2.5pt]{Node}
+& \multirow{1}{*}[-0.05em]{Individual} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} each node corresponds to a specific trace node with associated trace flows \\
+\raisebox{1.5pt}{$\centerdot$} session start time assigned from trace \\
+\raisebox{1.5pt}{$\centerdot$} overrides all distributions with per-node versions
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{M}arginal} &
-Use marginal empirical distributions from trace. \\
+& \multirow{1}{*}[-0.05em]{Nonparametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} generic flows assigned to nodes by sampling \\
+\raisebox{1.5pt}{$\centerdot$} session start time assigned from pool of start times
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{S}ynthetic} &
-Use synthetic distributions \textit{\`a la} Hern\'andez-Campos~\textit{et~al.}~\cite{Hernandez06:wlan-traffic}. \\
+& \multirow{1}{*}[-0.05em]{Parametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} generic flows assigned to nodes according to BiPareto model \\
+\raisebox{1.5pt}{$\centerdot$} session start time randomly chosen from viable range
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{U}niform} &
-Both end-points are uniformly randomly chosen from all of the wireless nodes. \\
+& \multirow{1}{*}[-0.05em]{Uniform} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} all nodes have average number of generic flows \\
+\raisebox{1.5pt}{$\centerdot$} session start times uniformly cover viable range
+\vspace{2pt}
+\end{minipage} \\
 \hline
-\multirow{2}[11]{*}{Flow}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
-Each flow has the actual start time, end time, total data sent, and number of packets from the trace. \\
+\multirow{11}{*}[2.5pt]{Flow}
+& \multirow{1}{*}[-0.05em]{Individual} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} each flow corresponds to a specific trace flow with an associated packet trace \\
+\raisebox{1.5pt}{$\centerdot$} flow size \& start time taken from trace \\
+\raisebox{1.5pt}{$\centerdot$} overrides distributions of size/interval packet behaviors with flow-specific versions
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}{\textbf{P}roportional} &
-Byte count taken from trace; sizes and intervals are CBR or VBR, on a per-flow basis. \\
+& \multirow{1}{*}[-0.05em]{Nonparametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} flow size is sampled from inherited distribution of flow sizes \\
+\raisebox{1.5pt}{$\centerdot$} start time relative to previous session flow sampled from inherited distribution \\
+\raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}{\textbf{U}niform} &
-All flows have the same duration, volume, packet rate, and data rate as the trace average. \\
+& \multirow{1}{*}[-0.05em]{Parametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} flow size is sampled from BiPareto model fitted to inherited distribution \\
+\raisebox{1.5pt}{$\centerdot$} start time relative to previous session flow sampled Lognormal fitted to inherited \\
+\raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node
+\vspace{2pt}
+\end{minipage} \\
+\cline{2-3}
+& \multirow{1}{*}[-0.05em]{Uniform} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} all flows have size equal to inherited mean \\
+\raisebox{1.5pt}{$\centerdot$} average inter-flow interval used between all flows \\
+\raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node
+\vspace{2pt}
+\end{minipage} \\
 \hline
-\multirow{2}[11]{*}{Packet}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{T}race} &
-The sequence of packet sizes and inter-transmission intervals is taken directly from the trace. \\
+\multirow{5}{*}[2.5pt]{Packet}
+& \multirow{1}{*}[-0.05em]{Individual} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} each packet corresponds to a specific trace packet
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{S}ample} &
-Packets sizes and intervals are randomly drawn from per-flow empirical distributions (VBR). \\
+& \multirow{1}{*}[-0.05em]{Nonparametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals are sampled from inherited distributions
+\vspace{2pt}
+\end{minipage} \\
 \cline{2-3}
-\bigstrut & \multirow{1}{*}[-0.05em]{\textbf{U}niform} &
-Packets sizes are uniform and the inter-packet interval is constant (CBR). \\
+& \multirow{1}{*}[-0.05em]{Parametric} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals are drawn from inherited parametric distributions
+\vspace{2pt}
+\end{minipage} \\
+\cline{2-3}
+& \multirow{1}{*}[-0.05em]{Uniform} &
+\begin{minipage}[l]{4.3in}
+\vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} all packet sizes and inter-packet intervals are equal to inherited mean
+\vspace{2pt}
+\end{minipage} \\
 \hline

### 2008-01-20 models.rb

@@ -14,8 +14,9 @@ class Hash
   def leaves; values.collect{|x| x.leaves}.flatten; end
 end
 
+bullet = '\\raisebox{1.5pt}{$\\centerdot$}'
 fixups = {
-  :Node   => [ 8,'3.0pt'],
+  :Node   => [ 8,'2.5pt'],
   :Flow   => [11,'2.5pt'],
   :Packet => [ 5,'2.5pt']
 }
@@ -41,8 +42,8 @@ models.each do |l|
     puts '\vspace{2pt}'
     items.each_with_index do |item,k|
       item.gsub!(/&/,'\\\&')
-      # puts "& & " if k > 0
-      puts "#{item}" + ((k < items.size-1) ? ' \\\\' : '')
+      line_end = (k < items.size-1) ? ' \\\\' : ''
+      puts "#{bullet} #{item}" + line_end
     end
     puts '\vspace{2pt}'
     puts '\end{minipage} \\\\'

### 2008-01-20 models.tex
Minor tweaks.

@@
 \multicolumn{1}{c}{\textbf{Model}} &
-\multicolumn{1}{c}{\textbf{Description}} \\[2pt]
+\multicolumn{1}{c}{\textbf{Overridden and inherited properties}} \\[2pt]
 \hline
@@
 \raisebox{1.5pt}{$\centerdot$} generic flows assigned to nodes according to BiPareto model \\
-\raisebox{1.5pt}{$\centerdot$} session start time randomly chosen from viable range
+\raisebox{1.5pt}{$\centerdot$} session start time assigned according to Poisson arrival model
 \vspace{2pt}
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals are drawn from inherited parametric distributions
+\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals are drawn from fitted Exponential distributions
 \vspace{2pt}

### 2008-01-20 models.rb

@@ -25,7 +25,7 @@ puts <<'HEADER'
 \begin{tabular}{|c|c|p{4.3in}|}
 \multicolumn{1}{c}{\textbf{Level}} &
 \multicolumn{1}{c}{\textbf{Model}} &
-\multicolumn{1}{c}{\textbf{Description}} \\[2pt]
+\multicolumn{1}{c}{\textbf{Overridden and inherited properties}} \\[2pt]
 HEADER
 models.each do |l|
   level = l.keys.first

### 2008-01-21 models.tex
Various edits:
  - models.rb:
  	- removed unnecessary #leaves method from Object, Array and Hash
  	- turn "*" at end of line into footnote mark
  - models.yaml:
    - regularized wording of Parametric model entries
    - added footnote markers ("*") at end of fitted Parametric entries
  - paper.tex:
    - changed title
    - work on introduction
    - added footnote text to models table (parametrics fitted with MLE)
  - references.bib:
    - added back Royer01 entry that got lost in the shuffle

@@
 \raisebox{1.5pt}{$\centerdot$} session start time assigned from trace \\
-\raisebox{1.5pt}{$\centerdot$} overrides all distributions with per-node versions
+\raisebox{1.5pt}{$\centerdot$} overrides all distributions with node-specific versions
 \vspace{2pt}
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} generic flows assigned to nodes by sampling \\
-\raisebox{1.5pt}{$\centerdot$} session start time assigned from pool of start times
+\raisebox{1.5pt}{$\centerdot$} generic flows assigned to nodes by sampling [TODO---more detail] \\
+\raisebox{1.5pt}{$\centerdot$} session start time randomly assigned from pool of trace session start times
 \vspace{2pt}
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} generic flows assigned to nodes according to BiPareto model \\
-\raisebox{1.5pt}{$\centerdot$} session start time assigned according to Poisson arrival model
+\raisebox{1.5pt}{$\centerdot$} generic flows assigned to nodes according to fitted BiPareto model\raisebox{1pt}{\footnotesize*} \\
+\raisebox{1.5pt}{$\centerdot$} session start time assigned according to Poisson arrival model\raisebox{1pt}{\footnotesize*}
 \vspace{2pt}
@@
 \raisebox{1.5pt}{$\centerdot$} each flow corresponds to a specific trace flow with an associated packet trace \\
-\raisebox{1.5pt}{$\centerdot$} flow size \& start time taken from trace \\
+\raisebox{1.5pt}{$\centerdot$} flow size and start time taken from trace \\
 \raisebox{1.5pt}{$\centerdot$} overrides distributions of size/interval packet behaviors with flow-specific versions
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} flow size is sampled from BiPareto model fitted to inherited distribution \\
-\raisebox{1.5pt}{$\centerdot$} start time relative to previous session flow sampled Lognormal fitted to inherited \\
+\raisebox{1.5pt}{$\centerdot$} flow size is sampled from fitted BiPareto model\raisebox{1pt}{\footnotesize*} \\
+\raisebox{1.5pt}{$\centerdot$} start time relative to previous flow in session sampled from fitted Lognormal model\raisebox{1pt}{\footnotesize*} \\
 \raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals are sampled from inherited distributions
+\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals sampled from inherited distributions
 \vspace{2pt}
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals are drawn from fitted Exponential distributions
+\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals drawn from fitted Exponential distributions\raisebox{1pt}{\footnotesize*}
 \vspace{2pt}

### 2008-01-21 paper.tex
Finally tweaked the models & properties tables to satisfaction.
Minor tweaks.
Various edits:
  - models.rb:
  	- removed unnecessary #leaves method from Object, Array and Hash
  	- turn "*" at end of line into footnote mark
  - models.yaml:
    - regularized wording of Parametric model entries
    - added footnote markers ("*") at end of fitted Parametric entries
  - paper.tex:
    - changed title
    - work on introduction
    - added footnote text to models table (parametrics fitted with MLE)
  - references.bib:
    - added back Royer01 entry that got lost in the shuffle

@@
 
-\newcommand{\thetitle}{Wireless Traffic: The Failure of CBR Modeling}
+\newcommand{\thetitle}{LAN Workload: The Failure of Uniform Modeling}
 
@@
 \usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle}]{hyperref}
+\usepackage{array}
 
@@
 \title{
-\thetitle\\
-\vspace{0.2em}
-\LARGE{On The Inadequacy of Uniform Modeling for WLAN Workloads}
+\thetitle%\\
+%\vspace{0.2em}
+%\LARGE{On The Inadequacy of Uniform Modeling for WLAN Workloads}
 }
@@
 }
+
+\newcommand{\Section}[1]{\S\ref{sec:#1}}
 
@@
 
-\vspace{1em}
 \section{Introduction}\label{sec:intro}
+
+More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and billions of computers on the Internet and in local area networks~(\caps{LAN}s).
+The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
+Other research~\cite{Hernandez06:dissertation} has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback (see \Section{related-work}) induced by the ubiquitous transmission control protocol~(\caps{TCP}).
+Recently, the high-level statistical properties of wireless \caps{LAN} traffic have been rigorously modeled with a variety of parametric distributions~\cite{Hernandez06:wlan-traffic}.
+
+Despite great breakthroughs in understanding and methodology, the networking community remains without a cookbook for generating realistic \caps{LAN} workloads.
+Worse still, it is unclear what ``realistic'' should mean in this context.
+Foremost, a clear and testable definition of realism for traffic models is needed.
+In the absence of such a definition, it remains standard in wireless experiments to generate traffic using the simplest possible model: all nodes, flows and packets are assumed to behave uniformly.
+Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily.
+While more complex approaches have been proposed~\cite{xxx}, the simplest, na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better.
+It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.% in exchange for their increased complexity.
+
+%In this paper, we propose objective criteria by which to judge whether
 
@@
 \begin{table}[t]
-\noindent
+\vspace{0.7em}
 \begin{tabular}[t]{|l|}
@@
 \begin{small}
-\vspace{0.8em}
+\vspace{0.6em}
 \input{properties.tex}
-\vspace{0.7em}
+\vspace{0.6em}
 \end{small}
@@
 \end{tabular}
-\caption{Summary of properties defined for the hierarchy of object types in the traffic object model ({\footnotesize{TOM}}).}
+\caption{Summary of classes and their attributes in the proposed Traffic Object Model ({\footnotesize{TOM}}) for {\footnotesize{WLAN}} workload generation.}
+\vspace{-2em}
 \end{table}
 
-\begin{table}[t]
-\noindent
-\begin{tabular}[t]{|l|}
-\hline
-\begin{minipage}[l]{3.315in}
-\begin{small}
-\vspace{0.8em}
+\begin{table*}[t]
+\begin{center}
+\small
 \input{models.tex}
-\vspace{0.7em}
-\end{small}
-\end{minipage}
-\\\hline
-\end{tabular}
-\caption{Summary of how the models at each behavioral level override or inherit properties in the traffic object model ({\footnotesize{TOM}}).}
-\end{table}
+\footnotesize\\[0.5em]
+\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
+\end{center}
+\vspace{-0.75em}
+\caption{Summary of how the models at each behavioral level override or inherit properties in the Traffic Object Model ({\footnotesize{TOM}}).}
+\vspace{-1.75em}
+\end{table*}
 
@@
 The rest of the paper is organized as follows.
-In Section~\ref{sec:motivation} we present motivation and related work.
-Our experimental and analytical methodology is presented in Section~\ref{sec:methodology}, while the results of our experiments are explained and analyzed in Section~\ref{sec:results}.
-The ramifications of these results are discussed in Section~\ref{sec:discussion}.
-Finally, in Section~\ref{sec:conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.
+In \Section{motivation} we present motivation and related work.
+Our experimental and analytical methodology is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
+The ramifications of these results are discussed in \Section{discussion}.
+Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.
 
@@
 Paxson and Floyd observed in~\cite{Paxson95} that the interplay between end-point behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback.
-Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06}.
+Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}.
 This presents a fundamental difficulty, however, in that it presumes that we know the intent of end-points: what \textit{would} they have done under different conditions?
@@
 
-There is a large and diverse body of work on traffic analysis, modeling, and generation~\cite{Paxson95,Paxson96,Sommers04,Avallone04,Hernandez06}.
+There is a large and diverse body of work on traffic analysis, modeling, and generation~\cite{Paxson95,Paxson96,Sommers04,Avallone04,Hernandez06:dissertation}.
 We are only able to discuss a small, but hopefully representative sampling of this work.
@@
 The traffic models range from a completely realistic trace-driven model, to a standard \caps{CBR} traffic model.
-Various partially synthetic intermediate models, described in Section~\ref{sec:models-metrics}, are simulated to study the impact of different aspects of traffic behavior on network performance.
+Various partially synthetic intermediate models, described in \Section{models-metrics}, are simulated to study the impact of different aspects of traffic behavior on network performance.
 To preserve the fairness of the performance comparison, we keep as many features as possible constant across different traffic models.
@@
 
-The simulations described in Section~\ref{sec:models-metrics} provide us with the raw data to compare performance metrics for synthetic traffic models with those for real traffic traces.
+The simulations described in \Section{models-metrics} provide us with the raw data to compare performance metrics for synthetic traffic models with those for real traffic traces.
 To assess the realism of these models, however, we need a measure of how inaccurate the synthetic performance values are when compared to the real values.
@@
 Complementing the visual display of summary statistics, Figure~\ref{fig:boxplots} also lists three $p$-values to the right of each box-and-whisker plot.
-These are, in order, the $p$-values for the median, mean and 3-mean tests described in Section~\ref{sec:statistical-equivalence}.
+These are, in order, the $p$-values for the median, mean and 3-mean tests described in \Section{statistical-equivalence}.
 Each test catches a different type of unrealistic statistical behavior.
@@
 %\pagebreak
-% Hernandez06: ``complete methodology for reproducing the traffic observed on a network link in a closed-loop manner, and proposed a number of metrics for studying the realism of the generated traffic.''
+% Hernandez06:dissertation: ``complete methodology for reproducing the traffic observed on a network link in a closed-loop manner, and proposed a number of metrics for studying the realism of the generated traffic.''
 \bibliography{IEEE,references}

### 2008-01-21 models.rb

@@ -4,16 +4,6 @@ require 'yaml'
 
 models = YAML.load_file("models.yaml")
 
-class Object
-  def leaves; self; end
-end
-class Array
-  def leaves; collect{|x| x.leaves}.flatten; end
-end
-class Hash
-  def leaves; values.collect{|x| x.leaves}.flatten; end
-end
-
 bullet = '\\raisebox{1.5pt}{$\\centerdot$}'
 fixups = {
   :Node   => [ 8,'2.5pt'],
@@ -42,6 +32,7 @@ models.each do |l|
     puts '\vspace{2pt}'
     items.each_with_index do |item,k|
       item.gsub!(/&/,'\\\&')
+      item.gsub!(/\*\s*$/,'\raisebox{1pt}{\footnotesize*}')
       line_end = (k < items.size-1) ? ' \\\\' : ''
       puts "#{bullet} #{item}" + line_end
     end

### 2008-01-21 models.tex
Change Indvidual models to Trace.
Redone intro; added paragraph to related work; etc.
In the process of overhauling the Methodology section.

@@
 \multicolumn{1}{c}{\textbf{Level}} &
-\multicolumn{1}{c}{\textbf{Model}} &
+\multicolumn{1}{c}{\textbf{Sub-model}} &
 \multicolumn{1}{c}{\textbf{Overridden and inherited properties}} \\[2pt]
 \hline
-\multirow{8}{*}[2.5pt]{Node}
-& \multirow{1}{*}[-0.05em]{Individual} &
+\multirow{8}{*}[2.5pt]{\class{Node}}
+& \multirow{1}{*}[-0.05em]{\submodel{Trace}} &
 \begin{minipage}[l]{4.3in}
@@
 \cline{2-3}
-& \multirow{1}{*}[-0.05em]{Nonparametric} &
+& \multirow{1}{*}[-0.05em]{\submodel{Nonparametric}} &
 \begin{minipage}[l]{4.3in}
@@
 \cline{2-3}
-& \multirow{1}{*}[-0.05em]{Parametric} &
+& \multirow{1}{*}[-0.05em]{\submodel{Parametric}} &
 \begin{minipage}[l]{4.3in}
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} generic flows assigned to nodes according to fitted BiPareto model\raisebox{1pt}{\footnotesize*} \\
-\raisebox{1.5pt}{$\centerdot$} session start time assigned according to Poisson arrival model\raisebox{1pt}{\footnotesize*}
+\raisebox{1.5pt}{$\centerdot$} generic flows assigned to nodes according to fitted BiPareto model\raisebox{1.5pt}{\scriptsize*} \\
+\raisebox{1.5pt}{$\centerdot$} session start time assigned according to Poisson arrival model\raisebox{1.5pt}{\scriptsize*}
 \vspace{2pt}
@@
 \cline{2-3}
-& \multirow{1}{*}[-0.05em]{Uniform} &
+& \multirow{1}{*}[-0.05em]{\submodel{Uniform}} &
 \begin{minipage}[l]{4.3in}
@@
 \hline
-\multirow{11}{*}[2.5pt]{Flow}
-& \multirow{1}{*}[-0.05em]{Individual} &
+\multirow{11}{*}[2.5pt]{\class{Flow}}
+& \multirow{1}{*}[-0.05em]{\submodel{Trace}} &
 \begin{minipage}[l]{4.3in}
@@
 \cline{2-3}
-& \multirow{1}{*}[-0.05em]{Parametric} &
+& \multirow{1}{*}[-0.05em]{\submodel{Parametric}} &
 \begin{minipage}[l]{4.3in}
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} flow size is sampled from fitted BiPareto model\raisebox{1pt}{\footnotesize*} \\
-\raisebox{1.5pt}{$\centerdot$} start time relative to previous flow in session sampled from fitted Lognormal model\raisebox{1pt}{\footnotesize*} \\
+\raisebox{1.5pt}{$\centerdot$} flow size is sampled from fitted BiPareto model\raisebox{1.5pt}{\scriptsize*} \\
+\raisebox{1.5pt}{$\centerdot$} start time relative to previous flow in session sampled from fitted Lognormal model\raisebox{1.5pt}{\scriptsize*} \\
 \raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node
@@
 \hline
-\multirow{5}{*}[2.5pt]{Packet}
-& \multirow{1}{*}[-0.05em]{Individual} &
+\multirow{5}{*}[2.5pt]{\class{Packet}}
+& \multirow{1}{*}[-0.05em]{\submodel{Trace}} &
 \begin{minipage}[l]{4.3in}
@@
 \cline{2-3}
-& \multirow{1}{*}[-0.05em]{Parametric} &
+& \multirow{1}{*}[-0.05em]{\submodel{Parametric}} &
 \begin{minipage}[l]{4.3in}
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals drawn from fitted Exponential distributions\raisebox{1pt}{\footnotesize*}
+\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals drawn from fitted Exponential distributions\raisebox{1.5pt}{\scriptsize*}
 \vspace{2pt}

### 2008-01-21 paper.tex
Redone intro; added paragraph to related work; etc.
In the process of overhauling the Methodology section.

@@
 
-\newcommand{\thetitle}{LAN Workload: The Failure of Uniform Modeling}
+\newcommand{\thetitle}{On the Inadequacy of Uniform Modeling for WLAN Workload Generation}
 
@@
 
-\title{
-\thetitle%\\
-%\vspace{0.2em}
-%\LARGE{On The Inadequacy of Uniform Modeling for WLAN Workloads}
-}
+\title{\thetitle}
 \author{
@@
 
-\newcommand{\Section}[1]{\S\ref{sec:#1}}
+\newcommand{\Section}[1]{\hyperref[sec:#1]{Section~\ref*{sec:#1}}}
+\newcommand{\FHC}{Hern\'andez-Campos~\textit{et~al.}}
+\newcommand{\class}[1]{\textsf{\small{#1}}}
+\newcommand{\submodel}[1]{\textsf{\small{#1}}}
+\newcommand{\model}[1]{#1}
+\newcommand{\RFC}[1]{\caps{RFC}~{#1}}
 
@@
 The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
-Other research~\cite{Hernandez06:dissertation} has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback (see \Section{related-work}) induced by the ubiquitous transmission control protocol~(\caps{TCP}).
+Other research~\cite{Hernandez06:dissertation} has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback induced by the ubiquitous transmission control protocol~(\caps{TCP}).
 Recently, the high-level statistical properties of wireless \caps{LAN} traffic have been rigorously modeled with a variety of parametric distributions~\cite{Hernandez06:wlan-traffic}.
@@
 Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily.
-While more complex approaches have been proposed~\cite{xxx}, the simplest, na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better.
-It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.% in exchange for their increased complexity.
+While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
+the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better.
+It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.
 
+To address this foundational problem in network traffic modeling, we have introduced a fundamentally different approach to evaluating the realism of workload models~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics.
+We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values similar to those observed using the original trace to generate traffic.
+Using this new approach, we explore the space of traffic models, establishing rigorously that the commonly used uniform models for node, flow and packet behavior drastically misrepresent important \caps{WLAN} performance metrics at all levels of the protocol stack.
+
+It is natural to compare these common, na\"ive models with the high-level parametric models proposed by \FHC~\cite{Hernandez06:wlan-traffic}.
+Here, however, we face another fundamental challenge in traffic modeling.
+Even with specific models for various aspects of network behavior, it remains non-trivial to combine them into a coherent composite model capable of generating concrete workloads.
+The {\FHC} models, for example, specify the distribution of flows per node in the network; since every flow has two end-points, however, it remains unclear how to assign end-points to flows such that the marginal distribution of flow per node matches the BiPareto model indicated.
+Similarly, the models dictate BiPareto marginal distribution of flow sizes across the entire network, but say nothing about how these sizes are to be distributed among the nodes.
+We could assume that the number of flows per node is independent from the size of the flows at that node, but, as we shall see, this assumption is invalid.
+Moreover, the details of flow behavior---packet sizes and inter-packet intervals---remain entirely undefined.
+
+Our point is not to denigrate the models proposed in~\cite{Hernandez06:wlan-traffic} in any way.
+On the contrary: these parametric models are the state of the art in modeling traffic patterns for wireless local area networks.
+The purpose, rather, is to point out how large a gap still remains between having accurate high-level models of marginal aspects of network behavior and being able to generate usable workloads with those models.
+Moreover, a great deal of ambiguity lies in that gap since there is no single obvious way to combine models.
+
+To clarify, disambiguate and standardize the specification of generative local area traffic models, we propose the \textit{Traffic Object Model} (\textit{\caps{TOM}}).
+The \caps{TOM} entails a hierarchy of abstract classes: \class{Node}, \class{Flow}, and \class{Packet} are the primary classes; the \class{Network} class collects everything.
+Each abstract class defines an interface of properties.
+A traffic model defines concrete classes at each level that provide the required interfaces; each class defines its own properties by delegation, inheritance, or overriding properties of other classes.
+Because of the separation of the \caps{TOM} into well-defined levels, each concrete class naturally defines a sub-model, specific to that level's behavior; sub-models can be mixed and matched combinatorially.
+
+We define a range of sub-models at each level of the \caps{TOM}, with decreasing similarity to the original trace behavior: \submodel{Trace} (reproduce original behavior), \submodel{Nonparametric}, \submodel{Parametric}, and \submodel{Uniform}.
+The control model uses the Trace sub-model at every level, exactly reproducing the behavior recorded from the original network.
+The standard na\"ive model uses the \submodel{Uniform} sub-model at all levels.
+The models proposed by {\FHC} are implemented using the \submodel{Parametric} sub-model at every level.
+The \submodel{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.
+
+%: objects and their properties together with rules for delegating, inheriting, and overriding the properties across the objects in the \caps{TOM} hierarchy. The \caps{TOM} allows
+
+%The {\FHC} models, like most network behavior models, are marginal, meaning that they are fitted to data aggregated of over all nodes in the network, and not fitted on a per-node basis.
+With marginal models, since all information about interactions between the models is lost, so we must apply each model independently and hope that the result is good enough.
+
+%Before we can define traffic models, however, we face the next major fundamental challenge in traffic modeling.
+
+%In this paper, we address this fundamental problem in network traffic modeling by presenting an 
 %In this paper, we propose objective criteria by which to judge whether
 
-The experimental evaluation of new wireless technologies inevitably requires network traffic.
+%The experimental evaluation of new wireless technologies inevitably requires network traffic.
 The most commonly used models for generating wireless local-area network (\caps{WLAN}) workload are simple, treating both nodes and flows uniformly, and are generally understood to be na\"ive.
@@
 
-One of the hardest problems in modeling phenomena as complex as network traffic patterns lies in determining which characteristics of any given exemplar of behavior are essential and which can safely be abstracted away.
+%One of the hardest problems in modeling phenomena as complex as network traffic patterns lies in determining which characteristics of any given exemplar of behavior are essential and which can safely be abstracted away.
 In \caps{WLAN} traffic modeling, important characteristics have largely been determined based on intuition.
@@
 
-\begin{table}[t]
-\vspace{0.7em}
-\begin{tabular}[t]{|l|}
-\hline
-\begin{minipage}[l]{3.315in}
-\begin{small}
-\vspace{0.6em}
-\input{properties.tex}
-\vspace{0.6em}
-\end{small}
-\end{minipage}
-\\\hline
-\end{tabular}
-\caption{Summary of classes and their attributes in the proposed Traffic Object Model ({\footnotesize{TOM}}) for {\footnotesize{WLAN}} workload generation.}
-\vspace{-2em}
-\end{table}
-
-\begin{table*}[t]
-\begin{center}
-\small
-\input{models.tex}
-\footnotesize\\[0.5em]
-\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
-\end{center}
-\vspace{-0.75em}
-\caption{Summary of how the models at each behavioral level override or inherit properties in the Traffic Object Model ({\footnotesize{TOM}}).}
-\vspace{-1.75em}
-\end{table*}
-
-Our analysis shows that the standard constant bit-rate (\caps{CBR}) traffic model, with uniformly chosen nodes and uniform flows of traffic, significantly misrepresents important performance metrics at every level of the network.
+%Our analysis shows that the standard constant bit-rate (\caps{CBR}) traffic model, with uniformly chosen nodes and uniform flows of traffic, significantly misrepresents important performance metrics at every level of the network.
 To qualify these results in more depth, we also investigate partially synthetic variants of this standard model, allowing us to reason about which aspects of workload models have the most impact.
@@
 
-The first major contribution of this work is the definition of ``sufficient realism,'' together with the analytical and statistical methodology to rigorously test whether synthetic traffic models meet this definition.
+%The first major contribution of this work is the definition of ``sufficient realism,'' together with the analytical and statistical methodology to rigorously test whether synthetic traffic models meet this definition.
 The second contribution is the conclusion that the most commonly used model of wireless traffic drastically and consistently misrepresents some of the most important metrics for performance evaluation of wireless protocols.
@@
 
+This work extends that published in~\cite{Karpinski07:realism}~and~\cite{Karpinski07:cbr-failure}, using the same methodology for defining and quantifying the realism of synthetic traffic models.
+The Traffic Object Model builds on the combinatorial approach taken in~\cite{Karpinski07:cbr-failure}, while providing an flexible and general object-oriented framework for constructing generative traffic models.
+The traffic models evaluated here overlap and augment those evaluated in both prior works; comparison with the models proposed in {\FHC} is entirely new.
+The Qualnet simulation methodology has been enhanced by the implementation of full-duplex, trace-driven raw \caps{IP}, \caps{TCP}, and \caps{UDP} flows, and performance metric collection based on the \caps{IETF} recommendations for \caps{IP} performance metrics framework~\cite{rfc:ip-metrics,rfc:jitter}.
+
 % TODO: add in the Hernandez-Cortes' dissertation.
@@
 \begin{samepage}
-\noindent\textbf{Definition.}~A model of user behavior is \textit{sufficiently realistic} if, when compared with actual user behavior, the model, with parameter values extracted from the real data, yields statistically equivalent performance results.
+\noindent\textbf{Definition.}~A model of user behavior is \textit{sufficiently realistic} if, compared with actual user behavior, the model, with parameter values extracted from the real data, yields statistically equivalent performance results.
 \end{samepage}
@@
 \begin{figure}
+\vspace{0.45em}
 \begin{center}
@@
 
-Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of synthetic traffic models, including the standard random, uniform \caps{CBR} model.
-For our analysis, we use a 24-hour trace recorded in an infrastructured 802.11g wireless \caps{LAN} with 18 access points, deployed at the 60th Internet Engineering Task Force meeting (\caps{IETF60}), held in San Diego during August of 2004.
+Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of synthetic traffic models defined using the Traffic Object Model (\caps{TOM}) framework.
+We use a 24-hour trace recorded in an infrastructured 802.11g wireless \caps{LAN} with 18 access points, deployed at the 60th Internet Engineering Task Force meeting (\caps{IETF}60), held in San Diego during August of 2004.
 The traffic trace was captured using \texttt{\small{tcpdump}} at a single router, through which all wireless traffic for the meeting was routed, including traffic between wireless nodes.
@@
 
-We do not assume or claim that the traffic found at \caps{IETF60} is representative of conference settings in general.
+We do not assume or claim that the traffic found at \caps{IETF}60 is representative of conference settings in general.
 The observed behaviors are also unlikely to resemble those found in a typical commercial or residential setting.
@@
 
-\subsection{Simulations}\label{sec:simulations}
+\begin{table}[t]
+\vspace{0.75em}
+\begin{center}
+\begin{tabular}[t]{|l|}
+\hline
+\begin{minipage}[l]{2.8in}
+\begin{small}
+\vspace{0.6em}
+\input{properties.tex}
+\vspace{0.6em}
+\end{small}
+\end{minipage}
+\\\hline
+\end{tabular}
+\end{center}
+\vspace{-0.5em}
+\caption{Classes and their properties in the Traffic Object Model.}
+\vspace{-2em}
+\end{table}
 
-We use the Qualnet wireless network simulator to perform our experiments.
-We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
-In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays.
-Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings.
-We do not attempt to reproduce the physical environment of the original wireless network, nor do we simulate mobility.
-The only aspect of the original network's behavior that is reproduced is the total pattern of network-wide traffic.
+\subsection{The Traffic Object Model~(\caps{TOM})}\label{sec:models}
 
-There are a number of potential objections to this approach.
-We use single-hop trace data to drive multi-hop simulations; the physical environment, node mobility, handover behavior, and closed-loop dynamics of the original wireless setting are not faithfully reproduced.
-One must keep in mind, however, that the goal of this research is \textit{not} to understand the conditions of the original network.
-Rather, we are using the traffic behaviors observed as examples to help us better understand how different types of workload can affect performance metrics.
-In particular, we aim to understand how real workload compares with common synthetic traffic models.
-Of course, the reason for such objections is that networking researchers understand that the many aspects of behavior interact with each other in a complex and nearly inextricable manner.
-However, before we can hope to understand the interaction between workload and other features affecting network behavior, we must study traffic patterns alone, and learn to model them with reasonable accuracy in the absence of additional complicating factors.
-Accordingly, in this study, we detach application level traffic patterns from the other factors influencing network conditions, and study them in isolation.
-
-The 24-hour trace is split into 144 10-minute segments, each of which serves as the basis for a set of simulations using different traffic models.
-The traffic models range from a completely realistic trace-driven model, to a standard \caps{CBR} traffic model.
-Various partially synthetic intermediate models, described in \Section{models-metrics}, are simulated to study the impact of different aspects of traffic behavior on network performance.
-To preserve the fairness of the performance comparison, we keep as many features as possible constant across different traffic models.
-The traffic generated by each synthetic model preserves as many characteristics from the original trace as possible, within the constraints of the model.
-Moreover, the following features are preserved across all models: the numbers of wireless nodes, the number of flows, the number of application-initiated data units sent, the total bytes of application data sent, and the average flow duration (and therefore the average data rate).
-
-\subsection{Traffic Models \& Performance Metrics}\label{sec:models-metrics}
-
-\begin{figure*}
+\begin{table*}[t]
 \begin{center}
-\begin{minipage}[c]{0.67\linewidth}
-\subfloat[Flow Topology: Trace]
-{\includegraphics[width=1.55in]{flow_topology_trace}\label{fig:flow-topology-trace}}
-\subfloat[Flow Topology: Sink]
-{\includegraphics[width=1.55in]{flow_topology_sink}\label{fig:flow-topology-sink}}
-\subfloat[Flow Topology: Uniform]
-{\includegraphics[width=1.55in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
-\subfloat[Flow Behavior: Trace]
-{\includegraphics[width=1.55in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
-\subfloat[Flow Behavior: Uniform]
-{\includegraphics[width=1.55in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
-\subfloat[Packet Behavior]
-{\includegraphics[width=1.55in]{flow_packet_behavior}\label{fig:packet-behavior}}
-\end{minipage}\hfill
-\begin{minipage}[c]{0.30\linewidth}
-\caption{Examples illustrating the different traffic models for the three levels of behavior.
-Figures \ref{fig:flow-topology-trace}, \ref{fig:flow-topology-sink} and \ref{fig:flow-topology-uniform}, show example flow topologies.
-The width of each line is proportional to the logarithm of the number of flows between the nodes it connects.
-In each graph, node zero is the gateway to the Internet.
-Uniform and trace flow behavior examples are plotted in Figures \ref{fig:flow-behavior-trace} and \ref{fig:flow-behavior-uniform}.
-The time axis indicates when the various flows start and end; the width of each flow line is proportional to the logarithm of its data rate.
-Flow numbers are assigned arbitrarily.
-Figure \ref{fig:packet-behavior} compares packet behavior for the uniform model (i.e.
-{\footnotesize{CBR}}), with the trace of an actual flow.
-In the uniform model, the cumulative data sent increases smoothly over time (gray diagonal line).
-In the actual packet trace, the data transmissions are variable both in size and in inter-transmission interval, leading to a ``lumpy'' cumulative data history (black step-function).}
-\label{fig:traffic-models}
-\end{minipage}
+\small
+\input{models.tex}
+\footnotesize\\[0.5em]
+\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
 \end{center}
-\vspace{-1em}
-\end{figure*}
+\vspace{-0.75em}
+\caption{Summary of how the sub-models at each behavioral level of the {\footnotesize{TOM}} delegate, override or inherit properties.}
+\vspace{-1.75em}
+\end{table*}
 
@@
 
+\begin{figure*}[tb]
+\begin{center}
+\subfloat[trace node behavior]
+{\includegraphics[width=1.392in]{flow_topology_trace}\label{fig:flow-topology-trace}}
+\subfloat[uniform node behavior]
+{\includegraphics[width=1.392in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
+\subfloat[trace flow behavior]
+{\includegraphics[width=1.392in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
+\subfloat[uniform flow behavior]
+{\includegraphics[width=1.392in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
+\subfloat[packet behavior]
+{\includegraphics[width=1.392in]{flow_packet_behavior}\label{fig:packet-behavior}}
+\caption{Examples illustrating trace versus uniform behaviors a the node, flow and packet levels.
+Figures \ref{fig:flow-topology-trace} and \ref{fig:flow-topology-uniform} show example node behaviors.
+The width of each line is proportional to the logarithm of the number of flows between the nodes it connects (zero is the Internet gateway).
+Uniform and trace flow behavior examples are plotted in Figures \ref{fig:flow-behavior-trace} and \ref{fig:flow-behavior-uniform}.
+The time axis indicates when flows start and end; the width of each flow line is proportional to the logarithm of its data rate.
+Figure \ref{fig:packet-behavior} compares \caps{CBR} packet behavior with the trace of an actual flow.
+In the uniform model, the cumulative data sent increases smoothly over time, whereas in the actual packet trace, the transmissions are variable both in size and in inter-packet interval, leading to a ``lumpy'' cumulative data plot.}
+\label{fig:traffic-models}
+\end{center}
+\vspace{-2em}
+\end{figure*}
+
 %\begin{enumerate}
@@
 This eliminates three combinations of models and leaves nine viable behaviors, abbreviated by the first letters of their flow topology, flow behavior, and packet behavior models: \caps{TTT} (fully trace-based), \caps{TTU}, \caps{TUU}, \caps{STT}, \caps{STU}, \caps{SUU} (entirely synthetic, sink topology), \caps{UTT}, \caps{UTU}, \caps{UUU} (entirely synthetic, uniform topology).
+
+\subsection{Performance Metrics}\label{sec:metrics}
 
@@
 
+\subsection{Simulations}\label{sec:simulations}
+
+We use the Qualnet wireless network simulator to perform our experiments.
+We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
+In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays.
+Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings.
+We do not attempt to reproduce the physical environment of the original wireless network, nor do we simulate mobility.
+The only aspect of the original network's behavior that is reproduced is the total pattern of network-wide traffic.
+
+There are a number of potential objections to this approach.
+We use single-hop trace data to drive multi-hop simulations; the physical environment, node mobility, handover behavior, and closed-loop dynamics of the original wireless setting are not faithfully reproduced.
+One must keep in mind, however, that the goal of this research is \textit{not} to understand the conditions of the original network.
+Rather, we are using the traffic behaviors observed as examples to help us better understand how different types of workload can affect performance metrics.
+In particular, we aim to understand how real workload compares with common synthetic traffic models.
+Of course, the reason for such objections is that networking researchers understand that the many aspects of behavior interact with each other in a complex and nearly inextricable manner.
+However, before we can hope to understand the interaction between workload and other features affecting network behavior, we must study traffic patterns alone, and learn to model them with reasonable accuracy in the absence of additional complicating factors.
+Accordingly, in this study, we detach application level traffic patterns from the other factors influencing network conditions, and study them in isolation.
+
+The 24-hour trace is split into 144 10-minute segments, each of which serves as the basis for a set of simulations using different traffic models.
+The traffic models range from a completely realistic trace-driven model, to a standard \caps{CBR} traffic model.
+Various partially synthetic intermediate models, described in \Section{models}, are simulated to study the impact of different aspects of traffic behavior on network performance.
+To preserve the fairness of the performance comparison, we keep as many features as possible constant across different traffic models.
+The traffic generated by each synthetic model preserves as many characteristics from the original trace as possible, within the constraints of the model.
+Moreover, the following features are preserved across all models: the numbers of wireless nodes, the number of flows, the number of application-initiated data units sent, the total bytes of application data sent, and the average flow duration (and therefore the average data rate).
+
 \subsection{Measures of Error}\label{sec:error-measures}
 
-The simulations described in \Section{models-metrics} provide us with the raw data to compare performance metrics for synthetic traffic models with those for real traffic traces.
+The simulations described in \Section{simulations} provide us with the raw data to compare performance metrics for synthetic traffic models with those for real traffic traces.
 To assess the realism of these models, however, we need a measure of how inaccurate the synthetic performance values are when compared to the real values.
@@
 
-The most significant limitation on the generality of this analysis is that it is based entirely on a single data set from \caps{IETF60}---albeit a large and varied one.
+The most significant limitation on the generality of this analysis is that it is based entirely on a single data set from \caps{IETF}60---albeit a large and varied one.
 It is possible that traffic in this trace happens to produce network performance that is unusually dissimilar to standard traffic models.

### 2008-01-21 models.rb

@@ -14,7 +14,7 @@ fixups = {
 puts <<'HEADER'
 \begin{tabular}{|c|c|p{4.3in}|}
 \multicolumn{1}{c}{\textbf{Level}} &
-\multicolumn{1}{c}{\textbf{Model}} &
+\multicolumn{1}{c}{\textbf{Sub-model}} &
 \multicolumn{1}{c}{\textbf{Overridden and inherited properties}} \\[2pt]
 HEADER
 models.each do |l|
@@ -22,17 +22,17 @@ models.each do |l|
   mods = l[level]
   fixup = fixups[level.to_sym]
   puts '\hline'
-  puts "\\multirow{#{fixup.first}}{*}[#{fixup.last}]{#{level}}"
+  puts "\\multirow{#{fixup.first}}{*}[#{fixup.last}]{\\class{#{level}}}"
   mods.each_with_index do |m,j|
     model = m.keys.first
     items = m[model]
     items = [items] unless items.is_a?(Array)
-    puts "& \\multirow{1}{*}[-0.05em]{#{model}} &"
+    puts "& \\multirow{1}{*}[-0.05em]{\\submodel{#{model}}} &"
     puts '\begin{minipage}[l]{4.3in}'
     puts '\vspace{2pt}'
     items.each_with_index do |item,k|
       item.gsub!(/&/,'\\\&')
-      item.gsub!(/\*\s*$/,'\raisebox{1pt}{\footnotesize*}')
+      item.gsub!(/\*\s*$/,'\raisebox{1.5pt}{\scriptsize*}')
       line_end = (k < items.size-1) ? ' \\\\' : ''
       puts "#{bullet} #{item}" + line_end
     end

### 2008-01-21 models.tex
The Traffic Object Model section is written now.
Had to rearrange the properties and models tables.

@@
 \vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} session start time assigned from trace \\
 \raisebox{1.5pt}{$\centerdot$} each node corresponds to a specific trace node with associated trace flows \\
-\raisebox{1.5pt}{$\centerdot$} session start time assigned from trace \\
 \raisebox{1.5pt}{$\centerdot$} overrides all distributions with node-specific versions
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} generic flows assigned to nodes by sampling [TODO---more detail] \\
-\raisebox{1.5pt}{$\centerdot$} session start time randomly assigned from pool of trace session start times
+\raisebox{1.5pt}{$\centerdot$} session start time randomly assigned from pool of trace session start times \\
+\raisebox{1.5pt}{$\centerdot$} nodes assigned as flow end-points by sampling from inherited src/dst distributions
 \vspace{2pt}
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} generic flows assigned to nodes according to fitted BiPareto model\raisebox{1.5pt}{\scriptsize*} \\
-\raisebox{1.5pt}{$\centerdot$} session start time assigned according to Poisson arrival model\raisebox{1.5pt}{\scriptsize*}
+\raisebox{1.5pt}{$\centerdot$} session start time assigned according to Poisson arrival model\raisebox{1.5pt}{\scriptsize*} \\
+\raisebox{1.5pt}{$\centerdot$} nodes assigned as flow end-points by sampling from fitted BiPareto model\raisebox{1.5pt}{\scriptsize*}
 \vspace{2pt}
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} all nodes have average number of generic flows \\
-\raisebox{1.5pt}{$\centerdot$} session start times uniformly cover viable range
+\raisebox{1.5pt}{$\centerdot$} session start times selected uniformly from viable range \\
+\raisebox{1.5pt}{$\centerdot$} nodes have the mean number of generic flows
 \vspace{2pt}
@@
 \hline
-\multirow{11}{*}[2.5pt]{\class{Flow}}
+\multirow{14}{*}[2.5pt]{\class{Flow}}
 & \multirow{1}{*}[-0.05em]{\submodel{Trace}} &
@@
 \vspace{2pt}
+\raisebox{1.5pt}{$\centerdot$} flow start time and size taken from trace \\
 \raisebox{1.5pt}{$\centerdot$} each flow corresponds to a specific trace flow with an associated packet trace \\
-\raisebox{1.5pt}{$\centerdot$} flow size and start time taken from trace \\
-\raisebox{1.5pt}{$\centerdot$} overrides distributions of size/interval packet behaviors with flow-specific versions
+\raisebox{1.5pt}{$\centerdot$} overrides distributions of size/interval packet behaviors with flow-specific versions \\
+\raisebox{1.5pt}{$\centerdot$} packet stream from trace file
 \vspace{2pt}
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} flow size is sampled from inherited distribution of flow sizes \\
-\raisebox{1.5pt}{$\centerdot$} start time relative to previous session flow sampled from inherited distribution \\
-\raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node
+\raisebox{1.5pt}{$\centerdot$} start time relative to previous flow sampled from inherited distribution \\
+\raisebox{1.5pt}{$\centerdot$} flow size sampled from inherited distribution of flow sizes \\
+\raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node \\
+\raisebox{1.5pt}{$\centerdot$} generic packets generated until flow size exhausted
 \vspace{2pt}
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} flow size is sampled from fitted BiPareto model\raisebox{1.5pt}{\scriptsize*} \\
-\raisebox{1.5pt}{$\centerdot$} start time relative to previous flow in session sampled from fitted Lognormal model\raisebox{1.5pt}{\scriptsize*} \\
-\raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node
+\raisebox{1.5pt}{$\centerdot$} start time relative to previous flow sampled from fitted Lognormal model\raisebox{1.5pt}{\scriptsize*} \\
+\raisebox{1.5pt}{$\centerdot$} flow size sampled from fitted BiPareto model\raisebox{1.5pt}{\scriptsize*} \\
+\raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node \\
+\raisebox{1.5pt}{$\centerdot$} generic packets generated until flow size exhausted
 \vspace{2pt}
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} all flows have size equal to inherited mean \\
-\raisebox{1.5pt}{$\centerdot$} average inter-flow interval used between all flows \\
-\raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node
+\raisebox{1.5pt}{$\centerdot$} start time relative to previous flow equals mean inherited inter-flow interval \\
+\raisebox{1.5pt}{$\centerdot$} all flows have size equal to inherited mean flow size \\
+\raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node \\
+\raisebox{1.5pt}{$\centerdot$} generic packets generated until flow size exhausted
 \vspace{2pt}

### 2008-01-21 paper.tex
The Traffic Object Model section is written now.
Had to rearrange the properties and models tables.

@@
 \newcommand{\FHC}{Hern\'andez-Campos~\textit{et~al.}}
-\newcommand{\class}[1]{\textsf{\small{#1}}}
-\newcommand{\submodel}[1]{\textsf{\small{#1}}}
+\newcommand{\class}[1]{\textsf{\footnotesize{#1}}}
+\newcommand{\submodel}[1]{\textsf{\footnotesize{#1}}}
 \newcommand{\model}[1]{#1}
@@
 
-\begin{figure}
-\vspace{0.45em}
-\begin{center}
-\includegraphics[width=3.5in]{nodes-flows}%
-\vspace{-0.75em}%
-\caption{The number of active nodes and flows over time.} 
-\label{fig:nodes-flows}
-\end{center}
-\vspace{-2em}
-\end{figure}
-
 Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of synthetic traffic models defined using the Traffic Object Model (\caps{TOM}) framework.
@@
 
-%\begin{table*}
-%\begin{center}\small
-%\input{models.tex}
-%\caption{The three levels of traffic behavior and the traffic models defined for each level.}\label{tab:traffic-models}
-%\end{center}
-%\vspace{-1.75em}
-%\end{table*}
+\begin{figure}
+\vspace{0.45em}
+\begin{center}
+\includegraphics[width=3.5in]{nodes-flows}%
+\vspace{-0.75em}%
+\caption{The number of active nodes and flows over time.} 
+\label{fig:nodes-flows}
+\end{center}
+\vspace{-2em}
+\end{figure}
 
@@
 \hline
-\begin{minipage}[l]{2.8in}
+\begin{minipage}[l]{2.7in}
 \begin{small}
@@
 \input{properties.tex}
-\vspace{0.6em}
+%\vspace{0.6em}
 \end{small}
@@
 \vspace{-0.5em}
-\caption{Classes and their properties in the Traffic Object Model.}
+\caption{Classes and their attributes in the Traffic Object Model.}
 \vspace{-2em}
@@
 
-\subsection{The Traffic Object Model~(\caps{TOM})}\label{sec:models}
+\subsection{The Traffic Object Model~(\caps{TOM})}\label{sec:tom}
 
@@
 \vspace{-0.75em}
+\label{fig:properties}
 \caption{Summary of how the sub-models at each behavioral level of the {\footnotesize{TOM}} delegate, override or inherit properties.}
@@
 \end{table*}
+
+In this section, we present a generic object-oriented framework for implementing traffic generation models.
+This framework is called the Traffic Object Model (\caps{TOM}).
+It defines four levels of responsibility: network, node, flow and packet.
+The framework is designed so that different behaviors at each level can be naturally implemented  as classes in object-oriented programming languages.
+The framework automatically gains the advantages of object orientation, in particular modularity and encapsulation.
+Every traffic model is naturally split into separate spheres of responsibility, which constitute sub-models for the aspects of the model's behavior at each layer.
+Thanks to encapsulation and polymorphism of objects, the sub-models from different models can immediately be recombined to create new hybrid behaviors.
+It is trivial, for example, to produce trace-based node behavior, coupled with parametric flow behavior, and uniform packet behavior.
+Not all combinations of sub-models are valid, however; which classes may be combined depends on the details of each class.
+The ability to mix-and-match aspects of behavior at different levels greatly enhances our ability to explore the space of possible traffic models.
+
+%For example, to use uniform flow behavior but non-uniform, trace-based node behavior, simply mix the the uniform implementation of the
+
+Abstract base classes define the properties at each level that traffic models must implement.\footnote{In Java or C\#, this role may be played either by an actual abstract class or by an interface.
+In duck-typed languages such as Ruby or Python, no actual abstract class need exist, since type is implicitly defined by usage; if inheritance is desired, the base class can be made abstract by defining stub methods that throw informative exceptions if called.} Figure~\ref{fig:properties} enumerates all the properties exposed by traffic models used in this work.
+Other traffic models may necessitate exposing additional properties, but this list suffices to implement the models used in this paper.
+Concrete classes implement the behavior of specific traffic models at each level.
+A complete synthetic workload is created as follows:
+\begin{enumerate}
+\item instantiate a \class{Network} object of the appropriate class;
+\item the top-level \class{Network} object in turn instantiates a collection of \class{Node} objects and associated \class{Flow} objects;
+\item each \class{Flow} object generates a stream of \class{Packet} objects.
+\end{enumerate}
+The hierarchy of objects may be used to directly drive a simulation, or may generate configuration files that in turn drive a simulation.
+Our implementation does the latter, generating a collection of Qualnet configuration files and trace files specific to each workload instantiation (see \Section{simulations}).
+
+\subsection{Traffic Models}\label{sec:models}
 

### 2008-01-21 properties.tex
Finally tweaked the models & properties tables to satisfaction.
The Traffic Object Model section is written now.
Had to rearrange the properties and models tables.

@@
-\noindent
-\textbf{Network:}
-\begin{itemize}
-  \item collection of nodes
-  \item collection of flows
-  \item distribution of flow sizes
-  \item distribution of intervals between flow start times
-  \item distribution of packet sizes
-  \item distribution of inter-packet intervals
-\end{itemize}
-\noindent
-\textbf{Node:}
-\begin{itemize}
-  \item collection of source flows
-  \item collection of destination flows
-  \item session start time
-  \item distribution of flow sizes
-  \item distribution of intervals between flow start times
-  \item distribution of packet sizes
-  \item distribution of inter-packet intervals
-\end{itemize}
-\noindent
-\textbf{Flow:}
-\begin{itemize}
-  \item source
-  \item destination
-  \item start time
-  \item size (byte count)
-  \item distribution of packet sizes
-  \item distribution of inter-packet intervals
-\end{itemize}
-\noindent
-\textbf{Packet:}
-\begin{itemize}
-  \item send time
-  \item payload size (bytes)
-\end{itemize}
+\textbf{\submodel{Network:}}\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~collection of nodes\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~collection of flows\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of flow sizes\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of flow-source counts\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of flow-destination counts\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of intervals between flow start times\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of packet sizes\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of inter-packet intervals\\
+\textbf{\submodel{Node:}}\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~session start time\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~collection of source flows\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~collection of destination flows\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of flow sizes\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of intervals between flow start times\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of packet sizes\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of inter-packet intervals\\
+\textbf{\submodel{Flow:}}\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~start time\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~source node\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~destination node\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~stream of packets\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of packet sizes\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~distribution of inter-packet intervals\\
+\textbf{\submodel{Packet:}}\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~send time\\
+\hspace*{1ex}\raisebox{1.5pt}{$\centerdot$}~payload size (bytes)\\

### 2008-01-21 models.rb

@@ -7,7 +7,7 @@ models = YAML.load_file("models.yaml")
 bullet = '\\raisebox{1.5pt}{$\\centerdot$}'
 fixups = {
   :Node   => [ 8,'2.5pt'],
-  :Flow   => [11,'2.5pt'],
+  :Flow   => [14,'2.5pt'],
   :Packet => [ 5,'2.5pt']
 }
 

### 2008-01-21 properties.rb

@@ -4,13 +4,13 @@ require 'yaml'
 
 properties = YAML.load_file("properties.yaml")
 
+bullet = '\\raisebox{1.5pt}{$\\centerdot$}'
+
 properties.each do |l|
   level = l.keys.first
   props = l[level]
-  puts "\\textbf{#{level}:}"
-  puts '\begin{itemize}'
+  puts "\\textbf{\\submodel{#{level}:}}\\\\"
   props.each do |prop|
-    puts "  \\item #{prop}"
+    puts "\\hspace*{1ex}#{bullet}~#{prop}\\\\"
   end
-  puts '\end{itemize}'
 end

### 2008-01-21 models.tex
Fixed up formatting of TOM classes.

@@
 \begin{tabular}{|c|c|p{4.3in}|}
-\multicolumn{1}{c}{\textbf{Level}} &
-\multicolumn{1}{c}{\textbf{Sub-model}} &
+\multicolumn{1}{c}{\textbf{Class}} &
+\multicolumn{1}{c}{\textbf{Subclass}} &
 \multicolumn{1}{c}{\textbf{Overridden and inherited properties}} \\[2pt]
@@
 \multirow{8}{*}[2.5pt]{\class{Node}}
-& \multirow{1}{*}[-0.05em]{\submodel{Trace}} &
+& \multirow{1}{*}[-0.05em]{\class{Trace}} &
 \begin{minipage}[l]{4.3in}
@@
 \cline{2-3}
-& \multirow{1}{*}[-0.05em]{\submodel{Nonparametric}} &
+& \multirow{1}{*}[-0.05em]{\class{Nonparametric}} &
 \begin{minipage}[l]{4.3in}
@@
 \cline{2-3}
-& \multirow{1}{*}[-0.05em]{\submodel{Parametric}} &
+& \multirow{1}{*}[-0.05em]{\class{Parametric}} &
 \begin{minipage}[l]{4.3in}
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} session start time assigned according to Poisson arrival model\raisebox{1.5pt}{\scriptsize*} \\
-\raisebox{1.5pt}{$\centerdot$} nodes assigned as flow end-points by sampling from fitted BiPareto model\raisebox{1.5pt}{\scriptsize*}
+\raisebox{1.5pt}{$\centerdot$} session start time assigned according to Poisson arrival model \\
+\raisebox{1.5pt}{$\centerdot$} nodes assigned as flow end-points by sampling from fitted BiPareto model
 \vspace{2pt}
@@
 \cline{2-3}
-& \multirow{1}{*}[-0.05em]{\submodel{Uniform}} &
+& \multirow{1}{*}[-0.05em]{\class{Uniform}} &
 \begin{minipage}[l]{4.3in}
@@
 \multirow{14}{*}[2.5pt]{\class{Flow}}
-& \multirow{1}{*}[-0.05em]{\submodel{Trace}} &
+& \multirow{1}{*}[-0.05em]{\class{Trace}} &
 \begin{minipage}[l]{4.3in}
@@
 \cline{2-3}
-& \multirow{1}{*}[-0.05em]{\submodel{Parametric}} &
+& \multirow{1}{*}[-0.05em]{\class{Parametric}} &
 \begin{minipage}[l]{4.3in}
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} start time relative to previous flow sampled from fitted Lognormal model\raisebox{1.5pt}{\scriptsize*} \\
-\raisebox{1.5pt}{$\centerdot$} flow size sampled from fitted BiPareto model\raisebox{1.5pt}{\scriptsize*} \\
+\raisebox{1.5pt}{$\centerdot$} start time relative to previous flow sampled from fitted Lognormal model \\
+\raisebox{1.5pt}{$\centerdot$} flow size sampled from fitted BiPareto model \\
 \raisebox{1.5pt}{$\centerdot$} inherits distributions of size/interval packet behaviors from destination node \\
@@
 \multirow{5}{*}[2.5pt]{\class{Packet}}
-& \multirow{1}{*}[-0.05em]{\submodel{Trace}} &
+& \multirow{1}{*}[-0.05em]{\class{Trace}} &
 \begin{minipage}[l]{4.3in}
@@
 \cline{2-3}
-& \multirow{1}{*}[-0.05em]{\submodel{Parametric}} &
+& \multirow{1}{*}[-0.05em]{\class{Parametric}} &
 \begin{minipage}[l]{4.3in}
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals drawn from fitted Exponential distributions\raisebox{1.5pt}{\scriptsize*}
+\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals drawn from fitted Exponential distributions
 \vspace{2pt}

### 2008-01-21 paper.tex
Fixed up formatting of TOM classes.

@@
 
-\newcommand{\thetitle}{On the Inadequacy of Uniform Modeling for WLAN Workload Generation}
+\newcommand{\thetitle}{On the Inadequacy of Uniform Modeling for Wireless LAN Workload Generation}
 
@@
 \newcommand{\FHC}{Hern\'andez-Campos~\textit{et~al.}}
-\newcommand{\class}[1]{\textsf{\footnotesize{#1}}}
-\newcommand{\submodel}[1]{\textsf{\footnotesize{#1}}}
+\newcommand{\class}[1]{\textsc{\small{#1}}}
 \newcommand{\model}[1]{#1}
@@
 
-We define a range of sub-models at each level of the \caps{TOM}, with decreasing similarity to the original trace behavior: \submodel{Trace} (reproduce original behavior), \submodel{Nonparametric}, \submodel{Parametric}, and \submodel{Uniform}.
+We define a range of sub-models at each level of the \caps{TOM}, with decreasing similarity to the original trace behavior: \class{Trace} (reproduce original behavior), \class{Nonparametric}, \class{Parametric}, and \class{Uniform}.
 The control model uses the Trace sub-model at every level, exactly reproducing the behavior recorded from the original network.
-The standard na\"ive model uses the \submodel{Uniform} sub-model at all levels.
-The models proposed by {\FHC} are implemented using the \submodel{Parametric} sub-model at every level.
-The \submodel{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.
+The standard na\"ive model uses the \class{Uniform} sub-model at all levels.
+The models proposed by {\FHC} are implemented using the \class{Parametric} sub-model at every level.
+The \class{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.
 
@@
 The traffic models evaluated here overlap and augment those evaluated in both prior works; comparison with the models proposed in {\FHC} is entirely new.
-The Qualnet simulation methodology has been enhanced by the implementation of full-duplex, trace-driven raw \caps{IP}, \caps{TCP}, and \caps{UDP} flows, and performance metric collection based on the \caps{IETF} recommendations for \caps{IP} performance metrics framework~\cite{rfc:ip-metrics,rfc:jitter}.
+The Qualnet simulation methodology has been enhanced by the implementation of full-duplex, client-server, trace-driven raw \caps{IP}, \caps{TCP}, and \caps{UDP} flows.
+We have also implemented performance metric aggregation based on the \caps{IETF} framework for \caps{IP} performance metrics~\cite{rfc:ip-metrics,rfc:jitter}.
 
@@
 \begin{small}
-\vspace{0.6em}
+\vspace{0.7em}
 \input{properties.tex}
@@
 \vspace{-0.5em}
-\caption{Classes and their attributes in the Traffic Object Model.}
+\caption{{\footnotesize{TOM}} base classes and their attributes.}
+\label{tab:properties}
 \vspace{-2em}
@@
 \footnotesize\\[0.5em]
-\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
+%\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
 \end{center}
 \vspace{-0.75em}
-\label{fig:properties}
-\caption{Summary of how the sub-models at each behavioral level of the {\footnotesize{TOM}} delegate, override or inherit properties.}
+\caption{Details of how the sub-models at each behavioral level of the {\footnotesize{TOM}} inherit, override and delegate properties.}
+\label{tab:models}
 \vspace{-1.75em}
@@
 Every traffic model is naturally split into separate spheres of responsibility, which constitute sub-models for the aspects of the model's behavior at each layer.
-Thanks to encapsulation and polymorphism of objects, the sub-models from different models can immediately be recombined to create new hybrid behaviors.
+Encapsulation and polymorphism of objects, allows us to mix-and-match the sub-models from different models to create new hybrid behaviors.
 It is trivial, for example, to produce trace-based node behavior, coupled with parametric flow behavior, and uniform packet behavior.
 Not all combinations of sub-models are valid, however; which classes may be combined depends on the details of each class.
-The ability to mix-and-match aspects of behavior at different levels greatly enhances our ability to explore the space of possible traffic models.
-
-%For example, to use uniform flow behavior but non-uniform, trace-based node behavior, simply mix the the uniform implementation of the
 
 Abstract base classes define the properties at each level that traffic models must implement.\footnote{In Java or C\#, this role may be played either by an actual abstract class or by an interface.
-In duck-typed languages such as Ruby or Python, no actual abstract class need exist, since type is implicitly defined by usage; if inheritance is desired, the base class can be made abstract by defining stub methods that throw informative exceptions if called.} Figure~\ref{fig:properties} enumerates all the properties exposed by traffic models used in this work.
+In duck-typed languages such as Ruby or Python, no actual abstract class need exist, since type is implicitly defined by usage; if inheritance is desired, the base class can be made abstract by defining stub methods that throw informative exceptions if called.}
+Table~\ref{tab:properties} enumerates all the properties exposed by traffic models used in this work.
 Other traffic models may necessitate exposing additional properties, but this list suffices to implement the models used in this paper.
 Concrete classes implement the behavior of specific traffic models at each level.
-A complete synthetic workload is created as follows:
+A complete synthetic workload is instantiated as follows:
 \begin{enumerate}
-\item instantiate a \class{Network} object of the appropriate class;
+\item create a \class{Network} object of the appropriate class;
 \item the top-level \class{Network} object in turn instantiates a collection of \class{Node} objects and associated \class{Flow} objects;
@@
 \subsection{Traffic Models}\label{sec:models}
+
+For each of the \class{Node}, \class{Flow} and \class{Packet} abstract classes, we create define four concrete subclasses: \class{Trace}, \class{Nonparametric}, \class{Parametric} and \class{Uniform}.
+The \class{Trace} class at each level takes its properties directly from the behavior of the original traffic.
+The \class{Nonparametric} and \class{Parametric} classes use nonparametric models of marginal behavior at each level to
 

### 2008-01-21 models.rb

@@ -13,8 +13,8 @@ fixups = {
 
 puts <<'HEADER'
 \begin{tabular}{|c|c|p{4.3in}|}
-\multicolumn{1}{c}{\textbf{Level}} &
-\multicolumn{1}{c}{\textbf{Sub-model}} &
+\multicolumn{1}{c}{\textbf{Class}} &
+\multicolumn{1}{c}{\textbf{Subclass}} &
 \multicolumn{1}{c}{\textbf{Overridden and inherited properties}} \\[2pt]
 HEADER
 models.each do |l|
@@ -27,7 +27,7 @@ models.each do |l|
     model = m.keys.first
     items = m[model]
     items = [items] unless items.is_a?(Array)
-    puts "& \\multirow{1}{*}[-0.05em]{\\submodel{#{model}}} &"
+    puts "& \\multirow{1}{*}[-0.05em]{\\class{#{model}}} &"
     puts '\begin{minipage}[l]{4.3in}'
     puts '\vspace{2pt}'
     items.each_with_index do |item,k|

### 2008-01-21 properties.rb

@@ -6,10 +6,11 @@ properties = YAML.load_file("properties.yaml")
 
 bullet = '\\raisebox{1.5pt}{$\\centerdot$}'
 
-properties.each do |l|
+properties.each_with_index do |l,i|
+  puts '[3pt]' if i > 0
   level = l.keys.first
   props = l[level]
-  puts "\\textbf{\\submodel{#{level}:}}\\\\"
+  puts "\\textbf{\\class{#{level}:}}\\\\"
   props.each do |prop|
     puts "\\hspace*{1ex}#{bullet}~#{prop}\\\\"
   end

### 2008-01-22 models.tex
Finished updating the methodology.

@@
 \raisebox{1.5pt}{$\centerdot$} session start time randomly assigned from pool of trace session start times \\
-\raisebox{1.5pt}{$\centerdot$} nodes assigned as flow end-points by sampling from inherited src/dst distributions
+\raisebox{1.5pt}{$\centerdot$} nodes assigned as flow end-points by sampling from inherited distributions
 \vspace{2pt}
@@
 \end{minipage} \\
+\hline
 \hline
@@
 \hline
+\hline
 \multirow{5}{*}[2.5pt]{\class{Packet}}
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals drawn from fitted Exponential distributions
+\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals sampled from fitted Exponential distributions
 \vspace{2pt}

### 2008-01-22 paper.tex
Finished updating the methodology.

@@
 The standard na\"ive model uses the \class{Uniform} sub-model at all levels.
-The models proposed by {\FHC} are implemented using the \class{Parametric} sub-model at every level.
+The models proposed by {\FHC} are implemented by applying the \class{Parametric} sub-models at all levels.
 The \class{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.
@@
 \label{tab:models}
-\vspace{-1.75em}
+\vspace{-1em}
 \end{table*}
@@
 
-Abstract base classes define the properties at each level that traffic models must implement.\footnote{In Java or C\#, this role may be played either by an actual abstract class or by an interface.
-In duck-typed languages such as Ruby or Python, no actual abstract class need exist, since type is implicitly defined by usage; if inheritance is desired, the base class can be made abstract by defining stub methods that throw informative exceptions if called.}
+% TODO: fix-up client-server vs. source-destination terminology.
+
+Abstract base classes define the properties at each level that traffic models must implement.\footnote{In Java or C\#, this role may be served either by an actual abstract class or by an interface.
+In duck-typed languages such as Ruby or Python, no actual abstract class need exist, since type is implicitly defined by usage; if abstract classes are desired, they can be simulated by creating base classes with stub methods that throw informative exceptions upon being called.}
 Table~\ref{tab:properties} enumerates all the properties exposed by traffic models used in this work.
-Other traffic models may necessitate exposing additional properties, but this list suffices to implement the models used in this paper.
-Concrete classes implement the behavior of specific traffic models at each level.
+Future traffic models may necessitate exposing additional properties, but this list suffices to implement the models used in this paper.
+Concrete subclasses implement the behavior of specific traffic models at each level.
 A complete synthetic workload is instantiated as follows:
 \begin{enumerate}
-\item create a \class{Network} object of the appropriate class;
-\item the top-level \class{Network} object in turn instantiates a collection of \class{Node} objects and associated \class{Flow} objects;
+\item create a \class{Network} object of the appropriate subclass;
+\item the top-level \class{Network} object in turn creates a collection of \class{Node} objects and associated \class{Flow} objects;
 \item each \class{Flow} object generates a stream of \class{Packet} objects.
@@
 
-For each of the \class{Node}, \class{Flow} and \class{Packet} abstract classes, we create define four concrete subclasses: \class{Trace}, \class{Nonparametric}, \class{Parametric} and \class{Uniform}.
-The \class{Trace} class at each level takes its properties directly from the behavior of the original traffic.
-The \class{Nonparametric} and \class{Parametric} classes use nonparametric models of marginal behavior at each level to
+For each of the \class{Node}, \class{Flow} and \class{Packet} abstract superclasses, we define four concrete subclasses: \class{Trace}, \class{Nonparametric}, \class{Parametric}, and \class{Uniform}.
+The \class{Trace} classes take properties of objects directly from the behavior of the corresponding object in the original trace.
+The \class{Nonparametric} and \class{Parametric} classes use models of the corresponding type to randomly generate the properties of objects.
+The \class{Uniform} classes generate identical objects with average properties.
+The specific details of how each the concrete classes at each level are defined are specified in Table~\ref{tab:models}.
+The parameters for the fitted models used in the \class{Parametric} subclasses are estimated from the corresponding nonparametric distribution using maximum likelihood estimation.
+Figure~\ref{fig:trace-vs-uniform} shows visual depictions comparing trace versus uniform behavior at various levels in example scenarios.
 
-We separate traffic generation into three levels of behavior: network, flow and packet.
-The {network} level determines properties pertaining to nodes and their interaction.
-For example, it determines which nodes communicate with each other, and how frequently.
-This level can be viewed as assigning network nodes to the source and destination roles of flows, as well as determining per-node properties which affect lower level models.
-The next level of behavior determines overall properties of flows.
-This includes each flow's start time, duration, packet count and total data volume.
-Finally, the packet level of behavior determines the individual sizes of packets and the durations of intervals between their transmission.
-Despite the concreteness of the examples mentioned here, there are also less obvious, abstract properties that models at the network and flow levels determine.
-Flow models, for example, dictate the distribution of each flow's packet sizes, a property which can be used to determine individual packet sizes, but which is not itself specific to any particular packet.
-
-%\subsubsection*{Network Behavior Models}
-
-%We consider four models of network-level behavior. The most faithful is the ``trace'' model, which assigns each flow
-
-For each level of behavior we define a spectrum of models.
-The gold standard of realism at each level is provided by the ``trace'' model for that level, which replicates the node, flow or packet behavior of the original trace exactly.
-The most rudimentary model at each level is the ``uniform'' model, which treats all entities identically---be they nodes, flows or packets.
-Despite being so simplistic (or perhaps because of it), the uniform behavior models are by far the most commonly for workload generation in experiments.
-Between these extremes on the spectrum of realism, we consider one or two models with intermediate degrees of complexity and fidelity.
-At the packet level, we define the ``sample'' model, which is in essence a per-flow variable bit-rate (\caps{VBR}) model: each packet's size is determined by randomly sampling a value from the packet size distribution given by the flow model; the interval until the next packet is then sampled from the given interval distribution.\footnote{Note that the uniform packet model (\textit{i.e.} {\scriptsize{CBR}}), can be considered to be a special case of the sample model where the size and time distributions are both constant.} The intermediate model at the flow level is designated as ``proportional'': the distributions of packet sizes and inter-packet intervals are taken from the trace and passed down to the packet model (which may use the actual distribution, via sampling, or simply use the mean size and intervals, to generate \caps{CBR} traffic).
+While the modularity of the Traffic Object Model allows us to mix and match sub-models, not all possible combinations are valid.
+In particular, note that the \class{Trace} classes are tied to a specific object from the underlying trace, whereas for other classes, there is no such correspondence to any specific object.
+Consequently, once a non-trace model has been used, trace models may not be used at the lower levels.
+This still leaves 40 valid composite combinations of sub-models; we only evaluate twelve of these.
+Composite models are abbreviated by the initials of their constituent sub-models.
+The fully trace-base model, \class{TTT}, serves as an experimental control.
+We consider \class{TTN}, \class{TTP}, and \class{TTU} in order to compare models of packet behavior.
+We evaluate different flow models, using the best possible model at the packet level: \class{TNN}, \class{TPN}, \class{TUN}.
+Similarly, we compare models for node behavior: \class{NNN}, \class{PNN}, \class{UNN}.
+Finally, we use the na\"ive uniform model, \class{UUU}, and the composite parametric model, \class{PPP}, proposed by {\FHC}~\cite{Hernandez06:wlan-traffic}, as external points of comparison.
 
@@
 The time axis indicates when flows start and end; the width of each flow line is proportional to the logarithm of its data rate.
-Figure \ref{fig:packet-behavior} compares \caps{CBR} packet behavior with the trace of an actual flow.
+Figure \ref{fig:packet-behavior} compares uniform (i.e.
+{\footnotesize{CBR}}) packet behavior with the trace of an actual flow.
 In the uniform model, the cumulative data sent increases smoothly over time, whereas in the actual packet trace, the transmissions are variable both in size and in inter-packet interval, leading to a ``lumpy'' cumulative data plot.}
-\label{fig:traffic-models}
+\label{fig:trace-vs-uniform}
 \end{center}
@@
 \end{figure*}
+
+%We separate traffic generation into three levels of behavior: network, flow and packet.
+The {network} level determines properties pertaining to nodes and their interaction.
+For example, it determines which nodes communicate with each other, and how frequently.
+This level can be viewed as assigning network nodes to the source and destination roles of flows, as well as determining per-node properties which affect lower level models.
+The next level of behavior determines overall properties of flows.
+This includes each flow's start time, duration, packet count and total data volume.
+Finally, the packet level of behavior determines the individual sizes of packets and the durations of intervals between their transmission.
+Despite the concreteness of the examples mentioned here, there are also less obvious, abstract properties that models at the network and flow levels determine.
+Flow models, for example, dictate the distribution of each flow's packet sizes, a property which can be used to determine individual packet sizes, but which is not itself specific to any particular packet.
+
+%\subsubsection*{Network Behavior Models}
+
+%We consider four models of network-level behavior. The most faithful is the ``trace'' model, which assigns each flow
+
+%For each level of behavior we define a spectrum of models.
+The gold standard of realism at each level is provided by the ``trace'' model for that level, which replicates the node, flow or packet behavior of the original trace exactly.
+The most rudimentary model at each level is the ``uniform'' model, which treats all entities identically---be they nodes, flows or packets.
+Despite being so simplistic (or perhaps because of it), the uniform behavior models are by far the most commonly for workload generation in experiments.
+Between these extremes on the spectrum of realism, we consider one or two models with intermediate degrees of complexity and fidelity.
+At the packet level, we define the ``sample'' model, which is in essence a per-flow variable bit-rate (\caps{VBR}) model: each packet's size is determined by randomly sampling a value from the packet size distribution given by the flow model; the interval until the next packet is then sampled from the given interval distribution.\footnote{Note that the uniform packet model (\textit{i.e.} {\scriptsize{CBR}}), can be considered to be a special case of the sample model where the size and time distributions are both constant.} The intermediate model at the flow level is designated as ``proportional'': the distributions of packet sizes and inter-packet intervals are taken from the trace and passed down to the packet model (which may use the actual distribution, via sampling, or simply use the mean size and intervals, to generate \caps{CBR} traffic).
 
@@
 
-The three levels of behavior are orthogonal and can be varied almost independently.
+%The three levels of behavior are orthogonal and can be varied almost independently.
 The exception to their independence is that the trace-based packet behavior model can only be used when the flow behavior model is also trace-based.
@@
 
-We have selected a variety of performance metrics at the application, network, and link layers of the protocol stack:
+We have selected seven performance metrics at the application, network, and link layers of the protocol stack:
 \begin{enumerate}
@@
 \end{enumerate}
-These metrics are commonly used to evaluate the performance of new wireless protocols.
-For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
+These metrics are commonly used to evaluate the performance of wireless protocols.%For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
 %Unfortunately, not all of the simulations were able to run to completion.
@@
 
-We use the Qualnet wireless network simulator to perform our experiments.
+We use the Qualnet wireless network simulator (version 4.0.1) to perform our experiments.
 We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
@@
 Moreover, the following features are preserved across all models: the numbers of wireless nodes, the number of flows, the number of application-initiated data units sent, the total bytes of application data sent, and the average flow duration (and therefore the average data rate).
+
+We have modified the standard Qualnet distribution in the two significant ways.
+First, In our previous work~\cite{Karpinski07:realism,Karpinski07:cbr-failure}, we approximated full-duplex \caps{TCP} with a pair of \caps{UDP} flows.
+For this work, we have implemented a new Qualnet application driver that allows trace-driven full-duplex \caps{TCP} flows.
+This allows us to accurately reproduce the full dynamics of \caps{TCP} feedback.
+Moreover, raw \caps{IP} and \caps{UDP} flows are implemented using the same code, guaranteeing that all traffic is simulated uniformly and that all performance metrics are aggregated in the same manner.
+The second modification is that we have re-instrumented Qualnet to collect \caps{IP} performance metrics in strict adherence to the recommendations of the \caps{IETF} working group recommendations~\cite{rfc:ip-metrics,rfc:jitter}.
 

### 2008-01-22 models.rb

@@ -40,8 +40,8 @@ models.each do |l|
     puts '\end{minipage} \\\\'
     puts '\cline{2-3}' if j < models.size
   end
+  puts '\hline'
 end
 puts <<'FOOTER'
-\hline
 \end{tabular}
 FOOTER

### 2008-01-22 abstract.tex
Rewrote the abstract.

@@
-When new wireless technologies are deployed and subjected to real usage patterns, unforeseen performance problems inevitably seem to arise, to be fixed only in later generations.
+Although our understanding of traffic patterns in computer networks is vastly deeper than it was twenty years ago, we still are missing a ``cookbook'' for generating realistic workload in local area networks ({\footnotesize{LAN}}s).
+Worse still, we are without a commonly agreed-upon definition of what it means for a traffic model to be realistic.
+This inability to produce realistic wireless {\footnotesize{LAN}} workload is no mere academic concern: we believe it is one of the major reasons that wireless experiments---simulated or otherwise---so rarely succeed in predicting real-world performance of new protocols.
+Our results indicate that na\"ive but commonly used traffic models can induce drastic distortion of vital performance metrics, such as end-to-end application-level network delay, and jitter, as well as important metrics at other layers of the network.
+These errors are great than an order of magnitude more than a quarter of the diverse simulation scenarios we have evaluated.
+In this work we present an objective, testable definition of ``sufficient realism'' for traffic models, and develop the theoretical methodology necessary to interpret experimental results using this definition.
+We also propose the Traffic Object Model~({\footnotesize{TOM}}) as a standard framework for implementing reusable, modular {\footnotesize{LAN}} traffic models.
+%
+%When new wireless technologies are deployed and subjected to real usage patterns, unforeseen performance problems inevitably seem to arise, to be fixed only in later generations.
 Why do these performance issues fail to appear in experimental settings before the technology is deployed?

### 2008-02-01 models.tex
Updates.

@@
 \hline
-\multirow{8}{*}[2.5pt]{\class{Node}}
+\multirow{8}{*}[2.5pt]{\class{Session}}
 & \multirow{1}{*}[-0.05em]{\class{Trace}} &
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} session start time assigned from trace \\
-\raisebox{1.5pt}{$\centerdot$} each node corresponds to a specific trace node with associated trace flows \\
+\raisebox{1.5pt}{$\centerdot$} start time assigned from trace \\
+\raisebox{1.5pt}{$\centerdot$} session corresponds to a specific trace node with associated trace flows \\
 \raisebox{1.5pt}{$\centerdot$} overrides all distributions with node-specific versions
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} session start time randomly assigned from pool of trace session start times \\
+\raisebox{1.5pt}{$\centerdot$} start time randomly assigned from pool of trace session start times \\
 \raisebox{1.5pt}{$\centerdot$} nodes assigned as flow end-points by sampling from inherited distributions
@@
 \vspace{2pt}
-\raisebox{1.5pt}{$\centerdot$} session start time assigned according to Poisson arrival model \\
+\raisebox{1.5pt}{$\centerdot$} start time assigned according to Poisson arrival model \\
 \raisebox{1.5pt}{$\centerdot$} nodes assigned as flow end-points by sampling from fitted BiPareto model
@@
 \raisebox{1.5pt}{$\centerdot$} session start times selected uniformly from viable range \\
-\raisebox{1.5pt}{$\centerdot$} nodes have the mean number of generic flows
+\raisebox{1.5pt}{$\centerdot$} sessions all have the mean number of generic flows
 \vspace{2pt}

### 2008-02-01 models.rb

@@ -6,9 +6,9 @@ models = YAML.load_file("models.yaml")
 
 bullet = '\\raisebox{1.5pt}{$\\centerdot$}'
 fixups = {
-  :Node   => [ 8,'2.5pt'],
-  :Flow   => [14,'2.5pt'],
-  :Packet => [ 5,'2.5pt']
+  :Session => [ 8,'2.5pt'],
+  :Flow    => [14,'2.5pt'],
+  :Packet  => [ 5,'2.5pt']
 }
 
 puts <<'HEADER'

### 2008-02-01 paper.tex
Modifed version from home that got overwritten in the last sync.

@@
 \section{Introduction}\label{sec:intro}
+
+We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in local area networks (\caps{LAN}s).
+Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematic precision and succinctness, while simultaneously being made clearer and more general.
+
+Most of what are called traffic models are actually only partial descriptions of one or a handful of aspects of a complete model.
+The ubiquitous constant bit-rate (\caps{CBR}) model, for example, merely describes the nature of the packets in each flow in a workload: in a given flow, the sizes of packets and the intervals between them are are all uniform.
+Hence, the transmitted bit-rate is constant.
+
+Two things are require for the direct empirical comparison of traffic models:
+\begin{itemize}
+\item An objective standard for comparing quality of models.
+% WinMee/Broadnets provide this
+\item A common framework in which different models can be implemented and compared.
+% this paper provides this
+\end{itemize}
 
@@
 {\includegraphics[width=1.392in]{flow_packet_behavior}\label{fig:packet-behavior}}
-\caption{Examples illustrating trace versus uniform behaviors a the node, flow and packet levels.
+\caption{Examples illustrating trace (real) versus uniform (na\"ive synthetic) behaviors at the node, flow and packet levels.
 Figures \ref{fig:flow-topology-trace} and \ref{fig:flow-topology-uniform} show example node behaviors.
@@
 Moreover, raw \caps{IP} and \caps{UDP} flows are implemented using the same code, guaranteeing that all traffic is simulated uniformly and that all performance metrics are aggregated in the same manner.
-The second modification is that we have re-instrumented Qualnet to collect \caps{IP} performance metrics in strict adherence to the recommendations of the \caps{IETF} working group recommendations~\cite{rfc:ip-metrics,rfc:jitter}.
+The second modification is that we have instrumented Qualnet to collect \caps{IP} performance metrics in strict adherence to the recommendations of the \caps{IETF} working group recommendations~\cite{rfc:ip-metrics,rfc:jitter}.
 

### 2008-02-02 paper.tex
Paragraph on algebraic representations of flow behavior.

@@
 
-\newcommand{\thetitle}{On the Inadequacy of Uniform Modeling for Wireless LAN Workload Generation}
+\newcommand{\thetitle}{Linear Algebra as a Unifying Theory of Workload Modeling}
 
@@
 
-\title{\thetitle}
+\title{Linear Algebra as a Unifying Theory\\of Wireless Workload Modeling}
 \author{
@@
 \end{abstract}
+
+\section{Algebraic Representations \\of Network Behavior}
+
+\subsection*{Flow Behavior}
+
+The behavior of a flow may be mapped into a vector space such that the two operations, namely vector addition and scalar multiplication, can be sensibly defined in a way that naturally corresponds with the aggregation of behaviors.
+This is best clarified with an intentionally simplistic example.
+Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and the total duration, $d$.
+This triplet can be naturally embedded in Euclidean space: $\vec{x}=(p,b,d)\in\R^3$.
+If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum behavior,'' $\vec{x}_1+\vec{x}_2$ has:
+\begin{align*}
+p_1+p_2 &\text{ packets,}\\
+b_1+b_2 &\text{ bytes,}\\
+d_1+d_2 &\text{ duration.}
+\end{align*}
+Their average, i.e. $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
+
+While this may seem trivial, note that the representation must be chosen carefully.
+Suppose instead, for example, we represented each flow's behavior by its packet count, $p$, average packet size, $b/p$, and average inter-packet interval, $d/(p-1)$.
+In this representation, the average of two behaviors has the average average packet size (the two averages are intentional), and the average average inter-packet interval.
+There lies the rub: this is still the case no matter how many packets are in each flow.
+Instead, we should more naturally weight the averages by the number of packets in each constituent.
+This is precisely what the packets-bytes-duration representation accomplishes.
 

### 2008-02-02 paper.tex
More description of flow behaviors.

@@
 
-\subsection*{Flow Behavior}
+\subsection{Flow Behavior}
 
@@
 This triplet can be naturally embedded in Euclidean space: $\vec{x}=(p,b,d)\in\R^3$.
-If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum behavior,'' $\vec{x}_1+\vec{x}_2$ has:
-\begin{align*}
-p_1+p_2 &\text{ packets,}\\
-b_1+b_2 &\text{ bytes,}\\
-d_1+d_2 &\text{ duration.}
-\end{align*}
-Their average, i.e. $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
+If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum'' behavior, $\vec{x}_1+\vec{x}_2$ has $p_1+p_2$ packets, $b_1+b_2$ bytes, $d_1+d_2$ duration.
+Their average, i.e.
+$\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
 
-While this may seem trivial, note that the representation must be chosen carefully.
+%While this may seem trivial, note that the representation must be chosen carefully.
 Suppose instead, for example, we represented each flow's behavior by its packet count, $p$, average packet size, $b/p$, and average inter-packet interval, $d/(p-1)$.
-In this representation, the average of two behaviors has the average average packet size (the two averages are intentional), and the average average inter-packet interval.
-There lies the rub: this is still the case no matter how many packets are in each flow.
+In this representation, the average of two behaviors has the \textit{average} average packet size, and the \textit{average} average inter-packet interval.
+Therein lies the rub: this is the case no matter how many packets are in each flow; if a flow with thousands of 1500-byte packets is averaged with a single-packet, one-byte flow, the average behavior has...
 Instead, we should more naturally weight the averages by the number of packets in each constituent.
 This is precisely what the packets-bytes-duration representation accomplishes.
+
+The {packets-bytes-duration} model representation of flow behavior is clearly quite simplistic.
+It captures nothing about variability in packet sizes, nor about the ``burstiness'' of packet transmissions.
+At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space.
+The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to aggregate behavior of flows.
+Instead, we take an intermediate approach, suggested by the main result of~\cite{Karpinski07:realism}: if the start time and end time of each flow is known, it suffices to know the empirical distribution of packet sizes and inter-packet intervals in order to reproduce network performance characteristics accurately.
+Accordingly, we represent a flow's behavior by these precise characteristics.
+
+Explicitly, we encode a flow's behavior with four vectors: $\vec{z} \in \R^{d_z}$ is a histogram of packet sizes; $\vec{v} \in \R^{d_v}$ is a histogram of inter-packet intervals; $\vec{\alpha}, \vec{\epsilon} \in \R^{d_t}$ are vectors indicating when the flow begins and ends.
+To expand on this, $z_i$ (i.e.
+the $i$th element of $\vec{z}$) is the number of packets in the flow such with $i$ bytes and $d_z$ is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes.
+The histogram of inter-packet intervals is similar, except that domain of possible intervals is a continuum, so we much map the range of possible intervals onto the discrete index values $\left\{1,\dots,d_v\right\}$.
+In our experiments, we use the following logarithmic quantization function:
+\begin{align}
+q(v) &= \left\lfloor(d_v-1)\frac{\ln v - \ln v_{\min}}{\ln v_{\max}-\ln v_{\min}}\right\rfloor+1.
+\end{align}
+with $v_{\min}=10^{-6}$, which is the minimal interval our trace data, and $v_{\max}=600$, which is the total duration of each simulated situation, and hence an upper bound on the inter-packet intervals.
+We use $d_v=1500$ to match the dimensionally of the packet size representation.
+
+The flow start and end time indicator vectors, $\vec{\alpha}$ and $\vec{\epsilon}$, also require some explanation.
+Start and end times are simple scalar values.
+Why not represent them as such?
+The answer lies in the requirement that behaviors must add meaningfully.
+If we represent the start time as a scalar, then adding the behavior of two flows produces a (fairly) meaningless value.
+It's certainly not a start time itself---it may even exceed the start time of any actual flow.
+Instead, we again quantize the continuum of time: we map the range of possible start times onto the indices $\left\{1,\dots,d_t\right\}$, where $d_t$ is some chosen dimensionality.
+If the start time is mapped to index, $i$, then $\vec{\alpha}=e_i$, the vector with 1 at index $i$ and zeros elsewhere.
+The benefit of this representation is that we can now appropriately represent aggregate behavior through vector addition: the sum of start vectors over any collection of flows represents is a histogram of start times for those flows.
+
+Finally, we collect all the pieces of flow behavior into matrix representations of behavior across an entire traffic trace.
+If there are $f$ flows in the trace, we define the size matrix $Z\in\R^{f \times d_z}$ to be the matrix whose row vectors are the size vectors $\vec{z}_i$ for $i=1,\dots,f$.
+In precisely the same manner, we define the interval matrix, $V\in\R^{f \times d_v}$, and the start time matrix, $A\in\R^{f \times d_t}$, and the end time matrix, $E\in\R^{f \times d_t}$.
+
+\subsection{Flow Sources \& Destinations}
 

### 2008-02-02 paper.tex
More work on explaining flow behavior.

@@
 
-\subsection{Flow Behavior}
+The traffic workload pattern in a network is simply a collection of sequences of packets between hosts in the network sent at certain times.
+It is standard in network analysis to aggregate sequences of packets sharing the same \caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP/UDP} port numbers.
+Such a sequence of packets is called a \textit{flow}.
+The traffic pattern is characterized completely by the detailed behavior of it's flows.
+The behavior of a is defined by six properties: its source and destination nodes, its start and end times, and the sizes of and intervals between its constituent packets.
+We propose to represent the behavior of each flow in terms of six vectors---one for each aspect of the flow's behavior.
+The aggregate behavior of all the flows in a workload pattern may then be represented by six matrices, each row of which gives the behavior vectors of one of the flows.
 
-The behavior of a flow may be mapped into a vector space such that the two operations, namely vector addition and scalar multiplication, can be sensibly defined in a way that naturally corresponds with the aggregation of behaviors.
-This is best clarified with an intentionally simplistic example.
+When mapping flow behaviors into a vector space, it is important that the mapping of behaviors into the vector spaces in question respect the vector space operations: addition of vectors and scalar multiplication.
+This requirement is best clarified with an intentionally simplistic example.
 Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and the total duration, $d$.
-This triplet can be naturally embedded in Euclidean space: $\vec{x}=(p,b,d)\in\R^3$.
+This triplet can be naturally embedded into Euclidean space: $\vec{x}=(p,b,d)\in\R^3$.
+If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum'' behavior, $\vec{x}_1+\vec{x}_2$ has $p_1+p_2$ packets, $b_1+b_2$ bytes, $d_1+d_2$ duration.
+Their average, i.e.
+$\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
+
+%The behavior of a flow may be mapped into a vector space such that the two operations, namely vector addition and scalar multiplication, naturally correspond with the aggregation of behaviors.
+This requirement is best clarified with an intentionally simplistic example.
+Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and the total duration, $d$.
+This triplet can be naturally embedded into Euclidean space: $\vec{x}=(p,b,d)\in\R^3$.
 If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum'' behavior, $\vec{x}_1+\vec{x}_2$ has $p_1+p_2$ packets, $b_1+b_2$ bytes, $d_1+d_2$ duration.
@@
 
-The {packets-bytes-duration} model representation of flow behavior is clearly quite simplistic.
-It captures nothing about variability in packet sizes, nor about the ``burstiness'' of packet transmissions.
+The {packets-bytes-duration} model representation is clearly quite simplistic, not to mention incomplete.
+It captures nothing about variability in packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time unspecified; and it says nothing about the source or destination of the flow.
 At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space.
 The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to aggregate behavior of flows.
-Instead, we take an intermediate approach, suggested by the main result of~\cite{Karpinski07:realism}: if the start time and end time of each flow is known, it suffices to know the empirical distribution of packet sizes and inter-packet intervals in order to reproduce network performance characteristics accurately.
+We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: if the end-points, start time and end time of each flow is known, it suffices to know the empirical distribution of packet sizes and inter-packet intervals in order to reproduce network performance characteristics accurately.
 Accordingly, we represent a flow's behavior by these precise characteristics.
+We encode a single flow's behavior with six component vectors:
+\begin{itemize}
+\item $\vec{z} \in \R^{d_z}$ is a histogram of packet sizes;
+\item $\vec{v} \in \R^{d_v}$ is a histogram of inter-packet intervals;
+\item $\vec{b} \in \R^{d_t}$ indicates when the flow begins;
+\item $\vec{e} \in \R^{d_t}$ indicates when the flow ends;
+\item $\vec{s}\in\R^n$ indicates the source node of the flow;
+\item $\vec{d}\in\R^n$ indicates the destination node of the flow.
+\end{itemize}
+The behavior vectors for all flows are collected into six corresponding matrices, whose rows are the behavior vectors above:
+\begin{itemize}
+\item $Z\in\R^{f \times d_z}$ the size matrix;
+\item $V\in\R^{f \times d_v}$ the interval matrix;
+\item $B\in\R^{f \times d_t}$ the start time matrix;
+\item $E\in\R^{f \times d_t}$ the end time matrix;
+\item $S\in\R^{f \times n}$ the source matrix;
+\item $D\in\R^{f \times n}$ the destination matrix.
+\end{itemize}
 
-Explicitly, we encode a flow's behavior with four vectors: $\vec{z} \in \R^{d_z}$ is a histogram of packet sizes; $\vec{v} \in \R^{d_v}$ is a histogram of inter-packet intervals; $\vec{\alpha}, \vec{\epsilon} \in \R^{d_t}$ are vectors indicating when the flow begins and ends.
+%We collect all the components of flow behavior into matrix representations of behavior across an entire traffic trace.
+If there are $f$ flows in the trace, we define the size matrix $Z\in\R^{f \times d_z}$ to be the matrix whose rows are the size vectors $\vec{z}_i$ for $i=1,\dots,f$.
+In precisely the same manner, we define the interval matrix, $V\in\R^{f \times d_v}$, and the start time matrix, $A\in\R^{f \times d_t}$, and the end time matrix, $E\in\R^{f \times d_t}$.
+
+\subsection{Packet Sizes \& Inter-Packet Intervals}
+
 To expand on this, $z_i$ (i.e.
@@
 We use $d_v=1500$ to match the dimensionally of the packet size representation.
+The packet size and inter-packet intervals are collected into matrices, $Z\in\R^{f \times d_z}$ and $V\in\R^{f \times d_v}$, whose rows are the behavior vectors $\vec{z}_i$ and $\vec{v}_i$ indexed from $i=1,\dots,f$ where $f$ is the number of flows in the collection.
+
+\subsection{Start \& End Times}
 
@@
 The answer lies in the requirement that behaviors must add meaningfully.
-If we represent the start time as a scalar, then adding the behavior of two flows produces a (fairly) meaningless value.
+If we represent the start time as a scalar, then adding the behavior of two flows produces a meaningless value.
 It's certainly not a start time itself---it may even exceed the start time of any actual flow.
@@
 If the start time is mapped to index, $i$, then $\vec{\alpha}=e_i$, the vector with 1 at index $i$ and zeros elsewhere.
-The benefit of this representation is that we can now appropriately represent aggregate behavior through vector addition: the sum of start vectors over any collection of flows represents is a histogram of start times for those flows.
+The benefit of this representation is that we can now appropriately represent aggregate behavior through vector addition: the sum of start vectors over any collection of flows is a histogram of start times for those flows.
 
-Finally, we collect all the pieces of flow behavior into matrix representations of behavior across an entire traffic trace.
-If there are $f$ flows in the trace, we define the size matrix $Z\in\R^{f \times d_z}$ to be the matrix whose row vectors are the size vectors $\vec{z}_i$ for $i=1,\dots,f$.
-In precisely the same manner, we define the interval matrix, $V\in\R^{f \times d_v}$, and the start time matrix, $A\in\R^{f \times d_t}$, and the end time matrix, $E\in\R^{f \times d_t}$.
-
-\subsection{Flow Sources \& Destinations}
+\subsection{Source \& Destination Nodes}
 

### 2008-02-02 paper.tex
More work on explaining flow behavior.

@@
 When mapping flow behaviors into a vector space, it is important that the mapping of behaviors into the vector spaces in question respect the vector space operations: addition of vectors and scalar multiplication.
-This requirement is best clarified with an intentionally simplistic example.
-Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and the total duration, $d$.
+This requirement is best clarified with a simple example.
+Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and its total duration, $d$.
 This triplet can be naturally embedded into Euclidean space: $\vec{x}=(p,b,d)\in\R^3$.
@@
 It captures nothing about variability in packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time unspecified; and it says nothing about the source or destination of the flow.
-At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space.
-The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to aggregate behavior of flows.
-We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: if the end-points, start time and end time of each flow is known, it suffices to know the empirical distribution of packet sizes and inter-packet intervals in order to reproduce network performance characteristics accurately.
+At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space, etc.
+The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to aggregate behaviors of flows.
+We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: if the end-points, start time and end time of each flow are known, it suffices to know empirical distributions of packet sizes and inter-packet intervals for each flow to reproduce network performance accurately.
 Accordingly, we represent a flow's behavior by these precise characteristics.

### 2008-02-02 paper.tex
Rewriting the packets-bytes-duration example paragraph.

@@
 
-\begin{abstract}
-\input{abstract.tex}
-\end{abstract}
+%\begin{abstract}
+%\input{abstract.tex}
+%\end{abstract}
 
@@
 
-The traffic workload pattern in a network is simply a collection of sequences of packets between hosts in the network sent at certain times.
-It is standard in network analysis to aggregate sequences of packets sharing the same \caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP/UDP} port numbers.
+The traffic workload pattern in a network is simply a collection of packets between hosts in the network sent at certain times.
+It is standard in network analysis, however, to aggregate sequences of packets sharing the same \caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP} or \caps{UDP} port numbers.
 Such a sequence of packets is called a \textit{flow}.
 The traffic pattern is characterized completely by the detailed behavior of it's flows.
-The behavior of a is defined by six properties: its source and destination nodes, its start and end times, and the sizes of and intervals between its constituent packets.
-We propose to represent the behavior of each flow in terms of six vectors---one for each aspect of the flow's behavior.
-The aggregate behavior of all the flows in a workload pattern may then be represented by six matrices, each row of which gives the behavior vectors of one of the flows.
+The behavior of a flow is defined by six properties: its source and destination nodes, its start and end times, and the specific sizes of and intervals between its constituent packets.
+We propose to represent the behavior of each flow in terms of six vectors---one for each aspect of its behavior.
+The aggregate behavior of all the flows in a workload pattern may then be represented by six matrices, the rows of which are the behavior vectors of the flows.
 
 When mapping flow behaviors into a vector space, it is important that the mapping of behaviors into the vector spaces in question respect the vector space operations: addition of vectors and scalar multiplication.
-This requirement is best clarified with a simple example.
+This requirement is best clarified with a simplified example.
 Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and its total duration, $d$.
-This triplet can be naturally embedded into Euclidean space: $\vec{x}=(p,b,d)\in\R^3$.
-If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum'' behavior, $\vec{x}_1+\vec{x}_2$ has $p_1+p_2$ packets, $b_1+b_2$ bytes, $d_1+d_2$ duration.
-Their average, i.e.
+Despite its simplicity, this representation is adequate to reproduce constant bit-rate (\caps{CBR}), time-randomized flow behavior.
+The triplet of representative values can be naturally embedded into Euclidean space: $\vec{x}=(p,b,d)\in\R^3$.
+If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their sum as vectors, $x_1+x_2=(p_1+p_2,b_1+b_2,d_1+d_2)$, provides an appropriate description of the aggregate behavior of the two flows together.
+Similarly, their average, i.e.
 $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
@@
 
-To expand on this, $z_i$ (i.e.
-the $i$th element of $\vec{z}$) is the number of packets in the flow such with $i$ bytes and $d_z$ is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes.
+The packets size behavior of a flow is represented by a histogram vector, $\vec{z}$, whose $i$th element, $z_i$, is the number of packets in the flow with $i$ bytes of payload data.
+The dimension, $d_z$, is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes.
 The histogram of inter-packet intervals is similar, except that domain of possible intervals is a continuum, so we much map the range of possible intervals onto the discrete index values $\left\{1,\dots,d_v\right\}$.

### 2008-02-02 paper.tex
Work on the paragraph introducint our standard model.

@@
 
-The {packets-bytes-duration} model representation is clearly quite simplistic, not to mention incomplete.
-It captures nothing about variability in packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time unspecified; and it says nothing about the source or destination of the flow.
+The {packets-bytes-duration} model representation is clearly overly simplistic and very incomplete.
+(Yet, it is more detailed than most commonly used experimental traffic models!) It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time unspecified; it says nothing about the source or destination of the flow.
 At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space, etc.
-The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to aggregate behaviors of flows.
-We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: if the end-points, start time and end time of each flow are known, it suffices to know empirical distributions of packet sizes and inter-packet intervals for each flow to reproduce network performance accurately.
+The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows.
+We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, and empirical distributions of packet sizes and inter-packet intervals for each flow.
 Accordingly, we represent a flow's behavior by these precise characteristics.
-We encode a single flow's behavior with six component vectors:
+We encode a single flow's behavior with six vectors:
 \begin{itemize}
-\item $\vec{z} \in \R^{d_z}$ is a histogram of packet sizes;
-\item $\vec{v} \in \R^{d_v}$ is a histogram of inter-packet intervals;
+\item $\vec{s}\in\R^n$ indicates the source of the flow;
+\item $\vec{d}\in\R^n$ indicates the destination of the flow.;
 \item $\vec{b} \in \R^{d_t}$ indicates when the flow begins;
 \item $\vec{e} \in \R^{d_t}$ indicates when the flow ends;
-\item $\vec{s}\in\R^n$ indicates the source node of the flow;
-\item $\vec{d}\in\R^n$ indicates the destination node of the flow.
+\item $\vec{z} \in \R^{d_z}$ is a histogram of packet sizes;
+\item $\vec{v} \in \R^{d_v}$ is a histogram of inter-packet intervals.
 \end{itemize}
-The behavior vectors for all flows are collected into six corresponding matrices, whose rows are the behavior vectors above:
+If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with six matrices:
 \begin{itemize}
-\item $Z\in\R^{f \times d_z}$ the size matrix;
-\item $V\in\R^{f \times d_v}$ the interval matrix;
+\item $S\in\R^{f \times n}$ the source matrix;
+\item $D\in\R^{f \times n}$ the destination matrix;
 \item $B\in\R^{f \times d_t}$ the start time matrix;
 \item $E\in\R^{f \times d_t}$ the end time matrix;
-\item $S\in\R^{f \times n}$ the source matrix;
-\item $D\in\R^{f \times n}$ the destination matrix.
+\item $Z\in\R^{f \times d_z}$ the size matrix;
+\item $V\in\R^{f \times d_v}$ the interval matrix.
 \end{itemize}
-
-%We collect all the components of flow behavior into matrix representations of behavior across an entire traffic trace.
-If there are $f$ flows in the trace, we define the size matrix $Z\in\R^{f \times d_z}$ to be the matrix whose rows are the size vectors $\vec{z}_i$ for $i=1,\dots,f$.
-In precisely the same manner, we define the interval matrix, $V\in\R^{f \times d_v}$, and the start time matrix, $A\in\R^{f \times d_t}$, and the end time matrix, $E\in\R^{f \times d_t}$.
+The rows of each matrix are behavior vectors of the $f$ flows in the workload, indexed in some chosen order.
 

### 2008-02-02 paper.tex
Combined vector/matrix notation lists into one.

@@
 \newcommand{\E}[1]{\left<#1\right>}
-\newcommand{\abs}[1]{\left|#1\right|}
 \newcommand{\R}{\mathbb{R}}
 \newcommand{\Q}{\mathcal{Q}}
+
+\newcommand{\abs}[1]{\left|#1\right|}
+\newcommand{\absx}[1]{|#1|}
+\newcommand{\set}[1]{\left\{#1\right\}}
+\newcommand{\setx}[1]{\{#1\}}
 
@@
 \newcommand{\tablename}{Table}
+
+\renewcommand{\labelitemi}{\raisebox{2pt}{$\centerdot$}}
 
@@
 We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, and empirical distributions of packet sizes and inter-packet intervals for each flow.
-Accordingly, we represent a flow's behavior by these precise characteristics.
-We encode a single flow's behavior with six vectors:
+Accordingly, we represent a flow's behavior by these precise characteristics.If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with six matrices; the rows of each matrix are behavior vectors of the $f$ flows in the workload, indexed in some chosen order.
+We use the following notation for the behavior matrices and their constituent vectors:
 \begin{itemize}
-\item $\vec{s}\in\R^n$ indicates the source of the flow;
-\item $\vec{d}\in\R^n$ indicates the destination of the flow.;
-\item $\vec{b} \in \R^{d_t}$ indicates when the flow begins;
-\item $\vec{e} \in \R^{d_t}$ indicates when the flow ends;
-\item $\vec{z} \in \R^{d_z}$ is a histogram of packet sizes;
-\item $\vec{v} \in \R^{d_v}$ is a histogram of inter-packet intervals.
+\item $S=\setx{\vec{s}_i}\in\R^{f \times n}$ --- the source matrix;
+\item $D=\setx{\vec{d}_i}\in\R^{f \times n}$ --- the destination matrix;
+\item $B=\setx{\vec{b}_i}\in\R^{f \times d_t}$ --- the start time matrix;
+\item $E=\setx{\vec{e}_i}\in\R^{f \times d_t}$ --- the end time matrix;
+\item $Z=\setx{\vec{z}_i}\in\R^{f \times d_z}$ --- the size matrix;
+\item $V=\setx{\vec{v}_i}\in\R^{f \times d_v}$ --- the interval matrix.
 \end{itemize}
-If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with six matrices:
-\begin{itemize}
-\item $S\in\R^{f \times n}$ the source matrix;
-\item $D\in\R^{f \times n}$ the destination matrix;
-\item $B\in\R^{f \times d_t}$ the start time matrix;
-\item $E\in\R^{f \times d_t}$ the end time matrix;
-\item $Z\in\R^{f \times d_z}$ the size matrix;
-\item $V\in\R^{f \times d_v}$ the interval matrix.
-\end{itemize}
-The rows of each matrix are behavior vectors of the $f$ flows in the workload, indexed in some chosen order.
 

### 2008-02-02 paper.tex
New hand-tabbed, aligned list of matrix/vector notations.
Insanely nit-picking fixups for alignments.
Source/destination node representation.

@@
 
-\renewcommand{\labelitemi}{\raisebox{2pt}{$\centerdot$}}
+\renewcommand{\bullet}{\raisebox{2pt}{$\centerdot$}}
+%\renewcommand{\labelitemi}{\bullet}
 
@@
 Accordingly, we represent a flow's behavior by these precise characteristics.If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with six matrices; the rows of each matrix are behavior vectors of the $f$ flows in the workload, indexed in some chosen order.
-We use the following notation for the behavior matrices and their constituent vectors:
-\begin{itemize}
-\item $S=\setx{\vec{s}_i}\in\R^{f \times n}$ --- the source matrix;
-\item $D=\setx{\vec{d}_i}\in\R^{f \times n}$ --- the destination matrix;
-\item $B=\setx{\vec{b}_i}\in\R^{f \times d_t}$ --- the start time matrix;
-\item $E=\setx{\vec{e}_i}\in\R^{f \times d_t}$ --- the end time matrix;
-\item $Z=\setx{\vec{z}_i}\in\R^{f \times d_z}$ --- the size matrix;
-\item $V=\setx{\vec{v}_i}\in\R^{f \times d_v}$ --- the interval matrix.
-\end{itemize}
+We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
+\begin{tabbing}
+\hspace{3em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
+$D\:$\=$=\setx{\vec{d}_i}\:$\=\kill
+\>\bullet\> sources: \>$S$\>$=\setx{\vec{s}_i}$\>$\in\R^{f \times n}$,\\
+\>\bullet\> destinations: \>$D$\>$=\setx{\vec{d}_i}$\>$\in\R^{f \times n}$,\\
+\>\bullet\> start times: \>$B$\>$=\setx{\vec{b}_i}$\>$\in\R^{f \times d_t}$,\\
+\>\bullet\> end times: \>$E$\>$=\setx{\vec{e}_i}$\>$\in\R^{f \times d_t}$,\\
+\>\bullet\> packet sizes: \>$Z$\>$=\setx{\vec{z}_i}$\>$\in\R^{f \times d_z}$,\\
+\>\bullet\> inter-packet intervals: \>$V$\>$=\setx{\vec{v}_i}$\>$\in\R^{f \times d_v}$.
+\end{tabbing}
+
+\subsection{Source \& Destination Nodes}
+
+Let $n$ be the number of nodes in the given traffic collection.
+We map nodes onto the standard basis vectors of $\R^n$: if the $j$th node is the source of the the $i$th flow, then $\vec{s}_i=\mathbf{e}_j$ (the vector with 1 at index $j$ and zeros elsewhere); if the $k$th node is the destination, then $\vec{d}_i=\mathbf{e}_k$.
+This may seem like a rather expansive representation, given that start and end nodes can be represented as a unidimensional indices from $\set{1,\dots,n}$.
+The advantage becomes clear when we consider vector operations.
+The sum of the source indices over a collection is meaningless; the sum of the the source vectors, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.
+
+\subsection{Start \& End Times}
+
+The flow start and end time indicator vectors, $\vec{\alpha}$ and $\vec{\epsilon}$, also require some explanation.
+Start and end times are simple scalar values.
+Why not represent them as such?
+The answer lies in the requirement that behaviors must add meaningfully.
+If we represent the start time as a scalar, then adding the behavior of two flows produces a meaningless value.
+It's certainly not a start time itself---it may even exceed the start time of any actual flow.
+Instead, we again quantize the continuum of time: we map the range of possible start times onto the indices $\left\{1,\dots,d_t\right\}$, where $d_t$ is some chosen dimensionality.
+If the start time is mapped to index, $i$, then $\vec{\alpha}=e_i$, the vector with 1 at index $i$ and zeros elsewhere.
+The benefit of this representation is that we can now appropriately represent aggregate behavior through vector addition: the sum of start vectors over any collection of flows is a histogram of start times for those flows.
 
@@
 The packet size and inter-packet intervals are collected into matrices, $Z\in\R^{f \times d_z}$ and $V\in\R^{f \times d_v}$, whose rows are the behavior vectors $\vec{z}_i$ and $\vec{v}_i$ indexed from $i=1,\dots,f$ where $f$ is the number of flows in the collection.
-
-\subsection{Start \& End Times}
-
-The flow start and end time indicator vectors, $\vec{\alpha}$ and $\vec{\epsilon}$, also require some explanation.
-Start and end times are simple scalar values.
-Why not represent them as such?
-The answer lies in the requirement that behaviors must add meaningfully.
-If we represent the start time as a scalar, then adding the behavior of two flows produces a meaningless value.
-It's certainly not a start time itself---it may even exceed the start time of any actual flow.
-Instead, we again quantize the continuum of time: we map the range of possible start times onto the indices $\left\{1,\dots,d_t\right\}$, where $d_t$ is some chosen dimensionality.
-If the start time is mapped to index, $i$, then $\vec{\alpha}=e_i$, the vector with 1 at index $i$ and zeros elsewhere.
-The benefit of this representation is that we can now appropriately represent aggregate behavior through vector addition: the sum of start vectors over any collection of flows is a histogram of start times for those flows.
-
-\subsection{Source \& Destination Nodes}
 

### 2008-02-02 paper.tex
Finished the time & interval behavior section.
Pretty much done with the algebraic representations section.

@@
 \newcommand{\Q}{\mathcal{Q}}
+\newcommand{\Qt}{Q_{\text{time}}}
+\newcommand{\Qi}{Q_{\text{ival}}}
 
@@
 \newcommand{\setx}[1]{\{#1\}}
+\newcommand{\floor}[1]{\left\lfloor#1\right\rfloor}
+\newcommand{\floorx}[1]{\lfloor#1\rfloor}
+\newcommand{\ceil}[1]{\left\lceil#1\right\rceil}
+\newcommand{\ceilx}[1]{\lceil#1\rceil}
+\newcommand{\round}[1]{\left[#1\right]}
+\newcommand{\roundx}[1]{[#1]}
+\newcommand{\fracx}[2]{#1/#2}
+\newcommand{\fracp}[2]{\left(\frac{#1}{#2}\right)}
+\newcommand{\fracpx}[2]{(#1/#2)}
 
@@
 
-The flow start and end time indicator vectors, $\vec{\alpha}$ and $\vec{\epsilon}$, also require some explanation.
-Start and end times are simple scalar values.
-Why not represent them as such?
-The answer lies in the requirement that behaviors must add meaningfully.
-If we represent the start time as a scalar, then adding the behavior of two flows produces a meaningless value.
-It's certainly not a start time itself---it may even exceed the start time of any actual flow.
-Instead, we again quantize the continuum of time: we map the range of possible start times onto the indices $\left\{1,\dots,d_t\right\}$, where $d_t$ is some chosen dimensionality.
-If the start time is mapped to index, $i$, then $\vec{\alpha}=e_i$, the vector with 1 at index $i$ and zeros elsewhere.
-The benefit of this representation is that we can now appropriately represent aggregate behavior through vector addition: the sum of start vectors over any collection of flows is a histogram of start times for those flows.
+We also represent start and end times using indicator vectors.
+Time, however, is continuous, so we must first map times onto a range of time indices $\set{1,\dots,d_t}$.
+We transform the range $[0,t_{\max}]$ of possible start times using a simple quantization function:
+\begin{align}
+\Qt(t) &= \max\set{1,\ceil{\fracx{d_t t}{t_{\max}}}}.
+\end{align}
+Let $j$ and $k$ be the quantized start and end indices of a flow.
+Then the start and end vectors for the flow are $\vec{b}=\mathbf{e}_j$ and $\vec{e}=\mathbf{e}_k$.
+Again, this may seem excessively complex, but the advantage becomes clear when take aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.
 
@@
 The dimension, $d_z$, is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes.
-The histogram of inter-packet intervals is similar, except that domain of possible intervals is a continuum, so we much map the range of possible intervals onto the discrete index values $\left\{1,\dots,d_v\right\}$.
-In our experiments, we use the following logarithmic quantization function:
+The histogram of inter-packet intervals is similar, except that the domain of possible intervals is a continuum, so we much map this range onto the discrete index values $\left\{1,\dots,d_v\right\}$.
+In our experiments, we use the following adaptive quantization function:
 \begin{align}
-q(v) &= \left\lfloor(d_v-1)\frac{\ln v - \ln v_{\min}}{\ln v_{\max}-\ln v_{\min}}\right\rfloor+1.
+\Qi(v) &= \max\set{1,\ceil{d_v\fracp{v-v_{\min}}{v_{\max}-v_{\min}}^{\frac{1}{2}}}}.
 \end{align}
-with $v_{\min}=10^{-6}$, which is the minimal interval our trace data, and $v_{\max}=600$, which is the total duration of each simulated situation, and hence an upper bound on the inter-packet intervals.
-We use $d_v=1500$ to match the dimensionally of the packet size representation.
-The packet size and inter-packet intervals are collected into matrices, $Z\in\R^{f \times d_z}$ and $V\in\R^{f \times d_v}$, whose rows are the behavior vectors $\vec{z}_i$ and $\vec{v}_i$ indexed from $i=1,\dots,f$ where $f$ is the number of flows in the collection.
+This specific function was chosen because, given the values of $d_v$, $v_{\min}$, and $v_{\max}$ in our experiments, it provides millisecond-scale to second-scale resolution for intervals on the orders of $v_{\min}$ to $v_{\max}$, respectively, adapting resolution smoothly between these extremes.
+%The values we use are $v_{\min}=10^{-6}$, which is the minimal interval our trace data, and $v_{\max}=600$, which is the total duration of each simulated situation, and using $d_v=1500$ to match the dimensionally of the packet size representation.
+The summation of behavior vectors over multiple flows yields a joint histogram giving the pooled frequencies of all packet sizes and inter-packet intervals.
 

### 2008-02-02 paper.tex
Added an individual element notation to the notations table.

@@
 We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, and empirical distributions of packet sizes and inter-packet intervals for each flow.
-Accordingly, we represent a flow's behavior by these precise characteristics.If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with six matrices; the rows of each matrix are behavior vectors of the $f$ flows in the workload, indexed in some chosen order.
+Accordingly, we represent a flow's behavior by these precise characteristics.
+If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with six matrices; the rows of each matrix are behavior vectors of all flows, indexed in some chosen order.
 We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
 \begin{tabbing}
-\hspace{3em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
-$D\:$\=$=\setx{\vec{d}_i}\:$\=\kill
-\>\bullet\> sources: \>$S$\>$=\setx{\vec{s}_i}$\>$\in\R^{f \times n}$,\\
-\>\bullet\> destinations: \>$D$\>$=\setx{\vec{d}_i}$\>$\in\R^{f \times n}$,\\
-\>\bullet\> start times: \>$B$\>$=\setx{\vec{b}_i}$\>$\in\R^{f \times d_t}$,\\
-\>\bullet\> end times: \>$E$\>$=\setx{\vec{e}_i}$\>$\in\R^{f \times d_t}$,\\
-\>\bullet\> packet sizes: \>$Z$\>$=\setx{\vec{z}_i}$\>$\in\R^{f \times d_z}$,\\
-\>\bullet\> inter-packet intervals: \>$V$\>$=\setx{\vec{v}_i}$\>$\in\R^{f \times d_v}$.
+\hspace{1.5em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
+$D\:$\=$=\setx{\vec{d}_i}\:$\=$=\setx{d_{ij}}\:$\=\kill
+\>\bullet\> sources: \>$S$\>$=\setx{\vec{s}_i}$\>$=\setx{s_{ij}}$\>$\in\R^{f \times n}$,\\
+\>\bullet\> destinations: \>$D$\>$=\setx{\vec{d}_i}$\>$=\setx{d_{ij}}$\>$\in\R^{f \times n}$,\\
+\>\bullet\> start times: \>$B$\>$=\setx{\vec{b}_i}$\>$=\setx{b_{ij}}$\>$\in\R^{f \times d_t}$,\\
+\>\bullet\> end times: \>$E$\>$=\setx{\vec{e}_i}$\>$=\setx{e_{ij}}$\>$\in\R^{f \times d_t}$,\\
+\>\bullet\> packet sizes: \>$Z$\>$=\setx{\vec{z}_i}$\>$=\setx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
+\>\bullet\> inter-packet intervals: \>$V$\>$=\setx{\vec{v}_i}$\>$=\setx{v_{ij}}$\>$\in\R^{f \times d_v}$.
 \end{tabbing}
@@
 The summation of behavior vectors over multiple flows yields a joint histogram giving the pooled frequencies of all packet sizes and inter-packet intervals.
+
+\section{Derivative Representations of Network Behavior}
+
+Now that we have defined our representation of network behavior, we derive useful expressions for commonly used network properties in terms of these basic representations.
+
+\subsection{Communication Topology}
+
+The number of flows between each pair of nodes in the network is immediately given by the matrix product $SD^t\in\R^{n \times n}$.
+The entry at row $i$ and column $j$ in this matrix gives the number of flows from
+
+\section{Expression of Common Models in Algebraic Form}
+
+Most traffic models are not explicit means of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
 

### 2008-02-02 paper.tex
Added a bad-ass footnote about flow pairing.
Working on adding the flow pairing stuff into the representation section.

@@
 The traffic pattern is characterized completely by the detailed behavior of it's flows.
-The behavior of a flow is defined by six properties: its source and destination nodes, its start and end times, and the specific sizes of and intervals between its constituent packets.
-We propose to represent the behavior of each flow in terms of six vectors---one for each aspect of its behavior.
-The aggregate behavior of all the flows in a workload pattern may then be represented by six matrices, the rows of which are the behavior vectors of the flows.
+The behavior of a flow is defined by seven properties: its source and destination nodes; its start and end times; the specific sizes of and intervals between its constituent packets; and finally, whether it is a response, paired with another flow.\footnote{Flow pairing is typically not addressed in traffic models.
+However, the most common transport protocol, {\scriptsize{TCP}}, is full-duplex, exhibiting tightly coupled pairs of client-server flows.
+Therefore, accurate representation of traffic behavior requires knowing which flows are request-response pairs.
+Moreover, although they are not strictly full-duplex, it is generally appropriate to view {\scriptsize{UDP}} and even {\scriptsize{ICMP}} flows as paired, since they are typically paired.
+For example, {\scriptsize{DNS}} responses are naturally associated with the corresponding request, and will have symmetrical {\scriptsize{IP}} addresses and port numbers.} We propose to represent the behavior of each flow in terms of seven vectors---one for each aspect of its behavior.
+The aggregate behavior of all the flows in a workload pattern may then be represented by these matrices, the rows of which are the behavior vectors of the flows.
 
@@
 The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows.
-We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, and empirical distributions of packet sizes and inter-packet intervals for each flow.
-Accordingly, we represent a flow's behavior by these precise characteristics.
-If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with six matrices; the rows of each matrix are behavior vectors of all flows, indexed in some chosen order.
+We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, empirical distributions of packet sizes and inter-packet intervals for each flow, and which other flow (if any) it is paired with.
+If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with seven matrices; the rows of each matrix are behavior vectors of all flows, indexed in some chosen order.
 We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
@@
 \>\bullet\> destinations: \>$D$\>$=\setx{\vec{d}_i}$\>$=\setx{d_{ij}}$\>$\in\R^{f \times n}$,\\
+\>\bullet\> flow pairs: \>$P$\>$=\setx{\vec{p}_i}$\>$=\setx{p_{ij}}$\>$\in\R^{f \times f}$,\\
 \>\bullet\> start times: \>$B$\>$=\setx{\vec{b}_i}$\>$=\setx{b_{ij}}$\>$\in\R^{f \times d_t}$,\\
@@
 We map nodes onto the standard basis vectors of $\R^n$: if the $j$th node is the source of the the $i$th flow, then $\vec{s}_i=\mathbf{e}_j$ (the vector with 1 at index $j$ and zeros elsewhere); if the $k$th node is the destination, then $\vec{d}_i=\mathbf{e}_k$.
-This may seem like a rather expansive representation, given that start and end nodes can be represented as a unidimensional indices from $\set{1,\dots,n}$.
+This may seem like an overly expansive representation, given that start and end nodes can be represented as a unidimensional indices from $\set{1,\dots,n}$.
 The advantage becomes clear when we consider vector operations.
@@
 
+\subsection{Flow Pairing: Clients \& Servers}
+
 \subsection{Start \& End Times}
 
-We also represent start and end times using indicator vectors.
-Time, however, is continuous, so we must first map times onto a range of time indices $\set{1,\dots,d_t}$.
+We represent start and end times using indicator vectors.
+Unlike node indices, however, time is continuous, so we must first map times onto a range of time indices $\set{1,\dots,d_t}$.
 We transform the range $[0,t_{\max}]$ of possible start times using a simple quantization function:
@@
 Then the start and end vectors for the flow are $\vec{b}=\mathbf{e}_j$ and $\vec{e}=\mathbf{e}_k$.
-Again, this may seem excessively complex, but the advantage becomes clear when take aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.
+Again, this may seem excessively complex, but the advantage becomes clear when we aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.
 
@@
 
-The packets size behavior of a flow is represented by a histogram vector, $\vec{z}$, whose $i$th element, $z_i$, is the number of packets in the flow with $i$ bytes of payload data.
+The packets size behavior of a flow is represented by a histogram vector, $\vec{z}$.
+Explicitly,  $z_{ij}$ is the number of packets in the $i$th flow with $j$ bytes of payload data.
 The dimension, $d_z$, is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes.
@@
 
-\section{Derivative Representations of Network Behavior}
-
-Now that we have defined our representation of network behavior, we derive useful expressions for commonly used network properties in terms of these basic representations.
+\section{Expressions for Common Workload Properties}
+
+Having defined our representation of network behavior, we turn to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.
 
@@
 
-The number of flows between each pair of nodes in the network is immediately given by the matrix product $SD^t\in\R^{n \times n}$.
-The entry at row $i$ and column $j$ in this matrix gives the number of flows from
+The number of flows between each pair of nodes in the network is immediately given by the matrix product $SD^t$: $(SD^t)_{ij}$ is the number flows from node $i$ to node $j$.
 

### 2008-02-02 paper.tex
Just realized that protocol type is not represented. Crap.

@@
 The traffic pattern is characterized completely by the detailed behavior of it's flows.
-The behavior of a flow is defined by seven properties: its source and destination nodes; its start and end times; the specific sizes of and intervals between its constituent packets; and finally, whether it is a response, paired with another flow.\footnote{Flow pairing is typically not addressed in traffic models.
+The behavior of a flow is defined by seven properties: its source and destination nodes; its start and end times; the specific sizes of and intervals between its constituent packets; and finally, whether it is paired with another flow in a request-response relationship.\footnote{Flow pairing is typically not addressed in traffic models.
 However, the most common transport protocol, {\scriptsize{TCP}}, is full-duplex, exhibiting tightly coupled pairs of client-server flows.
 Therefore, accurate representation of traffic behavior requires knowing which flows are request-response pairs.
-Moreover, although they are not strictly full-duplex, it is generally appropriate to view {\scriptsize{UDP}} and even {\scriptsize{ICMP}} flows as paired, since they are typically paired.
-For example, {\scriptsize{DNS}} responses are naturally associated with the corresponding request, and will have symmetrical {\scriptsize{IP}} addresses and port numbers.} We propose to represent the behavior of each flow in terms of seven vectors---one for each aspect of its behavior.
-The aggregate behavior of all the flows in a workload pattern may then be represented by these matrices, the rows of which are the behavior vectors of the flows.
+Moreover, although they are not strictly full-duplex, it is generally appropriate to view {\scriptsize{UDP}} and {\scriptsize{ICMP}} flows as paired, since the client-server paradigm is still used.
+Each {\scriptsize{DNS}} response, for example, is naturally associated with its corresponding request, and will have symmetrical {\scriptsize{IP}} addresses and port numbers.} We propose to represent the behavior of each flow in terms of seven vectors---one for each aspect of its behavior.
+The aggregate behavior of all the flows in a workload pattern are then represented as matrices, the rows of which are the behavior vectors of the flows.
 
@@
 
-The {packets-bytes-duration} model representation is clearly overly simplistic and very incomplete.
-(Yet, it is more detailed than most commonly used experimental traffic models!) It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time unspecified; it says nothing about the source or destination of the flow.
+The {packets-bytes-duration} representation is clearly overly simplistic and very incomplete.
+(Yet, it is more detailed than the most commonly used experimental traffic models!) It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time unspecified; it says nothing about the source or destination of the flow.
 At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space, etc.
 The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows.
-We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, empirical distributions of packet sizes and inter-packet intervals for each flow, and which other flow (if any) it is paired with.
-If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with seven matrices; the rows of each matrix are behavior vectors of all flows, indexed in some chosen order.
-We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
+We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, empirical distributions of packet sizes and inter-packet intervals for each flow, and which other flow (if any) it is a response to.
+If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with seven matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
+We use the following notation for the behavior matrices and their constituent vectors and elements:\vspace{-0.25em}
 \begin{tabbing}
@@
 
-\subsection{Flow Pairing: Clients \& Servers}
+\subsection{Flow Pairing: Requests \& Responses}
 

### 2008-02-03 paper.tex
I'm not sure that the flow pairing really belongs here.
The problem is that there's no way to make sense of summing flow pairings.

@@
 
-The traffic workload pattern in a network is simply a collection of packets between hosts in the network sent at certain times.
+%The traffic workload pattern in a network is the collection of actions taken by hosts, on behalf of users and applications, that affect the behavior of the network itself.
+By design, Internet Protocol (\caps{IP}) networks treat all data in the same manner~\cite{Clark88}.
+Therefore, the actual information transmitted is irrelevant, only the manner of transmission matters.
+Although networks are packet-switched, packets are logically organized into sequences of packets sharing the same \caps{IP} protocol, source and destination \caps{IP} address, and source and destination  \caps{TCP} or \caps{UDP} port numbers.
+Such a sequence of packets is called a \textit{flow}.
+Although there are a vast multitude different application types which use \caps{IP}, there are two fundamentally different behaviors of flows which we distinguish here: \caps{TCP}, which delivers data reliably through feedback and retransmissions; and \caps{UDP}, which delivers data only on a best-effort basis.
+While it is possible to implement reliable transmission similar to \caps{TCP} over \caps{UDP}, it is uncommon, so we make the generalization that network events, such as packet drops, do not significantly feedback to the behavior of upper layers.
+With \caps{TCP}, on the other hand, network events are known to affect the transport layer
+
+The traffic workload pattern in a network is simply a collection of \caps{IP} packets between hosts in the network sent at certain times.
 It is standard in network analysis, however, to aggregate sequences of packets sharing the same \caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP} or \caps{UDP} port numbers.
@@
 \>\bullet\> destinations: \>$D$\>$=\setx{\vec{d}_i}$\>$=\setx{d_{ij}}$\>$\in\R^{f \times n}$,\\
-\>\bullet\> flow pairs: \>$P$\>$=\setx{\vec{p}_i}$\>$=\setx{p_{ij}}$\>$\in\R^{f \times f}$,\\
+\>\bullet\> response: \>$R$\>$=\setx{\vec{r}_i}$\>$=\setx{r_{ij}}$\>$\in\R^{f \times f}$,\\
 \>\bullet\> start times: \>$B$\>$=\setx{\vec{b}_i}$\>$=\setx{b_{ij}}$\>$\in\R^{f \times d_t}$,\\
@@
 Let $n$ be the number of nodes in the given traffic collection.
-We map nodes onto the standard basis vectors of $\R^n$: if the $j$th node is the source of the the $i$th flow, then $\vec{s}_i=\mathbf{e}_j$ (the vector with 1 at index $j$ and zeros elsewhere); if the $k$th node is the destination, then $\vec{d}_i=\mathbf{e}_k$.
-This may seem like an overly expansive representation, given that start and end nodes can be represented as a unidimensional indices from $\set{1,\dots,n}$.
-The advantage becomes clear when we consider vector operations.
+We let $s_{ij}=1$ if the $j$th node is the source of the $i$th flow, and zero otherwise; similarly, $d_{ik}=1$ if the destination of the $i$th flow is the $k$th node.
+This sparse matrix may seem like an overly expansive representation, given that start and end nodes can be represented as unidimensional indices from $\set{1,\dots,n}$.
+The advantage becomes clear when we consider vector operations combining collections of behavior vectors.
 The sum of the source indices over a collection is meaningless; the sum of the the source vectors, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.
 
-\subsection{Flow Pairing: Requests \& Responses}
+\subsection{Requests \& Responses: Flow Pairing}
 

### 2008-02-03 paper.tex
In the midst of editing, but why not commit?

@@
 \newcommand{\Section}[1]{\hyperref[sec:#1]{Section~\ref*{sec:#1}}}
+\newcommand{\Equation}[1]{\hyperref[eqn:#1]{Equation~\ref*{eqn:#1}}}
+
 \newcommand{\FHC}{Hern\'andez-Campos~\textit{et~al.}}
@@
 \DeclareMathOperator{\Def}{def}
+\DeclareMathOperator{\Val}{val}
 
@@
 \newcommand{\Qi}{Q_{\text{ival}}}
+\newcommand{\ones}[1]{\mathbf{1}_{#1}}
 
@@
 \newcommand{\setx}[1]{\{#1\}}
+\newcommand{\seq}[1]{\left<#1\right>}
+\newcommand{\seqx}[1]{<#1>}
 \newcommand{\floor}[1]{\left\lfloor#1\right\rfloor}
@@
 
-\section{Algebraic Representations \\of Network Behavior}
+%\section{Algebraic Representations \\of Network Behavior}\label{sec:representations}
+\section{The General Matrix Model}
 
@@
 The traffic pattern is characterized completely by the detailed behavior of it's flows.
-The behavior of a flow is defined by seven properties: its source and destination nodes; its start and end times; the specific sizes of and intervals between its constituent packets; and finally, whether it is paired with another flow in a request-response relationship.\footnote{Flow pairing is typically not addressed in traffic models.
-However, the most common transport protocol, {\scriptsize{TCP}}, is full-duplex, exhibiting tightly coupled pairs of client-server flows.
-Therefore, accurate representation of traffic behavior requires knowing which flows are request-response pairs.
-Moreover, although they are not strictly full-duplex, it is generally appropriate to view {\scriptsize{UDP}} and {\scriptsize{ICMP}} flows as paired, since the client-server paradigm is still used.
-Each {\scriptsize{DNS}} response, for example, is naturally associated with its corresponding request, and will have symmetrical {\scriptsize{IP}} addresses and port numbers.} We propose to represent the behavior of each flow in terms of seven vectors---one for each aspect of its behavior.
+The behavior of a flow is defined by six properties: its source and destination nodes; its start and end times; and the specific sizes of and intervals between its constituent packets.
+We propose to represent the behavior of each flow in terms of six vectors---one for each aspect of its behavior.
 The aggregate behavior of all the flows in a workload pattern are then represented as matrices, the rows of which are the behavior vectors of the flows.
@@
 The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows.
-We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, empirical distributions of packet sizes and inter-packet intervals for each flow, and which other flow (if any) it is a response to.
-If there are $f$ flows in the entire workload pattern, then its collective behavior can be described with seven matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
+We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, and empirical distributions of packet sizes and inter-packet intervals for each flow.
+If there are $f$ flows in the entire workload pattern, then the collective behavior can be described with six matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
 We use the following notation for the behavior matrices and their constituent vectors and elements:\vspace{-0.25em}
@@
 \>\bullet\> destinations: \>$D$\>$=\setx{\vec{d}_i}$\>$=\setx{d_{ij}}$\>$\in\R^{f \times n}$,\\
-\>\bullet\> response: \>$R$\>$=\setx{\vec{r}_i}$\>$=\setx{r_{ij}}$\>$\in\R^{f \times f}$,\\
 \>\bullet\> start times: \>$B$\>$=\setx{\vec{b}_i}$\>$=\setx{b_{ij}}$\>$\in\R^{f \times d_t}$,\\
@@
 \end{tabbing}
+The total behavior matrix is the concatenation of these matrices: $F = (S,D,B,E,Z,V) \in \R^{f \times \Sigma}$ where $\Sigma=2(n+d_t)+d_z+d_v$.
+Many common traffic models can be expressed precisely and succinctly in terms of transformations or conditions on this behavior matrix and its constituents.
 
@@
 The advantage becomes clear when we consider vector operations combining collections of behavior vectors.
-The sum of the source indices over a collection is meaningless; the sum of the the source vectors, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.
-
-\subsection{Requests \& Responses: Flow Pairing}
+The sum of the source indices over a collection is meaningless; the sum of the the source vectors, $\set{\vec{s}_i}$, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.
 
@@
 \begin{align}
-\Qt(t) &= \max\set{1,\ceil{\fracx{d_t t}{t_{\max}}}}.
-\end{align}
+\label{eqn:time-quantization}
+\Qt(t) &= \max\set{1,\ceil{d_t\fracpx{t}{t_{\max}}}}.
+\end{align}
+%Suppose the $i$th flow has a quantized start index of $j$ and a end index of $k$. Then $b_{ij'}=\delta{jj'}$ and $e_{ik'}=\delta_{kk'}$.
 Let $j$ and $k$ be the quantized start and end indices of a flow.
-Then the start and end vectors for the flow are $\vec{b}=\mathbf{e}_j$ and $\vec{e}=\mathbf{e}_k$.
+Then the start and end vectors for the flow are $\vec{b}=\mathbf{e}_j$ and $\vec{e}=\mathbf{e}_k$ ($\mathbf{e}_j$ has unity at index $j$ and zeros elsewhere).
 Again, this may seem excessively complex, but the advantage becomes clear when we aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.
@@
 The dimension, $d_z$, is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes.
-The histogram of inter-packet intervals is similar, except that the domain of possible intervals is a continuum, so we much map this range onto the discrete index values $\left\{1,\dots,d_v\right\}$.
+The histogram of inter-packet intervals is similar, except that the domain of possible intervals is a continuum, so we must map this range onto the discrete index values $\left\{1,\dots,d_v\right\}$.
 In our experiments, we use the following adaptive quantization function:
 \begin{align}
+\label{eqn:interval-quantization}
 \Qi(v) &= \max\set{1,\ceil{d_v\fracp{v-v_{\min}}{v_{\max}-v_{\min}}^{\frac{1}{2}}}}.
@@
 
-The number of flows between each pair of nodes in the network is immediately given by the matrix product $SD^t$: $(SD^t)_{ij}$ is the number flows from node $i$ to node $j$.
-
-\section{Expression of Common Models in Algebraic Form}
-
-Most traffic models are not explicit means of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
+The number of flows between each pair of nodes in the network is immediately given by the matrix product $SD^t\in\R^{n \times n}$, where $*^t$ denotes the transpose.
+That is $(SD^t)_{ij}$ is the number of flows from node $i$ to node $j$.
+The total number of source-flows for each node is given by the vector $S^t\ones{f}\in\R^n$, while the total number of destination-flows is given by $D^t\ones{f}\in\R^n$ ($\ones{f}$ is the $f$-dimensional column vector consisting of all ones).
+
+\subsection{Average Flow Behaviors}
+
+The average packet sizes and inter-packet intervals for flows are computed by right multiplication by the vector of values corresponding to the column indices.
+For example, the packet sizes corresponding to the columns of the size matrix are simply the indices themselves: $\vec{\sigma}=(1,2,3,\dots,d_z)$.
+Since inter-packet intervals were quantized according to \Equation{interval-quantization}, we use the median value of the $k$th interval to represent it:
+\begin{align}
+\vec{\tau}(k) = v_{\min}+(v_{\max}-v_{\min})\fracp{k(k+1)+\frac{1}{2}}{d_v^2}.
+\end{align}
+With these index-value vectors, the average packet sizes and inter-packet intervals for each flow can be computed:
+\begin{align}
+\label{eqn:average-size}%
+\text{average sizes:}&~~~~~Z\vec{\sigma} \in \R^f,\\
+\label{eqn:average-interval}%
+\text{average intervals:}&~~~~~V\vec{\tau} \in \R^f.
+\end{align}
+
+%Computing the aggregate behavior is exceedingly simple. It is generally accomplished through
+
+\section{Expression of Common Models}
+
+Existing traffic models are typically not explicit methods of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
+In our algebraic modeling framework, such models can be interpreted as stipulations about the form of the behavior matrices, or as transforms that implement those stipulations given a real set of traffic matrices.
+
+%A generic traffic model is a way of generating the six matrices described in \Section{representations}.
+A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or regularized form.
+For experimental comparison, the dimensions of the matrices should all be preserved.
+If a trace-derived model produces experimental performance metrics that are similar to those produced by the original trace data, then simplifications made by the model are viable; if the performance is distorted, then the simplifications are discarding essential information about the traffic pattern.
+
+\subsection{Uniformity}
+
+It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees, largely because of the high complexity of doing otherwise and the lack of compelling alternatives.
+Here we examine how various notions of uniformity are represented in the matrix model.
+In general, \textit{intra-flow uniformity} is effected by right multiplication, which mixes the elements in each row, whereas \textit{inter-flow uniformity} is effected by left multiplication, which mixes elements across rows.
+The standard constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals.
+Each flow's bit-rate is therefore constant, and hence the name.
+The strict \caps{CBR} model can be effected by using the expressions for per-flow average packet sizes and inter-packet intervals given in Equations~\ref{eqn:average-size}~and~\ref{eqn:average-interval}.
+This essentially reduces the size and time representation to the
 

### 2008-02-04 paper.tex
Much work on paper.

@@
 \newcommand{\Qi}{Q_{\text{ival}}}
+\newcommand{\Di}{Q^{-1}_{\text{ival}}}
 \newcommand{\ones}[1]{\mathbf{1}_{#1}}
@@
 \newcommand{\setx}[1]{\{#1\}}
+\newcommand{\parens}[1]{\left(#1\right)}
+\newcommand{\parensx}[1]{(#1)}
 \newcommand{\seq}[1]{\left<#1\right>}
@@
 
-%\section{Algebraic Representations \\of Network Behavior}\label{sec:representations}
+%\section{Algebraic Representations \\of Network Behavior}
 \section{The General Matrix Model}
+\label{sec:representations}
 
@@
 Despite its simplicity, this representation is adequate to reproduce constant bit-rate (\caps{CBR}), time-randomized flow behavior.
-The triplet of representative values can be naturally embedded into Euclidean space: $\vec{x}=(p,b,d)\in\R^3$.
-If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their sum as vectors, $x_1+x_2=(p_1+p_2,b_1+b_2,d_1+d_2)$, provides an appropriate description of the aggregate behavior of the two flows together.
+The triplet of representative values can be naturally embedded into Euclidean space: $\vec{x}=\seq{p,b,d}\in\R^3$.
+If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their sum as vectors, $x_1+x_2=\seq{p_1+p_2,b_1+b_2,d_1+d_2}$, provides an appropriate description of the aggregate behavior of the two flows together.
 Similarly, their average, i.e.
@@
 \hspace{1.5em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
-$D\:$\=$=\setx{\vec{d}_i}\:$\=$=\setx{d_{ij}}\:$\=\kill
-\>\bullet\> sources: \>$S$\>$=\setx{\vec{s}_i}$\>$=\setx{s_{ij}}$\>$\in\R^{f \times n}$,\\
-\>\bullet\> destinations: \>$D$\>$=\setx{\vec{d}_i}$\>$=\setx{d_{ij}}$\>$\in\R^{f \times n}$,\\
-\>\bullet\> start times: \>$B$\>$=\setx{\vec{b}_i}$\>$=\setx{b_{ij}}$\>$\in\R^{f \times d_t}$,\\
-\>\bullet\> end times: \>$E$\>$=\setx{\vec{e}_i}$\>$=\setx{e_{ij}}$\>$\in\R^{f \times d_t}$,\\
-\>\bullet\> packet sizes: \>$Z$\>$=\setx{\vec{z}_i}$\>$=\setx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
-\>\bullet\> inter-packet intervals: \>$V$\>$=\setx{\vec{v}_i}$\>$=\setx{v_{ij}}$\>$\in\R^{f \times d_v}$.
+$D\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
+\>\bullet\> sources: \>$S$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$,\\
+\>\bullet\> destinations: \>$D$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$,\\
+\>\bullet\> start times: \>$B$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_t}$,\\
+\>\bullet\> end times: \>$E$\>$=\parensx{\vec{e}_i}$\>$=\parensx{e_{ij}}$\>$\in\R^{f \times d_t}$,\\
+\>\bullet\> packet sizes: \>$Z$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
+\>\bullet\> inter-packet intervals: \>$V$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
 \end{tabbing}
-The total behavior matrix is the concatenation of these matrices: $F = (S,D,B,E,Z,V) \in \R^{f \times \Sigma}$ where $\Sigma=2(n+d_t)+d_z+d_v$.
-Many common traffic models can be expressed precisely and succinctly in terms of transformations or conditions on this behavior matrix and its constituents.
+The total behavior is the direct product of these components: $F = \seq{S,D,B,E,Z,V} \in \R^{f \times \Sigma}$ with $\Sigma=2n+2d_t+d_z+d_v$.
+Many common traffic models can be expressed succinctly and precisely in terms of transformations or conditions on this behavior matrix and its constituents.
+In the following sections, we describe in detail how each flow's behavior is represented in these matrices.
+Finally, we explain how a model instance stochastically generates network traffic.
 
@@
 Explicitly,  $z_{ij}$ is the number of packets in the $i$th flow with $j$ bytes of payload data.
-The dimension, $d_z$, is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes.
+The dimension, $d_z$, is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium, typically 1500 bytes.
 The histogram of inter-packet intervals is similar, except that the domain of possible intervals is a continuum, so we must map this range onto the discrete index values $\left\{1,\dots,d_v\right\}$.
@@
 \label{eqn:interval-quantization}
-\Qi(v) &= \max\set{1,\ceil{d_v\fracp{v-v_{\min}}{v_{\max}-v_{\min}}^{\frac{1}{2}}}}.
-\end{align}
-This specific function was chosen because, given the values of $d_v$, $v_{\min}$, and $v_{\max}$ in our experiments, it provides millisecond-scale to second-scale resolution for intervals on the orders of $v_{\min}$ to $v_{\max}$, respectively, adapting resolution smoothly between these extremes.
-%The values we use are $v_{\min}=10^{-6}$, which is the minimal interval our trace data, and $v_{\max}=600$, which is the total duration of each simulated situation, and using $d_v=1500$ to match the dimensionally of the packet size representation.
-The summation of behavior vectors over multiple flows yields a joint histogram giving the pooled frequencies of all packet sizes and inter-packet intervals.
-
-\section{Expressions for Common Workload Properties}
+\Qi(v) &= \max\set{1,\ceil{d_v\fracp{v-v_{\min}}{v_{\max}-v_{\min}}^\gamma}}.
+\end{align}
+The parameter values function must be chosen.
+In our case, the time measurements had microsecond precision, so $v_{\min}=10^{-6}$.
+The length of each simulated scenario is ten minutes, so the upper limit on intervals is $v_{\max}=600$.
+The number of dimensions was chosen to match the dimensions for packet size: $d_v=1500$.
+Finally, the parameter $\gamma\in(0,1)$ allows us to fine tune how the resolution scales with interval size.
+We have chosen $\gamma=\fracx{1}{3}$ because this choice provides resolution that scales smoothly from the microseconds for small intervals to seconds for large intervals.
+
+\subsection{Traffic Generation}
+
+Given an instance of the general matrix model, how do we generate an actual workload form it?
+Each row of the model is used to randomly generate the behavior of a single flow.
+The source and destination nodes of the flow are randomly chosen using histogram vectors $\vec{s}$ and $\vec{d}$ the proportional probabilities of each node being chosen.\footnote{Since a flow must have distinct source and destination nodes, this sampling must be done jointly, avoiding the collisions.
+The simplest technique is the rejection method: if the same node is chosen for both source and destination, reject that choice and choose both nodes again.}
+For start and end times, a bin for each is chosen with weight given by the behavior histogram; within each bin, a random value is uniformly chosen from the range of possible values.\footnote{The start time must precede the end time, so these choices must also be made jointly, enforcing this restriction.
+The rejection method is used.}
+Each flow packet sequence is then generated, beginning at the chosen start time.
+The size of each packet is sampled from the histogram of packet sizes.
+The interval to wait before the next packet is determined from the behavior vector, $\vec{v}$: an index, $k$ is chosen, with probability proportional to the value, $\vec{v}_k$; a uniform value, $u$ is chosen from $[0,1)$, and $(k,u)$ is transformed back into a time value using the ``dequantization'' function:
+\begin{align}
+\label{eqn:interval-dequantization}
+\Di(k,u)=v_{\min}+(v_{\max}-v_{\min})\left(\frac{k+u}{d_v}\right)^{\frac{1}{\gamma}}.
+\end{align}
+
+%For example, if there are 3 nodes in a workload, and the source vector for a flow is $\vec{s}=(3,7,1)$, we choose a source node $k$ with $\Pr(k=1)=\frac{3}{11}$, $\Pr(k=2)=\frac{7}{11}$, and $\Pr(k=3)=\frac{1}{11}$.
+For inter-packet intervals, a bin is chosen in the same manner, and a random value from the range represented by the bin is generated.
+
+\section{Expression of Common Properties}
 
@@
 
-The number of flows between each pair of nodes in the network is immediately given by the matrix product $SD^t\in\R^{n \times n}$, where $*^t$ denotes the transpose.
+The number of flows between each pair of nodes in the network is expressed by the matrix product $SD^t\in\R^{n \times n}$ ($*^t$~denotes transpose).
 That is $(SD^t)_{ij}$ is the number of flows from node $i$ to node $j$.
-The total number of source-flows for each node is given by the vector $S^t\ones{f}\in\R^n$, while the total number of destination-flows is given by $D^t\ones{f}\in\R^n$ ($\ones{f}$ is the $f$-dimensional column vector consisting of all ones).
-
-\subsection{Average Flow Behaviors}
-
-The average packet sizes and inter-packet intervals for flows are computed by right multiplication by the vector of values corresponding to the column indices.
-For example, the packet sizes corresponding to the columns of the size matrix are simply the indices themselves: $\vec{\sigma}=(1,2,3,\dots,d_z)$.
-Since inter-packet intervals were quantized according to \Equation{interval-quantization}, we use the median value of the $k$th interval to represent it:
-\begin{align}
-\vec{\tau}(k) = v_{\min}+(v_{\max}-v_{\min})\fracp{k(k+1)+\frac{1}{2}}{d_v^2}.
-\end{align}
-With these index-value vectors, the average packet sizes and inter-packet intervals for each flow can be computed:
-\begin{align}
-\label{eqn:average-size}%
-\text{average sizes:}&~~~~~Z\vec{\sigma} \in \R^f,\\
-\label{eqn:average-interval}%
-\text{average intervals:}&~~~~~V\vec{\tau} \in \R^f.
-\end{align}
-
-%Computing the aggregate behavior is exceedingly simple. It is generally accomplished through
+The total number of source flows for each node is given by the vector $S^t\ones{f}$, and the total number of destination flows is given by $D^t\ones{f}$.
+($\ones{f}$ is the $f$-dimensional column vector consisting of all ones.)
+
+\subsection{Flow Packet Count, Size \& Duration}
+\label{sec:flow-totals}
+
+From the matrices $Z$ and $V$, we can compute the packet count, size (total bytes), and approximate duration for each flow.
+The computations for size and duration require knowing the values that each matrix column represents.
+The computed duration is only approximate because information is irrecoverably lost in the quantization process (see \Equation{interval-quantization}).
+The vectors of representative values for $Z$ and $V$ are:
+\begin{align}
+\vec{\sigma}&=(1,2,3,\dots,d_z),\\
+\vec{\tau}&=\seq{\Di\left(k,\fracx{1}{2}\right)}_{k=1}^{d_v}.
+\end{align}
+Using these vectors of representative values, the vector of packet-counts is $Z\ones{d_z}=V\ones{d_v}$, the vector of flow sizes vector is $Z\vec{\sigma}$, and the vector of durations is $V\vec{\tau}$.
 
@@
 Each flow's bit-rate is therefore constant, and hence the name.
-The strict \caps{CBR} model can be effected by using the expressions for per-flow average packet sizes and inter-packet intervals given in Equations~\ref{eqn:average-size}~and~\ref{eqn:average-interval}.
-This essentially reduces the size and time representation to the
+The strict \caps{CBR} model can be expressed simply using the packets-bytes-duration model discussed in \Section{representations}.
+Deriving the packet-bytes-duration model from the general matrix model is accomplished through right multiplication as described in \Section{flow-totals}.
+
+While constant bit-rate in the strict sense implies only that the packet behavior in each flow is uniform, it is extremely rare for it to be applied in this strict sense only; it is typical that \caps{CBR} models make assumptions about uniformity across flows, and not only within each flow.
+The most drastic simplification is to treat all flows in a synthetic workload identically.
+This is effected by left multiplication: $F'=\ones{f \times f} F$.
+In general, we can express uniformity through left-multiplication by some matrix.
+To make flow behavior uniform at the per-source node level, we left-multiply by $SS^t/f\in\R^{f \times f}$: $F'=(SS^t/f)F$.
+Per-destination node uniformity can be achieved similarly: $F'=(DD^t/f)F$.
+
+Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified.
+The confusion arises due to the lack of a unique way of associating flows with nodes.
+Each node can equally viably be associated with its source or its destination.
+Which association is more valid?
+Consider a web request: should the behavior of the request be determined by what the client typically does?
+Or should the requests that the server typically receives determine behavior?
+There is no armchair answer: only analysis of real data can give us the answer.
+
+The matrix model has the benefit of making the uniformization of behaviors completely explicit.
+It also allows us to easily define intermediate choices.
+If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
+\begin{align}
+F' = ((1-\alpha)(SS^t)+\alpha(DD^t))F.
+\end{align}
+The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors.
+Source-only behavior is given by $\alpha=0$, for destination-only set $\alpha=1$, and for the standard average, set $\alpha=\frac{1}{2}$.
+
+Another benefit of the matrix model is that all aspects of network behavior are determined concurrently.
+We are not left wondering what to do about source and destination nodes or flow start and end times after we have computed our simplified packet behaviors: the same transformation that average the packet behaviors also naturally averages the start and
 

### 2008-02-04 paper.tex
More massaging of the matrix model section.

@@
 The {packets-bytes-duration} representation is clearly overly simplistic and very incomplete.
-(Yet, it is more detailed than the most commonly used experimental traffic models!) It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time unspecified; it says nothing about the source or destination of the flow.
-At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space, etc.
+Yet, in giving a description of each flow, it is far more detailed than the most commonly used experimental traffic models.
+It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start and end time unspecified; it says nothing about the source or destination of the flow.
+At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets, etc.
+into a vector space.
 The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows.
-We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, and empirical distributions of packet sizes and inter-packet intervals for each flow.
+We take a highly detailed but intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, and empirical distributions of packet sizes and inter-packet intervals.
 If there are $f$ flows in the entire workload pattern, then the collective behavior can be described with six matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
-We use the following notation for the behavior matrices and their constituent vectors and elements:\vspace{-0.25em}
+We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
 \begin{tabbing}
@@
 Many common traffic models can be expressed succinctly and precisely in terms of transformations or conditions on this behavior matrix and its constituents.
-In the following sections, we describe in detail how each flow's behavior is represented in these matrices.
-Finally, we explain how a model instance stochastically generates network traffic.
+In the following sections, we explain in detail how each flow's behavior is represented in these matrices.
+Finally, we describe how a model instance stochastically generates network traffic.
 
@@
 
-Let $n$ be the number of nodes in the given traffic collection.
+Let $n$ be the number of nodes in the traffic collection to be modeled.
 We let $s_{ij}=1$ if the $j$th node is the source of the $i$th flow, and zero otherwise; similarly, $d_{ik}=1$ if the destination of the $i$th flow is the $k$th node.
-This sparse matrix may seem like an overly expansive representation, given that start and end nodes can be represented as unidimensional indices from $\set{1,\dots,n}$.
-The advantage becomes clear when we consider vector operations combining collections of behavior vectors.
-The sum of the source indices over a collection is meaningless; the sum of the the source vectors, $\set{\vec{s}_i}$, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.
+This sparse matrix may seem to be an overly expansive representation, given that start and end nodes can be represented simply as indices from $\set{1,\dots,n}$.
+The advantage becomes clear when we consider vector operations aggregating collections of behavior vectors.
+The sum of the source indices over a collection is meaningless; the sum of a set of source vectors, $\set{\vec{s}_i}$, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.
 
@@
 Let $j$ and $k$ be the quantized start and end indices of a flow.
-Then the start and end vectors for the flow are $\vec{b}=\mathbf{e}_j$ and $\vec{e}=\mathbf{e}_k$ ($\mathbf{e}_j$ has unity at index $j$ and zeros elsewhere).
+Then the start and end vectors for the flow are $\vec{b}=\mathbf{e}_j$ and $\vec{e}=\mathbf{e}_k$ ($\mathbf{e}_j$ has 1 at index $j$ and zeros elsewhere).
 Again, this may seem excessively complex, but the advantage becomes clear when we aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.
@@
 
-The packets size behavior of a flow is represented by a histogram vector, $\vec{z}$.
+The packets size behavior of a flow is represented by a histogram vector, $\vec{z}$, of packet sizes.
 Explicitly,  $z_{ij}$ is the number of packets in the $i$th flow with $j$ bytes of payload data.
 The dimension, $d_z$, is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium, typically 1500 bytes.
-The histogram of inter-packet intervals is similar, except that the domain of possible intervals is a continuum, so we must map this range onto the discrete index values $\left\{1,\dots,d_v\right\}$.
-In our experiments, we use the following adaptive quantization function:
+The representation of inter-packet intervals as a vector, $\vec{v}$, is similar, but the domain of possible intervals is a continuum, so we must map this range onto the discrete index values $\left\{1,\dots,d_v\right\}$.
+We use the following adaptive quantization function:
 \begin{align}
@@
 \end{align}
-The parameter values function must be chosen.
+The element $v_{ij}$ is the number of intervals, $v$, for the $i$th flow such that $\Qi(v)=i$.
+The parameter values defining $\Qi$ must be chosen by the experimenter.
 In our case, the time measurements had microsecond precision, so $v_{\min}=10^{-6}$.
@@
 The number of dimensions was chosen to match the dimensions for packet size: $d_v=1500$.
-Finally, the parameter $\gamma\in(0,1)$ allows us to fine tune how the resolution scales with interval size.
-We have chosen $\gamma=\fracx{1}{3}$ because this choice provides resolution that scales smoothly from the microseconds for small intervals to seconds for large intervals.
+Finally, the parameter $\gamma\in(0,1)$ allows us to fine-tune how the resolution scales with interval size; we have chosen $\gamma=\fracx{1}{3}$ because this choice provides resolution that scales smoothly from the microseconds for small intervals to seconds for large intervals.
 
@@
 
-Given an instance of the general matrix model, how do we generate an actual workload form it?
+Given an instance of the general matrix model, how do we produce an actual workload form it?
 Each row of the model is used to randomly generate the behavior of a single flow.
-The source and destination nodes of the flow are randomly chosen using histogram vectors $\vec{s}$ and $\vec{d}$ the proportional probabilities of each node being chosen.\footnote{Since a flow must have distinct source and destination nodes, this sampling must be done jointly, avoiding the collisions.
-The simplest technique is the rejection method: if the same node is chosen for both source and destination, reject that choice and choose both nodes again.}
-For start and end times, a bin for each is chosen with weight given by the behavior histogram; within each bin, a random value is uniformly chosen from the range of possible values.\footnote{The start time must precede the end time, so these choices must also be made jointly, enforcing this restriction.
+The source and destination nodes of the flow are randomly chosen, using histogram vectors $\vec{s}$ and $\vec{d}$ as the proportional probabilities of each node being chosen.\footnote{Since a flow must have distinct source and destination nodes, this sampling must be done jointly, avoiding the collisions.
+The simplest technique is the rejection method: if the same node is chosen for both source and destination, reject that pair and begin the selection process again.}
+For start and end times, a bin for each is chosen with probability proportional to the vector values; given a bin, a random value is uniformly chosen from the range of values in the bin.\footnote{The start time must precede the end time, so these choices must also be made jointly, enforcing this restriction.
 The rejection method is used.}
-Each flow packet sequence is then generated, beginning at the chosen start time.
-The size of each packet is sampled from the histogram of packet sizes.
-The interval to wait before the next packet is determined from the behavior vector, $\vec{v}$: an index, $k$ is chosen, with probability proportional to the value, $\vec{v}_k$; a uniform value, $u$ is chosen from $[0,1)$, and $(k,u)$ is transformed back into a time value using the ``dequantization'' function:
+The first packet is sent at chosen start time; the size of each packet is sampled from the histogram of packet sizes, $\vec{z}$; the interval until the next packet is determined from the interval vector, $\vec{v}$.
+Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}_k$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\seq{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
 \begin{align}
 \label{eqn:interval-dequantization}
-\Di(k,u)=v_{\min}+(v_{\max}-v_{\min})\left(\frac{k+u}{d_v}\right)^{\frac{1}{\gamma}}.
-\end{align}
-
-%For example, if there are 3 nodes in a workload, and the source vector for a flow is $\vec{s}=(3,7,1)$, we choose a source node $k$ with $\Pr(k=1)=\frac{3}{11}$, $\Pr(k=2)=\frac{7}{11}$, and $\Pr(k=3)=\frac{1}{11}$.
-For inter-packet intervals, a bin is chosen in the same manner, and a random value from the range represented by the bin is generated.
+\Di(k,u)=v_{\min}+(v_{\max}-v_{\min})\left(\frac{k-u}{d_v}\right)^{\frac{1}{\gamma}}.
+\end{align}
+The packet stream ends when a packet would be sent after the chosen end time of the flow.
 

### 2008-02-04 paper.tex
Work on common model expressions.

@@
 \newcommand{\ones}[1]{\mathbf{1}_{#1}}
+\newcommand{\zeros}[1]{\mathbf{0}_{#1}}
 
@@
 
-\subsection{Communication Topology}
+\subsection{Nodes: Communication Topology}
+\label{sec:communication-topology}
 
@@
 
-\subsection{Flow Packet Count, Size \& Duration}
+\subsection{Flows: Packet Count, Size \& Duration}
 \label{sec:flow-totals}
@@
 \begin{align}
-\vec{\sigma}&=(1,2,3,\dots,d_z),\\
+\vec{\sigma}&=\seq{1,2,3,\dots,d_z},\\
 \vec{\tau}&=\seq{\Di\left(k,\fracx{1}{2}\right)}_{k=1}^{d_v}.
 \end{align}
-Using these vectors of representative values, the vector of packet-counts is $Z\ones{d_z}=V\ones{d_v}$, the vector of flow sizes vector is $Z\vec{\sigma}$, and the vector of durations is $V\vec{\tau}$.
+Using these representative values, the vector of packet-counts is $Z\ones{d_z}=V\ones{d_v}$, the vector of flow sizes vector is $Z\vec{\sigma}$, and the vector of approximate durations is $V\vec{\tau}$.
+
+\subsection{Network: Marginal Behavior}
+\label{sec:marginal-behavior}
+
+\noindent
+...
 
@@
 It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees, largely because of the high complexity of doing otherwise and the lack of compelling alternatives.
-Here we examine how various notions of uniformity are represented in the matrix model.
+Here we examine how various notions of uniformity are expressed in the matrix model.
 In general, \textit{intra-flow uniformity} is effected by right multiplication, which mixes the elements in each row, whereas \textit{inter-flow uniformity} is effected by left multiplication, which mixes elements across rows.
-The standard constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals.
-Each flow's bit-rate is therefore constant, and hence the name.
-The strict \caps{CBR} model can be expressed simply using the packets-bytes-duration model discussed in \Section{representations}.
-Deriving the packet-bytes-duration model from the general matrix model is accomplished through right multiplication as described in \Section{flow-totals}.
+The constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
+The strict \caps{CBR} model can be expressed simply using the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
+The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma$, where
+\begin{align}
+\Gamma=\begin{pmatrix}
+\ones{} & \vec{\sigma} & \zeros{} \\
+\zeros{} & \zeros{} & \vec{\tau}
+\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
+\end{align}
+
+%Deriving the packet-bytes-duration model from the general matrix model is accomplished through right multiplication as described in \Section{flow-totals}.
+% TODO: this conclusion is pretty weak.
 

### 2008-02-04 paper.tex
Switching over to coding...

@@
 
-The traffic workload pattern in a network is simply a collection of \caps{IP} packets between hosts in the network sent at certain times.
+The traffic workload pattern in a network is a collection of \caps{IP} packets between hosts in the network sent at certain times.
 It is standard in network analysis, however, to aggregate sequences of packets sharing the same \caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP} or \caps{UDP} port numbers.
@@
 The constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
-The strict \caps{CBR} model can be expressed simply using the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma$, where
+The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
+The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma\in\R^{f \times 3}$, where
 \begin{align}
@@
 
-While constant bit-rate in the strict sense implies only that the packet behavior in each flow is uniform, it is extremely rare for it to be applied in this strict sense only; it is typical that \caps{CBR} models make assumptions about uniformity across flows, and not only within each flow.
+While constant bit-rate in the strict sense implies only that the packet behavior in each flow is uniform, it is very rare for it to be applied in this strict sense only; it is typical that ``\caps{CBR}'' models make assumptions about uniformity across flows, and not just within each flow.
 The most drastic simplification is to treat all flows in a synthetic workload identically.
-This is effected by left multiplication: $F'=\ones{f \times f} F$.
-In general, we can express uniformity through left-multiplication by some matrix.
+This is effected by left multiplication: $F'=(\ones{f \times f}/f)F$.
 To make flow behavior uniform at the per-source node level, we left-multiply by $SS^t/f\in\R^{f \times f}$: $F'=(SS^t/f)F$.
@@
 There is no armchair answer: only analysis of real data can give us the answer.
+The matrix model provides an analytical framework for doing so.
 

### 2008-02-04 paper.tex
Added a TODO; other minor edits.

@@
 With \caps{TCP}, on the other hand, network events are known to affect the transport layer
+
+% TODO: talk about other vector-space representation: perfect representation, markov models, ... others?
 
@@
 The first packet is sent at chosen start time; the size of each packet is sampled from the histogram of packet sizes, $\vec{z}$; the interval until the next packet is determined from the interval vector, $\vec{v}$.
-Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}_k$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\seq{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
+Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}(k)$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\seq{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
 \begin{align}

### 2008-02-05 paper.tex
Graphs!!!

@@
 
-%\begin{abstract}
-%\input{abstract.tex}
-%\end{abstract}
+\begin{abstract}
+We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in wireless networks.
+Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematical precision and succinctness, while simultaneously being made clearer and more general.
+We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to discredit the common constant bit-rate model.
+A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
+We conclude with future directions for modeling research based on the new algebraic framework.
+\end{abstract}
+
+\section{Introduction}\label{sec:intro}
+
+%Most of what are called traffic models are only partial descriptions of one or a handful of aspects of a total pattern of traffic in a network.
+The ubiquitous constant bit-rate (\caps{CBR}) model, for example, merely describes the nature of the packets in each flow in a workload: in a given flow, the sizes of packets and the intervals between them are are all uniform.
+Hence, the transmitted bit-rate is constant.
+
+%Two things are require for the direct empirical comparison of traffic models:
+%\begin{itemize}
+%\item An objective standard for comparing quality of models.
+%% WinMee/Broadnets provide this
+%\item A common framework in which different models can be implemented and compared.
+%% this paper provides this
+%\end{itemize}
+
+More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and billions of computers on the Internet and in local area networks~(\caps{LAN}s).
+The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
+Other research~\cite{Hernandez06:dissertation} has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback induced by the ubiquitous transmission control protocol~(\caps{TCP}).
+Recently, the high-level statistical properties of wireless \caps{LAN} traffic have been rigorously modeled with a variety of parametric distributions~\cite{Hernandez06:wlan-traffic}.
+
+Despite great breakthroughs in understanding and methodology, the networking community remains without a cookbook for generating realistic \caps{LAN} workloads.
+Drastic assumptions of uniformity abound in typical experimental workload generation.
+Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily.
+While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
+the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better.
+It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.
+
+We believe that two major breakthroughs are needed to bring a much needed revolution to the world of traffic modeling.
+The first requirement is an objective standard of realism by which to compare traffic models.
+Such a standard has already been introduced in~\cite{Karpinski07:realism} and used to demonstrate that uniform traffic modeling, including the ubiquitous \caps{CBR} model, drastically distort performance metrics at all levels of the network protocol stack~\cite{Karpinski07:cbr-failure}.
+The second required breakthrough is a unified framework for expressing traffic models that allows researchers to systematically express and understand the assumptions and transformations entailed in each model.
+
+In this paper, we present a highly general model for representing detailed full-network traffic patterns algebraically.
+The General Matrix Model (\caps{GMM}) represents the behavior or each network flow as a vector in a high-dimensional linear space, and represents total network behavior as a large matrix whose rows are the behavior vectors.
+The crucial property of this representation is that it respects vector space operations: summation of behavior vectors gives an accurate description of the aggregate behavior of the flows represented by those vectors.
+The benefit of this representation is that it makes it immediately possible to express a broad variety of assumptions and conditions on traffic patterns in terms of simple matrix operations.
+It also allows provides great mathematical insight and allows us to bring all the tools of modern linear algebra to bear on the problem of traffic modeling.
+
+%Worse still, it is unclear what ``realistic'' should mean in this context.
+Foremost, a clear and testable definition of realism for traffic models is needed.
+In the absence of such a definition, it remains standard in wireless experiments to generate traffic using the simplest possible model: all nodes, flows and packets are assumed to behave uniformly.
+Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily.
+While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
+%the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better.
+It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.
+
+%To address this foundational problem in network traffic modeling, we have introduced a fundamentally different approach to evaluating the realism of workload models~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics.
+We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values similar to those observed using the original trace to generate traffic.
+Using this new approach, we explore the space of traffic models, establishing rigorously that the commonly used uniform models for node, flow and packet behavior drastically misrepresent important \caps{WLAN} performance metrics at all levels of the protocol stack.
+
+%It is natural to compare these common, na\"ive models with the high-level parametric models proposed by \FHC~\cite{Hernandez06:wlan-traffic}.
+Here, however, we face another fundamental challenge in traffic modeling.
+Even with specific models for various aspects of network behavior, it remains non-trivial to combine them into a coherent composite model capable of generating concrete workloads.
+The {\FHC} models, for example, specify the distribution of flows per node in the network; since every flow has two end-points, however, it remains unclear how to assign end-points to flows such that the marginal distribution of flow per node matches the BiPareto model indicated.
+Similarly, the models dictate BiPareto marginal distribution of flow sizes across the entire network, but say nothing about how these sizes are to be distributed among the nodes.
+We could assume that the number of flows per node is independent from the size of the flows at that node, but, as we shall see, this assumption is invalid.
+Moreover, the details of flow behavior---packet sizes and inter-packet intervals---remain entirely undefined.
+
+%Our point is not to denigrate the models proposed in~\cite{Hernandez06:wlan-traffic} in any way.
+On the contrary: these parametric models are the state of the art in modeling traffic patterns for wireless local area networks.
+The purpose, rather, is to point out how large a gap still remains between having accurate high-level models of marginal aspects of network behavior and being able to generate usable workloads with those models.
+Moreover, a great deal of ambiguity lies in that gap since there is no single obvious way to combine models.
+
+%To clarify, disambiguate and standardize the specification of generative local area traffic models, we propose the \textit{Traffic Object Model} (\textit{\caps{TOM}}).
+The \caps{TOM} entails a hierarchy of abstract classes: \class{Node}, \class{Flow}, and \class{Packet} are the primary classes; the \class{Network} class collects everything.
+Each abstract class defines an interface of properties.
+A traffic model defines concrete classes at each level that provide the required interfaces; each class defines its own properties by delegation, inheritance, or overriding properties of other classes.
+Because of the separation of the \caps{TOM} into well-defined levels, each concrete class naturally defines a sub-model, specific to that level's behavior; sub-models can be mixed and matched combinatorially.
+
+%We define a range of sub-models at each level of the \caps{TOM}, with decreasing similarity to the original trace behavior: \class{Trace} (reproduce original behavior), \class{Nonparametric}, \class{Parametric}, and \class{Uniform}.
+The control model uses the Trace sub-model at every level, exactly reproducing the behavior recorded from the original network.
+The standard na\"ive model uses the \class{Uniform} sub-model at all levels.
+The models proposed by {\FHC} are implemented by applying the \class{Parametric} sub-models at all levels.
+The \class{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.
+
+%: objects and their properties together with rules for delegating, inheriting, and overriding the properties across the objects in the \caps{TOM} hierarchy. The \caps{TOM} allows
+
+%The {\FHC} models, like most network behavior models, are marginal, meaning that they are fitted to data aggregated of over all nodes in the network, and not fitted on a per-node basis.
+With marginal models, since all information about interactions between the models is lost, so we must apply each model independently and hope that the result is good enough.
+
+%Before we can define traffic models, however, we face the next major fundamental challenge in traffic modeling.
+
+%In this paper, we address this fundamental problem in network traffic modeling by presenting an 
+%In this paper, we propose objective criteria by which to judge whether
+
+%The experimental evaluation of new wireless technologies inevitably requires network traffic.
+The most commonly used models for generating wireless local-area network (\caps{WLAN}) workload are simple, treating both nodes and flows uniformly, and are generally understood to be na\"ive.
+They remain popular because of the lack of compelling alternatives.
+The scope of impact of different workloads on network performance is largely unresearched and thus, poorly understood.
+We believe that one of the major handicaps preventing wireless experiments from more accurately predicting real-world performance is the lack of realistic models of \caps{WLAN} workload.
+
+%The evaluation of wireless technology requires the generation of workload to test the viability and performance of the new protocol or technique being studied.
+We believe that lack of realism in traffic workload generation is one of the major limiting factors that prevents simulations and experimental test-bed deployments from accurately predicting the real-world performance of wireless technologies.
+Today, very little is understood about the impact of different workloads on network performance.
+Uniform constant bit-rate traffic (\caps{CBR}) is commonly used to evaluate protocols, but there is no evidence that behavior under such workloads is an accurate predictor of performance under real usage patterns.
+The inability to experimentally forecast real-world performance is a severe handicap to the entire networking community.
+It hinders the ability to effectively develop better solutions to the many difficult problems that face emerging wireless technologies.
+
+%One of the hardest problems in modeling phenomena as complex as network traffic patterns lies in determining which characteristics of any given exemplar of behavior are essential and which can safely be abstracted away.
+In \caps{WLAN} traffic modeling, important characteristics have largely been determined based on intuition.
+In previous work~\cite{Karpinski07:realism}, we introduced a fundamentally different approach to evaluating the realism of workload models.
+Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics.
+We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values that are similar to those observed using the original trace to generate traffic.
+Using this new approach, we explore the space of traffic models, establishing rigorously for the first time that common but simplistic models of \caps{WLAN} traffic drastically misrepresent important performance metrics at all levels of the protocol stack.
+
+%Our analysis shows that the standard constant bit-rate (\caps{CBR}) traffic model, with uniformly chosen nodes and uniform flows of traffic, significantly misrepresents important performance metrics at every level of the network.
+To qualify these results in more depth, we also investigate partially synthetic variants of this standard model, allowing us to reason about which aspects of workload models have the most impact.
+For the most commonly used model, the delay of application data traversing the network from sender to receiver is underestimated by more than a factor of 5, \textit{on average}, and by more than a factor of 11 in a quarter of usage scenarios.
+Network control overhead is overestimated, on average, by a factor of about 2.5, while link control overhead is underestimated by almost a factor of 2.
+Even for metrics where the average and median misrepresentations are not so extreme, the statistical characteristics of error values often indicate that something is fundamentally unrealistic.
+None of the partially synthetic models manage to accurately represent more than one or two performance characteristics.
+We further use the \caps{AODV}~\cite{rfc:aodv} and \caps{OLSR}~\cite{rfc:olsr} ad~hoc routing protocols to show that the relative performance of protocols can be inverted when changing from using a simplistic traffic model to using real traffic: using random end-point, uniform \caps{CBR} traffic, \caps{AODV} appears to induce less link-layer overhead than \caps{OLSR}, whereas using real traffic, it in fact induces more.
+
+%The first major contribution of this work is the definition of ``sufficient realism,'' together with the analytical and statistical methodology to rigorously test whether synthetic traffic models meet this definition.
+The second contribution is the conclusion that the most commonly used model of wireless traffic drastically and consistently misrepresents some of the most important metrics for performance evaluation of wireless protocols.
+Many performance comparisons based on this model may need revisiting.
+The last and most important contribution is the collection of ideas and analytical tools necessary to create more realistic synthetic traffic models in the future.
+By applying the methods developed in this work, it will be possible to discover precisely which aspects of network usage affect the realism of performance results.
+Once this is known, it will become feasible to create models that accurately reproduce those aspects of real wireless usage.
+
+The rest of the paper is organized as follows.
+In \Section{motivation} we present motivation and related work.
+The General Matrix Model is presented in \Section{general-matrix-model}.
+Our experimental and analytical methodology for evaluating traffic models presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
+The ramifications of these results are discussed in \Section{discussion}.
+Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.
+
+\section{Motivation \& Related Work}\label{sec:motivation}\label{sec:related-work}
+
+The history of networking research contains many examples of simplistic models that have proven not only to be inaccurate, but also to drastically skew important characteristics of network behavior.
+Paxson and Floyd showed that the Poisson packet arrival model, which had been standard for studying wide-area Internet traffic, failed to capture the burstiness and self-similarity of real traffic~\cite{Paxson95}.
+The equally common but simplistic Random Way-Point~(\caps{RWP}) mobility model was found to exhibit ``density waves'' and gradual slow-down of average node speed~\cite{Royer01,Yoon03:speed-decay}.
+In the worst cases, overly simplistic models can switch the relative performance of protocols, thereby invalidating the conclusions drawn from performance comparisons using those models.
+
+The interaction of wireless users and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted.
+The joint pattern of traffic generation and mobility through time and space completely determines the effect of wireless usage on the lower levels of the network.
+This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner.%~\cite{Clark88}.
+\footnote{This is violated by some quality of service (QoS) schemes.
+However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid.
+The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.} The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic and motion in the experiments are sufficiently realistic.
+
+Paxson and Floyd observed in~\cite{Paxson95} that the interplay between end-point behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback.
+Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}.
+This presents a fundamental difficulty, however, in that it presumes that we know the intent of end-points: what \textit{would} they have done under different conditions?
+While we can speculate about what an individual node might hypothetically do, we currently do not understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations.
+Especially in the wireless setting, a fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback.
+Accordingly, in this paper we attempt to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without multi-level feedback.
+It is important to realize that while this does not provide a final picture, we currently lack even a first-order understanding of the effect of different workloads on performance.
+This first-order understanding is an essential initial step.
+
+There have been a significant number of studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}.
+These analyses have described a wide variety of aspects of wireless network behavior, and provide much insight into the workings of real, deployed wireless networks.
+These studies present a broad analysis of general system features and trends of specific corporate wireless local-area networks (\caps{WLAN}s)~\cite{Tang99,Balazinska03}, university campus \caps{WLAN}s~\cite{Tang00,Kotz02,Chinchilla04,Schwab04,Henderson04,Tuduce05}, and temporary \caps{WLAN}s at conference venues~\cite{Balachandran02,Jardosh05:ewind}.
+They also provide a large body of raw data for subsequent analysis and modeling research.
+Our work provides the methodology for turning this rich foundation of field data into usable, realistic models of workload for a wide variety of networking situations.
+
+The choice of mobility models for mobile wireless simulations can have a drastic impact on important performance metrics~\cite{Camp02,Yoon03:speed-decay,Yoon03:sound-models,Jardosh03,Zheng04}.
+Moreover, commonly used but simplistic mobility models, such as \caps{RWP}, exhibit characteristics, including density waves and speed decay, that are categorically dissimilar from any known real-world behavior~\cite{Royer01,Yoon03:speed-decay,Yoon03:sound-models}.
+In response to this evidence, more realistic mobility models have been proposed~\cite{Jardosh03,Tuduce05,Jardosh05:voronoi}.
+While much of this work focuses on making models that are simply more intuitively appealing~\cite{Jardosh03,Jardosh05:voronoi}, some work has begun to capitalize on this newly created wealth of wireless field data by deriving models from observed usage behavior, rather than intuition alone~\cite{Balazinska03,Tuduce05}.
+%The analysis of Balazinska \textit{et al.}~\cite{Balazinska03} introduces the notions of the prevalence and persistence of an \caps{AP} with respect to mobile nodes.\footnote{These concepts are adapted from Paxson's work on Internet routing behavior: ``\textit{prevalence} meaning the overall likelihood that a particular route is encountered, and \textit{persistence}, the likelihood that a route remains unchanged over a long period of time''~\cite{Paxson96}.} Tuduce \textit{et al.}~\cite{Tuduce05} have based their mobility model on the persistence and prevalence characteristics of real trace data, which makes their work particularly relevant to our definition of sufficient realism.
+They do not, however, determine the impact of prevalence and persistence characteristics on performance metrics.
+
+In this paper, instead of mobility, we examine an even more fundamental aspect of user behavior in wireless networks: the pattern of traffic generated by users and applications.
+This aspect of behavior is more fundamental because it applies to all types of wireless networks, not just mobile and ad~hoc networks.
+Moreover, the effect of traffic patterns applies not only to simulations, but also to experimental deployments, which have become the gold standard for wireless protocol evaluation.
+Experimental deployments sidestep the problem of accurately modeling the lower layers of the network.
+Unless traffic and mobility are modeled realistically, however, the experimental results will still be unreliable.
+
+There is a large and diverse body of work on traffic analysis, modeling, and generation~\cite{Paxson95,Paxson96,Sommers04,Avallone04,Hernandez06:dissertation}.
+We are only able to discuss a small, but hopefully representative sampling of this work.
+Almost all of the traffic generation work has focused on wide-area Internet backbone traffic.
+%Because of the comparatively unlimited capacity of wired networks, traffic workload generally does not offer interesting technical challenges at the edge of the network.
+The two most prominent traffic generation frameworks are Harpoon and \caps{D-ITG}.
+Harpoon~\cite{Sommers04} uses a traffic trace for self-training, and can subsequently generate synthetic traffic with certain statistical properties based on the original trace.
+The properties reproduced are the empirical distributions of the following: ``file size, inter-connection time, source and destination IP ranges, number of active sessions.'' There is no criterion proposed to determine whether these properties characterize the original traffic adequately---we can only hope that this approximation is good enough.
+For many purposes, it likely is sufficient; in particular, it is probably appropriate for the intended use in generating traffic for Internet backbone simulations.
+Wireless networks, however, are particularly sensitive to workload conditions, and sampling from a limited set of empirical distributions does not suffice to reproduce realistic network-wide traffic.
+
+\caps{D-ITG}~\cite{Avallone04} generates flows using an independent sampling model for packet sizes and inter-packet intervals.
+The framework contains pre-made models for several common types of Internet traffic.
+The focus of this project, however, is on providing the infrastructure to generate very large volumes of synthetic Internet-like backbone traffic.
+No analysis is provided for determining the realism of traffic mixes, or for choosing flow end-points realistically.
+In wireless networks, these factors are of crucial importance to performance, and cannot be overlooked.
+Both Harpoon and \caps{D-ITG} provide excellent traffic generation platforms, but do not provide a systematic framework for understanding or reproducing realistic whole-network workload in the wireless setting.
+
+%This work extends that published in~\cite{Karpinski07:realism}~and~\cite{Karpinski07:cbr-failure}, using the same methodology for defining and quantifying the realism of synthetic traffic models.
+The Traffic Object Model builds on the combinatorial approach taken in~\cite{Karpinski07:cbr-failure}, while providing an flexible and general object-oriented framework for constructing generative traffic models.
+The traffic models evaluated here overlap and augment those evaluated in both prior works; comparison with the models proposed in {\FHC} is entirely new.
+The Qualnet simulation methodology has been enhanced by the implementation of full-duplex, client-server, trace-driven raw \caps{IP}, \caps{TCP}, and \caps{UDP} flows.
+We have also implemented performance metric aggregation based on the \caps{IETF} framework for \caps{IP} performance metrics~\cite{rfc:ip-metrics,rfc:jitter}.
+
+% TODO: add in the Hernandez-Cortes' dissertation.
+% TODO: concluding remarks about what's missing from traffic work in general.
+%   + there's no central goal or theme---because realism is ill-defined and subjective
+%   + there are no rigorous tests for whether realism is achieved since it's ill-defined
+%   + in wireless, where the impact is greater, there's been virtually no research of impact
+% TODO: cite a bunch of papers that use the CBR traffic model.
 
@@
 \section{The General Matrix Model}
+\label{sec:general-matrix-model}
 \label{sec:representations}
@@
 
-\subsection{Network: Marginal Behavior}
-\label{sec:marginal-behavior}
-
-\noindent
-...
+%\subsection{Network: Marginal Behavior}
+%\label{sec:marginal-behavior}
+
+%\noindent
+%...
 
@@
 
-\subsection{Uniformity}
+\subsection{Uniform Models}
 
@@
 \begin{align}
-F' = ((1-\alpha)(SS^t)+\alpha(DD^t))F.
+F' = \frac{1}{f}((1-\alpha)(SS^t)+\alpha(DD^t))F.
 \end{align}
 The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors.
-Source-only behavior is given by $\alpha=0$, for destination-only set $\alpha=1$, and for the standard average, set $\alpha=\frac{1}{2}$.
+Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.
 
@@
 
-\section{Introduction}\label{sec:intro}
-
-We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in local area networks (\caps{LAN}s).
-Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematic precision and succinctness, while simultaneously being made clearer and more general.
-
-Most of what are called traffic models are actually only partial descriptions of one or a handful of aspects of a complete model.
-The ubiquitous constant bit-rate (\caps{CBR}) model, for example, merely describes the nature of the packets in each flow in a workload: in a given flow, the sizes of packets and the intervals between them are are all uniform.
-Hence, the transmitted bit-rate is constant.
-
-Two things are require for the direct empirical comparison of traffic models:
-\begin{itemize}
-\item An objective standard for comparing quality of models.
-% WinMee/Broadnets provide this
-\item A common framework in which different models can be implemented and compared.
-% this paper provides this
-\end{itemize}
-
-More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and billions of computers on the Internet and in local area networks~(\caps{LAN}s).
-The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
-Other research~\cite{Hernandez06:dissertation} has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback induced by the ubiquitous transmission control protocol~(\caps{TCP}).
-Recently, the high-level statistical properties of wireless \caps{LAN} traffic have been rigorously modeled with a variety of parametric distributions~\cite{Hernandez06:wlan-traffic}.
-
-Despite great breakthroughs in understanding and methodology, the networking community remains without a cookbook for generating realistic \caps{LAN} workloads.
-Worse still, it is unclear what ``realistic'' should mean in this context.
-Foremost, a clear and testable definition of realism for traffic models is needed.
-In the absence of such a definition, it remains standard in wireless experiments to generate traffic using the simplest possible model: all nodes, flows and packets are assumed to behave uniformly.
-Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily.
-While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
-the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better.
-It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.
-
-To address this foundational problem in network traffic modeling, we have introduced a fundamentally different approach to evaluating the realism of workload models~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics.
-We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values similar to those observed using the original trace to generate traffic.
-Using this new approach, we explore the space of traffic models, establishing rigorously that the commonly used uniform models for node, flow and packet behavior drastically misrepresent important \caps{WLAN} performance metrics at all levels of the protocol stack.
-
-It is natural to compare these common, na\"ive models with the high-level parametric models proposed by \FHC~\cite{Hernandez06:wlan-traffic}.
-Here, however, we face another fundamental challenge in traffic modeling.
-Even with specific models for various aspects of network behavior, it remains non-trivial to combine them into a coherent composite model capable of generating concrete workloads.
-The {\FHC} models, for example, specify the distribution of flows per node in the network; since every flow has two end-points, however, it remains unclear how to assign end-points to flows such that the marginal distribution of flow per node matches the BiPareto model indicated.
-Similarly, the models dictate BiPareto marginal distribution of flow sizes across the entire network, but say nothing about how these sizes are to be distributed among the nodes.
-We could assume that the number of flows per node is independent from the size of the flows at that node, but, as we shall see, this assumption is invalid.
-Moreover, the details of flow behavior---packet sizes and inter-packet intervals---remain entirely undefined.
-
-Our point is not to denigrate the models proposed in~\cite{Hernandez06:wlan-traffic} in any way.
-On the contrary: these parametric models are the state of the art in modeling traffic patterns for wireless local area networks.
-The purpose, rather, is to point out how large a gap still remains between having accurate high-level models of marginal aspects of network behavior and being able to generate usable workloads with those models.
-Moreover, a great deal of ambiguity lies in that gap since there is no single obvious way to combine models.
-
-To clarify, disambiguate and standardize the specification of generative local area traffic models, we propose the \textit{Traffic Object Model} (\textit{\caps{TOM}}).
-The \caps{TOM} entails a hierarchy of abstract classes: \class{Node}, \class{Flow}, and \class{Packet} are the primary classes; the \class{Network} class collects everything.
-Each abstract class defines an interface of properties.
-A traffic model defines concrete classes at each level that provide the required interfaces; each class defines its own properties by delegation, inheritance, or overriding properties of other classes.
-Because of the separation of the \caps{TOM} into well-defined levels, each concrete class naturally defines a sub-model, specific to that level's behavior; sub-models can be mixed and matched combinatorially.
-
-We define a range of sub-models at each level of the \caps{TOM}, with decreasing similarity to the original trace behavior: \class{Trace} (reproduce original behavior), \class{Nonparametric}, \class{Parametric}, and \class{Uniform}.
-The control model uses the Trace sub-model at every level, exactly reproducing the behavior recorded from the original network.
-The standard na\"ive model uses the \class{Uniform} sub-model at all levels.
-The models proposed by {\FHC} are implemented by applying the \class{Parametric} sub-models at all levels.
-The \class{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.
-
-%: objects and their properties together with rules for delegating, inheriting, and overriding the properties across the objects in the \caps{TOM} hierarchy. The \caps{TOM} allows
-
-%The {\FHC} models, like most network behavior models, are marginal, meaning that they are fitted to data aggregated of over all nodes in the network, and not fitted on a per-node basis.
-With marginal models, since all information about interactions between the models is lost, so we must apply each model independently and hope that the result is good enough.
-
-%Before we can define traffic models, however, we face the next major fundamental challenge in traffic modeling.
-
-%In this paper, we address this fundamental problem in network traffic modeling by presenting an 
-%In this paper, we propose objective criteria by which to judge whether
-
-%The experimental evaluation of new wireless technologies inevitably requires network traffic.
-The most commonly used models for generating wireless local-area network (\caps{WLAN}) workload are simple, treating both nodes and flows uniformly, and are generally understood to be na\"ive.
-They remain popular because of the lack of compelling alternatives.
-The scope of impact of different workloads on network performance is largely unresearched and thus, poorly understood.
-We believe that one of the major handicaps preventing wireless experiments from more accurately predicting real-world performance is the lack of realistic models of \caps{WLAN} workload.
-
-%The evaluation of wireless technology requires the generation of workload to test the viability and performance of the new protocol or technique being studied.
-We believe that lack of realism in traffic workload generation is one of the major limiting factors that prevents simulations and experimental test-bed deployments from accurately predicting the real-world performance of wireless technologies.
-Today, very little is understood about the impact of different workloads on network performance.
-Uniform constant bit-rate traffic (\caps{CBR}) is commonly used to evaluate protocols, but there is no evidence that behavior under such workloads is an accurate predictor of performance under real usage patterns.
-The inability to experimentally forecast real-world performance is a severe handicap to the entire networking community.
-It hinders the ability to effectively develop better solutions to the many difficult problems that face emerging wireless technologies.
-
-%One of the hardest problems in modeling phenomena as complex as network traffic patterns lies in determining which characteristics of any given exemplar of behavior are essential and which can safely be abstracted away.
-In \caps{WLAN} traffic modeling, important characteristics have largely been determined based on intuition.
-In previous work~\cite{Karpinski07:realism}, we introduced a fundamentally different approach to evaluating the realism of workload models.
-Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics.
-We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values that are similar to those observed using the original trace to generate traffic.
-Using this new approach, we explore the space of traffic models, establishing rigorously for the first time that common but simplistic models of \caps{WLAN} traffic drastically misrepresent important performance metrics at all levels of the protocol stack.
-
-%Our analysis shows that the standard constant bit-rate (\caps{CBR}) traffic model, with uniformly chosen nodes and uniform flows of traffic, significantly misrepresents important performance metrics at every level of the network.
-To qualify these results in more depth, we also investigate partially synthetic variants of this standard model, allowing us to reason about which aspects of workload models have the most impact.
-For the most commonly used model, the delay of application data traversing the network from sender to receiver is underestimated by more than a factor of 5, \textit{on average}, and by more than a factor of 11 in a quarter of usage scenarios.
-Network control overhead is overestimated, on average, by a factor of about 2.5, while link control overhead is underestimated by almost a factor of 2.
-Even for metrics where the average and median misrepresentations are not so extreme, the statistical characteristics of error values often indicate that something is fundamentally unrealistic.
-None of the partially synthetic models manage to accurately represent more than one or two performance characteristics.
-We further use the \caps{AODV}~\cite{rfc:aodv} and \caps{OLSR}~\cite{rfc:olsr} ad~hoc routing protocols to show that the relative performance of protocols can be inverted when changing from using a simplistic traffic model to using real traffic: using random end-point, uniform \caps{CBR} traffic, \caps{AODV} appears to induce less link-layer overhead than \caps{OLSR}, whereas using real traffic, it in fact induces more.
-
-%The first major contribution of this work is the definition of ``sufficient realism,'' together with the analytical and statistical methodology to rigorously test whether synthetic traffic models meet this definition.
-The second contribution is the conclusion that the most commonly used model of wireless traffic drastically and consistently misrepresents some of the most important metrics for performance evaluation of wireless protocols.
-Many performance comparisons based on this model may need revisiting.
-The last and most important contribution is the collection of ideas and analytical tools necessary to create more realistic synthetic traffic models in the future.
-By applying the methods developed in this work, it will be possible to discover precisely which aspects of network usage affect the realism of performance results.
-Once this is known, it will become feasible to create models that accurately reproduce those aspects of real wireless usage.
-
-The rest of the paper is organized as follows.
-In \Section{motivation} we present motivation and related work.
-Our experimental and analytical methodology is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
-The ramifications of these results are discussed in \Section{discussion}.
-Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.
-
-\section{Motivation \& Related Work}\label{sec:motivation}\label{sec:related-work}
-
-The history of networking research contains many examples of simplistic models that have proven not only to be inaccurate, but also to drastically skew important characteristics of network behavior.
-Paxson and Floyd showed that the Poisson packet arrival model, which had been standard for studying wide-area Internet traffic, failed to capture the burstiness and self-similarity of real traffic~\cite{Paxson95}.
-The equally common but simplistic Random Way-Point~(\caps{RWP}) mobility model was found to exhibit ``density waves'' and gradual slow-down of average node speed~\cite{Royer01,Yoon03:speed-decay}.
-In the worst cases, overly simplistic models can switch the relative performance of protocols, thereby invalidating the conclusions drawn from performance comparisons using those models.
-
-The interaction of wireless users and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted.
-The joint pattern of traffic generation and mobility through time and space completely determines the effect of wireless usage on the lower levels of the network.
-This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner.%~\cite{Clark88}.
-\footnote{This is violated by some quality of service (QoS) schemes.
-However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid.
-The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.} The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic and motion in the experiments are sufficiently realistic.
-
-Paxson and Floyd observed in~\cite{Paxson95} that the interplay between end-point behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback.
-Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}.
-This presents a fundamental difficulty, however, in that it presumes that we know the intent of end-points: what \textit{would} they have done under different conditions?
-While we can speculate about what an individual node might hypothetically do, we currently do not understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations.
-Especially in the wireless setting, a fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback.
-Accordingly, in this paper we attempt to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without multi-level feedback.
-It is important to realize that while this does not provide a final picture, we currently lack even a first-order understanding of the effect of different workloads on performance.
-This first-order understanding is an essential initial step.
-
-There have been a significant number of studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}.
-These analyses have described a wide variety of aspects of wireless network behavior, and provide much insight into the workings of real, deployed wireless networks.
-These studies present a broad analysis of general system features and trends of specific corporate wireless local-area networks (\caps{WLAN}s)~\cite{Tang99,Balazinska03}, university campus \caps{WLAN}s~\cite{Tang00,Kotz02,Chinchilla04,Schwab04,Henderson04,Tuduce05}, and temporary \caps{WLAN}s at conference venues~\cite{Balachandran02,Jardosh05:ewind}.
-They also provide a large body of raw data for subsequent analysis and modeling research.
-Our work provides the methodology for turning this rich foundation of field data into usable, realistic models of workload for a wide variety of networking situations.
-
-The choice of mobility models for mobile wireless simulations can have a drastic impact on important performance metrics~\cite{Camp02,Yoon03:speed-decay,Yoon03:sound-models,Jardosh03,Zheng04}.
-Moreover, commonly used but simplistic mobility models, such as \caps{RWP}, exhibit characteristics, including density waves and speed decay, that are categorically dissimilar from any known real-world behavior~\cite{Royer01,Yoon03:speed-decay,Yoon03:sound-models}.
-In response to this evidence, more realistic mobility models have been proposed~\cite{Jardosh03,Tuduce05,Jardosh05:voronoi}.
-While much of this work focuses on making models that are simply more intuitively appealing~\cite{Jardosh03,Jardosh05:voronoi}, some work has begun to capitalize on this newly created wealth of wireless field data by deriving models from observed usage behavior, rather than intuition alone~\cite{Balazinska03,Tuduce05}.
-%The analysis of Balazinska \textit{et al.}~\cite{Balazinska03} introduces the notions of the prevalence and persistence of an \caps{AP} with respect to mobile nodes.\footnote{These concepts are adapted from Paxson's work on Internet routing behavior: ``\textit{prevalence} meaning the overall likelihood that a particular route is encountered, and \textit{persistence}, the likelihood that a route remains unchanged over a long period of time''~\cite{Paxson96}.} Tuduce \textit{et al.}~\cite{Tuduce05} have based their mobility model on the persistence and prevalence characteristics of real trace data, which makes their work particularly relevant to our definition of sufficient realism.
-They do not, however, determine the impact of prevalence and persistence characteristics on performance metrics.
-
-In this paper, instead of mobility, we examine an even more fundamental aspect of user behavior in wireless networks: the pattern of traffic generated by users and applications.
-This aspect of behavior is more fundamental because it applies to all types of wireless networks, not just mobile and ad~hoc networks.
-Moreover, the effect of traffic patterns applies not only to simulations, but also to experimental deployments, which have become the gold standard for wireless protocol evaluation.
-Experimental deployments sidestep the problem of accurately modeling the lower layers of the network.
-Unless traffic and mobility are modeled realistically, however, the experimental results will still be unreliable.
-
-There is a large and diverse body of work on traffic analysis, modeling, and generation~\cite{Paxson95,Paxson96,Sommers04,Avallone04,Hernandez06:dissertation}.
-We are only able to discuss a small, but hopefully representative sampling of this work.
-Almost all of the traffic generation work has focused on wide-area Internet backbone traffic.
-%Because of the comparatively unlimited capacity of wired networks, traffic workload generally does not offer interesting technical challenges at the edge of the network.
-The two most prominent traffic generation frameworks are Harpoon and \caps{D-ITG}.
-Harpoon~\cite{Sommers04} uses a traffic trace for self-training, and can subsequently generate synthetic traffic with certain statistical properties based on the original trace.
-The properties reproduced are the empirical distributions of the following: ``file size, inter-connection time, source and destination IP ranges, number of active sessions.'' There is no criterion proposed to determine whether these properties characterize the original traffic adequately---we can only hope that this approximation is good enough.
-For many purposes, it likely is sufficient; in particular, it is probably appropriate for the intended use in generating traffic for Internet backbone simulations.
-Wireless networks, however, are particularly sensitive to workload conditions, and sampling from a limited set of empirical distributions does not suffice to reproduce realistic network-wide traffic.
-
-\caps{D-ITG}~\cite{Avallone04} generates flows using an independent sampling model for packet sizes and inter-packet intervals.
-The framework contains pre-made models for several common types of Internet traffic.
-The focus of this project, however, is on providing the infrastructure to generate very large volumes of synthetic Internet-like backbone traffic.
-No analysis is provided for determining the realism of traffic mixes, or for choosing flow end-points realistically.
-In wireless networks, these factors are of crucial importance to performance, and cannot be overlooked.
-Both Harpoon and \caps{D-ITG} provide excellent traffic generation platforms, but do not provide a systematic framework for understanding or reproducing realistic whole-network workload in the wireless setting.
-
-This work extends that published in~\cite{Karpinski07:realism}~and~\cite{Karpinski07:cbr-failure}, using the same methodology for defining and quantifying the realism of synthetic traffic models.
-The Traffic Object Model builds on the combinatorial approach taken in~\cite{Karpinski07:cbr-failure}, while providing an flexible and general object-oriented framework for constructing generative traffic models.
-The traffic models evaluated here overlap and augment those evaluated in both prior works; comparison with the models proposed in {\FHC} is entirely new.
-The Qualnet simulation methodology has been enhanced by the implementation of full-duplex, client-server, trace-driven raw \caps{IP}, \caps{TCP}, and \caps{UDP} flows.
-We have also implemented performance metric aggregation based on the \caps{IETF} framework for \caps{IP} performance metrics~\cite{rfc:ip-metrics,rfc:jitter}.
-
-% TODO: add in the Hernandez-Cortes' dissertation.
-% TODO: concluding remarks about what's missing from traffic work in general.
-%   + there's no central goal or theme---because realism is ill-defined and subjective
-%   + there are no rigorous tests for whether realism is achieved since it's ill-defined
-%   + in wireless, where the impact is greater, there's been virtually no research of impact
-% TODO: cite a bunch of papers that use the CBR traffic model.
-
-\section{Methodology}\label{sec:methodology}
-
-The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results.
-Clearly, we need not simulate subatomic particle interactions in a wireless network simulator.
-Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same.
-Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior.
-This requirement leads us to the following definition:
+\section{Experimental Methodology}\label{sec:methodology}
+
+To evaluate whether traffic models are realistic or not, we use the differential simulation methodology introduced in~\cite{Karpinski07:realism} and used in~\cite{Karpinski07:cbr-failure} to compare a variety of uniform and trace-based models.
+This approach is based on the following definition, initially proposed in~\cite{Karpinski07:realism} and formalized in \cite{Karpinski07:cbr-failure}:
 
@@
 
-This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required.
+The essential idea behind the differential simulation technique is to compare traffic produced by synthetic models with known examples of actual real-world trace traffic in a variety of experimental scenarios.
+To accomplish this, we begin with a large 24-hour example of wireless trace traffic, break it into 144 ten-minute scenarios, and use each scenario's behavior as a control for a set synthetic traces, generated by various synthetic traffic models whose realism we with to evaluate.
+For each scenario, the synthetic models are configured to approximate the original trace behavior as closely as the models allow.
+Important network-wide performance metrics are then collected from simulations, and synthetic performance results compared with control results to determine the realism of the synthetic models.
+
+%The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results.
+Clearly, we need not simulate subatomic particle interactions in a wireless network simulator.
+Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same.
+Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior.
+This requirement leads us to the following definition:
+
+%This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required.
 We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.
+
+\subsection{Traffic Models}
+\label{sec:traffic-models}
+
+The original trace traffic constitutes the first model we simulate, although it is not really a model, {per se}.
+This serves as a control group against which to compare all other models.
+Before we can evaluate the General Matrix Model (\caps{GMM}) presented in \Section{general-matrix-model}, we must establish that one potentially important aspect of traffic behavior that is \textit{not} represented in the \caps{GMM} can legitimately be discarded.
+That aspect of behavior is \textit{\caps{TCP} feedback}.
+The \caps{GMM} does not specify the \caps{IP} protocol type of each flow.
+Therefore we must establish the extent to which using \caps{UDP} flows, without reliable delivery or congestion control, affects performance metrics.
+%\caps{GMM} also does not indicate which flows are paired with each other in a client-server relationship. While this pairing does not affect non-\caps{TCP} flows, we implement.
 
@@
 
-Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of synthetic traffic models defined using the Traffic Object Model (\caps{TOM}) framework.
+Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of trace-based models and synthetic traffic models defined using the General Matrix Model.
 We use a 24-hour trace recorded in an infrastructured 802.11g wireless \caps{LAN} with 18 access points, deployed at the 60th Internet Engineering Task Force meeting (\caps{IETF}60), held in San Diego during August of 2004.
@@
 % 2,085,563 flows, 58,377,994 packets of application data, and 51,988,245,537 bytes of application data.
+
+\begin{figure}
+\vspace{0.45em}
+\begin{center}
+\includegraphics[width=3.5in]{nodes-flows}%
+\vspace{-0.75em}%
+\caption{The number of active nodes and flows over time.} 
+\label{fig:nodes-flows}
+\end{center}
+\vspace{-2em}
+\end{figure}
 
@@
 
-\begin{figure}
-\vspace{0.45em}
-\begin{center}
-\includegraphics[width=3.5in]{nodes-flows}%
-\vspace{-0.75em}%
-\caption{The number of active nodes and flows over time.} 
-\label{fig:nodes-flows}
-\end{center}
-\vspace{-2em}
-\end{figure}
-
 Before using the traces, it is necessary to extract application-level behavior from the trace header data.
@@
 
-\begin{table}[t]
-\vspace{0.75em}
-\begin{center}
-\begin{tabular}[t]{|l|}
-\hline
-\begin{minipage}[l]{2.7in}
-\begin{small}
-\vspace{0.7em}
-\input{properties.tex}
-%\vspace{0.6em}
-\end{small}
-\end{minipage}
-\\\hline
-\end{tabular}
-\end{center}
-\vspace{-0.5em}
-\caption{{\footnotesize{TOM}} base classes and their attributes.}
-\label{tab:properties}
-\vspace{-2em}
-\end{table}
-
-\subsection{The Traffic Object Model~(\caps{TOM})}\label{sec:tom}
-
-\begin{table*}[t]
-\begin{center}
-\small
-\input{models.tex}
-\footnotesize\\[0.5em]
-%\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
-\end{center}
-\vspace{-0.75em}
-\caption{Details of how the sub-models at each behavioral level of the {\footnotesize{TOM}} inherit, override and delegate properties.}
-\label{tab:models}
-\vspace{-1em}
-\end{table*}
-
-In this section, we present a generic object-oriented framework for implementing traffic generation models.
+%\begin{table}[t]
+%\vspace{0.75em}
+%\begin{center}
+%\begin{tabular}[t]{|l|}
+%\hline
+%\begin{minipage}[l]{2.7in}
+%\begin{small}
+%\vspace{0.7em}
+%\input{properties.tex}
+%%\vspace{0.6em}
+%\end{small}
+%\end{minipage}
+%\\\hline
+%\end{tabular}
+%\end{center}
+%\vspace{-0.5em}
+%\caption{{\footnotesize{TOM}} base classes and their attributes.}
+%\label{tab:properties}
+%\vspace{-2em}
+%\end{table}
+
+%\subsection{The Traffic Object Model~(\caps{TOM})}\label{sec:tom}
+
+%\begin{table*}[t]
+%\begin{center}
+%\small
+%\input{models.tex}
+%\footnotesize\\[0.5em]
+%%\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
+%\end{center}
+%\vspace{-0.75em}
+%\caption{Details of how the sub-models at each behavioral level of the {\footnotesize{TOM}} inherit, override and delegate properties.}
+%\label{tab:models}
+%\vspace{-1em}
+%\end{table*}
+
+%In this section, we present a generic object-oriented framework for implementing traffic generation models.
 This framework is called the Traffic Object Model (\caps{TOM}).
@@
 
-% TODO: fix-up client-server vs. source-destination terminology.
-
-Abstract base classes define the properties at each level that traffic models must implement.\footnote{In Java or C\#, this role may be served either by an actual abstract class or by an interface.
+%% TODO: fix-up client-server vs. source-destination terminology.
+
+%Abstract base classes define the properties at each level that traffic models must implement.\footnote{In Java or C\#, this role may be served either by an actual abstract class or by an interface.
 In duck-typed languages such as Ruby or Python, no actual abstract class need exist, since type is implicitly defined by usage; if abstract classes are desired, they can be simulated by creating base classes with stub methods that throw informative exceptions upon being called.}
-Table~\ref{tab:properties} enumerates all the properties exposed by traffic models used in this work.
+%Table~\ref{tab:properties} enumerates all the properties exposed by traffic models used in this work.
 Future traffic models may necessitate exposing additional properties, but this list suffices to implement the models used in this paper.
@@
 A complete synthetic workload is instantiated as follows:
-\begin{enumerate}
-\item create a \class{Network} object of the appropriate subclass;
-\item the top-level \class{Network} object in turn creates a collection of \class{Node} objects and associated \class{Flow} objects;
-\item each \class{Flow} object generates a stream of \class{Packet} objects.
-\end{enumerate}
-The hierarchy of objects may be used to directly drive a simulation, or may generate configuration files that in turn drive a simulation.
+%\begin{enumerate}
+%\item create a \class{Network} object of the appropriate subclass;
+%\item the top-level \class{Network} object in turn creates a collection of \class{Node} objects and associated \class{Flow} objects;
+%\item each \class{Flow} object generates a stream of \class{Packet} objects.
+%\end{enumerate}
+%The hierarchy of objects may be used to directly drive a simulation, or may generate configuration files that in turn drive a simulation.
 Our implementation does the latter, generating a collection of Qualnet configuration files and trace files specific to each workload instantiation (see \Section{simulations}).
 
-\subsection{Traffic Models}\label{sec:models}
-
-For each of the \class{Node}, \class{Flow} and \class{Packet} abstract superclasses, we define four concrete subclasses: \class{Trace}, \class{Nonparametric}, \class{Parametric}, and \class{Uniform}.
+%\subsection{Traffic Models}\label{sec:models}
+
+%For each of the \class{Node}, \class{Flow} and \class{Packet} abstract superclasses, we define four concrete subclasses: \class{Trace}, \class{Nonparametric}, \class{Parametric}, and \class{Uniform}.
 The \class{Trace} classes take properties of objects directly from the behavior of the corresponding object in the original trace.
@@
 
-While the modularity of the Traffic Object Model allows us to mix and match sub-models, not all possible combinations are valid.
+%While the modularity of the Traffic Object Model allows us to mix and match sub-models, not all possible combinations are valid.
 In particular, note that the \class{Trace} classes are tied to a specific object from the underlying trace, whereas for other classes, there is no such correspondence to any specific object.
@@
 The traffic models range from a completely realistic trace-driven model, to a standard \caps{CBR} traffic model.
-Various partially synthetic intermediate models, described in \Section{models}, are simulated to study the impact of different aspects of traffic behavior on network performance.
+Various partially synthetic intermediate models, described in \Section{traffic-models}, are simulated to study the impact of different aspects of traffic behavior on network performance.
 To preserve the fairness of the performance comparison, we keep as many features as possible constant across different traffic models.
@@
 
-\begin{sidewaysfigure*}
-%\begin{centering}
-\noindent
-\subfloat[Application: Average Jitter]
-{\includegraphics[width=4.6in]{Application-Average-Jitter}
-\label{fig:Application-Average-Jitter}}
-\subfloat[Application: Average Delay]
-{\includegraphics[width=4.6in]{Application-Average-Delay}
-\label{fig:Application-Average-Delay}}
-\subfloat[Application: Received Throughput]
-{\includegraphics[width=4.6in]{Application-Received-Throughput}
- \label{fig:Application-Received-Throughput}}
-\subfloat[Network: Control Overhead]
-{\includegraphics[width=4.6in]{Network-Control-Overhead}
- \label{fig:Network-Control-Overhead}}
-\subfloat[Network: Packet Drop Rate]
-{\includegraphics[width=4.6in]{Network-Packet-Drop-Rate}
- \label{fig:Network-Packet-Drop-Rate}}
-\subfloat[Network: \caps{RREQ}s Initiated]
-{\includegraphics[width=4.6in]{Network-RREQs-Initiated}
- \label{fig:Network-RREQs-Initiated}}
-\subfloat[Link (MAC): Control Overhead]
-{\includegraphics[width=4.6in]{MAC-Control-Overhead}
- \label{fig:MAC-Control-Overhead}}
-\subfloat[Link (MAC): Retransmissions]
-{\includegraphics[width=4.6in]{MAC-Retransmissions}
- \label{fig:MAC-Retransmissions}}
-\subfloat[Link (\caps{MAC}): Retransmission Failures]
-{\includegraphics[width=4.6in]{MAC-Retransmission-Failures}
- \label{fig:MAC-Retransmission-Failures}}
-\hspace{0.035in}
-\begin{minipage}[t]{4.545in}
-\vspace{-2em}
-\caption{Box-and-whisker plots of log-ratio error values for all metrics and traffic models.
+%\begin{sidewaysfigure*}
+%%\begin{centering}
+%\noindent
+%\subfloat[Application: Average Jitter]
+%{\includegraphics[width=4.6in]{Application-Average-Jitter}
+%\label{fig:Application-Average-Jitter}}
+%\subfloat[Application: Average Delay]
+%{\includegraphics[width=4.6in]{Application-Average-Delay}
+%\label{fig:Application-Average-Delay}}
+%\subfloat[Application: Received Throughput]
+%{\includegraphics[width=4.6in]{Application-Received-Throughput}
+% \label{fig:Application-Received-Throughput}}
+%\subfloat[Network: Control Overhead]
+%{\includegraphics[width=4.6in]{Network-Control-Overhead}
+% \label{fig:Network-Control-Overhead}}
+%\subfloat[Network: Packet Drop Rate]
+%{\includegraphics[width=4.6in]{Network-Packet-Drop-Rate}
+% \label{fig:Network-Packet-Drop-Rate}}
+%\subfloat[Network: \caps{RREQ}s Initiated]
+%{\includegraphics[width=4.6in]{Network-RREQs-Initiated}
+% \label{fig:Network-RREQs-Initiated}}
+%\subfloat[Link (MAC): Control Overhead]
+%{\includegraphics[width=4.6in]{MAC-Control-Overhead}
+% \label{fig:MAC-Control-Overhead}}
+%\subfloat[Link (MAC): Retransmissions]
+%{\includegraphics[width=4.6in]{MAC-Retransmissions}
+% \label{fig:MAC-Retransmissions}}
+%\subfloat[Link (\caps{MAC}): Retransmission Failures]
+%{\includegraphics[width=4.6in]{MAC-Retransmission-Failures}
+% \label{fig:MAC-Retransmission-Failures}}
+%\hspace{0.035in}
+%\begin{minipage}[t]{4.545in}
+%\vspace{-2em}
+%\caption{Box-and-whisker plots of log-ratio error values for all metrics and traffic models.
 The lower axis indicates the log-ratio, while the upper axis shows raw ratio values.
@@
 Numbers to the right are $p$-values for the median, mean, and 3-mean tests, respectively; $p$-values above $\alpha=0.05$ appear in bold.}
-\label{fig:boxplots}
-\end{minipage}
-%\end{centering}
-\end{sidewaysfigure*}
+%\label{fig:boxplots}
+%\end{minipage}
+%%\end{centering}
+%\end{sidewaysfigure*}
 
@@
 
-\begin{figure*}
-\begin{centering}
-\subfloat[App. Average Jitter: {\footnotesize{STU}}]
-{\includegraphics[width=1.75in]{application-average-jitter-stu}%
-\label{fig:application-average-jitter-stu}}
-\subfloat[Net. Packet Drop Rate: {\footnotesize{UTU}}]
-{\includegraphics[width=1.75in]{network-packet-drop-rate-utu}%
-\label{fig:network-packet-drop-rate-utu}}
-\subfloat[Net. Packet Drop Rate: {\footnotesize{STU}}]
-{\includegraphics[width=1.75in]{network-packet-drop-rate-stu}%
-\label{fig:network-packet-drop-rate-stu}}
-\subfloat[Link Retrans. Failures: {\footnotesize{SUU}}]
-{\includegraphics[width=1.75in]{mac-retransmission-failures-suu}%
-\label{fig:mac-retransmission-failures-suu}}
-\caption{Scatter plots of various metrics for an alternate traffic model ({\footnotesize{STU}}, {\footnotesize{UTU}}, {\footnotesize{SUU}}) versus real trace traffic~({\footnotesize{TTT}}).
+%\begin{figure*}
+%\begin{centering}
+%\subfloat[App. Average Jitter: {\footnotesize{STU}}]
+%{\includegraphics[width=1.75in]{application-average-jitter-stu}%
+%\label{fig:application-average-jitter-stu}}
+%\subfloat[Net. Packet Drop Rate: {\footnotesize{UTU}}]
+%{\includegraphics[width=1.75in]{network-packet-drop-rate-utu}%
+%\label{fig:network-packet-drop-rate-utu}}
+%\subfloat[Net. Packet Drop Rate: {\footnotesize{STU}}]
+%{\includegraphics[width=1.75in]{network-packet-drop-rate-stu}%
+%\label{fig:network-packet-drop-rate-stu}}
+%\subfloat[Link Retrans. Failures: {\footnotesize{SUU}}]
+%{\includegraphics[width=1.75in]{mac-retransmission-failures-suu}%
+%\label{fig:mac-retransmission-failures-suu}}
+%\caption{Scatter plots of various metrics for an alternate traffic model ({\footnotesize{STU}}, {\footnotesize{UTU}}, {\footnotesize{SUU}}) versus real trace traffic~({\footnotesize{TTT}}).
 Each data point represents a pair of matched simulations: the $x$-coordinate is the metric value for trace traffic, the $y$-coordinate is the metric value for the synthetic model.
 Scenarios are plotted according to the amount of flows in their trace segment: the bottom third with filled dots; the middle third with circles; the top third with squares.}
-\label{fig:scatterplots}
-\end{centering}
-\vspace{-1em}
-\end{figure*}
+%\label{fig:scatterplots}
+%\end{centering}
+%\vspace{-1em}
+%\end{figure*}
 
@@
 \section{Results}\label{sec:results}
+
+\begin{figure*}[htb]
+\subfloat[\textbf{Application:} Average End-to-End Delay]%
+{\includegraphics[width=7in]{plots/Application/delay.pdf}}
+\subfloat[\textbf{Application:} Average Received Throughput]%
+{\includegraphics[width=7in]{plots/Application/throughput.pdf}}
+\subfloat[\textbf{Application:} Total Received Data]%
+{\includegraphics[width=7in]{plots/Application/total_recv_packets.pdf}}
+\subfloat[\textbf{Network:} AODV Control Overhead]%
+{\includegraphics[width=7in]{plots/Network/control_packets.pdf}}
+\subfloat[\textbf{Link:} 802.11 Control Overhead]%
+{\includegraphics[width=7in]{plots/Link/overhead.pdf}}%
+\caption{Box-and-whisker plots of log-ratio error values for all metrics and traffic models.
+The lower axis indicates the log-ratio, while the upper axis shows raw ratio values.
+Each box contains the central majority of log-ratio values: the left and right bounds are at the $25^\text{th}$ and $75^\text{th}$ percentiles.
+The dark middle line indicates the median value, while the diamond marks the mean.
+The whiskers (dotted lines) extend to the furthest non-outlier values, while the points beyond that are outliers.
+The notch in the middle of each bar indicates a $95\%$ confidence interval for the true underlying median value; if two notches do not overlap, they are very unlikely to have the same median.}
+\label{fig:box-plots}
+\vspace{-1.25em}
+\end{figure*}
 
@@
 
-\begin{figure}
-\vspace{0.6em}
-\begin{centering}
-\includegraphics[width=3.45in]{mac-control-overhead-cmp}
-\caption{Comparison of {\footnotesize{MAC}} control overhead for {\footnotesize{AODV}} and {\footnotesize{OLSR}} using both the {\footnotesize{UUU}} and {\footnotesize{TTT}} traffic models.
+%\begin{figure}
+%\vspace{0.6em}
+%\begin{centering}
+%\includegraphics[width=3.45in]{mac-control-overhead-cmp}
+%\caption{Comparison of {\footnotesize{MAC}} control overhead for {\footnotesize{AODV}} and {\footnotesize{OLSR}} using both the {\footnotesize{UUU}} and {\footnotesize{TTT}} traffic models.
 The fit lines show locally weighted, smoothed performance trends.}
-\label{fig:mac-control-overhead-cmp}
-\end{centering}
-\vspace{-1.5em}
-\end{figure}
+%\label{fig:mac-control-overhead-cmp}
+%\end{centering}
+%\vspace{-1.5em}
+%\end{figure}
 

### 2008-02-05 paper.tex
Paper!

@@
 \newcommand{\Equation}[1]{\hyperref[eqn:#1]{Equation~\ref*{eqn:#1}}}
+\newcommand{\Figure}[1]{\hyperref[fig:#1]{Figure~\ref*{fig:#1}}}
+\newcommand{\Table}[1]{\hyperref[tab:#1]{Table~\ref*{tab:#1}}}
 
@@
 
+\begin{table*}
+\begin{center}
+\small
+\begin{tabular}{|c|c|l|}
+\multicolumn{1}{c}{\textbf{model}} &
+\multicolumn{1}{c}{\textbf{transformation}} &
+\multicolumn{1}{c}{\textbf{description}} \\
+\hline
+{\footnotesize{GMM}} & --- & the general matrix model \\\hline\hline
+flow uniform & $F'=(\ones{f \times f}/f)F$ & flows are generated stochastically identically \\\hline
+node uniform & $S'=S(\ones{n \times n}/n)$~~;~~$D'=D(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
+time uniform & $B'=B(\ones{d_t \times d_t}/d_t)$~~;~~$E'=E(\ones{d_t \times d_t}/d_t)$ & all start times and end times are equally likely \\\hline
+full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above three models \\\hline\hline
+node marginal & $S'=(\ones{f \times f}/f)S$~~;~~$D'=(\ones{f \times f}/f)D$ & nodes behave distinctly, but behavior is independent of flows \\\hline
+packet marginal & $Z'=(\ones{f \times f}/f)Z$~~;~~$V'=(\ones{f \times f}/f)V$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
+\end{tabular}
+\caption{Matrix-based traffic models evaluated.}
+\label{tab:traffic-models}
+\end{center}
+\end{table*}
+
 The original trace traffic constitutes the first model we simulate, although it is not really a model, {per se}.
@@
 %\caps{GMM} also does not indicate which flows are paired with each other in a client-server relationship. While this pairing does not affect non-\caps{TCP} flows, we implement.
+Other synthetic models derived from algebraic transformations of the \caps{GMM} are listed in \Table{traffic-models}.
 
@@
 
-Our simulation results are summarized in Figure~\ref{fig:boxplots}.
+Our simulation results are summarized in Figure~\ref{fig:box-plots}.
 Each subfigure shows a single performance metric.
@@
 Additionally, the mean and median markers should be close to the center.
-Complementing the visual display of summary statistics, Figure~\ref{fig:boxplots} also lists three $p$-values to the right of each box-and-whisker plot.
-These are, in order, the $p$-values for the median, mean and 3-mean tests described in \Section{statistical-equivalence}.
-Each test catches a different type of unrealistic statistical behavior.
-
-The \caps{UTU} model, for example, which uses real flow behavior but synthetic flow topology and packet behavior, does a very good job of accurately reproducing realistic average end-to-end delay (Figure~\ref{fig:Application-Average-Delay}): the median and mean are both close to zero, and the $p$-values are all greater than $0.05$.
-The standard uniform \caps{CBR} model (\caps{UUU}), on the other hand, underestimates average end-to-end delay by between a factor of $2$ and $11$ half of the time, and by more than a factor of $11$ a quarter of the time.
-Its $p$-values for delay are all less than $0.005$.
-
-Figure~\ref{fig:Application-Average-Jitter} shows that the \caps{STU} traffic model passes the mean test but not the median or 3-mean tests.
-This result indicates that this model tends to overestimate jitter, but in some subset of cases it significantly underestimates instead, yielding a mean near zero, but a skewed median.
-This analysis is verified when we examine a scatterplot of metric values for \caps{STU} versus \caps{TTT} traffic models.
-In Figure~\ref{fig:application-average-jitter-stu} we can see that for the majority of simulations, the \caps{STU} value for jitter exceeds the \caps{TTT} value.
-In a significant minority, however, the \caps{STU} model yields very small jitter values, while the \caps{TTT} model gives large values.
-Even though the average error is near zero, this is not a model that reproduces realistic jitter.
-
-The \caps{UTU} traffic model accurately reproduces packet dropping behavior in routing queues.
-Figure~\ref{fig:network-packet-drop-rate-utu} shows a scatterplot for such a realistic model: the data points are well clustered symmetrically around the diagonal line.
-We conclude that the primary influence on the packet drop rate is flow behavior.
-However, if we use a sink topology model, as in Figure~\ref{fig:network-packet-drop-rate-stu}, the drop rate becomes inflated.
-The sink topology introduces an unrealistic routing bottleneck in the network, causing excessive queue overflows for all sizes of scenarios.
-The uniform topology model does not exhibit this bottleneck, and thereby avoids producing this unrealistic performance artifact.
-This demonstrates that while the uniform topology model is generally less realistic than the sink model, for certain metrics, their relative quality is the opposite.
-
-Figure~\ref{fig:mac-retransmission-failures-suu} illustrates a case where the 3-mean test catches unrealistic behavior that the median and mean tests do not catch.
-From Figure~\ref{fig:MAC-Retransmission-Failures} we can see that the mean and median error values for the \caps{SUU} model are fairly small.
-The scatter-plot, however, shows that this model significantly underestimates the failure rate for large scenarios with many flows (squares), while overestimating the rate in smaller scenarios.
-
-There are few, if any, positive conclusions that can be drawn from these results.
-The primary message is that these synthetic traffic models, especially the standard uniform \caps{CBR} model, consistently misrepresent the most important performance metrics.
-The traffic model that performs the best overall is the \caps{UTU} model, which uses real flow behavior with uniform flow topology and uniform packet behavior.
-This model, however, still fails statistical tests of realism for all but two of the metrics considered.
-Further development of traffic models is needed before it becomes possible to generate traffic in simulations or experimental deployments such that a single set of experiments can realistically evaluate all aspects of network performance.
+
+The primary result is that the General Matrix Model~(\caps{GMM}) is sufficiently realistic.
+It accurately reproduces important performance metrics at all levels of the protocol stack in simulation.
+This only moves us a step closer to being able to generate realistic, completely synthetic workloads, since the \caps{GMM} in these comparisons  derives almost all of its behavior directly from the trace.
+However, it means that we have effectively reduced the problem of generating realistic workload to the problem of classifying and understanding what sorts of traffic matrices occur in real workloads.
+Moreover, we can express a vast array of different approaches to traffic modeling simply and uniformly using matrix operations.
 
@@
 While using a real, physical network successfully sidesteps simulation problems below the application layer, without realistic traffic models, reliable, meaningful performance predictions remain beyond our reach.
-
-\subsection{Relative Performance Comparisons}
-
-Performance evaluations are primarily used to compare new protocols with existing ones.
-It remains possible that while misrepresenting important metrics, synthetic traffic models preserve the relative performance of protocols.
-In order to test this hypothesis, we have run further simulations, revisiting a classic comparison of ad~hoc routing protocols: \caps{AODV} \textit{vs.} \caps{OLSR}.
-% TODO: cite a previous comparison here.
-We focus on the trace segments with 75 or fewer active nodes, since published performance comparisons of ad hoc routing protocols have typically not used more nodes than this.
-For each scenario, we have run simulations with 10 different seed values using the \caps{AODV} and \caps{OLSR} routing protocols and the \caps{UUU} and \caps{TTT} traffic models, allowing us to compare relative performance results when switching traffic models.
-
-The key result of our comparison is that, for certain performance metrics, \textbf{the relative performance of \caps{AODV} and \caps{OLSR} is inverted, simply  by changing traffic models.} Figure~\ref{fig:mac-control-overhead-cmp} shows some of the results from this experiment.
-The graph plots \caps{MAC} control overhead against the number of nodes in each simulation scenario.
-Each data point shows the average overhead for the simulations of that scenario using ten seed values.
-With the standard \caps{UUU} traffic model, \caps{OLSR} consistently has much higher overhead, across the board, than \caps{AODV}.
-When real traffic is used, however, \caps{OLSR} performs significantly better than \caps{AODV}.
-A similar switch occurs for the number of \caps{MAC} retransmissions.
-This experiment demonstrates that the use of unrealistic traffic models can change the relative performance of different protocols, as well as skew absolute performance measurements.
-
-%\begin{figure}
-%\vspace{0.6em}
-%\begin{centering}
-%\includegraphics[width=3.45in]{mac-control-overhead-cmp}
-%\caption{Comparison of {\footnotesize{MAC}} control overhead for {\footnotesize{AODV}} and {\footnotesize{OLSR}} using both the {\footnotesize{UUU}} and {\footnotesize{TTT}} traffic models.
-The fit lines show locally weighted, smoothed performance trends.}
-%\label{fig:mac-control-overhead-cmp}
-%\end{centering}
-%\vspace{-1.5em}
-%\end{figure}
 
@@
 Traffic traces, on the other hand, must be used without significant alteration if they are to actually provide the desired realism.
-The ``messiness'' of the performance comparison from trace data in Figure~\ref{fig:mac-control-overhead-cmp} illustrates why using traces directly is not ideal: each data point differs not only in the number of nodes shown on the $x$-axis, but also in other dimensions, such as the number of flows and packets, and the average flow duration.
+%The ``messiness'' of the performance comparison from trace data in Figure~\ref{fig:mac-control-overhead-cmp} illustrates why using traces directly is not ideal: each data point differs not only in the number of nodes shown on the $x$-axis, but also in other dimensions, such as the number of flows and packets, and the average flow duration.
 The result is a highly noisy comparison, affected by many unseen parameters.
@@
 
-The next step towards better traffic models is to investigate which aspects of real traces may be altered without detrimentally affecting the resulting performance metrics.
+%The next step towards better traffic models is to investigate which aspects of real traces may be altered without detrimentally affecting the resulting performance metrics.
 For example, to test whether a complex time-series model of packet behavior is necessary, we randomize the order of the packet sizes and/or inter-packet intervals and compare performance using these randomized traces against performance using the original traces.

### 2008-02-07 paper.tex
Added plots!

@@
 
-\begin{figure*}[htb]
+\begin{sidewaysfigure*}
+%\begin{figure*}[htb]
 \subfloat[\textbf{Application:} Average End-to-End Delay]%
-{\includegraphics[width=7in]{plots/Application/delay.pdf}}
+{\includegraphics[width=4.6in]{plots/Application/delay.pdf}}
 \subfloat[\textbf{Application:} Average Received Throughput]%
-{\includegraphics[width=7in]{plots/Application/throughput.pdf}}
-\subfloat[\textbf{Application:} Total Received Data]%
-{\includegraphics[width=7in]{plots/Application/total_recv_packets.pdf}}
+{\includegraphics[width=4.6in]{plots/Application/throughput.pdf}}
+\subfloat[\textbf{Application:} Total Received Packets]%
+{\includegraphics[width=4.6in]{plots/Application/total_recv_packets.pdf}}
 \subfloat[\textbf{Network:} AODV Control Overhead]%
-{\includegraphics[width=7in]{plots/Network/control_packets.pdf}}
+{\includegraphics[width=4.6in]{plots/Network/control_packets.pdf}}
+\subfloat[\textbf{Network:} Packets Dropped in Routing Queue]%
+{\includegraphics[width=4.6in]{plots/Network/dropped.pdf}}
 \subfloat[\textbf{Link:} 802.11 Control Overhead]%
-{\includegraphics[width=7in]{plots/Link/overhead.pdf}}%
+{\includegraphics[width=4.6in]{plots/Link/overhead.pdf}}%
 \caption{Box-and-whisker plots of log-ratio error values for all metrics and traffic models.
@@
 \label{fig:box-plots}
-\vspace{-1.25em}
-\end{figure*}
+%\vspace{-1.25em}
+%\end{figure*}
+\end{sidewaysfigure*}
 

### 2008-02-07 paper.tex
Editing: abstract, intro.

@@
 
-\title{Linear Algebra as a Unifying Theory\\of Wireless Workload Modeling}
+\title{An Algebraic Representation of Wireless Workload}
 \author{
@@
 \begin{abstract}
-We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in wireless networks.
+We present a representation of wireless workload patterns as large, sparse matrices, together with a method for stochastically generating experimental workload from any given matrix.
+Each row vector represents a single flow of packets in the network, with components representing the \caps{IP} protocol type, source and destination nodes, flow size, start time, and packet behavior.
+The essential property of the algebraic representation is that the summation of rows naturally yields a faithful description of the aggregate behavior of the corresponding flows.
+This deceptively simple property allows us to express many common types of traffic models succinctly in terms of a few matrix multiplications and linear constraints.
+This approach has many benefits: 1) it makes the meaning of certain generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2) it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3) the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed.
+We use the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models, to experimentally validate the general matrix model.
+%We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models.
+A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
+%
+%We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in wireless networks.
 Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematical precision and succinctness, while simultaneously being made clearer and more general.
@@
 
-More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and billions of computers on the Internet and in local area networks~(\caps{LAN}s).
+More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local area networks~(\caps{LAN}s).
 The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
@@
 
-Despite great breakthroughs in understanding and methodology, the networking community remains without a cookbook for generating realistic \caps{LAN} workloads.
-Drastic assumptions of uniformity abound in typical experimental workload generation.
+Despite great breakthroughs in understanding and methodology, the networking community remains without a ``cookbook'' for generating realistic \caps{LAN} workloads.
+By ``cookbook,'' we mean an explicit method for generating network traffic---down to the size and send time of each application data unit.
+In the current standard practice for experimental workload generation, drastic assumptions of uniformity abound.
 Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily.
 While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
-the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better.
+the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that any one model is better than the rest.
 It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.
 
-We believe that two major breakthroughs are needed to bring a much needed revolution to the world of traffic modeling.
+We believe that two major breakthroughs are needed to improve traffic modeling.
 The first requirement is an objective standard of realism by which to compare traffic models.
@@
 The second required breakthrough is a unified framework for expressing traffic models that allows researchers to systematically express and understand the assumptions and transformations entailed in each model.
+The major contribution of this paper is in addressing this second requirement.
+
+%Our previous work on traffic modeling provided the means
 
 In this paper, we present a highly general model for representing detailed full-network traffic patterns algebraically.
-The General Matrix Model (\caps{GMM}) represents the behavior or each network flow as a vector in a high-dimensional linear space, and represents total network behavior as a large matrix whose rows are the behavior vectors.
+The General Matrix Model (\caps{GMM}) represents the behavior or each network flow as a vector in a high-dimensional linear space, and represents total network behavior as a large matrix whose rows are these flow-behavior vectors.
 The crucial property of this representation is that it respects vector space operations: summation of behavior vectors gives an accurate description of the aggregate behavior of the flows represented by those vectors.
 The benefit of this representation is that it makes it immediately possible to express a broad variety of assumptions and conditions on traffic patterns in terms of simple matrix operations.
-It also allows provides great mathematical insight and allows us to bring all the tools of modern linear algebra to bear on the problem of traffic modeling.
+It also provides vital mathematical clarity and allows us to apply all the tools of modern linear algebra to the problems of traffic modeling.
 
@@
 The number of dimensions was chosen to match the dimensions for packet size: $d_v=1500$.
-Finally, the parameter $\gamma\in(0,1)$ allows us to fine-tune how the resolution scales with interval size; we have chosen $\gamma=\fracx{1}{3}$ because this choice provides resolution that scales smoothly from the microseconds for small intervals to seconds for large intervals.
+Finally, the parameter $\gamma\in(0,1)$ allows us to fine-tune how the resolution scales with interval size; we have chosen $\gamma=\frac{1}{3}$ because this choice provides resolution that scales smoothly from the microseconds for small intervals to seconds for large intervals.
 
@@
 
-Given an instance of the general matrix model, how do we produce an actual workload form it?
+Given an instance of the general matrix model, how do we produce an actual workload from it?
 Each row of the model is used to randomly generate the behavior of a single flow.
@@
 
-\subsection{Tests of Statistical Equivalence}\label{sec:statistical-equivalence}
-
-%\begin{sidewaysfigure*}
-%%\begin{centering}
-%\noindent
-%\subfloat[Application: Average Jitter]
-%{\includegraphics[width=4.6in]{Application-Average-Jitter}
-%\label{fig:Application-Average-Jitter}}
-%\subfloat[Application: Average Delay]
-%{\includegraphics[width=4.6in]{Application-Average-Delay}
-%\label{fig:Application-Average-Delay}}
-%\subfloat[Application: Received Throughput]
-%{\includegraphics[width=4.6in]{Application-Received-Throughput}
-% \label{fig:Application-Received-Throughput}}
-%\subfloat[Network: Control Overhead]
-%{\includegraphics[width=4.6in]{Network-Control-Overhead}
-% \label{fig:Network-Control-Overhead}}
-%\subfloat[Network: Packet Drop Rate]
-%{\includegraphics[width=4.6in]{Network-Packet-Drop-Rate}
-% \label{fig:Network-Packet-Drop-Rate}}
-%\subfloat[Network: \caps{RREQ}s Initiated]
-%{\includegraphics[width=4.6in]{Network-RREQs-Initiated}
-% \label{fig:Network-RREQs-Initiated}}
-%\subfloat[Link (MAC): Control Overhead]
-%{\includegraphics[width=4.6in]{MAC-Control-Overhead}
-% \label{fig:MAC-Control-Overhead}}
-%\subfloat[Link (MAC): Retransmissions]
-%{\includegraphics[width=4.6in]{MAC-Retransmissions}
-% \label{fig:MAC-Retransmissions}}
-%\subfloat[Link (\caps{MAC}): Retransmission Failures]
-%{\includegraphics[width=4.6in]{MAC-Retransmission-Failures}
-% \label{fig:MAC-Retransmission-Failures}}
-%\hspace{0.035in}
-%\begin{minipage}[t]{4.545in}
-%\vspace{-2em}
-%\caption{Box-and-whisker plots of log-ratio error values for all metrics and traffic models.
+%\subsection{Tests of Statistical Equivalence}\label{sec:statistical-equivalence}
+
+%%\begin{sidewaysfigure*}
+%%%\begin{centering}
+%%\noindent
+%%\subfloat[Application: Average Jitter]
+%%{\includegraphics[width=4.6in]{Application-Average-Jitter}
+%%\label{fig:Application-Average-Jitter}}
+%%\subfloat[Application: Average Delay]
+%%{\includegraphics[width=4.6in]{Application-Average-Delay}
+%%\label{fig:Application-Average-Delay}}
+%%\subfloat[Application: Received Throughput]
+%%{\includegraphics[width=4.6in]{Application-Received-Throughput}
+%% \label{fig:Application-Received-Throughput}}
+%%\subfloat[Network: Control Overhead]
+%%{\includegraphics[width=4.6in]{Network-Control-Overhead}
+%% \label{fig:Network-Control-Overhead}}
+%%\subfloat[Network: Packet Drop Rate]
+%%{\includegraphics[width=4.6in]{Network-Packet-Drop-Rate}
+%% \label{fig:Network-Packet-Drop-Rate}}
+%%\subfloat[Network: \caps{RREQ}s Initiated]
+%%{\includegraphics[width=4.6in]{Network-RREQs-Initiated}
+%% \label{fig:Network-RREQs-Initiated}}
+%%\subfloat[Link (MAC): Control Overhead]
+%%{\includegraphics[width=4.6in]{MAC-Control-Overhead}
+%% \label{fig:MAC-Control-Overhead}}
+%%\subfloat[Link (MAC): Retransmissions]
+%%{\includegraphics[width=4.6in]{MAC-Retransmissions}
+%% \label{fig:MAC-Retransmissions}}
+%%\subfloat[Link (\caps{MAC}): Retransmission Failures]
+%%{\includegraphics[width=4.6in]{MAC-Retransmission-Failures}
+%% \label{fig:MAC-Retransmission-Failures}}
+%%\hspace{0.035in}
+%%\begin{minipage}[t]{4.545in}
+%%\vspace{-2em}
+%%\caption{Box-and-whisker plots of log-ratio error values for all metrics and traffic models.
 The lower axis indicates the log-ratio, while the upper axis shows raw ratio values.
@@
 Numbers to the right are $p$-values for the median, mean, and 3-mean tests, respectively; $p$-values above $\alpha=0.05$ appear in bold.}
-%\label{fig:boxplots}
-%\end{minipage}
-%%\end{centering}
-%\end{sidewaysfigure*}
-
-In this section, we consider the values of performance metrics as random variables, drawn from unknown distributions.
+%%\label{fig:boxplots}
+%%\end{minipage}
+%%%\end{centering}
+%%\end{sidewaysfigure*}
+
+%In this section, we consider the values of performance metrics as random variables, drawn from unknown distributions.
 We present three tests for the statistical equivalence of the metric values induced by synthetic and real traffic.
@@
 
-\textbf{The Median Test.} If $\M$ induces realistic performance, then the median of each log-ratio variable, $R_k^\M=\log(X_k^\M/X_k^\TTT)$, should be zero.
+%\textbf{The Median Test.} If $\M$ induces realistic performance, then the median of each log-ratio variable, $R_k^\M=\log(X_k^\M/X_k^\TTT)$, should be zero.
 The $k^\text{th}$ indicator variable is defined as
-\begin{align}
-	I_k^\M = \begin{cases}
-		0 & \text{if $R_k^\M < 0$,} \\
-		1 & \text{if $R_k^\M \ge 0$.}
-	\end{cases}
-\end{align}
-If the median value of $R_k^\M$ is truly zero, then $I_k^\M$ is a Bernoulli variable with probability parameter $p=\frac{1}{2}$.
+%\begin{align}
+%	I_k^\M = \begin{cases}
+%		0 & \text{if $R_k^\M < 0$,} \\
+%		1 & \text{if $R_k^\M \ge 0$.}
+%	\end{cases}
+%\end{align}
+%If the median value of $R_k^\M$ is truly zero, then $I_k^\M$ is a Bernoulli variable with probability parameter $p=\frac{1}{2}$.
 The variables $I_k^\M$ are all independent since they come from separate simulations, and cannot affect each other's outcomes.
@@
 The median test applies the exact binomial cumulative distribution function (\caps{CDF}) % TODO: cite this.
- for $n$ and $p$ to the observed value of $S_n^\M$, yielding a $p$-value: the probability that such an extreme value would occur by chance under the hypothesis that the median of each $R_k^\M$ is zero.
+% for $n$ and $p$ to the observed value of $S_n^\M$, yielding a $p$-value: the probability that such an extreme value would occur by chance under the hypothesis that the median of each $R_k^\M$ is zero.
 % TODO: fix the wording here.
 
-%\begin{figure*}
-%\begin{centering}
-%\subfloat[App. Average Jitter: {\footnotesize{STU}}]
-%{\includegraphics[width=1.75in]{application-average-jitter-stu}%
-%\label{fig:application-average-jitter-stu}}
-%\subfloat[Net. Packet Drop Rate: {\footnotesize{UTU}}]
-%{\includegraphics[width=1.75in]{network-packet-drop-rate-utu}%
-%\label{fig:network-packet-drop-rate-utu}}
-%\subfloat[Net. Packet Drop Rate: {\footnotesize{STU}}]
-%{\includegraphics[width=1.75in]{network-packet-drop-rate-stu}%
-%\label{fig:network-packet-drop-rate-stu}}
-%\subfloat[Link Retrans. Failures: {\footnotesize{SUU}}]
-%{\includegraphics[width=1.75in]{mac-retransmission-failures-suu}%
-%\label{fig:mac-retransmission-failures-suu}}
-%\caption{Scatter plots of various metrics for an alternate traffic model ({\footnotesize{STU}}, {\footnotesize{UTU}}, {\footnotesize{SUU}}) versus real trace traffic~({\footnotesize{TTT}}).
+%%\begin{figure*}
+%%\begin{centering}
+%%\subfloat[App. Average Jitter: {\footnotesize{STU}}]
+%%{\includegraphics[width=1.75in]{application-average-jitter-stu}%
+%%\label{fig:application-average-jitter-stu}}
+%%\subfloat[Net. Packet Drop Rate: {\footnotesize{UTU}}]
+%%{\includegraphics[width=1.75in]{network-packet-drop-rate-utu}%
+%%\label{fig:network-packet-drop-rate-utu}}
+%%\subfloat[Net. Packet Drop Rate: {\footnotesize{STU}}]
+%%{\includegraphics[width=1.75in]{network-packet-drop-rate-stu}%
+%%\label{fig:network-packet-drop-rate-stu}}
+%%\subfloat[Link Retrans. Failures: {\footnotesize{SUU}}]
+%%{\includegraphics[width=1.75in]{mac-retransmission-failures-suu}%
+%%\label{fig:mac-retransmission-failures-suu}}
+%%\caption{Scatter plots of various metrics for an alternate traffic model ({\footnotesize{STU}}, {\footnotesize{UTU}}, {\footnotesize{SUU}}) versus real trace traffic~({\footnotesize{TTT}}).
 Each data point represents a pair of matched simulations: the $x$-coordinate is the metric value for trace traffic, the $y$-coordinate is the metric value for the synthetic model.
 Scenarios are plotted according to the amount of flows in their trace segment: the bottom third with filled dots; the middle third with circles; the top third with squares.}
-%\label{fig:scatterplots}
-%\end{centering}
-%\vspace{-1em}
-%\end{figure*}
-
-\textbf{The Mean Test.} We use Lyapunov's generalization of the Central Limit Theorem (\caps{LCLT}) to test the hypothesis that the mean of each $R_k^\M$ is zero.
+%%\label{fig:scatterplots}
+%%\end{centering}
+%%\vspace{-1em}
+%%\end{figure*}
+
+%\textbf{The Mean Test.} We use Lyapunov's generalization of the Central Limit Theorem (\caps{LCLT}) to test the hypothesis that the mean of each $R_k^\M$ is zero.
 We present the theorem and its application to the series $R_k^\M$ in the Appendix.
@@
 If the mean of each $R_k^\M$ is zero, then \caps{LCLT} implies that the test statistic
-\begin{align}
-\hat{Z}_n^\M &= \frac{\sum_{k=1}^n{R_k^\M}}{\sqrt{\sum_{k=1}^n{(R_k^\M)^2}}}
-\end{align}
-converges to a standard normal distribution for large values of $n$, where $n$ is the number of simulated scenarios.
+%\begin{align}
+%\hat{Z}_n^\M &= \frac{\sum_{k=1}^n{R_k^\M}}{\sqrt{\sum_{k=1}^n{(R_k^\M)^2}}}
+%\end{align}
+%converges to a standard normal distribution for large values of $n$, where $n$ is the number of simulated scenarios.
 In this case, $n=144$, which is fairly large by traditional statistical standards.
@@
 
-\textbf{The 3-Mean Test.} Performance behavior in scenarios with a large number of nodes or flows is often very different from behavior in small scenarios.
+%\textbf{The 3-Mean Test.} Performance behavior in scenarios with a large number of nodes or flows is often very different from behavior in small scenarios.
 In some cases, $R_k^\M$ is skewed positively for one group, but negatively for another.
@@
 
-\section{Acknowledgments}
-This work was funded in part through NSF Career Award CNS-0347886.
+%\section{Acknowledgments}
+%This work was funded in part through NSF Career Award CNS-0347886.
 
@@
 
-\vspace{1em}\noindent
-\textbf{Theorem.}~[\textit{Lyapunov's Central Limit Theorem}] Let $\{R_k\}_{k=1}^{\infty}$ be a series of independent variables with $\E{R_k}=0$.
+%\vspace{1em}\noindent
+%\textbf{Theorem.}~[\textit{Lyapunov's Central Limit Theorem}] Let $\{R_k\}_{k=1}^{\infty}$ be a series of independent variables with $\E{R_k}=0$.
 Let $s_n^2 = \sum_{k=1}^n{\E{\abs{R_k^2}}}$ and $r_n^3 = \sum_{k=1}^n{\E{\abs{R_k^3}}}$.
@@
 
-\vspace{1em}
-To apply the LCLT to the series $R_k^\M=\log(X_k^\M/X_k^\TTT)$, we must show that under the null hypothesis, the assumptions of the theorem are satisfied by this series.
+%\vspace{1em}
+%To apply the LCLT to the series $R_k^\M=\log(X_k^\M/X_k^\TTT)$, we must show that under the null hypothesis, the assumptions of the theorem are satisfied by this series.
 First, the null hypothesis, implies that $\E{R_k}=0$.

### 2008-02-08 paper.tex
Lot of work...

@@
 
-\newcommand{\thetitle}{Linear Algebra as a Unifying Theory of Workload Modeling}
+\newcommand{\thetitle}{An Algebraic Representation of Wireless Workload}
 
@@
 \newcommand{\class}[1]{\textsc{\small{#1}}}
-\newcommand{\model}[1]{#1}
+\newcommand{\model}[1]{\textsf{\small{#1}}}
 \newcommand{\RFC}[1]{\caps{RFC}~{#1}}
@@
 \newcommand{\Qt}{Q_{\text{time}}}
+\newcommand{\Qb}{Q_{\text{byte}}}
 \newcommand{\Qi}{Q_{\text{ival}}}
@@
 
-\newcommand{\TTT}{\text{TTT}}
-\newcommand{\TTU}{\text{TTU}}
-\newcommand{\TUU}{\text{TUU}}
-\newcommand{\TUT}{\text{TUT}}
-\newcommand{\UTU}{\text{UTU}}
-\newcommand{\UTT}{\text{UTT}}
-
 \newcommand{\figurename}{Figure}
@@
 \begin{abstract}
-We present a representation of wireless workload patterns as large, sparse matrices, together with a method for stochastically generating experimental workload from any given matrix.
-Each row vector represents a single flow of packets in the network, with components representing the \caps{IP} protocol type, source and destination nodes, flow size, start time, and packet behavior.
-The essential property of the algebraic representation is that the summation of rows naturally yields a faithful description of the aggregate behavior of the corresponding flows.
-This deceptively simple property allows us to express many common types of traffic models succinctly in terms of a few matrix multiplications and linear constraints.
-This approach has many benefits: 1) it makes the meaning of certain generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2) it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3) the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed.
-We use the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models, to experimentally validate the general matrix model.
+We propose a representation of wireless workload patterns as large, sparse matrices and provide a method for stochastically generating experimental workload from a given matrix.
+The essential property of the algebraic representation is that the summation of vectors naturally yields a faithful description of the aggregate behavior of the corresponding flows.
+This deceptively simple property allows us to express many common concepts from traffic modeling succinctly in terms of a few linear transformations.
+The algebraic representation has many benefits: 1)~it makes the meaning of generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2)~it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3)~the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed; 4)~the vast body of algebraic theory and highly optimized numerical linear software may immediately be applied to traffic modeling.
+We use the paired differential simulation methodology introduced by the authors in previous work to experimentally demonstrate that the general matrix model accurately reproduces realistic network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+We use the same experimental methodology to explore the implications of various assumptions and simplifications that are commonly made in traffic modeling.
 %We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models.
@@
 The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
-Other research~\cite{Hernandez06:dissertation} has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback induced by the ubiquitous transmission control protocol~(\caps{TCP}).
+Other research has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback induced by the ubiquitous transmission control protocol~(\caps{TCP})~\cite{Hernandez06:dissertation}.
 Recently, the high-level statistical properties of wireless \caps{LAN} traffic have been rigorously modeled with a variety of parametric distributions~\cite{Hernandez06:wlan-traffic}.
@@
 Despite great breakthroughs in understanding and methodology, the networking community remains without a ``cookbook'' for generating realistic \caps{LAN} workloads.
-By ``cookbook,'' we mean an explicit method for generating network traffic---down to the size and send time of each application data unit.
+By ``cookbook,'' we mean an explicit method for generating network traffic---down to the size and transmission time of each application data unit.
+This lack is particularly relevant to wireless research, since the medium fundamentally resource-limited, and consequently local traffic patterns can have a huge impact on overall network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+
 In the current standard practice for experimental workload generation, drastic assumptions of uniformity abound.
-Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily.
+Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate.
+The flows in an experiment are commonly given the same duration and data rate.
+Nodes in the network are often selected as flow endpoints randomly or arbitrarily.
 While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
@@
 The first requirement is an objective standard of realism by which to compare traffic models.
-Such a standard has already been introduced in~\cite{Karpinski07:realism} and used to demonstrate that uniform traffic modeling, including the ubiquitous \caps{CBR} model, drastically distort performance metrics at all levels of the network protocol stack~\cite{Karpinski07:cbr-failure}.
-The second required breakthrough is a unified framework for expressing traffic models that allows researchers to systematically express and understand the assumptions and transformations entailed in each model.
-The major contribution of this paper is in addressing this second requirement.
-
-%Our previous work on traffic modeling provided the means
-
-In this paper, we present a highly general model for representing detailed full-network traffic patterns algebraically.
-The General Matrix Model (\caps{GMM}) represents the behavior or each network flow as a vector in a high-dimensional linear space, and represents total network behavior as a large matrix whose rows are these flow-behavior vectors.
+Such a standard has already been introduced and used to demonstrate that uniform traffic modeling, including the ubiquitous \caps{CBR} model, drastically distort performance metrics at all levels of the network protocol stack~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+This standard serves to demonstrate the harm of overly simplistic models, such as \caps{CBR}, and provides the means to test whether more complex models deliver the desired improvements.
+The second breakthrough that is needed to improve modeling is a unified framework for expressing a broad variety of traffic models, that allows researchers to systematically express and reason about the assumptions and simplifications entailed in each model.
+
+The major contribution of this paper is addressing this second requirement.
+We propose the General Matrix Model (\caps{GMM}) for local network traffic.
+In this model, the traffic pattern of an entire network is represented as a large, sparse matrix.
+Each row vector represents a single flow of packets in the network, with components representing the \caps{IP} protocol type, source and destination nodes, flow size, start time, and packet behavior.
 The crucial property of this representation is that it respects vector space operations: summation of behavior vectors gives an accurate description of the aggregate behavior of the flows represented by those vectors.
-The benefit of this representation is that it makes it immediately possible to express a broad variety of assumptions and conditions on traffic patterns in terms of simple matrix operations.
-It also provides vital mathematical clarity and allows us to apply all the tools of modern linear algebra to the problems of traffic modeling.
+One benefit of this representation is that it makes it immediately possible to express a broad variety of assumptions and conditions on traffic patterns in terms of simple matrix operations.
+It also provides vital mathematical clarity and allows us to apply the theoretical and computational tools of modern linear algebra to the problems of traffic analysis and modeling.
+
+After presenting \caps{GMM} and discussing how common concepts in traffic modeling can be expressed in precise mathematical terms within this framework, we proceed with experimental validation of \caps{GMM}.
+We use the technique of \textit{paired differential simulation} that the authors proposed and used in previous work~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+This technique uses real wireless trace traffic as an experimental control and conducts paired experiments comparing network performance from simulations using control traffic against others using synthetically generated traffic.
+The results are positive: \caps{GMM} accurately reproduces network performance characteristics at all levels of the protocol stack.
+%This does not solve the ``cookbook'' problem, however, because the \caps{GMM} must still be based on real traffic in order to reproduce its behavioral characteristics.
+It does however, reduce the problem of general traffic modeling to a more constrained problem of matrix modeling.
+
+Our experiments with various common assumptions of uniformity and regularity in network behavior yield mixed conclusions: most straightforward assumptions of homogenous behavior distort performance results significantly, but there are notable exceptions.
+The clearest exception is the time domain: traffic models appear to be relatively insensitive to local perturbation of the start time of flows---so long as the overall distribution remains approximately correct.
+Our conclusion is that traffic modeling needs to discard its assumptions of uniformity, regularity and independence.
+Instead we need to move ahead trying to understand the complex dependencies between the many aspects of traffic behaviors.
+By reducing traffic modeling to the well-understood domain of numerical linear analysis, the General Matrix Model brings us significantly closer to the ultimate goal of having a ``cookbook'' model for generating arbitrary, realistic experimental workload patterns.
 
@@
 Even with specific models for various aspects of network behavior, it remains non-trivial to combine them into a coherent composite model capable of generating concrete workloads.
-The {\FHC} models, for example, specify the distribution of flows per node in the network; since every flow has two end-points, however, it remains unclear how to assign end-points to flows such that the marginal distribution of flow per node matches the BiPareto model indicated.
+The {\FHC} models, for example, specify the distribution of flows per node in the network; since every flow has two endpoints, however, it remains unclear how to assign endpoints to flows such that the marginal distribution of flow per node matches the BiPareto model indicated.
 Similarly, the models dictate BiPareto marginal distribution of flow sizes across the entire network, but say nothing about how these sizes are to be distributed among the nodes.
@@
 
-%: objects and their properties together with rules for delegating, inheriting, and overriding the properties across the objects in the \caps{TOM} hierarchy. The \caps{TOM} allows
+% objects and their properties together with rules for delegating, inheriting, and overriding the properties across the objects in the \caps{TOM} hierarchy. The \caps{TOM} allows
 
@@
 None of the partially synthetic models manage to accurately represent more than one or two performance characteristics.
-We further use the \caps{AODV}~\cite{rfc:aodv} and \caps{OLSR}~\cite{rfc:olsr} ad~hoc routing protocols to show that the relative performance of protocols can be inverted when changing from using a simplistic traffic model to using real traffic: using random end-point, uniform \caps{CBR} traffic, \caps{AODV} appears to induce less link-layer overhead than \caps{OLSR}, whereas using real traffic, it in fact induces more.
+We further use the \caps{AODV}~\cite{rfc:aodv} and \caps{OLSR}~\cite{rfc:olsr} ad~hoc routing protocols to show that the relative performance of protocols can be inverted when changing from using a simplistic traffic model to using real traffic: using random endpoint, uniform \caps{CBR} traffic, \caps{AODV} appears to induce less link-layer overhead than \caps{OLSR}, whereas using real traffic, it in fact induces more.
 
@@
 The rest of the paper is organized as follows.
-In \Section{motivation} we present motivation and related work.
+In \Section{motivation} we discuss motivation and related work.
 The General Matrix Model is presented in \Section{general-matrix-model}.
-Our experimental and analytical methodology for evaluating traffic models presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
+Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
 The ramifications of these results are discussed in \Section{discussion}.
@@
 However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid.
-The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.} The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic and motion in the experiments are sufficiently realistic.
-
-Paxson and Floyd observed in~\cite{Paxson95} that the interplay between end-point behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback.
+The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.} The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic.
+
+Paxson and Floyd observed in~\cite{Paxson95} that the interplay between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback.
 Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}.
-This presents a fundamental difficulty, however, in that it presumes that we know the intent of end-points: what \textit{would} they have done under different conditions?
-While we can speculate about what an individual node might hypothetically do, we currently do not understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations.
+This presents a fundamental difficulty, however, in that it presumes that we know the intent of endpoints: what \textit{would} they have done under different conditions?
+While we can speculate about what an individual node might hypothetically do, we currently hardly understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations.
 Especially in the wireless setting, a fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback.
@@
 The focus of this project, however, is on providing the infrastructure to generate very large volumes of synthetic Internet-like backbone traffic.
-No analysis is provided for determining the realism of traffic mixes, or for choosing flow end-points realistically.
+No analysis is provided for determining the realism of traffic mixes, or for choosing flow endpoints realistically.
 In wireless networks, these factors are of crucial importance to performance, and cannot be overlooked.
@@
 Such a sequence of packets is called a \textit{flow}.
-The traffic pattern is characterized completely by the detailed behavior of it's flows.
-The behavior of a flow is defined by six properties: its source and destination nodes; its start and end times; and the specific sizes of and intervals between its constituent packets.
-We propose to represent the behavior of each flow in terms of six vectors---one for each aspect of its behavior.
+The traffic pattern of an entire network is simply the collected behaviors of all flows occurring in the network.
+The behavior of a flow as it affects the network is characterized by the following properties: its \caps{IP} protocol type (\caps{TCP}, \caps{UDP}, \caps{ICMP}, etc.); its source and destination nodes; its start time; and the specific sizes of and intervals between its constituent packets.
+We propose to represent the behavior of each flow in terms of seven vectors, which suffice to describe all of the above properties in enough detail to stochastically generate realistic traffic.
 The aggregate behavior of all the flows in a workload pattern are then represented as matrices, the rows of which are the behavior vectors of the flows.
@@
 The {packets-bytes-duration} representation is clearly overly simplistic and very incomplete.
-Yet, in giving a description of each flow, it is far more detailed than the most commonly used experimental traffic models.
-It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start and end time unspecified; it says nothing about the source or destination of the flow.
+It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time and flow size unspecified; it says nothing about the source or destination of the flow.
 At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets, etc.
@@
 The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows.
-We take a highly detailed but intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the correct source node, destination node, start time, end time, and empirical distributions of packet sizes and inter-packet intervals.
+We take a highly detailed but intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the flow type, source node, destination node, start time, flow size, and empirical distributions of packet sizes and inter-packet intervals.
 If there are $f$ flows in the entire workload pattern, then the collective behavior can be described with six matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
@@
 $D\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
+\>\bullet\> type: \>$T$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$,\\
 \>\bullet\> sources: \>$S$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$,\\
 \>\bullet\> destinations: \>$D$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$,\\
-\>\bullet\> start times: \>$B$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_t}$,\\
-\>\bullet\> end times: \>$E$\>$=\parensx{\vec{e}_i}$\>$=\parensx{e_{ij}}$\>$\in\R^{f \times d_t}$,\\
+\>\bullet\> start times: \>$A$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$,\\
+\>\bullet\> sizes in bytes: \>$B$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$,\\
 \>\bullet\> packet sizes: \>$Z$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
@@
 \end{tabbing}
-The total behavior is the direct product of these components: $F = \seq{S,D,B,E,Z,V} \in \R^{f \times \Sigma}$ with $\Sigma=2n+2d_t+d_z+d_v$.
+The total behavior is the direct product of these components: $F = \seq{T,S,D,A,B,Z,V} \in \R^{f \times \Sigma}$ where $\Sigma=256+2n+d_t+d_b+d_z+d_v$.
 Many common traffic models can be expressed succinctly and precisely in terms of transformations or conditions on this behavior matrix and its constituents.
@@
 
+\subsection{Flow Type}
+
+The type of a flow is its \caps{IP} protocol number.
+This ranges in possible values from 1 to 255 (0 is reserved for \caps{IPv6}).
+We represent the flow type as an indicator vector: a 255-dimensional vector consisting of all zeros except for the index which is the flow type.
+So a \caps{TCP} flow, having protocol type 6, would have the type vector $\vec{t}=\seq{0,0,0,0,0,1,0,\dots,0}$.
+This representation may immediately strike one as unnecessarily expansive and inefficient since a single byte suffices to represent the protocol type.
+The necessity of this representation arises when we represent aggregations of flows as sums of vectors.
+The sum of protocol numbers is meaningless: two \caps{TCP} flows plus five \caps{ICMP} flows does not equal a single \caps{UDP} flow, despite the fact that $2\cdot6+5\cdot1=17$.
+The summation of the corresponding indicator vectors, however, is $\seq{5,0,0,0,0,2,0,\dots}$.
+This vector is a histogram indicating the frequency with which each protocol type occurred in the collection of flows over which the sum was taken.
+
 \subsection{Source \& Destination Nodes}
@@
 We let $s_{ij}=1$ if the $j$th node is the source of the $i$th flow, and zero otherwise; similarly, $d_{ik}=1$ if the destination of the $i$th flow is the $k$th node.
-This sparse matrix may seem to be an overly expansive representation, given that start and end nodes can be represented simply as indices from $\set{1,\dots,n}$.
-The advantage becomes clear when we consider vector operations aggregating collections of behavior vectors.
-The sum of the source indices over a collection is meaningless; the sum of a set of source vectors, $\set{\vec{s}_i}$, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.
-
-\subsection{Start \& End Times}
-
-We represent start and end times using indicator vectors.
+This sparse matrix may again seem to be an overly expansive representation, given that start and end nodes can be represented simply as indices from $\set{1,\dots,n}$.
+The sum of a set of source vectors, $\set{\vec{s}_i}$, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.
+
+\subsection{Start Time}
+
+We represent start time of each flow using an indicator vector.
 Unlike node indices, however, time is continuous, so we must first map times onto a range of time indices $\set{1,\dots,d_t}$.
@@
 %Suppose the $i$th flow has a quantized start index of $j$ and a end index of $k$. Then $b_{ij'}=\delta{jj'}$ and $e_{ik'}=\delta_{kk'}$.
-Let $j$ and $k$ be the quantized start and end indices of a flow.
-Then the start and end vectors for the flow are $\vec{b}=\mathbf{e}_j$ and $\vec{e}=\mathbf{e}_k$ ($\mathbf{e}_j$ has 1 at index $j$ and zeros elsewhere).
-Again, this may seem excessively complex, but the advantage becomes clear when we aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.
+Let $k$ be the quantized start index of a flow.
+Then the start vectors for the flow is $\vec{a}=\mathbf{e}_k$ ($\mathbf{e}_k$ has 1 at index $k$ and zeros elsewhere).
+The advantage of this representation becomes clear when we aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.
+In practice, we use $d_t=1500$.
+
+\subsection{Flow Size}
+
+We represent flow size with another indicator vector.
+Although the set of possible flow sizes is discrete, for practical reasons we map the very large range of possible flow sizes down to a much smaller range of indices (some flows in our trace have 109\caps{MB} of data in a single ten-minute period, which would give an indicator vector of over 100 million dimensions).
+We use the following function to convert flow sizes to indexes:
+\begin{align*}
+%\label{eqn:byte-quantization}
+\Qb(b) &= \max\set{1,\ceil{d_b\fracp{b-1}{b_{\max}-1}^\beta}}.
+\end{align*}
+Here $b_{\max}$ is the maximum input flow size.
+The parameter $\beta$ is a non-linear scaling factor exponent that allows the quantization to smoothly shift from byte-sized resolution at the low end of the possible range of flow sizes to the a resolution of megabytes at the high end of the spectrum.
+In our example data, $b_{\max}\approx109\text{\caps{MB}}$ as we previously mentioned.
+We use a representation dimensionality of $d_b=1500$, to match our other quantized vectors, and $\beta=\fracx{1}{2.7}$ because it yields the desired dynamic range response of the possible flow sizes.
 
@@
 The representation of inter-packet intervals as a vector, $\vec{v}$, is similar, but the domain of possible intervals is a continuum, so we must map this range onto the discrete index values $\left\{1,\dots,d_v\right\}$.
-We use the following adaptive quantization function:
+We use similar non-linear quantization function as before:
 \begin{align}
@@
 \subsection{Traffic Generation}
+\label{sec:traffic-generation}
 
@@
 Each row of the model is used to randomly generate the behavior of a single flow.
-The source and destination nodes of the flow are randomly chosen, using histogram vectors $\vec{s}$ and $\vec{d}$ as the proportional probabilities of each node being chosen.\footnote{Since a flow must have distinct source and destination nodes, this sampling must be done jointly, avoiding the collisions.
+The type, source, and destination nodes of the flow are randomly chosen, using histogram vectors $\vec{t}$, $\vec{s}$, and $\vec{d}$ as the proportional probabilities of each type or node being chosen.\footnote{Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
 The simplest technique is the rejection method: if the same node is chosen for both source and destination, reject that pair and begin the selection process again.}
-For start and end times, a bin for each is chosen with probability proportional to the vector values; given a bin, a random value is uniformly chosen from the range of values in the bin.\footnote{The start time must precede the end time, so these choices must also be made jointly, enforcing this restriction.
-The rejection method is used.}
-The first packet is sent at chosen start time; the size of each packet is sampled from the histogram of packet sizes, $\vec{z}$; the interval until the next packet is determined from the interval vector, $\vec{v}$.
+Next we use the flow size vector, $\vec{b}$ to randomly sample the number of bytes in the flow.
+The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\dot{z}$ and $\dot{v}$.
+From these, together with the flow size in bytes, we estimate the expected duration of the flow: $\dot{d} = \fracx{b\dot{v}}{\dot{z}}$.
+We finally choose a start time for the flow, using $\vec{a}$ as a vector of relative probabilities for each bin of start times, but ensuring that the start time plus the expected duration, $\dot{d}$, does not exceed the maximum simulation time for the workflow.
+From the range of possible values mapping to the chosen start bin, we choose a random time uniformly.
+
+Having determined the overall parameters of the flow, we generate the actual packets of the flow.
+The first packet is sent at the chosen start time; the size of each packet is sampled from the histogram of packet sizes, $\vec{z}$; the interval until the next packet is determined from the interval vector, $\vec{v}$.
 Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}(k)$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\seq{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
@@
 \end{align}
-The packet stream ends when a packet would be sent after the chosen end time of the flow.
+The packet stream ends when the number of bytes allocated to the flow have been emitted. %The last packet is only the size of the remaining number of bytes.
 
@@
 
-\subsection{Nodes: Communication Topology}
+\subsection{Communication Topology}
 \label{sec:communication-topology}
@@
 
-\subsection{Flows: Packet Count, Size \& Duration}
+\subsection{Expected Average Packet Size \& Intervals}
 \label{sec:flow-totals}
 
-From the matrices $Z$ and $V$, we can compute the packet count, size (total bytes), and approximate duration for each flow.
-The computations for size and duration require knowing the values that each matrix column represents.
+From the matrices $Z$ and $V$, we can compute the expected average packet size, $\dot{z}$, and average inter-packet interval, $\dot{v}$, for each flow.
+The computations require knowing the values that each matrix column represents.
 The computed duration is only approximate because information is irrecoverably lost in the quantization process (see \Equation{interval-quantization}).
@@
 \end{align}
-Using these representative values, the vector of packet-counts is $Z\ones{d_z}=V\ones{d_v}$, the vector of flow sizes vector is $Z\vec{\sigma}$, and the vector of approximate durations is $V\vec{\tau}$.
-
-%\subsection{Network: Marginal Behavior}
-%\label{sec:marginal-behavior}
-
-%\noindent
-%...
-
-\section{Expression of Common Models}
-
-Existing traffic models are typically not explicit methods of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
-In our algebraic modeling framework, such models can be interpreted as stipulations about the form of the behavior matrices, or as transforms that implement those stipulations given a real set of traffic matrices.
-
-%A generic traffic model is a way of generating the six matrices described in \Section{representations}.
-A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or regularized form.
-For experimental comparison, the dimensions of the matrices should all be preserved.
-If a trace-derived model produces experimental performance metrics that are similar to those produced by the original trace data, then simplifications made by the model are viable; if the performance is distorted, then the simplifications are discarding essential information about the traffic pattern.
-
-\subsection{Uniform Models}
-
-It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees, largely because of the high complexity of doing otherwise and the lack of compelling alternatives.
+Using these representative values, the expected average packet size for a flow may be computed as $\dot{z}=\fracx{\vec{z}\cdot\vec{\sigma}}{\vec{z}\cdot\ones{d_z}}$, and the expected average inter-packet interval size as $\dot{v}=\fracx{\vec{v}\cdot\vec{\tau}}{\vec{v}\cdot\ones{d_v}}$.
+
+\section{Expression of Common Modeling Concepts}
+
+The simplifying assumptions made by most traffic models can be classified into three categories.
+We use the terms \textit{uniformity}, \textit{marginality}, and \textit{regularity} do describe these classes of assumptions.
+They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether the make stipulations about uniformity within each flow, across all flows, or within groups of flows.
+
+\subsection{Uniform Behavior}
+
+The concept of uniformity, as used here, entails that certain properties of each flow are assumed to be statistically or deterministically uniform.
+Examples include: assuming that each flow's packets have the same size and inter-packet interval (i.e.
+the \caps{CBR} flow model); assuming that each node is equally likely to be the source of a given flow; assuming that all start times for a flow are equally likely.
+
+Since uniform behaviors entail homogeneity of properties with respect to each single flow individually, a trace matrix can be transformed to the closest matrix satisfying the model's assumptions via right matrix multiplication.
+Right multiplication mixes the elements within each row, and if the mixing weights are all equal, the results is homogenized behavior.
+For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $S$ and $D$:
+\begin{align*}
+S' &= S\ones{n \times n}/n, & D' &= D\ones{n \times n}/n.
+\end{align*}
+
+%We use the term \textit{uniformity} do describe assumptions made about the ``sameness'' of various properties of flows.
+The constant bit-rate (\caps{CBR}) model, for example, is an assumption about the uniformity of all of a flow's packets: i.e.
+that they have the same size and are evenly spaced in time.
+While this is the strictest sense of the \caps{CBR} model, it is rare for this assumption to be made in isolation.
+Another common simplification is to make the sizes of flows identical, or chose them randomly from a uniform distribution; similarly for the starting time of each flow.
+Finally, when choosing endpoints for a given flow, lacking a better model, it is not unusual to treat all nodes identically, making each an equally likely source or destination for a given flow.
+
+It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees.
+This is largely because of the high complexity of doing otherwise and the lack of compelling alternatives.
 Here we examine how various notions of uniformity are expressed in the matrix model.
@@
 
-%Deriving the packet-bytes-duration model from the general matrix model is accomplished through right multiplication as described in \Section{flow-totals}.
-% TODO: this conclusion is pretty weak.
-
-While constant bit-rate in the strict sense implies only that the packet behavior in each flow is uniform, it is very rare for it to be applied in this strict sense only; it is typical that ``\caps{CBR}'' models make assumptions about uniformity across flows, and not just within each flow.
+While constant bit-rate in the strict sense implies only that the packet behavior in each flow is uniform, it is very rare for it to be applied in this strict sense only: it is typical that ``\caps{CBR}'' models make assumptions about uniformity across flows, and not just within each flow.
 The most drastic simplification is to treat all flows in a synthetic workload identically.
 This is effected by left multiplication: $F'=(\ones{f \times f}/f)F$.
-To make flow behavior uniform at the per-source node level, we left-multiply by $SS^t/f\in\R^{f \times f}$: $F'=(SS^t/f)F$.
-Per-destination node uniformity can be achieved similarly: $F'=(DD^t/f)F$.
+To make flow behavior uniform on a per-source-node basis, we left-multiply the ``source-uniformizing'' matrix: $F'=(SS^t/f)F$.
+Per-destination-node uniformity can be achieved similarly, with left-multiplication by a corresponding destination-uniformizing matrix: $F'=(DD^t/f)F$.
 
@@
 
-Another benefit of the matrix model is that all aspects of network behavior are determined concurrently.
-We are not left wondering what to do about source and destination nodes or flow start and end times after we have computed our simplified packet behaviors: the same transformation that average the packet behaviors also naturally averages the start and
-
-\section{Experimental Methodology}\label{sec:methodology}
-
-To evaluate whether traffic models are realistic or not, we use the differential simulation methodology introduced in~\cite{Karpinski07:realism} and used in~\cite{Karpinski07:cbr-failure} to compare a variety of uniform and trace-based models.
-This approach is based on the following definition, initially proposed in~\cite{Karpinski07:realism} and formalized in \cite{Karpinski07:cbr-failure}:
-
-\vspace{0.5em}
-\begin{samepage}
-\noindent\textbf{Definition.}~A model of user behavior is \textit{sufficiently realistic} if, compared with actual user behavior, the model, with parameter values extracted from the real data, yields statistically equivalent performance results.
-\end{samepage}
-\vspace{0.5em}
-
-The essential idea behind the differential simulation technique is to compare traffic produced by synthetic models with known examples of actual real-world trace traffic in a variety of experimental scenarios.
-To accomplish this, we begin with a large 24-hour example of wireless trace traffic, break it into 144 ten-minute scenarios, and use each scenario's behavior as a control for a set synthetic traces, generated by various synthetic traffic models whose realism we with to evaluate.
-For each scenario, the synthetic models are configured to approximate the original trace behavior as closely as the models allow.
-Important network-wide performance metrics are then collected from simulations, and synthetic performance results compared with control results to determine the realism of the synthetic models.
-
-%The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results.
-Clearly, we need not simulate subatomic particle interactions in a wireless network simulator.
-Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same.
-Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior.
-This requirement leads us to the following definition:
-
-%This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required.
-We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.
-
-\subsection{Traffic Models}
-\label{sec:traffic-models}
-
-\begin{table*}
-\begin{center}
-\small
-\begin{tabular}{|c|c|l|}
-\multicolumn{1}{c}{\textbf{model}} &
-\multicolumn{1}{c}{\textbf{transformation}} &
-\multicolumn{1}{c}{\textbf{description}} \\
-\hline
-{\footnotesize{GMM}} & --- & the general matrix model \\\hline\hline
-flow uniform & $F'=(\ones{f \times f}/f)F$ & flows are generated stochastically identically \\\hline
-node uniform & $S'=S(\ones{n \times n}/n)$~~;~~$D'=D(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
-time uniform & $B'=B(\ones{d_t \times d_t}/d_t)$~~;~~$E'=E(\ones{d_t \times d_t}/d_t)$ & all start times and end times are equally likely \\\hline
-full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above three models \\\hline\hline
-node marginal & $S'=(\ones{f \times f}/f)S$~~;~~$D'=(\ones{f \times f}/f)D$ & nodes behave distinctly, but behavior is independent of flows \\\hline
-packet marginal & $Z'=(\ones{f \times f}/f)Z$~~;~~$V'=(\ones{f \times f}/f)V$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
-\end{tabular}
-\caption{Matrix-based traffic models evaluated.}
-\label{tab:traffic-models}
-\end{center}
-\end{table*}
-
-The original trace traffic constitutes the first model we simulate, although it is not really a model, {per se}.
-This serves as a control group against which to compare all other models.
-Before we can evaluate the General Matrix Model (\caps{GMM}) presented in \Section{general-matrix-model}, we must establish that one potentially important aspect of traffic behavior that is \textit{not} represented in the \caps{GMM} can legitimately be discarded.
-That aspect of behavior is \textit{\caps{TCP} feedback}.
-The \caps{GMM} does not specify the \caps{IP} protocol type of each flow.
-Therefore we must establish the extent to which using \caps{UDP} flows, without reliable delivery or congestion control, affects performance metrics.
-%\caps{GMM} also does not indicate which flows are paired with each other in a client-server relationship. While this pairing does not affect non-\caps{TCP} flows, we implement.
-Other synthetic models derived from algebraic transformations of the \caps{GMM} are listed in \Table{traffic-models}.
-
-\subsection{Trace Data}\label{sec:trace-data}
-%\subsection{Trace Data \& Simulations}\label{sec:trace-data}\label{sec:simulations}
-
-Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of trace-based models and synthetic traffic models defined using the General Matrix Model.
-We use a 24-hour trace recorded in an infrastructured 802.11g wireless \caps{LAN} with 18 access points, deployed at the 60th Internet Engineering Task Force meeting (\caps{IETF}60), held in San Diego during August of 2004.
-The traffic trace was captured using \texttt{\small{tcpdump}} at a single router, through which all wireless traffic for the meeting was routed, including traffic between wireless nodes.
-The snap length of the capture was 100 bytes, allowing \caps{IP}, \caps{ICMP}, \caps{UDP} and \caps{TCP} headers to be analyzed.
-We limit our work to the 24-hour sub-trace recorded on Wednesday, August 4th.
-This trace contains a broad variety of behaviors and entails a very large volume of traffic: 2.1 million flows, 58 million packets, and 52 billion bytes.
-% 2,085,563 flows, 58,377,994 packets of application data, and 51,988,245,537 bytes of application data.
-
-\begin{figure}
-\vspace{0.45em}
-\begin{center}
-\includegraphics[width=3.5in]{nodes-flows}%
-\vspace{-0.75em}%
-\caption{The number of active nodes and flows over time.} 
-\label{fig:nodes-flows}
-\end{center}
-\vspace{-2em}
-\end{figure}
-
-We do not assume or claim that the traffic found at \caps{IETF}60 is representative of conference settings in general.
-The observed behaviors are also unlikely to resemble those found in a typical commercial or residential setting.
-We have chosen this trace, however, because within it can be found behaviors resembling many different types of wireless usage cases.
-Figure~\ref{fig:nodes-flows} shows the wide variations in the number of active flows and nodes over the course of the trace.
-In the night and morning hours, the traffic patterns are similar to those one might find in a moderately trafficked business or residential area.
-During working group sessions, we see highly concentrated, heavy usage patterns.
-At the zenith of activity, over 800 users, 33 thousand flows, and 1 million packets are seen in a single 10-minute trace segment.
-At the nadir, a lone node sent only a single 61-byte packet over the course of 10 minutes.
-All levels of activity between these extremes are represented.
-Moreover, the mix of traffic types observed changes dramatically over the course of the day, providing a wide representation of possible blends of behavior.
-This heterogeneity and extreme range of behaviors makes the \caps{IETF} data set ideal for this evaluation.
-The variety of activity gives us greater confidence that success or failure of traffic models is not tied to any specific network condition, but is broadly and generally applicable.
-%For a traffic model to be truly realistic, it must be realistic across many different usage cases.
-
-Before using the traces, it is necessary to extract application-level behavior from the trace header data.
-First, we split the trace into individual packet flows.
-A flow is a series of packets sharing the following five attributes: \caps{IP} and transport protocols (raw \caps{IP}, \caps{ICMP}, \caps{TCP}, \caps{UDP}); source and destination \caps{IP} addresses and \caps{TCP}/\caps{UDP} port numbers.
-Next, the quantity of application-initiated data contained in each packet is calculated.
-For non-\caps{TCP} packets, this quantity is simply the size of the transport-layer payload, but for \caps{TCP} the calculation is more complicated: only new data transfers, explicitly initiated by the application are counted.
-Data retransmitted by \caps{TCP} is disregarded, and empty \caps{ACK}s are ignored.
-\caps{SYN} and \caps{FIN} flags in packets (even empty ones) are counted as a single byte each, since they are explicitly signaled by the application.
-%After processing, we have a collection of flows, each with a sequence of timestamps and the amount of application-initiated data sent at that time.
-
-%\begin{table}[t]
-%\vspace{0.75em}
-%\begin{center}
-%\begin{tabular}[t]{|l|}
-%\hline
-%\begin{minipage}[l]{2.7in}
-%\begin{small}
-%\vspace{0.7em}
-%\input{properties.tex}
-%%\vspace{0.6em}
-%\end{small}
-%\end{minipage}
-%\\\hline
-%\end{tabular}
-%\end{center}
-%\vspace{-0.5em}
-%\caption{{\footnotesize{TOM}} base classes and their attributes.}
-%\label{tab:properties}
-%\vspace{-2em}
-%\end{table}
-
-%\subsection{The Traffic Object Model~(\caps{TOM})}\label{sec:tom}
-
-%\begin{table*}[t]
-%\begin{center}
-%\small
-%\input{models.tex}
-%\footnotesize\\[0.5em]
-%%\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
-%\end{center}
-%\vspace{-0.75em}
-%\caption{Details of how the sub-models at each behavioral level of the {\footnotesize{TOM}} inherit, override and delegate properties.}
-%\label{tab:models}
-%\vspace{-1em}
-%\end{table*}
-
-%In this section, we present a generic object-oriented framework for implementing traffic generation models.
-This framework is called the Traffic Object Model (\caps{TOM}).
-It defines four levels of responsibility: network, node, flow and packet.
-The framework is designed so that different behaviors at each level can be naturally implemented  as classes in object-oriented programming languages.
-The framework automatically gains the advantages of object orientation, in particular modularity and encapsulation.
-Every traffic model is naturally split into separate spheres of responsibility, which constitute sub-models for the aspects of the model's behavior at each layer.
-Encapsulation and polymorphism of objects, allows us to mix-and-match the sub-models from different models to create new hybrid behaviors.
-It is trivial, for example, to produce trace-based node behavior, coupled with parametric flow behavior, and uniform packet behavior.
-Not all combinations of sub-models are valid, however; which classes may be combined depends on the details of each class.
-
-%% TODO: fix-up client-server vs. source-destination terminology.
-
-%Abstract base classes define the properties at each level that traffic models must implement.\footnote{In Java or C\#, this role may be served either by an actual abstract class or by an interface.
-In duck-typed languages such as Ruby or Python, no actual abstract class need exist, since type is implicitly defined by usage; if abstract classes are desired, they can be simulated by creating base classes with stub methods that throw informative exceptions upon being called.}
-%Table~\ref{tab:properties} enumerates all the properties exposed by traffic models used in this work.
-Future traffic models may necessitate exposing additional properties, but this list suffices to implement the models used in this paper.
-Concrete subclasses implement the behavior of specific traffic models at each level.
-A complete synthetic workload is instantiated as follows:
-%\begin{enumerate}
-%\item create a \class{Network} object of the appropriate subclass;
-%\item the top-level \class{Network} object in turn creates a collection of \class{Node} objects and associated \class{Flow} objects;
-%\item each \class{Flow} object generates a stream of \class{Packet} objects.
-%\end{enumerate}
-%The hierarchy of objects may be used to directly drive a simulation, or may generate configuration files that in turn drive a simulation.
-Our implementation does the latter, generating a collection of Qualnet configuration files and trace files specific to each workload instantiation (see \Section{simulations}).
-
-%\subsection{Traffic Models}\label{sec:models}
-
-%For each of the \class{Node}, \class{Flow} and \class{Packet} abstract superclasses, we define four concrete subclasses: \class{Trace}, \class{Nonparametric}, \class{Parametric}, and \class{Uniform}.
-The \class{Trace} classes take properties of objects directly from the behavior of the corresponding object in the original trace.
-The \class{Nonparametric} and \class{Parametric} classes use models of the corresponding type to randomly generate the properties of objects.
-The \class{Uniform} classes generate identical objects with average properties.
-The specific details of how each the concrete classes at each level are defined are specified in Table~\ref{tab:models}.
-The parameters for the fitted models used in the \class{Parametric} subclasses are estimated from the corresponding nonparametric distribution using maximum likelihood estimation.
-Figure~\ref{fig:trace-vs-uniform} shows visual depictions comparing trace versus uniform behavior at various levels in example scenarios.
-
-%While the modularity of the Traffic Object Model allows us to mix and match sub-models, not all possible combinations are valid.
-In particular, note that the \class{Trace} classes are tied to a specific object from the underlying trace, whereas for other classes, there is no such correspondence to any specific object.
-Consequently, once a non-trace model has been used, trace models may not be used at the lower levels.
-This still leaves 40 valid composite combinations of sub-models; we only evaluate twelve of these.
-Composite models are abbreviated by the initials of their constituent sub-models.
-The fully trace-base model, \class{TTT}, serves as an experimental control.
-We consider \class{TTN}, \class{TTP}, and \class{TTU} in order to compare models of packet behavior.
-We evaluate different flow models, using the best possible model at the packet level: \class{TNN}, \class{TPN}, \class{TUN}.
-Similarly, we compare models for node behavior: \class{NNN}, \class{PNN}, \class{UNN}.
-Finally, we use the na\"ive uniform model, \class{UUU}, and the composite parametric model, \class{PPP}, proposed by {\FHC}~\cite{Hernandez06:wlan-traffic}, as external points of comparison.
+\subsection{Marginal Behavior}
+
+Marginality is an orthogonal concept to uniformity: it implies that the distribution of some property is identical \textit{across} all flows.
+For example, if we determine a distribution of flow sizes across the entire network and assume that this distribution applies \textit{independent of other flow properties}, this is a marginal model for flow size.
+The traffic models proposed by {\FHC}~\cite{Hernandez06:wlan-traffic} are all marginal: they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node.
+These models do not account for non-independence between these and other flow properties.
+We will discuss this limitation and its implications when we analyze our experimental results.
+
+\subsubsection{Regularized Behavior}
+
+The final form of sameness we consider, is regularization.
+This is a restricted form of marginality: we may assume that marginal distributions apply to all the flows in certain groupings.
+For example, we can provide a distribution of flow sizes on a per source-node basis; that is, each source node can have it's \textit{own} distribution, and the sizes of its flows are drawn from that distribution, independent of their other properties.
+
+%We use the terms here to distinguish three fundamentally different types of assumptions about behavioral homogeneity.
+
+%Existing traffic models are typically not explicit methods of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
+In our algebraic modeling framework, such models can be interpreted as stipulations about the form of the behavior matrices, or as transforms that implement those stipulations given a real set of traffic matrices.
+
+%A generic traffic model is a way of generating the six matrices described in \Section{representations}.
+A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or regularized form.
+For experimental comparison, the dimensions of the matrices should all be preserved.
+If a trace-derived model produces experimental performance metrics that are similar to those produced by the original trace data, then simplifications made by the model are viable; if the performance is distorted, then the simplifications are discarding essential information about the traffic pattern.
+
+\subsection{Uniformity}
 
@@
 
-%We separate traffic generation into three levels of behavior: network, flow and packet.
-The {network} level determines properties pertaining to nodes and their interaction.
-For example, it determines which nodes communicate with each other, and how frequently.
-This level can be viewed as assigning network nodes to the source and destination roles of flows, as well as determining per-node properties which affect lower level models.
-The next level of behavior determines overall properties of flows.
-This includes each flow's start time, duration, packet count and total data volume.
-Finally, the packet level of behavior determines the individual sizes of packets and the durations of intervals between their transmission.
-Despite the concreteness of the examples mentioned here, there are also less obvious, abstract properties that models at the network and flow levels determine.
-Flow models, for example, dictate the distribution of each flow's packet sizes, a property which can be used to determine individual packet sizes, but which is not itself specific to any particular packet.
-
-%\subsubsection*{Network Behavior Models}
-
-%We consider four models of network-level behavior. The most faithful is the ``trace'' model, which assigns each flow
-
-%For each level of behavior we define a spectrum of models.
-The gold standard of realism at each level is provided by the ``trace'' model for that level, which replicates the node, flow or packet behavior of the original trace exactly.
-The most rudimentary model at each level is the ``uniform'' model, which treats all entities identically---be they nodes, flows or packets.
-Despite being so simplistic (or perhaps because of it), the uniform behavior models are by far the most commonly for workload generation in experiments.
-Between these extremes on the spectrum of realism, we consider one or two models with intermediate degrees of complexity and fidelity.
-At the packet level, we define the ``sample'' model, which is in essence a per-flow variable bit-rate (\caps{VBR}) model: each packet's size is determined by randomly sampling a value from the packet size distribution given by the flow model; the interval until the next packet is then sampled from the given interval distribution.\footnote{Note that the uniform packet model (\textit{i.e.} {\scriptsize{CBR}}), can be considered to be a special case of the sample model where the size and time distributions are both constant.} The intermediate model at the flow level is designated as ``proportional'': the distributions of packet sizes and inter-packet intervals are taken from the trace and passed down to the packet model (which may use the actual distribution, via sampling, or simply use the mean size and intervals, to generate \caps{CBR} traffic).
-
-%\begin{enumerate}
-%\item \textbf{Network Behavior}: which nodes communicate with each other, and how frequently; \textit{i.e.} how flow end-points are mapped onto nodes in the network.
-%\item \textbf{Flow Behavior}: high-level parameters for each flow, including start time, end time, packets sent, bytes sent.
-%\item \textbf{Packet Behavior}: sizes of individual packets, and the intervals between their transmission.
-%\end{enumerate}
-
-%The different levels of models and variants at each level are listed and described in Table~\ref{tab:traffic-models} and illustrated with specific examples in Figure~\ref{fig:traffic-models}.
-
-% TODO: more here on models...
-% TODO: rewrite next paragraph too...
-
-%The three levels of behavior are orthogonal and can be varied almost independently.
-The exception to their independence is that the trace-based packet behavior model can only be used when the flow behavior model is also trace-based.
-Once flow behavior is decoupled from the actual trace, there is no natural way to preserve packet behavior.
-This eliminates three combinations of models and leaves nine viable behaviors, abbreviated by the first letters of their flow topology, flow behavior, and packet behavior models: \caps{TTT} (fully trace-based), \caps{TTU}, \caps{TUU}, \caps{STT}, \caps{STU}, \caps{SUU} (entirely synthetic, sink topology), \caps{UTT}, \caps{UTU}, \caps{UUU} (entirely synthetic, uniform topology).
-
-\subsection{Performance Metrics}\label{sec:metrics}
-
-We have selected seven performance metrics at the application, network, and link layers of the protocol stack:
-\begin{enumerate}
-\setlength{\itemsep}{0em}
-\item \textbf{Application:} average end-to-end delay, average jitter, total received throughput.
-\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), routing queue drop rate.
-\item \textbf{Link (\caps{MAC}):} control overhead (\caps{RTS/CTS/ACK}), packet retransmission rate.
-% TODO: are we actually using RTS/CTS?
-\end{enumerate}
-These metrics are commonly used to evaluate the performance of wireless protocols.%For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
-%Unfortunately, not all of the simulations were able to run to completion.
-In some scenarios, the number of nodes and flows caused Qualnet to simply fail.
-We were, however, able to get complete results for 72.5\% of the simulations, providing a clear view of overall trends.\!\footnote{That 27.5\% of simulations did not finish may seem like a cause for concern, but it does not in fact invalidate our results.
-The simulations that did not finish are those for which the number of flows and nodes overwhelmed the capacity of the simulator.
-Qualnet cannot simulate more than circa 650 nodes sending 35,000 flows.
-While it would be of great interest to examine results for these large-scale scenarios, they are essentially useless for comparison to other simulations studies: other studies do not simulate this quantity of nodes and flows precisely because even the best existing simulators do not support it.
-Once the simulators become more scalable and robust, our analysis can be extended to these scenarios as well.}
-
-\subsection{Simulations}\label{sec:simulations}
+% TODO: differentiate between restrictive versions of models and transformative versions.
+% TODO: distinguish uniform, marginal and regularized models.
+
+\subsection{Marginality}
+
+Most approaches to traffic modeling that go beyond uniformity assumptions, instead describe marginal properties of various aspects of traffic behavior.
+For example, {\FHC}~\cite{Hernandez06:wlan-traffic} provide highly convincing parametric models for six marginal properties of wireless network traffic: session (node) arrivals follow a time-varying Poisson process; per-session flow inter-arrival times are Lognormally distributed; flows per session (node) follow a BiPareto distribution; flow sizes follow a BiPareto distribution.
+This is certainly much greater detail than simply assuming that all nodes and flows are identical.
+
+%How do marginal models translate into the algebraic language of the General Matrix Model? Whereas uniformity assumptions for nodes,
+
+%Another benefit of the matrix model is that all aspects of network behavior are determined concurrently.
+We are not left wondering what to do about source and destination nodes or flow start and end times after we have computed our simplified packet behaviors: the same transformation that average the packet behaviors also naturally averages the start and
+
+\section{Experimental Methodology}\label{sec:methodology}
+
+To evaluate whether traffic models are realistic or not, we use the method of \textit{paired differential simulation} introduced in our previous work on traffic modeling~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+%This approach is based on the following definition:
+%
+%\begin{samepage}
+%\noindent\textbf{Definition.}~A model of user behavior is \textit{sufficiently realistic} if, compared with actual user behavior, the model, with parameter values extracted from the real data, yields statistically equivalent performance results.
+%\end{samepage}
+%\vspace{0.5em}
+%
+Paired differential simulation is based on the standard scientific technique of controlled experimentation. 
+The hypothesis we are testing is that a given synthetic model accurately reproduces the performance metrics exhibited by real network traffic.
+To test this hypothesis, we conduct a series of paired experiments and compare the results for each test subject with a corresponding control subject.
+In this case, the experiments are wireless network simulations, and the subjects are real or synthetic workload instances.
+The control group is the set of simulations where the workload is an actual traffic trace, recorded as described in \Section{trace-data}.
+The test group is a set of simulations where workload is synthetically generated using a simplified traffic model.
+The experiments are paired: for each control simulation, there is a synthetic workload, with behavior as similar to the control as model will allow, against which the results are compared.
+The output of the series of experiments are paired values of performance metrics for each simulated scenario: one from the control, one from the test model.
+
+The remainder of this section proceeds as follows.
+In \Section{traffic-models} we present the collection of traffic models we will evaluate.
+In \Section{trace-data} we describe the trace data that serves as the control for our experiments.
+Our simulation methodology is detailed in \Section{simulations}, and the performance metrics used for statistical comparison of control and test experiments are described in \Section{performance-metrics}.
+Finally, \Section{measures-of-error} is an in-depth mathematical discussion of the appropriate measure of error for the pairs of metric values that are produced by the paired differential simulation methodology.
+
+% to compare traffic produced by synthetic models with known examples of recorded trace traffic in a variety of experimental scenarios.
+To accomplish this, we begin with a large 24-hour example of wireless trace traffic, break it into 144 ten-minute scenarios, and use each scenario's simulated trace behavior as a control for a set synthetic traces, generated by various traffic models.
+For each scenario, the synthetic models are configured to approximate the original trace behavior as closely as the models allow.
+Important network-wide performance metrics are then collected from simulations, and synthetic performance results compared with control results to determine the realism of the synthetic models.
+
+%The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results.
+Clearly, we need not simulate subatomic particle interactions in a wireless network simulator.
+Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same.
+Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior.
+This requirement leads us to the following definition:
+
+%This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required.
+We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.
+
+\subsection{Traffic Models}
+\label{sec:traffic-models}
+
+\begin{table*}
+\begin{center}
+\small
+\begin{tabular}{|c|c|l|}
+\multicolumn{1}{c}{\textbf{model}} &
+\multicolumn{1}{c}{\textbf{transformation}} &
+\multicolumn{1}{c}{\textbf{description}} \\
+\hline
+{\footnotesize{GMM}} & --- & the general matrix model \\\hline\hline
+
+time uniform & $A'=A(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
+node uniform & $S'=S(\ones{n \times n}/n)$~~~~$D'=D(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
+full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above three models \\\hline
+\hline
+time marginal & $A'=(\ones{f \times f}/f)A$ & start time independently chosen from overall distribution of start times \\\hline
+size marginal & $B'=(\ones{f \times f}/f)B$ & flow size (bytes) is chosen independently of other properties \\\hline
+node marginal & $S'=(\ones{f \times f}/f)S$~~~~$D'=(\ones{f \times f}/f)D$ & nodes behave distinctly, but behavior is independent of flows \\\hline
+packet marginal & $Z'=(\ones{f \times f}/f)Z$~~~~$V'=(\ones{f \times f}/f)V$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
+full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline
+\hline
+source regular & $F'=(SS^t/f)F$ & flow behaviors are aggregated on a per source-node basis \\\hline
+dest regular & $F'=(DD^t/f)F$ & flow behaviors are aggregated on a per destination-node basis \\\hline
+mixed regular & $F'=\frac{1}{2f}(SS^t+DD^t)F$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
+
+\end{tabular}
+\caption{Matrix-based traffic models evaluated by paired differential simulation.}
+\label{tab:traffic-models}
+\end{center}
+\end{table*}
+
+The first model we evaluate is the General Matrix Model itself.
+We derive the matrix representation directly from the trace data and use it to generate workload as described in \Section{traffic-generation}.
+This serves to test whether the \caps{GMM} is sufficiently general: if it accurately reproduces all performance metrics across the collection of scenarios simulated, this provides strong evidence that \caps{GMM} captures enough of the detail of trace behavior to be considered a sufficiently general representation of network traffic patterns.
+%An affirmation of the generality of the \caps{GMM} acts as reduction of the
+
+After validating the \caps{GMM}, we explore the effects of various uniformity, marginality, and regularization assumptions on the realism of traffic models.
+The uniform models are the most common, being the simplest both to conceptualize and to implement: simply make no distinction between start times, source or destination nodes.
+Choose them at random, using no information from reality other than the number of nodes and the possible range of start times.
+We do not evaluate uniform packet behavior because it has been sufficiently discredited in our previous work.
+
+In more sophisticated modeling work, where uniformity is discarded as too simple a model for behavior, marginality is often applied instead.
+It is assumed---often implicitly---that marginal models are adequate to capture the essential aspects of network behavior.
+We put that implicit assumption to the test here.
+How well do perfect marginal models represent real behavior?
+When we say that our marginal models are ``perfect,'' we mean that they are nonparametric and calculated directly from trace behavior.
+These are the marginal models that most accurately describe the trace traffic, because they are directly derived from it and make no parametric simplifications.
+This provides an upper bound on the quality of marginal models in general, since parametric marginal models are derived, in turn, as simplifications of these nonparametric marginal behaviors.
+\Table{traffic-models} provides a complete list of all the traffic models we compare in our evaluation.
+
+%The original trace traffic serves as a control group against which to compare all other models.
+Before we can evaluate the General Matrix Model (\caps{GMM}), we must establish that one potentially important aspect of traffic behavior that is \textit{not} represented in the \caps{GMM} can legitimately be discarded.
+That aspect of behavior is \textit{\caps{TCP} feedback}.
+The \caps{GMM} does not specify the \caps{IP} protocol type of each flow.
+Therefore we must establish the extent to which using \caps{UDP} flows, without reliable delivery or congestion control, affects performance metrics.
+%%\caps{GMM} also does not indicate which flows are paired with each other in a client-server relationship. While this pairing does not affect non-\caps{TCP} flows, we implement.
+%Other synthetic models derived from algebraic transformations of the \caps{GMM} are listed in \Table{traffic-models}.
+
+\subsection{Trace Data}\label{sec:trace-data}
+%\subsection{Trace Data \& Simulations}\label{sec:trace-data}\label{sec:simulations}
+
+Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of trace-based models and synthetic traffic models defined using the General Matrix Model.
+We use a 24-hour trace recorded in an infrastructured 802.11g wireless \caps{LAN} with 18 access points, deployed at the 60th Internet Engineering Task Force meeting (\caps{IETF}60), held in San Diego during August of 2004.
+The traffic trace was captured using \texttt{\small{tcpdump}} at a single router, through which all wireless traffic for the meeting was routed, including traffic between wireless nodes.
+The snap length of the capture was 100 bytes, allowing \caps{IP}, \caps{ICMP}, \caps{UDP} and \caps{TCP} headers to be analyzed.
+We limit our work to the 24-hour sub-trace recorded on Wednesday, August 4th.
+This trace contains a broad variety of behaviors and entails a very large volume of traffic: 2.1 million flows, 58 million packets, and 52 billion bytes.
+% 2,085,563 flows, 58,377,994 packets of application data, and 51,988,245,537 bytes of application data.
+
+\begin{figure}
+\vspace{0.45em}
+\begin{center}
+\includegraphics[width=3.5in]{nodes-flows}%
+\vspace{-0.75em}%
+\caption{The number of active nodes and flows over time.} 
+\label{fig:nodes-flows}
+\end{center}
+\vspace{-2em}
+\end{figure}
+
+We do not assume or claim that the traffic found at \caps{IETF}60 is representative of conference settings in general.
+The observed behaviors are also unlikely to resemble those found in a typical commercial or residential setting.
+We have chosen this trace, however, because within it can be found behaviors resembling many different types of wireless usage cases.
+Figure~\ref{fig:nodes-flows} shows the wide variations in the number of active flows and nodes over the course of the trace.
+In the night and morning hours, the traffic patterns are similar to those one might find in a moderately trafficked business or residential area.
+During working group sessions, we see highly concentrated, heavy usage patterns.
+At the zenith of activity, over 800 users, 33 thousand flows, and 1 million packets are seen in a single 10-minute trace segment.
+At the nadir, a lone node sent only a single 61-byte packet over the course of 10 minutes.
+All levels of activity between these extremes are represented.
+Moreover, the mix of traffic types observed changes dramatically over the course of the day, providing a wide representation of possible blends of behavior.
+This heterogeneity and extreme range of behaviors makes the \caps{IETF} data set ideal for this evaluation.
+The variety of activity gives us greater confidence that success or failure of traffic models is not tied to any specific network condition, but is broadly and generally applicable.
+%For a traffic model to be truly realistic, it must be realistic across many different usage cases.
+
+Before using the traces, it is necessary to extract application-level behavior from the trace header data.
+First, we split the trace into individual packet flows.
+A flow is a series of packets sharing the following five attributes: \caps{IP} and transport protocols (raw \caps{IP}, \caps{ICMP}, \caps{TCP}, \caps{UDP}); source and destination \caps{IP} addresses and \caps{TCP}/\caps{UDP} port numbers.
+Next, the quantity of application-initiated data contained in each packet is calculated.
+For non-\caps{TCP} packets, this quantity is simply the size of the transport-layer payload, but for \caps{TCP} the calculation is more complicated: only new data transfers, explicitly initiated by the application are counted.
+Data retransmitted by \caps{TCP} is disregarded, and empty \caps{ACK}s are ignored.
+\caps{SYN} and \caps{FIN} flags in packets (even empty ones) are counted as a single byte each, since they are explicitly signaled by the application.
+%After processing, we have a collection of flows, each with a sequence of timestamps and the amount of application-initiated data sent at that time.
+
+\subsection{Simulations}
+\label{sec:simulations}
 
@@
 
-We have modified the standard Qualnet distribution in the two significant ways.
-First, In our previous work~\cite{Karpinski07:realism,Karpinski07:cbr-failure}, we approximated full-duplex \caps{TCP} with a pair of \caps{UDP} flows.
+In our previous work, we approximated \caps{TCP} with a pair of half-duplex \caps{UDP} flows~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
 For this work, we have implemented a new Qualnet application driver that allows trace-driven full-duplex \caps{TCP} flows.
 This allows us to accurately reproduce the full dynamics of \caps{TCP} feedback.
-Moreover, raw \caps{IP} and \caps{UDP} flows are implemented using the same code, guaranteeing that all traffic is simulated uniformly and that all performance metrics are aggregated in the same manner.
-The second modification is that we have instrumented Qualnet to collect \caps{IP} performance metrics in strict adherence to the recommendations of the \caps{IETF} working group recommendations~\cite{rfc:ip-metrics,rfc:jitter}.
-
-\subsection{Measures of Error}\label{sec:error-measures}
+Moreover, raw \caps{IP}, \caps{UDP} and \caps{TCP} flows are implemented using the same code, guaranteeing that all traffic is simulated uniformly and that all performance metrics are aggregated in the same manner.
+We have also instrumented Qualnet to collect \caps{IP} performance metrics in strict adherence to the recommendations of the \caps{IETF} working group on \caps{IP} performance metrics~\cite{rfc:ip-metrics}.
+
+\subsection{Performance Metrics}
+\label{sec:performance-metrics}
+
+We have selected six performance metrics to present here. They are commonly used as indicators of network performance at the application, network, and link layers of the protocol stack:
+\begin{enumerate}
+\setlength{\itemsep}{0em}
+\item \textbf{Application:} average end-to-end delay, packet delivery ratio, received throughput.
+\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), packet drops in \caps{IP} routing queues.
+\item \textbf{Link:} 802.11 control overhead (\caps{RTS/CTS/ACK}).
+\end{enumerate}
+These metrics are commonly used to evaluate wireless protocols.
+We have examined a broad variety of other wireless network performance metrics, and although we do not have room to present or discuss them here, the results shown are representative of the overall realism of the traffic models.
+%For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
+%Unfortunately, not all of the simulations were able to run to completion.
+In some scenarios, the number of nodes and flows caused Qualnet to simply fail.
+We were, however, able to get complete results for 72.5\% of the simulations, providing a clear view of overall trends.\!\footnote{That 27.5\% of simulations did not finish may seem like a cause for concern, but it does not in fact invalidate our results.
+The simulations that did not finish are those for which the number of flows and nodes overwhelmed the capacity of the simulator.
+Qualnet cannot simulate more than circa 650 nodes sending 35,000 flows.
+While it would be of great interest to examine results for these large-scale scenarios, they are essentially useless for comparison to other simulations studies: other studies do not simulate this quantity of nodes and flows precisely because even the best existing simulators do not support it.
+Once the simulators become more scalable and robust, our analysis can be extended to these scenarios as well.}
+
+\subsection{Measures of Error}
+\label{sec:measures-of-error}
 
@@
 \begin{align}
-E(t^\TTT,t^\M)
-	= E\left({\frac{r^\TTT}{fd},\frac{r^\M}{fd}}\right)
-	= E(r^\TTT,r^\M).
+E(t_{\text{trc}},t_{\text{mod}})
+	= E\left({\frac{r_{\text{trc}}}{fd},\frac{r_{\text{mod}}}{fd}}\right)
+	= E(r_{\text{trc}},r_{\text{mod}}).
 \end{align}
@@
 This property allows us to compare error values meaningfully across different traffic models.
-For example, if flow topology and packet behavior affected some performance metric independently with no interaction effects, we would expect that
-\begin{align}
-E(x^\TTT,x^\UTU) \approx E(x^\TTT,x^\TTU) + E(x^\TTT,x^\UTT).
-\end{align}
-If these two values differ significantly, there must be some interaction between the two levels of behavior that introduces more error than can be explained by each separately.
+%For example, if flow topology and packet behavior affected some performance metric independently with no interaction effects, we would expect that
+%\begin{align}
+%E(x^\TTT,x^\UTU) \approx E(x^\TTT,x^\TTU) + E(x^\TTT,x^\UTT).
+%\end{align}
+%If these two values differ significantly, there must be some interaction between the two levels of behavior that introduces more error than can be explained by each separately.
 Without the property of additivity given in Equation~\ref{eqn:linear-compounding}, such a comparison would not be possible or meaningful.
-
-Additivity of compounded errors also implies two desirable properties that are easily derived from Equation~\ref{eqn:linear-compounding}.
-It forces the error of an accurate representation to be zero: $E(x,x) = 0$.
-It also forces underestimation and overestimation to be treated symmetrically.
-The error of underestimating by some factor is opposite but equal to overestimating by the same factor:
-\begin{align}
-E(x,x/c) = - E(x,xc).
-\end{align}
-
-It is easily verified that the difference, ratio, and standard error measures do not satisfy Equation~\ref{eqn:linear-compounding}, and the difference, as noted, does not satisfy Equation~\ref{eqn:normalization-invariance}.
-The log-ratio is the only metric presented that satisfies both conditions.
-Moreover, it can be proved that $\log(y/x)$ is the \textit{only} differentiable function that satisfies both (up to a constant).
-In the Appendix, we present a proof of this claim.
-Throughout the rest of the paper, we use the log-ratio to measure the error of performance metrics.
-% TODO: should Appendix be capitalized?
-% TODO: hyperref the Appendix for PDF.
-
-%\subsection{Tests of Statistical Equivalence}\label{sec:statistical-equivalence}
-
-%%\begin{sidewaysfigure*}
-%%%\begin{centering}
-%%\noindent
-%%\subfloat[Application: Average Jitter]
-%%{\includegraphics[width=4.6in]{Application-Average-Jitter}
-%%\label{fig:Application-Average-Jitter}}
-%%\subfloat[Application: Average Delay]
-%%{\includegraphics[width=4.6in]{Application-Average-Delay}
-%%\label{fig:Application-Average-Delay}}
-%%\subfloat[Application: Received Throughput]
-%%{\includegraphics[width=4.6in]{Application-Received-Throughput}
-%% \label{fig:Application-Received-Throughput}}
-%%\subfloat[Network: Control Overhead]
-%%{\includegraphics[width=4.6in]{Network-Control-Overhead}
-%% \label{fig:Network-Control-Overhead}}
-%%\subfloat[Network: Packet Drop Rate]
-%%{\includegraphics[width=4.6in]{Network-Packet-Drop-Rate}
-%% \label{fig:Network-Packet-Drop-Rate}}
-%%\subfloat[Network: \caps{RREQ}s Initiated]
-%%{\includegraphics[width=4.6in]{Network-RREQs-Initiated}
-%% \label{fig:Network-RREQs-Initiated}}
-%%\subfloat[Link (MAC): Control Overhead]
-%%{\includegraphics[width=4.6in]{MAC-Control-Overhead}
-%% \label{fig:MAC-Control-Overhead}}
-%%\subfloat[Link (MAC): Retransmissions]
-%%{\includegraphics[width=4.6in]{MAC-Retransmissions}
-%% \label{fig:MAC-Retransmissions}}
-%%\subfloat[Link (\caps{MAC}): Retransmission Failures]
-%%{\includegraphics[width=4.6in]{MAC-Retransmission-Failures}
-%% \label{fig:MAC-Retransmission-Failures}}
-%%\hspace{0.035in}
-%%\begin{minipage}[t]{4.545in}
-%%\vspace{-2em}
-%%\caption{Box-and-whisker plots of log-ratio error values for all metrics and traffic models.
-The lower axis indicates the log-ratio, while the upper axis shows raw ratio values.
-Each box contains the central majority of log-ratio values: the left and right bounds are at the $25^\text{th}$ and $75^\text{th}$ percentiles.
-The dark middle line indicates the median value, while the diamond marks the mean.
-The whiskers (dotted lines) extend to the furthest non-outlier values, while the points beyond that are outliers.
-The notches in the middle show a $95\%$ confidence interval for the actual underlying median value; if two notches do not overlap, they are very unlikely to have the same median.
-Numbers to the right are $p$-values for the median, mean, and 3-mean tests, respectively; $p$-values above $\alpha=0.05$ appear in bold.}
-%%\label{fig:boxplots}
-%%\end{minipage}
-%%%\end{centering}
-%%\end{sidewaysfigure*}
-
-%In this section, we consider the values of performance metrics as random variables, drawn from unknown distributions.
-We present three tests for the statistical equivalence of the metric values induced by synthetic and real traffic.
-Let $\M$ be a traffic model as before and let $\X$ be a performance metric.
-Let $X_k^\M$ be a random variable representing the value of $\X$ in the $k^{\text{th}}$ scenario using the traffic model $\M$.
-If the distribution of $X_k^\M$ is the same as that of $X_k^\TTT$, then both the median and mean values of the log-ratio $R_k^\M=\log(X_k^\M/X_k^\TTT)$ should be zero.
-The first two tests check the plausibility of precisely these hypotheses.
-The third test separates small, medium, and large simulation scenarios, and test their means separately to catch any size-dependent performance bias.
-
-%\textbf{The Median Test.} If $\M$ induces realistic performance, then the median of each log-ratio variable, $R_k^\M=\log(X_k^\M/X_k^\TTT)$, should be zero.
-The $k^\text{th}$ indicator variable is defined as
-%\begin{align}
-%	I_k^\M = \begin{cases}
-%		0 & \text{if $R_k^\M < 0$,} \\
-%		1 & \text{if $R_k^\M \ge 0$.}
-%	\end{cases}
-%\end{align}
-%If the median value of $R_k^\M$ is truly zero, then $I_k^\M$ is a Bernoulli variable with probability parameter $p=\frac{1}{2}$.
-The variables $I_k^\M$ are all independent since they come from separate simulations, and cannot affect each other's outcomes.
-Therefore, the sum $S_n^\M = \sum_{k=1}^n {I_k^\M}$ should follow a binomial distribution of $n$ trials with $p=\frac{1}{2}$.
-The median test applies the exact binomial cumulative distribution function (\caps{CDF}) % TODO: cite this.
-% for $n$ and $p$ to the observed value of $S_n^\M$, yielding a $p$-value: the probability that such an extreme value would occur by chance under the hypothesis that the median of each $R_k^\M$ is zero.
-% TODO: fix the wording here.
-
-%%\begin{figure*}
-%%\begin{centering}
-%%\subfloat[App. Average Jitter: {\footnotesize{STU}}]
-%%{\includegraphics[width=1.75in]{application-average-jitter-stu}%
-%%\label{fig:application-average-jitter-stu}}
-%%\subfloat[Net. Packet Drop Rate: {\footnotesize{UTU}}]
-%%{\includegraphics[width=1.75in]{network-packet-drop-rate-utu}%
-%%\label{fig:network-packet-drop-rate-utu}}
-%%\subfloat[Net. Packet Drop Rate: {\footnotesize{STU}}]
-%%{\includegraphics[width=1.75in]{network-packet-drop-rate-stu}%
-%%\label{fig:network-packet-drop-rate-stu}}
-%%\subfloat[Link Retrans. Failures: {\footnotesize{SUU}}]
-%%{\includegraphics[width=1.75in]{mac-retransmission-failures-suu}%
-%%\label{fig:mac-retransmission-failures-suu}}
-%%\caption{Scatter plots of various metrics for an alternate traffic model ({\footnotesize{STU}}, {\footnotesize{UTU}}, {\footnotesize{SUU}}) versus real trace traffic~({\footnotesize{TTT}}).
-Each data point represents a pair of matched simulations: the $x$-coordinate is the metric value for trace traffic, the $y$-coordinate is the metric value for the synthetic model.
-Scenarios are plotted according to the amount of flows in their trace segment: the bottom third with filled dots; the middle third with circles; the top third with squares.}
-%%\label{fig:scatterplots}
-%%\end{centering}
-%%\vspace{-1em}
-%%\end{figure*}
-
-%\textbf{The Mean Test.} We use Lyapunov's generalization of the Central Limit Theorem (\caps{LCLT}) to test the hypothesis that the mean of each $R_k^\M$ is zero.
-We present the theorem and its application to the series $R_k^\M$ in the Appendix.
-Here we simply present the resulting test statistic and its usage.
-If the mean of each $R_k^\M$ is zero, then \caps{LCLT} implies that the test statistic
-%\begin{align}
-%\hat{Z}_n^\M &= \frac{\sum_{k=1}^n{R_k^\M}}{\sqrt{\sum_{k=1}^n{(R_k^\M)^2}}}
-%\end{align}
-%converges to a standard normal distribution for large values of $n$, where $n$ is the number of simulated scenarios.
-In this case, $n=144$, which is fairly large by traditional statistical standards.
-The $p$-value of the mean test is given by applying the standard normal \caps{CDF} to the test statistic, $\hat{Z}_n^\M$.
-
-%\textbf{The 3-Mean Test.} Performance behavior in scenarios with a large number of nodes or flows is often very different from behavior in small scenarios.
-In some cases, $R_k^\M$ is skewed positively for one group, but negatively for another.
-In such cases, $R_k^\M$ can pass the median and mean tests even though behavior in each case is unrealistic.
-To catch such situations, we split the simulations into three groups by the number of flows: the lower, middle, and upper thirds.
-The 3-mean test simply applies the mean test to each of these groups separately and uses the minimum $p$-value of the three.
-This reduces the power of the test, since each group is smaller, but can catch cases where the error has size-dependent biases that cancel out on average.
-
-\section{Results}\label{sec:results}
 
@@
 {\includegraphics[width=4.6in]{plots/Application/delay.pdf}}
+\subfloat[\textbf{Application:} Packet Delivery Ratio]%
+{\includegraphics[width=4.6in]{plots/Application/delivery_ratio.pdf}}
 \subfloat[\textbf{Application:} Average Received Throughput]%
 {\includegraphics[width=4.6in]{plots/Application/throughput.pdf}}
-\subfloat[\textbf{Application:} Total Received Packets]%
-{\includegraphics[width=4.6in]{plots/Application/total_recv_packets.pdf}}
 \subfloat[\textbf{Network:} AODV Control Overhead]%
@@
 
-Our simulation results are summarized in Figure~\ref{fig:box-plots}.
+Additivity of compounded errors also implies two desirable properties that are easily derived from Equation~\ref{eqn:linear-compounding}.
+It forces the error of an accurate representation to be zero: $E(x,x) = 0$.
+It also forces underestimation and overestimation to be treated symmetrically.
+The error of underestimating by some factor is opposite but equal to overestimating by the same factor:
+\begin{align}
+E(x,x/c) = - E(x,xc).
+\end{align}
+
+It is easily verified that the difference, ratio, and standard error measures do not satisfy \Equation{linear-compounding}, and the difference, as noted, does not satisfy \Equation{normalization-invariance}.
+The log-ratio is the only metric presented that satisfies both conditions.
+Moreover, it can be proved that $\log(y/x)$ is the \textit{only} differentiable function that satisfies both (up to a constant).
+In the Appendix, we present a proof of this claim.
+Throughout the rest of the paper, we use the log-ratio to measure the error of performance metrics.
+% TODO: should Appendix be capitalized?
+% TODO: hyperref the Appendix for PDF.
+
+\section{Results}\label{sec:results}
+
+Our simulation results are summarized in \Figure{box-plots}.
 Each subfigure shows a single performance metric.
-The distribution of errors for each traffic model is visualized with a box-and-whisker plot.
-These plots allow immediate assessment of realism: a good traffic model should have log-ratio values that are tightly clustered around the center, with a small, evenly balanced box.
+The distribution of log-ratio error values for each traffic model is visualized with a box-and-whisker plot.
+The box indicates the range from the 25th to 75th percentiles of values, while the ``whiskers'' indicate the full range, excluding outliners (which are shown as isolated points).
+These plots allow immediate assessment of realism: a good traffic model should have error values that are tightly clustered around the center, with a small, evenly balanced box, and relatively small whiskers.
 Additionally, the mean and median markers should be close to the center.
-
-The primary result is that the General Matrix Model~(\caps{GMM}) is sufficiently realistic.
-It accurately reproduces important performance metrics at all levels of the protocol stack in simulation.
+It should be noted that the error scale is logarithmic, so even a slight increase in spread or deviation from the center indicates a disproportionately large decrease in model quality.
+Non-logarithmic scale markers are shown above each plot; the values indicate the factor of over- or under-representation for the performance metric.
+
+The primary result is that the \caps{GMM} accurately reflects the performance of the trace data it is based on.
+It has small, centered error bars for every performance metric presented here, as well as for those we have looked at otherwise.
+This is an essential result, since otherwise, \caps{GMM} is at best a shaky foundation for further modeling work.
+With these results, we can be assured that if we can approximate the matrix for a given example of trace traffic, then we have also approximated the original trace behavior.
+In a sense, this result is a reduction of the general traffic modeling problem to a better-understood matrix modeling problem.
+
+It is worth noting that the \caps{GMM} has competition from a few of the time-simplified models: the \model{time marginal} and \model{time uniform} model both do approximately as well, and in some cases better, for every metric examined.
+{\FHC} found strong evidence that session arrivals followed a time-varying Poisson arrival process, meaning that other than long-scale time-varying arrival rate and very small-scale clustering of flows within sessions, the overall flow arrival rate is fairly smooth.
+We suspect that on the scale of minutes to tens of minutes in which our simulation exist, it is sufficient to model both flow and session arrivals uniformly.
+This result indicates that the precise temporal placement of flows is not highly sensitive: they tend to be relatively evenly spaced out at short time-scales, and network performance is not sensitive to changes in start-time on that scale.
+
+%We also believe that the fact that both the \model{time marginal} and \model{time uniform} models perform about as well as the \caps{GMM} indicates not that they are performing exceptionally well, but that the \caps{GMM} is performing worse than it should in this case.
+This may be an artifact of the interaction between packet behavior, flow size, flow duration and start time, as described in \Section{traffic-generation}.
+This remark may appear to be a non-sequitur, but these properties are all closely related.
+Consider that packet behavior (i.e.
+distribution of packet sizes and inter-packet intervals) determines the expected mean packet size and expected mean interval duration.
+The expected duration is then related to the flow size via the following equation:
+%\begin{align*}
+%\E{\text{duration}} = [\text{flow size}]\frac{\E{\text{mean interval}}}{\E{\text{mean size}}}
+%\end{align*}
+%The duration must also satisfy $t_{\text{start}}+[\text{duration}] \le t_{\max}$.
+
+%It accurately reproduces important performance metrics at all levels of the protocol stack in simulation.
 This only moves us a step closer to being able to generate realistic, completely synthetic workloads, since the \caps{GMM} in these comparisons  derives almost all of its behavior directly from the trace.
@@
 \section{Discussion}\label{sec:discussion}
+
+The story of network analysis and modeling is one of variety. Assumptions of ``sameness'' have fallen repeatedly. ...
+The matrix model gives new insight into
 

### 2008-02-08 paper.tex
Finalized through the common representations section.

@@
 
-\subsection{Traffic Generation}
-\label{sec:traffic-generation}
-
-Given an instance of the general matrix model, how do we produce an actual workload from it?
-Each row of the model is used to randomly generate the behavior of a single flow.
-The type, source, and destination nodes of the flow are randomly chosen, using histogram vectors $\vec{t}$, $\vec{s}$, and $\vec{d}$ as the proportional probabilities of each type or node being chosen.\footnote{Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
-The simplest technique is the rejection method: if the same node is chosen for both source and destination, reject that pair and begin the selection process again.}
-Next we use the flow size vector, $\vec{b}$ to randomly sample the number of bytes in the flow.
-The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\dot{z}$ and $\dot{v}$.
-From these, together with the flow size in bytes, we estimate the expected duration of the flow: $\dot{d} = \fracx{b\dot{v}}{\dot{z}}$.
-We finally choose a start time for the flow, using $\vec{a}$ as a vector of relative probabilities for each bin of start times, but ensuring that the start time plus the expected duration, $\dot{d}$, does not exceed the maximum simulation time for the workflow.
-From the range of possible values mapping to the chosen start bin, we choose a random time uniformly.
-
-Having determined the overall parameters of the flow, we generate the actual packets of the flow.
-The first packet is sent at the chosen start time; the size of each packet is sampled from the histogram of packet sizes, $\vec{z}$; the interval until the next packet is determined from the interval vector, $\vec{v}$.
-Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}(k)$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\seq{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
-\begin{align}
-\label{eqn:interval-dequantization}
-\Di(k,u)=v_{\min}+(v_{\max}-v_{\min})\left(\frac{k-u}{d_v}\right)^{\frac{1}{\gamma}}.
-\end{align}
-The packet stream ends when the number of bytes allocated to the flow have been emitted. %The last packet is only the size of the remaining number of bytes.
-
-\section{Expression of Common Properties}
-
-Having defined our representation of network behavior, we turn to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.
-
-\subsection{Communication Topology}
-\label{sec:communication-topology}
-
-The number of flows between each pair of nodes in the network is expressed by the matrix product $SD^t\in\R^{n \times n}$ ($*^t$~denotes transpose).
-That is $(SD^t)_{ij}$ is the number of flows from node $i$ to node $j$.
-The total number of source flows for each node is given by the vector $S^t\ones{f}$, and the total number of destination flows is given by $D^t\ones{f}$.
-($\ones{f}$ is the $f$-dimensional column vector consisting of all ones.)
-
-\subsection{Expected Average Packet Size \& Intervals}
-\label{sec:flow-totals}
-
-From the matrices $Z$ and $V$, we can compute the expected average packet size, $\dot{z}$, and average inter-packet interval, $\dot{v}$, for each flow.
-The computations require knowing the values that each matrix column represents.
-The computed duration is only approximate because information is irrecoverably lost in the quantization process (see \Equation{interval-quantization}).
-The vectors of representative values for $Z$ and $V$ are:
-\begin{align}
-\vec{\sigma}&=\seq{1,2,3,\dots,d_z},\\
-\vec{\tau}&=\seq{\Di\left(k,\fracx{1}{2}\right)}_{k=1}^{d_v}.
-\end{align}
-Using these representative values, the expected average packet size for a flow may be computed as $\dot{z}=\fracx{\vec{z}\cdot\vec{\sigma}}{\vec{z}\cdot\ones{d_z}}$, and the expected average inter-packet interval size as $\dot{v}=\fracx{\vec{v}\cdot\vec{\tau}}{\vec{v}\cdot\ones{d_v}}$.
-
-\section{Expression of Common Modeling Concepts}
-
-The simplifying assumptions made by most traffic models can be classified into three categories.
-We use the terms \textit{uniformity}, \textit{marginality}, and \textit{regularity} do describe these classes of assumptions.
-They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether the make stipulations about uniformity within each flow, across all flows, or within groups of flows.
-
-\subsection{Uniform Behavior}
-
-The concept of uniformity, as used here, entails that certain properties of each flow are assumed to be statistically or deterministically uniform.
-Examples include: assuming that each flow's packets have the same size and inter-packet interval (i.e.
-the \caps{CBR} flow model); assuming that each node is equally likely to be the source of a given flow; assuming that all start times for a flow are equally likely.
-
-Since uniform behaviors entail homogeneity of properties with respect to each single flow individually, a trace matrix can be transformed to the closest matrix satisfying the model's assumptions via right matrix multiplication.
-Right multiplication mixes the elements within each row, and if the mixing weights are all equal, the results is homogenized behavior.
-For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $S$ and $D$:
-\begin{align*}
-S' &= S\ones{n \times n}/n, & D' &= D\ones{n \times n}/n.
-\end{align*}
-
-%We use the term \textit{uniformity} do describe assumptions made about the ``sameness'' of various properties of flows.
-The constant bit-rate (\caps{CBR}) model, for example, is an assumption about the uniformity of all of a flow's packets: i.e.
-that they have the same size and are evenly spaced in time.
-While this is the strictest sense of the \caps{CBR} model, it is rare for this assumption to be made in isolation.
-Another common simplification is to make the sizes of flows identical, or chose them randomly from a uniform distribution; similarly for the starting time of each flow.
-Finally, when choosing endpoints for a given flow, lacking a better model, it is not unusual to treat all nodes identically, making each an equally likely source or destination for a given flow.
-
-It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees.
-This is largely because of the high complexity of doing otherwise and the lack of compelling alternatives.
-Here we examine how various notions of uniformity are expressed in the matrix model.
-In general, \textit{intra-flow uniformity} is effected by right multiplication, which mixes the elements in each row, whereas \textit{inter-flow uniformity} is effected by left multiplication, which mixes elements across rows.
-The constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
-The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma\in\R^{f \times 3}$, where
-\begin{align}
-\Gamma=\begin{pmatrix}
-\ones{} & \vec{\sigma} & \zeros{} \\
-\zeros{} & \zeros{} & \vec{\tau}
-\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
-\end{align}
-
-While constant bit-rate in the strict sense implies only that the packet behavior in each flow is uniform, it is very rare for it to be applied in this strict sense only: it is typical that ``\caps{CBR}'' models make assumptions about uniformity across flows, and not just within each flow.
-The most drastic simplification is to treat all flows in a synthetic workload identically.
-This is effected by left multiplication: $F'=(\ones{f \times f}/f)F$.
-To make flow behavior uniform on a per-source-node basis, we left-multiply the ``source-uniformizing'' matrix: $F'=(SS^t/f)F$.
-Per-destination-node uniformity can be achieved similarly, with left-multiplication by a corresponding destination-uniformizing matrix: $F'=(DD^t/f)F$.
-
-Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified.
-The confusion arises due to the lack of a unique way of associating flows with nodes.
-Each node can equally viably be associated with its source or its destination.
-Which association is more valid?
-Consider a web request: should the behavior of the request be determined by what the client typically does?
-Or should the requests that the server typically receives determine behavior?
-There is no armchair answer: only analysis of real data can give us the answer.
-The matrix model provides an analytical framework for doing so.
-
-The matrix model has the benefit of making the uniformization of behaviors completely explicit.
-It also allows us to easily define intermediate choices.
-If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
-\begin{align}
-F' = \frac{1}{f}((1-\alpha)(SS^t)+\alpha(DD^t))F.
-\end{align}
-The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors.
-Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.
-
-\subsection{Marginal Behavior}
-
-Marginality is an orthogonal concept to uniformity: it implies that the distribution of some property is identical \textit{across} all flows.
-For example, if we determine a distribution of flow sizes across the entire network and assume that this distribution applies \textit{independent of other flow properties}, this is a marginal model for flow size.
-The traffic models proposed by {\FHC}~\cite{Hernandez06:wlan-traffic} are all marginal: they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node.
-These models do not account for non-independence between these and other flow properties.
-We will discuss this limitation and its implications when we analyze our experimental results.
-
-\subsubsection{Regularized Behavior}
-
-The final form of sameness we consider, is regularization.
-This is a restricted form of marginality: we may assume that marginal distributions apply to all the flows in certain groupings.
-For example, we can provide a distribution of flow sizes on a per source-node basis; that is, each source node can have it's \textit{own} distribution, and the sizes of its flows are drawn from that distribution, independent of their other properties.
-
-%We use the terms here to distinguish three fundamentally different types of assumptions about behavioral homogeneity.
-
-%Existing traffic models are typically not explicit methods of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
-In our algebraic modeling framework, such models can be interpreted as stipulations about the form of the behavior matrices, or as transforms that implement those stipulations given a real set of traffic matrices.
-
-%A generic traffic model is a way of generating the six matrices described in \Section{representations}.
-A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or regularized form.
-For experimental comparison, the dimensions of the matrices should all be preserved.
-If a trace-derived model produces experimental performance metrics that are similar to those produced by the original trace data, then simplifications made by the model are viable; if the performance is distorted, then the simplifications are discarding essential information about the traffic pattern.
-
-\subsection{Uniformity}
-
 \begin{figure*}[tb]
@@
 \end{center}
-\vspace{-2em}
+\vspace{-1.25em}
 \end{figure*}
 
-% TODO: differentiate between restrictive versions of models and transformative versions.
-% TODO: distinguish uniform, marginal and regularized models.
-
-\subsection{Marginality}
-
-Most approaches to traffic modeling that go beyond uniformity assumptions, instead describe marginal properties of various aspects of traffic behavior.
+\subsection{Traffic Generation}
+\label{sec:traffic-generation}
+
+Given an instance of the general matrix model, how do we produce an actual workload from it?
+Each row of the model is used to randomly generate the behavior of a single flow.
+The type, source, and destination nodes of the flow are randomly chosen, using histogram vectors $\vec{t}$, $\vec{s}$, and $\vec{d}$ as the proportional probabilities of each type or node being chosen.\footnote{Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
+The simplest technique is the rejection method: if the same node is chosen for both source and destination, reject that pair and begin the selection process again.}
+Next we use the flow size vector, $\vec{b}$ to randomly sample the number of bytes in the flow.
+The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\dot{z}$ and $\dot{v}$.
+From these, together with the flow size in bytes, we estimate the expected duration of the flow: $\dot{d} = \fracx{b\dot{v}}{\dot{z}}$.
+We finally choose a start time for the flow, using $\vec{a}$ as a vector of relative probabilities for each bin of start times, but ensuring that the start time plus the expected duration, $\dot{d}$, does not exceed the maximum simulation time for the workflow.
+From the range of possible values mapping to the chosen start bin, we choose a random time uniformly.
+
+Having determined the overall parameters of the flow, we generate the actual packets of the flow.
+The first packet is sent at the chosen start time; the size of each packet is sampled from the histogram of packet sizes, $\vec{z}$; the interval until the next packet is determined from the interval vector, $\vec{v}$.
+Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}(k)$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\seq{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
+\begin{align}
+\label{eqn:interval-dequantization}
+\Di(k,u)=v_{\min}+(v_{\max}-v_{\min})\left(\frac{k-u}{d_v}\right)^{\frac{1}{\gamma}}.
+\end{align}
+The packet stream ends when the number of bytes allocated to the flow have been emitted. %The last packet is only the size of the remaining number of bytes.
+
+\section{Expression of Common Properties}
+
+Having defined our representation of network behavior, we turn to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.
+
+\subsection{Communication Topology}
+\label{sec:communication-topology}
+
+The number of flows between each pair of nodes in the network is expressed by the matrix product $SD^t\in\R^{n \times n}$ ($*^t$~denotes transpose).
+That is $(SD^t)_{ij}$ is the number of flows from node $i$ to node $j$.
+The total number of source flows for each node is given by the vector $S^t\ones{f}$, and the total number of destination flows is given by $D^t\ones{f}$.
+($\ones{f}$ is the $f$-dimensional column vector consisting of all ones.)
+
+\subsection{Expected Average Packet Size \& Intervals}
+\label{sec:flow-totals}
+
+From the matrices $Z$ and $V$, we can compute the expected average packet size, $\dot{z}$, and average inter-packet interval, $\dot{v}$, for each flow.
+The computations require knowing the values that each matrix column represents.
+The computed duration is only approximate because information is irrecoverably lost in the quantization process (see \Equation{interval-quantization}).
+The vectors of representative values for $Z$ and $V$ are:
+\begin{align}
+\vec{\sigma}&=\seq{1,2,3,\dots,d_z},\\
+\vec{\tau}&=\seq{\Di\left(k,\fracx{1}{2}\right)}_{k=1}^{d_v}.
+\end{align}
+Using these representative values, the expected average packet size for a flow may be computed as $\dot{z}=\fracx{\vec{z}\cdot\vec{\sigma}}{\vec{z}\cdot\ones{d_z}}$, and the expected average inter-packet interval size as $\dot{v}=\fracx{\vec{v}\cdot\vec{\tau}}{\vec{v}\cdot\ones{d_v}}$.
+
+\section{Expression of Common Modeling Concepts}
+
+The simplifying assumptions made by most traffic models can be classified into three categories.
+We use the terms \textit{uniformity}, \textit{marginality}, and \textit{regularity} do describe these classes of assumptions.
+They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether the make stipulations about uniformity within each flow, across all flows, or within groups of flows.
+In the following sections, we explore how various modeling concepts can be expressed in the framework of the General Matrix Model.
+Models are expressed as transformations that will convert arbitrary traffic matrix instances into similar ones that satisfy the model's assumptions.
+These transformations will later allow us to evaluate the effect of each model's simplifications on the realism of generated workloads.
+
+\subsection{Uniform Behavior}
+
+The concept of uniformity, as used here, entails that certain properties of each flow are assumed to be statistically or deterministically uniform.
+Examples include: assuming that each flow's packets have the same size and inter-packet interval (i.e.
+the \caps{CBR} flow model); assuming that each node is equally likely to be the source of a given flow; assuming that all start times for a flow are equally likely.
+
+Since uniform behaviors entail homogeneity of properties with respect to each flow individually, a trace matrix can be transformed to a matrix satisfying the model's assumptions via right matrix multiplication.
+Right multiplication mixes the elements within each row, and if the mixing weights are uniform, the result is uniform behavior.
+For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $S$ and $D$:
+\begin{align*}
+S' &= S\ones{n \times n}/n, & D' &= D\ones{n \times n}/n.
+\end{align*}
+This makes all the values in each flow's row identically the average of the original $n$ values.
+As another example, the constant bit-rate (\caps{CBR}) flow model, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
+The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
+The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma\in\R^{f \times 3}$, where
+\begin{align}
+\Gamma=\begin{pmatrix}
+\ones{} & \vec{\sigma} & \zeros{} \\
+\zeros{} & \zeros{} & \vec{\tau}
+\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
+\end{align}
+\Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to actual behaviors from traces.
+%By replacing $S'$ with $S$, this can also be viewed as a set of linear constraints on the matrix $S$, providing a proscriptive definition
+
+%We use the term \textit{uniformity} do describe assumptions made about the ``sameness'' of various properties of flows.
+The constant bit-rate (\caps{CBR}) model, for example, is an assumption about the uniformity of all of a flow's packets: i.e.
+that they have the same size and are evenly spaced in time.
+While this is the strictest sense of the \caps{CBR} model, it is rare for this assumption to be made in isolation.
+Another common simplification is to make the sizes of flows identical, or chose them randomly from a uniform distribution; similarly for the starting time of each flow.
+Finally, when choosing endpoints for a given flow, lacking a better model, it is not unusual to treat all nodes identically, making each an equally likely source or destination for a given flow.
+
+%It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees.
+This is largely because of the high complexity of doing otherwise and the lack of compelling alternatives.
+Here we examine how various notions of uniformity are expressed in the matrix model.
+In general, \textit{intra-flow uniformity} is effected by right multiplication, which mixes the elements in each row, whereas \textit{inter-flow uniformity} is effected by left multiplication, which mixes elements across rows.
+The constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
+The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
+The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma\in\R^{f \times 3}$, where
+%\begin{align}
+%\Gamma=\begin{pmatrix}
+%\ones{} & \vec{\sigma} & \zeros{} \\
+%\zeros{} & \zeros{} & \vec{\tau}
+%\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
+%\end{align}
+
+\subsection{Marginal Behavior}
+
+Marginality is an orthogonal concept to uniformity: the distribution of some property is identical \textit{across} all flows.
+For example, if we determine a distribution of flow sizes across the entire network and assume that this distribution applies \textit{independent of other flow properties}, this is a marginal model for flow size.
+Marginal behavior is effected by left multiplication by a matrix with uniform columns.
+This specific example is generated by left-uniformizing the flow size matrix: $B' = (\ones{f \times f}/f)B$.
+The traffic models proposed by {\FHC} are all marginal: they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node.
+These models do not account for non-independence between these and other flow properties.
+We will discuss this limitation and its implications when we analyze our experimental results.
+
+\subsection{Regularized Behavior}
+
+The final form of homogenization we consider, is regularization.
+This is a restricted form of marginality: we may assume that marginal distributions apply to all the flows in certain groupings.
+For example, we can provide a distribution of flow sizes on a per-source basis.
+That is, each source node can have its \textit{own} distribution, and the sizes of its flows are drawn from that distribution, independent of their other properties.
+This is simply accomplished by left-multiplication by a non-uniform matrix.
+For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-uniformizing'' matrix: $F'=(SS^t/f)F$.
+Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-uniformizing matrix: $F'=(DD^t/f)F$.
+%Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified.
+The confusion arises due to the lack of a unique way of associating flows with nodes.
+Each node can equally viably be associated with its source or its destination.
+Which association is more valid?
+Consider a web request: should the behavior of the request be determined by what the client typically does?
+Or should the requests that the server typically receives determine behavior?
+There is no armchair answer: only analysis of real data can give us the answer.
+The matrix model provides an analytical framework for doing so.
+The matrix model has the benefit of making the uniformization of behaviors completely explicit.
+It also allows us to easily define intermediate choices.
+If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
+\begin{align}
+F' = \frac{1}{f}((1-\alpha)(SS^t)+\alpha(DD^t))F.
+\end{align}
+The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors.
+Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.
+
+%We use the terms here to distinguish three fundamentally different types of assumptions about behavioral homogeneity.
+
+%Existing traffic models are typically not explicit methods of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
+In our algebraic modeling framework, such models can be interpreted as stipulations about the form of the behavior matrices, or as transforms that implement those stipulations given a real set of traffic matrices.
+
+%A generic traffic model is a way of generating the six matrices described in \Section{representations}.
+A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or regularized form.
+For experimental comparison, the dimensions of the matrices should all be preserved.
+If a trace-derived model produces experimental performance metrics that are similar to those produced by the original trace data, then simplifications made by the model are viable; if the performance is distorted, then the simplifications are discarding essential information about the traffic pattern.
+
+%\subsection{Uniformity}
+
+%% TODO: differentiate between restrictive versions of models and transformative versions.
+%% TODO: distinguish uniform, marginal and regularized models.
+
+%\subsection{Marginality}
+
+%Most approaches to traffic modeling that go beyond uniformity assumptions, instead describe marginal properties of various aspects of traffic behavior.
 For example, {\FHC}~\cite{Hernandez06:wlan-traffic} provide highly convincing parametric models for six marginal properties of wireless network traffic: session (node) arrivals follow a time-varying Poisson process; per-session flow inter-arrival times are Lognormally distributed; flows per session (node) follow a BiPareto distribution; flow sizes follow a BiPareto distribution.
@@
 
-\section{Experimental Methodology}\label{sec:methodology}
+\section{Experimental Methodology}
+\label{sec:methodology}
 
@@
 {\footnotesize{GMM}} & --- & the general matrix model \\\hline\hline
-
 time uniform & $A'=A(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
 node uniform & $S'=S(\ones{n \times n}/n)$~~~~$D'=D(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
-full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above three models \\\hline
-\hline
+full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above three models \\\hline\hline
 time marginal & $A'=(\ones{f \times f}/f)A$ & start time independently chosen from overall distribution of start times \\\hline
@@
 packet marginal & $Z'=(\ones{f \times f}/f)Z$~~~~$V'=(\ones{f \times f}/f)V$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
-full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline
-\hline
+full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline\hline
 source regular & $F'=(SS^t/f)F$ & flow behaviors are aggregated on a per source-node basis \\\hline
@@
 mixed regular & $F'=\frac{1}{2f}(SS^t+DD^t)F$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
-
 \end{tabular}
@@
 \end{center}
+\vspace{-2em}
 \end{table*}
@@
 
-\begin{figure}
-\vspace{0.45em}
+\begin{figure}[b]
+\vspace{-1em}
 \begin{center}
@@
 \end{center}
-\vspace{-2em}
+\vspace{-0.8em}
 \end{figure}

### 2008-02-08 paper.tex
Work on results. Almost done...

@@
 This technique uses real wireless trace traffic as an experimental control and conducts paired experiments comparing network performance from simulations using control traffic against others using synthetically generated traffic.
-The results are positive: \caps{GMM} accurately reproduces network performance characteristics at all levels of the protocol stack.
+The results are very positive: \caps{GMM} accurately reproduces network performance characteristics at all levels of the protocol stack.
 %This does not solve the ``cookbook'' problem, however, because the \caps{GMM} must still be based on real traffic in order to reproduce its behavioral characteristics.
@@
 
-Our experiments with various common assumptions of uniformity and regularity in network behavior yield mixed conclusions: most straightforward assumptions of homogenous behavior distort performance results significantly, but there are notable exceptions.
+%Our experiments with various common assumptions of uniformity and regularity in network behavior yield mixed conclusions: most straightforward assumptions of homogenous behavior distort performance results significantly, but there are notable exceptions.
 The clearest exception is the time domain: traffic models appear to be relatively insensitive to local perturbation of the start time of flows---so long as the overall distribution remains approximately correct.
@@
 The General Matrix Model is presented in \Section{general-matrix-model}.
+In Sections~\ref{sec:common-properties}~and~\ref{sec:modeling-concepts} we explore how common network properties and modeling concepts can be expressed in the the \caps{GMM} framework.
 Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
@@
 
-\section{Motivation \& Related Work}\label{sec:motivation}\label{sec:related-work}
+\section{Motivation \& Related Work}
+\label{sec:related-work}
+\label{sec:motivation}
 
@@
 Finally, the parameter $\gamma\in(0,1)$ allows us to fine-tune how the resolution scales with interval size; we have chosen $\gamma=\frac{1}{3}$ because this choice provides resolution that scales smoothly from the microseconds for small intervals to seconds for large intervals.
+
+\subsection{Traffic Generation}
+\label{sec:traffic-generation}
+
+Given an instance of the general matrix model, how do we produce an actual workload from it?
+Each row of the model is used to randomly generate the behavior of a single flow.
+The type, source, and destination nodes of the flow are randomly chosen, using histogram vectors $\vec{t}$, $\vec{s}$, and $\vec{d}$ as the proportional probabilities of each type or node being chosen.\footnote{Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
+The simplest technique is the rejection method: if the same node is chosen for both source and destination, reject that pair and begin the selection process again.}
+Next we use the flow size vector, $\vec{b}$ to randomly sample the number of bytes in the flow.
+The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\dot{z}$ and $\dot{v}$.
+From these, together with the flow size in bytes, we estimate the expected duration of the flow: $\dot{d} = \fracx{b\dot{v}}{\dot{z}}$.
+We finally choose a start time for the flow, using $\vec{a}$ as a vector of relative probabilities for each bin of start times, but ensuring that the start time plus the expected duration, $\dot{d}$, does not exceed the maximum simulation time for the workflow.
+From the range of possible values mapping to the chosen start bin, we choose a random time uniformly.
+
+Having determined the overall parameters of the flow, we generate the actual packets of the flow.
+The first packet is sent at the chosen start time; the size of each packet is sampled from the histogram of packet sizes, $\vec{z}$; the interval until the next packet is determined from the interval vector, $\vec{v}$.
+Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}(k)$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\seq{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
+\begin{align}
+\label{eqn:interval-dequantization}
+\Di(k,u)=v_{\min}+(v_{\max}-v_{\min})\left(\frac{k-u}{d_v}\right)^{\frac{1}{\gamma}}.
+\end{align}
+The packet stream ends when the number of bytes allocated to the flow have been emitted. %The last packet is only the size of the remaining number of bytes.
 
@@
 
-\subsection{Traffic Generation}
-\label{sec:traffic-generation}
-
-Given an instance of the general matrix model, how do we produce an actual workload from it?
-Each row of the model is used to randomly generate the behavior of a single flow.
-The type, source, and destination nodes of the flow are randomly chosen, using histogram vectors $\vec{t}$, $\vec{s}$, and $\vec{d}$ as the proportional probabilities of each type or node being chosen.\footnote{Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
-The simplest technique is the rejection method: if the same node is chosen for both source and destination, reject that pair and begin the selection process again.}
-Next we use the flow size vector, $\vec{b}$ to randomly sample the number of bytes in the flow.
-The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\dot{z}$ and $\dot{v}$.
-From these, together with the flow size in bytes, we estimate the expected duration of the flow: $\dot{d} = \fracx{b\dot{v}}{\dot{z}}$.
-We finally choose a start time for the flow, using $\vec{a}$ as a vector of relative probabilities for each bin of start times, but ensuring that the start time plus the expected duration, $\dot{d}$, does not exceed the maximum simulation time for the workflow.
-From the range of possible values mapping to the chosen start bin, we choose a random time uniformly.
-
-Having determined the overall parameters of the flow, we generate the actual packets of the flow.
-The first packet is sent at the chosen start time; the size of each packet is sampled from the histogram of packet sizes, $\vec{z}$; the interval until the next packet is determined from the interval vector, $\vec{v}$.
-Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}(k)$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\seq{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
-\begin{align}
-\label{eqn:interval-dequantization}
-\Di(k,u)=v_{\min}+(v_{\max}-v_{\min})\left(\frac{k-u}{d_v}\right)^{\frac{1}{\gamma}}.
-\end{align}
-The packet stream ends when the number of bytes allocated to the flow have been emitted. %The last packet is only the size of the remaining number of bytes.
-
 \section{Expression of Common Properties}
+\label{sec:common-properties}
 
@@
 \section{Expression of Common Modeling Concepts}
+\label{sec:modeling-concepts}
 
@@
 \end{align}
-\Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to actual behaviors from traces.
+\Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to realistic behaviors from traces.
 %By replacing $S'$ with $S$, this can also be viewed as a set of linear constraints on the matrix $S$, providing a proscriptive definition
@@
 \begin{figure}[b]
-\vspace{-1em}
+%\vspace{-1em}
 \begin{center}
@@
 \end{center}
-\vspace{-0.8em}
+\vspace{-0.9em}
 \end{figure}
@@
 We have chosen this trace, however, because within it can be found behaviors resembling many different types of wireless usage cases.
-Figure~\ref{fig:nodes-flows} shows the wide variations in the number of active flows and nodes over the course of the trace.
+\Figure{nodes-flows} shows the wide variations in the number of active flows and nodes over the course of the trace.
 In the night and morning hours, the traffic patterns are similar to those one might find in a moderately trafficked business or residential area.
@@
 
+\break
 Before using the traces, it is necessary to extract application-level behavior from the trace header data.
@@
 \item \textbf{Application:} average end-to-end delay, packet delivery ratio, received throughput.
-\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), packet drops in \caps{IP} routing queues.
+\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), packets dropped in routing queues.
 \item \textbf{Link:} 802.11 control overhead (\caps{RTS/CTS/ACK}).
@@
 
-\section{Results}\label{sec:results}
+\section{Results}
+\label{sec:results}
 
@@
 With these results, we can be assured that if we can approximate the matrix for a given example of trace traffic, then we have also approximated the original trace behavior.
-In a sense, this result is a reduction of the general traffic modeling problem to a better-understood matrix modeling problem.
-
-It is worth noting that the \caps{GMM} has competition from a few of the time-simplified models: the \model{time marginal} and \model{time uniform} model both do approximately as well, and in some cases better, for every metric examined.
+In a sense, this result is a reduction of the general traffic modeling problem to a more tractable matrix modeling problem.
+
+It is worth noting that \caps{GMM} has competition from a few of the time-simplified models: the \model{time marginal} and \model{time uniform} model both do approximately as well, and in some cases better, for the metrics examined.
 {\FHC} found strong evidence that session arrivals followed a time-varying Poisson arrival process, meaning that other than long-scale time-varying arrival rate and very small-scale clustering of flows within sessions, the overall flow arrival rate is fairly smooth.
@@
 This result indicates that the precise temporal placement of flows is not highly sensitive: they tend to be relatively evenly spaced out at short time-scales, and network performance is not sensitive to changes in start-time on that scale.
+
+Not all marginal models have such a benign effect on the accuracy of performance metrics.
+The worst offender by far is the \model{size marginal} model.
+This model distorts behavior at every level, by more than a factor 20 in 25\% of scenarios in the case of routing queue packet drops.
+This is a highly significant result because modeling flow size using a network-wide marginal distribution is precisely what {\FHC}, for example, attempt.
+This results informs us that this approach is doomed to failure: flow size cannot be assigned without considering its relation to other flow properties: source and destination nodes, and packet behavior at the very least.
+
+The \model{packet marginal} model is another offender, albeit not as egregiously.
+The direction of the misrepresentations in this case, however, is more dangerous: received throughput is overestimated by more than three times, 75\% of the time.
+On the other hand, \caps{AODV} control overhead is underestimated by half on average.
+Such serious misrepresentations are especially noteworthy given the apparently innocuous assumption applied by this model: the \model{packet marginal} model is essentially a standard variable bit-rate (\caps{VBR}) packet behavior model, using the empirical network-wide distributions of packet sizes and inter-packet intervals.
+Otherwise the model matches the original trace behavior exactly.
+And yet these drastic distortions occur.
+In our previous work, we have shown similarly drastic misrepresentations to occur when constant bit-rate packet behavior is assumed.
+This results indicates that to reproduce accurate network performance, we must provide packet behaviors on a more granular level: each node or even each flow must be assigned a ``custom'' packet behavior by some means.
+
+These strong negative results for two fundamental and very common marginal models provide severe limits on the realism that can be achieved through marginal modeling.
+Traffic modeling research \textit{must} move beyond simply finding parametric models for marginal distributions of network-wide properties.
+Realistic behavior cannot be generated using these distributions alone.
+Something must be captured about the interaction between the elements of flow behavior.
 

### 2008-02-08 paper.tex
Done with results; working on discussion.

@@
 
+Additivity of compounded errors also implies two desirable properties that are easily derived from Equation~\ref{eqn:linear-compounding}.
+It forces the error of an accurate representation to be zero: $E(x,x) = 0$.
+It also forces underestimation and overestimation to be treated symmetrically.
+The error of underestimating by some factor is opposite but equal to overestimating by the same factor:
+\begin{align}
+E(x,x/c) = - E(x,xc).
+\end{align}
+
+It is easily verified that the difference, ratio, and standard error measures do not satisfy \Equation{linear-compounding}, and the difference, as noted, does not satisfy \Equation{normalization-invariance}.
+The log-ratio is the only metric presented that satisfies both conditions.
+Moreover, it can be proved that $\log(y/x)$ is the \textit{only} differentiable function that satisfies both (up to a constant).
+In the Appendix, we present a proof of this claim.
+Throughout the rest of the paper, we use the log-ratio to measure the error of performance metrics.
+% TODO: should Appendix be capitalized?
+% TODO: hyperref the Appendix for PDF.
+
+\section{Results}
+\label{sec:results}
+
 \begin{sidewaysfigure*}
@@
 
-Additivity of compounded errors also implies two desirable properties that are easily derived from Equation~\ref{eqn:linear-compounding}.
-It forces the error of an accurate representation to be zero: $E(x,x) = 0$.
-It also forces underestimation and overestimation to be treated symmetrically.
-The error of underestimating by some factor is opposite but equal to overestimating by the same factor:
-\begin{align}
-E(x,x/c) = - E(x,xc).
-\end{align}
-
-It is easily verified that the difference, ratio, and standard error measures do not satisfy \Equation{linear-compounding}, and the difference, as noted, does not satisfy \Equation{normalization-invariance}.
-The log-ratio is the only metric presented that satisfies both conditions.
-Moreover, it can be proved that $\log(y/x)$ is the \textit{only} differentiable function that satisfies both (up to a constant).
-In the Appendix, we present a proof of this claim.
-Throughout the rest of the paper, we use the log-ratio to measure the error of performance metrics.
-% TODO: should Appendix be capitalized?
-% TODO: hyperref the Appendix for PDF.
-
-\section{Results}
-\label{sec:results}
-
 Our simulation results are summarized in \Figure{box-plots}.
@@
 Something must be captured about the interaction between the elements of flow behavior.
+
+Finally, we turn to the regularized models.
+These demonstrate neither exceptionally good nor exceptionally bad behavior.
+That the regularized models should be similar to the \model{full marginal} model but somewhat more realistic is entirely expected: regularization, the reader will recall, is a restricted form of marginal modeling, where each group of flows has its own marginal distribution.
+In these cases, the flows are grouped by source node or by destination node.
+The mixed case uses the average behavior of the other two cases.
+It is interesting to note that the mixed case is universally worse than either the source or destination case.
+Clearly, hybridizing the regularization here is not an effective technique.
+We believe that the cause of this effect is like mixing too many colors yielding an unpleasant shade of brown: any of the colors individually would be better than the non-descript average.
+As a more general conclusion, arbitrarily mixing model may not always be a good idea, despite the ease with which it is done in the matrix representation of workload behavior.
 
@@
 
-The story of network analysis and modeling is one of variety. Assumptions of ``sameness'' have fallen repeatedly. ...
-The matrix model gives new insight into
+%The story of network analysis and modeling is one of variety. Assumptions of ``sameness'' have fallen repeatedly. ...
+%The matrix model gives new insight into
 
 What are the ramifications of these results?
-The discovery that the most commonly used traffic model for wireless networks drastically misrepresents important performance metrics may shed some light on the lack of trust in results from wireless simulations.
-It is now well established that network usage behavior---both mobility, and, with this research, traffic patterns---have an impact on network performance that cannot be ignored.
+In our previous work, we had already demonstrated the drastic distortions that are introduced by uniformity assumptions made by the most common traffic models.
+This work serves to further critique established common assumptions in traffic modeling.
+The strongest message to be taken away is that the aspects of traffic behavior of both nodes and flows \textit{must} be considered jointly.
+It is not sufficient to continue modeling marginal distributions of network properties.
+These marginal models do not lend them selves easily to traffic generation in the first place, and when marginality assumptions are applied to trace traffic, the result is a severe distortion of important network performance characteristics.
+It is now well established that network traffic patterns have an impact on network performance that cannot be ignored.
 Even experimental deployments cannot avoid the need for more realistic traffic workload models.
@@
 
-\subsection{Generality of Results}
-
-% TODO: talk about other synthetic models not considered here.
-% The point of our analysis is not to show that no synthetic models can be accurate, but rather, that the commonly used models, and obvious variations thereof, are not accurate.
-
-The most significant limitation on the generality of this analysis is that it is based entirely on a single data set from \caps{IETF}60---albeit a large and varied one.
+%\subsection{Generality of Results}
+
+%% TODO: talk about other synthetic models not considered here.
+%% The point of our analysis is not to show that no synthetic models can be accurate, but rather, that the commonly used models, and obvious variations thereof, are not accurate.
+
+%The most significant limitation on the generality of this analysis is that it is based entirely on a single data set from \caps{IETF}60---albeit a large and varied one.
 It is possible that traffic in this trace happens to produce network performance that is unusually dissimilar to standard traffic models.
@@
 Traffic traces, on the other hand, must be used without significant alteration if they are to actually provide the desired realism.
+
+On the other hand, we have discredited both uniform and marginal modeling as viable approaches that can fully address the workload modeling requirements of the wireless research community.
+This, we believe, is where the General Matrix Model plays a vital role in the development of better models.
+
 %The ``messiness'' of the performance comparison from trace data in Figure~\ref{fig:mac-control-overhead-cmp} illustrates why using traces directly is not ideal: each data point differs not only in the number of nodes shown on the $x$-axis, but also in other dimensions, such as the number of flows and packets, and the average flow duration.
@@
 
-Instead of using trace data directly, it should be possible to configure a synthetic traffic model based on observations from a real data set, and then run side-by-side simulations using the synthetic model and the real data, producing statistically equivalent performance results.
+%Instead of using trace data directly, it should be possible to configure a synthetic traffic model based on observations from a real data set, and then run side-by-side simulations using the synthetic model and the real data, producing statistically equivalent performance results.
 This is precisely what our definition of sufficient realism entails.

### 2008-02-08 paper.tex
Updated plots; finished discussion.

@@
 Traffic traces, on the other hand, must be used without significant alteration if they are to actually provide the desired realism.
-
-On the other hand, we have discredited both uniform and marginal modeling as viable approaches that can fully address the workload modeling requirements of the wireless research community.
-This, we believe, is where the General Matrix Model plays a vital role in the development of better models.
+This inability to control all but one or two parameters is nearly crippling for experimental purposes.
+Therefore, the research community needs models whose parameters can be adjusted independently, but whose performance characteristics can be made to resemble whatever example of network behavior one might need.
+The General Matrix Model does not yet fulfill this role since its parameters cannot be freely varied.
+We believe, however, that \caps{GMM} plays a vital role in the ongoing process of developing  of better models.
+By reducing the general problem of traffic modeling to the domain of linear algebra, we have provided traffic researchers with a vast array of powerful theoretical and computational tools that can be immediately applied to the problem at hand.\footnote{The derived models used in this paper were all implemented in Matlab, each in only a few simple lines.
+All the models used here are implemented in only 19 lines of code.
+Of course, the framework to apply them to traffic traces and simulations is significantly larger.} What are the next steps?
+The authors are currently investigating a variety of matrix factorization and clustering techniques to extract hidden structure from a large and diverse body of real traffic.
+
+%The reader will note that the transformations applied in this paper are all quite simple.
+Our regularization matrices primarily consisted of ones and zeros.
+The algebraic representation of complete workload
 

### 2008-02-08 paper.tex
Done.

@@
 By reducing the general problem of traffic modeling to the domain of linear algebra, we have provided traffic researchers with a vast array of powerful theoretical and computational tools that can be immediately applied to the problem at hand.\footnote{The derived models used in this paper were all implemented in Matlab, each in only a few simple lines.
-All the models used here are implemented in only 19 lines of code.
-Of course, the framework to apply them to traffic traces and simulations is significantly larger.} What are the next steps?
+All the models used here are implemented in only 19 total lines of code.
+Of course, the framework to apply them to traffic traces and simulations is somewhat larger.} What are the next steps?
 The authors are currently investigating a variety of matrix factorization and clustering techniques to extract hidden structure from a large and diverse body of real traffic.
@@
 
-This research rigorously quantifies the impact of a variety of synthetic traffic models on performance metrics that wireless researchers use to evaluate new technologies and protocols.
+We have presented a fundamentally new, algebraic way of representing traffic patterns in local area networks.
+We call this representation the General Matrix Model.
+Each flow of a traffic collection is represented as a single high-dimensional vector with components representing the \caps{IP} protocol type, source and destination nodes, start time, flow size, and packet behavior.
+The essential property of the algebraic representation is that standard vector operations compute natural and useful descriptions of aggregate behavior over the collection of flows represented.
+Our experimental validation demonstrates clearly that the General Matrix Model accurately reproduces the performance characteristics of of real traffic.
+The benefits of a performance-preserving algebraic representation are multifold.
+The algebraic forms of the simplifying assumptions made by many common modeling approaches provide unprecedented clarity and coherence to a complex and confusing subject.
+For example, assumptions that various aspects of flow behavior are stochastically or deterministically uniform all take the same form algebraically: left matrix multiplication.
+The other major class of common simplifying assumptions made by traffic models is to apply some marginal behavior across the entire collection of flows or to disjoint subsets of the flows.
+Such simplifications correspond to right matrix multiplication in \caps{GMM}.
+Because of the simplicity and clarity that the algebraic structure brings to the subject, we can see clearly now that both common types of simplifying assumptions are na\"ive and simplistic.
+Our experimental results bolster this conclusion: both uniform and marginal models significantly misrepresent many important network performance metrics.
+With the General Matrix Model, however, we stand poised to use the powerful tools of modern linear algebra to develop general, elegant and powerful models of network behavior that just work.
+
+%This research rigorously quantifies the impact of a variety of synthetic traffic models on performance metrics that wireless researchers use to evaluate new technologies and protocols.
 The first step in this assessment process was to formally define what it means for a network usage model to be sufficiently realistic.
 In essence, a model is considered realistic if it produces performance results that are statistically equivalent to those produced by real usage.
-A well-defined, objective measure of realism for traffic models has not previously existed.
+%A well-defined, objective measure of realism for traffic models has not previously existed.
 Evaluations of realism have formerly relied on essentially arbitrary statistical measures of similarity to real traffic, which may or may not affect the performance metrics that researchers care about.
-The definition of sufficient realism leads us to our general experimental approach: we use differential analysis comparing performance metrics derived from real traffic with those derived from synthetic traffic models.
+%The definition of sufficient realism leads us to our general experimental approach: we use differential analysis comparing performance metrics derived from real traffic with those derived from synthetic traffic models.
 The theoretical contributions of this analysis are:
-\begin{enumerate}
-\item An in-depth analysis of the desirable mathematical properties of a measure of error for performance metrics.
-\item Proof that the unique measure of error that satisfies these properties is the log-ratio of metric values.
-\item Three rigorous tests of statistical equivalence between synthetic and real performance results.
-\end{enumerate}
-These analytical tools allow the evaluation of realism over a collection of drastically different usage scenarios.
+%\begin{enumerate}
+%\item An in-depth analysis of the desirable mathematical properties of a measure of error for performance metrics.
+%\item Proof that the unique measure of error that satisfies these properties is the log-ratio of metric values.
+%\item Three rigorous tests of statistical equivalence between synthetic and real performance results.
+%\end{enumerate}
+%These analytical tools allow the evaluation of realism over a collection of drastically different usage scenarios.
 Evaluation over a heterogeneous collection of scenarios is essential to establishing the credibility of usage models.
@@
 
-On the practical side, this paper gives crucial insight into why most researchers do not trust simulation results: with the traffic models commonly used, the results are unlikely to reflect real performance.
+%On the practical side, this paper gives crucial insight into why most researchers do not trust simulation results: with the traffic models commonly used, the results are unlikely to reflect real performance.
 Moreover, this problem will also hamper experiments in test-bed wireless networks, so long as the same na\"ive workload models are used.
@@
 The development of better traffic models should begin with real traces, and proceed by incremental changes, checked by differential analysis.
-%First, alter a small aspect of the trace, simulate, then compare.
+%%First, alter a small aspect of the trace, simulate, then compare.
 If the realism of the results is unaffected, the traffic feature altered was inessential.
 Otherwise, it is a feature of behavior that must be captured in a realistic traffic model.
-This approach will allow the precise mapping of which aspects of traffic patterns have an impact on performance and which ones can be safely abstracted away.
-% TODO: maybe cut out some of this last bit.
+%This approach will allow the precise mapping of which aspects of traffic patterns have an impact on performance and which ones can be safely abstracted away.
+%% TODO: maybe cut out some of this last bit.
 

### 2008-05-21 paper.tex
Finally found Springer journal format LaTeX files.
  - still complains about lots of fonts
  - also deleted .texshop files--using TextMate
  - deleted old IEEE format files

@@
-\documentclass[conference]{IEEEtran}
+\documentclass[twocolumn,final]{svjour3}
 
 \newcommand{\thetitle}{An Algebraic Representation of Wireless Workload}
-
-\usepackage[pdftex]{graphicx}
+\newcommand{\theauthors}{%
+  Stefan~Karpinski \and
+  Elizabeth~M.~Belding \and
+  Kevin~C.~Almeroth \and
+  John~R.~Gilbert%
+}
+
+%\usepackage[pdftex]{graphicx}
 \usepackage[labelfont=bf,small]{caption}
@@
 
-\hyphenation{net-works IEEEtran}
-\bibliographystyle{IEEEtran}
-
-\newcommand{\email}[1]{$\left<{\textit{#1}}\right>$}
 \newcommand{\caps}[1]{{\small{#1}}}
 
-\title{An Algebraic Representation of Wireless Workload}
-\author{
-{\large{Stefan~Karpinski, Elizabeth~M.~Belding, Kevin~C.~Almeroth}}\\
-\textit{\{sgk,ebelding,almeroth\}@cs.ucsb.edu}\vspace{0.5em}\\
+\title{\thetitle}
+\author{\theauthors}
+\institute{
+  \theauthors\\
   Department of Computer Science\\
-University of California, Santa Barbara
-\vspace{-0.5em}
+  University of California, Santa Barbara\\
+  Email: \texttt{\{sgk,ebelding,almeroth,gilbert\}@cs.ucsb.edu}
 }
@@
 \newcommand{\fracpx}[2]{(#1/#2)}
-
-\newcommand{\figurename}{Figure}
-\newcommand{\tablename}{Table}
 
@@
 
-\vspace{-0.5em}
-\appendix\footnotesize
+\section*{Appendix}
+\footnotesize
 
@@
 $f(z) = f'(1) \int z^{-1} dz + c = f'(1) \ln(z) + c$.
-By Eq.~\ref{eqn:linear-compounding}, $f(1)=0$, so $c=0$. Thus $f(e)=f'(1)\ln(e)=f'(1)$. We conclude that $f(z)=f(e)\ln(z)$, so $E(x,y) = f(y/x) = f(e)\ln(y/x) = E(1,e)\ln(y/x)$, as desired.\hfill\QED
+By Eq.~\ref{eqn:linear-compounding}, $f(1)=0$, so $c=0$. Thus $f(e)=f'(1)\ln(e)=f'(1)$. We conclude that $f(z)=f(e)\ln(z)$, so $E(x,y) = f(y/x) = f(e)\ln(y/x) = E(1,e)\ln(y/x)$, as desired.\hfill%\QED
 

### 2008-05-28 paper.tex
Fixed up figure sizes and some spacing issues.

@@
 
-\newcommand{\thetitle}{An Algebraic Representation of Wireless Workload}
+\newcommand{\thetitle}{Linear Representation of Wireless Workload}
 \newcommand{\theauthors}{%
@@
 \begin{tabbing}
-\hspace{1.5em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
+\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
 $D\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
@@
 \begin{center}
-\subfloat[trace node behavior]
-{\includegraphics[width=1.392in]{flow_topology_trace}\label{fig:flow-topology-trace}}
-\subfloat[uniform node behavior]
-{\includegraphics[width=1.392in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
-\subfloat[trace flow behavior]
-{\includegraphics[width=1.392in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
-\subfloat[uniform flow behavior]
-{\includegraphics[width=1.392in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
-\subfloat[packet behavior]
-{\includegraphics[width=1.392in]{flow_packet_behavior}\label{fig:packet-behavior}}
+\subfloat[node trace]
+{\includegraphics[width=1.335in]{flow_topology_trace}\label{fig:flow-topology-trace}}
+\subfloat[node uniform]
+{\includegraphics[width=1.335in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
+\subfloat[flow trace]
+{\includegraphics[width=1.335in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
+\subfloat[flow uniform]
+{\includegraphics[width=1.335in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
+\subfloat[packet]
+{\includegraphics[width=1.335in]{flow_packet_behavior}\label{fig:packet-behavior}}
 \caption{Examples illustrating trace (real) versus uniform (na\"ive synthetic) behaviors at the node, flow and packet levels.
@@
 We are not left wondering what to do about source and destination nodes or flow start and end times after we have computed our simplified packet behaviors: the same transformation that average the packet behaviors also naturally averages the start and
-
-\section{Experimental Methodology}
-\label{sec:methodology}
-
-To evaluate whether traffic models are realistic or not, we use the method of \textit{paired differential simulation} introduced in our previous work on traffic modeling~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-%This approach is based on the following definition:
-%
-%\begin{samepage}
-%\noindent\textbf{Definition.}~A model of user behavior is \textit{sufficiently realistic} if, compared with actual user behavior, the model, with parameter values extracted from the real data, yields statistically equivalent performance results.
-%\end{samepage}
-%\vspace{0.5em}
-%
-Paired differential simulation is based on the standard scientific technique of controlled experimentation. 
-The hypothesis we are testing is that a given synthetic model accurately reproduces the performance metrics exhibited by real network traffic.
-To test this hypothesis, we conduct a series of paired experiments and compare the results for each test subject with a corresponding control subject.
-In this case, the experiments are wireless network simulations, and the subjects are real or synthetic workload instances.
-The control group is the set of simulations where the workload is an actual traffic trace, recorded as described in \Section{trace-data}.
-The test group is a set of simulations where workload is synthetically generated using a simplified traffic model.
-The experiments are paired: for each control simulation, there is a synthetic workload, with behavior as similar to the control as model will allow, against which the results are compared.
-The output of the series of experiments are paired values of performance metrics for each simulated scenario: one from the control, one from the test model.
-
-The remainder of this section proceeds as follows.
-In \Section{traffic-models} we present the collection of traffic models we will evaluate.
-In \Section{trace-data} we describe the trace data that serves as the control for our experiments.
-Our simulation methodology is detailed in \Section{simulations}, and the performance metrics used for statistical comparison of control and test experiments are described in \Section{performance-metrics}.
-Finally, \Section{measures-of-error} is an in-depth mathematical discussion of the appropriate measure of error for the pairs of metric values that are produced by the paired differential simulation methodology.
-
-% to compare traffic produced by synthetic models with known examples of recorded trace traffic in a variety of experimental scenarios.
-To accomplish this, we begin with a large 24-hour example of wireless trace traffic, break it into 144 ten-minute scenarios, and use each scenario's simulated trace behavior as a control for a set synthetic traces, generated by various traffic models.
-For each scenario, the synthetic models are configured to approximate the original trace behavior as closely as the models allow.
-Important network-wide performance metrics are then collected from simulations, and synthetic performance results compared with control results to determine the realism of the synthetic models.
-
-%The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results.
-Clearly, we need not simulate subatomic particle interactions in a wireless network simulator.
-Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same.
-Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior.
-This requirement leads us to the following definition:
-
-%This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required.
-We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.
-
-\subsection{Traffic Models}
-\label{sec:traffic-models}
 
@@
 
+\section{Experimental Methodology}
+\label{sec:methodology}
+
+To evaluate whether traffic models are realistic or not, we use the method of \textit{paired differential simulation} introduced in our previous work on traffic modeling~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+%This approach is based on the following definition:
+%
+%\begin{samepage}
+%\noindent\textbf{Definition.}~A model of user behavior is \textit{sufficiently realistic} if, compared with actual user behavior, the model, with parameter values extracted from the real data, yields statistically equivalent performance results.
+%\end{samepage}
+%\vspace{0.5em}
+%
+Paired differential simulation is based on the standard scientific technique of controlled experimentation. 
+The hypothesis we are testing is that a given synthetic model accurately reproduces the performance metrics exhibited by real network traffic.
+To test this hypothesis, we conduct a series of paired experiments and compare the results for each test subject with a corresponding control subject.
+In this case, the experiments are wireless network simulations, and the subjects are real or synthetic workload instances.
+The control group is the set of simulations where the workload is an actual traffic trace, recorded as described in \Section{trace-data}.
+The test group is a set of simulations where workload is synthetically generated using a simplified traffic model.
+The experiments are paired: for each control simulation, there is a synthetic workload, with behavior as similar to the control as model will allow, against which the results are compared.
+The output of the series of experiments are paired values of performance metrics for each simulated scenario: one from the control, one from the test model.
+
+The remainder of this section proceeds as follows.
+In \Section{traffic-models} we present the collection of traffic models we will evaluate.
+In \Section{trace-data} we describe the trace data that serves as the control for our experiments.
+Our simulation methodology is detailed in \Section{simulations}, and the performance metrics used for statistical comparison of control and test experiments are described in \Section{performance-metrics}.
+Finally, \Section{measures-of-error} is an in-depth mathematical discussion of the appropriate measure of error for the pairs of metric values that are produced by the paired differential simulation methodology.
+
+% to compare traffic produced by synthetic models with known examples of recorded trace traffic in a variety of experimental scenarios.
+To accomplish this, we begin with a large 24-hour example of wireless trace traffic, break it into 144 ten-minute scenarios, and use each scenario's simulated trace behavior as a control for a set synthetic traces, generated by various traffic models.
+For each scenario, the synthetic models are configured to approximate the original trace behavior as closely as the models allow.
+Important network-wide performance metrics are then collected from simulations, and synthetic performance results compared with control results to determine the realism of the synthetic models.
+
+%The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results.
+Clearly, we need not simulate subatomic particle interactions in a wireless network simulator.
+Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same.
+Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior.
+This requirement leads us to the following definition:
+
+%This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required.
+We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.
+
+\subsection{Traffic Models}
+\label{sec:traffic-models}
+
 The first model we evaluate is the General Matrix Model itself.
@@
 \subfloat[\textbf{Application:} Average End-to-End Delay]%
-{\includegraphics[width=4.6in]{plots/Application/delay.pdf}}
+{\includegraphics[width=4.45in]{plots/Application/delay.pdf}}
 \subfloat[\textbf{Application:} Packet Delivery Ratio]%
-{\includegraphics[width=4.6in]{plots/Application/delivery_ratio.pdf}}
+{\includegraphics[width=4.45in]{plots/Application/delivery_ratio.pdf}}
 \subfloat[\textbf{Application:} Average Received Throughput]%
-{\includegraphics[width=4.6in]{plots/Application/throughput.pdf}}
+{\includegraphics[width=4.45in]{plots/Application/throughput.pdf}}
 \subfloat[\textbf{Network:} AODV Control Overhead]%
-{\includegraphics[width=4.6in]{plots/Network/control_packets.pdf}}
+{\includegraphics[width=4.45in]{plots/Network/control_packets.pdf}}
 \subfloat[\textbf{Network:} Packets Dropped in Routing Queue]%
-{\includegraphics[width=4.6in]{plots/Network/dropped.pdf}}
+{\includegraphics[width=4.45in]{plots/Network/dropped.pdf}}
 \subfloat[\textbf{Link:} 802.11 Control Overhead]%
-{\includegraphics[width=4.6in]{plots/Link/overhead.pdf}}%
+{\includegraphics[width=4.45in]{plots/Link/overhead.pdf}}%
 \caption{Box-and-whisker plots of log-ratio error values for all metrics and traffic models.

### 2008-05-28 paper.tex
More tweaking of sizes, figures and breaks.

@@
 full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above three models \\\hline\hline
-time marginal & $A'=(\ones{f \times f}/f)A$ & start time independently chosen from overall distribution of start times \\\hline
+time marginal & $A'=(\ones{f \times f}/f)A$ & start times chosen from marginal distribution of start times \\\hline
 size marginal & $B'=(\ones{f \times f}/f)B$ & flow size (bytes) is chosen independently of other properties \\\hline
@@
 
-\begin{figure}[b]
-%\vspace{-1em}
+\begin{figure}[t]
+\vspace{0.5em}
 \begin{center}
-\includegraphics[width=3.5in]{nodes-flows}%
-\vspace{-0.75em}%
+\includegraphics[width=3.3in]{nodes-flows}%
+\vspace{-1.25em}%
 \caption{The number of active nodes and flows over time.} 
@@
 \end{center}
-\vspace{-0.9em}
+\vspace{-2em}
 \end{figure}
@@
 
-\break
 Before using the traces, it is necessary to extract application-level behavior from the trace header data.

### 2008-05-28 paper.tex
Corrected a few typos & errors.

@@
 By ``cookbook,'' we mean an explicit method for generating network traffic---down to the size and transmission time of each application data unit.
-This lack is particularly relevant to wireless research, since the medium fundamentally resource-limited, and consequently local traffic patterns can have a huge impact on overall network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+This lack is particularly relevant to wireless research, since the medium is fundamentally resource-limited, and consequently local traffic patterns can have a huge impact on overall network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
 
@@
 
-Paxson and Floyd observed in~\cite{Paxson95} that the interplay between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback.
+Paxson and Floyd observed~\cite{Paxson95} that the interplay between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback.
 Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}.
@@
 
-There have been a significant number of studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}.
+There have been many studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}.
 These analyses have described a wide variety of aspects of wireless network behavior, and provide much insight into the workings of real, deployed wireless networks.
@@
 \end{tabbing}
-The total behavior is the direct product of these components: $F = \seq{T,S,D,A,B,Z,V} \in \R^{f \times \Sigma}$ where $\Sigma=256+2n+d_t+d_b+d_z+d_v$.
+The total behavior is the horizontal concatenation of these component matrices:
+\begin{align}
+F = \seq{T,S,D,A,B,Z,V} \in \R^{f \times \Sigma}
+\end{align}
+where $\Sigma=255+2n+d_t+d_b+d_z+d_v$.
 Many common traffic models can be expressed succinctly and precisely in terms of transformations or conditions on this behavior matrix and its constituents.

### 2008-05-29 paper.tex
Added type1cm font package.

--- 2008-05-28 paper.tex
@@
 
-\newcommand{\thetitle}{Linear Representation of Wireless Workload}
+\newcommand{\thetitle}{An Algebraic Representation of Wireless Workload}
 \newcommand{\theauthors}{%
@@
 
+\usepackage{type1cm}
 %\usepackage[pdftex]{graphicx}
@@
 \institute{
-  \theauthors\\
-  Department of Computer Science\\
-  University of California, Santa Barbara\\
+  \theauthors \\
+  Department of Computer Science \\
+  University of California, Santa Barbara \\
   Email: \texttt{\{sgk,ebelding,almeroth,gilbert\}@cs.ucsb.edu}
@@
 \begin{tabbing}
-\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
+\hspace{1.5em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
 $D\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
@@
 \begin{center}
-\subfloat[node trace]
-{\includegraphics[width=1.335in]{flow_topology_trace}\label{fig:flow-topology-trace}}
-\subfloat[node uniform]
-{\includegraphics[width=1.335in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
-\subfloat[flow trace]
-{\includegraphics[width=1.335in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
-\subfloat[flow uniform]
-{\includegraphics[width=1.335in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
-\subfloat[packet]
-{\includegraphics[width=1.335in]{flow_packet_behavior}\label{fig:packet-behavior}}
+\subfloat[trace node behavior]
+{\includegraphics[width=1.392in]{flow_topology_trace}\label{fig:flow-topology-trace}}
+\subfloat[uniform node behavior]
+{\includegraphics[width=1.392in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
+\subfloat[trace flow behavior]
+{\includegraphics[width=1.392in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
+\subfloat[uniform flow behavior]
+{\includegraphics[width=1.392in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
+\subfloat[packet behavior]
+{\includegraphics[width=1.392in]{flow_packet_behavior}\label{fig:packet-behavior}}
 \caption{Examples illustrating trace (real) versus uniform (na\"ive synthetic) behaviors at the node, flow and packet levels.
@@
 We are not left wondering what to do about source and destination nodes or flow start and end times after we have computed our simplified packet behaviors: the same transformation that average the packet behaviors also naturally averages the start and
+
+\section{Experimental Methodology}
+\label{sec:methodology}
+
+To evaluate whether traffic models are realistic or not, we use the method of \textit{paired differential simulation} introduced in our previous work on traffic modeling~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+%This approach is based on the following definition:
+%
+%\begin{samepage}
+%\noindent\textbf{Definition.}~A model of user behavior is \textit{sufficiently realistic} if, compared with actual user behavior, the model, with parameter values extracted from the real data, yields statistically equivalent performance results.
+%\end{samepage}
+%\vspace{0.5em}
+%
+Paired differential simulation is based on the standard scientific technique of controlled experimentation. 
+The hypothesis we are testing is that a given synthetic model accurately reproduces the performance metrics exhibited by real network traffic.
+To test this hypothesis, we conduct a series of paired experiments and compare the results for each test subject with a corresponding control subject.
+In this case, the experiments are wireless network simulations, and the subjects are real or synthetic workload instances.
+The control group is the set of simulations where the workload is an actual traffic trace, recorded as described in \Section{trace-data}.
+The test group is a set of simulations where workload is synthetically generated using a simplified traffic model.
+The experiments are paired: for each control simulation, there is a synthetic workload, with behavior as similar to the control as model will allow, against which the results are compared.
+The output of the series of experiments are paired values of performance metrics for each simulated scenario: one from the control, one from the test model.
+
+The remainder of this section proceeds as follows.
+In \Section{traffic-models} we present the collection of traffic models we will evaluate.
+In \Section{trace-data} we describe the trace data that serves as the control for our experiments.
+Our simulation methodology is detailed in \Section{simulations}, and the performance metrics used for statistical comparison of control and test experiments are described in \Section{performance-metrics}.
+Finally, \Section{measures-of-error} is an in-depth mathematical discussion of the appropriate measure of error for the pairs of metric values that are produced by the paired differential simulation methodology.
+
+% to compare traffic produced by synthetic models with known examples of recorded trace traffic in a variety of experimental scenarios.
+To accomplish this, we begin with a large 24-hour example of wireless trace traffic, break it into 144 ten-minute scenarios, and use each scenario's simulated trace behavior as a control for a set synthetic traces, generated by various traffic models.
+For each scenario, the synthetic models are configured to approximate the original trace behavior as closely as the models allow.
+Important network-wide performance metrics are then collected from simulations, and synthetic performance results compared with control results to determine the realism of the synthetic models.
+
+%The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results.
+Clearly, we need not simulate subatomic particle interactions in a wireless network simulator.
+Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same.
+Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior.
+This requirement leads us to the following definition:
+
+%This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required.
+We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.
+
+\subsection{Traffic Models}
+\label{sec:traffic-models}
 
@@
 
-\section{Experimental Methodology}
-\label{sec:methodology}
-
-To evaluate whether traffic models are realistic or not, we use the method of \textit{paired differential simulation} introduced in our previous work on traffic modeling~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-%This approach is based on the following definition:
-%
-%\begin{samepage}
-%\noindent\textbf{Definition.}~A model of user behavior is \textit{sufficiently realistic} if, compared with actual user behavior, the model, with parameter values extracted from the real data, yields statistically equivalent performance results.
-%\end{samepage}
-%\vspace{0.5em}
-%
-Paired differential simulation is based on the standard scientific technique of controlled experimentation. 
-The hypothesis we are testing is that a given synthetic model accurately reproduces the performance metrics exhibited by real network traffic.
-To test this hypothesis, we conduct a series of paired experiments and compare the results for each test subject with a corresponding control subject.
-In this case, the experiments are wireless network simulations, and the subjects are real or synthetic workload instances.
-The control group is the set of simulations where the workload is an actual traffic trace, recorded as described in \Section{trace-data}.
-The test group is a set of simulations where workload is synthetically generated using a simplified traffic model.
-The experiments are paired: for each control simulation, there is a synthetic workload, with behavior as similar to the control as model will allow, against which the results are compared.
-The output of the series of experiments are paired values of performance metrics for each simulated scenario: one from the control, one from the test model.
-
-The remainder of this section proceeds as follows.
-In \Section{traffic-models} we present the collection of traffic models we will evaluate.
-In \Section{trace-data} we describe the trace data that serves as the control for our experiments.
-Our simulation methodology is detailed in \Section{simulations}, and the performance metrics used for statistical comparison of control and test experiments are described in \Section{performance-metrics}.
-Finally, \Section{measures-of-error} is an in-depth mathematical discussion of the appropriate measure of error for the pairs of metric values that are produced by the paired differential simulation methodology.
-
-% to compare traffic produced by synthetic models with known examples of recorded trace traffic in a variety of experimental scenarios.
-To accomplish this, we begin with a large 24-hour example of wireless trace traffic, break it into 144 ten-minute scenarios, and use each scenario's simulated trace behavior as a control for a set synthetic traces, generated by various traffic models.
-For each scenario, the synthetic models are configured to approximate the original trace behavior as closely as the models allow.
-Important network-wide performance metrics are then collected from simulations, and synthetic performance results compared with control results to determine the realism of the synthetic models.
-
-%The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results.
-Clearly, we need not simulate subatomic particle interactions in a wireless network simulator.
-Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same.
-Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior.
-This requirement leads us to the following definition:
-
-%This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required.
-We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.
-
-\subsection{Traffic Models}
-\label{sec:traffic-models}
-
 The first model we evaluate is the General Matrix Model itself.
@@
 \subfloat[\textbf{Application:} Average End-to-End Delay]%
-{\includegraphics[width=4.45in]{plots/Application/delay.pdf}}
+{\includegraphics[width=4.6in]{plots/Application/delay.pdf}}
 \subfloat[\textbf{Application:} Packet Delivery Ratio]%
-{\includegraphics[width=4.45in]{plots/Application/delivery_ratio.pdf}}
+{\includegraphics[width=4.6in]{plots/Application/delivery_ratio.pdf}}
 \subfloat[\textbf{Application:} Average Received Throughput]%
-{\includegraphics[width=4.45in]{plots/Application/throughput.pdf}}
+{\includegraphics[width=4.6in]{plots/Application/throughput.pdf}}
 \subfloat[\textbf{Network:} AODV Control Overhead]%
-{\includegraphics[width=4.45in]{plots/Network/control_packets.pdf}}
+{\includegraphics[width=4.6in]{plots/Network/control_packets.pdf}}
 \subfloat[\textbf{Network:} Packets Dropped in Routing Queue]%
-{\includegraphics[width=4.45in]{plots/Network/dropped.pdf}}
+{\includegraphics[width=4.6in]{plots/Network/dropped.pdf}}
 \subfloat[\textbf{Link:} 802.11 Control Overhead]%
-{\includegraphics[width=4.45in]{plots/Link/overhead.pdf}}%
+{\includegraphics[width=4.6in]{plots/Link/overhead.pdf}}%
 \caption{Box-and-whisker plots of log-ratio error values for all metrics and traffic models.

### 2008-05-29 paper.tex
Tweaking notation formatting (vec/mat/mean) and font sizes.
The `type1cm' package actually makes all the captions, footnotes
and references much bigger, so I commented it out.

@@
 
-\usepackage{type1cm}
+%\usepackage{type1cm}
 %\usepackage[pdftex]{graphicx}
@@
 \newcommand{\zeros}[1]{\mathbf{0}_{#1}}
+\newcommand{\mat}[1]{\mathbf{#1}}
+\newcommand{\mean}[1]{\bar{#1}}
 
@@
 \hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
-$D\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
-\>\bullet\> type: \>$T$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$,\\
-\>\bullet\> sources: \>$S$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$,\\
-\>\bullet\> destinations: \>$D$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$,\\
-\>\bullet\> start times: \>$A$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$,\\
-\>\bullet\> sizes in bytes: \>$B$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$,\\
-\>\bullet\> packet sizes: \>$Z$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
-\>\bullet\> inter-packet intervals: \>$V$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
+$\mat{D}\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
+\>\bullet\> type: \>$\mat{T}$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$,\\
+\>\bullet\> sources: \>$\mat{S}$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$,\\
+\>\bullet\> destinations: \>$\mat{D}$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$,\\
+\>\bullet\> start times: \>$\mat{A}$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$,\\
+\>\bullet\> sizes in bytes: \>$\mat{B}$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$,\\
+\>\bullet\> packet sizes: \>$\mat{Z}$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
+\>\bullet\> inter-packet intervals: \>$\mat{V}$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
 \end{tabbing}
@@
 \begin{align}
-F = \seq{T,S,D,A,B,Z,V} \in \R^{f \times \Sigma}
+\mat{F} = \seq{\mat{T},\mat{S},\mat{D},\mat{A},\mat{B},\mat{Z},\mat{V}} \in \R^{f \times \Sigma},
 \end{align}
@@
 Next we use the flow size vector, $\vec{b}$ to randomly sample the number of bytes in the flow.
-The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\dot{z}$ and $\dot{v}$.
-From these, together with the flow size in bytes, we estimate the expected duration of the flow: $\dot{d} = \fracx{b\dot{v}}{\dot{z}}$.
-We finally choose a start time for the flow, using $\vec{a}$ as a vector of relative probabilities for each bin of start times, but ensuring that the start time plus the expected duration, $\dot{d}$, does not exceed the maximum simulation time for the workflow.
+The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\mean{z}$ and $\mean{v}$.
+From these, together with the flow size in bytes, we estimate the expected duration of the flow: $\mean{d} = \fracx{b\mean{v}}{\mean{z}}$.
+We finally choose a start time for the flow, using $\vec{a}$ as a vector of relative probabilities for each bin of start times, but ensuring that the start time plus the expected duration, $\mean{d}$, does not exceed the maximum simulation time for the workflow.
 From the range of possible values mapping to the chosen start bin, we choose a random time uniformly.
@@
 
-The number of flows between each pair of nodes in the network is expressed by the matrix product $SD^t\in\R^{n \times n}$ ($*^t$~denotes transpose).
-That is $(SD^t)_{ij}$ is the number of flows from node $i$ to node $j$.
-The total number of source flows for each node is given by the vector $S^t\ones{f}$, and the total number of destination flows is given by $D^t\ones{f}$.
+The number of flows between each pair of nodes in the network is expressed by the matrix product $\mat{SD}^t\in\R^{n \times n}$ ($*^t$~denotes transpose).
+That is $(\mat{SD}^t)_{ij}$ is the number of flows from node $i$ to node $j$.
+The total number of source flows for each node is given by the vector $\mat{S}^t\ones{f}$, and the total number of destination flows is given by $\mat{D}^t\ones{f}$.
 ($\ones{f}$ is the $f$-dimensional column vector consisting of all ones.)
@@
 
-From the matrices $Z$ and $V$, we can compute the expected average packet size, $\dot{z}$, and average inter-packet interval, $\dot{v}$, for each flow.
+From the matrices $\mat{Z}$ and $\mat{V}$, we can compute the expected average packet size, $\mean{z}$, and average inter-packet interval, $\mean{v}$, for each flow.
 The computations require knowing the values that each matrix column represents.
 The computed duration is only approximate because information is irrecoverably lost in the quantization process (see \Equation{interval-quantization}).
-The vectors of representative values for $Z$ and $V$ are:
+The vectors of representative values for $\mat{Z}$ and $\mat{V}$ are:
 \begin{align}
@@
 \end{align}
-Using these representative values, the expected average packet size for a flow may be computed as $\dot{z}=\fracx{\vec{z}\cdot\vec{\sigma}}{\vec{z}\cdot\ones{d_z}}$, and the expected average inter-packet interval size as $\dot{v}=\fracx{\vec{v}\cdot\vec{\tau}}{\vec{v}\cdot\ones{d_v}}$.
+Using these representative values, the expected average packet size for a flow may be computed as $\mean{z}=\fracx{\vec{z}\cdot\vec{\sigma}}{\vec{z}\cdot\ones{d_z}}$, and the expected average inter-packet interval size as $\mean{v}=\fracx{\vec{v}\cdot\vec{\tau}}{\vec{v}\cdot\ones{d_v}}$.
 
@@
 Right multiplication mixes the elements within each row, and if the mixing weights are uniform, the result is uniform behavior.
-For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $S$ and $D$:
+For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $\mat{S}$ and $\mat{D}$:
 \begin{align*}
-S' &= S\ones{n \times n}/n, & D' &= D\ones{n \times n}/n.
+\mat{S}' &= \mat{S}\ones{n \times n}/n, & \mat{D}' &= \mat{D}\ones{n \times n}/n.
 \end{align*}
@@
 The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma\in\R^{f \times 3}$, where
+The pair of matrices $\seq{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
 \begin{align}
@@
 \Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to realistic behaviors from traces.
-%By replacing $S'$ with $S$, this can also be viewed as a set of linear constraints on the matrix $S$, providing a proscriptive definition
+%By replacing $\mat{S}'$ with $\mat{S}$, this can also be viewed as a set of linear constraints on the matrix $\mat{S}$, providing a proscriptive definition
 
@@
 The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma\in\R^{f \times 3}$, where
+The pair of matrices $\seq{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
 %\begin{align}
@@
 Marginal behavior is effected by left multiplication by a matrix with uniform columns.
-This specific example is generated by left-uniformizing the flow size matrix: $B' = (\ones{f \times f}/f)B$.
+This specific example is generated by left-uniformizing the flow size matrix: $\mat{B}' = (\ones{f \times f}/f)\mat{B}$.
 The traffic models proposed by {\FHC} are all marginal: they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node.
@@
 This is simply accomplished by left-multiplication by a non-uniform matrix.
-For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-uniformizing'' matrix: $F'=(SS^t/f)F$.
-Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-uniformizing matrix: $F'=(DD^t/f)F$.
+For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-uniformizing'' matrix: $\mat{F}'=(\mat{SS}^t/f)\mat{F}$.
+Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-uniformizing matrix: $F'=(\mat{DD}^t/f)\mat{F}$.
 %Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified.
@@
 \begin{align}
-F' = \frac{1}{f}((1-\alpha)(SS^t)+\alpha(DD^t))F.
+\mat{F}' = \frac{1}{f}((1-\alpha)(\mat{SS}^t)+\alpha(\mat{DD}^t))\mat{F}.
 \end{align}
@@
 {\footnotesize{GMM}} & --- & the general matrix model \\\hline\hline
-time uniform & $A'=A(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
-node uniform & $S'=S(\ones{n \times n}/n)$~~~~$D'=D(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
+time uniform & $\mat{A}'=\mat{A}(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
+node uniform & $\mat{S}'=\mat{S}(\ones{n \times n}/n)$~~~~$\mat{D}'=\mat{D}(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
 full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above three models \\\hline\hline
-time marginal & $A'=(\ones{f \times f}/f)A$ & start times chosen from marginal distribution of start times \\\hline
-size marginal & $B'=(\ones{f \times f}/f)B$ & flow size (bytes) is chosen independently of other properties \\\hline
-node marginal & $S'=(\ones{f \times f}/f)S$~~~~$D'=(\ones{f \times f}/f)D$ & nodes behave distinctly, but behavior is independent of flows \\\hline
-packet marginal & $Z'=(\ones{f \times f}/f)Z$~~~~$V'=(\ones{f \times f}/f)V$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
+time marginal & $\mat{A}'=(\ones{f \times f}/f)\mat{A}$ & start times chosen from marginal distribution of start times \\\hline
+size marginal & $\mat{B}'=(\ones{f \times f}/f)\mat{B}$ & flow size (bytes) is chosen independently of other properties \\\hline
+node marginal & $\mat{S}'=(\ones{f \times f}/f)\mat{S}$~~~~$\mat{D}'=(\ones{f \times f}/f)\mat{D}$ & nodes behave distinctly, but behavior is independent of flows \\\hline
+packet marginal & $\mat{Z}'=(\ones{f \times f}/f)\mat{Z}$~~~~$\mat{V}'=(\ones{f \times f}/f)\mat{V}$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
 full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline\hline
-source regular & $F'=(SS^t/f)F$ & flow behaviors are aggregated on a per source-node basis \\\hline
-dest regular & $F'=(DD^t/f)F$ & flow behaviors are aggregated on a per destination-node basis \\\hline
-mixed regular & $F'=\frac{1}{2f}(SS^t+DD^t)F$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
+source regular & $\mat{F}'=(\mat{SS}^t/f)\mat{F}$ & flow behaviors are aggregated on a per source-node basis \\\hline
+dest regular & $\mat{F}'=(\mat{DD}^t/f)\mat{F}$ & flow behaviors are aggregated on a per destination-node basis \\\hline
+mixed regular & $\mat{F}'=\frac{1}{2f}(\mat{SS}^t+\mat{DD}^t)\mat{F}$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
 \end{tabular}
@@
 \begin{sidewaysfigure*}
-%\begin{figure*}[htb]
 \subfloat[\textbf{Application:} Average End-to-End Delay]%
-{\includegraphics[width=4.45in]{plots/Application/delay.pdf}}
+{\includegraphics[width=4.45in]{plots/Application/delay.pdf}}%
 \subfloat[\textbf{Application:} Packet Delivery Ratio]%
 {\includegraphics[width=4.45in]{plots/Application/delivery_ratio.pdf}}
+%
 \subfloat[\textbf{Application:} Average Received Throughput]%
-{\includegraphics[width=4.45in]{plots/Application/throughput.pdf}}
+{\includegraphics[width=4.45in]{plots/Application/throughput.pdf}}%
 \subfloat[\textbf{Network:} AODV Control Overhead]%
 {\includegraphics[width=4.45in]{plots/Network/control_packets.pdf}}
+%
 \subfloat[\textbf{Network:} Packets Dropped in Routing Queue]%
-{\includegraphics[width=4.45in]{plots/Network/dropped.pdf}}
+{\includegraphics[width=4.45in]{plots/Network/dropped.pdf}}%
 \subfloat[\textbf{Link:} 802.11 Control Overhead]%
 {\includegraphics[width=4.45in]{plots/Link/overhead.pdf}}%
-\caption{Box-and-whisker plots of log-ratio error values for all metrics and traffic models.
+%
+\caption{\footnotesize%
+Box-and-whisker plots of log-ratio error values for all metrics and traffic models.
 The lower axis indicates the log-ratio, while the upper axis shows raw ratio values.
@@
 The whiskers (dotted lines) extend to the furthest non-outlier values, while the points beyond that are outliers.
-The notch in the middle of each bar indicates a $95\%$ confidence interval for the true underlying median value; if two notches do not overlap, they are very unlikely to have the same median.}
+The notch in the middle of each bar indicates a $95\%$ confidence interval for the true underlying median value; if two notches do not overlap, they are very unlikely to have the same median.
+}
 \label{fig:box-plots}
-%\vspace{-1.25em}
-%\end{figure*}
 \end{sidewaysfigure*}

### 2008-05-29 paper.tex
break line for RREQ/RREP/RERR.
More notation fixup.

@@
 \newcommand{\mean}[1]{\bar{#1}}
+\newcommand{\trans}[1]{{#1}^T}
+\newcommand{\inner}[2]{{#1}\cdot{#2}}
 
@@
 Let $k$ be the quantized start index of a flow.
-Then the start vectors for the flow is $\vec{a}=\mathbf{e}_k$ ($\mathbf{e}_k$ has 1 at index $k$ and zeros elsewhere).
+Then the start vectors for the flow is $\vec{a}=\vec{e}_k$ (the $k$th standard basis vector, having 1 at index $k$ and zeros elsewhere).
 The advantage of this representation becomes clear when we aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.
@@
 We represent flow size with another indicator vector.
-Although the set of possible flow sizes is discrete, for practical reasons we map the very large range of possible flow sizes down to a much smaller range of indices (some flows in our trace have 109\caps{MB} of data in a single ten-minute period, which would give an indicator vector of over 100 million dimensions).
+Although the set of possible flow sizes is discrete, for practical reasons we map the very large range of possible flow sizes down to a much smaller range of indices (some flows in our trace have 109\caps{MB} of data in a single ten-minute period, which would give an indicator vector of over 100 million entries).
 We use the following function to convert flow sizes to indexes:
@@
 
-The packets size behavior of a flow is represented by a histogram vector, $\vec{z}$, of packet sizes.
+The packet size behavior of a flow is represented by a histogram vector, $\vec{z}$, of packet sizes.
 Explicitly,  $z_{ij}$ is the number of packets in the $i$th flow with $j$ bytes of payload data.
@@
 
-The number of flows between each pair of nodes in the network is expressed by the matrix product $\mat{SD}^t\in\R^{n \times n}$ ($*^t$~denotes transpose).
-That is $(\mat{SD}^t)_{ij}$ is the number of flows from node $i$ to node $j$.
-The total number of source flows for each node is given by the vector $\mat{S}^t\ones{f}$, and the total number of destination flows is given by $\mat{D}^t\ones{f}$.
+The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{SD}}\in\R^{n \times n}$ ($\trans{*}$~denotes transpose).
+That is $(\trans{\mat{SD}})_{ij}$ is the number of flows from node $i$ to node $j$.
+The total number of source flows for each node is given by the vector $\trans{\mat{S}}\ones{f}$, and the total number of destination flows is given by $\trans{\mat{D}}\ones{f}$.
 ($\ones{f}$ is the $f$-dimensional column vector consisting of all ones.)
@@
 \end{align}
-Using these representative values, the expected average packet size for a flow may be computed as $\mean{z}=\fracx{\vec{z}\cdot\vec{\sigma}}{\vec{z}\cdot\ones{d_z}}$, and the expected average inter-packet interval size as $\mean{v}=\fracx{\vec{v}\cdot\vec{\tau}}{\vec{v}\cdot\ones{d_v}}$.
+Using these representative values, the expected average packet size and expected average inter-packet interval size for a flow may be computed as
+\begin{align}
+  \mean{z}&=\fracx{\inner{\vec{z}}{\vec{\sigma}}}{\inner{\vec{z}}{\ones{d_z}}},\\
+  \mean{v}&=\fracx{\inner{\vec{v}}{\vec{\tau}}}{\inner{\vec{v}}{\ones{d_v}}},
+\end{align}
+respectively, where $\inner{\vec{x}}{\vec{y}}$ denotes the standard inner product of vectors.
 
@@
 This is simply accomplished by left-multiplication by a non-uniform matrix.
-For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-uniformizing'' matrix: $\mat{F}'=(\mat{SS}^t/f)\mat{F}$.
-Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-uniformizing matrix: $F'=(\mat{DD}^t/f)\mat{F}$.
+For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-uniformizing'' matrix: $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$.
+Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-uniformizing matrix: $F'=(\trans{\mat{DD}}/f)\mat{F}$.
 %Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified.
@@
 \begin{align}
-\mat{F}' = \frac{1}{f}((1-\alpha)(\mat{SS}^t)+\alpha(\mat{DD}^t))\mat{F}.
+\mat{F}' = \frac{1}{f}((1-\alpha)(\trans{\mat{SS}})+\alpha(\trans{\mat{DD}}))\mat{F}.
 \end{align}
@@
 full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline\hline
-source regular & $\mat{F}'=(\mat{SS}^t/f)\mat{F}$ & flow behaviors are aggregated on a per source-node basis \\\hline
-dest regular & $\mat{F}'=(\mat{DD}^t/f)\mat{F}$ & flow behaviors are aggregated on a per destination-node basis \\\hline
-mixed regular & $\mat{F}'=\frac{1}{2f}(\mat{SS}^t+\mat{DD}^t)\mat{F}$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
+source regular & $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$ & flow behaviors are aggregated on a per source-node basis \\\hline
+dest regular & $\mat{F}'=(\trans{\mat{DD}}/f)\mat{F}$ & flow behaviors are aggregated on a per destination-node basis \\\hline
+mixed regular & $\mat{F}'=\frac{1}{2f}(\trans{\mat{SS}}+\trans{\mat{DD}})\mat{F}$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
 \end{tabular}
@@
 \item \textbf{Application:} average end-to-end delay, packet delivery ratio, received throughput.
-\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), packets dropped in routing queues.
+\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/\\RERR}), packets dropped in routing queues.
 \item \textbf{Link:} 802.11 control overhead (\caps{RTS/CTS/ACK}).
@@
 \noindent
-\textbf{Theorem.} The unique differentiable function satisfying Eqs.~\ref{eqn:normalization-invariance}~\&~\ref{eqn:linear-compounding} is $E(1,e) \ln(y/x)$.
+\textbf{Theorem.} The unique differentiable function satisfying Eqs.~\ref{eqn:normalization-invariance} and \ref{eqn:linear-compounding} is $E(1,e) \ln(y/x)$.
 \textit{Proof:} Let $f(z) = E(1,z)$.

### 2008-05-29 paper.tex
Tweaking...

@@
 The simplifying assumptions made by most traffic models can be classified into three categories.
-We use the terms \textit{uniformity}, \textit{marginality}, and \textit{regularity} do describe these classes of assumptions.
+We use the terms \textit{uniformity}, \textit{marginality}, and \textit{regularity} to describe these classes of assumptions.
 They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether the make stipulations about uniformity within each flow, across all flows, or within groups of flows.
@@
 node uniform & $\mat{S}'=\mat{S}(\ones{n \times n}/n)$~~~~$\mat{D}'=\mat{D}(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
-full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above three models \\\hline\hline
+full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above models \\\hline\hline
 time marginal & $\mat{A}'=(\ones{f \times f}/f)\mat{A}$ & start times chosen from marginal distribution of start times \\\hline
@@
 The traffic generated by each synthetic model preserves as many characteristics from the original trace as possible, within the constraints of the model.
-Moreover, the following features are preserved across all models: the numbers of wireless nodes, the number of flows, the number of application-initiated data units sent, the total bytes of application data sent, and the average flow duration (and therefore the average data rate).
+Moreover, the following features are preserved across all models: the number of wireless nodes, the number of flows, the number of application-initiated data units sent, the total bytes of application data sent, and the average flow duration (and therefore the average data rate).
 
@@
 To assess the realism of these models, however, we need a measure of how inaccurate the synthetic performance values are when compared to the real values.
-Let $x$ be the value of a performance metric using real traffic, and $y$ the value of the same metric using an alternate traffic model, $\M$.
-Some common measures of error are the difference [$y-x$], the ratio [$y/x$], and the standard error [$(y-x)/x$].
-These are all reasonable measures of error; but which is most appropriate for assessing the realism of performance metrics?
+Let $x$ be the value of a performance metric using real traffic, and $y$ the value of the same metric using an alternate traffic model.
+Some common measures of error are the difference [$y-x$], the ratio [$y/x$], and the relative error [$(y-x)/x$], also commonly called the ``standard error.'' These are all reasonable measures of error; but which is most appropriate for assessing the realism of performance metrics?
 Instead of picking one arbitrarily, we will first consider the properties that an ideal error function should have, and then use those properties to determine the best measure of error.
@@
 \end{align}
-The difference measure does not satisfy Equation~\ref{eqn:normalization-invariance}, but the ratio, standard error, and log-ratio error measures all do.
+The difference measure does not satisfy Equation~\ref{eqn:normalization-invariance}, but the ratio, relative error, and log-ratio error measures all do.
 
@@
 
-It is easily verified that the difference, ratio, and standard error measures do not satisfy \Equation{linear-compounding}, and the difference, as noted, does not satisfy \Equation{normalization-invariance}.
+It is easily verified that the difference, ratio, and relative error measures do not satisfy \Equation{linear-compounding}, and the difference, as noted, does not satisfy \Equation{normalization-invariance}.
 The log-ratio is the only metric presented that satisfies both conditions.
@@
 \subfloat[\textbf{Application:} Average End-to-End Delay]%
-{\includegraphics[width=4.45in]{plots/Application/delay.pdf}}%
+{\includegraphics[width=4.45in]{plots/Application/delay.pdf}}\hfill
 \subfloat[\textbf{Application:} Packet Delivery Ratio]%
@@
 \subfloat[\textbf{Application:} Average Received Throughput]%
-{\includegraphics[width=4.45in]{plots/Application/throughput.pdf}}%
+{\includegraphics[width=4.45in]{plots/Application/throughput.pdf}}\hfill
 \subfloat[\textbf{Network:} AODV Control Overhead]%
@@
 \subfloat[\textbf{Network:} Packets Dropped in Routing Queue]%
-{\includegraphics[width=4.45in]{plots/Network/dropped.pdf}}%
+{\includegraphics[width=4.45in]{plots/Network/dropped.pdf}}\hfill
 \subfloat[\textbf{Link:} 802.11 Control Overhead]%

### 2008-05-29 paper.tex
Lots of intro work, some tweaking elsewhere.

@@
 \newcommand{\mat}[1]{\mathbf{#1}}
+\renewcommand{\vec}[1]{\mathbf{#1}}
 \newcommand{\mean}[1]{\bar{#1}}
@@
 More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local area networks~(\caps{LAN}s).
+The field of realistic workload generation, however, remains an active area of research with many unanswered questions.
+One of the major challenges of modeling full-network traffic is the manifest multilevel interdependence of behaviors.
+% The packet is the fundamental unit of traffic in any packet-switched network, including \caps{IP} networks.
+% Packets are aggregated at a variety of logical levels: by protocol type, by source and destination nodes, by flows sequences between communicating applications.
+% The behaviors of packets are specific to each aggregated group: the distribution of packet sizes is specific to each node as both source and destination, and to each flow as well.
+Because of the lack of adequate techniques for modeling this complex multilevel interdependence, three fundamental types of simplifying assumptions are commonly applied to both the analysis and generation of traffic:
+\begin{enumerate}
+\item \textbf{Uniformity:} that certain aspects of traffic behavior are uniformly distributed.
+For example, that  all nodes in a network have equal probability of being the source or destination of traffic.
+\item \textbf{Independence:} that traffic properties share a single marginal distribution across the entire network, and individual values are independently drawn from this marginal distribution.
+For example, assuming that a single distribution describes an entire network, as opposed to modeling each flow with its own distribution of packet sizes.
+\item \textbf{Subdivided Independence:} that traffic properties behave independently, but only within certain groupings.
+The assumption, for example, that all flows emanating from each single node share the same distribution of packet sizes.
+\end{enumerate}
+
 The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
@@
 
-The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{SD}}\in\R^{n \times n}$ ($\trans{*}$~denotes transpose).
+The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{SD}}\in\R^{n \times n}$ ($\trans{\,\cdot\,}$~denotes transpose).
 That is $(\trans{\mat{SD}})_{ij}$ is the number of flows from node $i$ to node $j$.

### 2008-05-29 paper.tex
More work on intro.

@@
 More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local area networks~(\caps{LAN}s).
-The field of realistic workload generation, however, remains an active area of research with many unanswered questions.
+Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
 One of the major challenges of modeling full-network traffic is the manifest multilevel interdependence of behaviors.
@@
 % The behaviors of packets are specific to each aggregated group: the distribution of packet sizes is specific to each node as both source and destination, and to each flow as well.
-Because of the lack of adequate techniques for modeling this complex multilevel interdependence, three fundamental types of simplifying assumptions are commonly applied to both the analysis and generation of traffic:
+% Because of the lack of adequate techniques for modeling this complex multilevel interdependence,
+Three types of simplifying assumptions are commonly applied to both analysis and generation of traffic:
 \begin{enumerate}
 \item \textbf{Uniformity:} that certain aspects of traffic behavior are uniformly distributed.
-For example, that  all nodes in a network have equal probability of being the source or destination of traffic.
+For example, that  all nodes in a network have equal probability of being the source or destination of packets or flows.
 \item \textbf{Independence:} that traffic properties share a single marginal distribution across the entire network, and individual values are independently drawn from this marginal distribution.
@@
 \end{enumerate}
-
-The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
+We would clearly be setting up a straw man if we implied that researchers who use these assumptions actually believe them to be true.
+They clearly are not: flows do not all share a single packet size distribution; nodes are not uniformly likely sources.
+These assumptions are applied because the drastically simplify both analysis and workload generation.
+The question remains, however, whether these assumptions are benign.
+Is the analysis and generation of network traffic drastically distorted by such assumptions?
+This is the first fundamental question that this paper addresses.
+
+While these assumptions certainly simplify both analysis and workload generation, they are quite certainly not true: flows do not all share a single packet size distribution; every node is not equally likely to be the source or destination of each packet or flow.
+
+% The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
 Other research has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback induced by the ubiquitous transmission control protocol~(\caps{TCP})~\cite{Hernandez06:dissertation}.

### 2008-05-29 paper.tex
More intro work...

@@
 \usepackage[font=small,labelfont=bf,position=top,nearskip=0em]{subfig}
-\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig}
+\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize}
 \usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle}]{hyperref}
@@
 
-\newcommand{\caps}[1]{{\small{#1}}}
+\newcommand{\caps}[1]{{\smaller{#1}}}
 
@@
 \newcommand{\FHC}{Hern\'andez-Campos~\textit{et~al.}}
-\newcommand{\class}[1]{\textsc{\small{#1}}}
-\newcommand{\model}[1]{\textsf{\small{#1}}}
+\newcommand{\class}[1]{\textsc{\smaller{#1}}}
+\newcommand{\model}[1]{\textsf{\smaller{#1}}}
 \newcommand{\RFC}[1]{\caps{RFC}~{#1}}
@@
 \item \textbf{Uniformity:} that certain aspects of traffic behavior are uniformly distributed.
-For example, that  all nodes in a network have equal probability of being the source or destination of packets or flows.
-\item \textbf{Independence:} that traffic properties share a single marginal distribution across the entire network, and individual values are independently drawn from this marginal distribution.
+For example, that  all nodes in a network have equal probability of being the source or destination of packets or flows.%
+\footnote{We use the common definition of a \textit{flow}: a sequence of packets sharing the same \caps{IP} protocol type, source and destination nodes, and \caps{TCP/UDP} port numbers.}
+\item \textbf{Independence:} that traffic properties share a single marginal%
+\footnote{The term ``marginal'' refers to the margins of actuarial tables formerly used for statistical computations.
+The rows and columns of the table are disjoint classes of two metrics.
+Each entry in a table contains a count of the number of events falling into the joint category for that row and column.
+The margins contain sums of the rows or columns, thus giving the distribution of classes along each dimension \emph{independent} of each other.}
+distribution across the entire network, and individual values are independently drawn from this marginal distribution.
 For example, assuming that a single distribution describes an entire network, as opposed to modeling each flow with its own distribution of packet sizes.
 \item \textbf{Subdivided Independence:} that traffic properties behave independently, but only within certain groupings.
-The assumption, for example, that all flows emanating from each single node share the same distribution of packet sizes.
+The assumption, for example, that all flows emanating from a node share the same distribution of packet sizes.
 \end{enumerate}
-We would clearly be setting up a straw man if we implied that researchers who use these assumptions actually believe them to be true.
+We would clearly be arguing against a straw man if we implied that researchers who apply these assumptions actually believe them to be true.
 They clearly are not: flows do not all share a single packet size distribution; nodes are not uniformly likely sources.
-These assumptions are applied because the drastically simplify both analysis and workload generation.
-The question remains, however, whether these assumptions are benign.
-Is the analysis and generation of network traffic drastically distorted by such assumptions?
-This is the first fundamental question that this paper addresses.
+These assumptions are applied because they drastically simplify both analysis and workload generation.
+There is no accepted, proven technique to represent or analyze the statistical interdependence of traffic patterns other than assuming that these interdependencies do not exist.
+Worse still, such assumptions are often made implicitly and never stated outright.
+
+The question remains whether assumptions of uniformity and independence are benign or whether they severely distort the properties of network behavior we are trying to analyze or replicate experimentally.
+%Is the analysis and generation of network traffic significantly distorted by such assumptions?
+This is the first fundamental question that this paper addresses. The that uniformity assumptions has already been made in earlier papers~\cite{Karpinski07:cbr-failure}.
+Therefore, we focus here on addressing assumptions of independence or subdivided independence.
+We find that assuming independent behaviors across entire networks is also extremely distorting of important measures of network performance.
+%This research has shown that assumptions of constant bit-rate (\caps{CBR}) flow behavior, as well as assumptions of uniformity at the flow and node levels of behavior drastically distort important performance metrics when used to generate wireless workload traffic for experiments.
 
@@
 Figure \ref{fig:packet-behavior} compares uniform (i.e.
-{\footnotesize{CBR}}) packet behavior with the trace of an actual flow.
+\caps{CBR}) packet behavior with the trace of an actual flow.
 In the uniform model, the cumulative data sent increases smoothly over time, whereas in the actual packet trace, the transmissions are variable both in size and in inter-packet interval, leading to a ``lumpy'' cumulative data plot.}
@@
 \hline
-{\footnotesize{GMM}} & --- & the general matrix model \\\hline\hline
+\caps{GMM} & --- & the general matrix model \\\hline\hline
 time uniform & $\mat{A}'=\mat{A}(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
@@
 %
-\caption{\footnotesize%
+\caption{%
 Box-and-whisker plots of log-ratio error values for all metrics and traffic models.

### 2008-05-29 paper.tex
...

@@
 
-\newcommand{\thetitle}{Linear Representation of Wireless Workload}
-\newcommand{\theauthors}{%
+\newcommand{\thetitle}{Linear Representation of Network Traffic}
+\newcommand{\theauthors}{
   Stefan~Karpinski \and
@@
   Kevin~C.~Almeroth \and
-  John~R.~Gilbert%
+  John~R.~Gilbert
 }
@@
 \usepackage[font=small,labelfont=bf,position=top,nearskip=0em]{subfig}
-\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize}
+\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize,paralist,array}
 \usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle}]{hyperref}
-\usepackage{array}
 
@@
 \institute{
-  \theauthors \\
+  S.~Karpinski \and
+  E.~Belding \and
+  K.~Almeroth \and
+  J.~Gilbert \\
   Department of Computer Science \\
-  University of California, Santa Barbara \\
+  University of California, Santa Barbara (USA) \\
   Email: \texttt{\{sgk,ebelding,almeroth,gilbert\}@cs.ucsb.edu}
@@
 
+\keywords{
+  traffic modeling \and
+  traffic analysis \and
+  workload generation \and
+  network simulation \and
+  linear algebra \and
+  matrices
+}
+
 \section{Introduction}\label{sec:intro}
@@
 Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
-One of the major challenges of modeling full-network traffic is the manifest multilevel interdependence of behaviors.
+One of the major challenges of modeling full-network traffic is the multilevel interdependence of behaviors.
 % The packet is the fundamental unit of traffic in any packet-switched network, including \caps{IP} networks.
@@
 \footnote{The term ``marginal'' refers to the margins of actuarial tables formerly used for statistical computations.
-The rows and columns of the table are disjoint classes of two metrics.
+The rows and columns of the table represent possible values of two properties.
 Each entry in a table contains a count of the number of events falling into the joint category for that row and column.
-The margins contain sums of the rows or columns, thus giving the distribution of classes along each dimension \emph{independent} of each other.}
+The margins contain sums of the rows or columns, thus giving the distribution of properties along each dimension \emph{independent} of each other.}
 distribution across the entire network, and individual values are independently drawn from this marginal distribution.
@@
 \end{enumerate}
-We would clearly be arguing against a straw man if we implied that researchers who apply these assumptions actually believe them to be true.
-They clearly are not: flows do not all share a single packet size distribution; nodes are not uniformly likely sources.
-These assumptions are applied because they drastically simplify both analysis and workload generation.
-There is no accepted, proven technique to represent or analyze the statistical interdependence of traffic patterns other than assuming that these interdependencies do not exist.
-Worse still, such assumptions are often made implicitly and never stated outright.
-
-The question remains whether assumptions of uniformity and independence are benign or whether they severely distort the properties of network behavior we are trying to analyze or replicate experimentally.
-%Is the analysis and generation of network traffic significantly distorted by such assumptions?
-This is the first fundamental question that this paper addresses. The that uniformity assumptions has already been made in earlier papers~\cite{Karpinski07:cbr-failure}.
-Therefore, we focus here on addressing assumptions of independence or subdivided independence.
+We would be arguing against a straw man if we implied that researchers who apply these assumptions actually believe them to be true.
+They clearly are not: flows have drastically different packet size distributions; different nodes initiate and receive drastically different numbers of packets and flows.
+Consider downloading a large file vs.
+typing in an \caps{SSH} session, and BitTorrent users vs.
+occasional email checkers.
+
+If no one actually believes that assumptions of uniformity and independence accurately describe network behavior, then why are these assumptions so common?
+The first reason is simply that these assumptions drastically simplify both traffic analysis and workload generation.
+If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate.
+% The answer is twofold:
+% \begin{inparaenum}[1)]
+% \item they drastically simplify both analysis and workload generation;
+% \item there is no generally accepted alternative for traffic analysis.
+% \end{inparaenum}
+% The first point is clear: if we ignore the interdependencies that make analysis difficult, the difficulties evaporate.
+This immediately begs the question of whether these assumptions are benign.
+% The question remains making these assumptions is benign.
+Does ignoring the non-uniformity and non-independence of network behavior distort the properties and metrics we are trying to analyze and replicate?
+The first contribution of this paper is to answer this question in the negative: we cannot stick our heads in the sand and pretend that network properties are uniform or independent without severely distorting crucial network performance metrics.
+
+The second reason why uniformity and independence assumptions are so commonly used is that there is no generally accepted alternative for analyzing traffic.
+%There is no proven, generally accepted approach to representing or analyzing the multilayered statistical interdependence of traffic patterns other than simply assuming that these interdependencies do not exist.
+%(Worse still, such assumptions are often made implicitly and their limitation never stated outright.)
+The difficulty lies in the high dimensional and stochastic nature of network behaviors.
+Suppose that we assume that packet sizes and inter-packet intervals are drawn from some distribution for each flow.
+Prior research has indicated that, given accurate per-flow distributions, this is a realistic generative model of network behavior~\cite{Karpinski07:realism}.%
+\footnote{It is realistic in the sense that traffic generated using this model accurately reproduces performance metrics of real traffic.}
+However, such empirical distributions are complex entities unto themselves---and worse still, each flow has its own distribution.
+Modeling the entire network's behavior thus entails producing a distribution of distributions---with only one sample value from each distribution.
+% How can we know if two flows' observed behaviors differ because they were fundamentally different or because the stochastic processes driving them happened to produce different-looking results?
+
+Uniform and independent modeling collapses network behavior (in a way that we will make precise later) to drastically reduce the dimensionality of the representation.
+This observation yields two immediate questions:
+\begin{enumerate}
+\item How can we represent the original ``uncollapsed'' network behavior?
+\item Is there a better way than assuming independence or uniformity to reduce dimensionality?
+\end{enumerate}
+It is already common practice in network modeling to think of and represent flow behaviors as vectors: the \caps{IP} protocol number, source and destination \caps{IP} addresses, and \caps{UDP/TCP} source and port numbers that define a packet flow are commonly referred to as a 5-tuple---i.e., a five-dimensional vector.
+The empirical distributions of packet sizes and inter-packet intervals that describe each flow's packet behavior are also naturally represented as histogram vectors: each dimension represents a bin and corresponding coordinate represents the number of packets falling into that bin.
+We argue that the natural generalization of this vectorial representation of flow behaviors to a collection of flows is to represent the entire collection as a matrix.
+
+The rest of the paper is organized as follows.
+In \Section{motivation} we discuss motivation and related work.
+The General Matrix Model is presented in \Section{general-matrix-model}.
+In Sections~\ref{sec:common-properties}~and~\ref{sec:modeling-concepts} we explore how common network properties and modeling concepts can be expressed in the the \caps{GMM} framework.
+Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
+The ramifications of these results are discussed in \Section{discussion}.
+Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.
+
+% The question remains whether assumptions of uniformity and independence are benign or whether they severely distort the properties of network behavior we are trying to analyze or replicate experimentally.
+% %Is the analysis and generation of network traffic significantly distorted by such assumptions?
+% This is the first fundamental question that this paper addresses. The that uniformity assumptions has already been made in earlier papers~\cite{Karpinski07:cbr-failure}.
+% Therefore, we focus here on addressing assumptions of independence or subdivided independence.
 We find that assuming independent behaviors across entire networks is also extremely distorting of important measures of network performance.
-%This research has shown that assumptions of constant bit-rate (\caps{CBR}) flow behavior, as well as assumptions of uniformity at the flow and node levels of behavior drastically distort important performance metrics when used to generate wireless workload traffic for experiments.
-
-While these assumptions certainly simplify both analysis and workload generation, they are quite certainly not true: flows do not all share a single packet size distribution; every node is not equally likely to be the source or destination of each packet or flow.
-
-% The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
+% %This research has shown that assumptions of constant bit-rate (\caps{CBR}) flow behavior, as well as assumptions of uniformity at the flow and node levels of behavior drastically distort important performance metrics when used to generate wireless workload traffic for experiments.
+% 
+% While these assumptions certainly simplify both analysis and workload generation, they are quite certainly not true: flows do not all share a single packet size distribution; every node is not equally likely to be the source or destination of each packet or flow.
+% 
+% % The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
 Other research has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback induced by the ubiquitous transmission control protocol~(\caps{TCP})~\cite{Hernandez06:dissertation}.
 Recently, the high-level statistical properties of wireless \caps{LAN} traffic have been rigorously modeled with a variety of parametric distributions~\cite{Hernandez06:wlan-traffic}.
-
-Despite great breakthroughs in understanding and methodology, the networking community remains without a ``cookbook'' for generating realistic \caps{LAN} workloads.
+% 
+% Despite great breakthroughs in understanding and methodology, the networking community remains without a ``cookbook'' for generating realistic \caps{LAN} workloads.
 By ``cookbook,'' we mean an explicit method for generating network traffic---down to the size and transmission time of each application data unit.
 This lack is particularly relevant to wireless research, since the medium is fundamentally resource-limited, and consequently local traffic patterns can have a huge impact on overall network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-
-In the current standard practice for experimental workload generation, drastic assumptions of uniformity abound.
+% 
+% In the current standard practice for experimental workload generation, drastic assumptions of uniformity abound.
 Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate.
@@
 While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
-the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that any one model is better than the rest.
+% the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that any one model is better than the rest.
 It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.
-
-We believe that two major breakthroughs are needed to improve traffic modeling.
+% 
+% We believe that two major breakthroughs are needed to improve traffic modeling.
 The first requirement is an objective standard of realism by which to compare traffic models.
@@
 This standard serves to demonstrate the harm of overly simplistic models, such as \caps{CBR}, and provides the means to test whether more complex models deliver the desired improvements.
-The second breakthrough that is needed to improve modeling is a unified framework for expressing a broad variety of traffic models, that allows researchers to systematically express and reason about the assumptions and simplifications entailed in each model.
-
-The major contribution of this paper is addressing this second requirement.
+% The second breakthrough that is needed to improve modeling is a unified framework for expressing a broad variety of traffic models, that allows researchers to systematically express and reason about the assumptions and simplifications entailed in each model.
+% 
+% The major contribution of this paper is addressing this second requirement.
 We propose the General Matrix Model (\caps{GMM}) for local network traffic.
@@
 It also provides vital mathematical clarity and allows us to apply the theoretical and computational tools of modern linear algebra to the problems of traffic analysis and modeling.
-
-After presenting \caps{GMM} and discussing how common concepts in traffic modeling can be expressed in precise mathematical terms within this framework, we proceed with experimental validation of \caps{GMM}.
+% 
+% After presenting \caps{GMM} and discussing how common concepts in traffic modeling can be expressed in precise mathematical terms within this framework, we proceed with experimental validation of \caps{GMM}.
 We use the technique of \textit{paired differential simulation} that the authors proposed and used in previous work~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
@@
 The results are very positive: \caps{GMM} accurately reproduces network performance characteristics at all levels of the protocol stack.
+
 %This does not solve the ``cookbook'' problem, however, because the \caps{GMM} must still be based on real traffic in order to reproduce its behavioral characteristics.
@@
 Once this is known, it will become feasible to create models that accurately reproduce those aspects of real wireless usage.
-
-The rest of the paper is organized as follows.
-In \Section{motivation} we discuss motivation and related work.
-The General Matrix Model is presented in \Section{general-matrix-model}.
-In Sections~\ref{sec:common-properties}~and~\ref{sec:modeling-concepts} we explore how common network properties and modeling concepts can be expressed in the the \caps{GMM} framework.
-Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
-The ramifications of these results are discussed in \Section{discussion}.
-Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.
 

### 2008-05-29 paper.tex
All your footnote are belong to us.

@@
 
+\newcommand{\newfootnote}[2]{\newcommand{#1}{\footnote{#2} }}
+
 \graphicspath{{graphs/}}
@@
   linear algebra \and
-  matrices
+  matrix analysis
 }
@@
 
-%Most of what are called traffic models are only partial descriptions of one or a handful of aspects of a total pattern of traffic in a network.
-The ubiquitous constant bit-rate (\caps{CBR}) model, for example, merely describes the nature of the packets in each flow in a workload: in a given flow, the sizes of packets and the intervals between them are are all uniform.
-Hence, the transmitted bit-rate is constant.
-
-%Two things are require for the direct empirical comparison of traffic models:
-%\begin{itemize}
-%\item An objective standard for comparing quality of models.
-%% WinMee/Broadnets provide this
-%\item A common framework in which different models can be implemented and compared.
-%% this paper provides this
-%\end{itemize}
-
-More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local area networks~(\caps{LAN}s).
-Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
-One of the major challenges of modeling full-network traffic is the multilevel interdependence of behaviors.
-% The packet is the fundamental unit of traffic in any packet-switched network, including \caps{IP} networks.
-% Packets are aggregated at a variety of logical levels: by protocol type, by source and destination nodes, by flows sequences between communicating applications.
-% The behaviors of packets are specific to each aggregated group: the distribution of packet sizes is specific to each node as both source and destination, and to each flow as well.
-% Because of the lack of adequate techniques for modeling this complex multilevel interdependence,
-Three types of simplifying assumptions are commonly applied to both analysis and generation of traffic:
-\begin{enumerate}
-\item \textbf{Uniformity:} that certain aspects of traffic behavior are uniformly distributed.
-For example, that  all nodes in a network have equal probability of being the source or destination of packets or flows.%
-\footnote{We use the common definition of a \textit{flow}: a sequence of packets sharing the same \caps{IP} protocol type, source and destination nodes, and \caps{TCP/UDP} port numbers.}
-\item \textbf{Independence:} that traffic properties share a single marginal%
-\footnote{The term ``marginal'' refers to the margins of actuarial tables formerly used for statistical computations.
+\newfootnote{\flownote}{We use the common definition of a \textit{flow} as a sequence of packets sharing the same  ``5-tuple'': \caps{IP} protocol type, source and destination nodes, and \caps{TCP/UDP} port numbers.}
+
+\newfootnote{\marginalnote}{The term ``marginal'' refers to the margins of actuarial tables formerly used for statistical computations.
 The rows and columns of the table represent possible values of two properties.
@@
 The margins contain sums of the rows or columns, thus giving the distribution of properties along each dimension \emph{independent} of each other.}
-distribution across the entire network, and individual values are independently drawn from this marginal distribution.
+
+More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local area networks~(\caps{LAN}s).
+Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
+One of the major challenges of modeling full-network traffic is the multilevel interdependence of packet, flow\flownote and node behaviors.
+Three types of simplifying assumptions are commonly applied for both analysis and generation:
+\begin{enumerate}
+\item \textbf{Uniformity:} that certain aspects of traffic behavior are uniformly distributed.
+For example, that  all nodes in a network have equal probability of being the source or destination of packets or flows.
+\item \textbf{Independence:} that traffic properties share a single marginal\marginalnote distribution across the entire network, and individual values are independently drawn from this marginal distribution.
 For example, assuming that a single distribution describes an entire network, as opposed to modeling each flow with its own distribution of packet sizes.
@@
 If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate.
-% The answer is twofold:
-% \begin{inparaenum}[1)]
-% \item they drastically simplify both analysis and workload generation;
-% \item there is no generally accepted alternative for traffic analysis.
-% \end{inparaenum}
-% The first point is clear: if we ignore the interdependencies that make analysis difficult, the difficulties evaporate.
 This immediately begs the question of whether these assumptions are benign.
-% The question remains making these assumptions is benign.
 Does ignoring the non-uniformity and non-independence of network behavior distort the properties and metrics we are trying to analyze and replicate?
@@
 
+\newfootnote{\realismnote}{It is realistic in the sense that traffic generated using this model accurately reproduces performance metrics of real traffic.}
+
 The second reason why uniformity and independence assumptions are so commonly used is that there is no generally accepted alternative for analyzing traffic.
-%There is no proven, generally accepted approach to representing or analyzing the multilayered statistical interdependence of traffic patterns other than simply assuming that these interdependencies do not exist.
-%(Worse still, such assumptions are often made implicitly and their limitation never stated outright.)
 The difficulty lies in the high dimensional and stochastic nature of network behaviors.
 Suppose that we assume that packet sizes and inter-packet intervals are drawn from some distribution for each flow.
-Prior research has indicated that, given accurate per-flow distributions, this is a realistic generative model of network behavior~\cite{Karpinski07:realism}.%
-\footnote{It is realistic in the sense that traffic generated using this model accurately reproduces performance metrics of real traffic.}
+Prior research has indicated that, given accurate per-flow distributions, this is a realistic generative model of network behavior~\cite{Karpinski07:realism}.\realismnote
 However, such empirical distributions are complex entities unto themselves---and worse still, each flow has its own distribution.
@@
 \end{enumerate}
-It is already common practice in network modeling to think of and represent flow behaviors as vectors: the \caps{IP} protocol number, source and destination \caps{IP} addresses, and \caps{UDP/TCP} source and port numbers that define a packet flow are commonly referred to as a 5-tuple---i.e., a five-dimensional vector.
-The empirical distributions of packet sizes and inter-packet intervals that describe each flow's packet behavior are also naturally represented as histogram vectors: each dimension represents a bin and corresponding coordinate represents the number of packets falling into that bin.
-We argue that the natural generalization of this vectorial representation of flow behaviors to a collection of flows is to represent the entire collection as a matrix.
+It is already common practice in network modeling to think of flow behaviors as vectors.
+The \caps{IP} protocol number, source and destination \caps{IP} addresses, and \caps{UDP/TCP} source and port numbers that define a packet flow are commonly referred to as a 5-tuple---i.e., a vector in five dimensions.
+Other aspects of a flow's behavior can be represented vectorially as well, but there is a one subtlety that may not be immediately obvious.
+
+By definition, vectors can be composed via addition, yielding a new vector representing the aggregate of the two constituent vectors.
+If we are to represent flow behaviors as vectors, then we must ensure that our representation respects addition of vectors: the sum of two flow representations should represent the aggregated behavior of the two flows.
+We demonstrate how this can be accomplished and argue that the natural generalization of this approach is to represent an entire network's behavior as a large matrix, with each row representing a single flow.
+We call such a representation of flow behavior a \emph{linear representation} since it possesses the fundamental characteristic of linearity:
+\begin{align}
+  f(x + y) = f(x) + f(y).
+\end{align}
+
+The matrix representation gives an uncollapsed representation of network behavior that preserves the non-uniformity and non-independence of flow behaviors.
 

### 2008-05-29 paper.tex
Another one bites the dust...

@@
 
-More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local area networks~(\caps{LAN}s).
+More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in packet-switched networks.
 Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
 One of the major challenges of modeling full-network traffic is the multilevel interdependence of packet, flow\flownote and node behaviors.
-Three types of simplifying assumptions are commonly applied for both analysis and generation:
+Three types of simplifying assumptions are commonly applied in both analysis and generation:
 \begin{enumerate}
@@
 \item \textbf{Subdivided Independence:} that traffic properties behave independently, but only within certain groupings.
-The assumption, for example, that all flows emanating from a node share the same distribution of packet sizes.
+The assumption, for example, that all flows emanating from a node share the same distribution of packet sizes is of this type.
 \end{enumerate}
@@
 They clearly are not: flows have drastically different packet size distributions; different nodes initiate and receive drastically different numbers of packets and flows.
-Consider downloading a large file vs.
-typing in an \caps{SSH} session, and BitTorrent users vs.
-occasional email checkers.
+Simple examples illustrate this: downloading a large file vs. typing in an \caps{SSH} session; BitTorrent users vs. occasional email checkers.
 
@@
 
-The second reason why uniformity and independence assumptions are so commonly used is that there is no generally accepted alternative for analyzing traffic.
+The second reason why uniformity and independence assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
 The difficulty lies in the high dimensional and stochastic nature of network behaviors.
 Suppose that we assume that packet sizes and inter-packet intervals are drawn from some distribution for each flow.
-Prior research has indicated that, given accurate per-flow distributions, this is a realistic generative model of network behavior~\cite{Karpinski07:realism}.\realismnote
-However, such empirical distributions are complex entities unto themselves---and worse still, each flow has its own distribution.
-Modeling the entire network's behavior thus entails producing a distribution of distributions---with only one sample value from each distribution.
+Prior research has indicated that, given accurate per-flow distributions, this is a fairly realistic model of network behavior~\cite{Karpinski07:realism}.\realismnote
+Such empirical distributions, however, are complex entities unto themselves---and worse still, each flow has its own distribution.
+Modeling the entire network's behavior thus entails producing a distribution of distributions.
 % How can we know if two flows' observed behaviors differ because they were fundamentally different or because the stochastic processes driving them happened to produce different-looking results?
 
-Uniform and independent modeling collapses network behavior (in a way that we will make precise later) to drastically reduce the dimensionality of the representation.
-This observation yields two immediate questions:
+Uniform and independent modeling reduce the dimensionality of network behavior drastically, making analysis and generation tractable problems.
+There are two immediate questions:
 \begin{enumerate}
-\item How can we represent the original ``uncollapsed'' network behavior?
-\item Is there a better way than assuming independence or uniformity to reduce dimensionality?
+\item Can we represent the original ``unreduced'' network behavior in such a way that analysis and generation are still tractable problems?
+\item Is there a better way to reduce dimensionality than assuming independence or uniformity of behaviors?
 \end{enumerate}
+The second contribution of this paper is to answer these questions and to propose a mathematical representation of network traffic that preserves the non-uniformity and non-independence of flow behaviors in a network, but makes traffic analysis and workload generation from this representation tractable problems.
+Here we will briefly explain our approach.
+
 It is already common practice in network modeling to think of flow behaviors as vectors.
 The \caps{IP} protocol number, source and destination \caps{IP} addresses, and \caps{UDP/TCP} source and port numbers that define a packet flow are commonly referred to as a 5-tuple---i.e., a vector in five dimensions.
-Other aspects of a flow's behavior can be represented vectorially as well, but there is a one subtlety that may not be immediately obvious.
-
-By definition, vectors can be composed via addition, yielding a new vector representing the aggregate of the two constituent vectors.
-If we are to represent flow behaviors as vectors, then we must ensure that our representation respects addition of vectors: the sum of two flow representations should represent the aggregated behavior of the two flows.
-We demonstrate how this can be accomplished and argue that the natural generalization of this approach is to represent an entire network's behavior as a large matrix, with each row representing a single flow.
-We call such a representation of flow behavior a \emph{linear representation} since it possesses the fundamental characteristic of linearity:
-\begin{align}
-  f(x + y) = f(x) + f(y).
-\end{align}
-
-The matrix representation gives an uncollapsed representation of network behavior that preserves the non-uniformity and non-independence of flow behaviors.
+Other aspects of a flow's behavior can be represented vectorially as well.
+We call such a mapping of flows into a vector space a \emph{linear representation} if it satisfies the following simple linearity criterion:
+\begin{quote}
+The sum of the representations of two flows must be another vector that represents the aggregate behavior of the two flows.
+\end{quote}
+We show that flows can be represented satisfying this criterion. This immediately allows us to perform useful computations on traffic patterns through simple vector addition and scaling.
+Finally, we argue that the natural generalization of this approach is to represent an entire network's behavior as a large matrix, with each row representing a single flow.
+The matrix representation gives a representation of network behavior that preserves the non-uniformity and non-independence of flow behaviors, but still is highly amenable to analysis using theoretical and numerical linear algebra.
+Moreover, we describe how to generate traffic from arbitrary traffic matrices, and demonstrate experimentally that the specific matrix representation we propose preserves the performance characteristics of real traffic.
+
+% TODO: update roadmap paragraph.
 
@@
 Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.
-
-% The question remains whether assumptions of uniformity and independence are benign or whether they severely distort the properties of network behavior we are trying to analyze or replicate experimentally.
-% %Is the analysis and generation of network traffic significantly distorted by such assumptions?
-% This is the first fundamental question that this paper addresses. The that uniformity assumptions has already been made in earlier papers~\cite{Karpinski07:cbr-failure}.
-% Therefore, we focus here on addressing assumptions of independence or subdivided independence.
-We find that assuming independent behaviors across entire networks is also extremely distorting of important measures of network performance.
-% %This research has shown that assumptions of constant bit-rate (\caps{CBR}) flow behavior, as well as assumptions of uniformity at the flow and node levels of behavior drastically distort important performance metrics when used to generate wireless workload traffic for experiments.
-% 
-% While these assumptions certainly simplify both analysis and workload generation, they are quite certainly not true: flows do not all share a single packet size distribution; every node is not equally likely to be the source or destination of each packet or flow.
-% 
-% % The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}.
-Other research has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback induced by the ubiquitous transmission control protocol~(\caps{TCP})~\cite{Hernandez06:dissertation}.
-Recently, the high-level statistical properties of wireless \caps{LAN} traffic have been rigorously modeled with a variety of parametric distributions~\cite{Hernandez06:wlan-traffic}.
-% 
-% Despite great breakthroughs in understanding and methodology, the networking community remains without a ``cookbook'' for generating realistic \caps{LAN} workloads.
-By ``cookbook,'' we mean an explicit method for generating network traffic---down to the size and transmission time of each application data unit.
-This lack is particularly relevant to wireless research, since the medium is fundamentally resource-limited, and consequently local traffic patterns can have a huge impact on overall network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-% 
-% In the current standard practice for experimental workload generation, drastic assumptions of uniformity abound.
-Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate.
-The flows in an experiment are commonly given the same duration and data rate.
-Nodes in the network are often selected as flow endpoints randomly or arbitrarily.
-While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
-% the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that any one model is better than the rest.
-It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.
-% 
-% We believe that two major breakthroughs are needed to improve traffic modeling.
-The first requirement is an objective standard of realism by which to compare traffic models.
-Such a standard has already been introduced and used to demonstrate that uniform traffic modeling, including the ubiquitous \caps{CBR} model, drastically distort performance metrics at all levels of the network protocol stack~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-This standard serves to demonstrate the harm of overly simplistic models, such as \caps{CBR}, and provides the means to test whether more complex models deliver the desired improvements.
-% The second breakthrough that is needed to improve modeling is a unified framework for expressing a broad variety of traffic models, that allows researchers to systematically express and reason about the assumptions and simplifications entailed in each model.
-% 
-% The major contribution of this paper is addressing this second requirement.
-We propose the General Matrix Model (\caps{GMM}) for local network traffic.
-In this model, the traffic pattern of an entire network is represented as a large, sparse matrix.
-Each row vector represents a single flow of packets in the network, with components representing the \caps{IP} protocol type, source and destination nodes, flow size, start time, and packet behavior.
-The crucial property of this representation is that it respects vector space operations: summation of behavior vectors gives an accurate description of the aggregate behavior of the flows represented by those vectors.
-One benefit of this representation is that it makes it immediately possible to express a broad variety of assumptions and conditions on traffic patterns in terms of simple matrix operations.
-It also provides vital mathematical clarity and allows us to apply the theoretical and computational tools of modern linear algebra to the problems of traffic analysis and modeling.
-% 
-% After presenting \caps{GMM} and discussing how common concepts in traffic modeling can be expressed in precise mathematical terms within this framework, we proceed with experimental validation of \caps{GMM}.
-We use the technique of \textit{paired differential simulation} that the authors proposed and used in previous work~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-This technique uses real wireless trace traffic as an experimental control and conducts paired experiments comparing network performance from simulations using control traffic against others using synthetically generated traffic.
-The results are very positive: \caps{GMM} accurately reproduces network performance characteristics at all levels of the protocol stack.
-
-%This does not solve the ``cookbook'' problem, however, because the \caps{GMM} must still be based on real traffic in order to reproduce its behavioral characteristics.
-It does however, reduce the problem of general traffic modeling to a more constrained problem of matrix modeling.
-
-%Our experiments with various common assumptions of uniformity and regularity in network behavior yield mixed conclusions: most straightforward assumptions of homogenous behavior distort performance results significantly, but there are notable exceptions.
-The clearest exception is the time domain: traffic models appear to be relatively insensitive to local perturbation of the start time of flows---so long as the overall distribution remains approximately correct.
-Our conclusion is that traffic modeling needs to discard its assumptions of uniformity, regularity and independence.
-Instead we need to move ahead trying to understand the complex dependencies between the many aspects of traffic behaviors.
-By reducing traffic modeling to the well-understood domain of numerical linear analysis, the General Matrix Model brings us significantly closer to the ultimate goal of having a ``cookbook'' model for generating arbitrary, realistic experimental workload patterns.
-
-%Worse still, it is unclear what ``realistic'' should mean in this context.
-Foremost, a clear and testable definition of realism for traffic models is needed.
-In the absence of such a definition, it remains standard in wireless experiments to generate traffic using the simplest possible model: all nodes, flows and packets are assumed to behave uniformly.
-Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily.
-While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
-%the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better.
-It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.
-
-%To address this foundational problem in network traffic modeling, we have introduced a fundamentally different approach to evaluating the realism of workload models~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics.
-We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values similar to those observed using the original trace to generate traffic.
-Using this new approach, we explore the space of traffic models, establishing rigorously that the commonly used uniform models for node, flow and packet behavior drastically misrepresent important \caps{WLAN} performance metrics at all levels of the protocol stack.
-
-%It is natural to compare these common, na\"ive models with the high-level parametric models proposed by \FHC~\cite{Hernandez06:wlan-traffic}.
-Here, however, we face another fundamental challenge in traffic modeling.
-Even with specific models for various aspects of network behavior, it remains non-trivial to combine them into a coherent composite model capable of generating concrete workloads.
-The {\FHC} models, for example, specify the distribution of flows per node in the network; since every flow has two endpoints, however, it remains unclear how to assign endpoints to flows such that the marginal distribution of flow per node matches the BiPareto model indicated.
-Similarly, the models dictate BiPareto marginal distribution of flow sizes across the entire network, but say nothing about how these sizes are to be distributed among the nodes.
-We could assume that the number of flows per node is independent from the size of the flows at that node, but, as we shall see, this assumption is invalid.
-Moreover, the details of flow behavior---packet sizes and inter-packet intervals---remain entirely undefined.
-
-%Our point is not to denigrate the models proposed in~\cite{Hernandez06:wlan-traffic} in any way.
-On the contrary: these parametric models are the state of the art in modeling traffic patterns for wireless local area networks.
-The purpose, rather, is to point out how large a gap still remains between having accurate high-level models of marginal aspects of network behavior and being able to generate usable workloads with those models.
-Moreover, a great deal of ambiguity lies in that gap since there is no single obvious way to combine models.
-
-%To clarify, disambiguate and standardize the specification of generative local area traffic models, we propose the \textit{Traffic Object Model} (\textit{\caps{TOM}}).
-The \caps{TOM} entails a hierarchy of abstract classes: \class{Node}, \class{Flow}, and \class{Packet} are the primary classes; the \class{Network} class collects everything.
-Each abstract class defines an interface of properties.
-A traffic model defines concrete classes at each level that provide the required interfaces; each class defines its own properties by delegation, inheritance, or overriding properties of other classes.
-Because of the separation of the \caps{TOM} into well-defined levels, each concrete class naturally defines a sub-model, specific to that level's behavior; sub-models can be mixed and matched combinatorially.
-
-%We define a range of sub-models at each level of the \caps{TOM}, with decreasing similarity to the original trace behavior: \class{Trace} (reproduce original behavior), \class{Nonparametric}, \class{Parametric}, and \class{Uniform}.
-The control model uses the Trace sub-model at every level, exactly reproducing the behavior recorded from the original network.
-The standard na\"ive model uses the \class{Uniform} sub-model at all levels.
-The models proposed by {\FHC} are implemented by applying the \class{Parametric} sub-models at all levels.
-The \class{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.
-
-% objects and their properties together with rules for delegating, inheriting, and overriding the properties across the objects in the \caps{TOM} hierarchy. The \caps{TOM} allows
-
-%The {\FHC} models, like most network behavior models, are marginal, meaning that they are fitted to data aggregated of over all nodes in the network, and not fitted on a per-node basis.
-With marginal models, since all information about interactions between the models is lost, so we must apply each model independently and hope that the result is good enough.
-
-%Before we can define traffic models, however, we face the next major fundamental challenge in traffic modeling.
-
-%In this paper, we address this fundamental problem in network traffic modeling by presenting an 
-%In this paper, we propose objective criteria by which to judge whether
-
-%The experimental evaluation of new wireless technologies inevitably requires network traffic.
-The most commonly used models for generating wireless local-area network (\caps{WLAN}) workload are simple, treating both nodes and flows uniformly, and are generally understood to be na\"ive.
-They remain popular because of the lack of compelling alternatives.
-The scope of impact of different workloads on network performance is largely unresearched and thus, poorly understood.
-We believe that one of the major handicaps preventing wireless experiments from more accurately predicting real-world performance is the lack of realistic models of \caps{WLAN} workload.
-
-%The evaluation of wireless technology requires the generation of workload to test the viability and performance of the new protocol or technique being studied.
-We believe that lack of realism in traffic workload generation is one of the major limiting factors that prevents simulations and experimental test-bed deployments from accurately predicting the real-world performance of wireless technologies.
-Today, very little is understood about the impact of different workloads on network performance.
-Uniform constant bit-rate traffic (\caps{CBR}) is commonly used to evaluate protocols, but there is no evidence that behavior under such workloads is an accurate predictor of performance under real usage patterns.
-The inability to experimentally forecast real-world performance is a severe handicap to the entire networking community.
-It hinders the ability to effectively develop better solutions to the many difficult problems that face emerging wireless technologies.
-
-%One of the hardest problems in modeling phenomena as complex as network traffic patterns lies in determining which characteristics of any given exemplar of behavior are essential and which can safely be abstracted away.
-In \caps{WLAN} traffic modeling, important characteristics have largely been determined based on intuition.
-In previous work~\cite{Karpinski07:realism}, we introduced a fundamentally different approach to evaluating the realism of workload models.
-Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics.
-We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values that are similar to those observed using the original trace to generate traffic.
-Using this new approach, we explore the space of traffic models, establishing rigorously for the first time that common but simplistic models of \caps{WLAN} traffic drastically misrepresent important performance metrics at all levels of the protocol stack.
-
-%Our analysis shows that the standard constant bit-rate (\caps{CBR}) traffic model, with uniformly chosen nodes and uniform flows of traffic, significantly misrepresents important performance metrics at every level of the network.
-To qualify these results in more depth, we also investigate partially synthetic variants of this standard model, allowing us to reason about which aspects of workload models have the most impact.
-For the most commonly used model, the delay of application data traversing the network from sender to receiver is underestimated by more than a factor of 5, \textit{on average}, and by more than a factor of 11 in a quarter of usage scenarios.
-Network control overhead is overestimated, on average, by a factor of about 2.5, while link control overhead is underestimated by almost a factor of 2.
-Even for metrics where the average and median misrepresentations are not so extreme, the statistical characteristics of error values often indicate that something is fundamentally unrealistic.
-None of the partially synthetic models manage to accurately represent more than one or two performance characteristics.
-We further use the \caps{AODV}~\cite{rfc:aodv} and \caps{OLSR}~\cite{rfc:olsr} ad~hoc routing protocols to show that the relative performance of protocols can be inverted when changing from using a simplistic traffic model to using real traffic: using random endpoint, uniform \caps{CBR} traffic, \caps{AODV} appears to induce less link-layer overhead than \caps{OLSR}, whereas using real traffic, it in fact induces more.
-
-%The first major contribution of this work is the definition of ``sufficient realism,'' together with the analytical and statistical methodology to rigorously test whether synthetic traffic models meet this definition.
-The second contribution is the conclusion that the most commonly used model of wireless traffic drastically and consistently misrepresents some of the most important metrics for performance evaluation of wireless protocols.
-Many performance comparisons based on this model may need revisiting.
-The last and most important contribution is the collection of ideas and analytical tools necessary to create more realistic synthetic traffic models in the future.
-By applying the methods developed in this work, it will be possible to discover precisely which aspects of network usage affect the realism of performance results.
-Once this is known, it will become feasible to create models that accurately reproduce those aspects of real wireless usage.
 

### 2008-05-29 paper.tex
Fixed bibliography problems.

@@
 \begin{quote}
-The sum of the representations of two flows must be another vector that represents the aggregate behavior of the two flows.
+The sum of the representations of two flows is a vector that represents the aggregate behavior of the two flows.
 \end{quote}
@@
 
-The history of networking research contains many examples of simplistic models that have proven not only to be inaccurate, but also to drastically skew important characteristics of network behavior.
+%The history of networking research contains many examples of simplistic models that have proven not only to be inaccurate, but also to drastically skew important characteristics of network behavior.
 Paxson and Floyd showed that the Poisson packet arrival model, which had been standard for studying wide-area Internet traffic, failed to capture the burstiness and self-similarity of real traffic~\cite{Paxson95}.
@@
 
+\newfootnote{\QoSnote}{This is violated by some quality of service (QoS) schemes.
+However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid.
+The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.}
+
 The interaction of wireless users and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted.
 The joint pattern of traffic generation and mobility through time and space completely determines the effect of wireless usage on the lower levels of the network.
-This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner.%~\cite{Clark88}.
-\footnote{This is violated by some quality of service (QoS) schemes.
-However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid.
-The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.} The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic.
+This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner~\cite{Clark88}.\QoSnote The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic.
 
@@
 % Hernandez06:dissertation: ``complete methodology for reproducing the traffic observed on a network link in a closed-loop manner, and proposed a number of metrics for studying the realism of the generated traffic.''
-\bibliography{IEEE,references}
+\bibliographystyle{plain}
+\bibliography{references}
 

### 2008-05-30 paper.tex
Moved trace vs. uniform figure up & out (of paper.tex).

@@
 \newcommand{\FHC}{Hern\'andez-Campos~\textit{et~al.}}
-\newcommand{\class}[1]{\textsc{\smaller{#1}}}
-\newcommand{\model}[1]{\textsf{\smaller{#1}}}
+\newcommand{\class}[1]{\textsc{#1}}
+\newcommand{\model}[1]{\textsf{#1}}
 \newcommand{\RFC}[1]{\caps{RFC}~{#1}}
@@
 \newcommand{\mat}[1]{\mathbf{#1}}
-\renewcommand{\vec}[1]{\mathbf{#1}}
 \newcommand{\mean}[1]{\bar{#1}}
@@
 
+\input{trace_vs_uniform}
+
 % TODO: update roadmap paragraph.
@@
 The ramifications of these results are discussed in \Section{discussion}.
-Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.
+Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies.
+%, and how it points the way to better traffic models for the future.
 
@@
 
-The interaction of wireless users and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted.
+The interaction of network users and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted.
 The joint pattern of traffic generation and mobility through time and space completely determines the effect of wireless usage on the lower levels of the network.
 This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner~\cite{Clark88}.\QoSnote The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic.
+\Figure{trace-vs-uniform} demonstrates how drastically different network traffic appears when assumptions of uniform behavior at various levels are imposed.
+Assumptions of independence are not as visually dramatic, but we will demonstrate that they distort the performance characteristics of networks almost as much---or more---than uniform models.
 
@@
 The packet stream ends when the number of bytes allocated to the flow have been emitted. %The last packet is only the size of the remaining number of bytes.
-
-\begin{figure*}[tb]
-\begin{center}
-\subfloat[node trace]
-{\includegraphics[width=1.335in]{flow_topology_trace}\label{fig:flow-topology-trace}}
-\subfloat[node uniform]
-{\includegraphics[width=1.335in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
-\subfloat[flow trace]
-{\includegraphics[width=1.335in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
-\subfloat[flow uniform]
-{\includegraphics[width=1.335in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
-\subfloat[packet]
-{\includegraphics[width=1.335in]{flow_packet_behavior}\label{fig:packet-behavior}}
-\caption{Examples illustrating trace (real) versus uniform (na\"ive synthetic) behaviors at the node, flow and packet levels.
-Figures \ref{fig:flow-topology-trace} and \ref{fig:flow-topology-uniform} show example node behaviors.
-The width of each line is proportional to the logarithm of the number of flows between the nodes it connects (zero is the Internet gateway).
-Uniform and trace flow behavior examples are plotted in Figures \ref{fig:flow-behavior-trace} and \ref{fig:flow-behavior-uniform}.
-The time axis indicates when flows start and end; the width of each flow line is proportional to the logarithm of its data rate.
-Figure \ref{fig:packet-behavior} compares uniform (i.e.
-\caps{CBR}) packet behavior with the trace of an actual flow.
-In the uniform model, the cumulative data sent increases smoothly over time, whereas in the actual packet trace, the transmissions are variable both in size and in inter-packet interval, leading to a ``lumpy'' cumulative data plot.}
-\label{fig:trace-vs-uniform}
-\end{center}
-\vspace{-1.25em}
-\end{figure*}
 

### 2008-05-30 trace_vs_uniform.tex
Moved trace vs. uniform figure up & out (of paper.tex).

\begin{figure*}[!t]
\begin{center}
\subfloat[node trace]
{\includegraphics[width=1.335in]{flow_topology_trace}\label{fig:flow-topology-trace}}
\subfloat[node uniform]
{\includegraphics[width=1.335in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
\subfloat[flow trace]
{\includegraphics[width=1.335in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
\subfloat[flow uniform]
{\includegraphics[width=1.335in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
\subfloat[packet]
{\includegraphics[width=1.335in]{flow_packet_behavior}\label{fig:packet-behavior}}
\caption{Examples illustrating trace (real) versus uniform (na\"ive synthetic) behaviors at the node, flow and packet levels. Figures \ref{fig:flow-topology-trace} and \ref{fig:flow-topology-uniform} show example node behaviors. The width of each line is proportional to the logarithm of the number of flows between the nodes it connects (zero is the Internet gateway). Uniform and trace flow behavior examples are plotted in Figures \ref{fig:flow-behavior-trace} and \ref{fig:flow-behavior-uniform}. The time axis indicates when flows start and end; the width of each flow line is proportional to the logarithm of its data rate. Figure \ref{fig:packet-behavior} compares uniform (i.e. \caps{CBR}) packet behavior with the trace of an actual flow. In the uniform model, the cumulative data sent increases smoothly over time, whereas in the actual packet trace, the transmissions are variable both in size and in inter-packet interval, leading to a ``lumpy'' cumulative data plot.}
\label{fig:trace-vs-uniform}
\end{center}
\vspace{-1.25em}
\end{figure*}

### 2008-05-30 paper.tex
...

@@
 While we can speculate about what an individual node might hypothetically do, we currently hardly understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations.
-Especially in the wireless setting, a fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback.
+A fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback.
 Accordingly, in this paper we attempt to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without multi-level feedback.
@@
 
-There have been many studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}.
+% There have been many studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}.
 These analyses have described a wide variety of aspects of wireless network behavior, and provide much insight into the workings of real, deployed wireless networks.
@@
 Our work provides the methodology for turning this rich foundation of field data into usable, realistic models of workload for a wide variety of networking situations.
-
-The choice of mobility models for mobile wireless simulations can have a drastic impact on important performance metrics~\cite{Camp02,Yoon03:speed-decay,Yoon03:sound-models,Jardosh03,Zheng04}.
+% 
+% The choice of mobility models for mobile wireless simulations can have a drastic impact on important performance metrics~\cite{Camp02,Yoon03:speed-decay,Yoon03:sound-models,Jardosh03,Zheng04}.
 Moreover, commonly used but simplistic mobility models, such as \caps{RWP}, exhibit characteristics, including density waves and speed decay, that are categorically dissimilar from any known real-world behavior~\cite{Royer01,Yoon03:speed-decay,Yoon03:sound-models}.
@@
 
-In this paper, instead of mobility, we examine an even more fundamental aspect of user behavior in wireless networks: the pattern of traffic generated by users and applications.
+% In this paper, instead of mobility, we examine an even more fundamental aspect of user behavior in wireless networks: the pattern of traffic generated by users and applications.
 This aspect of behavior is more fundamental because it applies to all types of wireless networks, not just mobile and ad~hoc networks.

### 2008-06-04 paper.tex
Independent => Marginal; Subdivided => Conditional.
Lots more work on intro.

@@
 %\renewcommand{\labelitemi}{\bullet}
+\newcommand{\latin}[1]{\textit{#1}}
 
@@
 
-More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in packet-switched networks.
+More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local-area networks (\caps{LAN}s).
 Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
@@
 \begin{enumerate}
-\item \textbf{Uniformity:} that certain aspects of traffic behavior are uniformly distributed.
-For example, that  all nodes in a network have equal probability of being the source or destination of packets or flows.
-\item \textbf{Independence:} that traffic properties share a single marginal\marginalnote distribution across the entire network, and individual values are independently drawn from this marginal distribution.
-For example, assuming that a single distribution describes an entire network, as opposed to modeling each flow with its own distribution of packet sizes.
-\item \textbf{Subdivided Independence:} that traffic properties behave independently, but only within certain groupings.
-The assumption, for example, that all flows emanating from a node share the same distribution of packet sizes is of this type.
+\item \textbf{Uniform:} that traffic behaviors are uniformly distributed. For example, assuming that  all nodes in a network have equal probability of being the source or destination of each flows.
+Another common example is the assumption that all packet sizes and inter-packet intervals are the same; this is commonly known as the ``constant bit-rate'' (\caps{CBR}) model.
+\item \textbf{Marginal:} that traffic properties share a single, common \emph{marginal}\marginalnote distribution across the entire network, and individual values are drawn independently and unconditionally from this shared distribution.
+For example, assuming that a single distribution of packet sizes describes all flows in a network, as compared to equipping each flow in the network with its own specific distribution.
+\item \textbf{Conditional:} that traffic behaves marginally, but only within equivalence classes defined by certain conditions.
+The assumption, for example, that all flows emanating from a common source node share the same packet size distribution would be a conditional model, conditioned on source nodes.
 \end{enumerate}
@@
 They clearly are not: flows have drastically different packet size distributions; different nodes initiate and receive drastically different numbers of packets and flows.
-Simple examples illustrate this: downloading a large file vs. typing in an \caps{SSH} session; BitTorrent users vs. occasional email checkers.
-
-If no one actually believes that assumptions of uniformity and independence accurately describe network behavior, then why are these assumptions so common?
+Simply considering intuitive examples of common behavior illustrate this:
+downloading a large file \latin{vs.} typing in an \caps{SSH} session;
+BitTorrent users \latin{vs.} occasional email checkers.
+
+If no one actually believes that assumptions of uniformity and marginality accurately describe network behavior, then why are these assumptions so common?
 The first reason is simply that these assumptions drastically simplify both traffic analysis and workload generation.
@@
 This immediately begs the question of whether these assumptions are benign.
-Does ignoring the non-uniformity and non-independence of network behavior distort the properties and metrics we are trying to analyze and replicate?
-The first contribution of this paper is to answer this question in the negative: we cannot stick our heads in the sand and pretend that network properties are uniform or independent without severely distorting crucial network performance metrics.
+Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to analyze and replicate?
+We answer this question definitively in the negative: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
 
@@
 
-The second reason why uniformity and independence assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
-The difficulty lies in the high dimensional and stochastic nature of network behaviors.
-Suppose that we assume that packet sizes and inter-packet intervals are drawn from some distribution for each flow.
-Prior research has indicated that, given accurate per-flow distributions, this is a fairly realistic model of network behavior~\cite{Karpinski07:realism}.\realismnote
-Such empirical distributions, however, are complex entities unto themselves---and worse still, each flow has its own distribution.
-Modeling the entire network's behavior thus entails producing a distribution of distributions.
-% How can we know if two flows' observed behaviors differ because they were fundamentally different or because the stochastic processes driving them happened to produce different-looking results?
+The second reason why uniformity and marginality assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
+The difficulty lies in the high-dimensional and stochastic nature of network behaviors.
+Suppose, for example, that we assume that packet sizes and inter-packet intervals are drawn independently from some given distribution for each flow.
+In the terminology introduced above, this is a flow-conditional packet behavior model.
+Prior research has indicated that, given accurate per-flow distributions, this is a sufficiently realistic model of network behavior~\cite{Karpinski07:realism}.\realismnote
+In this model, however, each flow must have its own specific, realistic pair of distributions for packet size and inter-packet intervals.
+Moreover, the distribution of distributions across the entire network must simultaneously be realistic.
+Modeling such ``distributions of distributions'' realistically for entire networks is currently an unsolved problem.
+This paper provides the mathematical framework in which to address this problem and demonstrates the effectiveness of this framework by taking the first steps towards fully understanding the complex interdependencies of real network behavior.
+% make a major first step towards being able to model the distribution of flow behaviors across an entire network, by providing the framework in which to address this problem mathematically, and by demonstrating that this framework actually allows us to meaningfully analyze non-uniform, non-independent cross-network behavior.
 

### 2008-06-04 paper.tex
Yet another intro rewrite.

@@
 Moreover, the distribution of distributions across the entire network must simultaneously be realistic.
-Modeling such ``distributions of distributions'' realistically for entire networks is currently an unsolved problem.
-This paper provides the mathematical framework in which to address this problem and demonstrates the effectiveness of this framework by taking the first steps towards fully understanding the complex interdependencies of real network behavior.
+Modeling such ``distributions of distributions'' realistically for entire networks is currently an unsolved problem, which we begin to address here.
+% This paper provides the mathematical framework in which to address this problem and demonstrates the effectiveness of this framework by taking the first steps towards fully understanding the complex interdependencies of real network behavior.
 % make a major first step towards being able to model the distribution of flow behaviors across an entire network, by providing the framework in which to address this problem mathematically, and by demonstrating that this framework actually allows us to meaningfully analyze non-uniform, non-independent cross-network behavior.
 
-Uniform and independent modeling reduce the dimensionality of network behavior drastically, making analysis and generation tractable problems.
-There are two immediate questions:
+Fundamentally, uniform, marginal and conditional models are applied to reduce the dimensionality of network behavior drastically, making analysis and generation tractable problems.
+From this perspective, two questions immediately present themselves:
 \begin{enumerate}
-\item Can we represent the original ``unreduced'' network behavior in such a way that analysis and generation are still tractable problems?
+\item Can we represent ``unreduced'' network behavior in such a way that analysis and generation are still tractable problems?
 \item Is there a better way to reduce dimensionality than assuming independence or uniformity of behaviors?
 \end{enumerate}
-The second contribution of this paper is to answer these questions and to propose a mathematical representation of network traffic that preserves the non-uniformity and non-independence of flow behaviors in a network, but makes traffic analysis and workload generation from this representation tractable problems.
-Here we will briefly explain our approach.
-
-It is already common practice in network modeling to think of flow behaviors as vectors.
-The \caps{IP} protocol number, source and destination \caps{IP} addresses, and \caps{UDP/TCP} source and port numbers that define a packet flow are commonly referred to as a 5-tuple---i.e., a vector in five dimensions.
-Other aspects of a flow's behavior can be represented vectorially as well.
-We call such a mapping of flows into a vector space a \emph{linear representation} if it satisfies the following simple linearity criterion:
+This paper answers both of these questions.
+To address the first question, we propose the concept of a \emph{linear representation} of network traffic.
+A linear representation of traffic represents each flow's behavior as a vector such that the following linearity condition is satisfied:
 \begin{quote}
@@
 \end{quote}
-We show that flows can be represented satisfying this criterion. This immediately allows us to perform useful computations on traffic patterns through simple vector addition and scaling.
-Finally, we argue that the natural generalization of this approach is to represent an entire network's behavior as a large matrix, with each row representing a single flow.
-The matrix representation gives a representation of network behavior that preserves the non-uniformity and non-independence of flow behaviors, but still is highly amenable to analysis using theoretical and numerical linear algebra.
-Moreover, we describe how to generate traffic from arbitrary traffic matrices, and demonstrate experimentally that the specific matrix representation we propose preserves the performance characteristics of real traffic.
+A complete traffic sample is expressed as a matrix of such vectors.
+Linear representations express full network behavior, without imposing assumptions of uniformity or marginality, yet allow effective traffic analysis and workload generation.
+We present a specific linear representation, called the general matrix model (\caps{GMM}), and demonstrate experimentally that in wireless networks this model generates workload that accurately reproduces the performance characteristics of real wireless network traffic.
+
+We address the second question in two parts.
+First, we observe that in linear representations of traffic, uniform, marginal and conditional assumptions are simply assertions about rows and columns of the representation matrix being identical to each other.
+Moreover, real traffic matrices can be transformed into the closest uniform, marginal or conditional approximation through simple matrix multiplications.
+Viewed this way, the cost of these assumptions becomes immediately apparent: these transformations discard vast amounts of information from the original trace.
+It is this wholesale destruction information that causes uniform and marginal traffic models to drastically distort the performance characteristics of real traffic.
+
+To show how traffic can be modeled less destructively, 
+% To demonstrate how linear representation allows drastically improved analysis and modeling of network traffic,
+we use non-negative matrix factorization to find an low-dimensional approximation of the packet size and inter-packet behavior matrices.
+% As an example of how the linear representation allows drastically improved analysis and modeling of network traffic, we apply non-negative matrix factorization to find an low-dimensional representation of packet size and inter-packet flow behaviors.
+% Each flow's behavior can be explained as a linear combination of only twelve ``basic behaviors.'' 
+% Although they are found purely algorithmically from packet size and inter-packet interval distributions, these behaviors correspond to intuitive mode of network activity.
+This factorization simultaneously extracts a small set of ``basic behaviors'' and explains each flow's observed behavior as a mixture of these basic behaviors.
+Although the basic behaviors are found algorithmically, they correspond very well with intuitive network activities such as file transfer, typing via \caps{SSH}, and pinging.
+
+% Via this factorization, each flow's behavior as a linear combination a handful of ``basic behaviors'' such as file transfer, typing via \caps{SSH}, ping traffic and other network activities. 
+% These intuitive behavior modes as well as the explanation of flows in terms of them are extracted from trace date purely algorithmically.
+
+% First we observe that in any linear representation of traffic, the assumption of uniformity corresponds the criterion that the coefficients in each dimension be the same.
+
+% To address the second question, we observe that in the general matrix model, 
+% 
+% Moreover, we apply state of the art matrix factorization techniques to this representation to explain cross-network flow behaviors as linear mixtures of only twelve basic behaviors.
+
+% To address the second question,
+
+% The second contribution of this paper is to answer these questions and to propose a mathematical representation of network traffic that preserves the non-uniformity and non-independence of flow behaviors in a network, but makes traffic analysis and workload generation from this representation tractable problems.
+% Here we will briefly explain our approach.
+% 
+% It is already common practice in network modeling to think of flow behaviors as vectors.
+% The \caps{IP} protocol number, source and destination \caps{IP} addresses, and \caps{UDP/TCP} source and port numbers that define a packet flow are commonly referred to as a 5-tuple---i.e., a vector in five dimensions.
+% Other aspects of a flow's behavior can be represented vectorially as well.
+We call such a mapping of flows into a vector space a \emph{linear representation} if it satisfies the following simple linearity criterion:
+% \begin{quote}
+% The sum of the representations of two flows is a vector that represents the aggregate behavior of the two flows.
+% \end{quote}
+% We show that flows can be represented satisfying this criterion. This immediately allows us to perform useful computations on traffic patterns through simple vector addition and scaling.
+% Finally, we argue that the natural generalization of this approach is to represent an entire network's behavior as a large matrix, with each row representing a single flow.
+% The matrix representation gives a representation of network behavior that preserves the non-uniformity and non-independence of flow behaviors, but still is highly amenable to analysis using theoretical and numerical linear algebra.
+% Moreover, we describe how to generate traffic from arbitrary traffic matrices, and demonstrate experimentally that the specific matrix representation we propose preserves the performance characteristics of real traffic.
 
@@
 \begin{align}
-  \mean{z}&=\fracx{\inner{\vec{z}}{\vec{\sigma}}}{\inner{\vec{z}}{\ones{d_z}}},\\
-  \mean{v}&=\fracx{\inner{\vec{v}}{\vec{\tau}}}{\inner{\vec{v}}{\ones{d_v}}},
+  \mean{z}&=\frac{\inner{\vec{z}}{\vec{\sigma}}}{\inner{\vec{z}}{\ones{}}}, &
+  \mean{v}&=\frac{\inner{\vec{v}}{\vec{\tau}}}{\inner{\vec{v}}{\ones{}}},
 \end{align}

### 2008-06-04 paper.tex
More work on intro.
The intro, once more.

@@
 
+% The model of flow behavior implicit in non-negative matrix factoriation technique is that flow behavior is a composite of some limited set of basic behaviors---and that fractions of each flow ``belong'' to each behavior.
+This is not an \latin{a priori} unreasonable hypothesis.
+Contrast that with the assumptions underlying uniform and marginal simplifications of traffic behavior, which can be seen to be intuitively unreasonable without ever conducting and actual experiment or calculation.
+
 \keywords{
@@
   linear algebra \and
-  matrix analysis
+  non-negative matrix factorization
 }
@@
 \begin{enumerate}
-\item \textbf{Uniform:} that traffic behaviors are uniformly distributed. For example, assuming that  all nodes in a network have equal probability of being the source or destination of each flows.
+\item \textbf{Uniform:} that traffic properties are uniformly distributed. For example, assuming that  all nodes in a network have equal probability of being the source or destination of each flows.
 Another common example is the assumption that all packet sizes and inter-packet intervals are the same; this is commonly known as the ``constant bit-rate'' (\caps{CBR}) model.
@@
 \item \textbf{Conditional:} that traffic behaves marginally, but only within equivalence classes defined by certain conditions.
-The assumption, for example, that all flows emanating from a common source node share the same packet size distribution would be a conditional model, conditioned on source nodes.
+The assumption, for example, that all flows emanating from a common source node share the same packet size distribution is a conditional model, conditioned on source nodes.
 \end{enumerate}
@@
 They clearly are not: flows have drastically different packet size distributions; different nodes initiate and receive drastically different numbers of packets and flows.
-Simply considering intuitive examples of common behavior illustrate this:
+Simply considering intuitive examples of common behavior illustrates this amply:
 downloading a large file \latin{vs.} typing in an \caps{SSH} session;
@@
 In this model, however, each flow must have its own specific, realistic pair of distributions for packet size and inter-packet intervals.
-Moreover, the distribution of distributions across the entire network must simultaneously be realistic.
-Modeling such ``distributions of distributions'' realistically for entire networks is currently an unsolved problem, which we begin to address here.
+Moreover, the ``distribution of distributions'' across the entire network must simultaneously be realistic.
+Modeling such distributions of distributions realistically for entire networks is currently an unsolved problem; we provide the means to address it.
 % This paper provides the mathematical framework in which to address this problem and demonstrates the effectiveness of this framework by taking the first steps towards fully understanding the complex interdependencies of real network behavior.
@@
 
-Fundamentally, uniform, marginal and conditional models are applied to reduce the dimensionality of network behavior drastically, making analysis and generation tractable problems.
+Fundamentally, uniform, marginal and conditional models are applied to reduce the dimensionality of network behaviors drastically, making analysis and generation tractable problems.
 From this perspective, two questions immediately present themselves:
 \begin{enumerate}
-\item Can we represent ``unreduced'' network behavior in such a way that analysis and generation are still tractable problems?
+\item Can we represent ``unreduced'' network behavior in such a manner that analysis and generation are still tractable problems?
 \item Is there a better way to reduce dimensionality than assuming independence or uniformity of behaviors?
@@
 To address the first question, we propose the concept of a \emph{linear representation} of network traffic.
-A linear representation of traffic represents each flow's behavior as a vector such that the following linearity condition is satisfied:
+A linear representation of traffic is a vector-valued function on flows such that the following linearity condition is satisfied:
 \begin{quote}
@@
 \end{quote}
-A complete traffic sample is expressed as a matrix of such vectors.
+The complete behavior of a network is expressed as a matrix of such vectors, having one row per flow.
 Linear representations express full network behavior, without imposing assumptions of uniformity or marginality, yet allow effective traffic analysis and workload generation.
@@
 
-We address the second question in two parts.
-First, we observe that in linear representations of traffic, uniform, marginal and conditional assumptions are simply assertions about rows and columns of the representation matrix being identical to each other.
-Moreover, real traffic matrices can be transformed into the closest uniform, marginal or conditional approximation through simple matrix multiplications.
-Viewed this way, the cost of these assumptions becomes immediately apparent: these transformations discard vast amounts of information from the original trace.
-It is this wholesale destruction information that causes uniform and marginal traffic models to drastically distort the performance characteristics of real traffic.
-
-To show how traffic can be modeled less destructively, 
+\newfootnote{\ranknote}{For uniform and marginal assumptions, the multiplication matrix has rank 1  since all rows and columns are identical.
+In the conditional case, the rank of the matrix is the number of conditioning classes.}
+
+To address the question of whether there are better ways to reduce the dimensionality of network behaviors, we first make a simple observation.
+In linear representations of traffic, uniform, marginal and conditional assumptions are assertions about rows and columns of the representation matrix being identical. 
+Moreover, real traffic matrices can be transformed into uniform, marginal or conditional forms through  very simple matrix multiplications.
+These transformations, however, are extremely destructive: the matrices are degenerate with very low rank,\ranknote so multiplication destroys almost all information in the original traffic matrix.
+
+\input{trace_vs_uniform}
+
+We argue that matrix factorization, which is nondestructive, is a far superior way to reduce the dimensionality of network traffic behaviors.
+% To show how traffic can be modeled less destructively, 
 % To demonstrate how linear representation allows drastically improved analysis and modeling of network traffic,
-we use non-negative matrix factorization to find an low-dimensional approximation of the packet size and inter-packet behavior matrices.
+To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF}) to find a low-dimensional approximation of the packet size and inter-packet interval behavior matrices.
 % As an example of how the linear representation allows drastically improved analysis and modeling of network traffic, we apply non-negative matrix factorization to find an low-dimensional representation of packet size and inter-packet flow behaviors.
@@
 % Although they are found purely algorithmically from packet size and inter-packet interval distributions, these behaviors correspond to intuitive mode of network activity.
-This factorization simultaneously extracts a small set of ``basic behaviors'' and explains each flow's observed behavior as a mixture of these basic behaviors.
-Although the basic behaviors are found algorithmically, they correspond very well with intuitive network activities such as file transfer, typing via \caps{SSH}, and pinging.
+This factorization extracts a small set of ``basic behaviors'' from a traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
+Although the basic behaviors are found algorithmically, they correspond to intuitive network activities such as file transfer, typing via \caps{SSH}, web surfing, and ping traffic.
 
@@
 
-\input{trace_vs_uniform}
-
 % TODO: update roadmap paragraph.
@@
 \Figure{trace-vs-uniform} demonstrates how drastically different network traffic appears when assumptions of uniform behavior at various levels are imposed.
-Assumptions of independence are not as visually dramatic, but we will demonstrate that they distort the performance characteristics of networks almost as much---or more---than uniform models.
+Assumptions of independence are not as visually dramatic, but we will demonstrate that they distort the performance characteristics of networks almost as much or more than uniform models.
 

### 2008-06-04 paper.tex
Work on related work & motivation.

@@
 \newcommand{\RFC}[1]{\caps{RFC}~{#1}}
+\newcommand{\tmix}{T\caps{MIX}}
 
@@
 Moreover, real traffic matrices can be transformed into uniform, marginal or conditional forms through  very simple matrix multiplications.
-These transformations, however, are extremely destructive: the matrices are degenerate with very low rank,\ranknote so multiplication destroys almost all information in the original traffic matrix.
+These transformations, however, are extremely destructive: the matrices are degenerate with very low rank, so multiplication destroys almost all information in the original traffic matrix.
 
@@
 
-We argue that matrix factorization, which is nondestructive, is a far superior way to reduce the dimensionality of network traffic behaviors.
+We argue that matrix factorization, which is nondestructive, is a far superior to multiplication as a way to reduce the dimensionality of network traffic behaviors.
 % To show how traffic can be modeled less destructively, 
@@
 This factorization extracts a small set of ``basic behaviors'' from a traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
-Although the basic behaviors are found algorithmically, they correspond to intuitive network activities such as file transfer, typing via \caps{SSH}, web surfing, and ping traffic.
+Although the basic behaviors are found without examining port numbers or protocols, they nevertheless correspond to intuitive network activities such as file transfer, typing via \caps{SSH}, web surfing, and ping traffic.
 
@@
 In \Section{motivation} we discuss motivation and related work.
-The General Matrix Model is presented in \Section{general-matrix-model}.
+The notion of linear representation and the general matrix model are presented in \Section{general-matrix-model}.
 In Sections~\ref{sec:common-properties}~and~\ref{sec:modeling-concepts} we explore how common network properties and modeling concepts can be expressed in the the \caps{GMM} framework.
@@
 This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner~\cite{Clark88}.\QoSnote The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic.
-\Figure{trace-vs-uniform} demonstrates how drastically different network traffic appears when assumptions of uniform behavior at various levels are imposed.
-Assumptions of independence are not as visually dramatic, but we will demonstrate that they distort the performance characteristics of networks almost as much or more than uniform models.
+\Figure{trace-vs-uniform} demonstrates how drastically different network traffic appears under uniformity assumptions at the node, flow and packet level of behavior.
+Marginality assumptions are not as visually dramatic, but we will demonstrate that they distort the performance characteristics of networks almost as much, and sometimes more, than uniform models.
 
 Paxson and Floyd observed~\cite{Paxson95} that the interplay between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback.
-Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}.
-This presents a fundamental difficulty, however, in that it presumes that we know the intent of endpoints: what \textit{would} they have done under different conditions?
+Traffic models typically attempt to preserve the closed-loop behavior of network traffic.
+Characterizing and modeling the feedback between the network and applications using \caps{TCP} is precisely the subject of Hern\'andez-Campos's dissertation~\cite{Hernandez06:dissertation}.
+Hern\'andez-Campos's addresses in great depth how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how most common applications \emph{would} react in response to other network conditions.
+This research has subsequently been turned into a \caps{TCP} flow generation tool for \caps{NS-2}, called \tmix~\cite{Weigle06}.
+It is important to note that {\tmix} does not provide models for application behavior or for mixtures of applications in networks.
+Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
+Our work complements this perfectly: it provides the means to analyze traffic and generate realistic multi-flow application workloads which can then serve as the necessary input to \tmix.
+
+% This presents a fundamental difficulty, however, in that it presumes that we know the intent of endpoints: what \textit{would} they have done under different conditions?
 While we can speculate about what an individual node might hypothetically do, we currently hardly understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations.
-A fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback.
-Accordingly, in this paper we attempt to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without multi-level feedback.
+A fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of application and user feedback.
+
+% Although we do not attempt to model application and user feedback, we do simulate the effect of the network layer and below on the dynamics of \caps{TCP} for flows that were originally transmitted using \caps{TCP}.
+Hern\'andez-Campos~\cite{Hernandez06:dissertation} addresses in great depth how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how many common applications would react in response to network conditions.
+This research has subsequently been turned into a \caps{TCP} flow generation tool for \caps{NS-2}, called \tmix~\cite{Weigle06}.
+This work is complementary to ours: {\tmix} does not provide a model for application behaviors; it only provides the ability to reproduce accurate closed-loop dynamics given an accurate application behavior model.
+Our work our work provides the means to analyze traffic and generate realistic application workloads which can then serve as the required inputs to \tmix.
+
+% This paper attempts to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without feedback from the network to applications and users.
 It is important to realize that while this does not provide a final picture, we currently lack even a first-order understanding of the effect of different workloads on performance.
@@
 They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether the make stipulations about uniformity within each flow, across all flows, or within groups of flows.
-In the following sections, we explore how various modeling concepts can be expressed in the framework of the General Matrix Model.
+In the following sections, we explore how various modeling concepts can be expressed in the framework of the general matrix model.
 Models are expressed as transformations that will convert arbitrary traffic matrix instances into similar ones that satisfy the model's assumptions.
@@
 
-%How do marginal models translate into the algebraic language of the General Matrix Model? Whereas uniformity assumptions for nodes,
+%How do marginal models translate into the algebraic language of the general matrix model? Whereas uniformity assumptions for nodes,
 
@@
 
-The first model we evaluate is the General Matrix Model itself.
+The first model we evaluate is the general matrix model itself.
 We derive the matrix representation directly from the trace data and use it to generate workload as described in \Section{traffic-generation}.
@@
 %The original trace traffic serves as a control group against which to compare all other models.
-Before we can evaluate the General Matrix Model (\caps{GMM}), we must establish that one potentially important aspect of traffic behavior that is \textit{not} represented in the \caps{GMM} can legitimately be discarded.
+Before we can evaluate the general matrix model (\caps{GMM}), we must establish that one potentially important aspect of traffic behavior that is \textit{not} represented in the \caps{GMM} can legitimately be discarded.
 That aspect of behavior is \textit{\caps{TCP} feedback}.
@@
 
-Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of trace-based models and synthetic traffic models defined using the General Matrix Model.
+Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of trace-based models and synthetic traffic models defined using the general matrix model.
 We use a 24-hour trace recorded in an infrastructured 802.11g wireless \caps{LAN} with 18 access points, deployed at the 60th Internet Engineering Task Force meeting (\caps{IETF}60), held in San Diego during August of 2004.
@@
 Therefore, the research community needs models whose parameters can be adjusted independently, but whose performance characteristics can be made to resemble whatever example of network behavior one might need.
-The General Matrix Model does not yet fulfill this role since its parameters cannot be freely varied.
+The general matrix model does not yet fulfill this role since its parameters cannot be freely varied.
 We believe, however, that \caps{GMM} plays a vital role in the ongoing process of developing  of better models.
@@
 We have presented a fundamentally new, algebraic way of representing traffic patterns in local area networks.
-We call this representation the General Matrix Model.
+We call this representation the general matrix model.
 Each flow of a traffic collection is represented as a single high-dimensional vector with components representing the \caps{IP} protocol type, source and destination nodes, start time, flow size, and packet behavior.
 The essential property of the algebraic representation is that standard vector operations compute natural and useful descriptions of aggregate behavior over the collection of flows represented.
-Our experimental validation demonstrates clearly that the General Matrix Model accurately reproduces the performance characteristics of of real traffic.
+Our experimental validation demonstrates clearly that the general matrix model accurately reproduces the performance characteristics of of real traffic.
 The benefits of a performance-preserving algebraic representation are multifold.
@@
 Our experimental results bolster this conclusion: both uniform and marginal models significantly misrepresent many important network performance metrics.
-With the General Matrix Model, however, we stand poised to use the powerful tools of modern linear algebra to develop general, elegant and powerful models of network behavior that just work.
+With the general matrix model, however, we stand poised to use the powerful tools of modern linear algebra to develop general, elegant and powerful models of network behavior that just work.
 

### 2008-06-04 trace_vs_uniform.tex
Work on related work & motivation.

@@
+%!TEX root = /Users/stefan/projects/papers/paper.4/paper.tex
 \begin{figure*}[!t]
 \begin{center}
-\subfloat[node trace]
+\subfloat[node real]
 {\includegraphics[width=1.335in]{flow_topology_trace}\label{fig:flow-topology-trace}}
@@
 {\includegraphics[width=1.335in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
-\subfloat[flow trace]
+\subfloat[flow real]
 {\includegraphics[width=1.335in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
@@
 {\includegraphics[width=1.335in]{flow_packet_behavior}\label{fig:packet-behavior}}
-\caption{Examples illustrating trace (real) versus uniform (na\"ive synthetic) behaviors at the node, flow and packet levels.
+\caption{Examples illustrating real versus uniform behaviors at the node, flow and packet levels.
 Figures \ref{fig:flow-topology-trace} and \ref{fig:flow-topology-uniform} show example node behaviors.
-The width of each line is proportional to the logarithm of the number of flows between the nodes it connects (zero is the Internet gateway).
+The width of each line is proportional to the logarithm of the number of flows between the nodes (zero is the Internet gateway).
 Uniform and trace flow behavior examples are plotted in Figures \ref{fig:flow-behavior-trace} and \ref{fig:flow-behavior-uniform}.

### 2008-06-05 paper.tex
Overhauling Related Work: Harpoon.

@@
 Modeling such distributions of distributions realistically for entire networks is currently an unsolved problem; we provide the means to address it.
-% This paper provides the mathematical framework in which to address this problem and demonstrates the effectiveness of this framework by taking the first steps towards fully understanding the complex interdependencies of real network behavior.
-% make a major first step towards being able to model the distribution of flow behaviors across an entire network, by providing the framework in which to address this problem mathematically, and by demonstrating that this framework actually allows us to meaningfully analyze non-uniform, non-independent cross-network behavior.
 
@@
 We argue that matrix factorization, which is nondestructive, is a far superior to multiplication as a way to reduce the dimensionality of network traffic behaviors.
-% To show how traffic can be modeled less destructively, 
-% To demonstrate how linear representation allows drastically improved analysis and modeling of network traffic,
 To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF}) to find a low-dimensional approximation of the packet size and inter-packet interval behavior matrices.
-% As an example of how the linear representation allows drastically improved analysis and modeling of network traffic, we apply non-negative matrix factorization to find an low-dimensional representation of packet size and inter-packet flow behaviors.
-% Each flow's behavior can be explained as a linear combination of only twelve ``basic behaviors.'' 
-% Although they are found purely algorithmically from packet size and inter-packet interval distributions, these behaviors correspond to intuitive mode of network activity.
 This factorization extracts a small set of ``basic behaviors'' from a traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
@@
 
-Paxson and Floyd observed~\cite{Paxson95} that the interplay between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback.
+\subsection{Traffic Generation}
+
+Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the exhibited behavior is governed by a closed feedback loop.
 Traffic models typically attempt to preserve the closed-loop behavior of network traffic.
-Characterizing and modeling the feedback between the network and applications using \caps{TCP} is precisely the subject of Hern\'andez-Campos's dissertation~\cite{Hernandez06:dissertation}.
-Hern\'andez-Campos's addresses in great depth how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how most common applications \emph{would} react in response to other network conditions.
-This research has subsequently been turned into a \caps{TCP} flow generation tool for \caps{NS-2}, called \tmix~\cite{Weigle06}.
-It is important to note that {\tmix} does not provide models for application behavior or for mixtures of applications in networks.
+In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
+Hern\'andez-Campos's dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and applications for most \caps{TCP} applications.
+Hern\'andez-Campos shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how most common applications would react in response to different network conditions.
+This research has subsequently been turned into a \caps{TCP} flow generation tool called \tmix~\cite{Weigle06}.
+It is important to note that {\tmix} does not provide models for application behaviors or for mixtures of applications in networks.
 Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
-Our work complements this perfectly: it provides the means to analyze traffic and generate realistic multi-flow application workloads which can then serve as the necessary input to \tmix.
+Our work complements this perfectly: this paper provides the means to analyze traffic and generate realistic whole-network application workloads which can then serve as the necessary input to \tmix.
 
@@
 
-There is a large and diverse body of work on traffic analysis, modeling, and generation~\cite{Paxson95,Paxson96,Sommers04,Avallone04,Hernandez06:dissertation}.
+% There is a large and diverse body of work on traffic analysis, modeling, and generation~\cite{Paxson95,Paxson96,Sommers04,Avallone04,Hernandez06:dissertation}.
 We are only able to discuss a small, but hopefully representative sampling of this work.
@@
 %Because of the comparatively unlimited capacity of wired networks, traffic workload generally does not offer interesting technical challenges at the edge of the network.
-The two most prominent traffic generation frameworks are Harpoon and \caps{D-ITG}.
+The two most prominent general traffic generation frameworks are Harpoon and \caps{D-ITG}.
 Harpoon~\cite{Sommers04} uses a traffic trace for self-training, and can subsequently generate synthetic traffic with certain statistical properties based on the original trace.
-The properties reproduced are the empirical distributions of the following: ``file size, inter-connection time, source and destination IP ranges, number of active sessions.'' There is no criterion proposed to determine whether these properties characterize the original traffic adequately---we can only hope that this approximation is good enough.
+The properties reproduced are the empirical distributions of the following: ``file size, inter-connection time, source and destination IP ranges, number of active sessions.''
+Harpoon exemplifies the uniform and marginal modeling approaches.
+Source and destination nodes are treated marginally: an empirically derived marginal frequencies of sources and destination nodes (independently) are used to choose endpoints for each flow.
+File sizes (\latin{i.e.} total flow sizes) are likewise treated marginally: each flow's total byte  count is sampled from a marginal distribution of flow sizes.
+\caps{UDP} flows are modeled as being constant bit-rate (\latin{i.e.} having uniform packet behavior), and \caps{TCP} flows are all treated as file transfers, making them effectively uniform with bandwidth limited by simulated \caps{TCP} dynamics.
+Inter-connection times (\latin{i.e.} per node-pair inter-flow intervals) are sampled from a single empirically derived marginal distribution.
+While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic, it is difficult to know this:
+generated traffic is only compared with trace traffic using metrics that are explicitly specified in the models.
+Realism is never double-checked using derived statistical metrics or actual performance metrics.
+For the purpose of generating local-area traffic for wireless experiments, it is certainly not sufficiently realistic.
+Previous research~\cite{Karpinski07:cbr-failure} has shown that \caps{CBR} packet behavior models, like those used in Harpoon, drastically distort vital performance metrics at all levels of the protocol stack in wireless network simulations.
+This paper gives equal cause for concern about marginal traffic models.
+
+% There is no criterion proposed to determine whether these properties characterize the original traffic adequately---we can only hope that this approximation is good enough.
 For many purposes, it likely is sufficient; in particular, it is probably appropriate for the intended use in generating traffic for Internet backbone simulations.
@@
 Clearly, hybridizing the regularization here is not an effective technique.
-We believe that the cause of this effect is like mixing too many colors yielding an unpleasant shade of brown: any of the colors individually would be better than the non-descript average.
+We believe that the cause of this effect is like mixing too many colors yielding an unpleasant shade of brown: any of the colors individually would be better than the nondescript average.
 As a more general conclusion, arbitrarily mixing model may not always be a good idea, despite the ease with which it is done in the matrix representation of workload behavior.

### 2008-06-05 paper.tex
More work on Related Work.

@@
 Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
-One of the major challenges of modeling full-network traffic is the multilevel interdependence of packet, flow\flownote and node behaviors.
+One of the major challenges of modeling whole-network traffic is the interdependence of packet, flow\flownote and node behaviors.
 Three types of simplifying assumptions are commonly applied in both analysis and generation:
 \begin{enumerate}
-\item \textbf{Uniform:} that traffic properties are uniformly distributed. For example, assuming that  all nodes in a network have equal probability of being the source or destination of each flows.
+\item \textbf{Uniform:} that traffic properties are uniformly distributed. For example, assuming that  all nodes in a network have equal probability of being the source or destination of each flow.
 Another common example is the assumption that all packet sizes and inter-packet intervals are the same; this is commonly known as the ``constant bit-rate'' (\caps{CBR}) model.
@@
 
-\subsection{Traffic Generation}
+While this work is not inherently specific to wireless networks, since the modeling techniques developed here equally applicable to traffic in wired networks, the results are of fundamental importance to wireless research.
+Realistic traffic models have a drastic impact on the experimental results in wireless networks~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+This is primarily because wireless networks are locally resource constrained.
+Compare this with the typical situation in wired \caps{LAN}s, where gigabit Ethernet and modern routers and switches can easily handle all but the most intense levels of traffic.
+If a wired network operator expects larger volumes of traffic on a regular basis, the solution is to upgrade the hardware, not to change the network protocols.
+In wireless networks, on the other hand, the physical medium cannot be upgraded: protocols must be improved to make better usage of the medium.
+Being able to perform experiments with realistic workload is a vital part of the ongoing process of improving wireless protocols.
+Without realistic workload models for experiments, performance comparisons may be misleading or even completely wrong~\cite{Karpinski07:cbr-failure}.% Better traffic models are necessary to avoid stumbling in the dark while attempting to improve wireless technologies.
 
 Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the exhibited behavior is governed by a closed feedback loop.
-Traffic models typically attempt to preserve the closed-loop behavior of network traffic.
+% Traffic models typically attempt to preserve the closed-loop behavior of network traffic.
 In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
-Hern\'andez-Campos's dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and applications for most \caps{TCP} applications.
-Hern\'andez-Campos shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how most common applications would react in response to different network conditions.
+Hern\'andez-Campos' dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and applications for typical \caps{TCP} applications.
+Hern\'andez-Campos shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
 This research has subsequently been turned into a \caps{TCP} flow generation tool called \tmix~\cite{Weigle06}.
-It is important to note that {\tmix} does not provide models for application behaviors or for mixtures of applications in networks.
+The acronym is somewhat misleading: {\tmix} does not provide models for application behaviors or for mixtures of applications in networks.
 Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
-Our work complements this perfectly: this paper provides the means to analyze traffic and generate realistic whole-network application workloads which can then serve as the necessary input to \tmix.
+Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary input to \tmix.
 
@@
 %Because of the comparatively unlimited capacity of wired networks, traffic workload generally does not offer interesting technical challenges at the edge of the network.
+
+\newfootnote{\harpoonnote}{Even this is unclear: generated traffic is only compared with trace traffic using the very same metrics that are explicitly specified as part of the model.
+Unsurprisingly, the distributions of sampled metrics closely match the distributions those metrics were sampled from.
+Realism is not double-checked using other statistical metrics or actual performance metrics.}
+
 The two most prominent general traffic generation frameworks are Harpoon and \caps{D-ITG}.
@@
 Harpoon exemplifies the uniform and marginal modeling approaches.
-Source and destination nodes are treated marginally: an empirically derived marginal frequencies of sources and destination nodes (independently) are used to choose endpoints for each flow.
+Source and destination nodes are treated marginally: empirically derived marginal frequencies of sources and destination nodes (independently) are used to choose endpoints for each flow.
 File sizes (\latin{i.e.} total flow sizes) are likewise treated marginally: each flow's total byte  count is sampled from a marginal distribution of flow sizes.
-\caps{UDP} flows are modeled as being constant bit-rate (\latin{i.e.} having uniform packet behavior), and \caps{TCP} flows are all treated as file transfers, making them effectively uniform with bandwidth limited by simulated \caps{TCP} dynamics.
+\caps{UDP} flows are modeled as being constant bit-rate (\latin{i.e.} having uniform packet behavior), and \caps{TCP} flows are all treated as file transfers, making them effectively uniformly behaved with bit-rate determined by \caps{TCP} dynamics.
 Inter-connection times (\latin{i.e.} per node-pair inter-flow intervals) are sampled from a single empirically derived marginal distribution.
-While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic, it is difficult to know this:
-generated traffic is only compared with trace traffic using metrics that are explicitly specified in the models.
-Realism is never double-checked using derived statistical metrics or actual performance metrics.
-For the purpose of generating local-area traffic for wireless experiments, it is certainly not sufficiently realistic.
-Previous research~\cite{Karpinski07:cbr-failure} has shown that \caps{CBR} packet behavior models, like those used in Harpoon, drastically distort vital performance metrics at all levels of the protocol stack in wireless network simulations.
-This paper gives equal cause for concern about marginal traffic models.
-
-% There is no criterion proposed to determine whether these properties characterize the original traffic adequately---we can only hope that this approximation is good enough.
-For many purposes, it likely is sufficient; in particular, it is probably appropriate for the intended use in generating traffic for Internet backbone simulations.
-Wireless networks, however, are particularly sensitive to workload conditions, and sampling from a limited set of empirical distributions does not suffice to reproduce realistic network-wide traffic.
-
-\caps{D-ITG}~\cite{Avallone04} generates flows using an independent sampling model for packet sizes and inter-packet intervals.
-The framework contains pre-made models for several common types of Internet traffic.
-The focus of this project, however, is on providing the infrastructure to generate very large volumes of synthetic Internet-like backbone traffic.
+While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic,\harpoonnote
+for the purpose of generating local-area traffic for wireless experiments, it is not sufficiently realistic.
+Previous research~\cite{Karpinski07:cbr-failure} has shown that \caps{CBR} packet behavior models used in Harpoon drastically distort vital performance metrics at all levels of the protocol stack in wireless network simulations.
+This paper shows that marginal traffic models also severely misrepresent performance metrics.
+
+Like Harpoon, \caps{D-ITG}~\cite{Avallone04} exemplifies the common uniform and marginal modeling assumptions.
+Packet behavior is modeled using a handful of standard statistical distributions (``constant, uniform, normal, cauchy, pareto, exponential, etc.'') for packet size and inter-packet interval distributions, from which packets are generated using repeated independent sampling.
+Guidance is not provided, however, for choosing among these distributions, choosing parameter values for individual flows, or selecting a mixture of distributions for flows across an entire network.
+Moreover, the mechanism for choosing end-points and start and end times for flows is not described; presumably it is explicitly specified by the user. 
+The focus of this project is primarily on the engineering challenge of generating very large volumes of synthetic Internet-like backbone traffic and injecting this traffic into real networks in a distributed fashion.
+While \caps{D-ITG} provides the mechanism for generating large volumes of network traffic, it does not provide any real guidance as to what traffic to generate.
+
+% The focus of this project, however, is on providing the infrastructure to generate very large volumes of synthetic Internet-like backbone traffic.
 No analysis is provided for determining the realism of traffic mixes, or for choosing flow endpoints realistically.

### 2008-06-05 paper.tex
Added basic behaviors figure.
Added caption, updated behavior figures.

@@
 
+\begin{sidewaysfigure*}
+{\includegraphics[height=5.95in]{plots/basic_behaviors_1.pdf}}\hfill%
+{\includegraphics[height=5.95in]{plots/basic_behaviors_2.pdf}}
+\caption{%
+The ten most prevalent components of flow behavior as pairs of packet payload size and of inter-packet interval distributions.
+The distributions are shown as cumulative distribution functions: the $x$-axis represents payload size (in kilobytes) or inter-packet interval duration (in log-seconds, base 10), respectively;
+the $y$-axis indicates the probability a value occurring, less than or equal to the $x$ value.
+The behaviors are ordered in descending rank of how many packets they explain;
+to the left, each behavior's rank is indicated, together with the percentage of total packets associated with it, and the cumulative percentage of packets explained.
+To the right of each behavior is a pie chart showing the breakdown of associated packets by protocol, as determined by well-known port numbers.
+}
+\label{fig:basic-behaviors}
+\end{sidewaysfigure*}
+
 Our simulation results are summarized in \Figure{box-plots}.

### 2008-06-05 paper.tex
Added a FHC paragraph in related work.

@@
 Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
-Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary input to \tmix.
+Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary input to {\tmix}.
 
@@
 Inter-connection times (\latin{i.e.} per node-pair inter-flow intervals) are sampled from a single empirically derived marginal distribution.
-While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic,\harpoonnote
-for the purpose of generating local-area traffic for wireless experiments, it is not sufficiently realistic.
+While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic,\harpoonnote for the purpose of generating local-area traffic for wireless experiments, it is not.
 Previous research~\cite{Karpinski07:cbr-failure} has shown that \caps{CBR} packet behavior models used in Harpoon drastically distort vital performance metrics at all levels of the protocol stack in wireless network simulations.
@@
 While \caps{D-ITG} provides the mechanism for generating large volumes of network traffic, it does not provide any real guidance as to what traffic to generate.
+
+\FHC~\cite{Hernandez06:wlan-traffic} have defined the state of the art in parametric modeling of wireless traffic patterns.
+They use rigorous statistical methods to fit parametric distributions to a number of properties of wireless traffic instances:
+% They propose the following models:
+flow arrival processes fit a non-stationary Poisson arrival model; flow inter-arrival intervals (per session) fit a Lognormal distribution;
+flows per session fit a BiPareto model; flow sizes fit a BiPareto model.
+However, all the proposed models are marginal:
+a single statistical model applies to all flows and nodes (packet behavior is not considered).
+% One intuitively obvious heterogeneous nature of network traffic:
+Each node has a different number of flows assigned to it, but otherwise nodes are identical. 
+The flows associated with each node are statistically identical:
+flow sizes are drawn from a single marginal distribution, and nothing else distinguishes them.
+%, unconditional of their associated node.
+% This is a classic example of the limits of marginal modeling.
+Yet, intuitively it is clear that different nodes in a network have different tendencies for flow size, flow initiation, and packet behavior.
+Such interdependent behaviors cannot be captured using marginal modeling.
+In this paper, we demonstrate that this inability is not harmless:
+treating all flows and nodes statistically identically severely misrepresents important wireless performance metrics in experiments.
 
@@
 We have examined a broad variety of other wireless network performance metrics, and although we do not have room to present or discuss them here, the results shown are representative of the overall realism of the traffic models.
-%For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
-%Unfortunately, not all of the simulations were able to run to completion.
-In some scenarios, the number of nodes and flows caused Qualnet to simply fail.
-We were, however, able to get complete results for 72.5\% of the simulations, providing a clear view of overall trends.\!\footnote{That 27.5\% of simulations did not finish may seem like a cause for concern, but it does not in fact invalidate our results.
-The simulations that did not finish are those for which the number of flows and nodes overwhelmed the capacity of the simulator.
-Qualnet cannot simulate more than circa 650 nodes sending 35,000 flows.
-While it would be of great interest to examine results for these large-scale scenarios, they are essentially useless for comparison to other simulations studies: other studies do not simulate this quantity of nodes and flows precisely because even the best existing simulators do not support it.
-Once the simulators become more scalable and robust, our analysis can be extended to these scenarios as well.}
-
-\subsection{Measures of Error}
-\label{sec:measures-of-error}
-
-The simulations described in \Section{simulations} provide us with the raw data to compare performance metrics for synthetic traffic models with those for real traffic traces.
-To assess the realism of these models, however, we need a measure of how inaccurate the synthetic performance values are when compared to the real values.
-Let $x$ be the value of a performance metric using real traffic, and $y$ the value of the same metric using an alternate traffic model.
-Some common measures of error are the difference [$y-x$], the ratio [$y/x$], and the relative error [$(y-x)/x$], also commonly called the ``standard error.'' These are all reasonable measures of error; but which is most appropriate for assessing the realism of performance metrics?
-Instead of picking one arbitrarily, we will first consider the properties that an ideal error function should have, and then use those properties to determine the best measure of error.
-Moreover, we will show that the unique measure of error that exhibits these ideal properties is the  \textit{log-ratio} of metric values:
-\begin{align}
-\log(y/x)=\log(y)-\log(x).
-\end{align}
-In this discussion, $E(x,y)$ is a generic error function applied to the synthetic value, $y$, with respect to the real value, $x$.
-
-The first property that an error function should have is  \textit{insensitivity to common factors}. That is, if both values are scaled by the same constant, the error should be unaffected:
-\begin{align}
-\label{eqn:normalization-invariance}
-\forall\: x,y,c: E(x c,y c) &= E(x,y)
-\end{align}
-There are three major motivations for this requirement:
-\begin{enumerate}
-\item Changing units should not affect error values.
-\item Error values for ``large'' and ``small'' scenarios should be directly comparable.
-Scenarios with large $x$ values will naturally have larger raw differences between $x$ and $y$.
-This requirement allows scenarios of different scales to be compared fairly and without bias.
-\item Changing between metrics that differ by a known constant for each scenario should not affect error values.
-\end{enumerate}
-The last point is best illustrated by an example.
-Consider two closely related performance metrics: average throughput, $t$, and total bytes received, $r$.
-Suppose that there are $f$ flows in a given scenario with average duration, $d$.
-Since $t =r/fd$, the metrics $t$ and $r$ contain the same information---they differ only by a known constant in each simulation scenario.
-Equation~\ref{eqn:normalization-invariance} ensures that the errors of these metrics are the same:
-\begin{align}
-E(t_{\text{trc}},t_{\text{mod}})
-	= E\left({\frac{r_{\text{trc}}}{fd},\frac{r_{\text{mod}}}{fd}}\right)
-	= E(r_{\text{trc}},r_{\text{mod}}).
-\end{align}
-The difference measure does not satisfy Equation~\ref{eqn:normalization-invariance}, but the ratio, relative error, and log-ratio error measures all do.
-
-The second property that an ideal error function should have is  \textit{additivity of compounded errors}.
-If two independent causes of error each induce some factor of misrepresentation, then the combined error should be the sum of the errors caused by each factor separately:
-\begin{align}
-\label{eqn:linear-compounding}
-\forall\: x,c_1,c_2:
-	E(x,x c_1 c_2) &= E(x,x c_1) + E(x,x c_2).
-\end{align}
-This property allows us to compare error values meaningfully across different traffic models.
-%For example, if flow topology and packet behavior affected some performance metric independently with no interaction effects, we would expect that
-%\begin{align}
-%E(x^\TTT,x^\UTU) \approx E(x^\TTT,x^\TTU) + E(x^\TTT,x^\UTT).
-%\end{align}
-%If these two values differ significantly, there must be some interaction between the two levels of behavior that introduces more error than can be explained by each separately.
-Without the property of additivity given in Equation~\ref{eqn:linear-compounding}, such a comparison would not be possible or meaningful.
-
-Additivity of compounded errors also implies two desirable properties that are easily derived from Equation~\ref{eqn:linear-compounding}.
-It forces the error of an accurate representation to be zero: $E(x,x) = 0$.
-It also forces underestimation and overestimation to be treated symmetrically.
-The error of underestimating by some factor is opposite but equal to overestimating by the same factor:
-\begin{align}
-E(x,x/c) = - E(x,xc).
-\end{align}
-
-It is easily verified that the difference, ratio, and relative error measures do not satisfy \Equation{linear-compounding}, and the difference, as noted, does not satisfy \Equation{normalization-invariance}.
-The log-ratio is the only metric presented that satisfies both conditions.
-Moreover, it can be proved that $\log(y/x)$ is the \textit{only} differentiable function that satisfies both (up to a constant).
-In the Appendix, we present a proof of this claim.
-Throughout the rest of the paper, we use the log-ratio to measure the error of performance metrics.
-% TODO: should Appendix be capitalized?
-% TODO: hyperref the Appendix for PDF.
 
@@
 
-\section*{Appendix}
-\footnotesize
-
-\noindent
-\textbf{Theorem.} The unique differentiable function satisfying Eqs.~\ref{eqn:normalization-invariance} and \ref{eqn:linear-compounding} is $E(1,e) \ln(y/x)$.
-\textit{Proof:} Let $f(z) = E(1,z)$.
-Eq.~\ref{eqn:normalization-invariance} gives $E(x,y) = E(1,y/x) = f(y/x)$.
-Eq.~\ref{eqn:linear-compounding} gives: $f(z) = f(z/w) + f(w)$.
-Differentiation by $z$ yields $f'(z) = w^{-1} f'(z/w)$.
-In particular, if we choose $w=z$, we get $f'(z) = z^{-1} f'(1)$.
-Integration by $z$ gives:
-$f(z) = f'(1) \int z^{-1} dz + c = f'(1) \ln(z) + c$.
-By Eq.~\ref{eqn:linear-compounding}, $f(1)=0$, so $c=0$. Thus $f(e)=f'(1)\ln(e)=f'(1)$. We conclude that $f(z)=f(e)\ln(z)$, so $E(x,y) = f(y/x) = f(e)\ln(y/x) = E(1,e)\ln(y/x)$, as desired.\hfill%\QED
-
-%\vspace{1em}\noindent
-%\textbf{Theorem.}~[\textit{Lyapunov's Central Limit Theorem}] Let $\{R_k\}_{k=1}^{\infty}$ be a series of independent variables with $\E{R_k}=0$.
-Let $s_n^2 = \sum_{k=1}^n{\E{\abs{R_k^2}}}$ and $r_n^3 = \sum_{k=1}^n{\E{\abs{R_k^3}}}$.
-For each $n$, let $Z_n$ be the normalized mean of $\{R_k\}_{k=1}^n$: $Z_n = \sum_{k=1}^n{R_k/s_n}$.
-If $\lim_{n\to\infty} {r_n/s_n} = 0$, then $\lim_{n\to\infty}{Z_n} \sim \mathcal{N}(0,1)$ (the standard normal distribution).
-(See~\cite{Feller68} page 229.)
-
-%\vspace{1em}
-%To apply the LCLT to the series $R_k^\M=\log(X_k^\M/X_k^\TTT)$, we must show that under the null hypothesis, the assumptions of the theorem are satisfied by this series.
-First, the null hypothesis, implies that $\E{R_k}=0$.
-Variables from separate simulations are independent since they cannot influence each other's values.
-Formally, $\Pr\left(X_k\middle|X_j\right)=\Pr\left(X_k\right)$.
-Therefore the log-ratios are also independent for different $k$.
-The last requirement is that $\lim_{n\to\infty}{r_n/s_n}=0$.
-To verify this, we use the estimators $\hat{s}_n^2=\sum{\abs{R_k^2}}$, and $\hat{r}_n^3=\sum{\abs{R_k^3}}$.
-When $\hat{r}_n/\hat{s}_n$ are plotted on a log-log scatter plot, with $n$ increasing up to the number of simulations, they asymptotically approach a downwardly sloped line as $n$ grows.
-Thus $\lim_{n\to\infty}{\ln(\hat{r}_n/\hat{s}_n)/\ln(n)}=c<0$.
-This implies that $\lim_{n\to\infty}{\ln(\hat{r}_n/\hat{s}_n)}=-\infty$, and therefore $\lim_{n\to\infty}{r_n/s_n}=\lim_{n\to\infty}{\hat{r}_n/\hat{s}_n}=0$.
-This test for the convergence of $r_n/s_n$ is applied to each model and metric pair.
-%The asymptotic approach of the ratio $\hat{r}_n/\hat{s}_n$ to a downwardly sloped line is verified visually and by comparison to a fitted regression line.
-
-%\vfill
-
-%\pagebreak
 % Hernandez06:dissertation: ``complete methodology for reproducing the traffic observed on a network link in a closed-loop manner, and proposed a number of metrics for studying the realism of the generated traffic.''

### 2008-06-06 paper.tex
More work on intro & motivation.

@@
   workload generation \and
-  network simulation \and
+  wireless networks \and
+  wireless simulation \and
   linear algebra \and
@@
 
-\newfootnote{\marginalnote}{The term ``marginal'' refers to the margins of actuarial tables formerly used for statistical computations.
+\newfootnote{\marginalnote}{The statistical term ``marginal'' refers to the margins of actuarial tables formerly used for statistical computations.
 The rows and columns of the table represent possible values of two properties.
 Each entry in a table contains a count of the number of events falling into the joint category for that row and column.
-The margins contain sums of the rows or columns, thus giving the distribution of properties along each dimension \emph{independent} of each other.}
+The margins contain sums of the rows and columns, thus giving the unconditional distributions of each property.}
 
@@
 \item \textbf{Marginal:} that traffic properties share a single, common \emph{marginal}\marginalnote distribution across the entire network, and individual values are drawn independently and unconditionally from this shared distribution.
-For example, assuming that a single distribution of packet sizes describes all flows in a network, as compared to equipping each flow in the network with its own specific distribution.
-\item \textbf{Conditional:} that traffic behaves marginally, but only within equivalence classes defined by certain conditions.
-The assumption, for example, that all flows emanating from a common source node share the same packet size distribution is a conditional model, conditioned on source nodes.
+For example, assuming that a single distribution of packet sizes describes all flows in a network, as compared to equipping each flow with its own specific distribution of sizes.
+\item \textbf{Conditional:} that traffic behaves marginally within equivalence classes defined by certain conditions.
+The assumption, for example, that all flows emanating from a common source node share the same packet size distribution is a conditional model, conditioned on the source nodes.
 \end{enumerate}
 We would be arguing against a straw man if we implied that researchers who apply these assumptions actually believe them to be true.
-They clearly are not: flows have drastically different packet size distributions; different nodes initiate and receive drastically different numbers of packets and flows.
+They clearly are not:
+different nodes initiate and receive drastically different numbers of packets and flows;
+flows have drastically different packet size distributions.
 Simply considering intuitive examples of common behavior illustrates this amply:
-downloading a large file \latin{vs.} typing in an \caps{SSH} session;
-BitTorrent users \latin{vs.} occasional email checkers.
+BitTorrent users \latin{vs.} occasional email checkers;
+downloading a large file \latin{vs.} typing in an \caps{SSH} session.
 
@@
 Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to analyze and replicate?
-We answer this question definitively in the negative: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
+This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
 
@@
 Moreover, the ``distribution of distributions'' across the entire network must simultaneously be realistic.
-Modeling such distributions of distributions realistically for entire networks is currently an unsolved problem; we provide the means to address it.
-
-Fundamentally, uniform, marginal and conditional models are applied to reduce the dimensionality of network behaviors drastically, making analysis and generation tractable problems.
+Modeling such distributions of distributions realistically for entire networks is currently an unsolved problem; we begin to address it here.
+
+Fundamentally, uniform, marginal and conditional models are applied so that we can represent network traffic behavior using a few empirical or analytical statistical distributions.
+Reduction in representation size and complexity makes analysis and generation tractable problems.
 From this perspective, two questions immediately present themselves:
 \begin{enumerate}
-\item Can we represent ``unreduced'' network behavior in such a manner that analysis and generation are still tractable problems?
-\item Is there a better way to reduce dimensionality than assuming independence or uniformity of behaviors?
+\item Can we represent ``unreduced'' network behavior such that analysis and generation are tractable?
+\item Are there better ways to reduce the representation of network behavior?
 \end{enumerate}
@@
 To address the first question, we propose the concept of a \emph{linear representation} of network traffic.
-A linear representation of traffic is a vector-valued function on flows such that the following linearity condition is satisfied:
+A linear representation of traffic maps each flow to a vector, such that the following linearity condition is satisfied:
 \begin{quote}
@@
 The complete behavior of a network is expressed as a matrix of such vectors, having one row per flow.
-Linear representations express full network behavior, without imposing assumptions of uniformity or marginality, yet allow effective traffic analysis and workload generation.
-We present a specific linear representation, called the general matrix model (\caps{GMM}), and demonstrate experimentally that in wireless networks this model generates workload that accurately reproduces the performance characteristics of real wireless network traffic.
+Linear representations express full network behavior, without imposing assumptions of uniformity or marginality, yet still allow effective traffic analysis and workload generation.
+We present a specific linear representation, called the general matrix model (\caps{GMM}), and demonstrate experimentally that this model generates wireless workload that accurately reproduces the performance characteristics of real wireless network traffic.
 
@@
 
-To address the question of whether there are better ways to reduce the dimensionality of network behaviors, we first make a simple observation.
-In linear representations of traffic, uniform, marginal and conditional assumptions are assertions about rows and columns of the representation matrix being identical. 
+To address the question of whether there are better ways to reduce the representation of network behaviors, we first make a simple observation.
+In linear representations of traffic, uniform, marginal and conditional assumptions are assertions of equality between the rows and columns of the traffic matrix.
 Moreover, real traffic matrices can be transformed into uniform, marginal or conditional forms through  very simple matrix multiplications.
@@
 
-We argue that matrix factorization, which is nondestructive, is a far superior to multiplication as a way to reduce the dimensionality of network traffic behaviors.
-To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF}) to find a low-dimensional approximation of the packet size and inter-packet interval behavior matrices.
-This factorization extracts a small set of ``basic behaviors'' from a traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
-Although the basic behaviors are found without examining port numbers or protocols, they nevertheless correspond to intuitive network activities such as file transfer, typing via \caps{SSH}, web surfing, and ping traffic.
+We argue that matrix factorization, which is nondestructive, is far superior to multiplication for reducing the representation of network traffic behaviors.
+To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF}) to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
+This factorization extracts a small set of ``basic behaviors'' from the traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
+Although basic behaviors are computed without examining port numbers, they nevertheless correspond to the intuitive behaviors expected for protocols.
+Basic behaviors are found corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.
 
@@
 
-The interaction of network users and application behavior with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted.
-The joint pattern of traffic generation and mobility through time and space completely determines the effect of wireless usage on the lower levels of the network.
+The interaction of network users and applications with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted.
+The joint pattern of traffic generation and mobility through time and space completely determines the effect of network usage on the lower levels of the stack.
 This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner~\cite{Clark88}.\QoSnote The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic.
-\Figure{trace-vs-uniform} demonstrates how drastically different network traffic appears under uniformity assumptions at the node, flow and packet level of behavior.
-Marginality assumptions are not as visually dramatic, but we will demonstrate that they distort the performance characteristics of networks almost as much, and sometimes more, than uniform models.
-
-While this work is not inherently specific to wireless networks, since the modeling techniques developed here equally applicable to traffic in wired networks, the results are of fundamental importance to wireless research.
+\Figure{trace-vs-uniform} illustrates through examples how drastically different network traffic appears under uniformity assumptions at different levels of behavior.
+Marginality assumptions are not as visually striking, but as we will demonstrate, they distort the performance characteristics of networks almost as much, and sometimes more, than uniform models.
+
+While this work is not inherently specific to wireless networks, since the modeling techniques developed are equally applicable to traffic in wired networks, the results are of specific importance to wireless research.
 Realistic traffic models have a drastic impact on the experimental results in wireless networks~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
@@
 If a wired network operator expects larger volumes of traffic on a regular basis, the solution is to upgrade the hardware, not to change the network protocols.
-In wireless networks, on the other hand, the physical medium cannot be upgraded: protocols must be improved to make better usage of the medium.
+In wireless networks, on the other hand, the physical medium cannot be upgraded; protocols must be improved to make better usage of the medium.
 Being able to perform experiments with realistic workload is a vital part of the ongoing process of improving wireless protocols.
@@
 
-Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the exhibited behavior is governed by a closed feedback loop.
+Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the emergent behavior is governed by a closed feedback loop.
 % Traffic models typically attempt to preserve the closed-loop behavior of network traffic.
 In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
-Hern\'andez-Campos' dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and applications for typical \caps{TCP} applications.
-Hern\'andez-Campos shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
+Hern\'andez-Campos' dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and typical \caps{TCP} applications.
+He shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
 This research has subsequently been turned into a \caps{TCP} flow generation tool called \tmix~\cite{Weigle06}.
@@
 Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
-Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary input to {\tmix}.
+Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary inputs to {\tmix}.
 
@@
 Harpoon exemplifies the uniform and marginal modeling approaches.
-Source and destination nodes are treated marginally: empirically derived marginal frequencies of sources and destination nodes (independently) are used to choose endpoints for each flow.
+Source and destination nodes are treated marginally: empirically derived marginal frequencies of sources and destinations are used to choose endpoints independently for each flow.
 File sizes (\latin{i.e.} total flow sizes) are likewise treated marginally: each flow's total byte  count is sampled from a marginal distribution of flow sizes.
-\caps{UDP} flows are modeled as being constant bit-rate (\latin{i.e.} having uniform packet behavior), and \caps{TCP} flows are all treated as file transfers, making them effectively uniformly behaved with bit-rate determined by \caps{TCP} dynamics.
-Inter-connection times (\latin{i.e.} per node-pair inter-flow intervals) are sampled from a single empirically derived marginal distribution.
+\caps{UDP} flows are modeled as being constant bit-rate; %(\latin{i.e.} having uniform packet behavior),
+\caps{TCP} flows are all treated as file transfers, making them effectively uniform with bit-rate determined by \caps{TCP} dynamics.
+Inter-connection times %(\latin{i.e.} per node-pair inter-flow intervals)
+are sampled from a single empirically derived marginal distribution.
 While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic,\harpoonnote for the purpose of generating local-area traffic for wireless experiments, it is not.
-Previous research~\cite{Karpinski07:cbr-failure} has shown that \caps{CBR} packet behavior models used in Harpoon drastically distort vital performance metrics at all levels of the protocol stack in wireless network simulations.
-This paper shows that marginal traffic models also severely misrepresent performance metrics.
-
-Like Harpoon, \caps{D-ITG}~\cite{Avallone04} exemplifies the common uniform and marginal modeling assumptions.
-Packet behavior is modeled using a handful of standard statistical distributions (``constant, uniform, normal, cauchy, pareto, exponential, etc.'') for packet size and inter-packet interval distributions, from which packets are generated using repeated independent sampling.
-Guidance is not provided, however, for choosing among these distributions, choosing parameter values for individual flows, or selecting a mixture of distributions for flows across an entire network.
-Moreover, the mechanism for choosing end-points and start and end times for flows is not described; presumably it is explicitly specified by the user. 
-The focus of this project is primarily on the engineering challenge of generating very large volumes of synthetic Internet-like backbone traffic and injecting this traffic into real networks in a distributed fashion.
+Previous research has shown that uniform packet behavior models used in Harpoon drastically distort vital performance metrics at all levels of the protocol stack~\cite{Karpinski07:cbr-failure}.
+% in wireless network simulations.
+This paper shows, moreover, that marginal models, like those used in Harpoon, also severely misrepresent performance metrics in wireless networks.
+
+Like Harpoon, \caps{D-ITG}~\cite{Avallone04} exemplifies uniform and marginal modeling assumptions. However, unlike Harpoon, \caps{D-ITG} does not derive models from real traffic.
+Packet behavior can be modeled using standard statistical distributions (``constant, uniform, normal, cauchy, pareto, exponential, etc.''). % for packet size and inter-packet intervals.
+Guidance is not provided, however, for choosing among these distributions, choosing parameter values, or selecting a mixture of distributions across a network.
+Flow and node behaviors are not specified at all, but rather left to the user's discretion.
+% Moreover, the mechanism for choosing end-points and start and end times for flows is not described; presumably it is explicitly specified by the user. 
+The focus of this project is primarily on the engineering challenge of generating very large volumes of synthetic traffic and injecting it into real networks in a distributed fashion.
 While \caps{D-ITG} provides the mechanism for generating large volumes of network traffic, it does not provide any real guidance as to what traffic to generate.
@@
 \FHC~\cite{Hernandez06:wlan-traffic} have defined the state of the art in parametric modeling of wireless traffic patterns.
-They use rigorous statistical methods to fit parametric distributions to a number of properties of wireless traffic instances:
+They use rigorous statistical methods to fit parametric models to a number of properties of wireless traffic.
 % They propose the following models:
-flow arrival processes fit a non-stationary Poisson arrival model; flow inter-arrival intervals (per session) fit a Lognormal distribution;
-flows per session fit a BiPareto model; flow sizes fit a BiPareto model.
-However, all the proposed models are marginal:
-a single statistical model applies to all flows and nodes (packet behavior is not considered).
+Flow arrivals fit a non-stationary Poisson model;
+flow inter-arrival times fit a Lognormal distribution;
+the number of flows per session fits a BiPareto model;
+flow sizes also fit a BiPareto model.
+The statistical analysis fitting these parametric models is thorough and convincing.
+The models, however, are marginal:
+a single statistical model applies to all flows and nodes; packet behavior is not considered.
 % One intuitively obvious heterogeneous nature of network traffic:
 Each node has a different number of flows assigned to it, but otherwise nodes are identical. 
-The flows associated with each node are statistically identical:
+Moreover, all flows, regardless of source node, are statistically identical:
 flow sizes are drawn from a single marginal distribution, and nothing else distinguishes them.
@@
 % This is a classic example of the limits of marginal modeling.
-Yet, intuitively it is clear that different nodes in a network have different tendencies for flow size, flow initiation, and packet behavior.
+Yet, intuitively it is clear that different nodes in real networks have drastically different tendencies for flow size, flow rates, and packet behavior.
 Such interdependent behaviors cannot be captured using marginal modeling.
 In this paper, we demonstrate that this inability is not harmless:
-treating all flows and nodes statistically identically severely misrepresents important wireless performance metrics in experiments.
+treating all flows and nodes statistically identically severely distorts crucial wireless performance metrics in experiments.
+We also provide sorely needed methods of representing and analyzing traffic patterns without making uniform or marginal assumptions about network behavior.
 

### 2008-06-06 trace_vs_uniform.tex
More work on intro & motivation.

@@
 \begin{center}
-\subfloat[node real]
+\subfloat[node: real]
 {\includegraphics[width=1.335in]{flow_topology_trace}\label{fig:flow-topology-trace}}
-\subfloat[node uniform]
+\subfloat[node: uniform]
 {\includegraphics[width=1.335in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
-\subfloat[flow real]
+\subfloat[flow: real]
 {\includegraphics[width=1.335in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
-\subfloat[flow uniform]
+\subfloat[flow: uniform]
 {\includegraphics[width=1.335in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}

### 2008-06-06 paper.tex
...
Merged thesis references. Worked over intro & related.

@@
 
-\newfootnote{\realismnote}{It is realistic in the sense that traffic generated using this model accurately reproduces performance metrics of real traffic.}
-
+% TODO: is this paragraph necessary?
 The second reason why uniformity and marginality assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
@@
 In the terminology introduced above, this is a flow-conditional packet behavior model.
-Prior research has indicated that, given accurate per-flow distributions, this is a sufficiently realistic model of network behavior~\cite{Karpinski07:realism}.\realismnote
+Prior research has indicated that, given accurate per-flow distributions, and keeping all other aspects of network behavior fixed, this simplification does not distort experimental performance metrics~\cite{Karpinski07:realism}.
 In this model, however, each flow must have its own specific, realistic pair of distributions for packet size and inter-packet intervals.
@@
 
-Fundamentally, uniform, marginal and conditional models are applied so that we can represent network traffic behavior using a few empirical or analytical statistical distributions.
-Reduction in representation size and complexity makes analysis and generation tractable problems.
+Fundamentally, uniform, marginal and conditional models are applied so that network traffic behavior can be represented using a manageable collection of empirical or analytical statistical distributions.
+Reduction in representation size and complexity makes analysis and generation problems tractable.
 From this perspective, two questions immediately present themselves:
@@
 We argue that matrix factorization, which is nondestructive, is far superior to multiplication for reducing the representation of network traffic behaviors.
-To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF}) to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
+To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF})~\cite{Lee01} to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
 This factorization extracts a small set of ``basic behaviors'' from the traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
@@
 Basic behaviors are found corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.
-
-% Via this factorization, each flow's behavior as a linear combination a handful of ``basic behaviors'' such as file transfer, typing via \caps{SSH}, ping traffic and other network activities. 
-% These intuitive behavior modes as well as the explanation of flows in terms of them are extracted from trace date purely algorithmically.
-
-% First we observe that in any linear representation of traffic, the assumption of uniformity corresponds the criterion that the coefficients in each dimension be the same.
-
-% To address the second question, we observe that in the general matrix model, 
-% 
-% Moreover, we apply state of the art matrix factorization techniques to this representation to explain cross-network flow behaviors as linear mixtures of only twelve basic behaviors.
-
-% To address the second question,
-
-% The second contribution of this paper is to answer these questions and to propose a mathematical representation of network traffic that preserves the non-uniformity and non-independence of flow behaviors in a network, but makes traffic analysis and workload generation from this representation tractable problems.
-% Here we will briefly explain our approach.
-% 
-% It is already common practice in network modeling to think of flow behaviors as vectors.
-% The \caps{IP} protocol number, source and destination \caps{IP} addresses, and \caps{UDP/TCP} source and port numbers that define a packet flow are commonly referred to as a 5-tuple---i.e., a vector in five dimensions.
-% Other aspects of a flow's behavior can be represented vectorially as well.
-We call such a mapping of flows into a vector space a \emph{linear representation} if it satisfies the following simple linearity criterion:
-% \begin{quote}
-% The sum of the representations of two flows is a vector that represents the aggregate behavior of the two flows.
-% \end{quote}
-% We show that flows can be represented satisfying this criterion. This immediately allows us to perform useful computations on traffic patterns through simple vector addition and scaling.
-% Finally, we argue that the natural generalization of this approach is to represent an entire network's behavior as a large matrix, with each row representing a single flow.
-% The matrix representation gives a representation of network behavior that preserves the non-uniformity and non-independence of flow behaviors, but still is highly amenable to analysis using theoretical and numerical linear algebra.
-% Moreover, we describe how to generate traffic from arbitrary traffic matrices, and demonstrate experimentally that the specific matrix representation we propose preserves the performance characteristics of real traffic.
 
@@
 Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies.
-%, and how it points the way to better traffic models for the future.
 
@@
 
-%The history of networking research contains many examples of simplistic models that have proven not only to be inaccurate, but also to drastically skew important characteristics of network behavior.
-Paxson and Floyd showed that the Poisson packet arrival model, which had been standard for studying wide-area Internet traffic, failed to capture the burstiness and self-similarity of real traffic~\cite{Paxson95}.
-The equally common but simplistic Random Way-Point~(\caps{RWP}) mobility model was found to exhibit ``density waves'' and gradual slow-down of average node speed~\cite{Royer01,Yoon03:speed-decay}.
-In the worst cases, overly simplistic models can switch the relative performance of protocols, thereby invalidating the conclusions drawn from performance comparisons using those models.
-
 \newfootnote{\QoSnote}{This is violated by some quality of service (QoS) schemes.
@@
 This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner~\cite{Clark88}.\QoSnote The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic.
-\Figure{trace-vs-uniform} illustrates through examples how drastically different network traffic appears under uniformity assumptions at different levels of behavior.
+\Figure{trace-vs-uniform} illustrates through examples how drastically different network traffic appears under assumptions of uniformity at different levels of behavior.
 Marginality assumptions are not as visually striking, but as we will demonstrate, they distort the performance characteristics of networks almost as much, and sometimes more, than uniform models.
 
-While this work is not inherently specific to wireless networks, since the modeling techniques developed are equally applicable to traffic in wired networks, the results are of specific importance to wireless research.
-Realistic traffic models have a drastic impact on the experimental results in wireless networks~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+While this work is not inherently specific to wireless networks, since the modeling techniques developed are equally applicable to traffic in wired networks, the results are of special importance to wireless research.
+Realistic traffic models have a drastic impact on experimental results in wireless networks~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
 This is primarily because wireless networks are locally resource constrained.
-Compare this with the typical situation in wired \caps{LAN}s, where gigabit Ethernet and modern routers and switches can easily handle all but the most intense levels of traffic.
+Compare this with the typical situation in wired \caps{LAN}s, where gigabit Ethernet and modern routers  can easily handle all but the most intense levels of traffic.
 If a wired network operator expects larger volumes of traffic on a regular basis, the solution is to upgrade the hardware, not to change the network protocols.
@@
 Being able to perform experiments with realistic workload is a vital part of the ongoing process of improving wireless protocols.
-Without realistic workload models for experiments, performance comparisons may be misleading or even completely wrong~\cite{Karpinski07:cbr-failure}.% Better traffic models are necessary to avoid stumbling in the dark while attempting to improve wireless technologies.
+Without realistic workload models for experiments, performance comparisons may be misleading or even completely wrong~\cite{Karpinski07:cbr-failure}.
 
 Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the emergent behavior is governed by a closed feedback loop.
-% Traffic models typically attempt to preserve the closed-loop behavior of network traffic.
 In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
@@
 Harpoon~\cite{Sommers04} uses a traffic trace for self-training, and can subsequently generate synthetic traffic with certain statistical properties based on the original trace.
-The properties reproduced are the empirical distributions of the following: ``file size, inter-connection time, source and destination IP ranges, number of active sessions.''
+The properties reproduced are the empirical distributions of the following: ``file size, inter-connection time, source and destination IP ranges, [and] number of active sessions.''
 Harpoon exemplifies the uniform and marginal modeling approaches.
 Source and destination nodes are treated marginally: empirically derived marginal frequencies of sources and destinations are used to choose endpoints independently for each flow.
-File sizes (\latin{i.e.} total flow sizes) are likewise treated marginally: each flow's total byte  count is sampled from a marginal distribution of flow sizes.
+Flow sizes (\latin{i.e.} file sizes) are likewise treated marginally: each flow's total byte  count is sampled from a marginal distribution of flow sizes.
 \caps{UDP} flows are modeled as being constant bit-rate; %(\latin{i.e.} having uniform packet behavior),
@@
 are sampled from a single empirically derived marginal distribution.
-While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic,\harpoonnote for the purpose of generating local-area traffic for wireless experiments, it is not.
+While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic,\harpoonnote for generating local-area traffic in wireless experiments, it is not.
 Previous research has shown that uniform packet behavior models used in Harpoon drastically distort vital performance metrics at all levels of the protocol stack~\cite{Karpinski07:cbr-failure}.
@@
 Like Harpoon, \caps{D-ITG}~\cite{Avallone04} exemplifies uniform and marginal modeling assumptions. However, unlike Harpoon, \caps{D-ITG} does not derive models from real traffic.
-Packet behavior can be modeled using standard statistical distributions (``constant, uniform, normal, cauchy, pareto, exponential, etc.''). % for packet size and inter-packet intervals.
+Packet behavior can be modeled using standard statistical distributions (``constant, uniform, normal, cauchy, pareto, exponential, etc.'').
 Guidance is not provided, however, for choosing among these distributions, choosing parameter values, or selecting a mixture of distributions across a network.
 Flow and node behaviors are not specified at all, but rather left to the user's discretion.
-% Moreover, the mechanism for choosing end-points and start and end times for flows is not described; presumably it is explicitly specified by the user. 
-The focus of this project is primarily on the engineering challenge of generating very large volumes of synthetic traffic and injecting it into real networks in a distributed fashion.
-While \caps{D-ITG} provides the mechanism for generating large volumes of network traffic, it does not provide any real guidance as to what traffic to generate.
-
-\FHC~\cite{Hernandez06:wlan-traffic} have defined the state of the art in parametric modeling of wireless traffic patterns.
+The focus of this project is primarily on the engineering challenge of generating very large volumes of synthetic traffic and injecting it into networks in a distributed fashion.
+While \caps{D-ITG} provides the mechanism for generating large volumes of traffic, it does not provide any real guidance as to what traffic to generate.
+
+\FHC~\cite{Hernandez06:spatio-temporal-modeling} have defined the state of the art in parametric modeling of wireless traffic patterns.
 They use rigorous statistical methods to fit parametric models to a number of properties of wireless traffic.
@@
 %Most approaches to traffic modeling that go beyond uniformity assumptions, instead describe marginal properties of various aspects of traffic behavior.
-For example, {\FHC}~\cite{Hernandez06:wlan-traffic} provide highly convincing parametric models for six marginal properties of wireless network traffic: session (node) arrivals follow a time-varying Poisson process; per-session flow inter-arrival times are Lognormally distributed; flows per session (node) follow a BiPareto distribution; flow sizes follow a BiPareto distribution.
+For example, {\FHC}~\cite{Hernandez06:spatio-temporal-modeling} provide highly convincing parametric models for six marginal properties of wireless network traffic: session (node) arrivals follow a time-varying Poisson process; per-session flow inter-arrival times are Lognormally distributed; flows per session (node) follow a BiPareto distribution; flow sizes follow a BiPareto distribution.
 This is certainly much greater detail than simply assuming that all nodes and flows are identical.

### 2008-06-06 paper.tex
Significant rewrite of intro.

@@
 
+% In short, the reason is that these assumptions drastically simplify both traffic analysis and workload generation.
+% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
+% This immediately begs the question of whether these assumptions are benign.
+% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to analyze and replicate?
+% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
+
+% TODO: is this paragraph necessary?
+% The second reason why uniformity and marginality assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
+% The difficulty lies in the high-dimensional and stochastic nature of network behaviors.
+% Suppose, for example, that we assume that packet sizes and inter-packet intervals are drawn independently from some given distribution for each flow.
+% In the terminology introduced above, this is a flow-conditional packet behavior model.
+% Prior research has indicated that, given accurate per-flow distributions, and keeping all other aspects of network behavior fixed, this simplification does not distort experimental performance metrics~\cite{Karpinski07:realism}.
+% In this model, however, each flow must have its own specific, realistic pair of distributions for packet size and inter-packet intervals.
+% Moreover, the ``distribution of distributions'' across the entire network must simultaneously be realistic.
+% Modeling such distributions of distributions realistically for entire networks is currently an unsolved problem; we begin to address it here.
+
 If no one actually believes that assumptions of uniformity and marginality accurately describe network behavior, then why are these assumptions so common?
-The first reason is simply that these assumptions drastically simplify both traffic analysis and workload generation.
-If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
-This immediately begs the question of whether these assumptions are benign.
-Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to analyze and replicate?
-This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
-
-% TODO: is this paragraph necessary?
-The second reason why uniformity and marginality assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
-The difficulty lies in the high-dimensional and stochastic nature of network behaviors.
-Suppose, for example, that we assume that packet sizes and inter-packet intervals are drawn independently from some given distribution for each flow.
-In the terminology introduced above, this is a flow-conditional packet behavior model.
-Prior research has indicated that, given accurate per-flow distributions, and keeping all other aspects of network behavior fixed, this simplification does not distort experimental performance metrics~\cite{Karpinski07:realism}.
-In this model, however, each flow must have its own specific, realistic pair of distributions for packet size and inter-packet intervals.
-Moreover, the ``distribution of distributions'' across the entire network must simultaneously be realistic.
-Modeling such distributions of distributions realistically for entire networks is currently an unsolved problem; we begin to address it here.
+There are two reasons.
+The first is that these assumptions drastically simplify both traffic analysis and workload generation, making them tractable problems.
+% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
+% This immediately begs the question of whether these assumptions are benign.
+% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to understand?
+% This paper answers that question.
+The second is that there are no generally accepted alternatives for representing and simplifying traffic patterns that allow effective analysis and generation without assuming uniformity or marginality.
+% Representing complex multi-level network behaviors in a way that does not make these simplifying assumptions but allows effective traffic analysis and workload generation is currently an unsolved problem;
+% we begin to address it here.
+
+% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
 
 Fundamentally, uniform, marginal and conditional models are applied so that network traffic behavior can be represented using a manageable collection of empirical or analytical statistical distributions.
-Reduction in representation size and complexity makes analysis and generation problems tractable.
-From this perspective, two questions immediately present themselves:
+Reduction in representation size and complexity makes analysis and generation feasible.
+From this perspective, three questions immediately present themselves:
 \begin{enumerate}
-\item Can we represent ``unreduced'' network behavior such that analysis and generation are tractable?
-\item Are there better ways to reduce the representation of network behavior?
+\item Do these assumptions distort the metrics we are trying to analyze, replicate, and predict?
+\item Can we represent ``unreduced'' network behavior such that analysis and generation remain tractable?
+\item Are there better ways to simplify and reduce the representation of network behavior?
 \end{enumerate}
-This paper answers both of these questions.
-To address the first question, we propose the concept of a \emph{linear representation} of network traffic.
+This paper answers all three questions. We begin, however by answering the second problem and  
+using our solution to address the other two questions.
+% transforming the unreduced representation of traffic into it's uniform, marginal or conditional counterparts.
+
+To represent unreduced network behavior practically, we propose the concept of a \emph{linear representation} of network traffic.
 A linear representation of traffic maps each flow to a vector, such that the following linearity condition is satisfied:
@@
 
-To address the question of whether there are better ways to reduce the representation of network behaviors, we first make a simple observation.
+To address the the other two questions, we begin with a simple observation.
 In linear representations of traffic, uniform, marginal and conditional assumptions are assertions of equality between the rows and columns of the traffic matrix.
-Moreover, real traffic matrices can be transformed into uniform, marginal or conditional forms through  very simple matrix multiplications.
+Moreover, real traffic matrices can be transformed into uniform, marginal and conditional forms through very simple matrix multiplications.
 These transformations, however, are extremely destructive: the matrices are degenerate with very low rank, so multiplication destroys almost all information in the original traffic matrix.
+We demonstrate experimentally, that this loss of information is not benign:
+uniform and marginal traffic models severely distorting crucial network performance metrics.
 
@@
 
-We argue that matrix factorization, which is nondestructive, is far superior to multiplication for reducing the representation of network traffic behaviors.
+Once traffic is represented linearly, alternatives to assuming uniformity or marginality become apparent.
+We argue that matrix factorization, which is nondestructive and does not assume uniformity or marginality, is far superior to multiplication for this problem.
 To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF})~\cite{Lee01} to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
@@
 Although basic behaviors are computed without examining port numbers, they nevertheless correspond to the intuitive behaviors expected for protocols.
-Basic behaviors are found corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.
+We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.
 

### 2008-06-06 paper.tex
A bit more rewriting.

@@
 This factorization extracts a small set of ``basic behaviors'' from the traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
-Although basic behaviors are computed without examining port numbers, they nevertheless correspond to the intuitive behaviors expected for protocols.
+Although the factorization is computed without examining port numbers, basic behaviors nevertheless correspond to intuitive expectations for protocols.
 We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.

### 2008-06-07 paper.tex
Related work para on WinMee & Broadnets papers.

@@
 
-% This presents a fundamental difficulty, however, in that it presumes that we know the intent of endpoints: what \textit{would} they have done under different conditions?
-While we can speculate about what an individual node might hypothetically do, we currently hardly understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations.
-A fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of application and user feedback.
-
-% Although we do not attempt to model application and user feedback, we do simulate the effect of the network layer and below on the dynamics of \caps{TCP} for flows that were originally transmitted using \caps{TCP}.
-Hern\'andez-Campos~\cite{Hernandez06:dissertation} addresses in great depth how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how many common applications would react in response to network conditions.
-This research has subsequently been turned into a \caps{TCP} flow generation tool for \caps{NS-2}, called \tmix~\cite{Weigle06}.
-This work is complementary to ours: {\tmix} does not provide a model for application behaviors; it only provides the ability to reproduce accurate closed-loop dynamics given an accurate application behavior model.
-Our work our work provides the means to analyze traffic and generate realistic application workloads which can then serve as the required inputs to \tmix.
-
-% This paper attempts to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without feedback from the network to applications and users.
-It is important to realize that while this does not provide a final picture, we currently lack even a first-order understanding of the effect of different workloads on performance.
-This first-order understanding is an essential initial step.
-
-% There have been many studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}.
-These analyses have described a wide variety of aspects of wireless network behavior, and provide much insight into the workings of real, deployed wireless networks.
-These studies present a broad analysis of general system features and trends of specific corporate wireless local-area networks (\caps{WLAN}s)~\cite{Tang99,Balazinska03}, university campus \caps{WLAN}s~\cite{Tang00,Kotz02,Chinchilla04,Schwab04,Henderson04,Tuduce05}, and temporary \caps{WLAN}s at conference venues~\cite{Balachandran02,Jardosh05:ewind}.
-They also provide a large body of raw data for subsequent analysis and modeling research.
-Our work provides the methodology for turning this rich foundation of field data into usable, realistic models of workload for a wide variety of networking situations.
-% 
-% The choice of mobility models for mobile wireless simulations can have a drastic impact on important performance metrics~\cite{Camp02,Yoon03:speed-decay,Yoon03:sound-models,Jardosh03,Zheng04}.
-Moreover, commonly used but simplistic mobility models, such as \caps{RWP}, exhibit characteristics, including density waves and speed decay, that are categorically dissimilar from any known real-world behavior~\cite{Royer01,Yoon03:speed-decay,Yoon03:sound-models}.
-In response to this evidence, more realistic mobility models have been proposed~\cite{Jardosh03,Tuduce05,Jardosh05:voronoi}.
-While much of this work focuses on making models that are simply more intuitively appealing~\cite{Jardosh03,Jardosh05:voronoi}, some work has begun to capitalize on this newly created wealth of wireless field data by deriving models from observed usage behavior, rather than intuition alone~\cite{Balazinska03,Tuduce05}.
-%The analysis of Balazinska \textit{et al.}~\cite{Balazinska03} introduces the notions of the prevalence and persistence of an \caps{AP} with respect to mobile nodes.\footnote{These concepts are adapted from Paxson's work on Internet routing behavior: ``\textit{prevalence} meaning the overall likelihood that a particular route is encountered, and \textit{persistence}, the likelihood that a route remains unchanged over a long period of time''~\cite{Paxson96}.} Tuduce \textit{et al.}~\cite{Tuduce05} have based their mobility model on the persistence and prevalence characteristics of real trace data, which makes their work particularly relevant to our definition of sufficient realism.
-They do not, however, determine the impact of prevalence and persistence characteristics on performance metrics.
-
-% In this paper, instead of mobility, we examine an even more fundamental aspect of user behavior in wireless networks: the pattern of traffic generated by users and applications.
-This aspect of behavior is more fundamental because it applies to all types of wireless networks, not just mobile and ad~hoc networks.
-Moreover, the effect of traffic patterns applies not only to simulations, but also to experimental deployments, which have become the gold standard for wireless protocol evaluation.
-Experimental deployments sidestep the problem of accurately modeling the lower layers of the network.
-Unless traffic and mobility are modeled realistically, however, the experimental results will still be unreliable.
-
-% There is a large and diverse body of work on traffic analysis, modeling, and generation~\cite{Paxson95,Paxson96,Sommers04,Avallone04,Hernandez06:dissertation}.
-We are only able to discuss a small, but hopefully representative sampling of this work.
-Almost all of the traffic generation work has focused on wide-area Internet backbone traffic.
-%Because of the comparatively unlimited capacity of wired networks, traffic workload generally does not offer interesting technical challenges at the edge of the network.
-
 \newfootnote{\harpoonnote}{Even this is unclear: generated traffic is only compared with trace traffic using the very same metrics that are explicitly specified as part of the model.
@@
 Flow sizes (\latin{i.e.} file sizes) are likewise treated marginally: each flow's total byte  count is sampled from a marginal distribution of flow sizes.
-\caps{UDP} flows are modeled as being constant bit-rate; %(\latin{i.e.} having uniform packet behavior),
+\caps{UDP} flows are modeled as being constant bit-rate;
 \caps{TCP} flows are all treated as file transfers, making them effectively uniform with bit-rate determined by \caps{TCP} dynamics.
-Inter-connection times %(\latin{i.e.} per node-pair inter-flow intervals)
-are sampled from a single empirically derived marginal distribution.
+Inter-connection times are sampled from a single empirically derived marginal distribution.
 While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic,\harpoonnote for generating local-area traffic in wireless experiments, it is not.
 Previous research has shown that uniform packet behavior models used in Harpoon drastically distort vital performance metrics at all levels of the protocol stack~\cite{Karpinski07:cbr-failure}.
-% in wireless network simulations.
 This paper shows, moreover, that marginal models, like those used in Harpoon, also severely misrepresent performance metrics in wireless networks.
@@
 They use rigorous statistical methods to fit parametric models to a number of properties of wireless traffic.
-% They propose the following models:
 Flow arrivals fit a non-stationary Poisson model;
@@
 a single statistical model applies to all flows and nodes; packet behavior is not considered.
-% One intuitively obvious heterogeneous nature of network traffic:
 Each node has a different number of flows assigned to it, but otherwise nodes are identical. 
@@
 flow sizes are drawn from a single marginal distribution, and nothing else distinguishes them.
-%, unconditional of their associated node.
-% This is a classic example of the limits of marginal modeling.
 Yet, intuitively it is clear that different nodes in real networks have drastically different tendencies for flow size, flow rates, and packet behavior.
@@
 
-% The focus of this project, however, is on providing the infrastructure to generate very large volumes of synthetic Internet-like backbone traffic.
-No analysis is provided for determining the realism of traffic mixes, or for choosing flow endpoints realistically.
-In wireless networks, these factors are of crucial importance to performance, and cannot be overlooked.
-Both Harpoon and \caps{D-ITG} provide excellent traffic generation platforms, but do not provide a systematic framework for understanding or reproducing realistic whole-network workload in the wireless setting.
-
-%This work extends that published in~\cite{Karpinski07:realism}~and~\cite{Karpinski07:cbr-failure}, using the same methodology for defining and quantifying the realism of synthetic traffic models.
-The Traffic Object Model builds on the combinatorial approach taken in~\cite{Karpinski07:cbr-failure}, while providing an flexible and general object-oriented framework for constructing generative traffic models.
-The traffic models evaluated here overlap and augment those evaluated in both prior works; comparison with the models proposed in {\FHC} is entirely new.
-The Qualnet simulation methodology has been enhanced by the implementation of full-duplex, client-server, trace-driven raw \caps{IP}, \caps{TCP}, and \caps{UDP} flows.
-We have also implemented performance metric aggregation based on the \caps{IETF} framework for \caps{IP} performance metrics~\cite{rfc:ip-metrics,rfc:jitter}.
-
-% TODO: add in the Hernandez-Cortes' dissertation.
-% TODO: concluding remarks about what's missing from traffic work in general.
-%   + there's no central goal or theme---because realism is ill-defined and subjective
-%   + there are no rigorous tests for whether realism is achieved since it's ill-defined
-%   + in wireless, where the impact is greater, there's been virtually no research of impact
-% TODO: cite a bunch of papers that use the CBR traffic model.
-
-%\section{Algebraic Representations \\of Network Behavior}
+This work uses the technique known as \emph{paired differential simulation} to evaluate the realism of modified or synthetically generated wireless workloads.
+This technique was proposed and subsequently refined by Karpinski~\latin{et~al.}~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+They introduce the notion of \emph{sufficient realism}, and propose paired differential simulation as an experimental methodology based on the standard technique of pairing control and test subjects.
+We describe the technique in \Section{methodology}.
+In the first paper~\cite{Karpinski07:realism}, they explore various ways of modifying trace traffic patterns, as a form of sensitivity analysis to determine which aspects of network behavior are essential and must be preserved in realistic models, and which aspects can be discarded as inessential.
+Of particular interest is their conclusion that detailed time-series packet behavior for flows is inessential:
+it suffices, for each flow, to have accurate unconditional distributions of packet sizes and inter-packet intervals;
+sufficiently realistic flow behavior can be reconstructed from these distributions by repeated independent sampling.
+In the second paper~\cite{Karpinski07:cbr-failure}, Karpinski~\latin{et~al.} explore the impact of a broad variety of uniformity assumptions at the packet, flow and node levels of behavior.
+They find that all uniformity assumptions are detrimental, inducing significant and often extreme misrepresentation of a broad variety of wireless performance metrics at every level of the protocol stack.
+Moreover, they show, using \caps{OLSR} and \caps{AODV}, that uniform traffic models can completely invert the relative performance of network protocols.
+
 \section{The General Matrix Model}

### 2008-06-09 paper.tex
Separated definitions; Linear Representation section.

@@
 \usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle}]{hyperref}
-
-\newcommand{\caps}[1]{{\smaller{#1}}}
+\usepackage{mathtools}
 
@@
 
-\newcommand{\Section}[1]{\hyperref[sec:#1]{Section~\ref*{sec:#1}}}
-\newcommand{\Equation}[1]{\hyperref[eqn:#1]{Equation~\ref*{eqn:#1}}}
-\newcommand{\Figure}[1]{\hyperref[fig:#1]{Figure~\ref*{fig:#1}}}
-\newcommand{\Table}[1]{\hyperref[tab:#1]{Table~\ref*{tab:#1}}}
-
-\newcommand{\FHC}{Hern\'andez-Campos~\textit{et~al.}}
-\newcommand{\class}[1]{\textsc{#1}}
-\newcommand{\model}[1]{\textsf{#1}}
-\newcommand{\RFC}[1]{\caps{RFC}~{#1}}
-\newcommand{\tmix}{T\caps{MIX}}
-
-\DeclareMathOperator{\var}{Var}
-\DeclareMathOperator{\N}{\mathcal{N}}
-\DeclareMathOperator{\Def}{def}
-\DeclareMathOperator{\Val}{val}
-
-\newcommand{\defeq}{\stackrel{\Def}{=}{}}
-\newcommand{\eqnote}[1]{\text{\footnotesize{[#1]}}}
-
-\newcommand{\X}{\mathsf{X}}
-\newcommand{\M}{\mathsf{M}}
-\newcommand{\E}[1]{\left<#1\right>}
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\Q}{\mathcal{Q}}
-\newcommand{\Qt}{Q_{\text{time}}}
-\newcommand{\Qb}{Q_{\text{byte}}}
-\newcommand{\Qi}{Q_{\text{ival}}}
-\newcommand{\Di}{Q^{-1}_{\text{ival}}}
-\newcommand{\ones}[1]{\mathbf{1}_{#1}}
-\newcommand{\zeros}[1]{\mathbf{0}_{#1}}
-\newcommand{\mat}[1]{\mathbf{#1}}
-\newcommand{\mean}[1]{\bar{#1}}
-\newcommand{\trans}[1]{{#1}^T}
-\newcommand{\inner}[2]{{#1}\cdot{#2}}
-
-\newcommand{\abs}[1]{\left|#1\right|}
-\newcommand{\absx}[1]{|#1|}
-\newcommand{\set}[1]{\left\{#1\right\}}
-\newcommand{\setx}[1]{\{#1\}}
-\newcommand{\parens}[1]{\left(#1\right)}
-\newcommand{\parensx}[1]{(#1)}
-\newcommand{\seq}[1]{\left<#1\right>}
-\newcommand{\seqx}[1]{<#1>}
-\newcommand{\floor}[1]{\left\lfloor#1\right\rfloor}
-\newcommand{\floorx}[1]{\lfloor#1\rfloor}
-\newcommand{\ceil}[1]{\left\lceil#1\right\rceil}
-\newcommand{\ceilx}[1]{\lceil#1\rceil}
-\newcommand{\round}[1]{\left[#1\right]}
-\newcommand{\roundx}[1]{[#1]}
-\newcommand{\fracx}[2]{#1/#2}
-\newcommand{\fracp}[2]{\left(\frac{#1}{#2}\right)}
-\newcommand{\fracpx}[2]{(#1/#2)}
-
-\renewcommand{\bullet}{\raisebox{2pt}{$\centerdot$}}
-%\renewcommand{\labelitemi}{\bullet}
-\newcommand{\latin}[1]{\textit{#1}}
-
-\newcommand{\newfootnote}[2]{\newcommand{#1}{\footnote{#2} }}
+\input{definitions}
 
@@
 
+\section{Linear Representation of Traffic}
+
+% In this section, we the concept of linear representation of network traffic.
+Before we introduce this concept, we discuss the characteristic features of network traffic from the traffic modeling perspective.
+
+The traffic workload pattern in a network is a collection of \caps{IP} packets between hosts in the network sent at certain times.
+It is standard in network analysis, however, to aggregate sequences of packets sharing the same ``5-tuple'':
+\caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP/UDP} port numbers.
+Such a sequence of packets is called a \textit{flow}.
+The traffic pattern of an entire network is simply the collected behaviors of all flows occurring in the network.
+
+The behavior of a flow as it affects the network is characterized by the following properties:
+its \caps{IP} protocol type (\caps{TCP}, \caps{UDP}, \caps{ICMP}, etc.);
+its source and destination nodes;
+its start time;
+and the specific sizes of and intervals between its constituent packets.
+We begin our exposition of traffic representation by formalizing these properties mathematically as sets:
+\begin{tabbing}
+\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{5em}\=\hspace{4em}\=$\in\Nodes\,$\=\kill
+\>\bullet\>types:\>$\type$\>$\in\Types$\>$= \set{1,\dots,255} \subset \N$ \\
+\>\bullet\>nodes:\>$\src,\dst$\>$\in\Nodes$\>$= \set{1,\dots,n} \subset \N$ \\
+\>\bullet\>times:\>$\start$\>$\in\Times$\>$= \R$ \\
+\>\bullet\>sizes:\>$\size$\>$\in\Sizes$\>$= \set{0,\dots,\text{\caps{MTU}}} \subset \N$ \\
+\>\bullet\>intervals:\>$\ival$\>$\in\Ivals$\>$ = \R^+ \union \set{\infty}$.
+\end{tabbing}
+There are 255 possible \caps{IP} types (0 is reserved for \caps{IPv6}).
+The number $n$ is simply the total number of nodes in the network being modeled.
+\caps{MTU} is the maximum transfer unit of the network (typically around 1500 bytes).
+We formalize the packet behavior of a flow as a pair of infinite sequences of packet sizes and inter-packet intervals:
+\begin{align}
+\sizes &= \seq{\size_1,\size_2,\dots} \in \SizeSeq, \\
+\ivals &= \seq{\ival_1,\ival_2,\dots} \in \IvalSeq.
+\end{align}
+Since flows contain only a finite number of packets, there exists a number of packets, $\pkts \in \N$, for each flow.
+Thus, we require that the sequences $\sizes$ and $\ivals$ satisfy the following requirements:
+\begin{align}
+&\forall\: i > \pkts:   \quad \size_i = 0,      \eqn{zero-packets}\\
+&\forall\: i < \pkts:   \quad \ival_i < \infty, \eqn{finite-intervals}\\
+&\forall\: i \ge \pkts: \quad \ival_i = \infty. \eqn{infinite-intervals}
+\end{align}
+% The send time of the $k$th packet can be calculated as
+% \begin{align}
+% \start + \sum_{i=1}^k \ival_i.
+% \end{align}
+Now we are ready to formalize the space of flows:
+\begin{align}
+  \Flows =
+    \Types    \cross
+    \Nodes^2  \cross
+    \Times    \cross
+    \N        \cross
+    \SizeSeq  \cross
+    \IvalSeq.
+\end{align}
+We write a generic element of $\Flows$ as
+\begin{align}
+  \flow = \tuple{
+    \type,
+    \src,
+    \dst,
+    \start,
+    \pkts,
+    \sizes,
+    \ivals
+  } \in \Flows.
+\end{align}
+% We define the function $\pkts:\flow \to \N$ as
+% \begin{align}
+%   \pkts(\flow) \defeq \min\set{i \in \N : \ival_i = \infty}.
+% \end{align}
+% Intuitively, by Equations~\ref{eqn:finite-intervals}~and~\ref{eqn:infinite-intervals}, this gives the number of packets in the flow.
+Let $2^\Flows$ denote the power set of $\Flows$, and $\Traffic \subseteq 2^\Flows$ the set of all finite sets of flows:
+\begin{align}
+  \Traffic = \set{X \in 2^\Flows : \card{X} < \infty}.
+\end{align}
+A traffic sample is simply an element of $\Traffic$. We will refer to elements of $\Traffic$ interchangeably as ``traffic samples'' or ``traffic instances.''
+
+Having chosen a mathematical formalization of network traffic, we may now define the notion of a linear representation of network traffic.
+
+\begin{definition}\defn{linear-representation}
+A \defterm{linear representation} of network traffic is a pair, $\tuple{V,\psi}$, where $V$ is a vector space and $\psi$ is a function, $\psi: \Traffic \to V$, that maps each traffic sample to the sum of the representations its constituent flows:
+\begin{align}\eqn{linearity}
+\forall~T \in \Traffic\!:\quad
+  \psi(T) &= \sum_{\mathclap{\flow \in T}}{\psi(\flow)}.
+\end{align}
+% When $T=\emptyset$, the right-hand side is a sum of zero items, defined to be $\vec{0}$, the identity element of $V$.
+\end{definition}
+
+\noindent
+In plain words, every finite collection of flows is represented by a vector that is the sum of the elements representing the individual flows in the collection.
+% Representing traffic using vectors allows us to analyze, model, and understand traffic patterns in terms of the thoroughly studied and well understood subject of linear algebra.
+The careful reader will notice that \emph{any} function $\Flows \to V$ uniquely determines a linear representation:
+\Equation{linearity} defines the extension to all of $\Traffic$.
+Representing flows by arbitrary vectors, however, is not particularly interesting or useful.
+Therefore, we focus on the meaningful representation of collections of flows, that can be factored or explained in terms of their constituent individual flows.
+In the next section, we clarify the concept and application of linear representation through examples.
+% we present and discuss specific examples of linear representations in order to clarify the concept and its application.
+
+\subsection{Examples of Linear Representations}
+
+Our first example is the packets-bytes-duration (\caps{PBD}) representation.
+The representation is $\pbd: \Traffic \to \R^3$, defined as
+$\pbd(\flow) = \tuple{\pkts,\mathsym{bytes},\mathsym{dur}}$, where
+\begin{align}
+\mathsym{bytes} &= \sum_{i=1}^{\mathclap{\pkts}}{\size_i} &
+\mathsym{dur} &= \sum_{i=1}^{\mathclap{\pkts-1}}{\ival_i}.
+\end{align}
+This representation encapsulates the information needed to reproduce a \caps{CBR} version of each flow:
+each packet should have a $\floor{\mathsym{bytes}/\pkts}$ byte payload, and intervals between packets should be $\mathsym{dur}/\pkts$ seconds.
+
 \section{The General Matrix Model}
@@
 The behavior of a flow as it affects the network is characterized by the following properties: its \caps{IP} protocol type (\caps{TCP}, \caps{UDP}, \caps{ICMP}, etc.); its source and destination nodes; its start time; and the specific sizes of and intervals between its constituent packets.
+
 We propose to represent the behavior of each flow in terms of seven vectors, which suffice to describe all of the above properties in enough detail to stochastically generate realistic traffic.
@@
 Despite its simplicity, this representation is adequate to reproduce constant bit-rate (\caps{CBR}), time-randomized flow behavior.
-The triplet of representative values can be naturally embedded into Euclidean space: $\vec{x}=\seq{p,b,d}\in\R^3$.
-If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their sum as vectors, $x_1+x_2=\seq{p_1+p_2,b_1+b_2,d_1+d_2}$, provides an appropriate description of the aggregate behavior of the two flows together.
+The triplet of representative values can be naturally embedded into Euclidean space: $\vec{x}=\tuple{p,b,d}\in\R^3$.
+If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their sum as vectors, $x_1+x_2=\tuple{p_1+p_2,b_1+b_2,d_1+d_2}$, provides an appropriate description of the aggregate behavior of the two flows together.
 Similarly, their average, i.e.
@@
 We take a highly detailed but intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the flow type, source node, destination node, start time, flow size, and empirical distributions of packet sizes and inter-packet intervals.
-If there are $f$ flows in the entire workload pattern, then the collective behavior can be described with six matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
+If there are $f$ flows in the entire workload pattern, then the collective behavior can be described with seven matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
 We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
@@
 $\mat{D}\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
-\>\bullet\> type: \>$\mat{T}$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$,\\
-\>\bullet\> sources: \>$\mat{S}$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$,\\
-\>\bullet\> destinations: \>$\mat{D}$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$,\\
-\>\bullet\> start times: \>$\mat{A}$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$,\\
-\>\bullet\> sizes in bytes: \>$\mat{B}$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$,\\
-\>\bullet\> packet sizes: \>$\mat{Z}$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
+\>\bullet\> type: \>$\mat{T}$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$\\
+\>\bullet\> sources: \>$\mat{S}$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$\\
+\>\bullet\> destinations: \>$\mat{D}$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$\\
+\>\bullet\> start times: \>$\mat{A}$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$\\
+\>\bullet\> sizes in bytes: \>$\mat{B}$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$\\
+\>\bullet\> packet sizes: \>$\mat{Z}$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$\\
 \>\bullet\> inter-packet intervals: \>$\mat{V}$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
@@
 \begin{align}
-\mat{F} = \seq{\mat{T},\mat{S},\mat{D},\mat{A},\mat{B},\mat{Z},\mat{V}} \in \R^{f \times \Sigma},
+\mat{F} = \left[\,\mat{T}~\mat{S}~\mat{D}~\mat{A}~\mat{B}~\mat{Z}~\mat{V}\,\right] \in \R^{f \times \Sigma},
 \end{align}
@@
 We represent the flow type as an indicator vector: a 255-dimensional vector consisting of all zeros except for the index which is the flow type.
-So a \caps{TCP} flow, having protocol type 6, would have the type vector $\vec{t}=\seq{0,0,0,0,0,1,0,\dots,0}$.
+So a \caps{TCP} flow, having protocol type 6, would have the type vector $\vec{t}=\tuple{0,0,0,0,0,1,0,\dots,0}$.
 This representation may immediately strike one as unnecessarily expansive and inefficient since a single byte suffices to represent the protocol type.
@@
 The sum of protocol numbers is meaningless: two \caps{TCP} flows plus five \caps{ICMP} flows does not equal a single \caps{UDP} flow, despite the fact that $2\cdot6+5\cdot1=17$.
-The summation of the corresponding indicator vectors, however, is $\seq{5,0,0,0,0,2,0,\dots}$.
+The summation of the corresponding indicator vectors, however, is $\tuple{5,0,0,0,0,2,0,\dots}$.
 This vector is a histogram indicating the frequency with which each protocol type occurred in the collection of flows over which the sum was taken.
@@
 The first packet is sent at the chosen start time; the size of each packet is sampled from the histogram of packet sizes, $\vec{z}$; the interval until the next packet is determined from the interval vector, $\vec{v}$.
-Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}(k)$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\seq{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
+Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}(k)$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\tuple{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
 \begin{align}
@@
 \begin{align}
-\vec{\sigma}&=\seq{1,2,3,\dots,d_z},\\
-\vec{\tau}&=\seq{\Di\left(k,\fracx{1}{2}\right)}_{k=1}^{d_v}.
+\vec{\sigma}&=\tuple{1,2,3,\dots,d_z},\\
+\vec{\tau}&=\tuple{\Di\left(k,\fracx{1}{2}\right)}_{k=1}^{d_v}.
 \end{align}
@@
 The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $\seq{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
+The pair of matrices $\tuple{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\tuple{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
 \begin{align}
@@
 The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $\seq{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
+The pair of matrices $\tuple{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\tuple{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
 %\begin{align}

### 2008-06-09 paper.tex
Finished Linear Representations section.

@@
 \begin{align}
-\sizes &= \seq{\size_1,\size_2,\dots} \in \SizeSeq, \\
+\sizes &= \seq{\size_1,\size_2,\dots} \in \SizeSeq \\
 \ivals &= \seq{\ival_1,\ival_2,\dots} \in \IvalSeq.
@@
 \begin{align}
-&\forall\: i > \pkts:   \quad \size_i = 0,      \eqn{zero-packets}\\
-&\forall\: i < \pkts:   \quad \ival_i < \infty, \eqn{finite-intervals}\\
-&\forall\: i \ge \pkts: \quad \ival_i = \infty. \eqn{infinite-intervals}
-\end{align}
-% The send time of the $k$th packet can be calculated as
-% \begin{align}
-% \start + \sum_{i=1}^k \ival_i.
-% \end{align}
+&\forall\: k> \pkts:   \quad \size_k = 0      \eqn{zero-packets}\\
+&\forall\: k < \pkts:   \quad \ival_k < \infty \eqn{finite-intervals}\\
+&\forall\: k \ge \pkts: \quad \ival_k = \infty. \eqn{infinite-intervals}
+\end{align}
+The send time of the $k$th packet can be expressed as
+\begin{align}
+\time_k = \start + \sum_{i=1}^{k-1} \ival_i.
+\end{align}
+This is guaranteed to be finite if and only if $k \le \pkts$.
 Now we are ready to formalize the space of flows:
@@
 \subsection{Examples of Linear Representations}
+\sec{linear-representation-examples}
+
+\newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $\pkts$ does not divide $\mathsym{bytes}$ evenly.
+The simplest is to ignore them; we leave leave more complex schemes to the reader.}
 
 Our first example is the packets-bytes-duration (\caps{PBD}) representation.
-The representation is $\pbd: \Traffic \to \R^3$, defined as
-$\pbd(\flow) = \tuple{\pkts,\mathsym{bytes},\mathsym{dur}}$, where
-\begin{align}
-\mathsym{bytes} &= \sum_{i=1}^{\mathclap{\pkts}}{\size_i} &
-\mathsym{dur} &= \sum_{i=1}^{\mathclap{\pkts-1}}{\ival_i}.
-\end{align}
-This representation encapsulates the information needed to reproduce a \caps{CBR} version of each flow:
-each packet should have a $\floor{\mathsym{bytes}/\pkts}$ byte payload, and intervals between packets should be $\mathsym{dur}/\pkts$ seconds.
+The representation is $\pbd: \Traffic \to \R^3$, mapping
+$\flow \mapsto \tuple{p,b,d}$, where
+\begin{align}
+p &= \pkts, &
+b &= \sum_{i=1}^{\mathclap{p}}{\size_i}, &
+d &= \sum_{i=1}^{\mathclap{p-1}}{\ival_i}.
+\end{align}
+This representation encapsulates the information necessary to reproduce a \caps{CBR} version of each flow:
+packets have payloads of $\floor{\fracx{b}{p}}$ bytes, and are sent every $\fracx{d}{p}$ seconds.\pbdnote
+Given two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, their sum as vectors, 
+\begin{align}
+\vec{x}_1+\vec{x}_2=\tuple{p_1+p_2,b_1+b_2,d_1+d_2},
+\end{align}
+provides an appropriate description of the aggregate behavior of the two flows together.
+Similarly, their average, \latin{i.e.} $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
+In contrast, a ``packets-payload-interval'' representation, expressing the \caps{CBR} parameters directly does not have this property:
+the sum of average payloads is not the overall average of payloads;
+likewise with intervals.
+In the \caps{PBD} representation, the linearity property leads to ``natural'' descriptions of aggregate behavior through vector addition, whereas in the packets-payload-interval representation, vector addition is ``unnatural.''
+
+The \caps{PBD} representation is highly lossy: it discards most of the information detailing the behavior of each flow.
+At the other extreme, we can give a ``perfect'' representation of each flow, that allows us to completely reconstruct each flow from its representation.
+With care, the sum of representations of flows can still naturally describe the aggregate behavior of  flows.
+We will define the representation $\Phi: \Traffic \to \FlowVec$, where
+\begin{align}
+\FlowVec = \R^{255} \cross \R^n \cross \R^n \cross L^2[\R] \cross L^2[\R].
+\end{align}
+% Here $\R^X$ is the vector space of functions $f: X \to \R$, which is isomorphic to the standard Euclidean vector space, $\R^\card{X}$.
+Here $L^2[\R]$ is the usual vector space of real functions~\cite{Rudin87}: $\R^\R=\set{f: \R \to \R}$ equipped with the norm,
+\begin{align}
+\norm{f}_2 = \Big[ \int_{-\infty}^{\infty} \abs{f}^2 dx \Big]^{\frac{1}{2}}.
+\end{align}
+We define $\Phi(\flow) = \tuple{\e_\type,\e_\src,\e_\dst,f,g}$, where $\e_k$ is the $k$th standard basis vector, and $f,g \in L^2[\R]$ are
+\begin{align}
+f(t) &= \sum_{k=1}^{\pkts}
+  \begin{cases}
+    1 & \text{if $t \ge \time_k$} \\
+    0 & \text{if $t <   \time_k$}
+  \end{cases} \\
+g(t) &= \sum_{k=1}^{\pkts}
+  \begin{cases}
+    \size_k & \text{if $t \ge \time_k$} \\
+    0       & \text{if $t <   \time_k$}.
+  \end{cases}
+\end{align}
+Thus, $f$ and $g$ are both integer-valued, monotonically increasing step functions, giving the cumulative number of packets sent and total bytes sent, respectively, at time $t$.
+From the representation via $\Phi$ of a single flow, we can readily recover its type, source and destination, and the transmission times and sizes of packets.
+
+Why is $\Phi$ natural in the same sense that the \caps{PBD} representation is natural?
+Suppose we have a traffic sample, $T\in\Traffic$.
+The first component of $\Phi(T)$ is a 255-dimensional vector, whose coordinates are a histogram of the number of times each flow type occurs in $T$.
+Similarly, the second and third components of $\Phi(T)$ are $n$-dimensional histograms of the number of times each node occurs as source and destination.
+The last components are step functions, like $f$ and $g$.
+Vector addition in $L^2[\R]$ dictates that these functions give the total number of cumulative packets and bytes, respectively, over all flows in $T$.
+Thus, summation of vectors under the representation $\Phi$ yields a natural and useful description of the aggregate behavior of many flows.
+
+Consider, for contrast, if we had represented flow type simply by $\type \in \N$ instead of $\e_\type \in \R^{255}$.
+The natural numbers containing $\type$ can readily be embedded into $\R^1$, which is a vector space.
+However, the addition of vectors is highly unnatural.
+The sum of protocol numbers is meaningless:
+two \caps{TCP} flows plus five \caps{ICMP} flows does not equal a single \caps{UDP} flow, despite the fact that $2\cdot6+5\cdot1=17$.
+The summation of vectors in the representation $\Phi$, on the other hand is $5\,\e_1+2\,\e_6$, indicating the frequency with which each type occurred.
+
+Both of the example representations discussed here can be applied in practice.
+The \caps{PBD} representation naturally allows us to compute aggregate and average \caps{CBR} behaviors of collections of flows.
+With a suitable machine encoding of step-functions, $\Phi$ can be implemented and used in practice.
+Representations that preserve even more information about flows can be constructed.
+We turn now in a different direction, however, and introduce a linear representation that strikes a practical balance in fidelity between $\pbd$ and $\Phi$.
 

### 2008-06-09 paper.tex
Revision of Linear Representations.

@@
 \section{Linear Representation of Traffic}
-
-% In this section, we the concept of linear representation of network traffic.
-Before we introduce this concept, we discuss the characteristic features of network traffic from the traffic modeling perspective.
+\label{sec:representations}
 
@@
 its start time;
-and the specific sizes of and intervals between its constituent packets.
+and the specific sizes packets of and intervals between their transmission.
 We begin our exposition of traffic representation by formalizing these properties mathematically as sets:
@@
 Since flows contain only a finite number of packets, there exists a number of packets, $\pkts \in \N$, for each flow.
-Thus, we require that the sequences $\sizes$ and $\ivals$ satisfy the following requirements:
+Since flows are finite, we require that the sequences $\sizes$ and $\ivals$ satisfy the following requirements:
 \begin{align}
@@
 \end{align}
-% We define the function $\pkts:\flow \to \N$ as
-% \begin{align}
-%   \pkts(\flow) \defeq \min\set{i \in \N : \ival_i = \infty}.
-% \end{align}
-% Intuitively, by Equations~\ref{eqn:finite-intervals}~and~\ref{eqn:infinite-intervals}, this gives the number of packets in the flow.
 Let $2^\Flows$ denote the power set of $\Flows$, and $\Traffic \subseteq 2^\Flows$ the set of all finite sets of flows:
@@
 \begin{definition}\defn{linear-representation}
-A \defterm{linear representation} of network traffic is a pair, $\tuple{V,\psi}$, where $V$ is a vector space and $\psi$ is a function, $\psi: \Traffic \to V$, that maps each traffic sample to the sum of the representations its constituent flows:
+A \defterm{linear representation} of network traffic is a pair, $\tuple{V,\psi}$, where $V$ is a vector space and $\psi$ is a function, $\psi: \Traffic \to V$, such that
 \begin{align}\eqn{linearity}
@@
 \end{align}
-% When $T=\emptyset$, the right-hand side is a sum of zero items, defined to be $\vec{0}$, the identity element of $V$.
 \end{definition}
@@
 The careful reader will notice that \emph{any} function $\Flows \to V$ uniquely determines a linear representation:
-\Equation{linearity} defines the extension to all of $\Traffic$.
-Representing flows by arbitrary vectors, however, is not particularly interesting or useful.
-Therefore, we focus on the meaningful representation of collections of flows, that can be factored or explained in terms of their constituent individual flows.
-In the next section, we clarify the concept and application of linear representation through examples.
+\Equation{linearity} uniquely extends the function to all of $\Traffic$.
+Representing flows by arbitrary vectors, however, is not particularly useful or interesting.
+We focus instead on meaningful representations, where collections of flows are represented by vectors that naturally describe the aggregate behavior in some manner.
+In the next section, we clarify these concepts through examples.
 % we present and discuss specific examples of linear representations in order to clarify the concept and its application.
@@
 \newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $\pkts$ does not divide $\mathsym{bytes}$ evenly.
-The simplest is to ignore them; we leave leave more complex schemes to the reader.}
+The simplest is to ignore it; we leave leave more complex schemes to the reader.}
 
@@
 The representation is $\pbd: \Traffic \to \R^3$, mapping
-$\flow \mapsto \tuple{p,b,d}$, where
+$\flow \mapsto \tuple{p,b,d} \in \R^3$, where
 \begin{align}
@@
 Similarly, their average, \latin{i.e.} $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
-In contrast, a ``packets-payload-interval'' representation, expressing the \caps{CBR} parameters directly does not have this property:
+In contrast, a ``packets-payload-interval'' representation, expressing the \caps{CBR} parameters directly, does not have this property:
 the sum of average payloads is not the overall average of payloads;
-likewise with intervals.
-In the \caps{PBD} representation, the linearity property leads to ``natural'' descriptions of aggregate behavior through vector addition, whereas in the packets-payload-interval representation, vector addition is ``unnatural.''
+likewise for intervals.
+In the \caps{PBD} representation, the linearity property leads to natural descriptions of aggregate behavior through vector addition, whereas in the packets-payload-interval representation, vector addition is ``unnatural.''
 
@@
 At the other extreme, we can give a ``perfect'' representation of each flow, that allows us to completely reconstruct each flow from its representation.
-With care, the sum of representations of flows can still naturally describe the aggregate behavior of  flows.
-We will define the representation $\Phi: \Traffic \to \FlowVec$, where
+With care, the sum of representations of flows can still naturally describe their aggregate behavior.
+Let the representation be $\Phi: \Traffic \to \FlowVec$, where
 \begin{align}
@@
 % Here $\R^X$ is the vector space of functions $f: X \to \R$, which is isomorphic to the standard Euclidean vector space, $\R^\card{X}$.
-Here $L^2[\R]$ is the usual vector space of real functions~\cite{Rudin87}: $\R^\R=\set{f: \R \to \R}$ equipped with the norm,
-\begin{align}
-\norm{f}_2 = \Big[ \int_{-\infty}^{\infty} \abs{f}^2 dx \Big]^{\frac{1}{2}}.
-\end{align}
-We define $\Phi(\flow) = \tuple{\e_\type,\e_\src,\e_\dst,f,g}$, where $\e_k$ is the $k$th standard basis vector, and $f,g \in L^2[\R]$ are
+Here $L^2[\R]$ is the vector space of real functions, $\R^\R=\set{f: \R \to \R}$, with the usual norm~\cite{Rudin87}.
+% \begin{align}
+% \norm{f}_2 = \Big[ \int_{-\infty}^{\infty} \abs{f}^2 dx \Big]^{\frac{1}{2}}.
+% \end{align}
+We define
+\begin{align}
+\Phi(\flow) = \tuple{\e_\type,\e_\src,\e_\dst,f,g},
+\end{align}
+where $\e_k$ is the $k$th standard basis vector, and $f,g \in L^2[\R]$ are functions defined by
 \begin{align}
@@
     1 & \text{if $t \ge \time_k$} \\
-    0 & \text{if $t <   \time_k$}
+    0 & \text{if $t <   \time_k$},
   \end{cases} \\
@@
 \end{align}
-Thus, $f$ and $g$ are both integer-valued, monotonically increasing step functions, giving the cumulative number of packets sent and total bytes sent, respectively, at time $t$.
-From the representation via $\Phi$ of a single flow, we can readily recover its type, source and destination, and the transmission times and sizes of packets.
+Thus, $f$ and $g$ are both integer-valued, monotonically increasing step functions, giving the cumulative number of packets and total bytes sent, respectively, at time $t$.
+From the representation, $\Phi(\flow)$, of a single flow, we can readily recover its type, source, destination, and the transmission times and sizes of packets.
 
@@
 The last components are step functions, like $f$ and $g$.
-Vector addition in $L^2[\R]$ dictates that these functions give the total number of cumulative packets and bytes, respectively, over all flows in $T$.
-Thus, summation of vectors under the representation $\Phi$ yields a natural and useful description of the aggregate behavior of many flows.
+Vector addition in $L^2[\R]$ dictates that these functions give the total number of cumulative packets and bytes sent, respectively, over all flows in $T$.
+Thus, summation of vectors under the representation $\Phi$ yields a natural and useful description of the aggregate behavior of flows.
 
@@
 With a suitable machine encoding of step-functions, $\Phi$ can be implemented and used in practice.
-Representations that preserve even more information about flows can be constructed.
-We turn now in a different direction, however, and introduce a linear representation that strikes a practical balance in fidelity between $\pbd$ and $\Phi$.
+It provides complete information about individual flows and highly detailed (though not complete) information about collections of flows.
+Representations that preserve even more information about traffic instances can readily be constructed.
+We turn now, however, in a different direction, and introduce a specific linear representation that strikes a practical balance in fidelity between $\pbd$ and $\Phi$.
 
@@
 \label{sec:general-matrix-model}
-\label{sec:representations}
 

### 2008-06-09 paper.tex
Work on GMM.

@@
 In the next section, we clarify these concepts through examples.
-% we present and discuss specific examples of linear representations in order to clarify the concept and its application.
 
@@
 
-%The traffic workload pattern in a network is the collection of actions taken by hosts, on behalf of users and applications, that affect the behavior of the network itself.
-By design, Internet Protocol (\caps{IP}) networks treat all data in the same manner~\cite{Clark88}.
-Therefore, the actual information transmitted is irrelevant, only the manner of transmission matters.
-Although networks are packet-switched, packets are logically organized into sequences of packets sharing the same \caps{IP} protocol, source and destination \caps{IP} address, and source and destination  \caps{TCP} or \caps{UDP} port numbers.
-Such a sequence of packets is called a \textit{flow}.
-Although there are a vast multitude different application types which use \caps{IP}, there are two fundamentally different behaviors of flows which we distinguish here: \caps{TCP}, which delivers data reliably through feedback and retransmissions; and \caps{UDP}, which delivers data only on a best-effort basis.
-While it is possible to implement reliable transmission similar to \caps{TCP} over \caps{UDP}, it is uncommon, so we make the generalization that network events, such as packet drops, do not significantly feedback to the behavior of upper layers.
-With \caps{TCP}, on the other hand, network events are known to affect the transport layer
-
-% TODO: talk about other vector-space representation: perfect representation, markov models, ... others?
-
-The traffic workload pattern in a network is a collection of \caps{IP} packets between hosts in the network sent at certain times.
-It is standard in network analysis, however, to aggregate sequences of packets sharing the same \caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP} or \caps{UDP} port numbers.
-Such a sequence of packets is called a \textit{flow}.
-The traffic pattern of an entire network is simply the collected behaviors of all flows occurring in the network.
-The behavior of a flow as it affects the network is characterized by the following properties: its \caps{IP} protocol type (\caps{TCP}, \caps{UDP}, \caps{ICMP}, etc.); its source and destination nodes; its start time; and the specific sizes of and intervals between its constituent packets.
-
-We propose to represent the behavior of each flow in terms of seven vectors, which suffice to describe all of the above properties in enough detail to stochastically generate realistic traffic.
-The aggregate behavior of all the flows in a workload pattern are then represented as matrices, the rows of which are the behavior vectors of the flows.
-
-When mapping flow behaviors into a vector space, it is important that the mapping of behaviors into the vector spaces in question respect the vector space operations: addition of vectors and scalar multiplication.
-This requirement is best clarified with a simplified example.
-Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and its total duration, $d$.
-Despite its simplicity, this representation is adequate to reproduce constant bit-rate (\caps{CBR}), time-randomized flow behavior.
-The triplet of representative values can be naturally embedded into Euclidean space: $\vec{x}=\tuple{p,b,d}\in\R^3$.
-If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their sum as vectors, $x_1+x_2=\tuple{p_1+p_2,b_1+b_2,d_1+d_2}$, provides an appropriate description of the aggregate behavior of the two flows together.
-Similarly, their average, i.e.
-$\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
-
-%The behavior of a flow may be mapped into a vector space such that the two operations, namely vector addition and scalar multiplication, naturally correspond with the aggregation of behaviors.
-This requirement is best clarified with an intentionally simplistic example.
-Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and the total duration, $d$.
-This triplet can be naturally embedded into Euclidean space: $\vec{x}=(p,b,d)\in\R^3$.
-If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum'' behavior, $\vec{x}_1+\vec{x}_2$ has $p_1+p_2$ packets, $b_1+b_2$ bytes, $d_1+d_2$ duration.
-Their average, i.e.
-$\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
-
-%While this may seem trivial, note that the representation must be chosen carefully.
-Suppose instead, for example, we represented each flow's behavior by its packet count, $p$, average packet size, $b/p$, and average inter-packet interval, $d/(p-1)$.
-In this representation, the average of two behaviors has the \textit{average} average packet size, and the \textit{average} average inter-packet interval.
-Therein lies the rub: this is the case no matter how many packets are in each flow; if a flow with thousands of 1500-byte packets is averaged with a single-packet, one-byte flow, the average behavior has...
-Instead, we should more naturally weight the averages by the number of packets in each constituent.
-This is precisely what the packets-bytes-duration representation accomplishes.
-
-The {packets-bytes-duration} representation is clearly overly simplistic and very incomplete.
-It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time and flow size unspecified; it says nothing about the source or destination of the flow.
-At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets, etc.
-into a vector space.
-The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows.
-We take a highly detailed but intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the flow type, source node, destination node, start time, flow size, and empirical distributions of packet sizes and inter-packet intervals.
+The \caps{PBD} representation of flows uses 3 dimensions to encode the behavior of each flow, while $\Phi$ uses $255+2n$ dimensions, plus the continuum of dimensions of two Banach spaces (\latin{i.e.} the two copies of $L^2[\R]$).
+In this section, we define a representation, $\phi$, that splits the difference and uses a large but finite number of dimensions to represent each flow.
+Once the vector representation of each flow is defined, we may construct matrices of representation vectors, with a row for each flow in a traffic sample.
+When network behavior is expresses as matrices, we can very simply and efficiently compute a broad variety of network behavior properties using standard matrix operations.
+Furthermore, we can express uniformity and marginality assumptions about behavior as equality constraints between the rows and columns of the traffic matrix.
+Moreover, we can transform unconstrained traffic instances into uniform, marginal, or conditional approximations via right or left multiplications by easily computed matrices.
+
+The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins, and represented each bin by a standard basis element of $\R^m$, where $m$ is the number of bins for that property.
+Representation vectors are histograms counting items falling into each bin.
+The histograms for packet sizes and inter-packet intervals count packets, while other histograms count flows.
+
 If there are $f$ flows in the entire workload pattern, then the collective behavior can be described with seven matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.

### 2008-06-09 paper.tex
GMM properties table.
Rewritten exposition of GMM.

@@
 \begin{align}
-&\forall\: k> \pkts:   \quad \size_k = 0      \eqn{zero-packets}\\
-&\forall\: k < \pkts:   \quad \ival_k < \infty \eqn{finite-intervals}\\
+&\forall\: k \le \pkts: \quad \size_k > 0       \eqn{positive-packets}  \\
+&\forall\: k >   \pkts: \quad \size_k = 0       \eqn{zero-packets}      \\
+&\forall\: k <   \pkts: \quad \ival_k < \infty  \eqn{finite-intervals}  \\
 &\forall\: k \ge \pkts: \quad \ival_k = \infty. \eqn{infinite-intervals}
 \end{align}
+Packet sizes are required to be positive because we are modeling application layer data transfer:
+an empty data transmission request is a non-event.
 The send time of the $k$th packet can be expressed as
@@
 \section{The General Matrix Model}
-\label{sec:general-matrix-model}
+\sec{general-matrix-model}
+
+\input{gmm_properties}
 
@@
 Once the vector representation of each flow is defined, we may construct matrices of representation vectors, with a row for each flow in a traffic sample.
-When network behavior is expresses as matrices, we can very simply and efficiently compute a broad variety of network behavior properties using standard matrix operations.
+We call the matrix representation of network traffic, the general matrix model~(\caps{GMM}).
+With network behavior is expresses as matrices, we can very simply and efficiently compute a broad variety of network behavior properties using standard matrix operations.
 Furthermore, we can express uniformity and marginality assumptions about behavior as equality constraints between the rows and columns of the traffic matrix.
@@
 
-The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins, and represented each bin by a standard basis element of $\R^m$, where $m$ is the number of bins for that property.
-Representation vectors are histograms counting items falling into each bin.
+The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins, and represented each bin by a standard basis element of $\R^d$, where $d$ is the number of bins for that property.
+In this scheme, representation vectors are histograms counting items falling into each bin.
 The histograms for packet sizes and inter-packet intervals count packets, while other histograms count flows.
-
-If there are $f$ flows in the entire workload pattern, then the collective behavior can be described with seven matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
-We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
+\Table{quantization} lists the properties that are used to define $\phi$, together with the quantization and dequantization functions.
+Each quantization function takes a possible property value as input and maps it to an index value in $\set{1,\dots,d}$, where $d$ is the dimension for the property in question.
+The index assigned determines the histogram bin of the input property.
+
+The histogram vector approach produces some representations of properties that are not immediately intuitive, and may seem to use an excessive number of representation dimensions.
+However, this approach automatically satisfies the ``naturalness'' criterion for linear representations.
+That is, the representation, $\phi(T)$, of some collections of flows, $T$, naturally describes the aggregate behavior of those flows.
+Perhaps even more importantly, arbitrary linear combinations of vectors can be meaningfully interpreted as describing some realizable traffic instance.
+In \Section{traffic-generation} we will describe how to choose, stochastically, a collection of flows matching the ``description'' of an arbitrary traffic matrix.
+% This is how we generate workload from transformed traffic matrices.
+
+The quantization functions specified in \Table{quantization} allow us to define the representation $\phi$: \begin{align}
+\phi(\flow) = \tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}},
+\end{align}
+where the component vectors are defined as
+\begin{align*}
+\vec{t} &= \e_k & k &= \Qf^{\text{t}}(\type) \\
+\vec{s} &= \e_k & k &= \Qf^{\text{s}}(\src) \\
+\vec{d} &= \e_k & k &= \Qf^{\text{d}}(\dst) \\
+\vec{a} &= \e_k & k &= \Qf^{\text{a}}(\start) \\
+\vec{b} &= \e_k & k &= \Qf^{\text{b}}(\bytes) \\
+\vec{z} &= {\textstyle \sum \e_k} & k &\in \set{\Qf^{\text{z}}(\size_i) : i \le \pkts} \\
+\vec{v} &= {\textstyle \sum \e_k} & k &\in \set{\Qf^{\text{v}}(\ival_i) : i <   \pkts}.
+\end{align*}
+Here, the functions $\Qf^\text{x}$ are the respective quantization functions for each property.
+If there are $f$ flows in a traffic instance, then the collective behavior can be described with seven matrices---one for each property;
+the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
+We use the following notations for the behavior matrices:
+\vspace{-0.25em}
 \begin{tabbing}
@@
 \end{tabbing}
-The total behavior is the horizontal concatenation of these component matrices:
-\begin{align}
-\mat{F} = \left[\,\mat{T}~\mat{S}~\mat{D}~\mat{A}~\mat{B}~\mat{Z}~\mat{V}\,\right] \in \R^{f \times \Sigma},
-\end{align}
-where $\Sigma=255+2n+d_t+d_b+d_z+d_v$.
-Many common traffic models can be expressed succinctly and precisely in terms of transformations or conditions on this behavior matrix and its constituents.
-In the following sections, we explain in detail how each flow's behavior is represented in these matrices.
-Finally, we describe how a model instance stochastically generates network traffic.
-
-\subsection{Flow Type}
-
-The type of a flow is its \caps{IP} protocol number.
-This ranges in possible values from 1 to 255 (0 is reserved for \caps{IPv6}).
-We represent the flow type as an indicator vector: a 255-dimensional vector consisting of all zeros except for the index which is the flow type.
-So a \caps{TCP} flow, having protocol type 6, would have the type vector $\vec{t}=\tuple{0,0,0,0,0,1,0,\dots,0}$.
-This representation may immediately strike one as unnecessarily expansive and inefficient since a single byte suffices to represent the protocol type.
-The necessity of this representation arises when we represent aggregations of flows as sums of vectors.
-The sum of protocol numbers is meaningless: two \caps{TCP} flows plus five \caps{ICMP} flows does not equal a single \caps{UDP} flow, despite the fact that $2\cdot6+5\cdot1=17$.
-The summation of the corresponding indicator vectors, however, is $\tuple{5,0,0,0,0,2,0,\dots}$.
-This vector is a histogram indicating the frequency with which each protocol type occurred in the collection of flows over which the sum was taken.
-
-\subsection{Source \& Destination Nodes}
-
-Let $n$ be the number of nodes in the traffic collection to be modeled.
-We let $s_{ij}=1$ if the $j$th node is the source of the $i$th flow, and zero otherwise; similarly, $d_{ik}=1$ if the destination of the $i$th flow is the $k$th node.
-This sparse matrix may again seem to be an overly expansive representation, given that start and end nodes can be represented simply as indices from $\set{1,\dots,n}$.
-The sum of a set of source vectors, $\set{\vec{s}_i}$, however, is an $n$-dimensional histogram indicating how often each source node occurs in the collection.
-
-\subsection{Start Time}
-
-We represent start time of each flow using an indicator vector.
-Unlike node indices, however, time is continuous, so we must first map times onto a range of time indices $\set{1,\dots,d_t}$.
-We transform the range $[0,t_{\max}]$ of possible start times using a simple quantization function:
-\begin{align}
-\label{eqn:time-quantization}
-\Qt(t) &= \max\set{1,\ceil{d_t\fracpx{t}{t_{\max}}}}.
-\end{align}
-%Suppose the $i$th flow has a quantized start index of $j$ and a end index of $k$. Then $b_{ij'}=\delta{jj'}$ and $e_{ik'}=\delta_{kk'}$.
-Let $k$ be the quantized start index of a flow.
-Then the start vectors for the flow is $\vec{a}=\vec{e}_k$ (the $k$th standard basis vector, having 1 at index $k$ and zeros elsewhere).
-The advantage of this representation becomes clear when we aggregate flows: the sum of start vectors over a collection of flows is a $d_t$-dimensional histogram indicating the frequency of start times occurring in each quantization bin.
-In practice, we use $d_t=1500$.
-
-\subsection{Flow Size}
-
-We represent flow size with another indicator vector.
-Although the set of possible flow sizes is discrete, for practical reasons we map the very large range of possible flow sizes down to a much smaller range of indices (some flows in our trace have 109\caps{MB} of data in a single ten-minute period, which would give an indicator vector of over 100 million entries).
-We use the following function to convert flow sizes to indexes:
-\begin{align*}
-%\label{eqn:byte-quantization}
-\Qb(b) &= \max\set{1,\ceil{d_b\fracp{b-1}{b_{\max}-1}^\beta}}.
-\end{align*}
-Here $b_{\max}$ is the maximum input flow size.
-The parameter $\beta$ is a non-linear scaling factor exponent that allows the quantization to smoothly shift from byte-sized resolution at the low end of the possible range of flow sizes to the a resolution of megabytes at the high end of the spectrum.
-In our example data, $b_{\max}\approx109\text{\caps{MB}}$ as we previously mentioned.
-We use a representation dimensionality of $d_b=1500$, to match our other quantized vectors, and $\beta=\fracx{1}{2.7}$ because it yields the desired dynamic range response of the possible flow sizes.
-
-\subsection{Packet Sizes \& Inter-Packet Intervals}
-
-The packet size behavior of a flow is represented by a histogram vector, $\vec{z}$, of packet sizes.
-Explicitly,  $z_{ij}$ is the number of packets in the $i$th flow with $j$ bytes of payload data.
-The dimension, $d_z$, is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium, typically 1500 bytes.
-The representation of inter-packet intervals as a vector, $\vec{v}$, is similar, but the domain of possible intervals is a continuum, so we must map this range onto the discrete index values $\left\{1,\dots,d_v\right\}$.
-We use similar non-linear quantization function as before:
-\begin{align}
-\label{eqn:interval-quantization}
-\Qi(v) &= \max\set{1,\ceil{d_v\fracp{v-v_{\min}}{v_{\max}-v_{\min}}^\gamma}}.
-\end{align}
-The element $v_{ij}$ is the number of intervals, $v$, for the $i$th flow such that $\Qi(v)=i$.
-The parameter values defining $\Qi$ must be chosen by the experimenter.
-In our case, the time measurements had microsecond precision, so $v_{\min}=10^{-6}$.
-The length of each simulated scenario is ten minutes, so the upper limit on intervals is $v_{\max}=600$.
-The number of dimensions was chosen to match the dimensions for packet size: $d_v=1500$.
-Finally, the parameter $\gamma\in(0,1)$ allows us to fine-tune how the resolution scales with interval size; we have chosen $\gamma=\frac{1}{3}$ because this choice provides resolution that scales smoothly from the microseconds for small intervals to seconds for large intervals.
+The total behavior matrix is the horizontal concatenation of these component matrices:
+\begin{align}
+\mat{F} = \left[\,\mat{T}~\mat{S}~\mat{D}~\mat{A}~\mat{B}~\mat{Z}~\mat{V}\,\right] \in \R^{f \times N},
+\end{align}
+where $N=255+2n+d_t+d_b+d_z+d_v$.
+% Many common traffic modeling assumptions can be expressed succinctly and precisely in terms of transformations or conditions on this behavior matrix and its components.
+
+Before continuing on to less mundane topics, we must briefly address our choices of constants appearing in \Table{quantization}.
+The externally predetermined constants are:
+$d_z=\text{\caps{MTU}}=1500$, determined by the wireless network medium;
+$b_{\max} \approx 109 \text{\caps{MB}}$, determined by the size of the largest flow in our trace data;
+$t_{\max}=v_{\max}=600$, determined by our experimental methodology which slices traces into 10-minute scenarios;
+and $v_{\min}=10^{-6}$, determined by the time resolution of our trace.
+For lack of a better choice, we have simply chosen to use the same number of quantization bins as packet size for the unfixed dimension constants:
+$d_t = d_b = d_v = 1500$.
+Arguments can certainly be made for other choices, but this choice appears to work adequately in practice.
+
+Two other unspecified constants appear in \Table{quantization}: $\beta$ and $\gamma$.
+These parameters are non-linear scaling exponents that allows the quantization to smoothly shift from fine-grained resolution at the low end of the spectrum to coarse-grained resolution at the high end.
+We have chosen $\beta=\fracx{1}{2.7}$ because it yields a desirable dynamic range response from bytes at the low end to megabytes at the high end.
+Similarly, we have chosen $\gamma=\fracx{1}{3}$ because this choice provides resolution that scales smoothly from microseconds for small intervals to seconds for large intervals.
 

### 2008-06-09 paper.tex
Finished Traffic Generation.

@@
 The quantization functions specified in \Table{quantization} allow us to define the representation $\phi$: \begin{align}
-\phi(\flow) = \tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}},
+\phi(\flow) = \vec{f} = \tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}},
 \end{align}
@@
 where $N=255+2n+d_t+d_b+d_z+d_v$.
-% Many common traffic modeling assumptions can be expressed succinctly and precisely in terms of transformations or conditions on this behavior matrix and its components.
-
-Before continuing on to less mundane topics, we must briefly address our choices of constants appearing in \Table{quantization}.
+We call the representation of whole-network traffic instances as such a matrix, the general matrix model.
+While it is by no means the only possible representation of traffic in terms of matrices, many common traffic modeling assumptions can be expressed succinctly and precisely in terms of transformations or conditions of the \caps{GMM} and its components.
+Furthermore, realistic traffic can be generated from trace-derived matrix instances.
+
+\subsection{Choosing Constants}
+
+Before continuing with less mundane topics, we must briefly address our choices of constants appearing in \Table{quantization}.
 The externally predetermined constants are:
@@
 \subsection{Traffic Generation}
-\label{sec:traffic-generation}
+\sec{traffic-generation}
 
@@
 Each row of the model is used to randomly generate the behavior of a single flow.
-The type, source, and destination nodes of the flow are randomly chosen, using histogram vectors $\vec{t}$, $\vec{s}$, and $\vec{d}$ as the proportional probabilities of each type or node being chosen.\footnote{Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
+To do this, we use the ``dequantization'' functions listed in \Table{quantization}.
+The dequantization procedure works as follows.
+Let $\vec{f}$
+% \begin{align}
+% \vec{f}=\tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}}
+% \end{align}
+be a row of the traffic matrix, and let $\prop$ be a property.
+Write $\vec{f}_\prop$ for the corresponding component vector of $\vec{f}$.
+Chose an index value $k\in\set{1,\dots,d_\prop}$, randomly, with weight given by the coefficients of $\vec{f}_\prop$.
+Choose $u \in [0,1]$ uniformly randomly.
+Then $x = k-u \in [0,d]$, with probability of lying in the interval $(i-1,i]$ proportional to the coefficient of $\e_i$ in $\vec{f}_\prop$.
+The dequantization function for $\prop$, call it $\Df_\prop$, will map the value $x$ back into the appropriate range of property values for $\prop$.
+
+The above procedure allows us to sample random values for each property in accordance with the behavioral description implied by $\vec{f}$.
+This allows us to stochastically generate a sequence of packets in accordance with a row from an arbitrary \caps{GMM} matrix instance using the following procedure:
+\begin{enumerate}
+\item Sample random source and destination node pairs.
+Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
+The simplest technique is the rejection method:
+if the same node is chosen for both source and destination, reject that pair and begin the selection process again.
+\item Next, randomly sample the number of bytes.
+\item The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\mean{z}$ and $\mean{v}$.
+From these, together with the flow size in bytes, we estimate the expected duration of the flow:
+$\mean{d} = \fracx{b\mean{v}}{\mean{z}}$.
+\item Randomly sample a start time for the flow.
+Ensure that the start time plus the expected duration, $\mean{d}$, does not exceed the maximum simulation time for the workflow by using the rejection method.
+\item The first packet is sent at the chosen start time.
+The size of each packet is determined by sampling a random size value.
+The interval until the next packet is determined by sampling a random interval value.
+The packet stream ends when the number of bytes allocated to the flow have been emitted or when the packet emission time exceeds $t_{\max}$.
+The last packet is only the size of the remaining number of bytes.
+\end{enumerate}
+Some remarks about the procedure:
+\begin{itemize}
+\item For some vectors in $\R^{f \cross N}$, it may be impossible to choose distinct source and destination nodes or flow sizes and start times satisfying the procedure.
+However, for any vector that is a sum of vectors representing valid flows---\latin{i.e.} ones having distinct source and destination nodes and feasible combinations of start time, duration, and packet behaviors---the procedure will terminate.
+\item Because the representation $\phi$ models the seven properties independent of each other, it is impossible to do better than to sample the properties independently, modulo the requirements of the procedure.
+In the case of a row which is simple a representation of an actual flow, this makes no difference, except for the generation of packets (which are modeled independently).
+In the case of composite rows, however, this implies that flow properties are modeled independent of each other.
+\end{itemize}
+Despite these remarks, we will show in our experimental evaluation, that this procedure generates sufficiently realistic wireless workloads.
+
+% The type, source, and destination nodes of the flow are randomly chosen, using histogram vectors $\vec{t}$, $\vec{s}$, and $\vec{d}$ as the proportional probabilities of each type or node being chosen.\footnote{Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
 The simplest technique is the rejection method: if the same node is chosen for both source and destination, reject that pair and begin the selection process again.}
-Next we use the flow size vector, $\vec{b}$ to randomly sample the number of bytes in the flow.
+% Next we use the flow size vector, $\vec{b}$ to randomly sample the number of bytes in the flow.
 The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\mean{z}$ and $\mean{v}$.
@@
 From the range of possible values mapping to the chosen start bin, we choose a random time uniformly.
-
-Having determined the overall parameters of the flow, we generate the actual packets of the flow.
+% 
+% Having determined the overall parameters of the flow, we generate the actual packets of the flow.
 The first packet is sent at the chosen start time; the size of each packet is sampled from the histogram of packet sizes, $\vec{z}$; the interval until the next packet is determined from the interval vector, $\vec{v}$.
 Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}(k)$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\tuple{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
-\begin{align}
-\label{eqn:interval-dequantization}
-\Di(k,u)=v_{\min}+(v_{\max}-v_{\min})\left(\frac{k-u}{d_v}\right)^{\frac{1}{\gamma}}.
-\end{align}
-The packet stream ends when the number of bytes allocated to the flow have been emitted. %The last packet is only the size of the remaining number of bytes.
+% \begin{align}
+% \label{eqn:interval-dequantization}
+% \Di(k,u)=v_{\min}+(v_{\max}-v_{\min})\left(\frac{k-u}{d_v}\right)^{\frac{1}{\gamma}}.
+% \end{align}
+% The packet stream ends when the number of bytes allocated to the flow have been emitted. %The last packet is only the size of the remaining number of bytes.
 
@@
 
-The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{SD}}\in\R^{n \times n}$ ($\trans{\,\cdot\,}$~denotes transpose).
+The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{SD}}\in\R^{n \times n}$ ($\trans{\,\diamond\,}$~denotes transpose).
 That is $(\trans{\mat{SD}})_{ij}$ is the number of flows from node $i$ to node $j$.

### 2008-06-09 paper.tex
Tweaking rest of paper.

@@
 Compare this with the typical situation in wired \caps{LAN}s, where gigabit Ethernet and modern routers  can easily handle all but the most intense levels of traffic.
-If a wired network operator expects larger volumes of traffic on a regular basis, the solution is to upgrade the hardware, not to change the network protocols.
+If a wired network operator expects larger volumes of traffic on a conditional basis, the solution is to upgrade the hardware, not to change the network protocols.
 In wireless networks, on the other hand, the physical medium cannot be upgraded; protocols must be improved to make better usage of the medium.
@@
 
-% The type, source, and destination nodes of the flow are randomly chosen, using histogram vectors $\vec{t}$, $\vec{s}$, and $\vec{d}$ as the proportional probabilities of each type or node being chosen.\footnote{Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
-The simplest technique is the rejection method: if the same node is chosen for both source and destination, reject that pair and begin the selection process again.}
-% Next we use the flow size vector, $\vec{b}$ to randomly sample the number of bytes in the flow.
-The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\mean{z}$ and $\mean{v}$.
-From these, together with the flow size in bytes, we estimate the expected duration of the flow: $\mean{d} = \fracx{b\mean{v}}{\mean{z}}$.
-We finally choose a start time for the flow, using $\vec{a}$ as a vector of relative probabilities for each bin of start times, but ensuring that the start time plus the expected duration, $\mean{d}$, does not exceed the maximum simulation time for the workflow.
-From the range of possible values mapping to the chosen start bin, we choose a random time uniformly.
-% 
-% Having determined the overall parameters of the flow, we generate the actual packets of the flow.
-The first packet is sent at the chosen start time; the size of each packet is sampled from the histogram of packet sizes, $\vec{z}$; the interval until the next packet is determined from the interval vector, $\vec{v}$.
-Specifically, an interval index, $k$, is chosen with probability proportional to $\vec{v}(k)$; a second value, $u$, is chosen uniformly from $(0,1]$; the pair $\tuple{k,u}$ is transformed back into a continuous time value using the ``dequantization'' function:
-% \begin{align}
-% \label{eqn:interval-dequantization}
-% \Di(k,u)=v_{\min}+(v_{\max}-v_{\min})\left(\frac{k-u}{d_v}\right)^{\frac{1}{\gamma}}.
-% \end{align}
-% The packet stream ends when the number of bytes allocated to the flow have been emitted. %The last packet is only the size of the remaining number of bytes.
-
-\section{Expression of Common Properties}
+\subsection{Expression of Common Properties}
 \label{sec:common-properties}
 
-Having defined our representation of network behavior, we turn to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.
-
-\subsection{Communication Topology}
+Having defined our representation of network behavior and presented the procedure for stochastically generating workload, we turn to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.
+
+\subsubsection{Communication Topology}
 \label{sec:communication-topology}
@@
 
-\subsection{Expected Average Packet Size \& Intervals}
+\subsubsection{Expected Average Packet Size \& Intervals}
 \label{sec:flow-totals}
@@
 The computations require knowing the values that each matrix column represents.
-The computed duration is only approximate because information is irrecoverably lost in the quantization process (see \Equation{interval-quantization}).
+The computed duration is approximate because information is irrecoverably lost in the quantization process.
 The vectors of representative values for $\mat{Z}$ and $\mat{V}$ are:
@@
 
-\section{Expression of Common Modeling Concepts}
+\subsection{Expression of Common Modeling Concepts}
 \label{sec:modeling-concepts}
@@
 The simplifying assumptions made by most traffic models can be classified into three categories.
-We use the terms \textit{uniformity}, \textit{marginality}, and \textit{regularity} to describe these classes of assumptions.
+We use the terms \textit{uniformity}, \textit{marginality}, and \textit{conditionality} to describe these classes of assumptions.
 They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether the make stipulations about uniformity within each flow, across all flows, or within groups of flows.
@@
 
-\subsection{Uniform Behavior}
+\subsubsection{Uniformity}
 
@@
 The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $\tuple{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\tuple{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
+The pair of matrices $[\mat{Z}~\mat{V}]$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $[\mat{Z}~\mat{V}]\,\Gamma\in\R^{f \times 3}$, where
 \begin{align}
@@
 
-\subsection{Marginal Behavior}
+\subsubsection{Marginality}
 
@@
 
-\subsection{Regularized Behavior}
-
-The final form of homogenization we consider, is regularization.
+\subsection{Conditionality}
+
+The final form of homogenization we consider, is conditionality.
 This is a restricted form of marginality: we may assume that marginal distributions apply to all the flows in certain groupings.
@@
 This is simply accomplished by left-multiplication by a non-uniform matrix.
-For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-uniformizing'' matrix: $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$.
-Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-uniformizing matrix: $F'=(\trans{\mat{DD}}/f)\mat{F}$.
+For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-conditioning'' matrix: $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$.
+Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-conditioning matrix: $F'=(\trans{\mat{DD}}/f)\mat{F}$.
 %Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified.
@@
 The matrix model provides an analytical framework for doing so.
-The matrix model has the benefit of making the uniformization of behaviors completely explicit.
+The matrix model has the benefit of making the conditioning of behaviors completely explicit.
 It also allows us to easily define intermediate choices.
@@
 %A generic traffic model is a way of generating the six matrices described in \Section{representations}.
-A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or regularized form.
+A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or conditionalized form.
 For experimental comparison, the dimensions of the matrices should all be preserved.
@@
 %% TODO: differentiate between restrictive versions of models and transformative versions.
-%% TODO: distinguish uniform, marginal and regularized models.
+%% TODO: distinguish uniform, marginal and conditionalized models.
 
@@
 full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline\hline
-source regular & $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$ & flow behaviors are aggregated on a per source-node basis \\\hline
-dest regular & $\mat{F}'=(\trans{\mat{DD}}/f)\mat{F}$ & flow behaviors are aggregated on a per destination-node basis \\\hline
-mixed regular & $\mat{F}'=\frac{1}{2f}(\trans{\mat{SS}}+\trans{\mat{DD}})\mat{F}$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
+source conditional & $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$ & flow behaviors are aggregated on a per source-node basis \\\hline
+dest conditional & $\mat{F}'=(\trans{\mat{DD}}/f)\mat{F}$ & flow behaviors are aggregated on a per destination-node basis \\\hline
+mixed conditional & $\mat{F}'=\frac{1}{2f}(\trans{\mat{SS}}+\trans{\mat{DD}})\mat{F}$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
 \end{tabular}
@@
 Our simulation methodology is detailed in \Section{simulations}, and the performance metrics used for statistical comparison of control and test experiments are described in \Section{performance-metrics}.
-Finally, \Section{measures-of-error} is an in-depth mathematical discussion of the appropriate measure of error for the pairs of metric values that are produced by the paired differential simulation methodology.
+% Finally, \Section{measures-of-error} is an in-depth mathematical discussion of the appropriate measure of error for the pairs of metric values that are produced by the paired differential simulation methodology.
 
@@
 
-After validating the \caps{GMM}, we explore the effects of various uniformity, marginality, and regularization assumptions on the realism of traffic models.
+After validating the \caps{GMM}, we explore the effects of various uniformity, marginality, and conditionalization assumptions on the realism of traffic models.
 The uniform models are the most common, being the simplest both to conceptualize and to implement: simply make no distinction between start times, source or destination nodes.
@@
 
-\begin{figure}[t]
-\vspace{0.5em}
+\begin{figure}[!b]
 \begin{center}
@@
 \end{center}
-\vspace{-2em}
+\vspace{-1.3em}
 \end{figure}
@@
 
+Our simulation results are summarized in \Figure{box-plots}.
+Each subfigure shows a single performance metric.
+The distribution of log-ratio error values for each traffic model is visualized with a box-and-whisker plot.
+The box indicates the range from the 25th to 75th percentiles of values, while the ``whiskers'' indicate the full range, excluding outliners (which are shown as isolated points).
+These plots allow immediate assessment of realism: a good traffic model should have error values that are tightly clustered around the center, with a small, evenly balanced box, and relatively small whiskers.
+Additionally, the mean and median markers should be close to the center.
+It should be noted that the error scale is logarithmic, so even a slight increase in spread or deviation from the center indicates a disproportionately large decrease in model quality.
+Non-logarithmic scale markers are shown above each plot; the values indicate the factor of over- or under-representation for the performance metric.
+
+The primary result is that the \caps{GMM} accurately reflects the performance of the trace data it is based on.
+It has small, centered error bars for every performance metric presented here, as well as for those we have looked at otherwise.
+This is an essential result, since otherwise, \caps{GMM} is at best a shaky foundation for further modeling work.
+With these results, we can be assured that if we can approximate the matrix for a given example of trace traffic, then we have also approximated the original trace behavior.
+In a sense, this result is a reduction of the general traffic modeling problem to a more tractable matrix modeling problem.
+
+It is worth noting that \caps{GMM} has competition from a few of the time-simplified models: the \model{time marginal} and \model{time uniform} model both do approximately as well, and in some cases better, for the metrics examined.
+{\FHC} found strong evidence that session arrivals followed a time-varying Poisson arrival process, meaning that other than long-scale time-varying arrival rate and very small-scale clustering of flows within sessions, the overall flow arrival rate is fairly smooth.
+We suspect that on the scale of minutes to tens of minutes in which our simulation exist, it is sufficient to model both flow and session arrivals uniformly.
+This result indicates that the precise temporal placement of flows is not highly sensitive: they tend to be relatively evenly spaced out at short time-scales, and network performance is not sensitive to changes in start-time on that scale.
+
+Not all marginal models have such a benign effect on the accuracy of performance metrics.
+The worst offender by far is the \model{size marginal} model.
+This model distorts behavior at every level, by more than a factor 20 in 25\% of scenarios in the case of routing queue packet drops.
+This is a highly significant result because modeling flow size using a network-wide marginal distribution is precisely what {\FHC}, for example, attempt.
+This results informs us that this approach is doomed to failure: flow size cannot be assigned without considering its relation to other flow properties: source and destination nodes, and packet behavior at the very least.
+
+The \model{packet marginal} model is another offender, albeit not as egregiously.
+The direction of the misrepresentations in this case, however, is more dangerous: received throughput is overestimated by more than three times, 75\% of the time.
+On the other hand, \caps{AODV} control overhead is underestimated by half on average.
+Such serious misrepresentations are especially noteworthy given the apparently innocuous assumption applied by this model: the \model{packet marginal} model is essentially a standard variable bit-rate (\caps{VBR}) packet behavior model, using the empirical network-wide distributions of packet sizes and inter-packet intervals.
+Otherwise the model matches the original trace behavior exactly.
+And yet these drastic distortions occur.
+In our previous work, we have shown similarly drastic misrepresentations to occur when constant bit-rate packet behavior is assumed.
+This results indicates that to reproduce accurate network performance, we must provide packet behaviors on a more granular level: each node or even each flow must be assigned a ``custom'' packet behavior by some means.
+
+These strong negative results for two fundamental and very common marginal models provide severe limits on the realism that can be achieved through marginal modeling.
+Traffic modeling research \textit{must} move beyond simply finding parametric models for marginal distributions of network-wide properties.
+Realistic behavior cannot be generated using these distributions alone.
+Something must be captured about the interaction between the elements of flow behavior.
+
+Finally, we turn to the conditionalized models.
+These demonstrate neither exceptionally good nor exceptionally bad behavior.
+That the conditionalized models should be similar to the \model{full marginal} model but somewhat more realistic is entirely expected: conditionalization, the reader will recall, is a restricted form of marginal modeling, where each group of flows has its own marginal distribution.
+In these cases, the flows are grouped by source node or by destination node.
+The mixed case uses the average behavior of the other two cases.
+It is interesting to note that the mixed case is universally worse than either the source or destination case.
+Clearly, hybridizing the conditionalization here is not an effective technique.
+We believe that the cause of this effect is like mixing too many colors yielding an unpleasant shade of brown: any of the colors individually would be better than the nondescript average.
+As a more general conclusion, arbitrarily mixing model may not always be a good idea, despite the ease with which it is done in the matrix representation of workload behavior.
+
+%We also believe that the fact that both the \model{time marginal} and \model{time uniform} models perform about as well as the \caps{GMM} indicates not that they are performing exceptionally well, but that the \caps{GMM} is performing worse than it should in this case.
+This may be an artifact of the interaction between packet behavior, flow size, flow duration and start time, as described in \Section{traffic-generation}.
+This remark may appear to be a non-sequitur, but these properties are all closely related.
+Consider that packet behavior (i.e.
+distribution of packet sizes and inter-packet intervals) determines the expected mean packet size and expected mean interval duration.
+The expected duration is then related to the flow size via the following equation:
+%\begin{align*}
+%\E{\text{duration}} = [\text{flow size}]\frac{\E{\text{mean interval}}}{\E{\text{mean size}}}
+%\end{align*}
+%The duration must also satisfy $t_{\text{start}}+[\text{duration}] \le t_{\max}$.
+
+%It accurately reproduces important performance metrics at all levels of the protocol stack in simulation.
+This only moves us a step closer to being able to generate realistic, completely synthetic workloads, since the \caps{GMM} in these comparisons  derives almost all of its behavior directly from the trace.
+However, it means that we have effectively reduced the problem of generating realistic workload to the problem of classifying and understanding what sorts of traffic matrices occur in real workloads.
+Moreover, we can express a vast array of different approaches to traffic modeling simply and uniformly using matrix operations.
+
+\section{Discussion}
+\label{sec:discussion}
+
+%The story of network analysis and modeling is one of variety. Assumptions of ``sameness'' have fallen repeatedly. ...
+%The matrix model gives new insight into
+
 \begin{sidewaysfigure*}
@@
 \end{sidewaysfigure*}
-
-Our simulation results are summarized in \Figure{box-plots}.
-Each subfigure shows a single performance metric.
-The distribution of log-ratio error values for each traffic model is visualized with a box-and-whisker plot.
-The box indicates the range from the 25th to 75th percentiles of values, while the ``whiskers'' indicate the full range, excluding outliners (which are shown as isolated points).
-These plots allow immediate assessment of realism: a good traffic model should have error values that are tightly clustered around the center, with a small, evenly balanced box, and relatively small whiskers.
-Additionally, the mean and median markers should be close to the center.
-It should be noted that the error scale is logarithmic, so even a slight increase in spread or deviation from the center indicates a disproportionately large decrease in model quality.
-Non-logarithmic scale markers are shown above each plot; the values indicate the factor of over- or under-representation for the performance metric.
-
-The primary result is that the \caps{GMM} accurately reflects the performance of the trace data it is based on.
-It has small, centered error bars for every performance metric presented here, as well as for those we have looked at otherwise.
-This is an essential result, since otherwise, \caps{GMM} is at best a shaky foundation for further modeling work.
-With these results, we can be assured that if we can approximate the matrix for a given example of trace traffic, then we have also approximated the original trace behavior.
-In a sense, this result is a reduction of the general traffic modeling problem to a more tractable matrix modeling problem.
-
-It is worth noting that \caps{GMM} has competition from a few of the time-simplified models: the \model{time marginal} and \model{time uniform} model both do approximately as well, and in some cases better, for the metrics examined.
-{\FHC} found strong evidence that session arrivals followed a time-varying Poisson arrival process, meaning that other than long-scale time-varying arrival rate and very small-scale clustering of flows within sessions, the overall flow arrival rate is fairly smooth.
-We suspect that on the scale of minutes to tens of minutes in which our simulation exist, it is sufficient to model both flow and session arrivals uniformly.
-This result indicates that the precise temporal placement of flows is not highly sensitive: they tend to be relatively evenly spaced out at short time-scales, and network performance is not sensitive to changes in start-time on that scale.
-
-Not all marginal models have such a benign effect on the accuracy of performance metrics.
-The worst offender by far is the \model{size marginal} model.
-This model distorts behavior at every level, by more than a factor 20 in 25\% of scenarios in the case of routing queue packet drops.
-This is a highly significant result because modeling flow size using a network-wide marginal distribution is precisely what {\FHC}, for example, attempt.
-This results informs us that this approach is doomed to failure: flow size cannot be assigned without considering its relation to other flow properties: source and destination nodes, and packet behavior at the very least.
-
-The \model{packet marginal} model is another offender, albeit not as egregiously.
-The direction of the misrepresentations in this case, however, is more dangerous: received throughput is overestimated by more than three times, 75\% of the time.
-On the other hand, \caps{AODV} control overhead is underestimated by half on average.
-Such serious misrepresentations are especially noteworthy given the apparently innocuous assumption applied by this model: the \model{packet marginal} model is essentially a standard variable bit-rate (\caps{VBR}) packet behavior model, using the empirical network-wide distributions of packet sizes and inter-packet intervals.
-Otherwise the model matches the original trace behavior exactly.
-And yet these drastic distortions occur.
-In our previous work, we have shown similarly drastic misrepresentations to occur when constant bit-rate packet behavior is assumed.
-This results indicates that to reproduce accurate network performance, we must provide packet behaviors on a more granular level: each node or even each flow must be assigned a ``custom'' packet behavior by some means.
-
-These strong negative results for two fundamental and very common marginal models provide severe limits on the realism that can be achieved through marginal modeling.
-Traffic modeling research \textit{must} move beyond simply finding parametric models for marginal distributions of network-wide properties.
-Realistic behavior cannot be generated using these distributions alone.
-Something must be captured about the interaction between the elements of flow behavior.
-
-Finally, we turn to the regularized models.
-These demonstrate neither exceptionally good nor exceptionally bad behavior.
-That the regularized models should be similar to the \model{full marginal} model but somewhat more realistic is entirely expected: regularization, the reader will recall, is a restricted form of marginal modeling, where each group of flows has its own marginal distribution.
-In these cases, the flows are grouped by source node or by destination node.
-The mixed case uses the average behavior of the other two cases.
-It is interesting to note that the mixed case is universally worse than either the source or destination case.
-Clearly, hybridizing the regularization here is not an effective technique.
-We believe that the cause of this effect is like mixing too many colors yielding an unpleasant shade of brown: any of the colors individually would be better than the nondescript average.
-As a more general conclusion, arbitrarily mixing model may not always be a good idea, despite the ease with which it is done in the matrix representation of workload behavior.
-
-%We also believe that the fact that both the \model{time marginal} and \model{time uniform} models perform about as well as the \caps{GMM} indicates not that they are performing exceptionally well, but that the \caps{GMM} is performing worse than it should in this case.
-This may be an artifact of the interaction between packet behavior, flow size, flow duration and start time, as described in \Section{traffic-generation}.
-This remark may appear to be a non-sequitur, but these properties are all closely related.
-Consider that packet behavior (i.e.
-distribution of packet sizes and inter-packet intervals) determines the expected mean packet size and expected mean interval duration.
-The expected duration is then related to the flow size via the following equation:
-%\begin{align*}
-%\E{\text{duration}} = [\text{flow size}]\frac{\E{\text{mean interval}}}{\E{\text{mean size}}}
-%\end{align*}
-%The duration must also satisfy $t_{\text{start}}+[\text{duration}] \le t_{\max}$.
-
-%It accurately reproduces important performance metrics at all levels of the protocol stack in simulation.
-This only moves us a step closer to being able to generate realistic, completely synthetic workloads, since the \caps{GMM} in these comparisons  derives almost all of its behavior directly from the trace.
-However, it means that we have effectively reduced the problem of generating realistic workload to the problem of classifying and understanding what sorts of traffic matrices occur in real workloads.
-Moreover, we can express a vast array of different approaches to traffic modeling simply and uniformly using matrix operations.
-
-\section{Discussion}\label{sec:discussion}
-
-%The story of network analysis and modeling is one of variety. Assumptions of ``sameness'' have fallen repeatedly. ...
-%The matrix model gives new insight into
 
@@
 %The reader will note that the transformations applied in this paper are all quite simple.
-Our regularization matrices primarily consisted of ones and zeros.
+Our conditionalization matrices primarily consisted of ones and zeros.
 The algebraic representation of complete workload

### 2008-06-09 paper.tex
Revised Discussion.

@@
 \subsection{Expression of Common Modeling Concepts}
-\label{sec:modeling-concepts}
+\sec{modeling-concepts}
 
@@
 %A generic traffic model is a way of generating the six matrices described in \Section{representations}.
-A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or conditionalized form.
+A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or conditional form.
 For experimental comparison, the dimensions of the matrices should all be preserved.
@@
 %% TODO: differentiate between restrictive versions of models and transformative versions.
-%% TODO: distinguish uniform, marginal and conditionalized models.
+%% TODO: distinguish uniform, marginal and conditional models.
 
@@
 
-After validating the \caps{GMM}, we explore the effects of various uniformity, marginality, and conditionalization assumptions on the realism of traffic models.
+After validating the \caps{GMM}, we explore the effects of various uniformity, marginality, and conditioning assumptions on the realism of traffic models.
 The uniform models are the most common, being the simplest both to conceptualize and to implement: simply make no distinction between start times, source or destination nodes.
@@
 
-Finally, we turn to the conditionalized models.
+Finally, we turn to the conditional models.
 These demonstrate neither exceptionally good nor exceptionally bad behavior.
-That the conditionalized models should be similar to the \model{full marginal} model but somewhat more realistic is entirely expected: conditionalization, the reader will recall, is a restricted form of marginal modeling, where each group of flows has its own marginal distribution.
+That the conditional models should be similar to the \model{full marginal} model but somewhat more realistic is entirely expected: conditioning, the reader will recall, is a restricted form of marginal modeling, where each group of flows has its own marginal distribution.
 In these cases, the flows are grouped by source node or by destination node.
-The mixed case uses the average behavior of the other two cases.
+
+% The mixed case uses the average behavior of the other two cases.
 It is interesting to note that the mixed case is universally worse than either the source or destination case.
-Clearly, hybridizing the conditionalization here is not an effective technique.
+Clearly, hybridizing the conditioning here is not an effective technique.
 We believe that the cause of this effect is like mixing too many colors yielding an unpleasant shade of brown: any of the colors individually would be better than the nondescript average.
@@
 %The matrix model gives new insight into
+
+Thus far, we have addressed two out of three of the originally posed questions regarding traffic modeling.
+First, we have proposed a general approach for representing whole-network traffic without making assumptions of uniformity or marginality.
+To this end, we introduced the concept of a linear representation of traffic.
+The general matrix model entails a specific linear representation, such that network traffic patterns can easily be expressed as matrices.
+Our experimental results show that the \caps{GMM} is sufficiently realistic for generating wireless workload derived from traces.
+When uniform or marginal assumptions are applied, however, sufficient realism eludes us, indicating that if we are to simplify and reduce the representation of network behaviors, we must do some through some other technique.
+In \Section{modeling-concepts}, we observed that application of uniformity and marginality assumptions to traffic instances is equivalent, in the \caps{GMM}, to right and left multiplication, respectively, by matrices of rank one.
+These transformations destroy almost all of the information in the original traffic matrix---only the zero matrix has lower rank than these.
+
+The question that remains to be answered is how we can simplify and reduce the representation of network behaviors without making unwarranted and intuitively clearly false assumptions of uniformity of marginality of network behaviors.
+We present here preliminary results from applying non-negative matrix factorization to the packet behaviors of collections of flows.
+Recall from \Section{general-matrix-model}, that the matrices $Z$ and $V$ represent the packet size and inter-packet interval histograms of a collection of flows, with each row corresponding to a flow, and each column corresponding to a packet size or a range of inter-packet intervals.
+We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}]$.
+\caps{NMF} attempts to approximate $\mat{P}$ as the product of three non-negative matrices:
+\begin{align}
+\mat{P} \approx \mat{W}\mat{\Sigma}\mat{H},
+\end{align}
+where $\mat{W} \in \R^{f \times K}$ is row-stochastic,
+$\mat{\Sigma} \in \R^{K \times K}$ is diagonal with descending diagonal entries, and
+$\mat{H} \in \R^{K \times (d_z+d_v)}$ is row-stochastic as well.
+$K\in\N$ is the rank of this factorization, which must be determined somehow as well.
+What is the interpretation of such factorization?
+The rows of $\mat{H}$ are \emph{basic behaviors}---pairs of packet size and inter-packet interval distributions, such that each flow's packet behavior can be approximated as a linear combination of these basic behaviors.
+Specifically, the matrix $\mat{M}\mat{\Sigma}$ gives the mixing coefficients for each flow.
+The diagonal entries of $\mat{\Sigma}$ have a special meaning as well:
+they are the number of packets associated with each basic behavior.
+Thus the basic behaviors are sorted in descending order of how much traffic they explain.
+Note that the factorization does not associate basic behaviors with individual packets;
+rather, total flow behavior is explained as a mixture of basic behaviors, so the fraction of packets in each flow explained by a given behavior is known, while the specific packets are unknown.
 
@@
 
-What are the ramifications of these results?
+% What are the ramifications of these results?
 In our previous work, we had already demonstrated the drastic distortions that are introduced by uniformity assumptions made by the most common traffic models.
@@
 
-\subsection{Towards Realistic Models of Wireless Workload}
-
-What would better traffic models look like?
+% \subsection{Towards Realistic Models of Wireless Workload}
+% 
+% What would better traffic models look like?
 How can we create them?
@@
 %The reader will note that the transformations applied in this paper are all quite simple.
-Our conditionalization matrices primarily consisted of ones and zeros.
+Our conditioning matrices primarily consisted of ones and zeros.
 The algebraic representation of complete workload

### 2008-06-09 paper.tex
Some work on Discussion.

@@
 
+The striking results of applying non-negative matrix factorization to $\mat{P}$ are shown in \Figure{basic-behaviors}.
+Twelve basic behaviors suffice to explain all the traffic in a ``toy''  sample of 4138 flows.
+The pairs of packet size and inter-packet interval distributions are almost immediately recognizable as resembling intuitive modes of network behavior.
+The first basic behavior, for example has all packets near the \caps{MTU}, while intervals are smoothly distributed around a fraction of a millisecond.
+This is what we would expect from downloading large files; and indeed, the breakdown of packets by port number bears out this interpretation---almost all of the traffic is \caps{HTTP} server traffic, with small slivers of \caps{IMAP} and \caps{SSH} server traffic.
+The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all \caps{SSH} traffic, interacting with a remote terminal via keystrokes and screen updates.
+
 \begin{sidewaysfigure*}

### 2008-06-12 paper.tex
Various edits.

@@
 Once the vector representation of each flow is defined, we may construct matrices of representation vectors, with a row for each flow in a traffic sample.
-We call the matrix representation of network traffic, the general matrix model~(\caps{GMM}).
-With network behavior is expresses as matrices, we can very simply and efficiently compute a broad variety of network behavior properties using standard matrix operations.
+We call the matrix representation of network traffic the \emph{general matrix model}~(\caps{GMM}).
+Having expressed network behavior in terms of matrices, we can very simply and efficiently compute a broad variety of network behavior properties using standard matrix operations.
 Furthermore, we can express uniformity and marginality assumptions about behavior as equality constraints between the rows and columns of the traffic matrix.
@@
 
-The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins, and represented each bin by a standard basis element of $\R^d$, where $d$ is the number of bins for that property.
-In this scheme, representation vectors are histograms counting items falling into each bin.
-The histograms for packet sizes and inter-packet intervals count packets, while other histograms count flows.
+The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins---%
+one per value for properties that are discrete already and require no compression---%
+and map values in each bin to a standard basis element of $\R^d$, where $d$ is the number of bins.
+In this scheme, representation vectors are effectively histograms, with coordinate values in each dimension counting items falling into the corresponding bin.
+% The histograms for packet sizes and inter-packet intervals count packets, while other histograms count flows.
 \Table{quantization} lists the properties that are used to define $\phi$, together with the quantization and dequantization functions.
-Each quantization function takes a possible property value as input and maps it to an index value in $\set{1,\dots,d}$, where $d$ is the dimension for the property in question.
-The index assigned determines the histogram bin of the input property.
-
-The histogram vector approach produces some representations of properties that are not immediately intuitive, and may seem to use an excessive number of representation dimensions.
+Each quantization function takes a possible property value as input and maps it to an index value in $\set{1,\dots,d}$, where $d$ is the number of bins, \latin{i.e.} dimension, for the property in question.
+% The index assigned determines the histogram bin of the input property.
+
+The histogram-vector approach produces some representations of properties that are not immediately intuitive, and may seem to use an excessive number of representation dimensions.
 However, this approach automatically satisfies the ``naturalness'' criterion for linear representations.
-That is, the representation, $\phi(T)$, of some collections of flows, $T$, naturally describes the aggregate behavior of those flows.
-Perhaps even more importantly, arbitrary linear combinations of vectors can be meaningfully interpreted as describing some realizable traffic instance.
-In \Section{traffic-generation} we will describe how to choose, stochastically, a collection of flows matching the ``description'' of an arbitrary traffic matrix.
+That is, the representation, $\phi(T)$, naturally describes the aggregate behavior of flows in $T$.
+Perhaps even more importantly, arbitrary linear combinations of vectors can be meaningfully interpreted as describing some realizable, hypothetical traffic instance.
+In \Section{traffic-generation} we will describe how to generate a flow for any sum of representation vectors.
 % This is how we generate workload from transformed traffic matrices.
 
-The quantization functions specified in \Table{quantization} allow us to define the representation $\phi$: \begin{align}
+The quantization functions specified in \Table{quantization} allow us to define the representation $\phi: \Traffic \to \R^N$: \begin{align}
 \phi(\flow) = \vec{f} = \tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}},
 \end{align}
-where the component vectors are defined as
+where $N=255+2n+d_t+d_b+d_z+d_v$ and vectors are
 \begin{align*}
@@
 \end{align*}
-Here, the functions $\Qf^\text{x}$ are the respective quantization functions for each property.
+The functions $\Qf^\text{x}$ are the respective quantization functions for each property.
 If there are $f$ flows in a traffic instance, then the collective behavior can be described with seven matrices---one for each property;
@@
 \begin{align}
-\mat{F} = \left[\,\mat{T}~\mat{S}~\mat{D}~\mat{A}~\mat{B}~\mat{Z}~\mat{V}\,\right] \in \R^{f \times N},
-\end{align}
-where $N=255+2n+d_t+d_b+d_z+d_v$.
-We call the representation of whole-network traffic instances as such a matrix, the general matrix model.
-While it is by no means the only possible representation of traffic in terms of matrices, many common traffic modeling assumptions can be expressed succinctly and precisely in terms of transformations or conditions of the \caps{GMM} and its components.
-Furthermore, realistic traffic can be generated from trace-derived matrix instances.
+\mat{F} = \left[\,\mat{T}~\mat{S}~\mat{D}~\mat{A}~\mat{B}~\mat{Z}~\mat{V}\,\right] \in \R^{f \times N}.
+\end{align}
+We call this total representation the general matrix model.
+While it is by no means the only possible representation of traffic in terms of matrices, as demonstrate in the following sections, many common traffic modeling assumptions can be expressed succinctly and precisely in terms of transformations or conditions of the \caps{GMM} and its components.
+Furthermore, realistic workload can be generated from matrix instances.
 

### 2008-06-12 paper.tex
Massive structural reorganization.

@@
 
-\graphicspath{{graphs/}}
+\graphicspath{{graphics/}}
 
@@
 
-\input{trace_vs_uniform}
+\input{figures/trace_vs_uniform}
 
@@
 
-\input{gmm_properties}
+\input{tables/gmm_properties}
 
@@
 \subfloat[\textbf{Application:} Average End-to-End Delay]%
-{\includegraphics[width=4.45in]{plots/Application/delay.pdf}}\hfill
+{\includegraphics[width=4.45in]{Application/delay.pdf}}\hfill
 \subfloat[\textbf{Application:} Packet Delivery Ratio]%
-{\includegraphics[width=4.45in]{plots/Application/delivery_ratio.pdf}}
+{\includegraphics[width=4.45in]{Application/delivery_ratio.pdf}}
 %
 \subfloat[\textbf{Application:} Average Received Throughput]%
-{\includegraphics[width=4.45in]{plots/Application/throughput.pdf}}\hfill
+{\includegraphics[width=4.45in]{Application/throughput.pdf}}\hfill
 \subfloat[\textbf{Network:} AODV Control Overhead]%
-{\includegraphics[width=4.45in]{plots/Network/control_packets.pdf}}
+{\includegraphics[width=4.45in]{Network/control_packets.pdf}}
 %
 \subfloat[\textbf{Network:} Packets Dropped in Routing Queue]%
-{\includegraphics[width=4.45in]{plots/Network/dropped.pdf}}\hfill
+{\includegraphics[width=4.45in]{Network/dropped.pdf}}\hfill
 \subfloat[\textbf{Link:} 802.11 Control Overhead]%
-{\includegraphics[width=4.45in]{plots/Link/overhead.pdf}}%
+{\includegraphics[width=4.45in]{Link/overhead.pdf}}%
 %
@@
 \begin{sidewaysfigure*}
-{\includegraphics[height=5.95in]{plots/basic_behaviors_1.pdf}}\hfill%
-{\includegraphics[height=5.95in]{plots/basic_behaviors_2.pdf}}
+{\includegraphics[height=5.95in]{basic_behaviors_1.pdf}}\hfill%
+{\includegraphics[height=5.95in]{basic_behaviors_2.pdf}}
 \caption{%

### 2008-06-12 models.rb
Deleting old, unused files.

@@ -1,47 +1,0 @@
-#!/usr/bin/ruby
-
-require 'yaml'
-
-models = YAML.load_file("models.yaml")
-
-bullet = '\\raisebox{1.5pt}{$\\centerdot$}'
-fixups = {
-  :Session => [ 8,'2.5pt'],
-  :Flow    => [14,'2.5pt'],
-  :Packet  => [ 5,'2.5pt']
-}
-
-puts <<'HEADER'
-\begin{tabular}{|c|c|p{4.3in}|}
-\multicolumn{1}{c}{\textbf{Class}} &
-\multicolumn{1}{c}{\textbf{Subclass}} &
-\multicolumn{1}{c}{\textbf{Overridden and inherited properties}} \\[2pt]
-HEADER
-models.each do |l|
-  level = l.keys.first
-  mods = l[level]
-  fixup = fixups[level.to_sym]
-  puts '\hline'
-  puts "\\multirow{#{fixup.first}}{*}[#{fixup.last}]{\\class{#{level}}}"
-  mods.each_with_index do |m,j|
-    model = m.keys.first
-    items = m[model]
-    items = [items] unless items.is_a?(Array)
-    puts "& \\multirow{1}{*}[-0.05em]{\\class{#{model}}} &"
-    puts '\begin{minipage}[l]{4.3in}'
-    puts '\vspace{2pt}'
-    items.each_with_index do |item,k|
-      item.gsub!(/&/,'\\\&')
-      item.gsub!(/\*\s*$/,'\raisebox{1.5pt}{\scriptsize*}')
-      line_end = (k < items.size-1) ? ' \\\\' : ''
-      puts "#{bullet} #{item}" + line_end
-    end
-    puts '\vspace{2pt}'
-    puts '\end{minipage} \\\\'
-    puts '\cline{2-3}' if j < models.size
-  end
-  puts '\hline'
-end
-puts <<'FOOTER'
-\end{tabular}
-FOOTER

### 2008-06-12 properties.rb

@@ -1,17 +1,0 @@
-#!/usr/bin/ruby
-
-require 'yaml'
-
-properties = YAML.load_file("properties.yaml")
-
-bullet = '\\raisebox{1.5pt}{$\\centerdot$}'
-
-properties.each_with_index do |l,i|
-  puts '[3pt]' if i > 0
-  level = l.keys.first
-  props = l[level]
-  puts "\\textbf{\\class{#{level}:}}\\\\"
-  props.each do |prop|
-    puts "\\hspace*{1ex}#{bullet}~#{prop}\\\\"
-  end
-end

### 2008-06-12 figures/basic_behaviors.tex
Moved all tables & figures into separate files.

--- 2008-06-12 paper.tex
@@
-\documentclass[twocolumn,final]{svjour3}
-
-\newcommand{\thetitle}{Linear Representation of Network Traffic}
-\newcommand{\theauthors}{
-  Stefan~Karpinski \and
-  Elizabeth~M.~Belding \and
-  Kevin~C.~Almeroth \and
-  John~R.~Gilbert
-}
-
-%\usepackage{type1cm}
-%\usepackage[pdftex]{graphicx}
-\usepackage[labelfont=bf,small]{caption}
-\usepackage[font=small,labelfont=bf,position=top,nearskip=0em]{subfig}
-\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize,paralist,array}
-\usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle}]{hyperref}
-\usepackage{mathtools}
-
-\title{\thetitle}
-\author{\theauthors}
-\institute{
-  S.~Karpinski \and
-  E.~Belding \and
-  K.~Almeroth \and
-  J.~Gilbert \\
-  Department of Computer Science \\
-  University of California, Santa Barbara (USA) \\
-  Email: \texttt{\{sgk,ebelding,almeroth,gilbert\}@cs.ucsb.edu}
-}
-
-\input{definitions}
-
-\graphicspath{{graphics/}}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-We propose a representation of wireless workload patterns as large, sparse matrices and provide a method for stochastically generating experimental workload from a given matrix.
-The essential property of the algebraic representation is that the summation of vectors naturally yields a faithful description of the aggregate behavior of the corresponding flows.
-This deceptively simple property allows us to express many common concepts from traffic modeling succinctly in terms of a few linear transformations.
-The algebraic representation has many benefits: 1)~it makes the meaning of generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2)~it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3)~the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed; 4)~the vast body of algebraic theory and highly optimized numerical linear software may immediately be applied to traffic modeling.
-We use the paired differential simulation methodology introduced by the authors in previous work to experimentally demonstrate that the general matrix model accurately reproduces realistic network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-We use the same experimental methodology to explore the implications of various assumptions and simplifications that are commonly made in traffic modeling.
-%We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models.
-A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
-%
-%We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in wireless networks.
-Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematical precision and succinctness, while simultaneously being made clearer and more general.
-We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to discredit the common constant bit-rate model.
-A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
-We conclude with future directions for modeling research based on the new algebraic framework.
-\end{abstract}
-
-% The model of flow behavior implicit in non-negative matrix factoriation technique is that flow behavior is a composite of some limited set of basic behaviors---and that fractions of each flow ``belong'' to each behavior.
-This is not an \latin{a priori} unreasonable hypothesis.
-Contrast that with the assumptions underlying uniform and marginal simplifications of traffic behavior, which can be seen to be intuitively unreasonable without ever conducting and actual experiment or calculation.
-
-\keywords{
-  traffic modeling \and
-  traffic analysis \and
-  workload generation \and
-  wireless networks \and
-  wireless simulation \and
-  linear algebra \and
-  non-negative matrix factorization
-}
-
-\section{Introduction}\label{sec:intro}
-
-\newfootnote{\flownote}{We use the common definition of a \textit{flow} as a sequence of packets sharing the same  ``5-tuple'': \caps{IP} protocol type, source and destination nodes, and \caps{TCP/UDP} port numbers.}
-
-\newfootnote{\marginalnote}{The statistical term ``marginal'' refers to the margins of actuarial tables formerly used for statistical computations.
-The rows and columns of the table represent possible values of two properties.
-Each entry in a table contains a count of the number of events falling into the joint category for that row and column.
-The margins contain sums of the rows and columns, thus giving the unconditional distributions of each property.}
-
-More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local-area networks (\caps{LAN}s).
-Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
-One of the major challenges of modeling whole-network traffic is the interdependence of packet, flow\flownote and node behaviors.
-Three types of simplifying assumptions are commonly applied in both analysis and generation:
-\begin{enumerate}
-\item \textbf{Uniform:} that traffic properties are uniformly distributed. For example, assuming that  all nodes in a network have equal probability of being the source or destination of each flow.
-Another common example is the assumption that all packet sizes and inter-packet intervals are the same; this is commonly known as the ``constant bit-rate'' (\caps{CBR}) model.
-\item \textbf{Marginal:} that traffic properties share a single, common \emph{marginal}\marginalnote distribution across the entire network, and individual values are drawn independently and unconditionally from this shared distribution.
-For example, assuming that a single distribution of packet sizes describes all flows in a network, as compared to equipping each flow with its own specific distribution of sizes.
-\item \textbf{Conditional:} that traffic behaves marginally within equivalence classes defined by certain conditions.
-The assumption, for example, that all flows emanating from a common source node share the same packet size distribution is a conditional model, conditioned on the source nodes.
-\end{enumerate}
-We would be arguing against a straw man if we implied that researchers who apply these assumptions actually believe them to be true.
-They clearly are not:
-different nodes initiate and receive drastically different numbers of packets and flows;
-flows have drastically different packet size distributions.
-Simply considering intuitive examples of common behavior illustrates this amply:
-BitTorrent users \latin{vs.} occasional email checkers;
-downloading a large file \latin{vs.} typing in an \caps{SSH} session.
-
-% In short, the reason is that these assumptions drastically simplify both traffic analysis and workload generation.
-% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
-% This immediately begs the question of whether these assumptions are benign.
-% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to analyze and replicate?
-% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
-
-% TODO: is this paragraph necessary?
-% The second reason why uniformity and marginality assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
-% The difficulty lies in the high-dimensional and stochastic nature of network behaviors.
-% Suppose, for example, that we assume that packet sizes and inter-packet intervals are drawn independently from some given distribution for each flow.
-% In the terminology introduced above, this is a flow-conditional packet behavior model.
-% Prior research has indicated that, given accurate per-flow distributions, and keeping all other aspects of network behavior fixed, this simplification does not distort experimental performance metrics~\cite{Karpinski07:realism}.
-% In this model, however, each flow must have its own specific, realistic pair of distributions for packet size and inter-packet intervals.
-% Moreover, the ``distribution of distributions'' across the entire network must simultaneously be realistic.
-% Modeling such distributions of distributions realistically for entire networks is currently an unsolved problem; we begin to address it here.
-
-If no one actually believes that assumptions of uniformity and marginality accurately describe network behavior, then why are these assumptions so common?
-There are two reasons.
-The first is that these assumptions drastically simplify both traffic analysis and workload generation, making them tractable problems.
-% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
-% This immediately begs the question of whether these assumptions are benign.
-% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to understand?
-% This paper answers that question.
-The second is that there are no generally accepted alternatives for representing and simplifying traffic patterns that allow effective analysis and generation without assuming uniformity or marginality.
-% Representing complex multi-level network behaviors in a way that does not make these simplifying assumptions but allows effective traffic analysis and workload generation is currently an unsolved problem;
-% we begin to address it here.
-
-% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
-
-Fundamentally, uniform, marginal and conditional models are applied so that network traffic behavior can be represented using a manageable collection of empirical or analytical statistical distributions.
-Reduction in representation size and complexity makes analysis and generation feasible.
-From this perspective, three questions immediately present themselves:
-\begin{enumerate}
-\item Do these assumptions distort the metrics we are trying to analyze, replicate, and predict?
-\item Can we represent ``unreduced'' network behavior such that analysis and generation remain tractable?
-\item Are there better ways to simplify and reduce the representation of network behavior?
-\end{enumerate}
-This paper answers all three questions. We begin, however by answering the second problem and  
-using our solution to address the other two questions.
-% transforming the unreduced representation of traffic into it's uniform, marginal or conditional counterparts.
-
-To represent unreduced network behavior practically, we propose the concept of a \emph{linear representation} of network traffic.
-A linear representation of traffic maps each flow to a vector, such that the following linearity condition is satisfied:
-\begin{quote}
-The sum of the representations of two flows is a vector that represents the aggregate behavior of the two flows.
-\end{quote}
-The complete behavior of a network is expressed as a matrix of such vectors, having one row per flow.
-Linear representations express full network behavior, without imposing assumptions of uniformity or marginality, yet still allow effective traffic analysis and workload generation.
-We present a specific linear representation, called the general matrix model (\caps{GMM}), and demonstrate experimentally that this model generates wireless workload that accurately reproduces the performance characteristics of real wireless network traffic.
-
-\newfootnote{\ranknote}{For uniform and marginal assumptions, the multiplication matrix has rank 1  since all rows and columns are identical.
-In the conditional case, the rank of the matrix is the number of conditioning classes.}
-
-To address the the other two questions, we begin with a simple observation.
-In linear representations of traffic, uniform, marginal and conditional assumptions are assertions of equality between the rows and columns of the traffic matrix.
-Moreover, real traffic matrices can be transformed into uniform, marginal and conditional forms through very simple matrix multiplications.
-These transformations, however, are extremely destructive: the matrices are degenerate with very low rank, so multiplication destroys almost all information in the original traffic matrix.
-We demonstrate experimentally, that this loss of information is not benign:
-uniform and marginal traffic models severely distorting crucial network performance metrics.
-
-\input{figures/trace_vs_uniform}
-
-Once traffic is represented linearly, alternatives to assuming uniformity or marginality become apparent.
-We argue that matrix factorization, which is nondestructive and does not assume uniformity or marginality, is far superior to multiplication for this problem.
-To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF})~\cite{Lee01} to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
-This factorization extracts a small set of ``basic behaviors'' from the traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
-Although the factorization is computed without examining port numbers, basic behaviors nevertheless correspond to intuitive expectations for protocols.
-We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.
-
-% TODO: update roadmap paragraph.
-
-The rest of the paper is organized as follows.
-In \Section{motivation} we discuss motivation and related work.
-The notion of linear representation and the general matrix model are presented in \Section{general-matrix-model}.
-In Sections~\ref{sec:common-properties}~and~\ref{sec:modeling-concepts} we explore how common network properties and modeling concepts can be expressed in the the \caps{GMM} framework.
-Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
-The ramifications of these results are discussed in \Section{discussion}.
-Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies.
-
-\section{Motivation \& Related Work}
-\label{sec:related-work}
-\label{sec:motivation}
-
-\newfootnote{\QoSnote}{This is violated by some quality of service (QoS) schemes.
-However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid.
-The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.}
-
-The interaction of network users and applications with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted.
-The joint pattern of traffic generation and mobility through time and space completely determines the effect of network usage on the lower levels of the stack.
-This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner~\cite{Clark88}.\QoSnote The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic.
-\Figure{trace-vs-uniform} illustrates through examples how drastically different network traffic appears under assumptions of uniformity at different levels of behavior.
-Marginality assumptions are not as visually striking, but as we will demonstrate, they distort the performance characteristics of networks almost as much, and sometimes more, than uniform models.
-
-While this work is not inherently specific to wireless networks, since the modeling techniques developed are equally applicable to traffic in wired networks, the results are of special importance to wireless research.
-Realistic traffic models have a drastic impact on experimental results in wireless networks~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-This is primarily because wireless networks are locally resource constrained.
-Compare this with the typical situation in wired \caps{LAN}s, where gigabit Ethernet and modern routers  can easily handle all but the most intense levels of traffic.
-If a wired network operator expects larger volumes of traffic on a conditional basis, the solution is to upgrade the hardware, not to change the network protocols.
-In wireless networks, on the other hand, the physical medium cannot be upgraded; protocols must be improved to make better usage of the medium.
-Being able to perform experiments with realistic workload is a vital part of the ongoing process of improving wireless protocols.
-Without realistic workload models for experiments, performance comparisons may be misleading or even completely wrong~\cite{Karpinski07:cbr-failure}.
-
-Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the emergent behavior is governed by a closed feedback loop.
-In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
-Hern\'andez-Campos' dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and typical \caps{TCP} applications.
-He shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
-This research has subsequently been turned into a \caps{TCP} flow generation tool called \tmix~\cite{Weigle06}.
-The acronym is somewhat misleading: {\tmix} does not provide models for application behaviors or for mixtures of applications in networks.
-Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
-Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary inputs to {\tmix}.
-
-\newfootnote{\harpoonnote}{Even this is unclear: generated traffic is only compared with trace traffic using the very same metrics that are explicitly specified as part of the model.
-Unsurprisingly, the distributions of sampled metrics closely match the distributions those metrics were sampled from.
-Realism is not double-checked using other statistical metrics or actual performance metrics.}
-
-The two most prominent general traffic generation frameworks are Harpoon and \caps{D-ITG}.
-Harpoon~\cite{Sommers04} uses a traffic trace for self-training, and can subsequently generate synthetic traffic with certain statistical properties based on the original trace.
-The properties reproduced are the empirical distributions of the following: ``file size, inter-connection time, source and destination IP ranges, [and] number of active sessions.''
-Harpoon exemplifies the uniform and marginal modeling approaches.
-Source and destination nodes are treated marginally: empirically derived marginal frequencies of sources and destinations are used to choose endpoints independently for each flow.
-Flow sizes (\latin{i.e.} file sizes) are likewise treated marginally: each flow's total byte  count is sampled from a marginal distribution of flow sizes.
-\caps{UDP} flows are modeled as being constant bit-rate;
-\caps{TCP} flows are all treated as file transfers, making them effectively uniform with bit-rate determined by \caps{TCP} dynamics.
-Inter-connection times are sampled from a single empirically derived marginal distribution.
-While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic,\harpoonnote for generating local-area traffic in wireless experiments, it is not.
-Previous research has shown that uniform packet behavior models used in Harpoon drastically distort vital performance metrics at all levels of the protocol stack~\cite{Karpinski07:cbr-failure}.
-This paper shows, moreover, that marginal models, like those used in Harpoon, also severely misrepresent performance metrics in wireless networks.
-
-Like Harpoon, \caps{D-ITG}~\cite{Avallone04} exemplifies uniform and marginal modeling assumptions. However, unlike Harpoon, \caps{D-ITG} does not derive models from real traffic.
-Packet behavior can be modeled using standard statistical distributions (``constant, uniform, normal, cauchy, pareto, exponential, etc.'').
-Guidance is not provided, however, for choosing among these distributions, choosing parameter values, or selecting a mixture of distributions across a network.
-Flow and node behaviors are not specified at all, but rather left to the user's discretion.
-The focus of this project is primarily on the engineering challenge of generating very large volumes of synthetic traffic and injecting it into networks in a distributed fashion.
-While \caps{D-ITG} provides the mechanism for generating large volumes of traffic, it does not provide any real guidance as to what traffic to generate.
-
-\FHC~\cite{Hernandez06:spatio-temporal-modeling} have defined the state of the art in parametric modeling of wireless traffic patterns.
-They use rigorous statistical methods to fit parametric models to a number of properties of wireless traffic.
-Flow arrivals fit a non-stationary Poisson model;
-flow inter-arrival times fit a Lognormal distribution;
-the number of flows per session fits a BiPareto model;
-flow sizes also fit a BiPareto model.
-The statistical analysis fitting these parametric models is thorough and convincing.
-The models, however, are marginal:
-a single statistical model applies to all flows and nodes; packet behavior is not considered.
-Each node has a different number of flows assigned to it, but otherwise nodes are identical. 
-Moreover, all flows, regardless of source node, are statistically identical:
-flow sizes are drawn from a single marginal distribution, and nothing else distinguishes them.
-Yet, intuitively it is clear that different nodes in real networks have drastically different tendencies for flow size, flow rates, and packet behavior.
-Such interdependent behaviors cannot be captured using marginal modeling.
-In this paper, we demonstrate that this inability is not harmless:
-treating all flows and nodes statistically identically severely distorts crucial wireless performance metrics in experiments.
-We also provide sorely needed methods of representing and analyzing traffic patterns without making uniform or marginal assumptions about network behavior.
-
-This work uses the technique known as \emph{paired differential simulation} to evaluate the realism of modified or synthetically generated wireless workloads.
-This technique was proposed and subsequently refined by Karpinski~\latin{et~al.}~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-They introduce the notion of \emph{sufficient realism}, and propose paired differential simulation as an experimental methodology based on the standard technique of pairing control and test subjects.
-We describe the technique in \Section{methodology}.
-In the first paper~\cite{Karpinski07:realism}, they explore various ways of modifying trace traffic patterns, as a form of sensitivity analysis to determine which aspects of network behavior are essential and must be preserved in realistic models, and which aspects can be discarded as inessential.
-Of particular interest is their conclusion that detailed time-series packet behavior for flows is inessential:
-it suffices, for each flow, to have accurate unconditional distributions of packet sizes and inter-packet intervals;
-sufficiently realistic flow behavior can be reconstructed from these distributions by repeated independent sampling.
-In the second paper~\cite{Karpinski07:cbr-failure}, Karpinski~\latin{et~al.} explore the impact of a broad variety of uniformity assumptions at the packet, flow and node levels of behavior.
-They find that all uniformity assumptions are detrimental, inducing significant and often extreme misrepresentation of a broad variety of wireless performance metrics at every level of the protocol stack.
-Moreover, they show, using \caps{OLSR} and \caps{AODV}, that uniform traffic models can completely invert the relative performance of network protocols.
-
-\section{Linear Representation of Traffic}
-\label{sec:representations}
-
-The traffic workload pattern in a network is a collection of \caps{IP} packets between hosts in the network sent at certain times.
-It is standard in network analysis, however, to aggregate sequences of packets sharing the same ``5-tuple'':
-\caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP/UDP} port numbers.
-Such a sequence of packets is called a \textit{flow}.
-The traffic pattern of an entire network is simply the collected behaviors of all flows occurring in the network.
-
-The behavior of a flow as it affects the network is characterized by the following properties:
-its \caps{IP} protocol type (\caps{TCP}, \caps{UDP}, \caps{ICMP}, etc.);
-its source and destination nodes;
-its start time;
-and the specific sizes packets of and intervals between their transmission.
-We begin our exposition of traffic representation by formalizing these properties mathematically as sets:
-\begin{tabbing}
-\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{5em}\=\hspace{4em}\=$\in\Nodes\,$\=\kill
-\>\bullet\>types:\>$\type$\>$\in\Types$\>$= \set{1,\dots,255} \subset \N$ \\
-\>\bullet\>nodes:\>$\src,\dst$\>$\in\Nodes$\>$= \set{1,\dots,n} \subset \N$ \\
-\>\bullet\>times:\>$\start$\>$\in\Times$\>$= \R$ \\
-\>\bullet\>sizes:\>$\size$\>$\in\Sizes$\>$= \set{0,\dots,\text{\caps{MTU}}} \subset \N$ \\
-\>\bullet\>intervals:\>$\ival$\>$\in\Ivals$\>$ = \R^+ \union \set{\infty}$.
-\end{tabbing}
-There are 255 possible \caps{IP} types (0 is reserved for \caps{IPv6}).
-The number $n$ is simply the total number of nodes in the network being modeled.
-\caps{MTU} is the maximum transfer unit of the network (typically around 1500 bytes).
-We formalize the packet behavior of a flow as a pair of infinite sequences of packet sizes and inter-packet intervals:
-\begin{align}
-\sizes &= \seq{\size_1,\size_2,\dots} \in \SizeSeq \\
-\ivals &= \seq{\ival_1,\ival_2,\dots} \in \IvalSeq.
-\end{align}
-Since flows contain only a finite number of packets, there exists a number of packets, $\pkts \in \N$, for each flow.
-Since flows are finite, we require that the sequences $\sizes$ and $\ivals$ satisfy the following requirements:
-\begin{align}
-&\forall\: k \le \pkts: \quad \size_k > 0       \eqn{positive-packets}  \\
-&\forall\: k >   \pkts: \quad \size_k = 0       \eqn{zero-packets}      \\
-&\forall\: k <   \pkts: \quad \ival_k < \infty  \eqn{finite-intervals}  \\
-&\forall\: k \ge \pkts: \quad \ival_k = \infty. \eqn{infinite-intervals}
-\end{align}
-Packet sizes are required to be positive because we are modeling application layer data transfer:
-an empty data transmission request is a non-event.
-The send time of the $k$th packet can be expressed as
-\begin{align}
-\time_k = \start + \sum_{i=1}^{k-1} \ival_i.
-\end{align}
-This is guaranteed to be finite if and only if $k \le \pkts$.
-Now we are ready to formalize the space of flows:
-\begin{align}
-  \Flows =
-    \Types    \cross
-    \Nodes^2  \cross
-    \Times    \cross
-    \N        \cross
-    \SizeSeq  \cross
-    \IvalSeq.
-\end{align}
-We write a generic element of $\Flows$ as
-\begin{align}
-  \flow = \tuple{
-    \type,
-    \src,
-    \dst,
-    \start,
-    \pkts,
-    \sizes,
-    \ivals
-  } \in \Flows.
-\end{align}
-Let $2^\Flows$ denote the power set of $\Flows$, and $\Traffic \subseteq 2^\Flows$ the set of all finite sets of flows:
-\begin{align}
-  \Traffic = \set{X \in 2^\Flows : \card{X} < \infty}.
-\end{align}
-A traffic sample is simply an element of $\Traffic$. We will refer to elements of $\Traffic$ interchangeably as ``traffic samples'' or ``traffic instances.''
-
-Having chosen a mathematical formalization of network traffic, we may now define the notion of a linear representation of network traffic.
-
-\begin{definition}\defn{linear-representation}
-A \defterm{linear representation} of network traffic is a pair, $\tuple{V,\psi}$, where $V$ is a vector space and $\psi$ is a function, $\psi: \Traffic \to V$, such that
-\begin{align}\eqn{linearity}
-\forall~T \in \Traffic\!:\quad
-  \psi(T) &= \sum_{\mathclap{\flow \in T}}{\psi(\flow)}.
-\end{align}
-\end{definition}
-
-\noindent
-In plain words, every finite collection of flows is represented by a vector that is the sum of the elements representing the individual flows in the collection.
-% Representing traffic using vectors allows us to analyze, model, and understand traffic patterns in terms of the thoroughly studied and well understood subject of linear algebra.
-The careful reader will notice that \emph{any} function $\Flows \to V$ uniquely determines a linear representation:
-\Equation{linearity} uniquely extends the function to all of $\Traffic$.
-Representing flows by arbitrary vectors, however, is not particularly useful or interesting.
-We focus instead on meaningful representations, where collections of flows are represented by vectors that naturally describe the aggregate behavior in some manner.
-In the next section, we clarify these concepts through examples.
-
-\subsection{Examples of Linear Representations}
-\sec{linear-representation-examples}
-
-\newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $\pkts$ does not divide $\mathsym{bytes}$ evenly.
-The simplest is to ignore it; we leave leave more complex schemes to the reader.}
-
-Our first example is the packets-bytes-duration (\caps{PBD}) representation.
-The representation is $\pbd: \Traffic \to \R^3$, mapping
-$\flow \mapsto \tuple{p,b,d} \in \R^3$, where
-\begin{align}
-p &= \pkts, &
-b &= \sum_{i=1}^{\mathclap{p}}{\size_i}, &
-d &= \sum_{i=1}^{\mathclap{p-1}}{\ival_i}.
-\end{align}
-This representation encapsulates the information necessary to reproduce a \caps{CBR} version of each flow:
-packets have payloads of $\floor{\fracx{b}{p}}$ bytes, and are sent every $\fracx{d}{p}$ seconds.\pbdnote
-Given two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, their sum as vectors, 
-\begin{align}
-\vec{x}_1+\vec{x}_2=\tuple{p_1+p_2,b_1+b_2,d_1+d_2},
-\end{align}
-provides an appropriate description of the aggregate behavior of the two flows together.
-Similarly, their average, \latin{i.e.} $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
-In contrast, a ``packets-payload-interval'' representation, expressing the \caps{CBR} parameters directly, does not have this property:
-the sum of average payloads is not the overall average of payloads;
-likewise for intervals.
-In the \caps{PBD} representation, the linearity property leads to natural descriptions of aggregate behavior through vector addition, whereas in the packets-payload-interval representation, vector addition is ``unnatural.''
-
-The \caps{PBD} representation is highly lossy: it discards most of the information detailing the behavior of each flow.
-At the other extreme, we can give a ``perfect'' representation of each flow, that allows us to completely reconstruct each flow from its representation.
-With care, the sum of representations of flows can still naturally describe their aggregate behavior.
-Let the representation be $\Phi: \Traffic \to \FlowVec$, where
-\begin{align}
-\FlowVec = \R^{255} \cross \R^n \cross \R^n \cross L^2[\R] \cross L^2[\R].
-\end{align}
-% Here $\R^X$ is the vector space of functions $f: X \to \R$, which is isomorphic to the standard Euclidean vector space, $\R^\card{X}$.
-Here $L^2[\R]$ is the vector space of real functions, $\R^\R=\set{f: \R \to \R}$, with the usual norm~\cite{Rudin87}.
-% \begin{align}
-% \norm{f}_2 = \Big[ \int_{-\infty}^{\infty} \abs{f}^2 dx \Big]^{\frac{1}{2}}.
-% \end{align}
-We define
-\begin{align}
-\Phi(\flow) = \tuple{\e_\type,\e_\src,\e_\dst,f,g},
-\end{align}
-where $\e_k$ is the $k$th standard basis vector, and $f,g \in L^2[\R]$ are functions defined by
-\begin{align}
-f(t) &= \sum_{k=1}^{\pkts}
-  \begin{cases}
-    1 & \text{if $t \ge \time_k$} \\
-    0 & \text{if $t <   \time_k$},
-  \end{cases} \\
-g(t) &= \sum_{k=1}^{\pkts}
-  \begin{cases}
-    \size_k & \text{if $t \ge \time_k$} \\
-    0       & \text{if $t <   \time_k$}.
-  \end{cases}
-\end{align}
-Thus, $f$ and $g$ are both integer-valued, monotonically increasing step functions, giving the cumulative number of packets and total bytes sent, respectively, at time $t$.
-From the representation, $\Phi(\flow)$, of a single flow, we can readily recover its type, source, destination, and the transmission times and sizes of packets.
-
-Why is $\Phi$ natural in the same sense that the \caps{PBD} representation is natural?
-Suppose we have a traffic sample, $T\in\Traffic$.
-The first component of $\Phi(T)$ is a 255-dimensional vector, whose coordinates are a histogram of the number of times each flow type occurs in $T$.
-Similarly, the second and third components of $\Phi(T)$ are $n$-dimensional histograms of the number of times each node occurs as source and destination.
-The last components are step functions, like $f$ and $g$.
-Vector addition in $L^2[\R]$ dictates that these functions give the total number of cumulative packets and bytes sent, respectively, over all flows in $T$.
-Thus, summation of vectors under the representation $\Phi$ yields a natural and useful description of the aggregate behavior of flows.
-
-Consider, for contrast, if we had represented flow type simply by $\type \in \N$ instead of $\e_\type \in \R^{255}$.
-The natural numbers containing $\type$ can readily be embedded into $\R^1$, which is a vector space.
-However, the addition of vectors is highly unnatural.
-The sum of protocol numbers is meaningless:
-two \caps{TCP} flows plus five \caps{ICMP} flows does not equal a single \caps{UDP} flow, despite the fact that $2\cdot6+5\cdot1=17$.
-The summation of vectors in the representation $\Phi$, on the other hand is $5\,\e_1+2\,\e_6$, indicating the frequency with which each type occurred.
-
-Both of the example representations discussed here can be applied in practice.
-The \caps{PBD} representation naturally allows us to compute aggregate and average \caps{CBR} behaviors of collections of flows.
-With a suitable machine encoding of step-functions, $\Phi$ can be implemented and used in practice.
-It provides complete information about individual flows and highly detailed (though not complete) information about collections of flows.
-Representations that preserve even more information about traffic instances can readily be constructed.
-We turn now, however, in a different direction, and introduce a specific linear representation that strikes a practical balance in fidelity between $\pbd$ and $\Phi$.
-
-\section{The General Matrix Model}
-\sec{general-matrix-model}
-
-\input{tables/gmm_properties}
-
-The \caps{PBD} representation of flows uses 3 dimensions to encode the behavior of each flow, while $\Phi$ uses $255+2n$ dimensions, plus the continuum of dimensions of two Banach spaces (\latin{i.e.} the two copies of $L^2[\R]$).
-In this section, we define a representation, $\phi$, that splits the difference and uses a large but finite number of dimensions to represent each flow.
-Once the vector representation of each flow is defined, we may construct matrices of representation vectors, with a row for each flow in a traffic sample.
-We call the matrix representation of network traffic the \emph{general matrix model}~(\caps{GMM}).
-Having expressed network behavior in terms of matrices, we can very simply and efficiently compute a broad variety of network behavior properties using standard matrix operations.
-Furthermore, we can express uniformity and marginality assumptions about behavior as equality constraints between the rows and columns of the traffic matrix.
-Moreover, we can transform unconstrained traffic instances into uniform, marginal, or conditional approximations via right or left multiplications by easily computed matrices.
-
-The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins---%
-one per value for properties that are discrete already and require no compression---%
-and map values in each bin to a standard basis element of $\R^d$, where $d$ is the number of bins.
-In this scheme, representation vectors are effectively histograms, with coordinate values in each dimension counting items falling into the corresponding bin.
-% The histograms for packet sizes and inter-packet intervals count packets, while other histograms count flows.
-\Table{quantization} lists the properties that are used to define $\phi$, together with the quantization and dequantization functions.
-Each quantization function takes a possible property value as input and maps it to an index value in $\set{1,\dots,d}$, where $d$ is the number of bins, \latin{i.e.} dimension, for the property in question.
-% The index assigned determines the histogram bin of the input property.
-
-The histogram-vector approach produces some representations of properties that are not immediately intuitive, and may seem to use an excessive number of representation dimensions.
-However, this approach automatically satisfies the ``naturalness'' criterion for linear representations.
-That is, the representation, $\phi(T)$, naturally describes the aggregate behavior of flows in $T$.
-Perhaps even more importantly, arbitrary linear combinations of vectors can be meaningfully interpreted as describing some realizable, hypothetical traffic instance.
-In \Section{traffic-generation} we will describe how to generate a flow for any sum of representation vectors.
-% This is how we generate workload from transformed traffic matrices.
-
-The quantization functions specified in \Table{quantization} allow us to define the representation $\phi: \Traffic \to \R^N$: \begin{align}
-\phi(\flow) = \vec{f} = \tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}},
-\end{align}
-where $N=255+2n+d_t+d_b+d_z+d_v$ and vectors are
-\begin{align*}
-\vec{t} &= \e_k & k &= \Qf^{\text{t}}(\type) \\
-\vec{s} &= \e_k & k &= \Qf^{\text{s}}(\src) \\
-\vec{d} &= \e_k & k &= \Qf^{\text{d}}(\dst) \\
-\vec{a} &= \e_k & k &= \Qf^{\text{a}}(\start) \\
-\vec{b} &= \e_k & k &= \Qf^{\text{b}}(\bytes) \\
-\vec{z} &= {\textstyle \sum \e_k} & k &\in \set{\Qf^{\text{z}}(\size_i) : i \le \pkts} \\
-\vec{v} &= {\textstyle \sum \e_k} & k &\in \set{\Qf^{\text{v}}(\ival_i) : i <   \pkts}.
-\end{align*}
-The functions $\Qf^\text{x}$ are the respective quantization functions for each property.
-If there are $f$ flows in a traffic instance, then the collective behavior can be described with seven matrices---one for each property;
-the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
-We use the following notations for the behavior matrices:
-\vspace{-0.25em}
-\begin{tabbing}
-\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
-$\mat{D}\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
-\>\bullet\> type: \>$\mat{T}$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$\\
-\>\bullet\> sources: \>$\mat{S}$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$\\
-\>\bullet\> destinations: \>$\mat{D}$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$\\
-\>\bullet\> start times: \>$\mat{A}$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$\\
-\>\bullet\> sizes in bytes: \>$\mat{B}$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$\\
-\>\bullet\> packet sizes: \>$\mat{Z}$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$\\
-\>\bullet\> inter-packet intervals: \>$\mat{V}$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
-\end{tabbing}
-The total behavior matrix is the horizontal concatenation of these component matrices:
-\begin{align}
-\mat{F} = \left[\,\mat{T}~\mat{S}~\mat{D}~\mat{A}~\mat{B}~\mat{Z}~\mat{V}\,\right] \in \R^{f \times N}.
-\end{align}
-We call this total representation the general matrix model.
-While it is by no means the only possible representation of traffic in terms of matrices, as demonstrate in the following sections, many common traffic modeling assumptions can be expressed succinctly and precisely in terms of transformations or conditions of the \caps{GMM} and its components.
-Furthermore, realistic workload can be generated from matrix instances.
-
-\subsection{Choosing Constants}
-
-Before continuing with less mundane topics, we must briefly address our choices of constants appearing in \Table{quantization}.
-The externally predetermined constants are:
-$d_z=\text{\caps{MTU}}=1500$, determined by the wireless network medium;
-$b_{\max} \approx 109 \text{\caps{MB}}$, determined by the size of the largest flow in our trace data;
-$t_{\max}=v_{\max}=600$, determined by our experimental methodology which slices traces into 10-minute scenarios;
-and $v_{\min}=10^{-6}$, determined by the time resolution of our trace.
-For lack of a better choice, we have simply chosen to use the same number of quantization bins as packet size for the unfixed dimension constants:
-$d_t = d_b = d_v = 1500$.
-Arguments can certainly be made for other choices, but this choice appears to work adequately in practice.
-
-Two other unspecified constants appear in \Table{quantization}: $\beta$ and $\gamma$.
-These parameters are non-linear scaling exponents that allows the quantization to smoothly shift from fine-grained resolution at the low end of the spectrum to coarse-grained resolution at the high end.
-We have chosen $\beta=\fracx{1}{2.7}$ because it yields a desirable dynamic range response from bytes at the low end to megabytes at the high end.
-Similarly, we have chosen $\gamma=\fracx{1}{3}$ because this choice provides resolution that scales smoothly from microseconds for small intervals to seconds for large intervals.
-
-\subsection{Traffic Generation}
-\sec{traffic-generation}
-
-Given an instance of the general matrix model, how do we produce an actual workload from it?
-Each row of the model is used to randomly generate the behavior of a single flow.
-To do this, we use the ``dequantization'' functions listed in \Table{quantization}.
-The dequantization procedure works as follows.
-Let $\vec{f}$
-% \begin{align}
-% \vec{f}=\tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}}
-% \end{align}
-be a row of the traffic matrix, and let $\prop$ be a property.
-Write $\vec{f}_\prop$ for the corresponding component vector of $\vec{f}$.
-Chose an index value $k\in\set{1,\dots,d_\prop}$, randomly, with weight given by the coefficients of $\vec{f}_\prop$.
-Choose $u \in [0,1]$ uniformly randomly.
-Then $x = k-u \in [0,d]$, with probability of lying in the interval $(i-1,i]$ proportional to the coefficient of $\e_i$ in $\vec{f}_\prop$.
-The dequantization function for $\prop$, call it $\Df_\prop$, will map the value $x$ back into the appropriate range of property values for $\prop$.
-
-The above procedure allows us to sample random values for each property in accordance with the behavioral description implied by $\vec{f}$.
-This allows us to stochastically generate a sequence of packets in accordance with a row from an arbitrary \caps{GMM} matrix instance using the following procedure:
-\begin{enumerate}
-\item Sample random source and destination node pairs.
-Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
-The simplest technique is the rejection method:
-if the same node is chosen for both source and destination, reject that pair and begin the selection process again.
-\item Next, randomly sample the number of bytes.
-\item The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\mean{z}$ and $\mean{v}$.
-From these, together with the flow size in bytes, we estimate the expected duration of the flow:
-$\mean{d} = \fracx{b\mean{v}}{\mean{z}}$.
-\item Randomly sample a start time for the flow.
-Ensure that the start time plus the expected duration, $\mean{d}$, does not exceed the maximum simulation time for the workflow by using the rejection method.
-\item The first packet is sent at the chosen start time.
-The size of each packet is determined by sampling a random size value.
-The interval until the next packet is determined by sampling a random interval value.
-The packet stream ends when the number of bytes allocated to the flow have been emitted or when the packet emission time exceeds $t_{\max}$.
-The last packet is only the size of the remaining number of bytes.
-\end{enumerate}
-Some remarks about the procedure:
-\begin{itemize}
-\item For some vectors in $\R^{f \cross N}$, it may be impossible to choose distinct source and destination nodes or flow sizes and start times satisfying the procedure.
-However, for any vector that is a sum of vectors representing valid flows---\latin{i.e.} ones having distinct source and destination nodes and feasible combinations of start time, duration, and packet behaviors---the procedure will terminate.
-\item Because the representation $\phi$ models the seven properties independent of each other, it is impossible to do better than to sample the properties independently, modulo the requirements of the procedure.
-In the case of a row which is simple a representation of an actual flow, this makes no difference, except for the generation of packets (which are modeled independently).
-In the case of composite rows, however, this implies that flow properties are modeled independent of each other.
-\end{itemize}
-Despite these remarks, we will show in our experimental evaluation, that this procedure generates sufficiently realistic wireless workloads.
-
-\subsection{Expression of Common Properties}
-\label{sec:common-properties}
-
-Having defined our representation of network behavior and presented the procedure for stochastically generating workload, we turn to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.
-
-\subsubsection{Communication Topology}
-\label{sec:communication-topology}
-
-The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{SD}}\in\R^{n \times n}$ ($\trans{\,\diamond\,}$~denotes transpose).
-That is $(\trans{\mat{SD}})_{ij}$ is the number of flows from node $i$ to node $j$.
-The total number of source flows for each node is given by the vector $\trans{\mat{S}}\ones{f}$, and the total number of destination flows is given by $\trans{\mat{D}}\ones{f}$.
-($\ones{f}$ is the $f$-dimensional column vector consisting of all ones.)
-
-\subsubsection{Expected Average Packet Size \& Intervals}
-\label{sec:flow-totals}
-
-From the matrices $\mat{Z}$ and $\mat{V}$, we can compute the expected average packet size, $\mean{z}$, and average inter-packet interval, $\mean{v}$, for each flow.
-The computations require knowing the values that each matrix column represents.
-The computed duration is approximate because information is irrecoverably lost in the quantization process.
-The vectors of representative values for $\mat{Z}$ and $\mat{V}$ are:
-\begin{align}
-\vec{\sigma}&=\tuple{1,2,3,\dots,d_z},\\
-\vec{\tau}&=\tuple{\Di\left(k,\fracx{1}{2}\right)}_{k=1}^{d_v}.
-\end{align}
-Using these representative values, the expected average packet size and expected average inter-packet interval size for a flow may be computed as
-\begin{align}
-  \mean{z}&=\frac{\inner{\vec{z}}{\vec{\sigma}}}{\inner{\vec{z}}{\ones{}}}, &
-  \mean{v}&=\frac{\inner{\vec{v}}{\vec{\tau}}}{\inner{\vec{v}}{\ones{}}},
-\end{align}
-respectively, where $\inner{\vec{x}}{\vec{y}}$ denotes the standard inner product of vectors.
-
-\subsection{Expression of Common Modeling Concepts}
-\sec{modeling-concepts}
-
-The simplifying assumptions made by most traffic models can be classified into three categories.
-We use the terms \textit{uniformity}, \textit{marginality}, and \textit{conditionality} to describe these classes of assumptions.
-They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether the make stipulations about uniformity within each flow, across all flows, or within groups of flows.
-In the following sections, we explore how various modeling concepts can be expressed in the framework of the general matrix model.
-Models are expressed as transformations that will convert arbitrary traffic matrix instances into similar ones that satisfy the model's assumptions.
-These transformations will later allow us to evaluate the effect of each model's simplifications on the realism of generated workloads.
-
-\subsubsection{Uniformity}
-
-The concept of uniformity, as used here, entails that certain properties of each flow are assumed to be statistically or deterministically uniform.
-Examples include: assuming that each flow's packets have the same size and inter-packet interval (i.e.
-the \caps{CBR} flow model); assuming that each node is equally likely to be the source of a given flow; assuming that all start times for a flow are equally likely.
-
-Since uniform behaviors entail homogeneity of properties with respect to each flow individually, a trace matrix can be transformed to a matrix satisfying the model's assumptions via right matrix multiplication.
-Right multiplication mixes the elements within each row, and if the mixing weights are uniform, the result is uniform behavior.
-For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $\mat{S}$ and $\mat{D}$:
-\begin{align*}
-\mat{S}' &= \mat{S}\ones{n \times n}/n, & \mat{D}' &= \mat{D}\ones{n \times n}/n.
-\end{align*}
-This makes all the values in each flow's row identically the average of the original $n$ values.
-As another example, the constant bit-rate (\caps{CBR}) flow model, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
-The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $[\mat{Z}~\mat{V}]$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $[\mat{Z}~\mat{V}]\,\Gamma\in\R^{f \times 3}$, where
-\begin{align}
-\Gamma=\begin{pmatrix}
-\ones{} & \vec{\sigma} & \zeros{} \\
-\zeros{} & \zeros{} & \vec{\tau}
-\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
-\end{align}
-\Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to realistic behaviors from traces.
-%By replacing $\mat{S}'$ with $\mat{S}$, this can also be viewed as a set of linear constraints on the matrix $\mat{S}$, providing a proscriptive definition
-
-%We use the term \textit{uniformity} do describe assumptions made about the ``sameness'' of various properties of flows.
-The constant bit-rate (\caps{CBR}) model, for example, is an assumption about the uniformity of all of a flow's packets: i.e.
-that they have the same size and are evenly spaced in time.
-While this is the strictest sense of the \caps{CBR} model, it is rare for this assumption to be made in isolation.
-Another common simplification is to make the sizes of flows identical, or chose them randomly from a uniform distribution; similarly for the starting time of each flow.
-Finally, when choosing endpoints for a given flow, lacking a better model, it is not unusual to treat all nodes identically, making each an equally likely source or destination for a given flow.
-
-%It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees.
-This is largely because of the high complexity of doing otherwise and the lack of compelling alternatives.
-Here we examine how various notions of uniformity are expressed in the matrix model.
-In general, \textit{intra-flow uniformity} is effected by right multiplication, which mixes the elements in each row, whereas \textit{inter-flow uniformity} is effected by left multiplication, which mixes elements across rows.
-The constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
-The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $\tuple{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\tuple{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
-%\begin{align}
-%\Gamma=\begin{pmatrix}
-%\ones{} & \vec{\sigma} & \zeros{} \\
-%\zeros{} & \zeros{} & \vec{\tau}
-%\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
-%\end{align}
-
-\subsubsection{Marginality}
-
-Marginality is an orthogonal concept to uniformity: the distribution of some property is identical \textit{across} all flows.
-For example, if we determine a distribution of flow sizes across the entire network and assume that this distribution applies \textit{independent of other flow properties}, this is a marginal model for flow size.
-Marginal behavior is effected by left multiplication by a matrix with uniform columns.
-This specific example is generated by left-uniformizing the flow size matrix: $\mat{B}' = (\ones{f \times f}/f)\mat{B}$.
-The traffic models proposed by {\FHC} are all marginal: they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node.
-These models do not account for non-independence between these and other flow properties.
-We will discuss this limitation and its implications when we analyze our experimental results.
-
-\subsection{Conditionality}
-
-The final form of homogenization we consider, is conditionality.
-This is a restricted form of marginality: we may assume that marginal distributions apply to all the flows in certain groupings.
-For example, we can provide a distribution of flow sizes on a per-source basis.
-That is, each source node can have its \textit{own} distribution, and the sizes of its flows are drawn from that distribution, independent of their other properties.
-This is simply accomplished by left-multiplication by a non-uniform matrix.
-For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-conditioning'' matrix: $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$.
-Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-conditioning matrix: $F'=(\trans{\mat{DD}}/f)\mat{F}$.
-%Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified.
-The confusion arises due to the lack of a unique way of associating flows with nodes.
-Each node can equally viably be associated with its source or its destination.
-Which association is more valid?
-Consider a web request: should the behavior of the request be determined by what the client typically does?
-Or should the requests that the server typically receives determine behavior?
-There is no armchair answer: only analysis of real data can give us the answer.
-The matrix model provides an analytical framework for doing so.
-The matrix model has the benefit of making the conditioning of behaviors completely explicit.
-It also allows us to easily define intermediate choices.
-If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
-\begin{align}
-\mat{F}' = \frac{1}{f}((1-\alpha)(\trans{\mat{SS}})+\alpha(\trans{\mat{DD}}))\mat{F}.
-\end{align}
-The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors.
-Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.
-
-%We use the terms here to distinguish three fundamentally different types of assumptions about behavioral homogeneity.
-
-%Existing traffic models are typically not explicit methods of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
-In our algebraic modeling framework, such models can be interpreted as stipulations about the form of the behavior matrices, or as transforms that implement those stipulations given a real set of traffic matrices.
-
-%A generic traffic model is a way of generating the six matrices described in \Section{representations}.
-A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or conditional form.
-For experimental comparison, the dimensions of the matrices should all be preserved.
-If a trace-derived model produces experimental performance metrics that are similar to those produced by the original trace data, then simplifications made by the model are viable; if the performance is distorted, then the simplifications are discarding essential information about the traffic pattern.
-
-%\subsection{Uniformity}
-
-%% TODO: differentiate between restrictive versions of models and transformative versions.
-%% TODO: distinguish uniform, marginal and conditional models.
-
-%\subsection{Marginality}
-
-%Most approaches to traffic modeling that go beyond uniformity assumptions, instead describe marginal properties of various aspects of traffic behavior.
-For example, {\FHC}~\cite{Hernandez06:spatio-temporal-modeling} provide highly convincing parametric models for six marginal properties of wireless network traffic: session (node) arrivals follow a time-varying Poisson process; per-session flow inter-arrival times are Lognormally distributed; flows per session (node) follow a BiPareto distribution; flow sizes follow a BiPareto distribution.
-This is certainly much greater detail than simply assuming that all nodes and flows are identical.
-
-%How do marginal models translate into the algebraic language of the general matrix model? Whereas uniformity assumptions for nodes,
-
-%Another benefit of the matrix model is that all aspects of network behavior are determined concurrently.
-We are not left wondering what to do about source and destination nodes or flow start and end times after we have computed our simplified packet behaviors: the same transformation that average the packet behaviors also naturally averages the start and
-
-\begin{table*}
-\begin{center}
-\small
-\begin{tabular}{|c|c|l|}
-\multicolumn{1}{c}{\textbf{model}} &
-\multicolumn{1}{c}{\textbf{transformation}} &
-\multicolumn{1}{c}{\textbf{description}} \\
-\hline
-\caps{GMM} & --- & the general matrix model \\\hline\hline
-time uniform & $\mat{A}'=\mat{A}(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
-node uniform & $\mat{S}'=\mat{S}(\ones{n \times n}/n)$~~~~$\mat{D}'=\mat{D}(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
-full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above models \\\hline\hline
-time marginal & $\mat{A}'=(\ones{f \times f}/f)\mat{A}$ & start times chosen from marginal distribution of start times \\\hline
-size marginal & $\mat{B}'=(\ones{f \times f}/f)\mat{B}$ & flow size (bytes) is chosen independently of other properties \\\hline
-node marginal & $\mat{S}'=(\ones{f \times f}/f)\mat{S}$~~~~$\mat{D}'=(\ones{f \times f}/f)\mat{D}$ & nodes behave distinctly, but behavior is independent of flows \\\hline
-packet marginal & $\mat{Z}'=(\ones{f \times f}/f)\mat{Z}$~~~~$\mat{V}'=(\ones{f \times f}/f)\mat{V}$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
-full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline\hline
-source conditional & $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$ & flow behaviors are aggregated on a per source-node basis \\\hline
-dest conditional & $\mat{F}'=(\trans{\mat{DD}}/f)\mat{F}$ & flow behaviors are aggregated on a per destination-node basis \\\hline
-mixed conditional & $\mat{F}'=\frac{1}{2f}(\trans{\mat{SS}}+\trans{\mat{DD}})\mat{F}$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
-\end{tabular}
-\caption{Matrix-based traffic models evaluated by paired differential simulation.}
-\label{tab:traffic-models}
-\end{center}
-\vspace{-2em}
-\end{table*}
-
-\section{Experimental Methodology}
-\label{sec:methodology}
-
-To evaluate whether traffic models are realistic or not, we use the method of \textit{paired differential simulation} introduced in our previous work on traffic modeling~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-%This approach is based on the following definition:
-%
-%\begin{samepage}
-%\noindent\textbf{Definition.}~A model of user behavior is \textit{sufficiently realistic} if, compared with actual user behavior, the model, with parameter values extracted from the real data, yields statistically equivalent performance results.
-%\end{samepage}
-%\vspace{0.5em}
-%
-Paired differential simulation is based on the standard scientific technique of controlled experimentation. 
-The hypothesis we are testing is that a given synthetic model accurately reproduces the performance metrics exhibited by real network traffic.
-To test this hypothesis, we conduct a series of paired experiments and compare the results for each test subject with a corresponding control subject.
-In this case, the experiments are wireless network simulations, and the subjects are real or synthetic workload instances.
-The control group is the set of simulations where the workload is an actual traffic trace, recorded as described in \Section{trace-data}.
-The test group is a set of simulations where workload is synthetically generated using a simplified traffic model.
-The experiments are paired: for each control simulation, there is a synthetic workload, with behavior as similar to the control as model will allow, against which the results are compared.
-The output of the series of experiments are paired values of performance metrics for each simulated scenario: one from the control, one from the test model.
-
-The remainder of this section proceeds as follows.
-In \Section{traffic-models} we present the collection of traffic models we will evaluate.
-In \Section{trace-data} we describe the trace data that serves as the control for our experiments.
-Our simulation methodology is detailed in \Section{simulations}, and the performance metrics used for statistical comparison of control and test experiments are described in \Section{performance-metrics}.
-% Finally, \Section{measures-of-error} is an in-depth mathematical discussion of the appropriate measure of error for the pairs of metric values that are produced by the paired differential simulation methodology.
-
-% to compare traffic produced by synthetic models with known examples of recorded trace traffic in a variety of experimental scenarios.
-To accomplish this, we begin with a large 24-hour example of wireless trace traffic, break it into 144 ten-minute scenarios, and use each scenario's simulated trace behavior as a control for a set synthetic traces, generated by various traffic models.
-For each scenario, the synthetic models are configured to approximate the original trace behavior as closely as the models allow.
-Important network-wide performance metrics are then collected from simulations, and synthetic performance results compared with control results to determine the realism of the synthetic models.
-
-%The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results.
-Clearly, we need not simulate subatomic particle interactions in a wireless network simulator.
-Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same.
-Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior.
-This requirement leads us to the following definition:
-
-%This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required.
-We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.
-
-\subsection{Traffic Models}
-\label{sec:traffic-models}
-
-The first model we evaluate is the general matrix model itself.
-We derive the matrix representation directly from the trace data and use it to generate workload as described in \Section{traffic-generation}.
-This serves to test whether the \caps{GMM} is sufficiently general: if it accurately reproduces all performance metrics across the collection of scenarios simulated, this provides strong evidence that \caps{GMM} captures enough of the detail of trace behavior to be considered a sufficiently general representation of network traffic patterns.
-%An affirmation of the generality of the \caps{GMM} acts as reduction of the
-
-After validating the \caps{GMM}, we explore the effects of various uniformity, marginality, and conditioning assumptions on the realism of traffic models.
-The uniform models are the most common, being the simplest both to conceptualize and to implement: simply make no distinction between start times, source or destination nodes.
-Choose them at random, using no information from reality other than the number of nodes and the possible range of start times.
-We do not evaluate uniform packet behavior because it has been sufficiently discredited in our previous work.
-
-In more sophisticated modeling work, where uniformity is discarded as too simple a model for behavior, marginality is often applied instead.
-It is assumed---often implicitly---that marginal models are adequate to capture the essential aspects of network behavior.
-We put that implicit assumption to the test here.
-How well do perfect marginal models represent real behavior?
-When we say that our marginal models are ``perfect,'' we mean that they are nonparametric and calculated directly from trace behavior.
-These are the marginal models that most accurately describe the trace traffic, because they are directly derived from it and make no parametric simplifications.
-This provides an upper bound on the quality of marginal models in general, since parametric marginal models are derived, in turn, as simplifications of these nonparametric marginal behaviors.
-\Table{traffic-models} provides a complete list of all the traffic models we compare in our evaluation.
-
-%The original trace traffic serves as a control group against which to compare all other models.
-Before we can evaluate the general matrix model (\caps{GMM}), we must establish that one potentially important aspect of traffic behavior that is \textit{not} represented in the \caps{GMM} can legitimately be discarded.
-That aspect of behavior is \textit{\caps{TCP} feedback}.
-The \caps{GMM} does not specify the \caps{IP} protocol type of each flow.
-Therefore we must establish the extent to which using \caps{UDP} flows, without reliable delivery or congestion control, affects performance metrics.
-%%\caps{GMM} also does not indicate which flows are paired with each other in a client-server relationship. While this pairing does not affect non-\caps{TCP} flows, we implement.
-%Other synthetic models derived from algebraic transformations of the \caps{GMM} are listed in \Table{traffic-models}.
-
-\subsection{Trace Data}\label{sec:trace-data}
-%\subsection{Trace Data \& Simulations}\label{sec:trace-data}\label{sec:simulations}
-
-Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of trace-based models and synthetic traffic models defined using the general matrix model.
-We use a 24-hour trace recorded in an infrastructured 802.11g wireless \caps{LAN} with 18 access points, deployed at the 60th Internet Engineering Task Force meeting (\caps{IETF}60), held in San Diego during August of 2004.
-The traffic trace was captured using \texttt{\small{tcpdump}} at a single router, through which all wireless traffic for the meeting was routed, including traffic between wireless nodes.
-The snap length of the capture was 100 bytes, allowing \caps{IP}, \caps{ICMP}, \caps{UDP} and \caps{TCP} headers to be analyzed.
-We limit our work to the 24-hour sub-trace recorded on Wednesday, August 4th.
-This trace contains a broad variety of behaviors and entails a very large volume of traffic: 2.1 million flows, 58 million packets, and 52 billion bytes.
-% 2,085,563 flows, 58,377,994 packets of application data, and 51,988,245,537 bytes of application data.
-
-\begin{figure}[!b]
-\begin{center}
-\includegraphics[width=3.3in]{nodes-flows}%
-\vspace{-1.25em}%
-\caption{The number of active nodes and flows over time.} 
-\label{fig:nodes-flows}
-\end{center}
-\vspace{-1.3em}
-\end{figure}
-
-We do not assume or claim that the traffic found at \caps{IETF}60 is representative of conference settings in general.
-The observed behaviors are also unlikely to resemble those found in a typical commercial or residential setting.
-We have chosen this trace, however, because within it can be found behaviors resembling many different types of wireless usage cases.
-\Figure{nodes-flows} shows the wide variations in the number of active flows and nodes over the course of the trace.
-In the night and morning hours, the traffic patterns are similar to those one might find in a moderately trafficked business or residential area.
-During working group sessions, we see highly concentrated, heavy usage patterns.
-At the zenith of activity, over 800 users, 33 thousand flows, and 1 million packets are seen in a single 10-minute trace segment.
-At the nadir, a lone node sent only a single 61-byte packet over the course of 10 minutes.
-All levels of activity between these extremes are represented.
-Moreover, the mix of traffic types observed changes dramatically over the course of the day, providing a wide representation of possible blends of behavior.
-This heterogeneity and extreme range of behaviors makes the \caps{IETF} data set ideal for this evaluation.
-The variety of activity gives us greater confidence that success or failure of traffic models is not tied to any specific network condition, but is broadly and generally applicable.
-%For a traffic model to be truly realistic, it must be realistic across many different usage cases.
-
-Before using the traces, it is necessary to extract application-level behavior from the trace header data.
-First, we split the trace into individual packet flows.
-A flow is a series of packets sharing the following five attributes: \caps{IP} and transport protocols (raw \caps{IP}, \caps{ICMP}, \caps{TCP}, \caps{UDP}); source and destination \caps{IP} addresses and \caps{TCP}/\caps{UDP} port numbers.
-Next, the quantity of application-initiated data contained in each packet is calculated.
-For non-\caps{TCP} packets, this quantity is simply the size of the transport-layer payload, but for \caps{TCP} the calculation is more complicated: only new data transfers, explicitly initiated by the application are counted.
-Data retransmitted by \caps{TCP} is disregarded, and empty \caps{ACK}s are ignored.
-\caps{SYN} and \caps{FIN} flags in packets (even empty ones) are counted as a single byte each, since they are explicitly signaled by the application.
-%After processing, we have a collection of flows, each with a sequence of timestamps and the amount of application-initiated data sent at that time.
-
-\subsection{Simulations}
-\label{sec:simulations}
-
-We use the Qualnet wireless network simulator (version 4.0.1) to perform our experiments.
-We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
-In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays.
-Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings.
-We do not attempt to reproduce the physical environment of the original wireless network, nor do we simulate mobility.
-The only aspect of the original network's behavior that is reproduced is the total pattern of network-wide traffic.
-
-There are a number of potential objections to this approach.
-We use single-hop trace data to drive multi-hop simulations; the physical environment, node mobility, handover behavior, and closed-loop dynamics of the original wireless setting are not faithfully reproduced.
-One must keep in mind, however, that the goal of this research is \textit{not} to understand the conditions of the original network.
-Rather, we are using the traffic behaviors observed as examples to help us better understand how different types of workload can affect performance metrics.
-In particular, we aim to understand how real workload compares with common synthetic traffic models.
-Of course, the reason for such objections is that networking researchers understand that the many aspects of behavior interact with each other in a complex and nearly inextricable manner.
-However, before we can hope to understand the interaction between workload and other features affecting network behavior, we must study traffic patterns alone, and learn to model them with reasonable accuracy in the absence of additional complicating factors.
-Accordingly, in this study, we detach application level traffic patterns from the other factors influencing network conditions, and study them in isolation.
-
-The 24-hour trace is split into 144 10-minute segments, each of which serves as the basis for a set of simulations using different traffic models.
-The traffic models range from a completely realistic trace-driven model, to a standard \caps{CBR} traffic model.
-Various partially synthetic intermediate models, described in \Section{traffic-models}, are simulated to study the impact of different aspects of traffic behavior on network performance.
-To preserve the fairness of the performance comparison, we keep as many features as possible constant across different traffic models.
-The traffic generated by each synthetic model preserves as many characteristics from the original trace as possible, within the constraints of the model.
-Moreover, the following features are preserved across all models: the number of wireless nodes, the number of flows, the number of application-initiated data units sent, the total bytes of application data sent, and the average flow duration (and therefore the average data rate).
-
-In our previous work, we approximated \caps{TCP} with a pair of half-duplex \caps{UDP} flows~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-For this work, we have implemented a new Qualnet application driver that allows trace-driven full-duplex \caps{TCP} flows.
-This allows us to accurately reproduce the full dynamics of \caps{TCP} feedback.
-Moreover, raw \caps{IP}, \caps{UDP} and \caps{TCP} flows are implemented using the same code, guaranteeing that all traffic is simulated uniformly and that all performance metrics are aggregated in the same manner.
-We have also instrumented Qualnet to collect \caps{IP} performance metrics in strict adherence to the recommendations of the \caps{IETF} working group on \caps{IP} performance metrics~\cite{rfc:ip-metrics}.
-
-\subsection{Performance Metrics}
-\label{sec:performance-metrics}
-
-We have selected six performance metrics to present here. They are commonly used as indicators of network performance at the application, network, and link layers of the protocol stack:
-\begin{enumerate}
-\setlength{\itemsep}{0em}
-\item \textbf{Application:} average end-to-end delay, packet delivery ratio, received throughput.
-\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/\\RERR}), packets dropped in routing queues.
-\item \textbf{Link:} 802.11 control overhead (\caps{RTS/CTS/ACK}).
-\end{enumerate}
-These metrics are commonly used to evaluate wireless protocols.
-We have examined a broad variety of other wireless network performance metrics, and although we do not have room to present or discuss them here, the results shown are representative of the overall realism of the traffic models.
-
-\section{Results}
-\label{sec:results}
-
-\begin{sidewaysfigure*}
-\subfloat[\textbf{Application:} Average End-to-End Delay]%
-{\includegraphics[width=4.45in]{Application/delay.pdf}}\hfill
-\subfloat[\textbf{Application:} Packet Delivery Ratio]%
-{\includegraphics[width=4.45in]{Application/delivery_ratio.pdf}}
-%
-\subfloat[\textbf{Application:} Average Received Throughput]%
-{\includegraphics[width=4.45in]{Application/throughput.pdf}}\hfill
-\subfloat[\textbf{Network:} AODV Control Overhead]%
-{\includegraphics[width=4.45in]{Network/control_packets.pdf}}
-%
-\subfloat[\textbf{Network:} Packets Dropped in Routing Queue]%
-{\includegraphics[width=4.45in]{Network/dropped.pdf}}\hfill
-\subfloat[\textbf{Link:} 802.11 Control Overhead]%
-{\includegraphics[width=4.45in]{Link/overhead.pdf}}%
-%
-\caption{%
-Box-and-whisker plots of log-ratio error values for all metrics and traffic models.
-The lower axis indicates the log-ratio, while the upper axis shows raw ratio values.
-Each box contains the central majority of log-ratio values: the left and right bounds are at the $25^\text{th}$ and $75^\text{th}$ percentiles.
-The dark middle line indicates the median value, while the diamond marks the mean.
-The whiskers (dotted lines) extend to the furthest non-outlier values, while the points beyond that are outliers.
-The notch in the middle of each bar indicates a $95\%$ confidence interval for the true underlying median value; if two notches do not overlap, they are very unlikely to have the same median.
-}
-\label{fig:box-plots}
-\end{sidewaysfigure*}
-
-Our simulation results are summarized in \Figure{box-plots}.
-Each subfigure shows a single performance metric.
-The distribution of log-ratio error values for each traffic model is visualized with a box-and-whisker plot.
-The box indicates the range from the 25th to 75th percentiles of values, while the ``whiskers'' indicate the full range, excluding outliners (which are shown as isolated points).
-These plots allow immediate assessment of realism: a good traffic model should have error values that are tightly clustered around the center, with a small, evenly balanced box, and relatively small whiskers.
-Additionally, the mean and median markers should be close to the center.
-It should be noted that the error scale is logarithmic, so even a slight increase in spread or deviation from the center indicates a disproportionately large decrease in model quality.
-Non-logarithmic scale markers are shown above each plot; the values indicate the factor of over- or under-representation for the performance metric.
-
-The primary result is that the \caps{GMM} accurately reflects the performance of the trace data it is based on.
-It has small, centered error bars for every performance metric presented here, as well as for those we have looked at otherwise.
-This is an essential result, since otherwise, \caps{GMM} is at best a shaky foundation for further modeling work.
-With these results, we can be assured that if we can approximate the matrix for a given example of trace traffic, then we have also approximated the original trace behavior.
-In a sense, this result is a reduction of the general traffic modeling problem to a more tractable matrix modeling problem.
-
-It is worth noting that \caps{GMM} has competition from a few of the time-simplified models: the \model{time marginal} and \model{time uniform} model both do approximately as well, and in some cases better, for the metrics examined.
-{\FHC} found strong evidence that session arrivals followed a time-varying Poisson arrival process, meaning that other than long-scale time-varying arrival rate and very small-scale clustering of flows within sessions, the overall flow arrival rate is fairly smooth.
-We suspect that on the scale of minutes to tens of minutes in which our simulation exist, it is sufficient to model both flow and session arrivals uniformly.
-This result indicates that the precise temporal placement of flows is not highly sensitive: they tend to be relatively evenly spaced out at short time-scales, and network performance is not sensitive to changes in start-time on that scale.
-
-Not all marginal models have such a benign effect on the accuracy of performance metrics.
-The worst offender by far is the \model{size marginal} model.
-This model distorts behavior at every level, by more than a factor 20 in 25\% of scenarios in the case of routing queue packet drops.
-This is a highly significant result because modeling flow size using a network-wide marginal distribution is precisely what {\FHC}, for example, attempt.
-This results informs us that this approach is doomed to failure: flow size cannot be assigned without considering its relation to other flow properties: source and destination nodes, and packet behavior at the very least.
-
-The \model{packet marginal} model is another offender, albeit not as egregiously.
-The direction of the misrepresentations in this case, however, is more dangerous: received throughput is overestimated by more than three times, 75\% of the time.
-On the other hand, \caps{AODV} control overhead is underestimated by half on average.
-Such serious misrepresentations are especially noteworthy given the apparently innocuous assumption applied by this model: the \model{packet marginal} model is essentially a standard variable bit-rate (\caps{VBR}) packet behavior model, using the empirical network-wide distributions of packet sizes and inter-packet intervals.
-Otherwise the model matches the original trace behavior exactly.
-And yet these drastic distortions occur.
-In our previous work, we have shown similarly drastic misrepresentations to occur when constant bit-rate packet behavior is assumed.
-This results indicates that to reproduce accurate network performance, we must provide packet behaviors on a more granular level: each node or even each flow must be assigned a ``custom'' packet behavior by some means.
-
-These strong negative results for two fundamental and very common marginal models provide severe limits on the realism that can be achieved through marginal modeling.
-Traffic modeling research \textit{must} move beyond simply finding parametric models for marginal distributions of network-wide properties.
-Realistic behavior cannot be generated using these distributions alone.
-Something must be captured about the interaction between the elements of flow behavior.
-
-Finally, we turn to the conditional models.
-These demonstrate neither exceptionally good nor exceptionally bad behavior.
-That the conditional models should be similar to the \model{full marginal} model but somewhat more realistic is entirely expected: conditioning, the reader will recall, is a restricted form of marginal modeling, where each group of flows has its own marginal distribution.
-In these cases, the flows are grouped by source node or by destination node.
-
-% The mixed case uses the average behavior of the other two cases.
-It is interesting to note that the mixed case is universally worse than either the source or destination case.
-Clearly, hybridizing the conditioning here is not an effective technique.
-We believe that the cause of this effect is like mixing too many colors yielding an unpleasant shade of brown: any of the colors individually would be better than the nondescript average.
-As a more general conclusion, arbitrarily mixing model may not always be a good idea, despite the ease with which it is done in the matrix representation of workload behavior.
-
-%We also believe that the fact that both the \model{time marginal} and \model{time uniform} models perform about as well as the \caps{GMM} indicates not that they are performing exceptionally well, but that the \caps{GMM} is performing worse than it should in this case.
-This may be an artifact of the interaction between packet behavior, flow size, flow duration and start time, as described in \Section{traffic-generation}.
-This remark may appear to be a non-sequitur, but these properties are all closely related.
-Consider that packet behavior (i.e.
-distribution of packet sizes and inter-packet intervals) determines the expected mean packet size and expected mean interval duration.
-The expected duration is then related to the flow size via the following equation:
-%\begin{align*}
-%\E{\text{duration}} = [\text{flow size}]\frac{\E{\text{mean interval}}}{\E{\text{mean size}}}
-%\end{align*}
-%The duration must also satisfy $t_{\text{start}}+[\text{duration}] \le t_{\max}$.
-
-%It accurately reproduces important performance metrics at all levels of the protocol stack in simulation.
-This only moves us a step closer to being able to generate realistic, completely synthetic workloads, since the \caps{GMM} in these comparisons  derives almost all of its behavior directly from the trace.
-However, it means that we have effectively reduced the problem of generating realistic workload to the problem of classifying and understanding what sorts of traffic matrices occur in real workloads.
-Moreover, we can express a vast array of different approaches to traffic modeling simply and uniformly using matrix operations.
-
-\section{Discussion}
-\label{sec:discussion}
-
-%The story of network analysis and modeling is one of variety. Assumptions of ``sameness'' have fallen repeatedly. ...
-%The matrix model gives new insight into
-
-Thus far, we have addressed two out of three of the originally posed questions regarding traffic modeling.
-First, we have proposed a general approach for representing whole-network traffic without making assumptions of uniformity or marginality.
-To this end, we introduced the concept of a linear representation of traffic.
-The general matrix model entails a specific linear representation, such that network traffic patterns can easily be expressed as matrices.
-Our experimental results show that the \caps{GMM} is sufficiently realistic for generating wireless workload derived from traces.
-When uniform or marginal assumptions are applied, however, sufficient realism eludes us, indicating that if we are to simplify and reduce the representation of network behaviors, we must do some through some other technique.
-In \Section{modeling-concepts}, we observed that application of uniformity and marginality assumptions to traffic instances is equivalent, in the \caps{GMM}, to right and left multiplication, respectively, by matrices of rank one.
-These transformations destroy almost all of the information in the original traffic matrix---only the zero matrix has lower rank than these.
-
-The question that remains to be answered is how we can simplify and reduce the representation of network behaviors without making unwarranted and intuitively clearly false assumptions of uniformity of marginality of network behaviors.
-We present here preliminary results from applying non-negative matrix factorization to the packet behaviors of collections of flows.
-Recall from \Section{general-matrix-model}, that the matrices $Z$ and $V$ represent the packet size and inter-packet interval histograms of a collection of flows, with each row corresponding to a flow, and each column corresponding to a packet size or a range of inter-packet intervals.
-We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}]$.
-\caps{NMF} attempts to approximate $\mat{P}$ as the product of three non-negative matrices:
-\begin{align}
-\mat{P} \approx \mat{W}\mat{\Sigma}\mat{H},
-\end{align}
-where $\mat{W} \in \R^{f \times K}$ is row-stochastic,
-$\mat{\Sigma} \in \R^{K \times K}$ is diagonal with descending diagonal entries, and
-$\mat{H} \in \R^{K \times (d_z+d_v)}$ is row-stochastic as well.
-$K\in\N$ is the rank of this factorization, which must be determined somehow as well.
-What is the interpretation of such factorization?
-The rows of $\mat{H}$ are \emph{basic behaviors}---pairs of packet size and inter-packet interval distributions, such that each flow's packet behavior can be approximated as a linear combination of these basic behaviors.
-Specifically, the matrix $\mat{M}\mat{\Sigma}$ gives the mixing coefficients for each flow.
-The diagonal entries of $\mat{\Sigma}$ have a special meaning as well:
-they are the number of packets associated with each basic behavior.
-Thus the basic behaviors are sorted in descending order of how much traffic they explain.
-Note that the factorization does not associate basic behaviors with individual packets;
-rather, total flow behavior is explained as a mixture of basic behaviors, so the fraction of packets in each flow explained by a given behavior is known, while the specific packets are unknown.
-
-The striking results of applying non-negative matrix factorization to $\mat{P}$ are shown in \Figure{basic-behaviors}.
-Twelve basic behaviors suffice to explain all the traffic in a ``toy''  sample of 4138 flows.
-The pairs of packet size and inter-packet interval distributions are almost immediately recognizable as resembling intuitive modes of network behavior.
-The first basic behavior, for example has all packets near the \caps{MTU}, while intervals are smoothly distributed around a fraction of a millisecond.
-This is what we would expect from downloading large files; and indeed, the breakdown of packets by port number bears out this interpretation---almost all of the traffic is \caps{HTTP} server traffic, with small slivers of \caps{IMAP} and \caps{SSH} server traffic.
-The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all \caps{SSH} traffic, interacting with a remote terminal via keystrokes and screen updates.
+%!TEX root = ../paper.tex
 
@@
 \end{sidewaysfigure*}
-
-% What are the ramifications of these results?
-In our previous work, we had already demonstrated the drastic distortions that are introduced by uniformity assumptions made by the most common traffic models.
-This work serves to further critique established common assumptions in traffic modeling.
-The strongest message to be taken away is that the aspects of traffic behavior of both nodes and flows \textit{must} be considered jointly.
-It is not sufficient to continue modeling marginal distributions of network properties.
-These marginal models do not lend them selves easily to traffic generation in the first place, and when marginality assumptions are applied to trace traffic, the result is a severe distortion of important network performance characteristics.
-It is now well established that network traffic patterns have an impact on network performance that cannot be ignored.
-Even experimental deployments cannot avoid the need for more realistic traffic workload models.
-While using a real, physical network successfully sidesteps simulation problems below the application layer, without realistic traffic models, reliable, meaningful performance predictions remain beyond our reach.
-
-%\subsection{Generality of Results}
-
-%% TODO: talk about other synthetic models not considered here.
-%% The point of our analysis is not to show that no synthetic models can be accurate, but rather, that the commonly used models, and obvious variations thereof, are not accurate.
-
-%The most significant limitation on the generality of this analysis is that it is based entirely on a single data set from \caps{IETF}60---albeit a large and varied one.
-It is possible that traffic in this trace happens to produce network performance that is unusually dissimilar to standard traffic models.
-This data set, however, represents a highly heterogeneous collection of network usage behaviors, from slow and steady off-peak usage, to extremely heavy peak usage: over 800 users, 33 thousand flows, and 1 million packets in a single 10-minute trace segment.
-Despite the broad variety of behaviors, the results are consistent: in all types of usage scenarios, simplistic traffic models, like uniform \caps{CBR}, systematically skew important performance measurements at all levels of the network.
-While the precise results for other data sets might differ, it is very unlikely that \caps{CBR} traffic models will happen to accurately reproduce realistic performance in other experiments.
-This paper provides strong evidence that better traffic models are needed for performance evaluations.
-
-% \subsection{Towards Realistic Models of Wireless Workload}
-% 
-% What would better traffic models look like?
-How can we create them?
-One possible approach is to use actual traffic traces as we have done.
-This approach is unsatisfactory, however, because it provides the experimenter with almost no control over experiments.
-Synthetic models have parameters, which can be tweaked as necessary---adjusting, for example, the number of active nodes in a simulation, without affecting other parameters.
-Traffic traces, on the other hand, must be used without significant alteration if they are to actually provide the desired realism.
-This inability to control all but one or two parameters is nearly crippling for experimental purposes.
-Therefore, the research community needs models whose parameters can be adjusted independently, but whose performance characteristics can be made to resemble whatever example of network behavior one might need.
-The general matrix model does not yet fulfill this role since its parameters cannot be freely varied.
-We believe, however, that \caps{GMM} plays a vital role in the ongoing process of developing  of better models.
-By reducing the general problem of traffic modeling to the domain of linear algebra, we have provided traffic researchers with a vast array of powerful theoretical and computational tools that can be immediately applied to the problem at hand.\footnote{The derived models used in this paper were all implemented in Matlab, each in only a few simple lines.
-All the models used here are implemented in only 19 total lines of code.
-Of course, the framework to apply them to traffic traces and simulations is somewhat larger.} What are the next steps?
-The authors are currently investigating a variety of matrix factorization and clustering techniques to extract hidden structure from a large and diverse body of real traffic.
-
-%The reader will note that the transformations applied in this paper are all quite simple.
-Our conditioning matrices primarily consisted of ones and zeros.
-The algebraic representation of complete workload
-
-%The ``messiness'' of the performance comparison from trace data in Figure~\ref{fig:mac-control-overhead-cmp} illustrates why using traces directly is not ideal: each data point differs not only in the number of nodes shown on the $x$-axis, but also in other dimensions, such as the number of flows and packets, and the average flow duration.
-The result is a highly noisy comparison, affected by many unseen parameters.
-Only by applying a local smoothing algorithm are trends somewhat elucidated.
-
-%Instead of using trace data directly, it should be possible to configure a synthetic traffic model based on observations from a real data set, and then run side-by-side simulations using the synthetic model and the real data, producing statistically equivalent performance results.
-This is precisely what our definition of sufficient realism entails.
-The work in this paper provides the tools to measure how close to this ideal a model is and in what areas it needs improvement.
-Without this feedback, any improvements in realism are purely guesswork.
-Our breakdown of traffic behavior into three orthogonal levels also allows the problem to be approached in smaller pieces, rather than being solved all at once.
-
-%The next step towards better traffic models is to investigate which aspects of real traces may be altered without detrimentally affecting the resulting performance metrics.
-For example, to test whether a complex time-series model of packet behavior is necessary, we randomize the order of the packet sizes and/or inter-packet intervals and compare performance using these randomized traces against performance using the original traces.
-If the performance is unchanged, we can conclude that no complex time-series model of packet behavior is necessary: sampling the packet sizes and inter-packet intervals from empirical distributions is sufficiently realistic.
-If, on the other hand, the performance characteristics are altered by shuffling packets, then some time-series model of packet behavior is needed.
-By partially randomizing the packet order in specific ways, the exact limits of realism necessary can be found.
-A similar approach will allow the development of realistic models for the other levels of network usage behavior.
-
-\section{Conclusions}\label{sec:conclusions}
-
-%The real significance of this paper lies not in showing that the commonly used \caps{CBR} traffic model is unrealistic.
-Networking researchers understand that \caps{CBR} is at best a na\"ive approximation of real network traffic.
-Nor is the real significance that we have quantized the high degree of inaccuracy that can be induced by using this model.
-The real significance of this paper lies in providing an well-defined, objective measure of realism for traffic models.
-This has never previously existed: evaluations of realism have relied on essentially arbitrary statistical measures of similarity to real traffic.
-Existing traffic generators use empirical distributions of such quantities as file size, inter-connection time, packet size, or inter-packet intervals.
-They can only hope that reproducing these distributions as good enough.
-Without an external measure of accuracy, there's simply no way to know.
-
-%attempting to showing that existing models are unrealistic, or that new models are more realistic have always focused on arbitrary statistical measures. A
-
-We have presented a fundamentally new, algebraic way of representing traffic patterns in local area networks.
-We call this representation the general matrix model.
-Each flow of a traffic collection is represented as a single high-dimensional vector with components representing the \caps{IP} protocol type, source and destination nodes, start time, flow size, and packet behavior.
-The essential property of the algebraic representation is that standard vector operations compute natural and useful descriptions of aggregate behavior over the collection of flows represented.
-Our experimental validation demonstrates clearly that the general matrix model accurately reproduces the performance characteristics of of real traffic.
-The benefits of a performance-preserving algebraic representation are multifold.
-The algebraic forms of the simplifying assumptions made by many common modeling approaches provide unprecedented clarity and coherence to a complex and confusing subject.
-For example, assumptions that various aspects of flow behavior are stochastically or deterministically uniform all take the same form algebraically: left matrix multiplication.
-The other major class of common simplifying assumptions made by traffic models is to apply some marginal behavior across the entire collection of flows or to disjoint subsets of the flows.
-Such simplifications correspond to right matrix multiplication in \caps{GMM}.
-Because of the simplicity and clarity that the algebraic structure brings to the subject, we can see clearly now that both common types of simplifying assumptions are na\"ive and simplistic.
-Our experimental results bolster this conclusion: both uniform and marginal models significantly misrepresent many important network performance metrics.
-With the general matrix model, however, we stand poised to use the powerful tools of modern linear algebra to develop general, elegant and powerful models of network behavior that just work.
-
-%This research rigorously quantifies the impact of a variety of synthetic traffic models on performance metrics that wireless researchers use to evaluate new technologies and protocols.
-The first step in this assessment process was to formally define what it means for a network usage model to be sufficiently realistic.
-In essence, a model is considered realistic if it produces performance results that are statistically equivalent to those produced by real usage.
-%A well-defined, objective measure of realism for traffic models has not previously existed.
-Evaluations of realism have formerly relied on essentially arbitrary statistical measures of similarity to real traffic, which may or may not affect the performance metrics that researchers care about.
-%The definition of sufficient realism leads us to our general experimental approach: we use differential analysis comparing performance metrics derived from real traffic with those derived from synthetic traffic models.
-The theoretical contributions of this analysis are:
-%\begin{enumerate}
-%\item An in-depth analysis of the desirable mathematical properties of a measure of error for performance metrics.
-%\item Proof that the unique measure of error that satisfies these properties is the log-ratio of metric values.
-%\item Three rigorous tests of statistical equivalence between synthetic and real performance results.
-%\end{enumerate}
-%These analytical tools allow the evaluation of realism over a collection of drastically different usage scenarios.
-Evaluation over a heterogeneous collection of scenarios is essential to establishing the credibility of usage models.
-Moreover, these theoretical results are equally applicable to other types of usage models---for example, mobility.
-
-%On the practical side, this paper gives crucial insight into why most researchers do not trust simulation results: with the traffic models commonly used, the results are unlikely to reflect real performance.
-Moreover, this problem will also hamper experiments in test-bed wireless networks, so long as the same na\"ive workload models are used.
-The only way to address this fundamental lack of realism is to develop usage models that reproduce important performance metrics more accurately.
-Our theoretical results provide the tools necessary to do this.
-The development of better traffic models should begin with real traces, and proceed by incremental changes, checked by differential analysis.
-%%First, alter a small aspect of the trace, simulate, then compare.
-If the realism of the results is unaffected, the traffic feature altered was inessential.
-Otherwise, it is a feature of behavior that must be captured in a realistic traffic model.
-%This approach will allow the precise mapping of which aspects of traffic patterns have an impact on performance and which ones can be safely abstracted away.
-%% TODO: maybe cut out some of this last bit.
-
-%\section{Acknowledgments}
-%This work was funded in part through NSF Career Award CNS-0347886.
-
-% Hernandez06:dissertation: ``complete methodology for reproducing the traffic observed on a network link in a closed-loop manner, and proposed a number of metrics for studying the realism of the generated traffic.''
-\bibliographystyle{plain}
-\bibliography{references}
-
-\end{document}

### 2008-06-12 figures/box_plots.tex
Moved all tables & figures into separate files.

--- 2008-06-12 paper.tex
@@
-\documentclass[twocolumn,final]{svjour3}
-
-\newcommand{\thetitle}{Linear Representation of Network Traffic}
-\newcommand{\theauthors}{
-  Stefan~Karpinski \and
-  Elizabeth~M.~Belding \and
-  Kevin~C.~Almeroth \and
-  John~R.~Gilbert
-}
-
-%\usepackage{type1cm}
-%\usepackage[pdftex]{graphicx}
-\usepackage[labelfont=bf,small]{caption}
-\usepackage[font=small,labelfont=bf,position=top,nearskip=0em]{subfig}
-\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize,paralist,array}
-\usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle}]{hyperref}
-\usepackage{mathtools}
-
-\title{\thetitle}
-\author{\theauthors}
-\institute{
-  S.~Karpinski \and
-  E.~Belding \and
-  K.~Almeroth \and
-  J.~Gilbert \\
-  Department of Computer Science \\
-  University of California, Santa Barbara (USA) \\
-  Email: \texttt{\{sgk,ebelding,almeroth,gilbert\}@cs.ucsb.edu}
-}
-
-\input{definitions}
-
-\graphicspath{{graphics/}}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-We propose a representation of wireless workload patterns as large, sparse matrices and provide a method for stochastically generating experimental workload from a given matrix.
-The essential property of the algebraic representation is that the summation of vectors naturally yields a faithful description of the aggregate behavior of the corresponding flows.
-This deceptively simple property allows us to express many common concepts from traffic modeling succinctly in terms of a few linear transformations.
-The algebraic representation has many benefits: 1)~it makes the meaning of generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2)~it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3)~the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed; 4)~the vast body of algebraic theory and highly optimized numerical linear software may immediately be applied to traffic modeling.
-We use the paired differential simulation methodology introduced by the authors in previous work to experimentally demonstrate that the general matrix model accurately reproduces realistic network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-We use the same experimental methodology to explore the implications of various assumptions and simplifications that are commonly made in traffic modeling.
-%We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models.
-A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
-%
-%We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in wireless networks.
-Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematical precision and succinctness, while simultaneously being made clearer and more general.
-We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to discredit the common constant bit-rate model.
-A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
-We conclude with future directions for modeling research based on the new algebraic framework.
-\end{abstract}
-
-% The model of flow behavior implicit in non-negative matrix factoriation technique is that flow behavior is a composite of some limited set of basic behaviors---and that fractions of each flow ``belong'' to each behavior.
-This is not an \latin{a priori} unreasonable hypothesis.
-Contrast that with the assumptions underlying uniform and marginal simplifications of traffic behavior, which can be seen to be intuitively unreasonable without ever conducting and actual experiment or calculation.
-
-\keywords{
-  traffic modeling \and
-  traffic analysis \and
-  workload generation \and
-  wireless networks \and
-  wireless simulation \and
-  linear algebra \and
-  non-negative matrix factorization
-}
-
-\section{Introduction}\label{sec:intro}
-
-\newfootnote{\flownote}{We use the common definition of a \textit{flow} as a sequence of packets sharing the same  ``5-tuple'': \caps{IP} protocol type, source and destination nodes, and \caps{TCP/UDP} port numbers.}
-
-\newfootnote{\marginalnote}{The statistical term ``marginal'' refers to the margins of actuarial tables formerly used for statistical computations.
-The rows and columns of the table represent possible values of two properties.
-Each entry in a table contains a count of the number of events falling into the joint category for that row and column.
-The margins contain sums of the rows and columns, thus giving the unconditional distributions of each property.}
-
-More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local-area networks (\caps{LAN}s).
-Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
-One of the major challenges of modeling whole-network traffic is the interdependence of packet, flow\flownote and node behaviors.
-Three types of simplifying assumptions are commonly applied in both analysis and generation:
-\begin{enumerate}
-\item \textbf{Uniform:} that traffic properties are uniformly distributed. For example, assuming that  all nodes in a network have equal probability of being the source or destination of each flow.
-Another common example is the assumption that all packet sizes and inter-packet intervals are the same; this is commonly known as the ``constant bit-rate'' (\caps{CBR}) model.
-\item \textbf{Marginal:} that traffic properties share a single, common \emph{marginal}\marginalnote distribution across the entire network, and individual values are drawn independently and unconditionally from this shared distribution.
-For example, assuming that a single distribution of packet sizes describes all flows in a network, as compared to equipping each flow with its own specific distribution of sizes.
-\item \textbf{Conditional:} that traffic behaves marginally within equivalence classes defined by certain conditions.
-The assumption, for example, that all flows emanating from a common source node share the same packet size distribution is a conditional model, conditioned on the source nodes.
-\end{enumerate}
-We would be arguing against a straw man if we implied that researchers who apply these assumptions actually believe them to be true.
-They clearly are not:
-different nodes initiate and receive drastically different numbers of packets and flows;
-flows have drastically different packet size distributions.
-Simply considering intuitive examples of common behavior illustrates this amply:
-BitTorrent users \latin{vs.} occasional email checkers;
-downloading a large file \latin{vs.} typing in an \caps{SSH} session.
-
-% In short, the reason is that these assumptions drastically simplify both traffic analysis and workload generation.
-% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
-% This immediately begs the question of whether these assumptions are benign.
-% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to analyze and replicate?
-% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
-
-% TODO: is this paragraph necessary?
-% The second reason why uniformity and marginality assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
-% The difficulty lies in the high-dimensional and stochastic nature of network behaviors.
-% Suppose, for example, that we assume that packet sizes and inter-packet intervals are drawn independently from some given distribution for each flow.
-% In the terminology introduced above, this is a flow-conditional packet behavior model.
-% Prior research has indicated that, given accurate per-flow distributions, and keeping all other aspects of network behavior fixed, this simplification does not distort experimental performance metrics~\cite{Karpinski07:realism}.
-% In this model, however, each flow must have its own specific, realistic pair of distributions for packet size and inter-packet intervals.
-% Moreover, the ``distribution of distributions'' across the entire network must simultaneously be realistic.
-% Modeling such distributions of distributions realistically for entire networks is currently an unsolved problem; we begin to address it here.
-
-If no one actually believes that assumptions of uniformity and marginality accurately describe network behavior, then why are these assumptions so common?
-There are two reasons.
-The first is that these assumptions drastically simplify both traffic analysis and workload generation, making them tractable problems.
-% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
-% This immediately begs the question of whether these assumptions are benign.
-% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to understand?
-% This paper answers that question.
-The second is that there are no generally accepted alternatives for representing and simplifying traffic patterns that allow effective analysis and generation without assuming uniformity or marginality.
-% Representing complex multi-level network behaviors in a way that does not make these simplifying assumptions but allows effective traffic analysis and workload generation is currently an unsolved problem;
-% we begin to address it here.
-
-% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
-
-Fundamentally, uniform, marginal and conditional models are applied so that network traffic behavior can be represented using a manageable collection of empirical or analytical statistical distributions.
-Reduction in representation size and complexity makes analysis and generation feasible.
-From this perspective, three questions immediately present themselves:
-\begin{enumerate}
-\item Do these assumptions distort the metrics we are trying to analyze, replicate, and predict?
-\item Can we represent ``unreduced'' network behavior such that analysis and generation remain tractable?
-\item Are there better ways to simplify and reduce the representation of network behavior?
-\end{enumerate}
-This paper answers all three questions. We begin, however by answering the second problem and  
-using our solution to address the other two questions.
-% transforming the unreduced representation of traffic into it's uniform, marginal or conditional counterparts.
-
-To represent unreduced network behavior practically, we propose the concept of a \emph{linear representation} of network traffic.
-A linear representation of traffic maps each flow to a vector, such that the following linearity condition is satisfied:
-\begin{quote}
-The sum of the representations of two flows is a vector that represents the aggregate behavior of the two flows.
-\end{quote}
-The complete behavior of a network is expressed as a matrix of such vectors, having one row per flow.
-Linear representations express full network behavior, without imposing assumptions of uniformity or marginality, yet still allow effective traffic analysis and workload generation.
-We present a specific linear representation, called the general matrix model (\caps{GMM}), and demonstrate experimentally that this model generates wireless workload that accurately reproduces the performance characteristics of real wireless network traffic.
-
-\newfootnote{\ranknote}{For uniform and marginal assumptions, the multiplication matrix has rank 1  since all rows and columns are identical.
-In the conditional case, the rank of the matrix is the number of conditioning classes.}
-
-To address the the other two questions, we begin with a simple observation.
-In linear representations of traffic, uniform, marginal and conditional assumptions are assertions of equality between the rows and columns of the traffic matrix.
-Moreover, real traffic matrices can be transformed into uniform, marginal and conditional forms through very simple matrix multiplications.
-These transformations, however, are extremely destructive: the matrices are degenerate with very low rank, so multiplication destroys almost all information in the original traffic matrix.
-We demonstrate experimentally, that this loss of information is not benign:
-uniform and marginal traffic models severely distorting crucial network performance metrics.
-
-\input{figures/trace_vs_uniform}
-
-Once traffic is represented linearly, alternatives to assuming uniformity or marginality become apparent.
-We argue that matrix factorization, which is nondestructive and does not assume uniformity or marginality, is far superior to multiplication for this problem.
-To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF})~\cite{Lee01} to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
-This factorization extracts a small set of ``basic behaviors'' from the traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
-Although the factorization is computed without examining port numbers, basic behaviors nevertheless correspond to intuitive expectations for protocols.
-We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.
-
-% TODO: update roadmap paragraph.
-
-The rest of the paper is organized as follows.
-In \Section{motivation} we discuss motivation and related work.
-The notion of linear representation and the general matrix model are presented in \Section{general-matrix-model}.
-In Sections~\ref{sec:common-properties}~and~\ref{sec:modeling-concepts} we explore how common network properties and modeling concepts can be expressed in the the \caps{GMM} framework.
-Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
-The ramifications of these results are discussed in \Section{discussion}.
-Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies.
-
-\section{Motivation \& Related Work}
-\label{sec:related-work}
-\label{sec:motivation}
-
-\newfootnote{\QoSnote}{This is violated by some quality of service (QoS) schemes.
-However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid.
-The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.}
-
-The interaction of network users and applications with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted.
-The joint pattern of traffic generation and mobility through time and space completely determines the effect of network usage on the lower levels of the stack.
-This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner~\cite{Clark88}.\QoSnote The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic.
-\Figure{trace-vs-uniform} illustrates through examples how drastically different network traffic appears under assumptions of uniformity at different levels of behavior.
-Marginality assumptions are not as visually striking, but as we will demonstrate, they distort the performance characteristics of networks almost as much, and sometimes more, than uniform models.
-
-While this work is not inherently specific to wireless networks, since the modeling techniques developed are equally applicable to traffic in wired networks, the results are of special importance to wireless research.
-Realistic traffic models have a drastic impact on experimental results in wireless networks~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-This is primarily because wireless networks are locally resource constrained.
-Compare this with the typical situation in wired \caps{LAN}s, where gigabit Ethernet and modern routers  can easily handle all but the most intense levels of traffic.
-If a wired network operator expects larger volumes of traffic on a conditional basis, the solution is to upgrade the hardware, not to change the network protocols.
-In wireless networks, on the other hand, the physical medium cannot be upgraded; protocols must be improved to make better usage of the medium.
-Being able to perform experiments with realistic workload is a vital part of the ongoing process of improving wireless protocols.
-Without realistic workload models for experiments, performance comparisons may be misleading or even completely wrong~\cite{Karpinski07:cbr-failure}.
-
-Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the emergent behavior is governed by a closed feedback loop.
-In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
-Hern\'andez-Campos' dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and typical \caps{TCP} applications.
-He shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
-This research has subsequently been turned into a \caps{TCP} flow generation tool called \tmix~\cite{Weigle06}.
-The acronym is somewhat misleading: {\tmix} does not provide models for application behaviors or for mixtures of applications in networks.
-Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
-Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary inputs to {\tmix}.
-
-\newfootnote{\harpoonnote}{Even this is unclear: generated traffic is only compared with trace traffic using the very same metrics that are explicitly specified as part of the model.
-Unsurprisingly, the distributions of sampled metrics closely match the distributions those metrics were sampled from.
-Realism is not double-checked using other statistical metrics or actual performance metrics.}
-
-The two most prominent general traffic generation frameworks are Harpoon and \caps{D-ITG}.
-Harpoon~\cite{Sommers04} uses a traffic trace for self-training, and can subsequently generate synthetic traffic with certain statistical properties based on the original trace.
-The properties reproduced are the empirical distributions of the following: ``file size, inter-connection time, source and destination IP ranges, [and] number of active sessions.''
-Harpoon exemplifies the uniform and marginal modeling approaches.
-Source and destination nodes are treated marginally: empirically derived marginal frequencies of sources and destinations are used to choose endpoints independently for each flow.
-Flow sizes (\latin{i.e.} file sizes) are likewise treated marginally: each flow's total byte  count is sampled from a marginal distribution of flow sizes.
-\caps{UDP} flows are modeled as being constant bit-rate;
-\caps{TCP} flows are all treated as file transfers, making them effectively uniform with bit-rate determined by \caps{TCP} dynamics.
-Inter-connection times are sampled from a single empirically derived marginal distribution.
-While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic,\harpoonnote for generating local-area traffic in wireless experiments, it is not.
-Previous research has shown that uniform packet behavior models used in Harpoon drastically distort vital performance metrics at all levels of the protocol stack~\cite{Karpinski07:cbr-failure}.
-This paper shows, moreover, that marginal models, like those used in Harpoon, also severely misrepresent performance metrics in wireless networks.
-
-Like Harpoon, \caps{D-ITG}~\cite{Avallone04} exemplifies uniform and marginal modeling assumptions. However, unlike Harpoon, \caps{D-ITG} does not derive models from real traffic.
-Packet behavior can be modeled using standard statistical distributions (``constant, uniform, normal, cauchy, pareto, exponential, etc.'').
-Guidance is not provided, however, for choosing among these distributions, choosing parameter values, or selecting a mixture of distributions across a network.
-Flow and node behaviors are not specified at all, but rather left to the user's discretion.
-The focus of this project is primarily on the engineering challenge of generating very large volumes of synthetic traffic and injecting it into networks in a distributed fashion.
-While \caps{D-ITG} provides the mechanism for generating large volumes of traffic, it does not provide any real guidance as to what traffic to generate.
-
-\FHC~\cite{Hernandez06:spatio-temporal-modeling} have defined the state of the art in parametric modeling of wireless traffic patterns.
-They use rigorous statistical methods to fit parametric models to a number of properties of wireless traffic.
-Flow arrivals fit a non-stationary Poisson model;
-flow inter-arrival times fit a Lognormal distribution;
-the number of flows per session fits a BiPareto model;
-flow sizes also fit a BiPareto model.
-The statistical analysis fitting these parametric models is thorough and convincing.
-The models, however, are marginal:
-a single statistical model applies to all flows and nodes; packet behavior is not considered.
-Each node has a different number of flows assigned to it, but otherwise nodes are identical. 
-Moreover, all flows, regardless of source node, are statistically identical:
-flow sizes are drawn from a single marginal distribution, and nothing else distinguishes them.
-Yet, intuitively it is clear that different nodes in real networks have drastically different tendencies for flow size, flow rates, and packet behavior.
-Such interdependent behaviors cannot be captured using marginal modeling.
-In this paper, we demonstrate that this inability is not harmless:
-treating all flows and nodes statistically identically severely distorts crucial wireless performance metrics in experiments.
-We also provide sorely needed methods of representing and analyzing traffic patterns without making uniform or marginal assumptions about network behavior.
-
-This work uses the technique known as \emph{paired differential simulation} to evaluate the realism of modified or synthetically generated wireless workloads.
-This technique was proposed and subsequently refined by Karpinski~\latin{et~al.}~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-They introduce the notion of \emph{sufficient realism}, and propose paired differential simulation as an experimental methodology based on the standard technique of pairing control and test subjects.
-We describe the technique in \Section{methodology}.
-In the first paper~\cite{Karpinski07:realism}, they explore various ways of modifying trace traffic patterns, as a form of sensitivity analysis to determine which aspects of network behavior are essential and must be preserved in realistic models, and which aspects can be discarded as inessential.
-Of particular interest is their conclusion that detailed time-series packet behavior for flows is inessential:
-it suffices, for each flow, to have accurate unconditional distributions of packet sizes and inter-packet intervals;
-sufficiently realistic flow behavior can be reconstructed from these distributions by repeated independent sampling.
-In the second paper~\cite{Karpinski07:cbr-failure}, Karpinski~\latin{et~al.} explore the impact of a broad variety of uniformity assumptions at the packet, flow and node levels of behavior.
-They find that all uniformity assumptions are detrimental, inducing significant and often extreme misrepresentation of a broad variety of wireless performance metrics at every level of the protocol stack.
-Moreover, they show, using \caps{OLSR} and \caps{AODV}, that uniform traffic models can completely invert the relative performance of network protocols.
-
-\section{Linear Representation of Traffic}
-\label{sec:representations}
-
-The traffic workload pattern in a network is a collection of \caps{IP} packets between hosts in the network sent at certain times.
-It is standard in network analysis, however, to aggregate sequences of packets sharing the same ``5-tuple'':
-\caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP/UDP} port numbers.
-Such a sequence of packets is called a \textit{flow}.
-The traffic pattern of an entire network is simply the collected behaviors of all flows occurring in the network.
-
-The behavior of a flow as it affects the network is characterized by the following properties:
-its \caps{IP} protocol type (\caps{TCP}, \caps{UDP}, \caps{ICMP}, etc.);
-its source and destination nodes;
-its start time;
-and the specific sizes packets of and intervals between their transmission.
-We begin our exposition of traffic representation by formalizing these properties mathematically as sets:
-\begin{tabbing}
-\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{5em}\=\hspace{4em}\=$\in\Nodes\,$\=\kill
-\>\bullet\>types:\>$\type$\>$\in\Types$\>$= \set{1,\dots,255} \subset \N$ \\
-\>\bullet\>nodes:\>$\src,\dst$\>$\in\Nodes$\>$= \set{1,\dots,n} \subset \N$ \\
-\>\bullet\>times:\>$\start$\>$\in\Times$\>$= \R$ \\
-\>\bullet\>sizes:\>$\size$\>$\in\Sizes$\>$= \set{0,\dots,\text{\caps{MTU}}} \subset \N$ \\
-\>\bullet\>intervals:\>$\ival$\>$\in\Ivals$\>$ = \R^+ \union \set{\infty}$.
-\end{tabbing}
-There are 255 possible \caps{IP} types (0 is reserved for \caps{IPv6}).
-The number $n$ is simply the total number of nodes in the network being modeled.
-\caps{MTU} is the maximum transfer unit of the network (typically around 1500 bytes).
-We formalize the packet behavior of a flow as a pair of infinite sequences of packet sizes and inter-packet intervals:
-\begin{align}
-\sizes &= \seq{\size_1,\size_2,\dots} \in \SizeSeq \\
-\ivals &= \seq{\ival_1,\ival_2,\dots} \in \IvalSeq.
-\end{align}
-Since flows contain only a finite number of packets, there exists a number of packets, $\pkts \in \N$, for each flow.
-Since flows are finite, we require that the sequences $\sizes$ and $\ivals$ satisfy the following requirements:
-\begin{align}
-&\forall\: k \le \pkts: \quad \size_k > 0       \eqn{positive-packets}  \\
-&\forall\: k >   \pkts: \quad \size_k = 0       \eqn{zero-packets}      \\
-&\forall\: k <   \pkts: \quad \ival_k < \infty  \eqn{finite-intervals}  \\
-&\forall\: k \ge \pkts: \quad \ival_k = \infty. \eqn{infinite-intervals}
-\end{align}
-Packet sizes are required to be positive because we are modeling application layer data transfer:
-an empty data transmission request is a non-event.
-The send time of the $k$th packet can be expressed as
-\begin{align}
-\time_k = \start + \sum_{i=1}^{k-1} \ival_i.
-\end{align}
-This is guaranteed to be finite if and only if $k \le \pkts$.
-Now we are ready to formalize the space of flows:
-\begin{align}
-  \Flows =
-    \Types    \cross
-    \Nodes^2  \cross
-    \Times    \cross
-    \N        \cross
-    \SizeSeq  \cross
-    \IvalSeq.
-\end{align}
-We write a generic element of $\Flows$ as
-\begin{align}
-  \flow = \tuple{
-    \type,
-    \src,
-    \dst,
-    \start,
-    \pkts,
-    \sizes,
-    \ivals
-  } \in \Flows.
-\end{align}
-Let $2^\Flows$ denote the power set of $\Flows$, and $\Traffic \subseteq 2^\Flows$ the set of all finite sets of flows:
-\begin{align}
-  \Traffic = \set{X \in 2^\Flows : \card{X} < \infty}.
-\end{align}
-A traffic sample is simply an element of $\Traffic$. We will refer to elements of $\Traffic$ interchangeably as ``traffic samples'' or ``traffic instances.''
-
-Having chosen a mathematical formalization of network traffic, we may now define the notion of a linear representation of network traffic.
-
-\begin{definition}\defn{linear-representation}
-A \defterm{linear representation} of network traffic is a pair, $\tuple{V,\psi}$, where $V$ is a vector space and $\psi$ is a function, $\psi: \Traffic \to V$, such that
-\begin{align}\eqn{linearity}
-\forall~T \in \Traffic\!:\quad
-  \psi(T) &= \sum_{\mathclap{\flow \in T}}{\psi(\flow)}.
-\end{align}
-\end{definition}
-
-\noindent
-In plain words, every finite collection of flows is represented by a vector that is the sum of the elements representing the individual flows in the collection.
-% Representing traffic using vectors allows us to analyze, model, and understand traffic patterns in terms of the thoroughly studied and well understood subject of linear algebra.
-The careful reader will notice that \emph{any} function $\Flows \to V$ uniquely determines a linear representation:
-\Equation{linearity} uniquely extends the function to all of $\Traffic$.
-Representing flows by arbitrary vectors, however, is not particularly useful or interesting.
-We focus instead on meaningful representations, where collections of flows are represented by vectors that naturally describe the aggregate behavior in some manner.
-In the next section, we clarify these concepts through examples.
-
-\subsection{Examples of Linear Representations}
-\sec{linear-representation-examples}
-
-\newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $\pkts$ does not divide $\mathsym{bytes}$ evenly.
-The simplest is to ignore it; we leave leave more complex schemes to the reader.}
-
-Our first example is the packets-bytes-duration (\caps{PBD}) representation.
-The representation is $\pbd: \Traffic \to \R^3$, mapping
-$\flow \mapsto \tuple{p,b,d} \in \R^3$, where
-\begin{align}
-p &= \pkts, &
-b &= \sum_{i=1}^{\mathclap{p}}{\size_i}, &
-d &= \sum_{i=1}^{\mathclap{p-1}}{\ival_i}.
-\end{align}
-This representation encapsulates the information necessary to reproduce a \caps{CBR} version of each flow:
-packets have payloads of $\floor{\fracx{b}{p}}$ bytes, and are sent every $\fracx{d}{p}$ seconds.\pbdnote
-Given two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, their sum as vectors, 
-\begin{align}
-\vec{x}_1+\vec{x}_2=\tuple{p_1+p_2,b_1+b_2,d_1+d_2},
-\end{align}
-provides an appropriate description of the aggregate behavior of the two flows together.
-Similarly, their average, \latin{i.e.} $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
-In contrast, a ``packets-payload-interval'' representation, expressing the \caps{CBR} parameters directly, does not have this property:
-the sum of average payloads is not the overall average of payloads;
-likewise for intervals.
-In the \caps{PBD} representation, the linearity property leads to natural descriptions of aggregate behavior through vector addition, whereas in the packets-payload-interval representation, vector addition is ``unnatural.''
-
-The \caps{PBD} representation is highly lossy: it discards most of the information detailing the behavior of each flow.
-At the other extreme, we can give a ``perfect'' representation of each flow, that allows us to completely reconstruct each flow from its representation.
-With care, the sum of representations of flows can still naturally describe their aggregate behavior.
-Let the representation be $\Phi: \Traffic \to \FlowVec$, where
-\begin{align}
-\FlowVec = \R^{255} \cross \R^n \cross \R^n \cross L^2[\R] \cross L^2[\R].
-\end{align}
-% Here $\R^X$ is the vector space of functions $f: X \to \R$, which is isomorphic to the standard Euclidean vector space, $\R^\card{X}$.
-Here $L^2[\R]$ is the vector space of real functions, $\R^\R=\set{f: \R \to \R}$, with the usual norm~\cite{Rudin87}.
-% \begin{align}
-% \norm{f}_2 = \Big[ \int_{-\infty}^{\infty} \abs{f}^2 dx \Big]^{\frac{1}{2}}.
-% \end{align}
-We define
-\begin{align}
-\Phi(\flow) = \tuple{\e_\type,\e_\src,\e_\dst,f,g},
-\end{align}
-where $\e_k$ is the $k$th standard basis vector, and $f,g \in L^2[\R]$ are functions defined by
-\begin{align}
-f(t) &= \sum_{k=1}^{\pkts}
-  \begin{cases}
-    1 & \text{if $t \ge \time_k$} \\
-    0 & \text{if $t <   \time_k$},
-  \end{cases} \\
-g(t) &= \sum_{k=1}^{\pkts}
-  \begin{cases}
-    \size_k & \text{if $t \ge \time_k$} \\
-    0       & \text{if $t <   \time_k$}.
-  \end{cases}
-\end{align}
-Thus, $f$ and $g$ are both integer-valued, monotonically increasing step functions, giving the cumulative number of packets and total bytes sent, respectively, at time $t$.
-From the representation, $\Phi(\flow)$, of a single flow, we can readily recover its type, source, destination, and the transmission times and sizes of packets.
-
-Why is $\Phi$ natural in the same sense that the \caps{PBD} representation is natural?
-Suppose we have a traffic sample, $T\in\Traffic$.
-The first component of $\Phi(T)$ is a 255-dimensional vector, whose coordinates are a histogram of the number of times each flow type occurs in $T$.
-Similarly, the second and third components of $\Phi(T)$ are $n$-dimensional histograms of the number of times each node occurs as source and destination.
-The last components are step functions, like $f$ and $g$.
-Vector addition in $L^2[\R]$ dictates that these functions give the total number of cumulative packets and bytes sent, respectively, over all flows in $T$.
-Thus, summation of vectors under the representation $\Phi$ yields a natural and useful description of the aggregate behavior of flows.
-
-Consider, for contrast, if we had represented flow type simply by $\type \in \N$ instead of $\e_\type \in \R^{255}$.
-The natural numbers containing $\type$ can readily be embedded into $\R^1$, which is a vector space.
-However, the addition of vectors is highly unnatural.
-The sum of protocol numbers is meaningless:
-two \caps{TCP} flows plus five \caps{ICMP} flows does not equal a single \caps{UDP} flow, despite the fact that $2\cdot6+5\cdot1=17$.
-The summation of vectors in the representation $\Phi$, on the other hand is $5\,\e_1+2\,\e_6$, indicating the frequency with which each type occurred.
-
-Both of the example representations discussed here can be applied in practice.
-The \caps{PBD} representation naturally allows us to compute aggregate and average \caps{CBR} behaviors of collections of flows.
-With a suitable machine encoding of step-functions, $\Phi$ can be implemented and used in practice.
-It provides complete information about individual flows and highly detailed (though not complete) information about collections of flows.
-Representations that preserve even more information about traffic instances can readily be constructed.
-We turn now, however, in a different direction, and introduce a specific linear representation that strikes a practical balance in fidelity between $\pbd$ and $\Phi$.
-
-\section{The General Matrix Model}
-\sec{general-matrix-model}
-
-\input{tables/gmm_properties}
-
-The \caps{PBD} representation of flows uses 3 dimensions to encode the behavior of each flow, while $\Phi$ uses $255+2n$ dimensions, plus the continuum of dimensions of two Banach spaces (\latin{i.e.} the two copies of $L^2[\R]$).
-In this section, we define a representation, $\phi$, that splits the difference and uses a large but finite number of dimensions to represent each flow.
-Once the vector representation of each flow is defined, we may construct matrices of representation vectors, with a row for each flow in a traffic sample.
-We call the matrix representation of network traffic the \emph{general matrix model}~(\caps{GMM}).
-Having expressed network behavior in terms of matrices, we can very simply and efficiently compute a broad variety of network behavior properties using standard matrix operations.
-Furthermore, we can express uniformity and marginality assumptions about behavior as equality constraints between the rows and columns of the traffic matrix.
-Moreover, we can transform unconstrained traffic instances into uniform, marginal, or conditional approximations via right or left multiplications by easily computed matrices.
-
-The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins---%
-one per value for properties that are discrete already and require no compression---%
-and map values in each bin to a standard basis element of $\R^d$, where $d$ is the number of bins.
-In this scheme, representation vectors are effectively histograms, with coordinate values in each dimension counting items falling into the corresponding bin.
-% The histograms for packet sizes and inter-packet intervals count packets, while other histograms count flows.
-\Table{quantization} lists the properties that are used to define $\phi$, together with the quantization and dequantization functions.
-Each quantization function takes a possible property value as input and maps it to an index value in $\set{1,\dots,d}$, where $d$ is the number of bins, \latin{i.e.} dimension, for the property in question.
-% The index assigned determines the histogram bin of the input property.
-
-The histogram-vector approach produces some representations of properties that are not immediately intuitive, and may seem to use an excessive number of representation dimensions.
-However, this approach automatically satisfies the ``naturalness'' criterion for linear representations.
-That is, the representation, $\phi(T)$, naturally describes the aggregate behavior of flows in $T$.
-Perhaps even more importantly, arbitrary linear combinations of vectors can be meaningfully interpreted as describing some realizable, hypothetical traffic instance.
-In \Section{traffic-generation} we will describe how to generate a flow for any sum of representation vectors.
-% This is how we generate workload from transformed traffic matrices.
-
-The quantization functions specified in \Table{quantization} allow us to define the representation $\phi: \Traffic \to \R^N$: \begin{align}
-\phi(\flow) = \vec{f} = \tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}},
-\end{align}
-where $N=255+2n+d_t+d_b+d_z+d_v$ and vectors are
-\begin{align*}
-\vec{t} &= \e_k & k &= \Qf^{\text{t}}(\type) \\
-\vec{s} &= \e_k & k &= \Qf^{\text{s}}(\src) \\
-\vec{d} &= \e_k & k &= \Qf^{\text{d}}(\dst) \\
-\vec{a} &= \e_k & k &= \Qf^{\text{a}}(\start) \\
-\vec{b} &= \e_k & k &= \Qf^{\text{b}}(\bytes) \\
-\vec{z} &= {\textstyle \sum \e_k} & k &\in \set{\Qf^{\text{z}}(\size_i) : i \le \pkts} \\
-\vec{v} &= {\textstyle \sum \e_k} & k &\in \set{\Qf^{\text{v}}(\ival_i) : i <   \pkts}.
-\end{align*}
-The functions $\Qf^\text{x}$ are the respective quantization functions for each property.
-If there are $f$ flows in a traffic instance, then the collective behavior can be described with seven matrices---one for each property;
-the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
-We use the following notations for the behavior matrices:
-\vspace{-0.25em}
-\begin{tabbing}
-\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
-$\mat{D}\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
-\>\bullet\> type: \>$\mat{T}$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$\\
-\>\bullet\> sources: \>$\mat{S}$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$\\
-\>\bullet\> destinations: \>$\mat{D}$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$\\
-\>\bullet\> start times: \>$\mat{A}$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$\\
-\>\bullet\> sizes in bytes: \>$\mat{B}$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$\\
-\>\bullet\> packet sizes: \>$\mat{Z}$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$\\
-\>\bullet\> inter-packet intervals: \>$\mat{V}$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
-\end{tabbing}
-The total behavior matrix is the horizontal concatenation of these component matrices:
-\begin{align}
-\mat{F} = \left[\,\mat{T}~\mat{S}~\mat{D}~\mat{A}~\mat{B}~\mat{Z}~\mat{V}\,\right] \in \R^{f \times N}.
-\end{align}
-We call this total representation the general matrix model.
-While it is by no means the only possible representation of traffic in terms of matrices, as demonstrate in the following sections, many common traffic modeling assumptions can be expressed succinctly and precisely in terms of transformations or conditions of the \caps{GMM} and its components.
-Furthermore, realistic workload can be generated from matrix instances.
-
-\subsection{Choosing Constants}
-
-Before continuing with less mundane topics, we must briefly address our choices of constants appearing in \Table{quantization}.
-The externally predetermined constants are:
-$d_z=\text{\caps{MTU}}=1500$, determined by the wireless network medium;
-$b_{\max} \approx 109 \text{\caps{MB}}$, determined by the size of the largest flow in our trace data;
-$t_{\max}=v_{\max}=600$, determined by our experimental methodology which slices traces into 10-minute scenarios;
-and $v_{\min}=10^{-6}$, determined by the time resolution of our trace.
-For lack of a better choice, we have simply chosen to use the same number of quantization bins as packet size for the unfixed dimension constants:
-$d_t = d_b = d_v = 1500$.
-Arguments can certainly be made for other choices, but this choice appears to work adequately in practice.
-
-Two other unspecified constants appear in \Table{quantization}: $\beta$ and $\gamma$.
-These parameters are non-linear scaling exponents that allows the quantization to smoothly shift from fine-grained resolution at the low end of the spectrum to coarse-grained resolution at the high end.
-We have chosen $\beta=\fracx{1}{2.7}$ because it yields a desirable dynamic range response from bytes at the low end to megabytes at the high end.
-Similarly, we have chosen $\gamma=\fracx{1}{3}$ because this choice provides resolution that scales smoothly from microseconds for small intervals to seconds for large intervals.
-
-\subsection{Traffic Generation}
-\sec{traffic-generation}
-
-Given an instance of the general matrix model, how do we produce an actual workload from it?
-Each row of the model is used to randomly generate the behavior of a single flow.
-To do this, we use the ``dequantization'' functions listed in \Table{quantization}.
-The dequantization procedure works as follows.
-Let $\vec{f}$
-% \begin{align}
-% \vec{f}=\tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}}
-% \end{align}
-be a row of the traffic matrix, and let $\prop$ be a property.
-Write $\vec{f}_\prop$ for the corresponding component vector of $\vec{f}$.
-Chose an index value $k\in\set{1,\dots,d_\prop}$, randomly, with weight given by the coefficients of $\vec{f}_\prop$.
-Choose $u \in [0,1]$ uniformly randomly.
-Then $x = k-u \in [0,d]$, with probability of lying in the interval $(i-1,i]$ proportional to the coefficient of $\e_i$ in $\vec{f}_\prop$.
-The dequantization function for $\prop$, call it $\Df_\prop$, will map the value $x$ back into the appropriate range of property values for $\prop$.
-
-The above procedure allows us to sample random values for each property in accordance with the behavioral description implied by $\vec{f}$.
-This allows us to stochastically generate a sequence of packets in accordance with a row from an arbitrary \caps{GMM} matrix instance using the following procedure:
-\begin{enumerate}
-\item Sample random source and destination node pairs.
-Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
-The simplest technique is the rejection method:
-if the same node is chosen for both source and destination, reject that pair and begin the selection process again.
-\item Next, randomly sample the number of bytes.
-\item The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\mean{z}$ and $\mean{v}$.
-From these, together with the flow size in bytes, we estimate the expected duration of the flow:
-$\mean{d} = \fracx{b\mean{v}}{\mean{z}}$.
-\item Randomly sample a start time for the flow.
-Ensure that the start time plus the expected duration, $\mean{d}$, does not exceed the maximum simulation time for the workflow by using the rejection method.
-\item The first packet is sent at the chosen start time.
-The size of each packet is determined by sampling a random size value.
-The interval until the next packet is determined by sampling a random interval value.
-The packet stream ends when the number of bytes allocated to the flow have been emitted or when the packet emission time exceeds $t_{\max}$.
-The last packet is only the size of the remaining number of bytes.
-\end{enumerate}
-Some remarks about the procedure:
-\begin{itemize}
-\item For some vectors in $\R^{f \cross N}$, it may be impossible to choose distinct source and destination nodes or flow sizes and start times satisfying the procedure.
-However, for any vector that is a sum of vectors representing valid flows---\latin{i.e.} ones having distinct source and destination nodes and feasible combinations of start time, duration, and packet behaviors---the procedure will terminate.
-\item Because the representation $\phi$ models the seven properties independent of each other, it is impossible to do better than to sample the properties independently, modulo the requirements of the procedure.
-In the case of a row which is simple a representation of an actual flow, this makes no difference, except for the generation of packets (which are modeled independently).
-In the case of composite rows, however, this implies that flow properties are modeled independent of each other.
-\end{itemize}
-Despite these remarks, we will show in our experimental evaluation, that this procedure generates sufficiently realistic wireless workloads.
-
-\subsection{Expression of Common Properties}
-\label{sec:common-properties}
-
-Having defined our representation of network behavior and presented the procedure for stochastically generating workload, we turn to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.
-
-\subsubsection{Communication Topology}
-\label{sec:communication-topology}
-
-The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{SD}}\in\R^{n \times n}$ ($\trans{\,\diamond\,}$~denotes transpose).
-That is $(\trans{\mat{SD}})_{ij}$ is the number of flows from node $i$ to node $j$.
-The total number of source flows for each node is given by the vector $\trans{\mat{S}}\ones{f}$, and the total number of destination flows is given by $\trans{\mat{D}}\ones{f}$.
-($\ones{f}$ is the $f$-dimensional column vector consisting of all ones.)
-
-\subsubsection{Expected Average Packet Size \& Intervals}
-\label{sec:flow-totals}
-
-From the matrices $\mat{Z}$ and $\mat{V}$, we can compute the expected average packet size, $\mean{z}$, and average inter-packet interval, $\mean{v}$, for each flow.
-The computations require knowing the values that each matrix column represents.
-The computed duration is approximate because information is irrecoverably lost in the quantization process.
-The vectors of representative values for $\mat{Z}$ and $\mat{V}$ are:
-\begin{align}
-\vec{\sigma}&=\tuple{1,2,3,\dots,d_z},\\
-\vec{\tau}&=\tuple{\Di\left(k,\fracx{1}{2}\right)}_{k=1}^{d_v}.
-\end{align}
-Using these representative values, the expected average packet size and expected average inter-packet interval size for a flow may be computed as
-\begin{align}
-  \mean{z}&=\frac{\inner{\vec{z}}{\vec{\sigma}}}{\inner{\vec{z}}{\ones{}}}, &
-  \mean{v}&=\frac{\inner{\vec{v}}{\vec{\tau}}}{\inner{\vec{v}}{\ones{}}},
-\end{align}
-respectively, where $\inner{\vec{x}}{\vec{y}}$ denotes the standard inner product of vectors.
-
-\subsection{Expression of Common Modeling Concepts}
-\sec{modeling-concepts}
-
-The simplifying assumptions made by most traffic models can be classified into three categories.
-We use the terms \textit{uniformity}, \textit{marginality}, and \textit{conditionality} to describe these classes of assumptions.
-They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether the make stipulations about uniformity within each flow, across all flows, or within groups of flows.
-In the following sections, we explore how various modeling concepts can be expressed in the framework of the general matrix model.
-Models are expressed as transformations that will convert arbitrary traffic matrix instances into similar ones that satisfy the model's assumptions.
-These transformations will later allow us to evaluate the effect of each model's simplifications on the realism of generated workloads.
-
-\subsubsection{Uniformity}
-
-The concept of uniformity, as used here, entails that certain properties of each flow are assumed to be statistically or deterministically uniform.
-Examples include: assuming that each flow's packets have the same size and inter-packet interval (i.e.
-the \caps{CBR} flow model); assuming that each node is equally likely to be the source of a given flow; assuming that all start times for a flow are equally likely.
-
-Since uniform behaviors entail homogeneity of properties with respect to each flow individually, a trace matrix can be transformed to a matrix satisfying the model's assumptions via right matrix multiplication.
-Right multiplication mixes the elements within each row, and if the mixing weights are uniform, the result is uniform behavior.
-For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $\mat{S}$ and $\mat{D}$:
-\begin{align*}
-\mat{S}' &= \mat{S}\ones{n \times n}/n, & \mat{D}' &= \mat{D}\ones{n \times n}/n.
-\end{align*}
-This makes all the values in each flow's row identically the average of the original $n$ values.
-As another example, the constant bit-rate (\caps{CBR}) flow model, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
-The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $[\mat{Z}~\mat{V}]$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $[\mat{Z}~\mat{V}]\,\Gamma\in\R^{f \times 3}$, where
-\begin{align}
-\Gamma=\begin{pmatrix}
-\ones{} & \vec{\sigma} & \zeros{} \\
-\zeros{} & \zeros{} & \vec{\tau}
-\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
-\end{align}
-\Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to realistic behaviors from traces.
-%By replacing $\mat{S}'$ with $\mat{S}$, this can also be viewed as a set of linear constraints on the matrix $\mat{S}$, providing a proscriptive definition
-
-%We use the term \textit{uniformity} do describe assumptions made about the ``sameness'' of various properties of flows.
-The constant bit-rate (\caps{CBR}) model, for example, is an assumption about the uniformity of all of a flow's packets: i.e.
-that they have the same size and are evenly spaced in time.
-While this is the strictest sense of the \caps{CBR} model, it is rare for this assumption to be made in isolation.
-Another common simplification is to make the sizes of flows identical, or chose them randomly from a uniform distribution; similarly for the starting time of each flow.
-Finally, when choosing endpoints for a given flow, lacking a better model, it is not unusual to treat all nodes identically, making each an equally likely source or destination for a given flow.
-
-%It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees.
-This is largely because of the high complexity of doing otherwise and the lack of compelling alternatives.
-Here we examine how various notions of uniformity are expressed in the matrix model.
-In general, \textit{intra-flow uniformity} is effected by right multiplication, which mixes the elements in each row, whereas \textit{inter-flow uniformity} is effected by left multiplication, which mixes elements across rows.
-The constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
-The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $\tuple{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\tuple{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
-%\begin{align}
-%\Gamma=\begin{pmatrix}
-%\ones{} & \vec{\sigma} & \zeros{} \\
-%\zeros{} & \zeros{} & \vec{\tau}
-%\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
-%\end{align}
-
-\subsubsection{Marginality}
-
-Marginality is an orthogonal concept to uniformity: the distribution of some property is identical \textit{across} all flows.
-For example, if we determine a distribution of flow sizes across the entire network and assume that this distribution applies \textit{independent of other flow properties}, this is a marginal model for flow size.
-Marginal behavior is effected by left multiplication by a matrix with uniform columns.
-This specific example is generated by left-uniformizing the flow size matrix: $\mat{B}' = (\ones{f \times f}/f)\mat{B}$.
-The traffic models proposed by {\FHC} are all marginal: they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node.
-These models do not account for non-independence between these and other flow properties.
-We will discuss this limitation and its implications when we analyze our experimental results.
-
-\subsection{Conditionality}
-
-The final form of homogenization we consider, is conditionality.
-This is a restricted form of marginality: we may assume that marginal distributions apply to all the flows in certain groupings.
-For example, we can provide a distribution of flow sizes on a per-source basis.
-That is, each source node can have its \textit{own} distribution, and the sizes of its flows are drawn from that distribution, independent of their other properties.
-This is simply accomplished by left-multiplication by a non-uniform matrix.
-For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-conditioning'' matrix: $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$.
-Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-conditioning matrix: $F'=(\trans{\mat{DD}}/f)\mat{F}$.
-%Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified.
-The confusion arises due to the lack of a unique way of associating flows with nodes.
-Each node can equally viably be associated with its source or its destination.
-Which association is more valid?
-Consider a web request: should the behavior of the request be determined by what the client typically does?
-Or should the requests that the server typically receives determine behavior?
-There is no armchair answer: only analysis of real data can give us the answer.
-The matrix model provides an analytical framework for doing so.
-The matrix model has the benefit of making the conditioning of behaviors completely explicit.
-It also allows us to easily define intermediate choices.
-If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
-\begin{align}
-\mat{F}' = \frac{1}{f}((1-\alpha)(\trans{\mat{SS}})+\alpha(\trans{\mat{DD}}))\mat{F}.
-\end{align}
-The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors.
-Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.
-
-%We use the terms here to distinguish three fundamentally different types of assumptions about behavioral homogeneity.
-
-%Existing traffic models are typically not explicit methods of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
-In our algebraic modeling framework, such models can be interpreted as stipulations about the form of the behavior matrices, or as transforms that implement those stipulations given a real set of traffic matrices.
-
-%A generic traffic model is a way of generating the six matrices described in \Section{representations}.
-A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or conditional form.
-For experimental comparison, the dimensions of the matrices should all be preserved.
-If a trace-derived model produces experimental performance metrics that are similar to those produced by the original trace data, then simplifications made by the model are viable; if the performance is distorted, then the simplifications are discarding essential information about the traffic pattern.
-
-%\subsection{Uniformity}
-
-%% TODO: differentiate between restrictive versions of models and transformative versions.
-%% TODO: distinguish uniform, marginal and conditional models.
-
-%\subsection{Marginality}
-
-%Most approaches to traffic modeling that go beyond uniformity assumptions, instead describe marginal properties of various aspects of traffic behavior.
-For example, {\FHC}~\cite{Hernandez06:spatio-temporal-modeling} provide highly convincing parametric models for six marginal properties of wireless network traffic: session (node) arrivals follow a time-varying Poisson process; per-session flow inter-arrival times are Lognormally distributed; flows per session (node) follow a BiPareto distribution; flow sizes follow a BiPareto distribution.
-This is certainly much greater detail than simply assuming that all nodes and flows are identical.
-
-%How do marginal models translate into the algebraic language of the general matrix model? Whereas uniformity assumptions for nodes,
-
-%Another benefit of the matrix model is that all aspects of network behavior are determined concurrently.
-We are not left wondering what to do about source and destination nodes or flow start and end times after we have computed our simplified packet behaviors: the same transformation that average the packet behaviors also naturally averages the start and
-
-\begin{table*}
-\begin{center}
-\small
-\begin{tabular}{|c|c|l|}
-\multicolumn{1}{c}{\textbf{model}} &
-\multicolumn{1}{c}{\textbf{transformation}} &
-\multicolumn{1}{c}{\textbf{description}} \\
-\hline
-\caps{GMM} & --- & the general matrix model \\\hline\hline
-time uniform & $\mat{A}'=\mat{A}(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
-node uniform & $\mat{S}'=\mat{S}(\ones{n \times n}/n)$~~~~$\mat{D}'=\mat{D}(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
-full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above models \\\hline\hline
-time marginal & $\mat{A}'=(\ones{f \times f}/f)\mat{A}$ & start times chosen from marginal distribution of start times \\\hline
-size marginal & $\mat{B}'=(\ones{f \times f}/f)\mat{B}$ & flow size (bytes) is chosen independently of other properties \\\hline
-node marginal & $\mat{S}'=(\ones{f \times f}/f)\mat{S}$~~~~$\mat{D}'=(\ones{f \times f}/f)\mat{D}$ & nodes behave distinctly, but behavior is independent of flows \\\hline
-packet marginal & $\mat{Z}'=(\ones{f \times f}/f)\mat{Z}$~~~~$\mat{V}'=(\ones{f \times f}/f)\mat{V}$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
-full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline\hline
-source conditional & $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$ & flow behaviors are aggregated on a per source-node basis \\\hline
-dest conditional & $\mat{F}'=(\trans{\mat{DD}}/f)\mat{F}$ & flow behaviors are aggregated on a per destination-node basis \\\hline
-mixed conditional & $\mat{F}'=\frac{1}{2f}(\trans{\mat{SS}}+\trans{\mat{DD}})\mat{F}$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
-\end{tabular}
-\caption{Matrix-based traffic models evaluated by paired differential simulation.}
-\label{tab:traffic-models}
-\end{center}
-\vspace{-2em}
-\end{table*}
-
-\section{Experimental Methodology}
-\label{sec:methodology}
-
-To evaluate whether traffic models are realistic or not, we use the method of \textit{paired differential simulation} introduced in our previous work on traffic modeling~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-%This approach is based on the following definition:
-%
-%\begin{samepage}
-%\noindent\textbf{Definition.}~A model of user behavior is \textit{sufficiently realistic} if, compared with actual user behavior, the model, with parameter values extracted from the real data, yields statistically equivalent performance results.
-%\end{samepage}
-%\vspace{0.5em}
-%
-Paired differential simulation is based on the standard scientific technique of controlled experimentation. 
-The hypothesis we are testing is that a given synthetic model accurately reproduces the performance metrics exhibited by real network traffic.
-To test this hypothesis, we conduct a series of paired experiments and compare the results for each test subject with a corresponding control subject.
-In this case, the experiments are wireless network simulations, and the subjects are real or synthetic workload instances.
-The control group is the set of simulations where the workload is an actual traffic trace, recorded as described in \Section{trace-data}.
-The test group is a set of simulations where workload is synthetically generated using a simplified traffic model.
-The experiments are paired: for each control simulation, there is a synthetic workload, with behavior as similar to the control as model will allow, against which the results are compared.
-The output of the series of experiments are paired values of performance metrics for each simulated scenario: one from the control, one from the test model.
-
-The remainder of this section proceeds as follows.
-In \Section{traffic-models} we present the collection of traffic models we will evaluate.
-In \Section{trace-data} we describe the trace data that serves as the control for our experiments.
-Our simulation methodology is detailed in \Section{simulations}, and the performance metrics used for statistical comparison of control and test experiments are described in \Section{performance-metrics}.
-% Finally, \Section{measures-of-error} is an in-depth mathematical discussion of the appropriate measure of error for the pairs of metric values that are produced by the paired differential simulation methodology.
-
-% to compare traffic produced by synthetic models with known examples of recorded trace traffic in a variety of experimental scenarios.
-To accomplish this, we begin with a large 24-hour example of wireless trace traffic, break it into 144 ten-minute scenarios, and use each scenario's simulated trace behavior as a control for a set synthetic traces, generated by various traffic models.
-For each scenario, the synthetic models are configured to approximate the original trace behavior as closely as the models allow.
-Important network-wide performance metrics are then collected from simulations, and synthetic performance results compared with control results to determine the realism of the synthetic models.
-
-%The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results.
-Clearly, we need not simulate subatomic particle interactions in a wireless network simulator.
-Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same.
-Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior.
-This requirement leads us to the following definition:
-
-%This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required.
-We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.
-
-\subsection{Traffic Models}
-\label{sec:traffic-models}
-
-The first model we evaluate is the general matrix model itself.
-We derive the matrix representation directly from the trace data and use it to generate workload as described in \Section{traffic-generation}.
-This serves to test whether the \caps{GMM} is sufficiently general: if it accurately reproduces all performance metrics across the collection of scenarios simulated, this provides strong evidence that \caps{GMM} captures enough of the detail of trace behavior to be considered a sufficiently general representation of network traffic patterns.
-%An affirmation of the generality of the \caps{GMM} acts as reduction of the
-
-After validating the \caps{GMM}, we explore the effects of various uniformity, marginality, and conditioning assumptions on the realism of traffic models.
-The uniform models are the most common, being the simplest both to conceptualize and to implement: simply make no distinction between start times, source or destination nodes.
-Choose them at random, using no information from reality other than the number of nodes and the possible range of start times.
-We do not evaluate uniform packet behavior because it has been sufficiently discredited in our previous work.
-
-In more sophisticated modeling work, where uniformity is discarded as too simple a model for behavior, marginality is often applied instead.
-It is assumed---often implicitly---that marginal models are adequate to capture the essential aspects of network behavior.
-We put that implicit assumption to the test here.
-How well do perfect marginal models represent real behavior?
-When we say that our marginal models are ``perfect,'' we mean that they are nonparametric and calculated directly from trace behavior.
-These are the marginal models that most accurately describe the trace traffic, because they are directly derived from it and make no parametric simplifications.
-This provides an upper bound on the quality of marginal models in general, since parametric marginal models are derived, in turn, as simplifications of these nonparametric marginal behaviors.
-\Table{traffic-models} provides a complete list of all the traffic models we compare in our evaluation.
-
-%The original trace traffic serves as a control group against which to compare all other models.
-Before we can evaluate the general matrix model (\caps{GMM}), we must establish that one potentially important aspect of traffic behavior that is \textit{not} represented in the \caps{GMM} can legitimately be discarded.
-That aspect of behavior is \textit{\caps{TCP} feedback}.
-The \caps{GMM} does not specify the \caps{IP} protocol type of each flow.
-Therefore we must establish the extent to which using \caps{UDP} flows, without reliable delivery or congestion control, affects performance metrics.
-%%\caps{GMM} also does not indicate which flows are paired with each other in a client-server relationship. While this pairing does not affect non-\caps{TCP} flows, we implement.
-%Other synthetic models derived from algebraic transformations of the \caps{GMM} are listed in \Table{traffic-models}.
-
-\subsection{Trace Data}\label{sec:trace-data}
-%\subsection{Trace Data \& Simulations}\label{sec:trace-data}\label{sec:simulations}
-
-Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of trace-based models and synthetic traffic models defined using the general matrix model.
-We use a 24-hour trace recorded in an infrastructured 802.11g wireless \caps{LAN} with 18 access points, deployed at the 60th Internet Engineering Task Force meeting (\caps{IETF}60), held in San Diego during August of 2004.
-The traffic trace was captured using \texttt{\small{tcpdump}} at a single router, through which all wireless traffic for the meeting was routed, including traffic between wireless nodes.
-The snap length of the capture was 100 bytes, allowing \caps{IP}, \caps{ICMP}, \caps{UDP} and \caps{TCP} headers to be analyzed.
-We limit our work to the 24-hour sub-trace recorded on Wednesday, August 4th.
-This trace contains a broad variety of behaviors and entails a very large volume of traffic: 2.1 million flows, 58 million packets, and 52 billion bytes.
-% 2,085,563 flows, 58,377,994 packets of application data, and 51,988,245,537 bytes of application data.
-
-\begin{figure}[!b]
-\begin{center}
-\includegraphics[width=3.3in]{nodes-flows}%
-\vspace{-1.25em}%
-\caption{The number of active nodes and flows over time.} 
-\label{fig:nodes-flows}
-\end{center}
-\vspace{-1.3em}
-\end{figure}
-
-We do not assume or claim that the traffic found at \caps{IETF}60 is representative of conference settings in general.
-The observed behaviors are also unlikely to resemble those found in a typical commercial or residential setting.
-We have chosen this trace, however, because within it can be found behaviors resembling many different types of wireless usage cases.
-\Figure{nodes-flows} shows the wide variations in the number of active flows and nodes over the course of the trace.
-In the night and morning hours, the traffic patterns are similar to those one might find in a moderately trafficked business or residential area.
-During working group sessions, we see highly concentrated, heavy usage patterns.
-At the zenith of activity, over 800 users, 33 thousand flows, and 1 million packets are seen in a single 10-minute trace segment.
-At the nadir, a lone node sent only a single 61-byte packet over the course of 10 minutes.
-All levels of activity between these extremes are represented.
-Moreover, the mix of traffic types observed changes dramatically over the course of the day, providing a wide representation of possible blends of behavior.
-This heterogeneity and extreme range of behaviors makes the \caps{IETF} data set ideal for this evaluation.
-The variety of activity gives us greater confidence that success or failure of traffic models is not tied to any specific network condition, but is broadly and generally applicable.
-%For a traffic model to be truly realistic, it must be realistic across many different usage cases.
-
-Before using the traces, it is necessary to extract application-level behavior from the trace header data.
-First, we split the trace into individual packet flows.
-A flow is a series of packets sharing the following five attributes: \caps{IP} and transport protocols (raw \caps{IP}, \caps{ICMP}, \caps{TCP}, \caps{UDP}); source and destination \caps{IP} addresses and \caps{TCP}/\caps{UDP} port numbers.
-Next, the quantity of application-initiated data contained in each packet is calculated.
-For non-\caps{TCP} packets, this quantity is simply the size of the transport-layer payload, but for \caps{TCP} the calculation is more complicated: only new data transfers, explicitly initiated by the application are counted.
-Data retransmitted by \caps{TCP} is disregarded, and empty \caps{ACK}s are ignored.
-\caps{SYN} and \caps{FIN} flags in packets (even empty ones) are counted as a single byte each, since they are explicitly signaled by the application.
-%After processing, we have a collection of flows, each with a sequence of timestamps and the amount of application-initiated data sent at that time.
-
-\subsection{Simulations}
-\label{sec:simulations}
-
-We use the Qualnet wireless network simulator (version 4.0.1) to perform our experiments.
-We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
-In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays.
-Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings.
-We do not attempt to reproduce the physical environment of the original wireless network, nor do we simulate mobility.
-The only aspect of the original network's behavior that is reproduced is the total pattern of network-wide traffic.
-
-There are a number of potential objections to this approach.
-We use single-hop trace data to drive multi-hop simulations; the physical environment, node mobility, handover behavior, and closed-loop dynamics of the original wireless setting are not faithfully reproduced.
-One must keep in mind, however, that the goal of this research is \textit{not} to understand the conditions of the original network.
-Rather, we are using the traffic behaviors observed as examples to help us better understand how different types of workload can affect performance metrics.
-In particular, we aim to understand how real workload compares with common synthetic traffic models.
-Of course, the reason for such objections is that networking researchers understand that the many aspects of behavior interact with each other in a complex and nearly inextricable manner.
-However, before we can hope to understand the interaction between workload and other features affecting network behavior, we must study traffic patterns alone, and learn to model them with reasonable accuracy in the absence of additional complicating factors.
-Accordingly, in this study, we detach application level traffic patterns from the other factors influencing network conditions, and study them in isolation.
-
-The 24-hour trace is split into 144 10-minute segments, each of which serves as the basis for a set of simulations using different traffic models.
-The traffic models range from a completely realistic trace-driven model, to a standard \caps{CBR} traffic model.
-Various partially synthetic intermediate models, described in \Section{traffic-models}, are simulated to study the impact of different aspects of traffic behavior on network performance.
-To preserve the fairness of the performance comparison, we keep as many features as possible constant across different traffic models.
-The traffic generated by each synthetic model preserves as many characteristics from the original trace as possible, within the constraints of the model.
-Moreover, the following features are preserved across all models: the number of wireless nodes, the number of flows, the number of application-initiated data units sent, the total bytes of application data sent, and the average flow duration (and therefore the average data rate).
-
-In our previous work, we approximated \caps{TCP} with a pair of half-duplex \caps{UDP} flows~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-For this work, we have implemented a new Qualnet application driver that allows trace-driven full-duplex \caps{TCP} flows.
-This allows us to accurately reproduce the full dynamics of \caps{TCP} feedback.
-Moreover, raw \caps{IP}, \caps{UDP} and \caps{TCP} flows are implemented using the same code, guaranteeing that all traffic is simulated uniformly and that all performance metrics are aggregated in the same manner.
-We have also instrumented Qualnet to collect \caps{IP} performance metrics in strict adherence to the recommendations of the \caps{IETF} working group on \caps{IP} performance metrics~\cite{rfc:ip-metrics}.
-
-\subsection{Performance Metrics}
-\label{sec:performance-metrics}
-
-We have selected six performance metrics to present here. They are commonly used as indicators of network performance at the application, network, and link layers of the protocol stack:
-\begin{enumerate}
-\setlength{\itemsep}{0em}
-\item \textbf{Application:} average end-to-end delay, packet delivery ratio, received throughput.
-\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/\\RERR}), packets dropped in routing queues.
-\item \textbf{Link:} 802.11 control overhead (\caps{RTS/CTS/ACK}).
-\end{enumerate}
-These metrics are commonly used to evaluate wireless protocols.
-We have examined a broad variety of other wireless network performance metrics, and although we do not have room to present or discuss them here, the results shown are representative of the overall realism of the traffic models.
-
-\section{Results}
-\label{sec:results}
+%!TEX root = ../paper.tex
 
@@
 The lower axis indicates the log-ratio, while the upper axis shows raw ratio values.
-Each box contains the central majority of log-ratio values: the left and right bounds are at the $25^\text{th}$ and $75^\text{th}$ percentiles.
+Each box contains the central majority of log-ratio values:
+the left and right bounds are at the $25^\text{th}$ and $75^\text{th}$ percentiles.
 The dark middle line indicates the median value, while the diamond marks the mean.
 The whiskers (dotted lines) extend to the furthest non-outlier values, while the points beyond that are outliers.
-The notch in the middle of each bar indicates a $95\%$ confidence interval for the true underlying median value; if two notches do not overlap, they are very unlikely to have the same median.
+The notch in the middle of each bar indicates a $95\%$ confidence interval for the true underlying median value;
+if two notches do not overlap, they are very unlikely to have the same median.
 }
@@
 \end{sidewaysfigure*}
-
-Our simulation results are summarized in \Figure{box-plots}.
-Each subfigure shows a single performance metric.
-The distribution of log-ratio error values for each traffic model is visualized with a box-and-whisker plot.
-The box indicates the range from the 25th to 75th percentiles of values, while the ``whiskers'' indicate the full range, excluding outliners (which are shown as isolated points).
-These plots allow immediate assessment of realism: a good traffic model should have error values that are tightly clustered around the center, with a small, evenly balanced box, and relatively small whiskers.
-Additionally, the mean and median markers should be close to the center.
-It should be noted that the error scale is logarithmic, so even a slight increase in spread or deviation from the center indicates a disproportionately large decrease in model quality.
-Non-logarithmic scale markers are shown above each plot; the values indicate the factor of over- or under-representation for the performance metric.
-
-The primary result is that the \caps{GMM} accurately reflects the performance of the trace data it is based on.
-It has small, centered error bars for every performance metric presented here, as well as for those we have looked at otherwise.
-This is an essential result, since otherwise, \caps{GMM} is at best a shaky foundation for further modeling work.
-With these results, we can be assured that if we can approximate the matrix for a given example of trace traffic, then we have also approximated the original trace behavior.
-In a sense, this result is a reduction of the general traffic modeling problem to a more tractable matrix modeling problem.
-
-It is worth noting that \caps{GMM} has competition from a few of the time-simplified models: the \model{time marginal} and \model{time uniform} model both do approximately as well, and in some cases better, for the metrics examined.
-{\FHC} found strong evidence that session arrivals followed a time-varying Poisson arrival process, meaning that other than long-scale time-varying arrival rate and very small-scale clustering of flows within sessions, the overall flow arrival rate is fairly smooth.
-We suspect that on the scale of minutes to tens of minutes in which our simulation exist, it is sufficient to model both flow and session arrivals uniformly.
-This result indicates that the precise temporal placement of flows is not highly sensitive: they tend to be relatively evenly spaced out at short time-scales, and network performance is not sensitive to changes in start-time on that scale.
-
-Not all marginal models have such a benign effect on the accuracy of performance metrics.
-The worst offender by far is the \model{size marginal} model.
-This model distorts behavior at every level, by more than a factor 20 in 25\% of scenarios in the case of routing queue packet drops.
-This is a highly significant result because modeling flow size using a network-wide marginal distribution is precisely what {\FHC}, for example, attempt.
-This results informs us that this approach is doomed to failure: flow size cannot be assigned without considering its relation to other flow properties: source and destination nodes, and packet behavior at the very least.
-
-The \model{packet marginal} model is another offender, albeit not as egregiously.
-The direction of the misrepresentations in this case, however, is more dangerous: received throughput is overestimated by more than three times, 75\% of the time.
-On the other hand, \caps{AODV} control overhead is underestimated by half on average.
-Such serious misrepresentations are especially noteworthy given the apparently innocuous assumption applied by this model: the \model{packet marginal} model is essentially a standard variable bit-rate (\caps{VBR}) packet behavior model, using the empirical network-wide distributions of packet sizes and inter-packet intervals.
-Otherwise the model matches the original trace behavior exactly.
-And yet these drastic distortions occur.
-In our previous work, we have shown similarly drastic misrepresentations to occur when constant bit-rate packet behavior is assumed.
-This results indicates that to reproduce accurate network performance, we must provide packet behaviors on a more granular level: each node or even each flow must be assigned a ``custom'' packet behavior by some means.
-
-These strong negative results for two fundamental and very common marginal models provide severe limits on the realism that can be achieved through marginal modeling.
-Traffic modeling research \textit{must} move beyond simply finding parametric models for marginal distributions of network-wide properties.
-Realistic behavior cannot be generated using these distributions alone.
-Something must be captured about the interaction between the elements of flow behavior.
-
-Finally, we turn to the conditional models.
-These demonstrate neither exceptionally good nor exceptionally bad behavior.
-That the conditional models should be similar to the \model{full marginal} model but somewhat more realistic is entirely expected: conditioning, the reader will recall, is a restricted form of marginal modeling, where each group of flows has its own marginal distribution.
-In these cases, the flows are grouped by source node or by destination node.
-
-% The mixed case uses the average behavior of the other two cases.
-It is interesting to note that the mixed case is universally worse than either the source or destination case.
-Clearly, hybridizing the conditioning here is not an effective technique.
-We believe that the cause of this effect is like mixing too many colors yielding an unpleasant shade of brown: any of the colors individually would be better than the nondescript average.
-As a more general conclusion, arbitrarily mixing model may not always be a good idea, despite the ease with which it is done in the matrix representation of workload behavior.
-
-%We also believe that the fact that both the \model{time marginal} and \model{time uniform} models perform about as well as the \caps{GMM} indicates not that they are performing exceptionally well, but that the \caps{GMM} is performing worse than it should in this case.
-This may be an artifact of the interaction between packet behavior, flow size, flow duration and start time, as described in \Section{traffic-generation}.
-This remark may appear to be a non-sequitur, but these properties are all closely related.
-Consider that packet behavior (i.e.
-distribution of packet sizes and inter-packet intervals) determines the expected mean packet size and expected mean interval duration.
-The expected duration is then related to the flow size via the following equation:
-%\begin{align*}
-%\E{\text{duration}} = [\text{flow size}]\frac{\E{\text{mean interval}}}{\E{\text{mean size}}}
-%\end{align*}
-%The duration must also satisfy $t_{\text{start}}+[\text{duration}] \le t_{\max}$.
-
-%It accurately reproduces important performance metrics at all levels of the protocol stack in simulation.
-This only moves us a step closer to being able to generate realistic, completely synthetic workloads, since the \caps{GMM} in these comparisons  derives almost all of its behavior directly from the trace.
-However, it means that we have effectively reduced the problem of generating realistic workload to the problem of classifying and understanding what sorts of traffic matrices occur in real workloads.
-Moreover, we can express a vast array of different approaches to traffic modeling simply and uniformly using matrix operations.
-
-\section{Discussion}
-\label{sec:discussion}
-
-%The story of network analysis and modeling is one of variety. Assumptions of ``sameness'' have fallen repeatedly. ...
-%The matrix model gives new insight into
-
-Thus far, we have addressed two out of three of the originally posed questions regarding traffic modeling.
-First, we have proposed a general approach for representing whole-network traffic without making assumptions of uniformity or marginality.
-To this end, we introduced the concept of a linear representation of traffic.
-The general matrix model entails a specific linear representation, such that network traffic patterns can easily be expressed as matrices.
-Our experimental results show that the \caps{GMM} is sufficiently realistic for generating wireless workload derived from traces.
-When uniform or marginal assumptions are applied, however, sufficient realism eludes us, indicating that if we are to simplify and reduce the representation of network behaviors, we must do some through some other technique.
-In \Section{modeling-concepts}, we observed that application of uniformity and marginality assumptions to traffic instances is equivalent, in the \caps{GMM}, to right and left multiplication, respectively, by matrices of rank one.
-These transformations destroy almost all of the information in the original traffic matrix---only the zero matrix has lower rank than these.
-
-The question that remains to be answered is how we can simplify and reduce the representation of network behaviors without making unwarranted and intuitively clearly false assumptions of uniformity of marginality of network behaviors.
-We present here preliminary results from applying non-negative matrix factorization to the packet behaviors of collections of flows.
-Recall from \Section{general-matrix-model}, that the matrices $Z$ and $V$ represent the packet size and inter-packet interval histograms of a collection of flows, with each row corresponding to a flow, and each column corresponding to a packet size or a range of inter-packet intervals.
-We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}]$.
-\caps{NMF} attempts to approximate $\mat{P}$ as the product of three non-negative matrices:
-\begin{align}
-\mat{P} \approx \mat{W}\mat{\Sigma}\mat{H},
-\end{align}
-where $\mat{W} \in \R^{f \times K}$ is row-stochastic,
-$\mat{\Sigma} \in \R^{K \times K}$ is diagonal with descending diagonal entries, and
-$\mat{H} \in \R^{K \times (d_z+d_v)}$ is row-stochastic as well.
-$K\in\N$ is the rank of this factorization, which must be determined somehow as well.
-What is the interpretation of such factorization?
-The rows of $\mat{H}$ are \emph{basic behaviors}---pairs of packet size and inter-packet interval distributions, such that each flow's packet behavior can be approximated as a linear combination of these basic behaviors.
-Specifically, the matrix $\mat{M}\mat{\Sigma}$ gives the mixing coefficients for each flow.
-The diagonal entries of $\mat{\Sigma}$ have a special meaning as well:
-they are the number of packets associated with each basic behavior.
-Thus the basic behaviors are sorted in descending order of how much traffic they explain.
-Note that the factorization does not associate basic behaviors with individual packets;
-rather, total flow behavior is explained as a mixture of basic behaviors, so the fraction of packets in each flow explained by a given behavior is known, while the specific packets are unknown.
-
-The striking results of applying non-negative matrix factorization to $\mat{P}$ are shown in \Figure{basic-behaviors}.
-Twelve basic behaviors suffice to explain all the traffic in a ``toy''  sample of 4138 flows.
-The pairs of packet size and inter-packet interval distributions are almost immediately recognizable as resembling intuitive modes of network behavior.
-The first basic behavior, for example has all packets near the \caps{MTU}, while intervals are smoothly distributed around a fraction of a millisecond.
-This is what we would expect from downloading large files; and indeed, the breakdown of packets by port number bears out this interpretation---almost all of the traffic is \caps{HTTP} server traffic, with small slivers of \caps{IMAP} and \caps{SSH} server traffic.
-The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all \caps{SSH} traffic, interacting with a remote terminal via keystrokes and screen updates.
-
-\begin{sidewaysfigure*}
-{\includegraphics[height=5.95in]{basic_behaviors_1.pdf}}\hfill%
-{\includegraphics[height=5.95in]{basic_behaviors_2.pdf}}
-\caption{%
-The ten most prevalent components of flow behavior as pairs of packet payload size and of inter-packet interval distributions.
-The distributions are shown as cumulative distribution functions: the $x$-axis represents payload size (in kilobytes) or inter-packet interval duration (in log-seconds, base 10), respectively;
-the $y$-axis indicates the probability a value occurring, less than or equal to the $x$ value.
-The behaviors are ordered in descending rank of how many packets they explain;
-to the left, each behavior's rank is indicated, together with the percentage of total packets associated with it, and the cumulative percentage of packets explained.
-To the right of each behavior is a pie chart showing the breakdown of associated packets by protocol, as determined by well-known port numbers.
-}
-\label{fig:basic-behaviors}
-\end{sidewaysfigure*}
-
-% What are the ramifications of these results?
-In our previous work, we had already demonstrated the drastic distortions that are introduced by uniformity assumptions made by the most common traffic models.
-This work serves to further critique established common assumptions in traffic modeling.
-The strongest message to be taken away is that the aspects of traffic behavior of both nodes and flows \textit{must} be considered jointly.
-It is not sufficient to continue modeling marginal distributions of network properties.
-These marginal models do not lend them selves easily to traffic generation in the first place, and when marginality assumptions are applied to trace traffic, the result is a severe distortion of important network performance characteristics.
-It is now well established that network traffic patterns have an impact on network performance that cannot be ignored.
-Even experimental deployments cannot avoid the need for more realistic traffic workload models.
-While using a real, physical network successfully sidesteps simulation problems below the application layer, without realistic traffic models, reliable, meaningful performance predictions remain beyond our reach.
-
-%\subsection{Generality of Results}
-
-%% TODO: talk about other synthetic models not considered here.
-%% The point of our analysis is not to show that no synthetic models can be accurate, but rather, that the commonly used models, and obvious variations thereof, are not accurate.
-
-%The most significant limitation on the generality of this analysis is that it is based entirely on a single data set from \caps{IETF}60---albeit a large and varied one.
-It is possible that traffic in this trace happens to produce network performance that is unusually dissimilar to standard traffic models.
-This data set, however, represents a highly heterogeneous collection of network usage behaviors, from slow and steady off-peak usage, to extremely heavy peak usage: over 800 users, 33 thousand flows, and 1 million packets in a single 10-minute trace segment.
-Despite the broad variety of behaviors, the results are consistent: in all types of usage scenarios, simplistic traffic models, like uniform \caps{CBR}, systematically skew important performance measurements at all levels of the network.
-While the precise results for other data sets might differ, it is very unlikely that \caps{CBR} traffic models will happen to accurately reproduce realistic performance in other experiments.
-This paper provides strong evidence that better traffic models are needed for performance evaluations.
-
-% \subsection{Towards Realistic Models of Wireless Workload}
-% 
-% What would better traffic models look like?
-How can we create them?
-One possible approach is to use actual traffic traces as we have done.
-This approach is unsatisfactory, however, because it provides the experimenter with almost no control over experiments.
-Synthetic models have parameters, which can be tweaked as necessary---adjusting, for example, the number of active nodes in a simulation, without affecting other parameters.
-Traffic traces, on the other hand, must be used without significant alteration if they are to actually provide the desired realism.
-This inability to control all but one or two parameters is nearly crippling for experimental purposes.
-Therefore, the research community needs models whose parameters can be adjusted independently, but whose performance characteristics can be made to resemble whatever example of network behavior one might need.
-The general matrix model does not yet fulfill this role since its parameters cannot be freely varied.
-We believe, however, that \caps{GMM} plays a vital role in the ongoing process of developing  of better models.
-By reducing the general problem of traffic modeling to the domain of linear algebra, we have provided traffic researchers with a vast array of powerful theoretical and computational tools that can be immediately applied to the problem at hand.\footnote{The derived models used in this paper were all implemented in Matlab, each in only a few simple lines.
-All the models used here are implemented in only 19 total lines of code.
-Of course, the framework to apply them to traffic traces and simulations is somewhat larger.} What are the next steps?
-The authors are currently investigating a variety of matrix factorization and clustering techniques to extract hidden structure from a large and diverse body of real traffic.
-
-%The reader will note that the transformations applied in this paper are all quite simple.
-Our conditioning matrices primarily consisted of ones and zeros.
-The algebraic representation of complete workload
-
-%The ``messiness'' of the performance comparison from trace data in Figure~\ref{fig:mac-control-overhead-cmp} illustrates why using traces directly is not ideal: each data point differs not only in the number of nodes shown on the $x$-axis, but also in other dimensions, such as the number of flows and packets, and the average flow duration.
-The result is a highly noisy comparison, affected by many unseen parameters.
-Only by applying a local smoothing algorithm are trends somewhat elucidated.
-
-%Instead of using trace data directly, it should be possible to configure a synthetic traffic model based on observations from a real data set, and then run side-by-side simulations using the synthetic model and the real data, producing statistically equivalent performance results.
-This is precisely what our definition of sufficient realism entails.
-The work in this paper provides the tools to measure how close to this ideal a model is and in what areas it needs improvement.
-Without this feedback, any improvements in realism are purely guesswork.
-Our breakdown of traffic behavior into three orthogonal levels also allows the problem to be approached in smaller pieces, rather than being solved all at once.
-
-%The next step towards better traffic models is to investigate which aspects of real traces may be altered without detrimentally affecting the resulting performance metrics.
-For example, to test whether a complex time-series model of packet behavior is necessary, we randomize the order of the packet sizes and/or inter-packet intervals and compare performance using these randomized traces against performance using the original traces.
-If the performance is unchanged, we can conclude that no complex time-series model of packet behavior is necessary: sampling the packet sizes and inter-packet intervals from empirical distributions is sufficiently realistic.
-If, on the other hand, the performance characteristics are altered by shuffling packets, then some time-series model of packet behavior is needed.
-By partially randomizing the packet order in specific ways, the exact limits of realism necessary can be found.
-A similar approach will allow the development of realistic models for the other levels of network usage behavior.
-
-\section{Conclusions}\label{sec:conclusions}
-
-%The real significance of this paper lies not in showing that the commonly used \caps{CBR} traffic model is unrealistic.
-Networking researchers understand that \caps{CBR} is at best a na\"ive approximation of real network traffic.
-Nor is the real significance that we have quantized the high degree of inaccuracy that can be induced by using this model.
-The real significance of this paper lies in providing an well-defined, objective measure of realism for traffic models.
-This has never previously existed: evaluations of realism have relied on essentially arbitrary statistical measures of similarity to real traffic.
-Existing traffic generators use empirical distributions of such quantities as file size, inter-connection time, packet size, or inter-packet intervals.
-They can only hope that reproducing these distributions as good enough.
-Without an external measure of accuracy, there's simply no way to know.
-
-%attempting to showing that existing models are unrealistic, or that new models are more realistic have always focused on arbitrary statistical measures. A
-
-We have presented a fundamentally new, algebraic way of representing traffic patterns in local area networks.
-We call this representation the general matrix model.
-Each flow of a traffic collection is represented as a single high-dimensional vector with components representing the \caps{IP} protocol type, source and destination nodes, start time, flow size, and packet behavior.
-The essential property of the algebraic representation is that standard vector operations compute natural and useful descriptions of aggregate behavior over the collection of flows represented.
-Our experimental validation demonstrates clearly that the general matrix model accurately reproduces the performance characteristics of of real traffic.
-The benefits of a performance-preserving algebraic representation are multifold.
-The algebraic forms of the simplifying assumptions made by many common modeling approaches provide unprecedented clarity and coherence to a complex and confusing subject.
-For example, assumptions that various aspects of flow behavior are stochastically or deterministically uniform all take the same form algebraically: left matrix multiplication.
-The other major class of common simplifying assumptions made by traffic models is to apply some marginal behavior across the entire collection of flows or to disjoint subsets of the flows.
-Such simplifications correspond to right matrix multiplication in \caps{GMM}.
-Because of the simplicity and clarity that the algebraic structure brings to the subject, we can see clearly now that both common types of simplifying assumptions are na\"ive and simplistic.
-Our experimental results bolster this conclusion: both uniform and marginal models significantly misrepresent many important network performance metrics.
-With the general matrix model, however, we stand poised to use the powerful tools of modern linear algebra to develop general, elegant and powerful models of network behavior that just work.
-
-%This research rigorously quantifies the impact of a variety of synthetic traffic models on performance metrics that wireless researchers use to evaluate new technologies and protocols.
-The first step in this assessment process was to formally define what it means for a network usage model to be sufficiently realistic.
-In essence, a model is considered realistic if it produces performance results that are statistically equivalent to those produced by real usage.
-%A well-defined, objective measure of realism for traffic models has not previously existed.
-Evaluations of realism have formerly relied on essentially arbitrary statistical measures of similarity to real traffic, which may or may not affect the performance metrics that researchers care about.
-%The definition of sufficient realism leads us to our general experimental approach: we use differential analysis comparing performance metrics derived from real traffic with those derived from synthetic traffic models.
-The theoretical contributions of this analysis are:
-%\begin{enumerate}
-%\item An in-depth analysis of the desirable mathematical properties of a measure of error for performance metrics.
-%\item Proof that the unique measure of error that satisfies these properties is the log-ratio of metric values.
-%\item Three rigorous tests of statistical equivalence between synthetic and real performance results.
-%\end{enumerate}
-%These analytical tools allow the evaluation of realism over a collection of drastically different usage scenarios.
-Evaluation over a heterogeneous collection of scenarios is essential to establishing the credibility of usage models.
-Moreover, these theoretical results are equally applicable to other types of usage models---for example, mobility.
-
-%On the practical side, this paper gives crucial insight into why most researchers do not trust simulation results: with the traffic models commonly used, the results are unlikely to reflect real performance.
-Moreover, this problem will also hamper experiments in test-bed wireless networks, so long as the same na\"ive workload models are used.
-The only way to address this fundamental lack of realism is to develop usage models that reproduce important performance metrics more accurately.
-Our theoretical results provide the tools necessary to do this.
-The development of better traffic models should begin with real traces, and proceed by incremental changes, checked by differential analysis.
-%%First, alter a small aspect of the trace, simulate, then compare.
-If the realism of the results is unaffected, the traffic feature altered was inessential.
-Otherwise, it is a feature of behavior that must be captured in a realistic traffic model.
-%This approach will allow the precise mapping of which aspects of traffic patterns have an impact on performance and which ones can be safely abstracted away.
-%% TODO: maybe cut out some of this last bit.
-
-%\section{Acknowledgments}
-%This work was funded in part through NSF Career Award CNS-0347886.
-
-% Hernandez06:dissertation: ``complete methodology for reproducing the traffic observed on a network link in a closed-loop manner, and proposed a number of metrics for studying the realism of the generated traffic.''
-\bibliographystyle{plain}
-\bibliography{references}
-
-\end{document}

### 2008-06-12 paper.tex
Moved all tables & figures into separate files.

@@
 
-\begin{table*}
-\begin{center}
-\small
-\begin{tabular}{|c|c|l|}
-\multicolumn{1}{c}{\textbf{model}} &
-\multicolumn{1}{c}{\textbf{transformation}} &
-\multicolumn{1}{c}{\textbf{description}} \\
-\hline
-\caps{GMM} & --- & the general matrix model \\\hline\hline
-time uniform & $\mat{A}'=\mat{A}(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
-node uniform & $\mat{S}'=\mat{S}(\ones{n \times n}/n)$~~~~$\mat{D}'=\mat{D}(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
-full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above models \\\hline\hline
-time marginal & $\mat{A}'=(\ones{f \times f}/f)\mat{A}$ & start times chosen from marginal distribution of start times \\\hline
-size marginal & $\mat{B}'=(\ones{f \times f}/f)\mat{B}$ & flow size (bytes) is chosen independently of other properties \\\hline
-node marginal & $\mat{S}'=(\ones{f \times f}/f)\mat{S}$~~~~$\mat{D}'=(\ones{f \times f}/f)\mat{D}$ & nodes behave distinctly, but behavior is independent of flows \\\hline
-packet marginal & $\mat{Z}'=(\ones{f \times f}/f)\mat{Z}$~~~~$\mat{V}'=(\ones{f \times f}/f)\mat{V}$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
-full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline\hline
-source conditional & $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$ & flow behaviors are aggregated on a per source-node basis \\\hline
-dest conditional & $\mat{F}'=(\trans{\mat{DD}}/f)\mat{F}$ & flow behaviors are aggregated on a per destination-node basis \\\hline
-mixed conditional & $\mat{F}'=\frac{1}{2f}(\trans{\mat{SS}}+\trans{\mat{DD}})\mat{F}$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
-\end{tabular}
-\caption{Matrix-based traffic models evaluated by paired differential simulation.}
-\label{tab:traffic-models}
-\end{center}
-\vspace{-2em}
-\end{table*}
+\input{tables/traffic_models}
 
@@
 
-\begin{figure}[!b]
-\begin{center}
-\includegraphics[width=3.3in]{nodes-flows}%
-\vspace{-1.25em}%
-\caption{The number of active nodes and flows over time.} 
-\label{fig:nodes-flows}
-\end{center}
-\vspace{-1.3em}
-\end{figure}
+\input{figures/nodes_flows}
 
@@
 
-\begin{sidewaysfigure*}
-\subfloat[\textbf{Application:} Average End-to-End Delay]%
-{\includegraphics[width=4.45in]{Application/delay.pdf}}\hfill
-\subfloat[\textbf{Application:} Packet Delivery Ratio]%
-{\includegraphics[width=4.45in]{Application/delivery_ratio.pdf}}
-%
-\subfloat[\textbf{Application:} Average Received Throughput]%
-{\includegraphics[width=4.45in]{Application/throughput.pdf}}\hfill
-\subfloat[\textbf{Network:} AODV Control Overhead]%
-{\includegraphics[width=4.45in]{Network/control_packets.pdf}}
-%
-\subfloat[\textbf{Network:} Packets Dropped in Routing Queue]%
-{\includegraphics[width=4.45in]{Network/dropped.pdf}}\hfill
-\subfloat[\textbf{Link:} 802.11 Control Overhead]%
-{\includegraphics[width=4.45in]{Link/overhead.pdf}}%
-%
-\caption{%
-Box-and-whisker plots of log-ratio error values for all metrics and traffic models.
-The lower axis indicates the log-ratio, while the upper axis shows raw ratio values.
-Each box contains the central majority of log-ratio values: the left and right bounds are at the $25^\text{th}$ and $75^\text{th}$ percentiles.
-The dark middle line indicates the median value, while the diamond marks the mean.
-The whiskers (dotted lines) extend to the furthest non-outlier values, while the points beyond that are outliers.
-The notch in the middle of each bar indicates a $95\%$ confidence interval for the true underlying median value; if two notches do not overlap, they are very unlikely to have the same median.
-}
-\label{fig:box-plots}
-\end{sidewaysfigure*}
+\input{figures/box_plots}
 
@@
 
-\begin{sidewaysfigure*}
-{\includegraphics[height=5.95in]{basic_behaviors_1.pdf}}\hfill%
-{\includegraphics[height=5.95in]{basic_behaviors_2.pdf}}
-\caption{%
-The ten most prevalent components of flow behavior as pairs of packet payload size and of inter-packet interval distributions.
-The distributions are shown as cumulative distribution functions: the $x$-axis represents payload size (in kilobytes) or inter-packet interval duration (in log-seconds, base 10), respectively;
-the $y$-axis indicates the probability a value occurring, less than or equal to the $x$ value.
-The behaviors are ordered in descending rank of how many packets they explain;
-to the left, each behavior's rank is indicated, together with the percentage of total packets associated with it, and the cumulative percentage of packets explained.
-To the right of each behavior is a pie chart showing the breakdown of associated packets by protocol, as determined by well-known port numbers.
-}
-\label{fig:basic-behaviors}
-\end{sidewaysfigure*}
+\input{figures/basic_behaviors}
 

### 2008-06-12 tables/traffic_models.tex
Moved all tables & figures into separate files.

%!TEX root = ../paper.tex

\begin{table*}
\begin{center}
\small
\begin{tabular}{|c|c|l|}
\multicolumn{1}{c}{\textbf{model}} &
\multicolumn{1}{c}{\textbf{transformation}} &
\multicolumn{1}{c}{\textbf{description}} \\
\hline
\caps{GMM} & --- & the general matrix model \\\hline\hline
time uniform & $\mat{A}'=\mat{A}(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
node uniform & $\mat{S}'=\mat{S}(\ones{n \times n}/n)$~~~~$\mat{D}'=\mat{D}(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above models \\\hline\hline
time marginal & $\mat{A}'=(\ones{f \times f}/f)\mat{A}$ & start times chosen from marginal distribution of start times \\\hline
size marginal & $\mat{B}'=(\ones{f \times f}/f)\mat{B}$ & flow size (bytes) is chosen independently of other properties \\\hline
node marginal & $\mat{S}'=(\ones{f \times f}/f)\mat{S}$~~~~$\mat{D}'=(\ones{f \times f}/f)\mat{D}$ & nodes behave distinctly, but behavior is independent of flows \\\hline
packet marginal & $\mat{Z}'=(\ones{f \times f}/f)\mat{Z}$~~~~$\mat{V}'=(\ones{f \times f}/f)\mat{V}$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline\hline
source conditional & $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$ & flow behaviors are aggregated on a per source-node basis \\\hline
dest conditional & $\mat{F}'=(\trans{\mat{DD}}/f)\mat{F}$ & flow behaviors are aggregated on a per destination-node basis \\\hline
mixed conditional & $\mat{F}'=\frac{1}{2f}(\trans{\mat{SS}}+\trans{\mat{DD}})\mat{F}$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
\end{tabular}
\caption{Matrix-based traffic models evaluated by paired differential simulation.}
\label{tab:traffic-models}
\end{center}
\vspace{-2em}
\end{table*}

### 2008-06-12 figures/trace_vs_uniform.tex
Massive structural reorganization.
Moved all tables & figures into separate files.
Clean input formatting for some tables & figures.

--- 2008-06-06 trace_vs_uniform.tex
@@
-%!TEX root = /Users/stefan/projects/papers/paper.4/paper.tex
+%!TEX root = ../paper.tex
+
 \begin{figure*}[!t]
@@
 {\includegraphics[width=1.335in]{flow_packet_behavior}\label{fig:packet-behavior}}
-\caption{Examples illustrating real versus uniform behaviors at the node, flow and packet levels.
+\caption{%
+Examples illustrating real versus uniform behaviors at the node, flow and packet levels.
 Figures \ref{fig:flow-topology-trace} and \ref{fig:flow-topology-uniform} show example node behaviors.
@@
 Uniform and trace flow behavior examples are plotted in Figures \ref{fig:flow-behavior-trace} and \ref{fig:flow-behavior-uniform}.
-The time axis indicates when flows start and end; the width of each flow line is proportional to the logarithm of its data rate.
-Figure \ref{fig:packet-behavior} compares uniform (i.e.
-\caps{CBR}) packet behavior with the trace of an actual flow.
-In the uniform model, the cumulative data sent increases smoothly over time, whereas in the actual packet trace, the transmissions are variable both in size and in inter-packet interval, leading to a ``lumpy'' cumulative data plot.}
+The time axis indicates when flows start and end;
+the width of each flow line is proportional to the logarithm of its data rate.
+Figure \ref{fig:packet-behavior} compares uniform (i.e. \caps{CBR}) packet behavior with the trace of an actual flow.
+In the uniform model, the cumulative data sent increases smoothly over time, whereas in the actual packet trace, the transmissions are variable both in size and in inter-packet interval, leading to a ``lumpy'' cumulative data plot.
+}
 \label{fig:trace-vs-uniform}

### 2008-06-12 tables/traffic_models.tex
Clean input formatting for some tables & figures.

@@
 \hline
-\caps{GMM} & --- & the general matrix model \\\hline\hline
-time uniform & $\mat{A}'=\mat{A}(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
-node uniform & $\mat{S}'=\mat{S}(\ones{n \times n}/n)$~~~~$\mat{D}'=\mat{D}(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
-full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above models \\\hline\hline
-time marginal & $\mat{A}'=(\ones{f \times f}/f)\mat{A}$ & start times chosen from marginal distribution of start times \\\hline
-size marginal & $\mat{B}'=(\ones{f \times f}/f)\mat{B}$ & flow size (bytes) is chosen independently of other properties \\\hline
-node marginal & $\mat{S}'=(\ones{f \times f}/f)\mat{S}$~~~~$\mat{D}'=(\ones{f \times f}/f)\mat{D}$ & nodes behave distinctly, but behavior is independent of flows \\\hline
-packet marginal & $\mat{Z}'=(\ones{f \times f}/f)\mat{Z}$~~~~$\mat{V}'=(\ones{f \times f}/f)\mat{V}$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
-full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline\hline
-source conditional & $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$ & flow behaviors are aggregated on a per source-node basis \\\hline
-dest conditional & $\mat{F}'=(\trans{\mat{DD}}/f)\mat{F}$ & flow behaviors are aggregated on a per destination-node basis \\\hline
-mixed conditional & $\mat{F}'=\frac{1}{2f}(\trans{\mat{SS}}+\trans{\mat{DD}})\mat{F}$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
+\caps{GMM} &
+--- &
+the general matrix model
+\\\hline\hline
+time uniform &
+$\mat{A}'=\mat{A}(\ones{d_t \times d_t}/d_t)$ &
+all start times are equally likely
+\\\hline
+node uniform &
+$\mat{S}'=\mat{S}(\ones{n \times n}/n)$~~~~$\mat{D}'=\mat{D}(\ones{n \times n}/n)$ &
+nodes behave stochastically identically with respect to flows
+\\\hline
+full uniform &
+\textit{all uniform transformations} &
+applies all the uniformity assumptions of the above models
+\\\hline\hline
+time marginal &
+$\mat{A}'=(\ones{f \times f}/f)\mat{A}$ &
+start times chosen from marginal distribution of start times
+\\\hline
+size marginal &
+$\mat{B}'=(\ones{f \times f}/f)\mat{B}$ &
+flow size (bytes) is chosen independently of other properties
+\\\hline
+node marginal &
+$\mat{S}'=(\ones{f \times f}/f)\mat{S}$~~~~$\mat{D}'=(\ones{f \times f}/f)\mat{D}$ &
+nodes behave distinctly, but behavior is independent of flows
+\\\hline
+packet marginal &
+$\mat{Z}'=(\ones{f \times f}/f)\mat{Z}$~~~~$\mat{V}'=(\ones{f \times f}/f)\mat{V}$ &
+packet behavior is variable bit-rate, aggregated across all flows
+\\\hline
+full marginal &
+\textit{all marginal transformations} &
+applies all the marginality assumptions of the above four models
+\\\hline\hline
+source conditional &
+$\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$ &
+flow behaviors are aggregated on a per source-node basis
+\\\hline
+dest conditional &
+$\mat{F}'=(\trans{\mat{DD}}/f)\mat{F}$ &
+flow behaviors are aggregated on a per destination-node basis
+\\\hline
+mixed conditional &
+$\mat{F}'=\frac{1}{2f}(\trans{\mat{SS}}+\trans{\mat{DD}})\mat{F}$ &
+flow behaviors are aggregated on a mixed source/destination basis
+\\\hline
 \end{tabular}

### 2008-06-12 paper.tex
Cleaning up some math in Linear Reps.

@@
 \begin{align}
-&\forall\: k \le \pkts: \quad \size_k > 0       \eqn{positive-packets}  \\
-&\forall\: k >   \pkts: \quad \size_k = 0       \eqn{zero-packets}      \\
-&\forall\: k <   \pkts: \quad \ival_k < \infty  \eqn{finite-intervals}  \\
-&\forall\: k \ge \pkts: \quad \ival_k = \infty. \eqn{infinite-intervals}
-\end{align}
-Packet sizes are required to be positive because we are modeling application layer data transfer:
-an empty data transmission request is a non-event.
+\forall\: k: \quad
+  \size_k >  \mathrlap{0} \phantom{\infty}
+  &\iff k \le \pkts
+  \eqn{packet-positivity} \\
+\forall\: k: \quad
+  \ival_k < \infty
+  &\iff k < \pkts.
+  \eqn{interval-finiteness}
+% \forall\: k: \quad k \le \pkts &\iff \size_k > 0       \eqn{packet-positivity} \\
+% \forall\: k: \quad k <   \pkts &\iff \ival_k < \infty. \eqn{interval-finiteness}
+\end{align}
+Actual packet sizes are required to be positive because we are modeling application layer data transfer:
+empty data transmission requests are non-events.
 The send time of the $k$th packet can be expressed as
@@
 \end{align}
-This is guaranteed to be finite if and only if $k \le \pkts$.
-Now we are ready to formalize the space of flows:
+\Equation{interval-finiteness} implies that this will be finite if and only if $k \le \pkts$.
+Finally, we formalize the space of flows:
 \begin{align}

### 2008-06-13 paper.tex
Heavy revision to Common Properties.
Notational consistency work, LinRep section.

@@
 Perhaps even more importantly, arbitrary linear combinations of vectors can be meaningfully interpreted as describing some realizable, hypothetical traffic instance.
-In \Section{traffic-generation} we will describe how to generate a flow for any sum of representation vectors.
+In \Section{traffic-generation} we will describe how to generate flows from arbitrary composite representation vectors.
 % This is how we generate workload from transformed traffic matrices.
@@
 \begin{align*}
-\vec{t} &= \e_k & k &= \Qf^{\text{t}}(\type) \\
-\vec{s} &= \e_k & k &= \Qf^{\text{s}}(\src) \\
-\vec{d} &= \e_k & k &= \Qf^{\text{d}}(\dst) \\
-\vec{a} &= \e_k & k &= \Qf^{\text{a}}(\start) \\
-\vec{b} &= \e_k & k &= \Qf^{\text{b}}(\bytes) \\
-\vec{z} &= {\textstyle \sum \e_k} & k &\in \set{\Qf^{\text{z}}(\size_i) : i \le \pkts} \\
-\vec{v} &= {\textstyle \sum \e_k} & k &\in \set{\Qf^{\text{v}}(\ival_i) : i <   \pkts}.
+\vec{t} &= \e_k & k &= \Qf_\type(\type) \\
+\vec{s} &= \e_k & k &= \Qf_\node(\src) \\
+\vec{d} &= \e_k & k &= \Qf_\node(\dst) \\
+\vec{a} &= \e_k & k &= \Qf_\time(\start) \\
+\vec{b} &= \e_k & k &= \Qf_\bytes(\bytes) \\
+\vec{z} &= {\textstyle \sum \e_k} & k &\in \set{\Qf_\size(\size_i) : i \le \pkts} \\
+\vec{v} &= {\textstyle \sum \e_k} & k &\in \set{\Qf_\ival(\ival_i) : i <   \pkts}.
 \end{align*}
-The functions $\Qf^\text{x}$ are the respective quantization functions for each property.
+The functions $\Qf_\prop$ are the respective quantization functions for each property.
 If there are $f$ flows in a traffic instance, then the collective behavior can be described with seven matrices---one for each property;
 the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
-We use the following notations for the behavior matrices:
+We use the following notations for the matrices and indexed flow vectors:
 \vspace{-0.25em}
@@
 $\mat{D}\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
-\>\bullet\> type: \>$\mat{T}$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$\\
+\>\bullet\> types: \>$\mat{T}$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$\\
 \>\bullet\> sources: \>$\mat{S}$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$\\
@@
 \end{align}
-We call this total representation the general matrix model.
-While it is by no means the only possible representation of traffic in terms of matrices, as demonstrate in the following sections, many common traffic modeling assumptions can be expressed succinctly and precisely in terms of transformations or conditions of the \caps{GMM} and its components.
+We call this total representation the \emph{general matrix model}.
+While it is by no means the only possible representation of traffic in terms of matrices, as we demonstrate in the following sections, many common traffic modeling assumptions can be expressed succinctly and precisely in terms of transformations or conditions of the \caps{GMM} and its components.
 Furthermore, realistic workload can be generated from matrix instances.
@@
 
-\subsubsection{Communication Topology}
+\subsubsection{Packet Size \& Inter-Packet Interval Statistics}
+\label{sec:flow-totals}
+
+From the matrices $\mat{Z}$ and $\mat{V}$ we can compute a variety of statistics about packet sizes and inter-packet intervals.
+The flows vectors in the matrices may represent individual flows or composites of many flows.
+% about packet sizes, total packet count, inter-packet intervals, and total duration for each representation vector, which may represent a single actual flow, or the composite behavior of many flows.
+% We begin by expressing the vectors for total bytes, total packets and total duration for a matrix of flow representations, with size and interval matrices $\mat{Z}$ and $\mat{V}$.
+The computations require knowing the values that each size of interval matrix column represents.
+% The computed duration is approximate because information is irretrievably lost in the quantization process.
+The vectors of representative values for $\mat{Z}$ and $\mat{V}$ respectively are:
+\begin{align}
+\vec{\sigma} &= \tuple{1,2,3,\dots,d_z} \\
+\vec{\tau}   &= \tuple{\Df_\ival\left(k-\fracx{1}{2}\right)}_{k=1}^{d_v}.
+\end{align}
+The packet count, total byte and total duration vectors are computed as
+\begin{align}
+\vec{\Sigma_{\pkts}}  &= \mat{Z}\trans{\ones{}} = \mat{V}\trans{\ones{}} \\
+\vec{\Sigma_{\bytes}} &= \mat{Z}\trans{\vec{\sigma}} \\
+\vec{\Sigma_{\ivals}} &= \mat{V}\trans{\vec{\tau}}.
+\end{align}
+Packet count and total bytes values are exact;
+the total durations are only approximate because information is irretrievably lost in the quantization process.
+Using these vectors of totals, we may readily compute the average packet sizes and inter-packet intervals:
+\begin{align}
+\vec{\mean{z}} &=
+  \vec{\Sigma_{\bytes}} \eldiv \vec{\Sigma_{\pkts}}
+  \phantom{= \mat{V}\trans{\vec{\tau}}   \eldiv \mat{V}\trans{\ones{}}}
+  \mathllap{= \mat{Z}\trans{\vec{\sigma}} \eldiv \mat{Z}\trans{\ones{}}} \\
+\vec{\mean{v}} &=
+  \phantom{\vec{\Sigma_{\bytes}} \eldiv \vec{\Sigma_{\pkts}}}
+  \mathllap{\vec{\Sigma_{\ivals}} \eldiv \vec{\Sigma_{\pkts}}}
+  = \mat{V}\trans{\vec{\tau}}   \eldiv \mat{V}\trans{\ones{}}.
+\end{align}
+The symbol $\eldiv$ here represents element-wise division of vector coordinates, as it does in Matlab.
+
+\subsubsection{Inter-Node Communication Graphs}
 \label{sec:communication-topology}
 
+The total number of source flows for each node is given by the vector $\trans{\mat{S}}\ones{f}$, and the total number of destination flows is given by $\trans{\mat{D}}\ones{f}$.
+($\ones{f}$ is the $f$-dimensional column vector consisting of all ones.)
 The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{SD}}\in\R^{n \times n}$ ($\trans{\,\diamond\,}$~denotes transpose).
 That is $(\trans{\mat{SD}})_{ij}$ is the number of flows from node $i$ to node $j$.
-The total number of source flows for each node is given by the vector $\trans{\mat{S}}\ones{f}$, and the total number of destination flows is given by $\trans{\mat{D}}\ones{f}$.
-($\ones{f}$ is the $f$-dimensional column vector consisting of all ones.)
-
-\subsubsection{Expected Average Packet Size \& Intervals}
-\label{sec:flow-totals}
-
-From the matrices $\mat{Z}$ and $\mat{V}$, we can compute the expected average packet size, $\mean{z}$, and average inter-packet interval, $\mean{v}$, for each flow.
-The computations require knowing the values that each matrix column represents.
-The computed duration is approximate because information is irrecoverably lost in the quantization process.
-The vectors of representative values for $\mat{Z}$ and $\mat{V}$ are:
-\begin{align}
-\vec{\sigma}&=\tuple{1,2,3,\dots,d_z},\\
-\vec{\tau}&=\tuple{\Di\left(k,\fracx{1}{2}\right)}_{k=1}^{d_v}.
-\end{align}
-Using these representative values, the expected average packet size and expected average inter-packet interval size for a flow may be computed as
-\begin{align}
-  \mean{z}&=\frac{\inner{\vec{z}}{\vec{\sigma}}}{\inner{\vec{z}}{\ones{}}}, &
-  \mean{v}&=\frac{\inner{\vec{v}}{\vec{\tau}}}{\inner{\vec{v}}{\ones{}}},
-\end{align}
-respectively, where $\inner{\vec{x}}{\vec{y}}$ denotes the standard inner product of vectors.
+Interpreting this sparse matrix as a weighted graph immediately yields a simple and efficient means to compute this and other useful communication graphs.
+Other graphs can readily be computed with a few additional matrix multiplications.
+For example, using the computations given
 
@@
 
-\subsection{Conditionality}
+\subsubsection{Conditionality}
 

### 2008-06-13 paper.tex
Various fidgetting.

@@
 \vec{t} &= \e_k & k &= \Qf_\type(\type) \\
-\vec{s} &= \e_k & k &= \Qf_\node(\src) \\
-\vec{d} &= \e_k & k &= \Qf_\node(\dst) \\
-\vec{a} &= \e_k & k &= \Qf_\time(\start) \\
+\vec{s} &= \e_k & k &= \Qf_\src(\src) \\
+\vec{d} &= \e_k & k &= \Qf_\dst(\dst) \\
+\vec{a} &= \e_k & k &= \Qf_\start(\start) \\
 \vec{b} &= \e_k & k &= \Qf_\bytes(\bytes) \\
@@
 Choose $u \in [0,1]$ uniformly randomly.
-Then $x = k-u \in [0,d]$, with probability of lying in the interval $(i-1,i]$ proportional to the coefficient of $\e_i$ in $\vec{f}_\prop$.
-The dequantization function for $\prop$, call it $\Df_\prop$, will map the value $x$ back into the appropriate range of property values for $\prop$.
+Then $x = k-u \in [0,d]$, with probability of lying in the interval $[i-1,i]$ proportional to the coefficient of $\e_i$ in $\vec{f}_\prop$.
+The dequantization function for $\prop$, namely $\Df_\prop$, will map the value $x$ back into the appropriate range of property values for $\prop$.
 
@@
 \item Sample random source and destination node pairs.
-Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding the collisions.
+Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding collisions.
 The simplest technique is the rejection method:
@@
 
+\newfootnote{\onesnote}{%
+Through out this paper, $\ones{}$ denotes a matrix of all ones, but the dimensions of the matrix must be inferred by context.
+Where necessary, we disambiguate the dimensions by explicitly specifying the dimensions of the final matrix product.
+}
+
 From the matrices $\mat{Z}$ and $\mat{V}$ we can compute a variety of statistics about packet sizes and inter-packet intervals.
@@
 \end{align}
-The packet count, total byte and total duration vectors are computed as
-\begin{align}
-\vec{\Sigma_{\pkts}}  &= \mat{Z}\trans{\ones{}} = \mat{V}\trans{\ones{}} \\
-\vec{\Sigma_{\bytes}} &= \mat{Z}\trans{\vec{\sigma}} \\
-\vec{\Sigma_{\ivals}} &= \mat{V}\trans{\vec{\tau}}.
-\end{align}
+The packet count, total byte and total duration vectors are computed as\onesnote
+\begin{align}
+\vec{\Sigma}_\pkts  &= \mat{Z}\trans{\ones{}} = \mat{V}\trans{\ones{}} \in \R^f \eqn{packet-count}\\
+\vec{\Sigma}_\bytes &= \mat{Z}\trans{\vec{\sigma}} \in \R^f \eqn{total-bytes} \\
+\vec{\Sigma}_\ivals &= \mat{V}\trans{\vec{\tau}} \in \R^f \eqn{total-duration}.
+\end{align}
+Here $\ones{}$ is a row vector of all ones.
 Packet count and total bytes values are exact;
@@
 \vec{\mean{z}} &=
-  \vec{\Sigma_{\bytes}} \eldiv \vec{\Sigma_{\pkts}}
+  \vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts
   \phantom{= \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones{}}}
@@
 \vec{\mean{v}} &=
-  \phantom{\vec{\Sigma_{\bytes}} \eldiv \vec{\Sigma_{\pkts}}}
+  \phantom{\vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts}
   \mathllap{\vec{\Sigma_{\ivals}} \eldiv \vec{\Sigma_{\pkts}}}
@@
 
-The total number of source flows for each node is given by the vector $\trans{\mat{S}}\ones{f}$, and the total number of destination flows is given by $\trans{\mat{D}}\ones{f}$.
-($\ones{f}$ is the $f$-dimensional column vector consisting of all ones.)
-The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{SD}}\in\R^{n \times n}$ ($\trans{\,\diamond\,}$~denotes transpose).
-That is $(\trans{\mat{SD}})_{ij}$ is the number of flows from node $i$ to node $j$.
-Interpreting this sparse matrix as a weighted graph immediately yields a simple and efficient means to compute this and other useful communication graphs.
-Other graphs can readily be computed with a few additional matrix multiplications.
-For example, using the computations given
+The total number of source flows for each node is given by the vector $\trans{\mat{S}}\ones{f} \in \R^n$, and the total number of destination flows is given by $\trans{\mat{D}}\ones{f} \in \R^n$.
+The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{S}}\mat{D}\in\R^{n \times n}$.
+That is $(\trans{\mat{S}}\mat{D})_{ij}$ is the number of flows from node $i$ to node $j$.
+Interpreting this sparse matrix as a weighted graph immediately yields the source-destination flow communication graph.
+Similar graphs counting other quantities can readily be computed by inserting the appropriate weighting matrix.
+For example, we may compute the packet transmission graph:
+\begin{align}
+\trans{\mat{S}}\diag(\vec{\Sigma}_\pkts)\mat{D}\in\R^{n \times n}.
+\end{align}
+Here $\vec{\Sigma}_\pkts$ is the byte vector defined in \Equation{packet-count}, and the ``$\diag$'' operator gives the square, diagonal matrix with the given vector of values on its diagonal.
+Similar graphs for total bytes and total duration of flows may be computed using $\mat{\Sigma}_\bytes$ or $\mat{\Sigma}_\ival$ in place of $\mat{\Sigma}_\pkts$.
 
@@
 We use the terms \textit{uniformity}, \textit{marginality}, and \textit{conditionality} to describe these classes of assumptions.
-They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether the make stipulations about uniformity within each flow, across all flows, or within groups of flows.
+They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether they make stipulations about uniformity within each flow, across all flows, or within groups of flows.
 In the following sections, we explore how various modeling concepts can be expressed in the framework of the general matrix model.
@@
 For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $\mat{S}$ and $\mat{D}$:
-\begin{align*}
+\begin{align}
 \mat{S}' &= \mat{S}\ones{n \times n}/n, & \mat{D}' &= \mat{D}\ones{n \times n}/n.
-\end{align*}
+\end{align}
 This makes all the values in each flow's row identically the average of the original $n$ values.
@@
 The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $[\mat{Z}~\mat{V}]$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $[\mat{Z}~\mat{V}]\,\Gamma\in\R^{f \times 3}$, where
-\begin{align}
-\Gamma=\begin{pmatrix}
+The pair of matrices $[\mat{Z}~\mat{V}]$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $[\mat{Z}~\mat{V}]\,\mat{\Gamma}\in\R^{f \times 3}$, where
+\begin{align}
+\mat{\Gamma}=\begin{pmatrix}
 \ones{} & \vec{\sigma} & \zeros{} \\

### 2008-06-13 paper.tex
Moved TrafGen after CommonProps & CommonModels.

@@
 These parameters are non-linear scaling exponents that allows the quantization to smoothly shift from fine-grained resolution at the low end of the spectrum to coarse-grained resolution at the high end.
-We have chosen $\beta=\fracx{1}{2.7}$ because it yields a desirable dynamic range response from bytes at the low end to megabytes at the high end.
-Similarly, we have chosen $\gamma=\fracx{1}{3}$ because this choice provides resolution that scales smoothly from microseconds for small intervals to seconds for large intervals.
+We have chosen $\beta=\fracx{1}{2.7}$ and $\gamma=\fracx{1}{3}$ because these values yield desirable dynamic ranges: from single bytes at the lower end to megabytes at the high end for flow sizes; from microseconds to seconds for intervals.
+
+\subsection{Computation of Common Properties}
+\label{sec:common-properties}
+
+Having defined our representation of network behavior and presented the procedure for stochastically generating workload, we turn to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.
+
+\subsubsection{Packet Size \& Inter-Packet Interval Statistics}
+\label{sec:flow-totals}
+
+\newfootnote{\onesnote}{%
+Through out this paper, $\ones{}$ denotes a matrix of all ones; dimensions of the matrix are inferred by context.
+Where necessary, we disambiguate the dimensions by explicitly specifying the dimensions of the final matrix product, as we do here.
+}
+
+From the matrices $\mat{Z}$ and $\mat{V}$ we can compute a variety of statistics about packet sizes and inter-packet intervals.
+The flows vectors in the matrices may represent individual flows or composites of many flows.
+% about packet sizes, total packet count, inter-packet intervals, and total duration for each representation vector, which may represent a single actual flow, or the composite behavior of many flows.
+% We begin by expressing the vectors for total bytes, total packets and total duration for a matrix of flow representations, with size and interval matrices $\mat{Z}$ and $\mat{V}$.
+The computations require knowing the values that each size of interval matrix column represents.
+% The computed duration is approximate because information is irretrievably lost in the quantization process.
+The vectors of representative values for $\mat{Z}$ and $\mat{V}$ respectively are:
+\begin{align}
+\vec{\sigma} &= \tuple{1,2,3,\dots,d_z} \\
+\vec{\tau}   &= \tuple{\Df_\ival\left(k-\fracx{1}{2}\right)}_{k=1}^{d_v}.
+\end{align}
+The packet count, total byte and total duration vectors are computed as
+\begin{align}
+\vec{\Sigma}_\pkts  &= \mat{Z}\trans{\ones{}} = \mat{V}\trans{\ones{}} \in \R^f \eqn{packet-count}\\
+\vec{\Sigma}_\bytes &= \mat{Z}\trans{\vec{\sigma}} \in \R^f \eqn{total-bytes} \\
+\vec{\Sigma}_\ivals &= \mat{V}\trans{\vec{\tau}} \in \R^f \eqn{total-duration}.
+\end{align}
+Here $\ones{}$ is a row vector of all ones.\onesnote
+Packet count and total bytes values are exact;
+the total durations are only approximate because information is irretrievably lost in the quantization process.
+Using these vectors of totals, we may readily compute the average packet sizes and inter-packet intervals:
+\begin{align}
+\vec{\mean{z}} &=
+  \vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts
+  \phantom{= \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones{}}}
+  \mathllap{= \mat{Z}\trans{\vec{\sigma}} \eldiv \mat{Z}\trans{\ones{}}} \\
+\vec{\mean{v}} &=
+  \phantom{\vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts}
+  \mathllap{\vec{\Sigma_{\ivals}} \eldiv \vec{\Sigma_{\pkts}}}
+  = \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones{}}.
+\end{align}
+The symbol $\eldiv$ here represents element-wise division of vector coordinates, as it does in Matlab.
+
+\subsubsection{Inter-Node Communication Graphs}
+\label{sec:communication-topology}
+
+The total number of source flows for each node is given by the vector $\trans{\mat{S}}\ones{f} \in \R^n$, and the total number of destination flows is given by $\trans{\mat{D}}\ones{f} \in \R^n$.
+The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{S}}\mat{D}\in\R^{n \times n}$.
+That is $(\trans{\mat{S}}\mat{D})_{ij}$ is the number of flows from node $i$ to node $j$.
+Interpreting this sparse matrix as a weighted graph immediately yields the source-destination flow communication graph.
+\Figure{trace-vs-uniform} gives two visual examples of such graphs.
+Similar graphs counting other quantities can readily be computed by inserting the appropriate weighting matrix.
+For example, we may compute the inter-node packet transmission graph:
+\begin{align}
+\trans{\mat{S}}\diag(\vec{\Sigma}_\pkts)\mat{D}\in\R^{n \times n}.
+\end{align}
+Here $\vec{\Sigma}_\pkts$ is the byte vector defined in \Equation{packet-count}, and the ``$\diag$'' operator gives the square, diagonal matrix with the given vector of values on its diagonal.
+Similar graphs for total bytes and total duration of flows may be computed using $\mat{\Sigma}_\bytes$ or $\mat{\Sigma}_\ival$ in place of $\mat{\Sigma}_\pkts$.
+
+\subsection{Expression of Common Modeling Concepts}
+\sec{modeling-concepts}
+
+The simplifying assumptions made by most traffic models can be classified into three categories.
+We use the terms \textit{uniformity}, \textit{marginality}, and \textit{conditionality} to describe these classes of assumptions.
+They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether they make stipulations about uniformity within each flow, across all flows, or within groups of flows.
+In the following sections, we explore how various modeling concepts can be expressed in the framework of the general matrix model.
+Models are expressed as transformations that convert arbitrary traffic matrix instances into similar ones that satisfy the model's assumptions.
+These transformations later allow us to evaluate the effect of each model's simplifications on the realism of generated workloads.
+
+\subsubsection{Uniformity}
+
+The concept of uniformity, as used here, entails that certain properties of each flow are assumed to be statistically or deterministically uniform.
+Examples include: assuming that each flow's packets have the same size and inter-packet interval (i.e.
+the \caps{CBR} flow model); assuming that each node is equally likely to be the source of a given flow; assuming that all start times for a flow are equally likely.
+
+Since uniform behaviors entail homogeneity of properties with respect to each flow individually, a trace matrix can be transformed to a matrix satisfying the model's assumptions via right matrix multiplication.
+Right multiplication mixes the elements within each row, and if the mixing weights are uniform, the result is uniform behavior.
+For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $\mat{S}$ and $\mat{D}$:
+\begin{align}
+\mat{S}' &= \mat{S}\ones{n \times n}/n, & \mat{D}' &= \mat{D}\ones{n \times n}/n.
+\end{align}
+This makes all the values in each flow's row identically the average of the original $n$ values.
+As another example, the constant bit-rate (\caps{CBR}) flow model, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
+The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
+The pair of matrices $[\mat{Z}~\mat{V}]$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $[\mat{Z}~\mat{V}]\,\mat{\Gamma}\in\R^{f \times 3}$, where
+\begin{align}
+\mat{\Gamma}=\begin{pmatrix}
+\ones{} & \vec{\sigma} & \zeros{} \\
+\zeros{} & \zeros{} & \vec{\tau}
+\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
+\end{align}
+\Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to realistic behaviors from traces.
+%By replacing $\mat{S}'$ with $\mat{S}$, this can also be viewed as a set of linear constraints on the matrix $\mat{S}$, providing a proscriptive definition
+
+%We use the term \textit{uniformity} do describe assumptions made about the ``sameness'' of various properties of flows.
+The constant bit-rate (\caps{CBR}) model, for example, is an assumption about the uniformity of all of a flow's packets: i.e.
+that they have the same size and are evenly spaced in time.
+While this is the strictest sense of the \caps{CBR} model, it is rare for this assumption to be made in isolation.
+Another common simplification is to make the sizes of flows identical, or chose them randomly from a uniform distribution; similarly for the starting time of each flow.
+Finally, when choosing endpoints for a given flow, lacking a better model, it is not unusual to treat all nodes identically, making each an equally likely source or destination for a given flow.
+
+%It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees.
+This is largely because of the high complexity of doing otherwise and the lack of compelling alternatives.
+Here we examine how various notions of uniformity are expressed in the matrix model.
+In general, \textit{intra-flow uniformity} is effected by right multiplication, which mixes the elements in each row, whereas \textit{inter-flow uniformity} is effected by left multiplication, which mixes elements across rows.
+The constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
+The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
+The pair of matrices $\tuple{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\tuple{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
+%\begin{align}
+%\Gamma=\begin{pmatrix}
+%\ones{} & \vec{\sigma} & \zeros{} \\
+%\zeros{} & \zeros{} & \vec{\tau}
+%\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
+%\end{align}
+
+\subsubsection{Marginality}
+
+Marginality is an orthogonal concept to uniformity: the distribution of some property is identical \textit{across} all flows.
+For example, if we determine a distribution of flow sizes across the entire network and assume that this distribution applies \textit{independent of other flow properties}, this is a marginal model for flow size.
+Marginal behavior is effected by left multiplication by a matrix with uniform columns.
+This specific example is generated by left-uniformizing the flow size matrix: $\mat{B}' = (\ones{f \times f}/f)\mat{B}$.
+The traffic models proposed by {\FHC} are all marginal: they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node.
+These models do not account for non-independence between these and other flow properties.
+We will discuss this limitation and its implications when we analyze our experimental results.
+
+\subsubsection{Conditionality}
+
+The final form of homogenization we consider, is conditionality.
+This is a restricted form of marginality: we may assume that marginal distributions apply to all the flows in certain groupings.
+For example, we can provide a distribution of flow sizes on a per-source basis.
+That is, each source node can have its \textit{own} distribution, and the sizes of its flows are drawn from that distribution, independent of their other properties.
+This is simply accomplished by left-multiplication by a non-uniform matrix.
+For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-conditioning'' matrix: $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$.
+Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-conditioning matrix: $F'=(\trans{\mat{DD}}/f)\mat{F}$.
+%Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified.
+The confusion arises due to the lack of a unique way of associating flows with nodes.
+Each node can equally viably be associated with its source or its destination.
+Which association is more valid?
+Consider a web request: should the behavior of the request be determined by what the client typically does?
+Or should the requests that the server typically receives determine behavior?
+There is no armchair answer: only analysis of real data can give us the answer.
+The matrix model provides an analytical framework for doing so.
+The matrix model has the benefit of making the conditioning of behaviors completely explicit.
+It also allows us to easily define intermediate choices.
+If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
+\begin{align}
+\mat{F}' = \frac{1}{f}((1-\alpha)(\trans{\mat{SS}})+\alpha(\trans{\mat{DD}}))\mat{F}.
+\end{align}
+The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors.
+Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.
+
+%We use the terms here to distinguish three fundamentally different types of assumptions about behavioral homogeneity.
+
+%Existing traffic models are typically not explicit methods of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
+In our algebraic modeling framework, such models can be interpreted as stipulations about the form of the behavior matrices, or as transforms that implement those stipulations given a real set of traffic matrices.
+
+%A generic traffic model is a way of generating the six matrices described in \Section{representations}.
+A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or conditional form.
+For experimental comparison, the dimensions of the matrices should all be preserved.
+If a trace-derived model produces experimental performance metrics that are similar to those produced by the original trace data, then simplifications made by the model are viable; if the performance is distorted, then the simplifications are discarding essential information about the traffic pattern.
+
+%\subsection{Uniformity}
+
+%% TODO: differentiate between restrictive versions of models and transformative versions.
+%% TODO: distinguish uniform, marginal and conditional models.
+
+%\subsection{Marginality}
+
+%Most approaches to traffic modeling that go beyond uniformity assumptions, instead describe marginal properties of various aspects of traffic behavior.
+For example, {\FHC}~\cite{Hernandez06:spatio-temporal-modeling} provide highly convincing parametric models for six marginal properties of wireless network traffic: session (node) arrivals follow a time-varying Poisson process; per-session flow inter-arrival times are Lognormally distributed; flows per session (node) follow a BiPareto distribution; flow sizes follow a BiPareto distribution.
+This is certainly much greater detail than simply assuming that all nodes and flows are identical.
+
+%How do marginal models translate into the algebraic language of the general matrix model? Whereas uniformity assumptions for nodes,
+
+%Another benefit of the matrix model is that all aspects of network behavior are determined concurrently.
+We are not left wondering what to do about source and destination nodes or flow start and end times after we have computed our simplified packet behaviors: the same transformation that average the packet behaviors also naturally averages the start and
 
@@
 
-\subsection{Expression of Common Properties}
-\label{sec:common-properties}
-
-Having defined our representation of network behavior and presented the procedure for stochastically generating workload, we turn to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.
-
-\subsubsection{Packet Size \& Inter-Packet Interval Statistics}
-\label{sec:flow-totals}
-
-\newfootnote{\onesnote}{%
-Through out this paper, $\ones{}$ denotes a matrix of all ones, but the dimensions of the matrix must be inferred by context.
-Where necessary, we disambiguate the dimensions by explicitly specifying the dimensions of the final matrix product.
-}
-
-From the matrices $\mat{Z}$ and $\mat{V}$ we can compute a variety of statistics about packet sizes and inter-packet intervals.
-The flows vectors in the matrices may represent individual flows or composites of many flows.
-% about packet sizes, total packet count, inter-packet intervals, and total duration for each representation vector, which may represent a single actual flow, or the composite behavior of many flows.
-% We begin by expressing the vectors for total bytes, total packets and total duration for a matrix of flow representations, with size and interval matrices $\mat{Z}$ and $\mat{V}$.
-The computations require knowing the values that each size of interval matrix column represents.
-% The computed duration is approximate because information is irretrievably lost in the quantization process.
-The vectors of representative values for $\mat{Z}$ and $\mat{V}$ respectively are:
-\begin{align}
-\vec{\sigma} &= \tuple{1,2,3,\dots,d_z} \\
-\vec{\tau}   &= \tuple{\Df_\ival\left(k-\fracx{1}{2}\right)}_{k=1}^{d_v}.
-\end{align}
-The packet count, total byte and total duration vectors are computed as\onesnote
-\begin{align}
-\vec{\Sigma}_\pkts  &= \mat{Z}\trans{\ones{}} = \mat{V}\trans{\ones{}} \in \R^f \eqn{packet-count}\\
-\vec{\Sigma}_\bytes &= \mat{Z}\trans{\vec{\sigma}} \in \R^f \eqn{total-bytes} \\
-\vec{\Sigma}_\ivals &= \mat{V}\trans{\vec{\tau}} \in \R^f \eqn{total-duration}.
-\end{align}
-Here $\ones{}$ is a row vector of all ones.
-Packet count and total bytes values are exact;
-the total durations are only approximate because information is irretrievably lost in the quantization process.
-Using these vectors of totals, we may readily compute the average packet sizes and inter-packet intervals:
-\begin{align}
-\vec{\mean{z}} &=
-  \vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts
-  \phantom{= \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones{}}}
-  \mathllap{= \mat{Z}\trans{\vec{\sigma}} \eldiv \mat{Z}\trans{\ones{}}} \\
-\vec{\mean{v}} &=
-  \phantom{\vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts}
-  \mathllap{\vec{\Sigma_{\ivals}} \eldiv \vec{\Sigma_{\pkts}}}
-  = \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones{}}.
-\end{align}
-The symbol $\eldiv$ here represents element-wise division of vector coordinates, as it does in Matlab.
-
-\subsubsection{Inter-Node Communication Graphs}
-\label{sec:communication-topology}
-
-The total number of source flows for each node is given by the vector $\trans{\mat{S}}\ones{f} \in \R^n$, and the total number of destination flows is given by $\trans{\mat{D}}\ones{f} \in \R^n$.
-The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{S}}\mat{D}\in\R^{n \times n}$.
-That is $(\trans{\mat{S}}\mat{D})_{ij}$ is the number of flows from node $i$ to node $j$.
-Interpreting this sparse matrix as a weighted graph immediately yields the source-destination flow communication graph.
-Similar graphs counting other quantities can readily be computed by inserting the appropriate weighting matrix.
-For example, we may compute the packet transmission graph:
-\begin{align}
-\trans{\mat{S}}\diag(\vec{\Sigma}_\pkts)\mat{D}\in\R^{n \times n}.
-\end{align}
-Here $\vec{\Sigma}_\pkts$ is the byte vector defined in \Equation{packet-count}, and the ``$\diag$'' operator gives the square, diagonal matrix with the given vector of values on its diagonal.
-Similar graphs for total bytes and total duration of flows may be computed using $\mat{\Sigma}_\bytes$ or $\mat{\Sigma}_\ival$ in place of $\mat{\Sigma}_\pkts$.
-
-\subsection{Expression of Common Modeling Concepts}
-\sec{modeling-concepts}
-
-The simplifying assumptions made by most traffic models can be classified into three categories.
-We use the terms \textit{uniformity}, \textit{marginality}, and \textit{conditionality} to describe these classes of assumptions.
-They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether they make stipulations about uniformity within each flow, across all flows, or within groups of flows.
-In the following sections, we explore how various modeling concepts can be expressed in the framework of the general matrix model.
-Models are expressed as transformations that will convert arbitrary traffic matrix instances into similar ones that satisfy the model's assumptions.
-These transformations will later allow us to evaluate the effect of each model's simplifications on the realism of generated workloads.
-
-\subsubsection{Uniformity}
-
-The concept of uniformity, as used here, entails that certain properties of each flow are assumed to be statistically or deterministically uniform.
-Examples include: assuming that each flow's packets have the same size and inter-packet interval (i.e.
-the \caps{CBR} flow model); assuming that each node is equally likely to be the source of a given flow; assuming that all start times for a flow are equally likely.
-
-Since uniform behaviors entail homogeneity of properties with respect to each flow individually, a trace matrix can be transformed to a matrix satisfying the model's assumptions via right matrix multiplication.
-Right multiplication mixes the elements within each row, and if the mixing weights are uniform, the result is uniform behavior.
-For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $\mat{S}$ and $\mat{D}$:
-\begin{align}
-\mat{S}' &= \mat{S}\ones{n \times n}/n, & \mat{D}' &= \mat{D}\ones{n \times n}/n.
-\end{align}
-This makes all the values in each flow's row identically the average of the original $n$ values.
-As another example, the constant bit-rate (\caps{CBR}) flow model, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
-The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $[\mat{Z}~\mat{V}]$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $[\mat{Z}~\mat{V}]\,\mat{\Gamma}\in\R^{f \times 3}$, where
-\begin{align}
-\mat{\Gamma}=\begin{pmatrix}
-\ones{} & \vec{\sigma} & \zeros{} \\
-\zeros{} & \zeros{} & \vec{\tau}
-\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
-\end{align}
-\Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to realistic behaviors from traces.
-%By replacing $\mat{S}'$ with $\mat{S}$, this can also be viewed as a set of linear constraints on the matrix $\mat{S}$, providing a proscriptive definition
-
-%We use the term \textit{uniformity} do describe assumptions made about the ``sameness'' of various properties of flows.
-The constant bit-rate (\caps{CBR}) model, for example, is an assumption about the uniformity of all of a flow's packets: i.e.
-that they have the same size and are evenly spaced in time.
-While this is the strictest sense of the \caps{CBR} model, it is rare for this assumption to be made in isolation.
-Another common simplification is to make the sizes of flows identical, or chose them randomly from a uniform distribution; similarly for the starting time of each flow.
-Finally, when choosing endpoints for a given flow, lacking a better model, it is not unusual to treat all nodes identically, making each an equally likely source or destination for a given flow.
-
-%It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees.
-This is largely because of the high complexity of doing otherwise and the lack of compelling alternatives.
-Here we examine how various notions of uniformity are expressed in the matrix model.
-In general, \textit{intra-flow uniformity} is effected by right multiplication, which mixes the elements in each row, whereas \textit{inter-flow uniformity} is effected by left multiplication, which mixes elements across rows.
-The constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
-The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $\tuple{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\tuple{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
-%\begin{align}
-%\Gamma=\begin{pmatrix}
-%\ones{} & \vec{\sigma} & \zeros{} \\
-%\zeros{} & \zeros{} & \vec{\tau}
-%\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
-%\end{align}
-
-\subsubsection{Marginality}
-
-Marginality is an orthogonal concept to uniformity: the distribution of some property is identical \textit{across} all flows.
-For example, if we determine a distribution of flow sizes across the entire network and assume that this distribution applies \textit{independent of other flow properties}, this is a marginal model for flow size.
-Marginal behavior is effected by left multiplication by a matrix with uniform columns.
-This specific example is generated by left-uniformizing the flow size matrix: $\mat{B}' = (\ones{f \times f}/f)\mat{B}$.
-The traffic models proposed by {\FHC} are all marginal: they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node.
-These models do not account for non-independence between these and other flow properties.
-We will discuss this limitation and its implications when we analyze our experimental results.
-
-\subsubsection{Conditionality}
-
-The final form of homogenization we consider, is conditionality.
-This is a restricted form of marginality: we may assume that marginal distributions apply to all the flows in certain groupings.
-For example, we can provide a distribution of flow sizes on a per-source basis.
-That is, each source node can have its \textit{own} distribution, and the sizes of its flows are drawn from that distribution, independent of their other properties.
-This is simply accomplished by left-multiplication by a non-uniform matrix.
-For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-conditioning'' matrix: $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$.
-Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-conditioning matrix: $F'=(\trans{\mat{DD}}/f)\mat{F}$.
-%Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified.
-The confusion arises due to the lack of a unique way of associating flows with nodes.
-Each node can equally viably be associated with its source or its destination.
-Which association is more valid?
-Consider a web request: should the behavior of the request be determined by what the client typically does?
-Or should the requests that the server typically receives determine behavior?
-There is no armchair answer: only analysis of real data can give us the answer.
-The matrix model provides an analytical framework for doing so.
-The matrix model has the benefit of making the conditioning of behaviors completely explicit.
-It also allows us to easily define intermediate choices.
-If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
-\begin{align}
-\mat{F}' = \frac{1}{f}((1-\alpha)(\trans{\mat{SS}})+\alpha(\trans{\mat{DD}}))\mat{F}.
-\end{align}
-The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors.
-Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.
-
-%We use the terms here to distinguish three fundamentally different types of assumptions about behavioral homogeneity.
-
-%Existing traffic models are typically not explicit methods of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
-In our algebraic modeling framework, such models can be interpreted as stipulations about the form of the behavior matrices, or as transforms that implement those stipulations given a real set of traffic matrices.
-
-%A generic traffic model is a way of generating the six matrices described in \Section{representations}.
-A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or conditional form.
-For experimental comparison, the dimensions of the matrices should all be preserved.
-If a trace-derived model produces experimental performance metrics that are similar to those produced by the original trace data, then simplifications made by the model are viable; if the performance is distorted, then the simplifications are discarding essential information about the traffic pattern.
-
-%\subsection{Uniformity}
-
-%% TODO: differentiate between restrictive versions of models and transformative versions.
-%% TODO: distinguish uniform, marginal and conditional models.
-
-%\subsection{Marginality}
-
-%Most approaches to traffic modeling that go beyond uniformity assumptions, instead describe marginal properties of various aspects of traffic behavior.
-For example, {\FHC}~\cite{Hernandez06:spatio-temporal-modeling} provide highly convincing parametric models for six marginal properties of wireless network traffic: session (node) arrivals follow a time-varying Poisson process; per-session flow inter-arrival times are Lognormally distributed; flows per session (node) follow a BiPareto distribution; flow sizes follow a BiPareto distribution.
-This is certainly much greater detail than simply assuming that all nodes and flows are identical.
-
-%How do marginal models translate into the algebraic language of the general matrix model? Whereas uniformity assumptions for nodes,
-
-%Another benefit of the matrix model is that all aspects of network behavior are determined concurrently.
-We are not left wondering what to do about source and destination nodes or flow start and end times after we have computed our simplified packet behaviors: the same transformation that average the packet behaviors also naturally averages the start and
-
 \input{tables/traffic_models}

### 2008-06-13 paper.tex
Lots of work on Common Assumptions section.

@@
 \newfootnote{\onesnote}{%
-Through out this paper, $\ones{}$ denotes a matrix of all ones; dimensions of the matrix are inferred by context.
+Through out this paper, $\ones$ denotes a matrix of all ones; dimensions of the matrix are inferred by context.
 Where necessary, we disambiguate the dimensions by explicitly specifying the dimensions of the final matrix product, as we do here.
@@
 \begin{align}
-\vec{\Sigma}_\pkts  &= \mat{Z}\trans{\ones{}} = \mat{V}\trans{\ones{}} \in \R^f \eqn{packet-count}\\
+\vec{\Sigma}_\pkts  &= \mat{Z}\trans{\ones} = \mat{V}\trans{\ones} \in \R^f \eqn{packet-count}\\
 \vec{\Sigma}_\bytes &= \mat{Z}\trans{\vec{\sigma}} \in \R^f \eqn{total-bytes} \\
@@
 \end{align}
-Here $\ones{}$ is a row vector of all ones.\onesnote
+Here $\ones$ is a row vector of all ones.\onesnote
 Packet count and total bytes values are exact;
@@
   \vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts
-  \phantom{= \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones{}}}
-  \mathllap{= \mat{Z}\trans{\vec{\sigma}} \eldiv \mat{Z}\trans{\ones{}}} \\
+  \phantom{= \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones}}
+  \mathllap{= \mat{Z}\trans{\vec{\sigma}} \eldiv \mat{Z}\trans{\ones}} \\
 \vec{\mean{v}} &=
@@
   \mathllap{\vec{\Sigma_{\ivals}} \eldiv \vec{\Sigma_{\pkts}}}
-  = \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones{}}.
+  = \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones}.
 \end{align}
@@
 
-The total number of source flows for each node is given by the vector $\trans{\mat{S}}\ones{f} \in \R^n$, and the total number of destination flows is given by $\trans{\mat{D}}\ones{f} \in \R^n$.
+The total number of source flows for each node is given by the vector $\trans{\mat{S}}\ones \in \R^n$, and the total number of destination flows is given by $\trans{\mat{D}}\ones \in \R^n$.
 The number of flows between each pair of nodes in the network is expressed by the matrix product $\trans{\mat{S}}\mat{D}\in\R^{n \times n}$.
@@
 The concept of uniformity, as used here, entails that certain properties of each flow are assumed to be statistically or deterministically uniform.
-Examples include: assuming that each flow's packets have the same size and inter-packet interval (i.e.
-the \caps{CBR} flow model); assuming that each node is equally likely to be the source of a given flow; assuming that all start times for a flow are equally likely.
+Examples include:
+assuming that each flow's packets have the same size and inter-packet interval (i.e. the \caps{CBR} flow model);
+assuming that each node is equally likely to be the source of a given flow; assuming that all start times for a flow are equally likely.
 
@@
 \begin{align}
-\mat{S}' &= \mat{S}\ones{n \times n}/n, & \mat{D}' &= \mat{D}\ones{n \times n}/n.
-\end{align}
-This makes all the values in each flow's row identically the average of the original $n$ values.
-As another example, the constant bit-rate (\caps{CBR}) flow model, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
+\mat{S}' &= \rStoch(\mat{S}\ones) = \smallfrac{1}{n} \mat{S}\ones \in \R^{f \times n} \\
+\mat{D}' &= \rStoch(\mat{D}\ones) = \smallfrac{1}{n} \mat{D}\ones\in \R^{f \times n}.
+\end{align}
+The ``$\rStoch$'' operator scales each row to sum to unity, thus making the entire matrix row-stochastic;
+in this case, this is equivalent to scaling by $\smallfrac{1}{n}$.
+These transformations of $\mat{S}$ and $\mat{D}$ make all the values in each flow's row identically the average of the original $n$ values.
+As another example, the constant bit-rate (\caps{CBR}) flow model, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals;
+thus each flow's bit-rate is constant, as the name implies.
 The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $[\mat{Z}~\mat{V}]$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $[\mat{Z}~\mat{V}]\,\mat{\Gamma}\in\R^{f \times 3}$, where
+The pair of matrices $[\mat{Z}~\mat{V}]$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}:
+$[\mat{Z}~\mat{V}]\,\mat{\Gamma}\in\R^{f \times 3}$, where
 \begin{align}
 \mat{\Gamma}=\begin{pmatrix}
-\ones{} & \vec{\sigma} & \zeros{} \\
-\zeros{} & \zeros{} & \vec{\tau}
+\ones & \vec{\sigma} & \zeros \\
+\zeros & \zeros & \vec{\tau}
 \end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
@@
 \Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to realistic behaviors from traces.
-%By replacing $\mat{S}'$ with $\mat{S}$, this can also be viewed as a set of linear constraints on the matrix $\mat{S}$, providing a proscriptive definition
-
-%We use the term \textit{uniformity} do describe assumptions made about the ``sameness'' of various properties of flows.
-The constant bit-rate (\caps{CBR}) model, for example, is an assumption about the uniformity of all of a flow's packets: i.e.
-that they have the same size and are evenly spaced in time.
-While this is the strictest sense of the \caps{CBR} model, it is rare for this assumption to be made in isolation.
-Another common simplification is to make the sizes of flows identical, or chose them randomly from a uniform distribution; similarly for the starting time of each flow.
-Finally, when choosing endpoints for a given flow, lacking a better model, it is not unusual to treat all nodes identically, making each an equally likely source or destination for a given flow.
-
-%It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees.
-This is largely because of the high complexity of doing otherwise and the lack of compelling alternatives.
-Here we examine how various notions of uniformity are expressed in the matrix model.
-In general, \textit{intra-flow uniformity} is effected by right multiplication, which mixes the elements in each row, whereas \textit{inter-flow uniformity} is effected by left multiplication, which mixes elements across rows.
-The constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies.
-The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}.
-The pair of matrices $\tuple{\mat{Z},\mat{V}}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\tuple{\mat{Z},\mat{V}}\Gamma\in\R^{f \times 3}$, where
-%\begin{align}
-%\Gamma=\begin{pmatrix}
-%\ones{} & \vec{\sigma} & \zeros{} \\
-%\zeros{} & \zeros{} & \vec{\tau}
-%\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
-%\end{align}
 
@@
 
-Marginality is an orthogonal concept to uniformity: the distribution of some property is identical \textit{across} all flows.
-For example, if we determine a distribution of flow sizes across the entire network and assume that this distribution applies \textit{independent of other flow properties}, this is a marginal model for flow size.
+Marginality is an orthogonal concept to uniformity:
+the distribution of some property is identical \textit{across} all flows.
+For example, if we determine a distribution of flow sizes across the entire network and assume that this distribution applies independent of other flow properties, this is a marginal model for flow size.
 Marginal behavior is effected by left multiplication by a matrix with uniform columns.
-This specific example is generated by left-uniformizing the flow size matrix: $\mat{B}' = (\ones{f \times f}/f)\mat{B}$.
-The traffic models proposed by {\FHC} are all marginal: they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node.
+This specific example is generated by left-uniformizing the flow size matrix:
+$\mat{B}' = \smallfrac{1}{f}\ones\mat{B}$.
+The traffic models proposed by {\FHC} are all marginal:
+they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node.
 These models do not account for non-independence between these and other flow properties.
@@
 The final form of homogenization we consider, is conditionality.
-This is a restricted form of marginality: we may assume that marginal distributions apply to all the flows in certain groupings.
+This is a restricted form of marginality:
+we may assume that marginal distributions apply to all the flows in certain groupings.
 For example, we can provide a distribution of flow sizes on a per-source basis.
@@
 This is simply accomplished by left-multiplication by a non-uniform matrix.
-For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-conditioning'' matrix: $\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$.
-Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-conditioning matrix: $F'=(\trans{\mat{DD}}/f)\mat{F}$.
-%Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified.
-The confusion arises due to the lack of a unique way of associating flows with nodes.
-Each node can equally viably be associated with its source or its destination.
-Which association is more valid?
-Consider a web request: should the behavior of the request be determined by what the client typically does?
-Or should the requests that the server typically receives determine behavior?
-There is no armchair answer: only analysis of real data can give us the answer.
-The matrix model provides an analytical framework for doing so.
-The matrix model has the benefit of making the conditioning of behaviors completely explicit.
+For example, to make flow behavior uniform per source, consider the ``source-conditioning'' matrix:
+$\trans{\mat{S}}\mat{S} \in \R^{n \times n}$.
+This matrix has $(\trans{\mat{S}}\mat{S})_{ij} = 1$ if and only if flows $i$ and $j$ have the same source node.
+To condition flow behavior by source, we left-multiply by the stochasticized source-condtioning matrix:
+\begin{align}
+  \mat{F}'= \rStoch(\trans{\mat{S}}\mat{S})\mat{F}.
+\end{align}
+In this case, rows do not have the same sums, so the ``$\rStoch$'' operator is not equivalent to scalar multiplication.
+Per destination uniformity can be achieved similarly, replacing $\mat{S}$ with $\mat{D}$.
+
+The \caps{GMM} has the benefit of making the conditioning of behaviors completely explicit.
 It also allows us to easily define intermediate choices.
@@
 \begin{align}
-\mat{F}' = \frac{1}{f}((1-\alpha)(\trans{\mat{SS}})+\alpha(\trans{\mat{DD}}))\mat{F}.
+\mat{F}' = ((1-\alpha)\,\rStoch(\trans{\mat{S}}\mat{S})+(\alpha)\,\rStoch(\trans{\mat{D}}\mat{D}))\mat{F}.
 \end{align}
@@
 Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.
-
-%We use the terms here to distinguish three fundamentally different types of assumptions about behavioral homogeneity.
-
-%Existing traffic models are typically not explicit methods of generating complete synthetic traces of network traffic, but rather descriptions of some aspect of of the behavior of idealized traffic.
-In our algebraic modeling framework, such models can be interpreted as stipulations about the form of the behavior matrices, or as transforms that implement those stipulations given a real set of traffic matrices.
-
-%A generic traffic model is a way of generating the six matrices described in \Section{representations}.
-A trace-derived traffic model transforms the matrices for an actual trace of network behavior into a simplified or conditional form.
-For experimental comparison, the dimensions of the matrices should all be preserved.
-If a trace-derived model produces experimental performance metrics that are similar to those produced by the original trace data, then simplifications made by the model are viable; if the performance is distorted, then the simplifications are discarding essential information about the traffic pattern.
-
-%\subsection{Uniformity}
-
-%% TODO: differentiate between restrictive versions of models and transformative versions.
-%% TODO: distinguish uniform, marginal and conditional models.
-
-%\subsection{Marginality}
-
-%Most approaches to traffic modeling that go beyond uniformity assumptions, instead describe marginal properties of various aspects of traffic behavior.
-For example, {\FHC}~\cite{Hernandez06:spatio-temporal-modeling} provide highly convincing parametric models for six marginal properties of wireless network traffic: session (node) arrivals follow a time-varying Poisson process; per-session flow inter-arrival times are Lognormally distributed; flows per session (node) follow a BiPareto distribution; flow sizes follow a BiPareto distribution.
-This is certainly much greater detail than simply assuming that all nodes and flows are identical.
-
-%How do marginal models translate into the algebraic language of the general matrix model? Whereas uniformity assumptions for nodes,
-
-%Another benefit of the matrix model is that all aspects of network behavior are determined concurrently.
-We are not left wondering what to do about source and destination nodes or flow start and end times after we have computed our simplified packet behaviors: the same transformation that average the packet behaviors also naturally averages the start and
 
@@
 \end{itemize}
-Despite these remarks, we will show in our experimental evaluation, that this procedure generates sufficiently realistic wireless workloads.
+Despite these remarks, we will show in our experimental evaluation, that this procedure generates \emph{sufficiently realistic} wireless workloads~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
 

### 2008-06-13 tables/traffic_models.tex
Lots of work on Common Assumptions section.

@@
 time uniform &
-$\mat{A}'=\mat{A}(\ones{d_t \times d_t}/d_t)$ &
+$\mat{A}'=\mat{A}(\ones/d_t)$ &
 all start times are equally likely
@@
 node uniform &
-$\mat{S}'=\mat{S}(\ones{n \times n}/n)$~~~~$\mat{D}'=\mat{D}(\ones{n \times n}/n)$ &
+$\mat{S}'=\mat{S}(\ones/n)$~~~~$\mat{D}'=\mat{D}(\ones/n)$ &
 nodes behave stochastically identically with respect to flows
@@
 time marginal &
-$\mat{A}'=(\ones{f \times f}/f)\mat{A}$ &
+$\mat{A}'=(\ones/f)\mat{A}$ &
 start times chosen from marginal distribution of start times
@@
 size marginal &
-$\mat{B}'=(\ones{f \times f}/f)\mat{B}$ &
+$\mat{B}'=(\ones/f)\mat{B}$ &
 flow size (bytes) is chosen independently of other properties
@@
 node marginal &
-$\mat{S}'=(\ones{f \times f}/f)\mat{S}$~~~~$\mat{D}'=(\ones{f \times f}/f)\mat{D}$ &
+$\mat{S}'=(\ones/f)\mat{S}$~~~~$\mat{D}'=(\ones/f)\mat{D}$ &
 nodes behave distinctly, but behavior is independent of flows
@@
 packet marginal &
-$\mat{Z}'=(\ones{f \times f}/f)\mat{Z}$~~~~$\mat{V}'=(\ones{f \times f}/f)\mat{V}$ &
+$\mat{Z}'=(\ones/f)\mat{Z}$~~~~$\mat{V}'=(\ones/f)\mat{V}$ &
 packet behavior is variable bit-rate, aggregated across all flows

### 2008-06-13 paper.tex
Lots of work on Common Assumptions section.

@@
 It also allows us to easily define intermediate choices.
-If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
+If we wish make the behavior of each flow a weighted average of the composite behaviors of its source and destination nodes, we can express this easily:
 \begin{align}

### 2008-06-13 paper.tex
Applied all of E's changes, except where noted.

@@
 \newcommand{\thetitle}{Linear Representation of Network Traffic}
+\newcommand{\thesubtitle}{With Applications to Generating Realistic Wireless Workload}
 \newcommand{\theauthors}{
@@
 
-%\usepackage{type1cm}
-%\usepackage[pdftex]{graphicx}
 \usepackage[labelfont=bf,small]{caption}
 \usepackage[font=small,labelfont=bf,position=top,nearskip=0em]{subfig}
-\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize,paralist,array}
-\usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle}]{hyperref}
-\usepackage{mathtools}
+\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize,paralist,array,mathtools}
+\usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle: \thesubtitle}]{hyperref}
 
 \title{\thetitle}
+\subtitle{\thesubtitle}
 \author{\theauthors}
@@
 We use the same experimental methodology to explore the implications of various assumptions and simplifications that are commonly made in traffic modeling.
-%We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models.
-A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
-%
-%We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in wireless networks.
-Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematical precision and succinctness, while simultaneously being made clearer and more general.
-We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to discredit the common constant bit-rate model.
-A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
-We conclude with future directions for modeling research based on the new algebraic framework.
 \end{abstract}
@@
 \section{Introduction}\label{sec:intro}
+
+% TODO: add wireless application into this.
 
@@
 
-% In short, the reason is that these assumptions drastically simplify both traffic analysis and workload generation.
-% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
-% This immediately begs the question of whether these assumptions are benign.
-% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to analyze and replicate?
-% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
-
-% TODO: is this paragraph necessary?
-% The second reason why uniformity and marginality assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
-% The difficulty lies in the high-dimensional and stochastic nature of network behaviors.
-% Suppose, for example, that we assume that packet sizes and inter-packet intervals are drawn independently from some given distribution for each flow.
-% In the terminology introduced above, this is a flow-conditional packet behavior model.
-% Prior research has indicated that, given accurate per-flow distributions, and keeping all other aspects of network behavior fixed, this simplification does not distort experimental performance metrics~\cite{Karpinski07:realism}.
-% In this model, however, each flow must have its own specific, realistic pair of distributions for packet size and inter-packet intervals.
-% Moreover, the ``distribution of distributions'' across the entire network must simultaneously be realistic.
-% Modeling such distributions of distributions realistically for entire networks is currently an unsolved problem; we begin to address it here.
-
 If no one actually believes that assumptions of uniformity and marginality accurately describe network behavior, then why are these assumptions so common?
@@
 The first is that these assumptions drastically simplify both traffic analysis and workload generation, making them tractable problems.
-% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
-% This immediately begs the question of whether these assumptions are benign.
-% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to understand?
-% This paper answers that question.
 The second is that there are no generally accepted alternatives for representing and simplifying traffic patterns that allow effective analysis and generation without assuming uniformity or marginality.
-% Representing complex multi-level network behaviors in a way that does not make these simplifying assumptions but allows effective traffic analysis and workload generation is currently an unsolved problem;
-% we begin to address it here.
-
-% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.
 
@@
 using our solution to address the other two questions.
-% transforming the unreduced representation of traffic into it's uniform, marginal or conditional counterparts.
 
@@
 These transformations, however, are extremely destructive: the matrices are degenerate with very low rank, so multiplication destroys almost all information in the original traffic matrix.
-We demonstrate experimentally, that this loss of information is not benign:
-uniform and marginal traffic models severely distorting crucial network performance metrics.
+We demonstrate experimentally that this loss of information is not benign:
+uniform and marginal traffic models severely distort crucial network performance metrics.
 
@@
 
-While this work is not inherently specific to wireless networks, since the modeling techniques developed are equally applicable to traffic in wired networks, the results are of special importance to wireless research.
+While our work is not inherently specific to wireless networks, since the modeling techniques developed are equally applicable to traffic in wired networks, the results are of special importance to wireless research.
 Realistic traffic models have a drastic impact on experimental results in wireless networks~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
@@
 In wireless networks, on the other hand, the physical medium cannot be upgraded; protocols must be improved to make better usage of the medium.
-Being able to perform experiments with realistic workload is a vital part of the ongoing process of improving wireless protocols.
+The ability to perform experiments with realistic workload is a vital part of the ongoing process of improving wireless protocols.
 Without realistic workload models for experiments, performance comparisons may be misleading or even completely wrong~\cite{Karpinski07:cbr-failure}.
@@
 Hern\'andez-Campos' dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and typical \caps{TCP} applications.
-He shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
+The author shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
 This research has subsequently been turned into a \caps{TCP} flow generation tool called \tmix~\cite{Weigle06}.
@@
 its start time;
-and the specific sizes packets of and intervals between their transmission.
+and the specific sizes of packets and intervals between their transmission.
 We begin our exposition of traffic representation by formalizing these properties mathematically as sets:
@@
 \newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $\pkts$ does not divide $\mathsym{bytes}$ evenly.
-The simplest is to ignore it; we leave leave more complex schemes to the reader.}
+The simplest is to ignore it; we leave more complex schemes to the reader.}
 
@@
 The \caps{PBD} representation is highly lossy: it discards most of the information detailing the behavior of each flow.
-At the other extreme, we can give a ``perfect'' representation of each flow, that allows us to completely reconstruct each flow from its representation.
+At the other extreme, we can give a ``perfect'' representation of flows, that allows us to completely reconstruct each flow from its representation.
 With care, the sum of representations of flows can still naturally describe their aggregate behavior.
@@
 
-The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins---%
-one per value for properties that are discrete already and require no compression---%
+The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins
+(one per value for properties that are discrete already and require no compression),
 and map values in each bin to a standard basis element of $\R^d$, where $d$ is the number of bins.
@@
 
-The final form of homogenization we consider, is conditionality.
+The final form of homogenization we consider is conditionality.
 This is a restricted form of marginality:
@@
 \item Because the representation $\phi$ models the seven properties independent of each other, it is impossible to do better than to sample the properties independently, modulo the requirements of the procedure.
-In the case of a row which is simple a representation of an actual flow, this makes no difference, except for the generation of packets (which are modeled independently).
+In the case of a row which is simply a representation of an actual flow, this makes no difference, except for the generation of packets (which are modeled independently).
 In the case of composite rows, however, this implies that flow properties are modeled independent of each other.
@@
 The traffic trace was captured using \texttt{\small{tcpdump}} at a single router, through which all wireless traffic for the meeting was routed, including traffic between wireless nodes.
-The snap length of the capture was 100 bytes, allowing \caps{IP}, \caps{ICMP}, \caps{UDP} and \caps{TCP} headers to be analyzed.
+The snap length of the capture for each packet was 100 bytes, allowing \caps{IP}, \caps{ICMP}, \caps{UDP} and \caps{TCP} headers to be analyzed.
 We limit our work to the 24-hour sub-trace recorded on Wednesday, August 4th.
@@
 Next, the quantity of application-initiated data contained in each packet is calculated.
-For non-\caps{TCP} packets, this quantity is simply the size of the transport-layer payload, but for \caps{TCP} the calculation is more complicated: only new data transfers, explicitly initiated by the application are counted.
+For non-\caps{TCP} packets, this quantity is simply the size of the transport-layer payload, but for \caps{TCP} the calculation is more complicated: only new data transfers, explicitly initiated by the application, are counted.
 Data retransmitted by \caps{TCP} is disregarded, and empty \caps{ACK}s are ignored.
@@
 We use the Qualnet wireless network simulator (version 4.0.1) to perform our experiments.
-We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
+We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters but a radio range of only 20 meters---both in accordance with an indoor network environment.
 In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays.
@@
 Moreover, raw \caps{IP}, \caps{UDP} and \caps{TCP} flows are implemented using the same code, guaranteeing that all traffic is simulated uniformly and that all performance metrics are aggregated in the same manner.
-We have also instrumented Qualnet to collect \caps{IP} performance metrics in strict adherence to the recommendations of the \caps{IETF} working group on \caps{IP} performance metrics~\cite{rfc:ip-metrics}.
+% We have also instrumented Qualnet to collect \caps{IP} performance metrics in strict adherence to the recommendations of the \caps{IETF} working group on \caps{IP} performance metrics~\cite{rfc:ip-metrics}.
 
@@
 
+% TODO: remove mixed conditional from plots.
 \input{figures/box_plots}
@@
 {\FHC} found strong evidence that session arrivals followed a time-varying Poisson arrival process, meaning that other than long-scale time-varying arrival rate and very small-scale clustering of flows within sessions, the overall flow arrival rate is fairly smooth.
-We suspect that on the scale of minutes to tens of minutes in which our simulation exist, it is sufficient to model both flow and session arrivals uniformly.
+We suspect that on the scale of minutes to tens of minutes in which our simulations exist, it is sufficient to model both flow and session arrivals uniformly.
 This result indicates that the precise temporal placement of flows is not highly sensitive: they tend to be relatively evenly spaced out at short time-scales, and network performance is not sensitive to changes in start-time on that scale.
@@
 This is a highly significant result because modeling flow size using a network-wide marginal distribution is precisely what {\FHC}, for example, attempt.
-This results informs us that this approach is doomed to failure: flow size cannot be assigned without considering its relation to other flow properties: source and destination nodes, and packet behavior at the very least.
+This result informs us that this approach is doomed to failure: flow size cannot be assigned without considering its relation to other flow properties: source and destination nodes, and packet behavior at the very least.
 
@@
 And yet these drastic distortions occur.
-In our previous work, we have shown similarly drastic misrepresentations to occur when constant bit-rate packet behavior is assumed.
-This results indicates that to reproduce accurate network performance, we must provide packet behaviors on a more granular level: each node or even each flow must be assigned a ``custom'' packet behavior by some means.
+In our previous work, we have shown similarly drastic misrepresentations to occur when \caps{CBR} packet behavior is assumed~\cite{Karpinski07:cbr-failure}.
+This result indicates that to reproduce accurate network performance, we must provide packet behaviors on a more granular level: each node or even each flow must be assigned a ``custom'' packet behavior by some means.
 
@@
 These demonstrate neither exceptionally good nor exceptionally bad behavior.
-That the conditional models should be similar to the \model{full marginal} model but somewhat more realistic is entirely expected: conditioning, the reader will recall, is a restricted form of marginal modeling, where each group of flows has its own marginal distribution.
+That the conditional models should be similar to the \model{full marginal} model but somewhat more realistic is entirely expected: conditioning is a restricted form of marginal modeling, where each group of flows has its own marginal distribution.
 In these cases, the flows are grouped by source node or by destination node.
@@
 The general matrix model entails a specific linear representation, such that network traffic patterns can easily be expressed as matrices.
-Our experimental results show that the \caps{GMM} is sufficiently realistic for generating wireless workload derived from traces.
+Our experimental results show that the \caps{GMM} is sufficiently realistic for generating wireless workloads derived from traces.
 When uniform or marginal assumptions are applied, however, sufficient realism eludes us, indicating that if we are to simplify and reduce the representation of network behaviors, we must do some through some other technique.
@@
 These transformations destroy almost all of the information in the original traffic matrix---only the zero matrix has lower rank than these.
+
+\input{figures/basic_behaviors}
 
@@
 The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all \caps{SSH} traffic, interacting with a remote terminal via keystrokes and screen updates.
-
-\input{figures/basic_behaviors}
 
@@
 Our experimental validation demonstrates clearly that the general matrix model accurately reproduces the performance characteristics of of real traffic.
+
 The benefits of a performance-preserving algebraic representation are multifold.
@@
 % Hernandez06:dissertation: ``complete methodology for reproducing the traffic observed on a network link in a closed-loop manner, and proposed a number of metrics for studying the realism of the generated traffic.''
+
+\vfill
+\pagebreak
 \bibliographystyle{plain}

### 2008-06-14 paper.tex
Work on discussion.

@@
 
-% The model of flow behavior implicit in non-negative matrix factoriation technique is that flow behavior is a composite of some limited set of basic behaviors---and that fractions of each flow ``belong'' to each behavior.
+% The model of flow behavior implicit in nonnegative matrix factoriation technique is that flow behavior is a composite of some limited set of basic behaviors---and that fractions of each flow ``belong'' to each behavior.
 This is not an \latin{a priori} unreasonable hypothesis.
@@
   linear algebra \and
-  non-negative matrix factorization
+  nonnegative matrix factorization
 }
@@
 We argue that matrix factorization, which is nondestructive and does not assume uniformity or marginality, is far superior to multiplication for this problem.
-To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF})~\cite{Lee01} to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
+To demonstrate this approach, we use nonnegative matrix factorization (\caps{NMF})~\cite{Lee01} to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
 This factorization extracts a small set of ``basic behaviors'' from the traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
@@
 Our experimental results show that the \caps{GMM} is sufficiently realistic for generating wireless workloads derived from traces.
-When uniform or marginal assumptions are applied, however, sufficient realism eludes us, indicating that if we are to simplify and reduce the representation of network behaviors, we must do some through some other technique.
+When uniform or marginal assumptions are applied, however, sufficient realism eludes us, indicating that if we are to simplify and reduce the representation of network behaviors, we must do so by some other technique.
 In \Section{modeling-concepts}, we observed that application of uniformity and marginality assumptions to traffic instances is equivalent, in the \caps{GMM}, to right and left multiplication, respectively, by matrices of rank one.
-These transformations destroy almost all of the information in the original traffic matrix---only the zero matrix has lower rank than these.
+These transformations destroy almost all of the information in the original traffic matrix.
+Conditional models are equivalent to left-multiplication by somewhat less degenerate matrices, but they achieve realism only by approaching the \caps{GMM} itself.
 
@@
 
-The question that remains to be answered is how we can simplify and reduce the representation of network behaviors without making unwarranted and intuitively clearly false assumptions of uniformity of marginality of network behaviors.
-We present here preliminary results from applying non-negative matrix factorization to the packet behaviors of collections of flows.
-Recall from \Section{general-matrix-model}, that the matrices $Z$ and $V$ represent the packet size and inter-packet interval histograms of a collection of flows, with each row corresponding to a flow, and each column corresponding to a packet size or a range of inter-packet intervals.
-We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}]$.
-\caps{NMF} attempts to approximate $\mat{P}$ as the product of three non-negative matrices:
+It remains to show how we can simplify and reduce the representation of network behaviors without making unwarranted and manifestly false assumptions of uniformity of marginality of network behaviors.
+We present here preliminary results from applying nonnegative matrix factorization (\caps{NMF}) to the packet behaviors of real flows.
+Recall from \Section{general-matrix-model}, that the matrices $\mat{Z}$ and $\mat{V}$ represent the packet size and inter-packet interval histograms of a collection of flows, with each row corresponding to a flow, and each column corresponding to a packet size or a range of inter-packet intervals.
+We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}] \in \R^{f \times (d_z+d_v)}$.
+\caps{NMF} attempts to approximate $\mat{P}$ as the product of three nonnegative matrices:
 \begin{align}
@@
 \end{align}
-where $\mat{W} \in \R^{f \times K}$ is row-stochastic,
-$\mat{\Sigma} \in \R^{K \times K}$ is diagonal with descending diagonal entries, and
-$\mat{H} \in \R^{K \times (d_z+d_v)}$ is row-stochastic as well.
-$K\in\N$ is the rank of this factorization, which must be determined somehow as well.
-What is the interpretation of such factorization?
+where $\mat{W} \in \R^{f \times b}$ and $\mat{H} \in \R^{b \times (d_z+d_v)}$ are both row-stochastic, while
+$\mat{\Sigma} \in \R^{b \times b}$ is diagonal with descending diagonal entries.
+$b\in\N$ is the rank of this factorization, which must be determined somehow as well.
+What is the interpretation of such a factorization?
 The rows of $\mat{H}$ are \emph{basic behaviors}---pairs of packet size and inter-packet interval distributions, such that each flow's packet behavior can be approximated as a linear combination of these basic behaviors.
@@
 Thus the basic behaviors are sorted in descending order of how much traffic they explain.
-Note that the factorization does not associate basic behaviors with individual packets;
-rather, total flow behavior is explained as a mixture of basic behaviors, so the fraction of packets in each flow explained by a given behavior is known, while the specific packets are unknown.
-
-The striking results of applying non-negative matrix factorization to $\mat{P}$ are shown in \Figure{basic-behaviors}.
+% Note that the factorization does not associate basic behaviors with individual packets;
+% rather, total flow behavior is explained as a mixture of basic behaviors, so the fraction of packets in each flow explained by a given behavior is known, while the specific packets are unknown.
+
+The striking results of applying nonnegative matrix factorization to $\mat{P}$ are shown in \Figure{basic-behaviors}.
 Twelve basic behaviors suffice to explain all the traffic in a ``toy''  sample of 4138 flows.
@@
 This is what we would expect from downloading large files; and indeed, the breakdown of packets by port number bears out this interpretation---almost all of the traffic is \caps{HTTP} server traffic, with small slivers of \caps{IMAP} and \caps{SSH} server traffic.
-The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all \caps{SSH} traffic, interacting with a remote terminal via keystrokes and screen updates.
+The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all \caps{SSH} traffic, as indicated by the associate pie chart.
+This is precisely the behavior we expect from interacting with a remote terminal via keystrokes and screen updates.
+
+% TODO: say something here...
 
@@
 Our experimental results bolster this conclusion: both uniform and marginal models significantly misrepresent many important network performance metrics.
-With the general matrix model, however, we stand poised to use the powerful tools of modern linear algebra to develop general, elegant and powerful models of network behavior that just work.
+With the general matrix model, however, we stand poised to use the powerful tools of modern linear algebra to develop general, elegant and effective models of network behavior.
 

### 2008-06-14 paper.tex
New subtitle.
More work on appearance of models table.

@@
 \newcommand{\thetitle}{Linear Representation of Network Traffic}
-\newcommand{\thesubtitle}{With Applications to Generating Realistic Wireless Workload}
+\newcommand{\thesubtitle}{With Special Application to Wireless Workload Generation}
 \newcommand{\theauthors}{
@@
 \usepackage[font=small,labelfont=bf,position=top,nearskip=0em]{subfig}
-\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize,paralist,array,mathtools}
+\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize,paralist,array,mathtools,units}
 \usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle: \thesubtitle}]{hyperref}
@@
 \end{align}
-The symbol $\eldiv$ here represents element-wise division of vector coordinates, as it does in Matlab.
+The symbol $\eldiv$ here represents element-wise division of vectors, as it does in Matlab.
 
@@
 The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors.
-Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.
+Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\nicefrac{1}{2}$.
 

### 2008-06-14 tables/traffic_models.tex
More work on appearance of models table.

@@
 time uniform &
-$\mat{A}'=\mat{A}(\ones/d_t)$ &
+$\mat{A}' = (\nicefrac{1}{n})\,\mat{A}\ones$ &
 all start times are equally likely
@@
 node uniform &
-$\mat{S}'=\mat{S}(\ones/n)$~~~~$\mat{D}'=\mat{D}(\ones/n)$ &
-nodes behave stochastically identically with respect to flows
+$\mat{S}' = (\nicefrac{1}{n})\,\mat{S}\ones \quad
+ \mat{D}' = (\nicefrac{1}{n})\,\mat{D}\ones$ &
+all nodes equally likely as source or destination
 \\\hline
@@
 \textit{all uniform transformations} &
-applies all the uniformity assumptions of the above models
+all start times, sources and destinations equally likely
 \\\hline\hline
 time marginal &
-$\mat{A}'=(\ones/f)\mat{A}$ &
+$\mat{A}' = (\nicefrac{1}{f})\,\ones\mat{A}$ &
 start times chosen from marginal distribution of start times
@@
 size marginal &
-$\mat{B}'=(\ones/f)\mat{B}$ &
+$\mat{B}' = (\nicefrac{1}{f})\,\ones\mat{B}$ &
 flow size (bytes) is chosen independently of other properties
@@
 node marginal &
-$\mat{S}'=(\ones/f)\mat{S}$~~~~$\mat{D}'=(\ones/f)\mat{D}$ &
+$\mat{S}' = (\nicefrac{1}{f})\,\ones\mat{S} \quad
+ \mat{D}' = (\nicefrac{1}{f})\,\ones\mat{D}$ &
 nodes behave distinctly, but behavior is independent of flows
@@
 packet marginal &
-$\mat{Z}'=(\ones/f)\mat{Z}$~~~~$\mat{V}'=(\ones/f)\mat{V}$ &
+$\mat{Z}' = (\nicefrac{1}{f})\,\ones\mat{Z} \quad
+ \mat{V}' = (\nicefrac{1}{f})\,\ones\mat{V}$ &
 packet behavior is variable bit-rate, aggregated across all flows
@@
 source conditional &
-$\mat{F}'=(\trans{\mat{SS}}/f)\mat{F}$ &
+$\mat{F}' = \rStoch(\trans{\mat{S}}\mat{S})\,\mat{F}$ &
 flow behaviors are aggregated on a per source-node basis
@@
 dest conditional &
-$\mat{F}'=(\trans{\mat{DD}}/f)\mat{F}$ &
+$\mat{F}' = \rStoch(\trans{\mat{D}}\mat{D})\,\mat{F}$ &
 flow behaviors are aggregated on a per destination-node basis
@@
 mixed conditional &
-$\mat{F}'=\frac{1}{2f}(\trans{\mat{SS}}+\trans{\mat{DD}})\mat{F}$ &
+$\mat{F}' = \nicefrac{1}{2}\,(
+              \rStoch(\trans{\mat{S}}\mat{S}) +
+              \rStoch(\trans{\mat{D}}\mat{D})
+            )\,\mat{F}$ &
 flow behaviors are aggregated on a mixed source/destination basis
@@
 \end{center}
-\vspace{-2em}
+\vspace{-1em}
 \end{table*}

### 2008-06-15 paper.tex
Work on results.

@@
 These metrics are commonly used to evaluate wireless protocols.
-We have examined a broad variety of other wireless network performance metrics, and although we do not have room to present or discuss them here, the results shown are representative of the overall realism of the traffic models.
+We have examined a broad variety of additional wireless network performance metrics, and although we do not have room to present or discuss them here, the results shown are representative of the overall realism of the traffic models.
 
@@
 It is worth noting that \caps{GMM} has competition from a few of the time-simplified models: the \model{time marginal} and \model{time uniform} model both do approximately as well, and in some cases better, for the metrics examined.
-{\FHC} found strong evidence that session arrivals followed a time-varying Poisson arrival process, meaning that other than long-scale time-varying arrival rate and very small-scale clustering of flows within sessions, the overall flow arrival rate is fairly smooth.
+{\FHC} found strong evidence that session arrivals followed a time-varying Poisson arrival process, meaning that other than large scale time-vary\-ing arrival rate and very small scale clustering of flows within sessions, the overall flow arrival rate is fairly smooth.
 We suspect that on the scale of minutes to tens of minutes in which our simulations exist, it is sufficient to model both flow and session arrivals uniformly.
@@
 That the conditional models should be similar to the \model{full marginal} model but somewhat more realistic is entirely expected: conditioning is a restricted form of marginal modeling, where each group of flows has its own marginal distribution.
-In these cases, the flows are grouped by source node or by destination node.
+In these cases, the flows are grouped by source or by destination.
+Neither source nor destination conditioning appears to out-perform the other:
+they perform identically in each case;
+neither is perfect, but both are better in general than any marginal or uniform model.
+The node-conditional models tend to err in the same direction as the \model{full marginal} model, but in a few cases where the aspects of marginality lead to errors in different directions, the outcome follows one of the aspects more than others.
+In the case of received throughput, for example, the \model{packet marginal} tendency to overestimate error overshadows the tendency of the other marginal aspects to underestimate.
+
+% TODO: rerun the mixed model simulations.
 

### 2008-06-15 paper.tex
Added basic behaviors table; sending to John.
Miscellaneous tweaking.

@@
 provides an appropriate description of the aggregate behavior of the two flows together.
-Similarly, their average, \latin{i.e.} $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
+Similarly, their average, \latin{i.e.} $\smallfrac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
 In contrast, a ``packets-payload-interval'' representation, expressing the \caps{CBR} parameters directly, does not have this property:
@@
 These parameters are non-linear scaling exponents that allows the quantization to smoothly shift from fine-grained resolution at the low end of the spectrum to coarse-grained resolution at the high end.
-We have chosen $\beta=\fracx{1}{2.7}$ and $\gamma=\fracx{1}{3}$ because these values yield desirable dynamic ranges: from single bytes at the lower end to megabytes at the high end for flow sizes; from microseconds to seconds for intervals.
+We have chosen $\beta=\nicefrac{1}{2.7}$ and $\gamma=\nicefrac{1}{3}$ because these values yield desirable dynamic ranges: from single bytes at the lower end to megabytes at the high end for flow sizes; from microseconds to seconds for intervals.
 
@@
 \vec{\sigma} &= \tuple{1,2,3,\dots,d_z} \\
-\vec{\tau}   &= \tuple{\Df_\ival\left(k-\fracx{1}{2}\right)}_{k=1}^{d_v}.
+\vec{\tau}   &= \tuple{\Df_\ival\left(k-\smallfrac{1}{2}\right)}_{k=1}^{d_v}.
 \end{align}
@@
 The ``$\rStoch$'' operator scales each row to sum to unity, thus making the entire matrix row-stochastic;
-in this case, this is equivalent to scaling by $\smallfrac{1}{n}$.
+in this case, this is equivalent to scaling by $\nicefrac{1}{n}$.
 These transformations of $\mat{S}$ and $\mat{D}$ make all the values in each flow's row identically the average of the original $n$ values.
@@
 This specific example is generated by left-uniformizing the flow size matrix:
-$\mat{B}' = \smallfrac{1}{f}\ones\mat{B}$.
+$\mat{B}' = (\nicefrac{1}{f})\,\ones\mat{B}$.
 The traffic models proposed by {\FHC} are all marginal:
@@
 These models do not account for non-independence between these and other flow properties.
-We will discuss this limitation and its implications when we analyze our experimental results.
+We discuss this limitation and its implications when we analyze our experimental results.
 
@@
 The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all \caps{SSH} traffic, as indicated by the associate pie chart.
-This is precisely the behavior we expect from interacting with a remote terminal via keystrokes and screen updates.
-
-% TODO: say something here...
-
-% What are the ramifications of these results?
-In our previous work, we had already demonstrated the drastic distortions that are introduced by uniformity assumptions made by the most common traffic models.
-This work serves to further critique established common assumptions in traffic modeling.
-The strongest message to be taken away is that the aspects of traffic behavior of both nodes and flows \textit{must} be considered jointly.
-It is not sufficient to continue modeling marginal distributions of network properties.
-These marginal models do not lend them selves easily to traffic generation in the first place, and when marginality assumptions are applied to trace traffic, the result is a severe distortion of important network performance characteristics.
-It is now well established that network traffic patterns have an impact on network performance that cannot be ignored.
-Even experimental deployments cannot avoid the need for more realistic traffic workload models.
-While using a real, physical network successfully sidesteps simulation problems below the application layer, without realistic traffic models, reliable, meaningful performance predictions remain beyond our reach.
-
-%\subsection{Generality of Results}
-
-%% TODO: talk about other synthetic models not considered here.
-%% The point of our analysis is not to show that no synthetic models can be accurate, but rather, that the commonly used models, and obvious variations thereof, are not accurate.
-
-%The most significant limitation on the generality of this analysis is that it is based entirely on a single data set from \caps{IETF}60---albeit a large and varied one.
-It is possible that traffic in this trace happens to produce network performance that is unusually dissimilar to standard traffic models.
-This data set, however, represents a highly heterogeneous collection of network usage behaviors, from slow and steady off-peak usage, to extremely heavy peak usage: over 800 users, 33 thousand flows, and 1 million packets in a single 10-minute trace segment.
-Despite the broad variety of behaviors, the results are consistent: in all types of usage scenarios, simplistic traffic models, like uniform \caps{CBR}, systematically skew important performance measurements at all levels of the network.
-While the precise results for other data sets might differ, it is very unlikely that \caps{CBR} traffic models will happen to accurately reproduce realistic performance in other experiments.
-This paper provides strong evidence that better traffic models are needed for performance evaluations.
-
-% \subsection{Towards Realistic Models of Wireless Workload}
-% 
-% What would better traffic models look like?
-How can we create them?
-One possible approach is to use actual traffic traces as we have done.
-This approach is unsatisfactory, however, because it provides the experimenter with almost no control over experiments.
-Synthetic models have parameters, which can be tweaked as necessary---adjusting, for example, the number of active nodes in a simulation, without affecting other parameters.
-Traffic traces, on the other hand, must be used without significant alteration if they are to actually provide the desired realism.
-This inability to control all but one or two parameters is nearly crippling for experimental purposes.
-Therefore, the research community needs models whose parameters can be adjusted independently, but whose performance characteristics can be made to resemble whatever example of network behavior one might need.
-The general matrix model does not yet fulfill this role since its parameters cannot be freely varied.
-We believe, however, that \caps{GMM} plays a vital role in the ongoing process of developing  of better models.
-By reducing the general problem of traffic modeling to the domain of linear algebra, we have provided traffic researchers with a vast array of powerful theoretical and computational tools that can be immediately applied to the problem at hand.\footnote{The derived models used in this paper were all implemented in Matlab, each in only a few simple lines.
-All the models used here are implemented in only 19 total lines of code.
-Of course, the framework to apply them to traffic traces and simulations is somewhat larger.} What are the next steps?
-The authors are currently investigating a variety of matrix factorization and clustering techniques to extract hidden structure from a large and diverse body of real traffic.
-
-%The reader will note that the transformations applied in this paper are all quite simple.
-Our conditioning matrices primarily consisted of ones and zeros.
-The algebraic representation of complete workload
-
-%The ``messiness'' of the performance comparison from trace data in Figure~\ref{fig:mac-control-overhead-cmp} illustrates why using traces directly is not ideal: each data point differs not only in the number of nodes shown on the $x$-axis, but also in other dimensions, such as the number of flows and packets, and the average flow duration.
-The result is a highly noisy comparison, affected by many unseen parameters.
-Only by applying a local smoothing algorithm are trends somewhat elucidated.
-
-%Instead of using trace data directly, it should be possible to configure a synthetic traffic model based on observations from a real data set, and then run side-by-side simulations using the synthetic model and the real data, producing statistically equivalent performance results.
-This is precisely what our definition of sufficient realism entails.
-The work in this paper provides the tools to measure how close to this ideal a model is and in what areas it needs improvement.
-Without this feedback, any improvements in realism are purely guesswork.
-Our breakdown of traffic behavior into three orthogonal levels also allows the problem to be approached in smaller pieces, rather than being solved all at once.
-
-%The next step towards better traffic models is to investigate which aspects of real traces may be altered without detrimentally affecting the resulting performance metrics.
-For example, to test whether a complex time-series model of packet behavior is necessary, we randomize the order of the packet sizes and/or inter-packet intervals and compare performance using these randomized traces against performance using the original traces.
-If the performance is unchanged, we can conclude that no complex time-series model of packet behavior is necessary: sampling the packet sizes and inter-packet intervals from empirical distributions is sufficiently realistic.
-If, on the other hand, the performance characteristics are altered by shuffling packets, then some time-series model of packet behavior is needed.
-By partially randomizing the packet order in specific ways, the exact limits of realism necessary can be found.
-A similar approach will allow the development of realistic models for the other levels of network usage behavior.
+The other behaviors are listed in \Table{basic-behaviors} with interpretations.
+% This is precisely the behavior we expect from interacting with a remote terminal via keystrokes and screen updates.
+
+\input{tables/basic_behaviors}
 
 \section{Conclusions}\label{sec:conclusions}
-
-%The real significance of this paper lies not in showing that the commonly used \caps{CBR} traffic model is unrealistic.
-Networking researchers understand that \caps{CBR} is at best a na\"ive approximation of real network traffic.
-Nor is the real significance that we have quantized the high degree of inaccuracy that can be induced by using this model.
-The real significance of this paper lies in providing an well-defined, objective measure of realism for traffic models.
-This has never previously existed: evaluations of realism have relied on essentially arbitrary statistical measures of similarity to real traffic.
-Existing traffic generators use empirical distributions of such quantities as file size, inter-connection time, packet size, or inter-packet intervals.
-They can only hope that reproducing these distributions as good enough.
-Without an external measure of accuracy, there's simply no way to know.
-
-%attempting to showing that existing models are unrealistic, or that new models are more realistic have always focused on arbitrary statistical measures. A
 
@@
 
-%This research rigorously quantifies the impact of a variety of synthetic traffic models on performance metrics that wireless researchers use to evaluate new technologies and protocols.
-The first step in this assessment process was to formally define what it means for a network usage model to be sufficiently realistic.
-In essence, a model is considered realistic if it produces performance results that are statistically equivalent to those produced by real usage.
-%A well-defined, objective measure of realism for traffic models has not previously existed.
-Evaluations of realism have formerly relied on essentially arbitrary statistical measures of similarity to real traffic, which may or may not affect the performance metrics that researchers care about.
-%The definition of sufficient realism leads us to our general experimental approach: we use differential analysis comparing performance metrics derived from real traffic with those derived from synthetic traffic models.
-The theoretical contributions of this analysis are:
-%\begin{enumerate}
-%\item An in-depth analysis of the desirable mathematical properties of a measure of error for performance metrics.
-%\item Proof that the unique measure of error that satisfies these properties is the log-ratio of metric values.
-%\item Three rigorous tests of statistical equivalence between synthetic and real performance results.
-%\end{enumerate}
-%These analytical tools allow the evaluation of realism over a collection of drastically different usage scenarios.
-Evaluation over a heterogeneous collection of scenarios is essential to establishing the credibility of usage models.
-Moreover, these theoretical results are equally applicable to other types of usage models---for example, mobility.
-
-%On the practical side, this paper gives crucial insight into why most researchers do not trust simulation results: with the traffic models commonly used, the results are unlikely to reflect real performance.
-Moreover, this problem will also hamper experiments in test-bed wireless networks, so long as the same na\"ive workload models are used.
-The only way to address this fundamental lack of realism is to develop usage models that reproduce important performance metrics more accurately.
-Our theoretical results provide the tools necessary to do this.
-The development of better traffic models should begin with real traces, and proceed by incremental changes, checked by differential analysis.
-%%First, alter a small aspect of the trace, simulate, then compare.
-If the realism of the results is unaffected, the traffic feature altered was inessential.
-Otherwise, it is a feature of behavior that must be captured in a realistic traffic model.
-%This approach will allow the precise mapping of which aspects of traffic patterns have an impact on performance and which ones can be safely abstracted away.
-%% TODO: maybe cut out some of this last bit.
-
 %\section{Acknowledgments}
@@
 
-% Hernandez06:dissertation: ``complete methodology for reproducing the traffic observed on a network link in a closed-loop manner, and proposed a number of metrics for studying the realism of the generated traffic.''
-
-\vfill
-\pagebreak
 \bibliographystyle{plain}

### 2008-06-25 letter.tex
Letter addressing reviewer comments.

\documentclass{letter}

\oddsidemargin=.2in
\evensidemargin=.2in
\textwidth=5.9in
\topmargin=-.5in
\textheight=9in

\usepackage{cite,amsmath,amssymb,relsize,paralist,mathtools,units}

\input{definitions}

\newenvironment{thebibliography}[1]{%
\section{References}%
\begin{enumerate}%
}{%
\end{enumerate}%
}

\name{%
Stefan~Karpinski\\
Elizabeth~M.~Belding\\
Kevin~C.~Almeroth\\
John~R.~Gilbert%
}

\begin{document}
\begin{letter}{Reviewers of \caps{ACM}/Monet Manuscript \#MONE-95}

\opening{Dear Reviewers,}

Thank you for your thoughtful and helpful reviews of out \caps{ACM}/Monet submission.
From your responses and feedback from other peers, we believe that the first version of our paper did not adequately make clear the motivation, purpose or contribution of our work.
Therefore, we have significantly revised the introductory and motivational material of our paper as well as presentation of the technical concepts regarding the linear representation of network traffic as vectors and matrices.
The simulation results that support our claim that the general matrix model captures enough of the behavior of real traffic to be sufficiently realistic as a basis for generative models of network traffic remains the same except for the \model{mixed conditional} model (previously called \model{mixed regular}).
There was an error in the code generating traffic for this model, which caused the simulation results to be incorrect.
We have corrected this error.
In the new version, we have omitted the discussion of measures of error and statistical tests, which already appeared in our previous work~\cite{Karpinski07:cbr-failure}; it was deemed too far from the central topic of the paper.
Since the utility and significance of the general matrix model was not made clear in our previous version of the paper, we have added in the discussion section preliminary results from using the numerical linear technique of nonnegative matrix factorization on the \caps{GMM} matrices describing the packet behaviors of flows.
We believe these results are very exciting and promising, and indicate the utility and potential of the representing network traffic linearly.
In what follows, we indicate where in the paper revision, we address specific comments, questions and concerns raised by our collective reviewers.

\begin{enumerate}

\item \textbf{Information destruction in uniform, marginal and conditional (n\'e regular) models, seen through linear representation.}
It was mentioned that it was unclear how the transformations allow to understand the effect and drawbacks of each simplified uniform, marginal or conditional model.
In the revised paper, we have observed that the uniform, marginal and conditional transformations are all multiplications by specific low-rank matrices.
Since low-rank matrices are by definition uninvertible, the information content of the original traffic matrix is irrevocably destroyed by these transformations.
The revised paper discusses this issue in several places:
\begin{enumerate}
\item Introduction, page 2, column 2, paragraph 2;
% TODO: add another paragraph about this?
\item Discussion, page 15, column 2, paragraph 1.
\end{enumerate}
In the discussion section, moreover, we propose matrix factorization, rather matrix multiplication as a less destructive and thus more faithful method of reducing the complexity, size and dimension of traffic representations.

\item \textbf{Computational complexity of \caps{GMM} compared with simplified models.}
% TODO: address complexity in discussion.
This issue is now addressed in our discussion section.
We admit that large matrix representations are more computationally costly than simpler models.
The purpose of this work, however, is to establish how traffic can be represented and generated realistically at all, not how it can be done most efficiently.
Currently, realistic traffic modeling is an unsolved problem.
We observe, however, that generating application traffic for a simulation takes orders of magnitude less \caps{CPU}-time than running the simulation itself.
Therefore, traffic model complexity is considered a non-issue.

\item \textbf{Comprehensiveness of performance metrics used in evaluation.}
% TODO: address comprehensiveness of metrics in discussion.
We already briefly addressed this issue in the original version of the paper:
we have evaluated a variety of other performance metrics, but omit them here due to lack of space.
Moreover, the performance comparisons for the metrics presented here are indicative of other performance metrics.
We address this now in our discussion section in greater detail.

\item \textbf{Miscellaneous other questions and concerns.}
% TODO: make sure this is addressed adequately.
We have addressed the inter-relation between flow size, packet behavior, duration and start time in the section describing traffic generation (page 10, section 4.4).
The adequacy of the 24-hour trace is discussed in the discussion section: in short, 24 hours of data is not enough to draw general conclusions about all traffic, but is enough to demonstrate the potential effectiveness, or provide counter-examples to the effectiveness of modeling techniques; this is all we attempt to do here.
Further validation of methods on larger sets of data is required in the future.
We provide a concrete example of theoretical and numerical tools and their application to the \caps{GMM} in the use of nonnegative matrix factorization to understand and explain flow packet behaviors.
This technique and preliminary results from its application are shown in the discussion section.
The seven properties used in the \caps{GMM} are 
We note in our revised related work section that the work of \FHC~\cite{Hernandez06:dissertation} perfectly complements ours in that it provides the means to emulate the closed-loop behavior of applications using \caps{TCP}, but does not provide any models of application behavior;
our work provides the necessary models of applications behavior.
See Related Work, page 3, column 2, paragraph 2 for our discussion of this issue.

\end{enumerate}

We hope that the revised version addresses your concerns and questions adequately and that it meets your expectations for professional scholarly publications.

\closing{Yours Respectfully,}
\signature{Stefan~Karpinski}

\vfill
\bibliographystyle{plain}
\bibliography{references}

\end{letter}

\end{document}

### 2008-06-25 letter.tex
New letter version with quotes of reviewers.

@@
 
-\newenvironment{thebibliography}[1]{%
-\section{References}%
-\begin{enumerate}%
-}{%
-\end{enumerate}%
+\newcounter{topic}
+\newcommand{\topic}[1]{
+\addtocounter{topic}{1}
+\textbf{\arabic{topic}. #1}
 }
+
+\newenvironment{reviewer}[1]%
+{\begin{quote}\textbf{Reviewer #1:} \it}{\end{quote}}
+
+\newenvironment{thebibliography}[1]{\textbf{References:}\smaller\begin{enumerate}}{\end{enumerate}}
 
@@
 Thank you for your thoughtful and helpful reviews of out \caps{ACM}/Monet submission.
-From your responses and feedback from other peers, we believe that the first version of our paper did not adequately make clear the motivation, purpose or contribution of our work.
-Therefore, we have significantly revised the introductory and motivational material of our paper as well as presentation of the technical concepts regarding the linear representation of network traffic as vectors and matrices.
-The simulation results that support our claim that the general matrix model captures enough of the behavior of real traffic to be sufficiently realistic as a basis for generative models of network traffic remains the same except for the \model{mixed conditional} model (previously called \model{mixed regular}).
-There was an error in the code generating traffic for this model, which caused the simulation results to be incorrect.
-We have corrected this error.
-In the new version, we have omitted the discussion of measures of error and statistical tests, which already appeared in our previous work~\cite{Karpinski07:cbr-failure}; it was deemed too far from the central topic of the paper.
-Since the utility and significance of the general matrix model was not made clear in our previous version of the paper, we have added in the discussion section preliminary results from using the numerical linear technique of nonnegative matrix factorization on the \caps{GMM} matrices describing the packet behaviors of flows.
-We believe these results are very exciting and promising, and indicate the utility and potential of the representing network traffic linearly.
+Based on your responses and feedback from other colleagues, we believe that the first version of our paper did not adequately make clear the motivation, purpose or contribution of our work.
+Therefore, we have significantly revised the introductory and motivational material as well as the presentation of the technical concepts regarding the linear representation of network traffic as vectors and matrices.
+We hope that the new version more clearly conveys our overall motivation, contribution and technical approach.
+
+The simulation results supporting our claim that the general matrix model sufficiently realistic as a generative workload model remains the unchanged, with one exception:
+the \model{mixed conditional} model (previously called \model{mixed regular}).
+There was an error in the code generating traffic for this model, which caused the simulation results to be incorrect;
+we have corrected this error.
+
+In the new version, we have omitted the discussion of measures of error and statistical tests, which already appeared in our previous work~\cite{Karpinski07:cbr-failure}.
+It was deemed too far from the central topic of the paper; since it has already been published, we have omitted it and referenced the earlier work.
+
+Since the utility and significance of the general matrix model was not made clear in our previous version of the paper, we have added in the discussion section preliminary results from using nonnegative matrix factorization on the \caps{GMM} matrices describing the packet behaviors of flows.
+This recently developed numerical linear technique~\cite{Lee01} produces results that we believe are very exciting and promising.
+We hope that this example demonstrates the utility and potential of the representing network traffic linearly.
+
 In what follows, we indicate where in the paper revision, we address specific comments, questions and concerns raised by our collective reviewers.
 
-\begin{enumerate}
+\topic{Information destruction in uniform, marginal and conditional models (previously called regular), seen through the lens of linear representation.}
 
-\item \textbf{Information destruction in uniform, marginal and conditional (n\'e regular) models, seen through linear representation.}
-It was mentioned that it was unclear how the transformations allow to understand the effect and drawbacks of each simplified uniform, marginal or conditional model.
+\begin{reviewer}{1}
+It is clearly shown in section V that how common modeling concepts can be expressed using matrix transformation.
+However, it is not very clear how this transformation allow us to evaluate the effect of each model’s simplistics on the realism of generated workload as the authors claimed.
+\end{reviewer}
+
+% It was mentioned that it was unclear how the transformations allow to understand the effect and drawbacks of each simplified uniform, marginal or conditional model.
 In the revised paper, we have observed that the uniform, marginal and conditional transformations are all multiplications by specific low-rank matrices.
 Since low-rank matrices are by definition uninvertible, the information content of the original traffic matrix is irrevocably destroyed by these transformations.
-The revised paper discusses this issue in several places:
+The revised paper discusses this issue in two places:
 \begin{enumerate}
@@
 
-\item \textbf{Computational complexity of \caps{GMM} compared with simplified models.}
+\topic{Computational complexity of \caps{GMM} compared with simplified models.}
+
+\begin{reviewer}{1}
+Due to the vast size of the matrix and its sparseness, the requirement of processing and computational power might be quite demanding in order to smoothly express and generate traffic based on the GMM.
+There is no detail about how the matrix representation from the trace data is derived.
+Therefore, it is not clear how GMM can efficiently express the
+characteristics of total 2.1 million flows traced for 24 hours.
+\end{reviewer}
+
+\begin{reviewer}{2}
+[S]ince GMM is a more complicated model, I would assume that it is computationally more complicated to generate GMM-based traffic. There is no discussion on this issue in the paper.
+\end{reviewer}
+
 % TODO: address complexity in discussion.
@@
 We observe, however, that generating application traffic for a simulation takes orders of magnitude less \caps{CPU}-time than running the simulation itself.
-Therefore, traffic model complexity is considered a non-issue.
+Therefore, traffic model generation complexity is considered a non-issue.
 
-\item \textbf{Comprehensiveness of performance metrics used in evaluation.}
+\topic{Comprehensiveness of performance metrics used in evaluation.}
+
+\begin{reviewer}{1}
+The authors used the six performance metrics to prove how accurately GMM capture the characteristics of the real network.
+However, what if some other performance metrics, e.g., delay jitter, MAC-layer bandwidth share, etc., are concerned?
+Can we say these metrics also be accurately reflected with high confidence?
+\end{reviewer}
+
 % TODO: address comprehensiveness of metrics in discussion.
@@
 
-\item \textbf{Miscellaneous other questions and concerns.}
+\topic{Miscellaneous other questions and concerns.}
+
+% \begin{reviewer}{1}
+% \begin{enumerate}
+% \item What is a flow size exactly?
+Based on my understanding, it is the sum of all the bytes ﬂown between a pair of host.
+If so, flow size, packet size, and inter-packet interval are closely related to each other.
+Is there any inconsistency problem when considering them separately?
+% \item Does 24-hour trace data used in the paper really good enough? 
+% \item What theoretical and computational tools of modern linear algebra can be adopted and used for traffic generation specifically?
+One thing they mentioned as their current work is using matrix factorization and clustering technique to extract hidden structure from a large and diverse body of real traffic.
+But this haven’t verified yet and isn’t it too soon to claim this?
+% \item Wouldn‘t it be better if the authors come up with the criteria to determine whether the seven properties characterize the traced traffic adequately instead of proving the adequacy of GMM model by showing the six performance metrics closely resembled those of traced traffic?
+For example, by comparing distribution of packet size of GMM model and that of traced data.
+% \item Paxson and Floyd observed TCP traffic which protocol itself has closed-loop behavior. However, can we generalized this to the entire network as the authors mentioned?
+% \end{enumerate}
+% \end{reviewer}
+
+\begin{reviewer}{1}
+What is a flow size exactly?
+Based on my understanding, it is the sum of all the bytes ﬂown between a pair of host.
+If so, flow size, packet size, and inter-packet interval are closely related to each other.
+Is there any inconsistency problem when considering them separately?
+\end{reviewer}
+We have addressed the inter-relation between flow size, packet behavior, duration and start time in the section describing traffic generation (page 10, section 4.4).
+
+\begin{reviewer}{1}
+Does 24-hour trace data used in the paper really good enough?
+\end{reviewer}
 % TODO: make sure this is addressed adequately.
-We have addressed the inter-relation between flow size, packet behavior, duration and start time in the section describing traffic generation (page 10, section 4.4).
-The adequacy of the 24-hour trace is discussed in the discussion section: in short, 24 hours of data is not enough to draw general conclusions about all traffic, but is enough to demonstrate the potential effectiveness, or provide counter-examples to the effectiveness of modeling techniques; this is all we attempt to do here.
+The adequacy of the 24-hour trace is discussed in the discussion section: in short, 24 hours of data is not enough to draw general conclusions about all traffic, but is enough to demonstrate the potential effectiveness of \caps{GMM}, and provide counter-examples to the effectiveness of uniform, marginal and conditional modeling techniques;
+this is all we attempt to do here.
 Further validation of methods on larger sets of data is required in the future.
-We provide a concrete example of theoretical and numerical tools and their application to the \caps{GMM} in the use of nonnegative matrix factorization to understand and explain flow packet behaviors.
-This technique and preliminary results from its application are shown in the discussion section.
-The seven properties used in the \caps{GMM} are 
-We note in our revised related work section that the work of \FHC~\cite{Hernandez06:dissertation} perfectly complements ours in that it provides the means to emulate the closed-loop behavior of applications using \caps{TCP}, but does not provide any models of application behavior;
-our work provides the necessary models of applications behavior.
+
+\begin{reviewer}{1}
+What theoretical and computational tools of modern linear algebra can be adopted and used for traffic generation specifically?
+One thing they mentioned as their current work is using matrix factorization and clustering technique to extract hidden structure from a large and diverse body of real traffic.
+But this haven’t verified yet and isn’t it too soon to claim this?
+\end{reviewer}
+We provide a concrete example of a powerful technique that can be applied immediately to the \caps{GMM}: we use nonnegative matrix factorization to understand and explain flow packet behaviors.
+This technique and some exemplary results from its application are shown in the discussion section.
+It is also noted that all the traffic models used in this paper are written in only 19 lines of Matlab code, demonstrating the power of and simplicity of linear representation and the \caps{GMM} for traffic modeling explorations.
+
+\begin{reviewer}{1}
+Wouldn‘t it be better if the authors come up with the criteria to determine whether the seven properties characterize the traced traffic adequately instead of proving the adequacy of GMM model by showing the six performance metrics closely resembled those of traced traffic?
+For example, by comparing distribution of packet size of GMM model and that of traced data.
+\end{reviewer}
+The seven properties used in the \caps{GMM} are % TODO finish this...
+
+\begin{reviewer}{1}
+Paxson and Floyd observed TCP traffic which protocol itself has closed-loop behavior. However, can we generalized this to the entire network as the authors mentioned?
+\end{reviewer}
+We note in our revised related work section that the work of \FHC~\cite{Hernandez06:dissertation} perfectly complements ours in that it provides the means to emulate the closed-loop behavior of applications using \caps{TCP}, but does not provide any models of application behavior.
+Our work, on the other hand, provides the necessary models of applications behavior.
+Taken together, the \caps{TCP} flow generation technique of \FHC and our application-level workload modeling address both halves of the closed-loop network traffic modeling problem.
 See Related Work, page 3, column 2, paragraph 2 for our discussion of this issue.
 
-\end{enumerate}
+% The adequacy of the 24-hour trace is discussed in the discussion section: in short, 24 hours of data is not enough to draw general conclusions about all traffic, but is enough to demonstrate the potential effectiveness, or provide counter-examples to the effectiveness of modeling techniques; this is all we attempt to do here.
+% Further validation of methods on larger sets of data is required in the future.
+% We provide a concrete example of theoretical and numerical tools and their application to the \caps{GMM} in the use of nonnegative matrix factorization to understand and explain flow packet behaviors.
+% This technique and preliminary results from its application are shown in the discussion section.
+% The seven properties used in the \caps{GMM} are 
+% 
+% We note in our revised related work section that the work of \FHC~\cite{Hernandez06:dissertation} perfectly complements ours in that it provides the means to emulate the closed-loop behavior of applications using \caps{TCP}, but does not provide any models of application behavior;
+% our work provides the necessary models of applications behavior.
+% See Related Work, page 3, column 2, paragraph 2 for our discussion of this issue.
 

### 2008-06-26 paper.tex
Work.

@@
 
-\usepackage[labelfont=bf,small]{caption}
-\usepackage[font=small,labelfont=bf,position=top,nearskip=0em]{subfig}
-\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize,paralist,array,mathtools,units}
-\usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle: \thesubtitle}]{hyperref}
+\input{packages}
+\input{definitions}
 
@@
 }
-
-\input{definitions}
 

### 2008-06-26 paper.tex
Most of John's comments.

@@
 \begin{abstract}
-We propose a representation of wireless workload patterns as large, sparse matrices and provide a method for stochastically generating experimental workload from a given matrix.
+We propose a representation of wireless workload patterns as large, sparse matrices and provide a method for stochastically generating experimental workloads from a given matrix.
 The essential property of the algebraic representation is that the summation of vectors naturally yields a faithful description of the aggregate behavior of the corresponding flows.
 This deceptively simple property allows us to express many common concepts from traffic modeling succinctly in terms of a few linear transformations.
-The algebraic representation has many benefits: 1)~it makes the meaning of generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2)~it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3)~the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed; 4)~the vast body of algebraic theory and highly optimized numerical linear software may immediately be applied to traffic modeling.
+The algebraic representation has many benefits: 1)~it makes the meaning of generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2)~it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3)~the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed; 4)~the vast body of algebraic theory and highly optimized numerical  software may immediately be applied to traffic modeling.
 We use the paired differential simulation methodology introduced by the authors in previous work to experimentally demonstrate that the general matrix model accurately reproduces realistic network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
@@
 Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
-One of the major challenges of modeling whole-network traffic is the interdependence of packet, flow\flownote and node behaviors.
+One of the major challenges of modeling whole-network traffic is the interdependence of packet, flow,\flownote and node behaviors.
 Three types of simplifying assumptions are commonly applied in both analysis and generation:
@@
 \item \textbf{Marginal:} that traffic properties share a single, common \emph{marginal}\marginalnote distribution across the entire network, and individual values are drawn independently and unconditionally from this shared distribution.
-For example, assuming that a single distribution of packet sizes describes all flows in a network, as compared to equipping each flow with its own specific distribution of sizes.
+An example is assuming that a single distribution of packet sizes describes all flows in a network, as compared to equipping each flow with its own specific distribution of sizes.
 \item \textbf{Conditional:} that traffic behaves marginally within equivalence classes defined by certain conditions.
@@
 Although the factorization is computed without examining port numbers, basic behaviors nevertheless correspond to intuitive expectations for protocols.
-We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.
+We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer, and others.
 
@@
 In \Section{motivation} we discuss motivation and related work.
-The notion of linear representation and the general matrix model are presented in \Section{general-matrix-model}.
-In Sections~\ref{sec:common-properties}~and~\ref{sec:modeling-concepts} we explore how common network properties and modeling concepts can be expressed in the the \caps{GMM} framework.
+The notion of linear representation is presented in \Section{representations} and the general matrix model is presented in \Section{general-matrix-model}.
 Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
 The ramifications of these results are discussed in \Section{discussion}.
-Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies.
+Finally, in \Section{conclusions}, we conclude with a summary of this research and its impact on the future of wireless networking.
 
@@
 \Figure{trace-vs-uniform} illustrates through examples how drastically different network traffic appears under assumptions of uniformity at different levels of behavior.
+These examples are take from a wireless trace recorded at an \caps{IETF} meeting; details of the trace are given in \Section{trace-data}.
 Marginality assumptions are not as visually striking, but as we will demonstrate, they distort the performance characteristics of networks almost as much, and sometimes more, than uniform models.
@@
 In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
-Hern\'andez-Campos' dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and typical \caps{TCP} applications.
+Hern\'andez-Campos's dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and typical \caps{TCP} applications.
 The author shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
@@
 Like Harpoon, \caps{D-ITG}~\cite{Avallone04} exemplifies uniform and marginal modeling assumptions. However, unlike Harpoon, \caps{D-ITG} does not derive models from real traffic.
-Packet behavior can be modeled using standard statistical distributions (``constant, uniform, normal, cauchy, pareto, exponential, etc.'').
+Packet behavior can be modeled using standard statistical distributions (``constant, uniform, normal, Cauchy, Pareto, exponential, etc.'').
 Guidance is not provided, however, for choosing among these distributions, choosing parameter values, or selecting a mixture of distributions across a network.
@@
 provides an appropriate description of the aggregate behavior of the two flows together.
-Similarly, their average, \latin{i.e.} $\smallfrac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
+Similarly, their average, $\smallfrac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
 In contrast, a ``packets-payload-interval'' representation, expressing the \caps{CBR} parameters directly, does not have this property:
@@
 
-The \caps{PBD} representation of flows uses 3 dimensions to encode the behavior of each flow, while $\Phi$ uses $255+2n$ dimensions, plus the continuum of dimensions of two Banach spaces (\latin{i.e.} the two copies of $L^2[\R]$).
+The \caps{PBD} representation of flows uses three dimensions to encode the behavior of each flow, while $\Phi$ uses $255+2n$ dimensions, plus the continuum of dimensions of two Banach spaces (\latin{i.e.} the two copies of $L^2[\R]$).
 In this section, we define a representation, $\phi$, that splits the difference and uses a large but finite number of dimensions to represent each flow.
@@
 
+\subsection{Model Specification}
+
 The quantization functions specified in \Table{quantization} allow us to define the representation $\phi: \Traffic \to \R^N$: \begin{align}
@@
 \end{align}
-where $N=255+2n+d_t+d_b+d_z+d_v$ and vectors are
+where $N=255+2n+d_a+d_b+d_z+d_v$ and vectors are
 \begin{align*}
@@
 \vec{s} &= \e_k & k &= \Qf_\src(\src) \\
-\vec{d} &= \e_k & k &= \Qf_\dst(\dst) \\
+\vec{d} &= \e_k & k &= \Qf_\dst(\dst)
+\end{align*}
+\begin{align*}
 \vec{a} &= \e_k & k &= \Qf_\start(\start) \\
@@
 \>\bullet\> destinations: \>$\mat{D}$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$\\
-\>\bullet\> start times: \>$\mat{A}$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$\\
+\>\bullet\> start times: \>$\mat{A}$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_a}$\\
 \>\bullet\> sizes in bytes: \>$\mat{B}$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$\\
@@
 For lack of a better choice, we have simply chosen to use the same number of quantization bins as packet size for the unfixed dimension constants:
-$d_t = d_b = d_v = 1500$.
+$d_a = d_b = d_v = 1500$.
 Arguments can certainly be made for other choices, but this choice appears to work adequately in practice.
@@
 From the matrices $\mat{Z}$ and $\mat{V}$ we can compute a variety of statistics about packet sizes and inter-packet intervals.
-The flows vectors in the matrices may represent individual flows or composites of many flows.
+The row vectors in the matrices may represent individual flows or composites of many flows.
 % about packet sizes, total packet count, inter-packet intervals, and total duration for each representation vector, which may represent a single actual flow, or the composite behavior of many flows.
@@
 \begin{align}
-\vec{\Sigma}_\pkts  &= \mat{Z}\trans{\ones} = \mat{V}\trans{\ones} \in \R^f \eqn{packet-count}\\
+\vec{\Sigma}_\pkts  &= \mat{Z}\ones = \mat{V}\ones \in \R^f \eqn{packet-count}\\
 \vec{\Sigma}_\bytes &= \mat{Z}\trans{\vec{\sigma}} \in \R^f \eqn{total-bytes} \\
@@
   \vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts
-  \phantom{= \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones}}
-  \mathllap{= \mat{Z}\trans{\vec{\sigma}} \eldiv \mat{Z}\trans{\ones}} \\
+  \phantom{= \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\ones}
+  \mathllap{= \mat{Z}\trans{\vec{\sigma}} \eldiv \mat{Z}\ones} \\
 \vec{\mean{v}} &=
@@
   \mathllap{\vec{\Sigma_{\ivals}} \eldiv \vec{\Sigma_{\pkts}}}
-  = \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones}.
+  = \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\ones.
 \end{align}
@@
 Chose an index value $k\in\set{1,\dots,d_\prop}$, randomly, with weight given by the coefficients of $\vec{f}_\prop$.
-Choose $u \in [0,1]$ uniformly randomly.
+Choose $u \in [0,1]$ uniformly at random.
 Then $x = k-u \in [0,d]$, with probability of lying in the interval $[i-1,i]$ proportional to the coefficient of $\e_i$ in $\vec{f}_\prop$.
@@
 
-The above procedure allows us to sample random values for each property in accordance with the behavioral description implied by $\vec{f}$.
+The procedure above allows us to sample random values for each property in accordance with the behavioral description implied by $\vec{f}$.
 This allows us to stochastically generate a sequence of packets in accordance with a row from an arbitrary \caps{GMM} matrix instance using the following procedure:
@@
 \item Next, randomly sample the number of bytes.
-\item The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\mean{z}$ and $\mean{v}$.
+\item The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals});
+call them $\mean{z}$ and $\mean{v}$.
 From these, together with the flow size in bytes, we estimate the expected duration of the flow:
@@
 However, for any vector that is a sum of vectors representing valid flows---\latin{i.e.} ones having distinct source and destination nodes and feasible combinations of start time, duration, and packet behaviors---the procedure will terminate.
-\item Because the representation $\phi$ models the seven properties independent of each other, it is impossible to do better than to sample the properties independently, modulo the requirements of the procedure.
+\item Because the representation $\phi$ models the seven properties independently of each other, it is impossible to do better than to sample the properties independently, modulo the requirements of the procedure.
 In the case of a row which is simply a representation of an actual flow, this makes no difference, except for the generation of packets (which are modeled independently).
-In the case of composite rows, however, this implies that flow properties are modeled independent of each other.
+In the case of composite rows, however, this implies that flow properties are modeled independently of each other.
 \end{itemize}
-Despite these remarks, we will show in our experimental evaluation, that this procedure generates \emph{sufficiently realistic} wireless workloads~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+Despite these remarks, we will show in our experimental evaluation that this procedure generates \emph{sufficiently realistic} wireless workloads~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
 
@@
 This is a highly significant result because modeling flow size using a network-wide marginal distribution is precisely what {\FHC}, for example, attempt.
-This result informs us that this approach is doomed to failure: flow size cannot be assigned without considering its relation to other flow properties: source and destination nodes, and packet behavior at the very least.
+This result informs us that this approach is doomed to failure: flow size cannot be assigned without considering its relation to other flow properties; source and destination nodes, and packet behavior at the very least.
 
@@
 We present here preliminary results from applying nonnegative matrix factorization (\caps{NMF}) to the packet behaviors of real flows.
-Recall from \Section{general-matrix-model}, that the matrices $\mat{Z}$ and $\mat{V}$ represent the packet size and inter-packet interval histograms of a collection of flows, with each row corresponding to a flow, and each column corresponding to a packet size or a range of inter-packet intervals.
+Recall from \Section{general-matrix-model} that the matrices $\mat{Z}$ and $\mat{V}$ represent the packet size and inter-packet interval histograms of a collection of flows, with each row corresponding to a flow, and each column corresponding to a packet size or a range of inter-packet intervals.
 We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}] \in \R^{f \times (d_z+d_v)}$.

### 2008-06-26 paper.tex
Added sentence about significance of NMF.

@@
 We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}] \in \R^{f \times (d_z+d_v)}$.
-\caps{NMF} attempts to approximate $\mat{P}$ as the product of three nonnegative matrices:
+\caps{NMF} attempts to approximate $\mat{P}$ as the product of nonnegative matrices:
 \begin{align}
@@
 % This is precisely the behavior we expect from interacting with a remote terminal via keystrokes and screen updates.
+The true significance of this result lies in the fact that \caps{NMF} succeeds in classifying flows by network behavior completely automatically, not only without using port numbers, but also without \emph{any} information about the number or kind of plausible interpretations.´
+The algorithm not only agrees with our intuition about network behavior, but exceeds it by discovering real but unknown behaviors awaiting interpretation.
 

### 2008-06-26 figures/box_plots.tex
Removed mixed conditional; regular => conditional.

@@
 \subfloat[\textbf{Application:} Average End-to-End Delay]%
-{\includegraphics[width=4.45in]{Application/delay.pdf}}\hfill
+{\includegraphics[width=4.5in]{Application/delay.pdf}}\hfill
 \subfloat[\textbf{Application:} Packet Delivery Ratio]%
-{\includegraphics[width=4.45in]{Application/delivery_ratio.pdf}}
+{\includegraphics[width=4.5in]{Application/delivery_ratio.pdf}}
 %
 \subfloat[\textbf{Application:} Average Received Throughput]%
-{\includegraphics[width=4.45in]{Application/throughput.pdf}}\hfill
+{\includegraphics[width=4.5in]{Application/throughput.pdf}}\hfill
 \subfloat[\textbf{Network:} AODV Control Overhead]%
-{\includegraphics[width=4.45in]{Network/control_packets.pdf}}
+{\includegraphics[width=4.5in]{Network/control_packets.pdf}}
 %
 \subfloat[\textbf{Network:} Packets Dropped in Routing Queue]%
-{\includegraphics[width=4.45in]{Network/dropped.pdf}}\hfill
+{\includegraphics[width=4.5in]{Network/dropped.pdf}}\hfill
 \subfloat[\textbf{Link:} 802.11 Control Overhead]%
-{\includegraphics[width=4.45in]{Link/overhead.pdf}}%
+{\includegraphics[width=4.5in]{Link/overhead.pdf}}%
 %

### 2008-06-26 tables/traffic_models.tex
Removed mixed conditional; regular => conditional.

@@
 \\\hline
-dest conditional &
+dest. conditional &
 $\mat{F}' = \rStoch(\trans{\mat{D}}\mat{D})\,\mat{F}$ &
@@
 \\\hline
-mixed conditional &
-$\mat{F}' = \nicefrac{1}{2}\,(
-              \rStoch(\trans{\mat{S}}\mat{S}) +
-              \rStoch(\trans{\mat{D}}\mat{D})
-            )\,\mat{F}$ &
-flow behaviors are aggregated on a mixed source/destination basis
-\\\hline
+% mixed conditional &
+% $\mat{F}' = \nicefrac{1}{2}\,(
+%               \rStoch(\trans{\mat{S}}\mat{S}) +
+%               \rStoch(\trans{\mat{D}}\mat{D})
+%             )\,\mat{F}$ &
+% flow behaviors are aggregated on a mixed source/destination basis
+% \\\hline
 \end{tabular}

### 2008-06-26 paper.tex
Removing done TODOs, work on results.

@@
 We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer, and others.
-
-% TODO: update roadmap paragraph.
 
@@
 It also allows us to easily define intermediate choices.
-If we wish make the behavior of each flow a weighted average of the composite behaviors of its source and destination nodes, we can express this easily:
+If we wish, for example, to make the behavior of each flow a weighted average of the composite behaviors of its source and destination nodes, we can express this easily:
 \begin{align}
@@
 
-% TODO: remove mixed conditional from plots.
 \input{figures/box_plots}
@@
 neither is perfect, but both are better in general than any marginal or uniform model.
-The node-conditional models tend to err in the same direction as the \model{full marginal} model, but in a few cases where the aspects of marginality lead to errors in different directions, the outcome follows one of the aspects more than others.
+The \model{source/dest.
+conditional} models tend to err in the same direction as the \model{full marginal} model, but in a few cases where the aspects of marginality lead to errors in different directions, the outcome follows one of the aspects more than others.
 In the case of received throughput, for example, the \model{packet marginal} tendency to overestimate error overshadows the tendency of the other marginal aspects to underestimate.
-
-% TODO: rerun the mixed model simulations.
+Overall, we see that the conditional models are more realistic than other models, but this is effect occurs only as the number of nodes---\latin{i.e.} the number of conditioning classes---approaches the full rank of the original traffic matrix.
+In our discussion, we present and demonstrate a far more compelling method of reducing and simplifying the representation of the behavior of large collections of flows.
 

### 2008-06-26 letter.tex
Removing done TODOs, work on results.
Addressed all reviewer issues.

@@
 \item Introduction, page 2, column 2, paragraph 2;
-% TODO: add another paragraph about this?
 \item Discussion, page 15, column 2, paragraph 1.
@@
 
-% TODO: address comprehensiveness of metrics in discussion.
 We already briefly addressed this issue in the original version of the paper:
@@
 \topic{Miscellaneous other questions and concerns.}
-
-% \begin{reviewer}{1}
-% \begin{enumerate}
-% \item What is a flow size exactly?
-Based on my understanding, it is the sum of all the bytes ﬂown between a pair of host.
-If so, flow size, packet size, and inter-packet interval are closely related to each other.
-Is there any inconsistency problem when considering them separately?
-% \item Does 24-hour trace data used in the paper really good enough? 
-% \item What theoretical and computational tools of modern linear algebra can be adopted and used for traffic generation specifically?
-One thing they mentioned as their current work is using matrix factorization and clustering technique to extract hidden structure from a large and diverse body of real traffic.
-But this haven’t verified yet and isn’t it too soon to claim this?
-% \item Wouldn‘t it be better if the authors come up with the criteria to determine whether the seven properties characterize the traced traffic adequately instead of proving the adequacy of GMM model by showing the six performance metrics closely resembled those of traced traffic?
-For example, by comparing distribution of packet size of GMM model and that of traced data.
-% \item Paxson and Floyd observed TCP traffic which protocol itself has closed-loop behavior. However, can we generalized this to the entire network as the authors mentioned?
-% \end{enumerate}
-% \end{reviewer}
 
@@
 \end{reviewer}
-% TODO: make sure this is addressed adequately.
 The adequacy of the 24-hour trace is discussed in the discussion section: in short, 24 hours of data is not enough to draw general conclusions about all traffic, but is enough to demonstrate the potential effectiveness of \caps{GMM}, and provide counter-examples to the effectiveness of uniform, marginal and conditional modeling techniques;
@@
 \end{reviewer}
-The seven properties used in the \caps{GMM} are % TODO finish this...
+The seven properties used in the \caps{GMM} are not meant to be comprehensive, but rather representative.
+In the section in which we introduce the metrics, we state that we have collected and analyzed other performance metrics in our, experiments, but these performance metrics are representative of the overall results.
+At a higher level, comparing empirical distributions of models to the originals is meaningless:
+if the model is based on the trace, and includes packet size distribution then the two will necessarily match;
+running experiments is not necessary to show that.
+The real question is whether this matters or not:
+does it affect experimental results?
+We refer the reviewer to our earlier work for a more in-depth discussion of the issue~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
 
@@
 \closing{Yours Respectfully,}
-\signature{Stefan~Karpinski}
 

### 2008-06-26 paper.tex
Addressed all reviewer issues.

@@
 
-%The story of network analysis and modeling is one of variety. Assumptions of ``sameness'' have fallen repeatedly. ...
-%The matrix model gives new insight into
-
-Thus far, we have addressed two out of three of the originally posed questions regarding traffic modeling.
+Although we have shown it to be sufficiently realistic for driving wireless simulations, the general matrix model is not intended to be a final model used by networking researchers.
+Nor is it intended to be the ultimate perfect model of network behavior.
+Rather it serves as both a research tool and a stepping stone to better understanding of network traffic.
+Because it captures so much detail about traffic sample---%
+without making assumptions of uniformity or marginality of flow behavior---%
+it serves as a basis for deriving more advanced models that preserve more of the original behavior of networks than any uniform or marginal model ever can.
+The first important role it plays is in allowing us to succinctly mathematically express common models.
+All of the traffic models included in this paper are implemented in only 19 lines of Matlab code, applying simple matrix transformations to the \caps{GMM} representation of traffic.
+Through the lens of linear algebra, we see why uniform, marginal and even conditional models fail to adequately capture the gestalt behavior of networks:
+these simplifying assumptions force us to ignore most of the information present in a trace.
+In what follows, we address limitations of the the results presented in this paper, but conclude with an exciting and promising new technique for traffic analysis that amply demonstrates the potential of linear representation of network traffic.
+
+\subsection{Generality of Results}
+
+% The point of our analysis is not to show that no synthetic models can be accurate, but rather, that the commonly used models, and obvious variations thereof, are not accurate.
+
+A significant limitation on the generality of our results is that they are based on a single data set from \caps{IETF}60---albeit a large and varied one.
+It is possible that traffic in this trace happens to produce network performance that is unusually dissimilar to standard traffic behavior.
+This data set, however, represents a highly heterogeneous collection of network usage behaviors, from slow and steady off-peak usage, to extremely heavy peak usage:
+over 800 users, 33 thousand flows, and 1 million packets in a single 10-minute trace segment.
+Despite the broad variety of behaviors, the results are consistent:
+in all types of usage scenarios, uniform and marginal models, and to a lesser degree conditional ones, systematically skew important performance measurements at all levels of the network.
+While the precise results for other data sets might differ, it is very unlikely that these traffic models will happen to accurately reproduce realistic performance in other experiments.
+This paper provides strong evidence that better traffic models are needed for performance evaluations.
+Thus, while it is necessarily limited, our evaluation points us strongly in the direction away from uniform and marginal models.
+
+Another potential concern about generality is the relatively limited set of performance metrics evaluated.
+As mentioned in \Section{performance-metrics}, we have evaluated a broad variety of performance metrics---28 in all.
+The metrics evaluated here were selected both because they are among the most commonly used in evaluations, and because they are representative of the results for other metrics.
+In general, metrics at the same level of the protocol stack tend to behave in related ways:
+they may be positively or negatively correlated, but if one metric is severely distorted, it is unlikely that others are left intact.
+The direction of the distortion of metrics is highly sensitive to the experimental setup;
+the presence of error, however, is not.
+In other words, models that are sufficiently realistic in one experimental setup tend to be sufficiently realistic in other experiments;
+models that misrepresent performance metrics in certain cases tend to do so in other cases as well.
+The direction of the distortion may change, but the fact that they are inaccurate does not.
+See previous work using the same technique for evaluating realism for further examples with different experimental conditions and sets of performance metrics considered~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+
+\subsection{Computational Complexity and Overhead}
+
+Another potential concern about the approach of linear representation is the additional computational overhead it may induce in experimental settings.
+Large matrix representations of traffic models are certainly more computationally costly than simpler models.
+Currently, however, realistic local-area traffic modeling is an unsolved problem.
+The purpose of this work is to establish how traffic can be represented and generated realistically at all, not how it can be done most efficiently.
+We observe, however, that generating application traffic for a simulation takes orders of magnitude less \caps{CPU}-time than running the simulation itself.
+Therefore, traffic model generation complexity is considered a non-issue.
+A related issue is the significant simulation time incurred when experiments use realistic volumes of traffic.
+Indeed, this is a problem:
+we have run our simulations on a 64-node Itanium cluster, and the experiments still take weeks to complete.
+However, with the current state of the art in simulators and \caps{CPU} power, this problem is unavoidable.
+Including model preparation times, running simulations with uniform or marginal models takes negligibly less time than running simulations using realistic models like the \caps{GMM}.
+One of the possible effects of this research, however, is the potential ability to classify and catalog an realistic set of ``exemplary'' traffic scenarios for various applications.
+Such a catalog is far from existing, but its existence would allow researchers to run far fewer simulations to achieve infinitely more reliable results.
+
+\subsection{Nonnegative Factorization of Packet Behavior}
+
+We have addressed two out of three of the originally posed questions regarding traffic modeling; in this section, we address the third.
 First, we have proposed a general approach for representing whole-network traffic without making assumptions of uniformity or marginality.

### 2008-06-26 paper.tex
New NMF intro paragraph.

@@
 
-We have addressed two out of three of the originally posed questions regarding traffic modeling; in this section, we address the third.
-First, we have proposed a general approach for representing whole-network traffic without making assumptions of uniformity or marginality.
+% We have addressed two out of three of the originally posed questions regarding traffic modeling; in this section, we address the third.
+% First, we have proposed a general approach for representing whole-network traffic without making assumptions of uniformity or marginality.
 To this end, we introduced the concept of a linear representation of traffic.
@@
 These transformations destroy almost all of the information in the original traffic matrix.
+% Conditional models are equivalent to left-multiplication by somewhat less degenerate matrices, but they achieve realism only by approaching the \caps{GMM} itself.
+
+\input{figures/basic_behaviors}
+
+It remains to show how we can simplify and reduce the representation of network behaviors without making unwarranted and manifestly false assumptions of uniformity of marginality of network behaviors.
+In \Section{modeling-concepts}, we observed that application of uniformity and marginality assumptions to traffic instances is equivalent, in the \caps{GMM}, to right and left multiplication, respectively, by matrices of rank one.
+These transformations destroy almost all of the information in the original traffic matrix.
 Conditional models are equivalent to left-multiplication by somewhat less degenerate matrices, but they achieve realism only by approaching the \caps{GMM} itself.
-
-\input{figures/basic_behaviors}
-
-It remains to show how we can simplify and reduce the representation of network behaviors without making unwarranted and manifestly false assumptions of uniformity of marginality of network behaviors.
+In this section we describe matrix factorization, which provides an approximate low-rank representation of the original traffic.
+This technique not only allows us to model and analyze traffic without destroying the information content, but also leads us to a deep analytical understanding of network behaviors that matches our intuitive understanding of typical modes of network behavior.
+
 We present here preliminary results from applying nonnegative matrix factorization (\caps{NMF}) to the packet behaviors of real flows.

### 2008-06-26 letter.tex
Letter now explains why mixed cond. model was omitted.

@@
 the \model{mixed conditional} model (previously called \model{mixed regular}).
-There was an error in the code generating traffic for this model, which caused the simulation results to be incorrect;
-we have corrected this error.
+There was an error in the code generating traffic for this model, which caused the simulation results to be incorrect.
+Upon correcting this error, we found that the mixed model performed identically to the source and destination based models.
+These results were therefore deemed uninteresting and omitted.
 

### 2008-06-29 figures/basic_behaviors.tex
Fixed pie charts.
Changes for auto publication.

@@
 \begin{sidewaysfigure*}
-{\includegraphics[height=5.95in]{basic_behaviors_1.pdf}}\hfill%
-{\includegraphics[height=5.95in]{basic_behaviors_2.pdf}}
+{\includegraphics[height=5.95in]{basic_behaviors_1}}\hfill%
+{\includegraphics[height=5.95in]{basic_behaviors_2}}
 \caption{%
@@
 To the right of each behavior is a pie chart showing the breakdown of associated packets by protocol, as determined by well-known port numbers.
+Pie slices are only labeled if they constitute at least 3\% of associated packets.
 }

### 2008-06-29 figures/box_plots.tex
Changes for auto publication.

@@
 \subfloat[\textbf{Application:} Average End-to-End Delay]%
-{\includegraphics[width=4.5in]{Application/delay.pdf}}\hfill
+{\includegraphics[width=4.5in]{Application/delay}}\hfill
 \subfloat[\textbf{Application:} Packet Delivery Ratio]%
-{\includegraphics[width=4.5in]{Application/delivery_ratio.pdf}}
+{\includegraphics[width=4.5in]{Application/delivery_ratio}}
 %
 \subfloat[\textbf{Application:} Average Received Throughput]%
-{\includegraphics[width=4.5in]{Application/throughput.pdf}}\hfill
+{\includegraphics[width=4.5in]{Application/throughput}}\hfill
 \subfloat[\textbf{Network:} AODV Control Overhead]%
-{\includegraphics[width=4.5in]{Network/control_packets.pdf}}
+{\includegraphics[width=4.5in]{Network/control_packets}}
 %
 \subfloat[\textbf{Network:} Packets Dropped in Routing Queue]%
-{\includegraphics[width=4.5in]{Network/dropped.pdf}}\hfill
+{\includegraphics[width=4.5in]{Network/dropped}}\hfill
 \subfloat[\textbf{Link:} 802.11 Control Overhead]%
-{\includegraphics[width=4.5in]{Link/overhead.pdf}}%
+{\includegraphics[width=4.5in]{Link/overhead}}%
 %

### 2008-06-29 paper.tex
Final tweaking: now down to 18 pages.

@@
 \item Do these assumptions distort the metrics we are trying to analyze, replicate, and predict?
-\item Can we represent ``unreduced'' network behavior such that analysis and generation remain tractable?
+\item Can we represent ``unreduced'' network behavior so that analysis and generation remain tractable?
 \item Are there better ways to simplify and reduce the representation of network behavior?
@@
 
-\newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $\pkts$ does not divide $\mathsym{bytes}$ evenly.
-The simplest is to ignore it; we leave more complex schemes to the reader.}
+\newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $p$ does not divide $b$ evenly. The simplest is to ignore it; we leave more complex schemes to the reader.}
 
@@
 
-\subsubsection{Uniformity}
+\subsubsection{Uniform Modeling}
 
@@
 
-\subsubsection{Marginality}
+\subsubsection{Marginal Modeling}
 
@@
 
-\subsubsection{Conditionality}
+\subsubsection{Conditional Modeling}
 
@@
 
-The remainder of this section proceeds as follows.
+%The remainder of this section proceeds as follows.
 In \Section{traffic-models} we present the collection of traffic models we will evaluate.
@@
 % 2,085,563 flows, 58,377,994 packets of application data, and 51,988,245,537 bytes of application data.
-
-\input{figures/nodes_flows}
 
@@
 
+\input{figures/nodes_flows}
+
 Before using the traces, it is necessary to extract application-level behavior from the trace header data.
@@
 We observe, however, that generating application traffic for a simulation takes orders of magnitude less \caps{CPU}-time than running the simulation itself.
-Therefore, traffic model generation complexity is considered a non-issue.
+Therefore, traffic generation complexity is considered a non-issue.
+
 A related issue is the significant simulation time incurred when experiments use realistic volumes of traffic.
@@
 we have run our simulations on a 64-node Itanium cluster, and the experiments still take weeks to complete.
-However, with the current state of the art in simulators and \caps{CPU} power, this problem is unavoidable.
+Given the current state of the art in simulators and \caps{CPU} power, however, this problem is unavoidable.
 Including model preparation times, running simulations with uniform or marginal models takes negligibly less time than running simulations using realistic models like the \caps{GMM}.
-One of the possible effects of this research, however, is the potential ability to classify and catalog an realistic set of ``exemplary'' traffic scenarios for various applications.
-Such a catalog is far from existing, but its existence would allow researchers to run far fewer simulations to achieve infinitely more reliable results.
+One of the possible effects of this research, however, may be the ability to classify and catalog an realistic set of ``exemplary'' traffic scenarios for various applications.
+Such a catalog is far from existing, but its existence would allow researchers to run far fewer simulations to achieve incomparably more general and reliable results.
 
@@
 
-It remains to show how we can simplify and reduce the representation of network behaviors without making unwarranted and manifestly false assumptions of uniformity of marginality of network behaviors.
+It remains to show how we can simplify and reduce the representation of network behaviors without making unwarranted and manifestly false assumptions of uniformity or marginality of network behaviors.
 In \Section{modeling-concepts}, we observed that application of uniformity and marginality assumptions to traffic instances is equivalent, in the \caps{GMM}, to right and left multiplication, respectively, by matrices of rank one.
@@
 We present here preliminary results from applying nonnegative matrix factorization (\caps{NMF}) to the packet behaviors of real flows.
-Recall from \Section{general-matrix-model} that the matrices $\mat{Z}$ and $\mat{V}$ represent the packet size and inter-packet interval histograms of a collection of flows, with each row corresponding to a flow, and each column corresponding to a packet size or a range of inter-packet intervals.
+Recall from \Section{general-matrix-model} that the matrices $\mat{Z}$ and $\mat{V}$ represent the packet size and inter-packet interval histograms of a collection of flows.
 We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}] \in \R^{f \times (d_z+d_v)}$.
@@
 This is what we would expect from downloading large files; and indeed, the breakdown of packets by port number bears out this interpretation---almost all of the traffic is \caps{HTTP} server traffic, with small slivers of \caps{IMAP} and \caps{SSH} server traffic.
-The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all \caps{SSH} traffic, as indicated by the associate pie chart.
+The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all interactive \caps{SSH} traffic, as indicated by the associate pie chart.
 The other behaviors are listed in \Table{basic-behaviors} with interpretations.

### 2008-06-29 letter.tex
Generate tarball; finalized sections in references.

@@
 \newenvironment{thebibliography}[1]{\textbf{References:}\smaller\begin{enumerate}}{\end{enumerate}}
-
-\name{%
-Stefan~Karpinski\\
-Elizabeth~M.~Belding\\
-Kevin~C.~Almeroth\\
-John~R.~Gilbert%
-}
 
@@
 Since low-rank matrices are by definition uninvertible, the information content of the original traffic matrix is irrevocably destroyed by these transformations.
-The revised paper discusses this issue in two places:
-\begin{enumerate}
-\item Introduction, page 2, column 2, paragraph 2;
-\item Discussion, page 15, column 2, paragraph 1.
-\end{enumerate}
-In the discussion section, moreover, we propose matrix factorization, rather matrix multiplication as a less destructive and thus more faithful method of reducing the complexity, size and dimension of traffic representations.
+The revised paper discusses this issue in Section 1 (p. 2, col. 2, par. 2), Section 4.4, and Section 7 (p. 2, col. 2, par. 2) sections.
+In the discussion section (Section 7), moreover, we propose matrix factorization, rather matrix multiplication as a less destructive and thus more faithful method of reducing the complexity, size and dimension of traffic representations.
 
@@
 
-% TODO: address complexity in discussion.
-This issue is now addressed in our discussion section.
+This issue is now addressed in our discussion: Section 7.2.
 We admit that large matrix representations are more computationally costly than simpler models.
@@
 Moreover, the performance comparisons for the metrics presented here are indicative of other performance metrics.
-We address this now in our discussion section in greater detail.
+We address this now in our discussion section in greater detail: see Section 7.1.
 
@@
 Further validation of methods on larger sets of data is required in the future.
+We now address this issue in the discussion: see Section 7.1.
 
@@
 We provide a concrete example of a powerful technique that can be applied immediately to the \caps{GMM}: we use nonnegative matrix factorization to understand and explain flow packet behaviors.
-This technique and some exemplary results from its application are shown in the discussion section.
+This technique and some exemplary results from its application are shown in the discussion section (Section 7.3).
 It is also noted that all the traffic models used in this paper are written in only 19 lines of Matlab code, demonstrating the power of and simplicity of linear representation and the \caps{GMM} for traffic modeling explorations.
@@
 The seven properties used in the \caps{GMM} are not meant to be comprehensive, but rather representative.
-In the section in which we introduce the metrics, we state that we have collected and analyzed other performance metrics in our, experiments, but these performance metrics are representative of the overall results.
-At a higher level, comparing empirical distributions of models to the originals is meaningless:
+In Section 5.4, we state that we have collected and analyzed other performance metrics in our, experiments, but these performance metrics are representative of the overall results.
+We now further address this issue in Section 7.1 as well.
+To address the reviewer's specific point, at a higher level, comparing empirical distributions of models to the originals is meaningless:
 if the model is based on the trace, and includes packet size distribution then the two will necessarily match;
@@
 Taken together, the \caps{TCP} flow generation technique of \FHC and our application-level workload modeling address both halves of the closed-loop network traffic modeling problem.
-See Related Work, page 3, column 2, paragraph 2 for our discussion of this issue.
+See Section 2 for our discussion of this issue.
 
@@
 
-\closing{Yours Respectfully,}
+Yours Respectfully,
+
+\begin{quote}
+Stefan~Karpinski\\
+Elizabeth~M.~Belding\\
+Kevin~C.~Almeroth\\
+John~R.~Gilbert
+\end{quote}
 

### 2008-06-30 figures/box_plots.tex
Final changes for the submission.

@@
 \begin{sidewaysfigure*}
-\subfloat[\textbf{Application:} Average End-to-End Delay]%
-{\includegraphics[width=4.5in]{Application/delay}}\hfill
-\subfloat[\textbf{Application:} Packet Delivery Ratio]%
-{\includegraphics[width=4.5in]{Application/delivery_ratio}}
-%
-\subfloat[\textbf{Application:} Average Received Throughput]%
-{\includegraphics[width=4.5in]{Application/throughput}}\hfill
-\subfloat[\textbf{Network:} AODV Control Overhead]%
-{\includegraphics[width=4.5in]{Network/control_packets}}
-%
-\subfloat[\textbf{Network:} Packets Dropped in Routing Queue]%
-{\includegraphics[width=4.5in]{Network/dropped}}\hfill
-\subfloat[\textbf{Link:} 802.11 Control Overhead]%
-{\includegraphics[width=4.5in]{Link/overhead}}%
-%
+\subfloat[\textbf{Application:} Average End-to-End Delay]
+{\includegraphics[width=4.6in]{Application/delay}}
+\subfloat[\textbf{Application:} Packet Delivery Ratio]
+{\includegraphics[width=4.6in]{Application/delivery_ratio}}
+
+\subfloat[\textbf{Application:} Average Received Throughput]
+{\includegraphics[width=4.6in]{Application/throughput}}
+\subfloat[\textbf{Network:} AODV Control Overhead]
+{\includegraphics[width=4.6in]{Network/control_packets}}
+
+\subfloat[\textbf{Network:} Packets Dropped in Routing Queue]
+{\includegraphics[width=4.6in]{Network/dropped}}
+\subfloat[\textbf{Link:} 802.11 Control Overhead]
+{\includegraphics[width=4.6in]{Link/overhead}}
 \caption{%

### 2008-11-04 paper.tex
Corrections to the proof text.

@@
 
-Having defined our representation of network behavior and presented the procedure for stochastically generating workload, we turn to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.
+Having defined a specific matrix representation of network behavior,
+% and a procedure for stochastically generating workload from a given matrix,
+we turn now to deriving useful expressions for commonly used network properties in terms of these algebraic building blocks.
 
@@
 % We begin by expressing the vectors for total bytes, total packets and total duration for a matrix of flow representations, with size and interval matrices $\mat{Z}$ and $\mat{V}$.
-The computations require knowing the values that each size of interval matrix column represents.
+The computations require knowing the values that each matrix column represents, whether size or interval.
 % The computed duration is approximate because information is irretrievably lost in the quantization process.
@@
 The simplifying assumptions made by most traffic models can be classified into three categories.
-We use the terms \textit{uniformity}, \textit{marginality}, and \textit{conditionality} to describe these classes of assumptions.
+We use the terms \textit{uniform}, \textit{marginal}, and \textit{conditional} to describe these classes of assumptions.
 They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether they make stipulations about uniformity within each flow, across all flows, or within groups of flows.
@@
 When we say that our marginal models are ``perfect,'' we mean that they are nonparametric and calculated directly from trace behavior.
-These are the marginal models that most accurately describe the trace traffic, because they are directly derived from it and make no parametric simplifications.
+These are the marginal models that most accurately describe the trace traffic, because they are directly derived from it and make no parametric approximations.
 This provides an upper bound on the quality of marginal models in general, since parametric marginal models are derived, in turn, as simplifications of these nonparametric marginal behaviors.
@@
 models that misrepresent performance metrics in certain cases tend to do so in other cases as well.
-The direction of the distortion may change, but the fact that they are inaccurate does not.
-See previous work using the same technique for evaluating realism for further examples with different experimental conditions and sets of performance metrics considered~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+The direction of the distortion may change, but the presence or absence of accuracy does not.
+See previous work using the same technique for evaluating realism for further examples with different experimental conditions and more performance metrics considered~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
 
@@
 Including model preparation times, running simulations with uniform or marginal models takes negligibly less time than running simulations using realistic models like the \caps{GMM}.
-One of the possible effects of this research, however, may be the ability to classify and catalog an realistic set of ``exemplary'' traffic scenarios for various applications.
+One of the possible effects of this research, however, may be the ability to classify and catalog a realistic set of ``exemplary'' traffic scenarios for various applications.
 Such a catalog is far from existing, but its existence would allow researchers to run far fewer simulations to achieve incomparably more general and reliable results.
@@
 Conditional models are equivalent to left-multiplication by somewhat less degenerate matrices, but they achieve realism only by approaching the \caps{GMM} itself.
-In this section we describe matrix factorization, which provides an approximate low-rank representation of the original traffic.
-This technique not only allows us to model and analyze traffic without destroying the information content, but also leads us to a deep analytical understanding of network behaviors that matches our intuitive understanding of typical modes of network behavior.
+In this section we describe a matrix factorization technique that can provide an approximate low-rank representation of the original traffic without assuming uniformity or marginality.
+This technique not only allows us to simplify and analyze traffic without destroying the information content, but also leads us to a deep analytical understanding of network behaviors that matches our intuitive understanding of typical modes of network behavior.
 
@@
 \end{align}
-where $\mat{W} \in \R^{f \times b}$ and $\mat{H} \in \R^{b \times (d_z+d_v)}$ are both row-stochastic, while
-$\mat{\Sigma} \in \R^{b \times b}$ is diagonal with descending diagonal entries.
-$b\in\N$ is the rank of this factorization, which must be determined somehow as well.
+$\mat{W} \in \R^{f \times b}$ is column-stochastic, $\mat{\Sigma} \in \R^{b \times b}$ is diagonal with descending diagonal entries, and $\mat{H} \in \R^{b \times (d_z+d_v)}$ is row-stochastic.
+The inner dimension, $b\in\N$, must be determined as well.
 What is the interpretation of such a factorization?
@@
 Twelve basic behaviors suffice to explain all the traffic in a ``toy''  sample of 4138 flows.
-The pairs of packet size and inter-packet interval distributions are almost immediately recognizable as resembling intuitive modes of network behavior.
-The first basic behavior, for example has all packets near the \caps{MTU}, while intervals are smoothly distributed around a fraction of a millisecond.
+The pairs of packet size and inter-packet interval distributions are readily recognizable as resembling intuitive modes of network behavior.
+The first basic behavior, for example, has all packet sizes near the \caps{MTU}, while intervals are smoothly distributed around a fraction of a millisecond.
 This is what we would expect from downloading large files; and indeed, the breakdown of packets by port number bears out this interpretation---almost all of the traffic is \caps{HTTP} server traffic, with small slivers of \caps{IMAP} and \caps{SSH} server traffic.
@@
 The essential property of the algebraic representation is that standard vector operations compute natural and useful descriptions of aggregate behavior over the collection of flows represented.
-Our experimental validation demonstrates clearly that the general matrix model accurately reproduces the performance characteristics of of real traffic.
+Our experimental validation demonstrates that the general matrix model accurately reproduces performance characteristics of of real traffic.
 
@@
 Such simplifications correspond to right matrix multiplication in \caps{GMM}.
-Because of the simplicity and clarity that the algebraic structure brings to the subject, we can see clearly now that both common types of simplifying assumptions are na\"ive and simplistic.
+Because of the simplicity and clarity that the algebraic structure brings to the subject, we can see that both common types of simplifying assumptions are na\"ive and simplistic.
 Our experimental results bolster this conclusion: both uniform and marginal models significantly misrepresent many important network performance metrics.

### 2008-11-04 tables/traffic_models.tex
Corrections to the proof text.

@@
  \mat{D}' = (\nicefrac{1}{n})\,\mat{D}\ones$ &
-all nodes equally likely as source or destination
+all nodes are equally likely as source or destination
 \\\hline
@@
 \textit{all uniform transformations} &
-all start times, sources and destinations equally likely
+all start times, sources and destinations are equally likely
 \\\hline\hline
@@
 $\mat{B}' = (\nicefrac{1}{f})\,\ones\mat{B}$ &
-flow size (bytes) is chosen independently of other properties
+flow size (bytes) is chosen from marginal distribution of sizes
 \\\hline
@@
  \mat{D}' = (\nicefrac{1}{f})\,\ones\mat{D}$ &
-nodes behave distinctly, but behavior is independent of flows
+source and dest. chosen marginally, independent of other properties
 \\\hline
@@
  \mat{V}' = (\nicefrac{1}{f})\,\ones\mat{V}$ &
-packet behavior is variable bit-rate, aggregated across all flows
+packet behavior is variable bit-rate, marginal across all flows
 \\\hline
@@
 \end{tabular}
-\caption{Matrix-based traffic models evaluated by paired differential simulation.}
+\caption{Matrix-based traffic models used in paired differential simulation experiments.}
 \label{tab:traffic-models}
